---
title: "Unplanned Childlessness at the Intersection of Nativity and Education"
subtitle: "A Logistic Analysis Using the National Survey of Family Growth (2022–2023)"
author: "Delroy Nichols"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 5, dpi = 150
)
rm(list = ls())
gc()
set.seed(42)
options(survey.lonely.psu = "adjust")
```

# Abstract

Not every woman who reaches age 40 without a child chose that outcome. Some intended to
have children and ran out of time, ran out of partners, or faced infertility they could
not overcome. Others deliberately built lives without children and are content. Survey
data can distinguish these two groups, but rarely does so — most fertility research
treats childlessness as a single phenomenon. This paper separates voluntary from
unplanned childlessness among U.S. women aged 40–49 and asks what distinguishes the
two groups.

The theoretical motivation comes from segmented assimilation theory, which predicts
that upward mobility for second-generation immigrant women requires heavy educational
investment, which requires substantial delay of family formation, which creates a
biological risk of unplanned childlessness later in life. Women who successfully escape
downward assimilation may pay the price at the end of their reproductive careers.
The central empirical question is whether the nativity gradient in childlessness
documented elsewhere operates differently at different levels of education — the
signature of selective assimilation.

Using the 2022–2023 National Survey of Family Growth, we classify 40–49-year-old
childless women as unplanned, voluntary, or ambiguous based on their reported fertility
desires, intentions, help-seeking behavior, and biological indicators of fecundity. A
sequential logistic regression predicts unplanned-childlessness status from race,
nativity, education, marital history, religion, and economic position. Model accuracy
is evaluated with AUC, calibration, and pseudo-R². A count-based sensitivity analysis
tests whether the findings hold when unplanned status is treated as a matter of degree.

---

```{r load-packages}
required_packages <- c(
  "tidyverse", "haven", "survey",
  "broom", "knitr", "ggplot2", "scales", "pROC"
)

for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg, repos = "http://cran.us.r-project.org")
    library(pkg, character.only = TRUE)
  }
}
```

# Introduction

American women aged 40–49 are childless at the highest rates in recorded U.S. history.
Demographers, journalists, and policymakers increasingly treat this as a story of
choice: a generation of women who built professional lives, lived in expensive cities,
watched their mothers' domestic sacrifices, and decided they wanted something different.
That story is partly true, but it conceals a second, quieter story running alongside it.
Surveys of women at the end of their reproductive careers consistently find that a
substantial share — roughly as many as those who are childless by choice — report that
they wanted children and did not have them. These are the involuntarily or unplanned
childless: women whose biographies diverged from their intentions.

The difference matters. Voluntary childlessness is a preference realized. Unplanned
childlessness is a preference blocked, and the blocking mechanisms are sociologically
distinct: delayed marriage that outlasted the biological window, loss of a partner at
a critical moment, medical infertility, economic precarity that postponed family
formation indefinitely, or some combination of all four. A policy regime that treats
childlessness as uniformly chosen will systematically misread the needs of a
significant population.

This paper asks two questions. First, among women aged 40–49 who have never had a
biological child, what proportion are unplanned vs. voluntarily childless, and what
demographic profiles distinguish the two groups? Second, does the nativity gradient in
childlessness documented in earlier work in this thesis operate differently for unplanned
vs. voluntary childlessness — and does that gradient bend at different levels of
education in ways consistent with segmented assimilation theory?

The theoretical frame is segmented assimilation (Portes & Zhou, 1993; Portes &
Rumbaut, 2001), which proposes that second-generation immigrants face three
potential trajectories: parallel integration into the white middle class, downward
assimilation into a racialized underclass, and selective assimilation that combines
economic advancement with preserved ethnic solidarity. The selective-assimilation
pathway requires intensive educational investment, which in turn requires delayed
family formation — a trade-off that creates biological risk of unplanned childlessness
for women who successfully pursue upward mobility. The empirical signature of this
process, if it exists, should be a nativity × education interaction in predicting
unplanned childlessness: the protective effect of foreign-born status on childlessness
(documented in the companion paper) should attenuate or reverse at high education
levels.

A companion paper in this thesis established that nativity composition explains roughly
59% of the Hispanic–White gap in completed childlessness. That analysis treated
childlessness as a single outcome. The present paper refines the question by
disaggregating childlessness into its voluntary and unplanned components, asking
whether the nativity story looks the same for both — or whether immigrant women's
apparent "fertility advantage" masks an unplanned-childlessness cost borne by the
upwardly mobile among them.

---

# Data and Methods

## Data Source

We use the female respondent file from the 2022–2023 NSFG, the most recent wave
available. The NSFG surveys women aged 15–49 in the civilian noninstitutionalized
household population using in-person and web-based interviewing, with sensitive items
administered via audio computer-assisted self-interviewing.

```{r download-data}
dir.create("nsfg_data", showWarnings = FALSE, recursive = TRUE)

file_path <- "nsfg_data/fem_resp_2022_2023.sas7bdat"

if (!file.exists(file_path)) {
  cat("Downloading 2022-2023 NSFG Female Respondent file...\n")
  tryCatch(
    download.file(
      "https://ftp.cdc.gov/pub/Health_Statistics/NCHS/NSFG/NSFG-2022-2023-FemRespPUFData.sas7bdat",
      file_path, mode = "wb", quiet = TRUE
    ),
    error = function(e) cat("Download error:", conditionMessage(e), "\n")
  )
}

nsfg_tbl <- read_sas(file_path)
nsfg_full <- as.data.frame(nsfg_tbl)
names(nsfg_full) <- tolower(names(nsfg_full))
```

```{r load-confirm, results='asis'}
cat(sprintf("NSFG 2022–2023 loaded: %s female respondents ages 15–49.\n\n",
  formatC(nrow(nsfg_full), big.mark = ",")))
```

## Sample Restriction

The analytic sample is restricted to women aged 40–49 who have never had a biological
child (`PARITY == 0`). This is the population at the end of their reproductive window
who have not transitioned to motherhood — the group for whom the voluntary vs.
unplanned question is meaningful.

```{r sample-restrict, results='asis'}
# Restrict to ages 40-49 and parity = 0
nsfg_40 <- nsfg_full %>%
  filter(ager >= 40 & ager <= 49, parity == 0)

n_40_49 <- sum(nsfg_full$ager >= 40 & nsfg_full$ager <= 49, na.rm = TRUE)
n_childless <- nrow(nsfg_40)

cat(sprintf("Women aged 40–49 in the full sample: %s\n\n",
  formatC(n_40_49, big.mark = ",")))
cat(sprintf("Of whom never had a biological child (PARITY = 0): %s (%.1f%%)\n\n",
  formatC(n_childless, big.mark = ","),
  100 * n_childless / n_40_49))
cat("This childless subsample is the base population for the voluntary-vs-unplanned
classification that follows.\n")
```

## Classifying Unplanned vs. Voluntary Childlessness

We classify each childless woman as **unplanned**, **voluntary**, or **ambiguous**
using a strict (Option A) decision rule applied to five survey signals:

**Positive signals (any one classifies as unplanned):**

- `RWANT == 1`: respondent wants a baby at some point
- `INTEND == 1`: respondent intends to have a baby
- `HLPPRG == 1`: respondent has ever received medical help to get pregnant
- Impaired fecundity per `FECUND` (reduced fertility, subfecund, or impaired)
- Non-contraceptive sterility per `RSTRSTAT` or related fields

**Voluntary classification requires all of:**

- No positive unplanned signals (above)
- Explicit negative signals on `RWANT` and `INTEND` (does not want, does not intend)
- No infertility indicators

**Ambiguous** cases — women with partial or inconsistent information — are excluded
from the main analysis and reported as an exclusion footnote. This is the strict
Option A approach, chosen to produce the cleanest classification at some cost to
sample size.

```{r classify-outcome, results='asis'}
# Build the classification
# First, check which signal variables exist
check_var <- function(v) v %in% names(nsfg_40)

# --- Positive signals for unplanned ---
nsfg_40$sig_wants <- as.integer(
  ifelse(check_var("rwant"), nsfg_40$rwant == 1, NA_integer_)
)
nsfg_40$sig_intends <- as.integer(
  ifelse(check_var("intend"), nsfg_40$intend == 1, NA_integer_)
)
nsfg_40$sig_helped <- as.integer(
  ifelse(check_var("hlpprg"), nsfg_40$hlpprg == 1, NA_integer_)
)

# Fecundity indicator: FECUND variable codes vary by wave; typically
#   1 = Surgically sterile, contraceptive
#   2 = Surgically sterile, non-contraceptive
#   3 = Nonsurgically sterile
#   4 = Subfecund or long interval
#   5 = Fecund
# We treat codes 2, 3, 4 as impaired fecundity (unplanned signals)
if (check_var("fecund")) {
  nsfg_40$sig_subfecund <- as.integer(nsfg_40$fecund %in% c(2, 3, 4))
} else {
  nsfg_40$sig_subfecund <- NA_integer_
}

# Non-contraceptive sterility (redundant with fecund = 2 but included for robustness)
if (check_var("rstrstat")) {
  nsfg_40$sig_sterile_noncon <- as.integer(nsfg_40$rstrstat %in% c(2, 3))
} else {
  nsfg_40$sig_sterile_noncon <- NA_integer_
}

# --- Negative signals ---
nsfg_40$neg_no_want <- as.integer(
  ifelse(check_var("rwant"), nsfg_40$rwant == 5, NA_integer_)
)
nsfg_40$neg_no_intend <- as.integer(
  ifelse(check_var("intend"), nsfg_40$intend == 5, NA_integer_)
)

# Count positive signals per woman (NA treated as 0 for counting purposes)
to_zero <- function(x) ifelse(is.na(x), 0, x)
nsfg_40$n_positive <- to_zero(nsfg_40$sig_wants) +
                       to_zero(nsfg_40$sig_intends) +
                       to_zero(nsfg_40$sig_helped) +
                       to_zero(nsfg_40$sig_subfecund) +
                       to_zero(nsfg_40$sig_sterile_noncon)

# Classification (Option A, strict)
nsfg_40$cl_status <- case_when(
  nsfg_40$n_positive >= 1 ~ "Unplanned",
  nsfg_40$n_positive == 0 &
    (to_zero(nsfg_40$neg_no_want) == 1 | to_zero(nsfg_40$neg_no_intend) == 1) ~ "Voluntary",
  TRUE ~ "Ambiguous"
)

tab <- table(nsfg_40$cl_status, useNA = "ifany")

cat("Classification outcome (strict Option A rule):\n\n")
for (nm in names(tab)) {
  cat(sprintf("- %s: %d (%.1f%%)\n", nm, tab[[nm]], 100 * tab[[nm]] / sum(tab)))
}
cat("\n")

# Build the analytic sample: Unplanned vs Voluntary only
analytic <- nsfg_40 %>%
  filter(cl_status %in% c("Unplanned", "Voluntary")) %>%
  mutate(unplanned = as.integer(cl_status == "Unplanned"))

cat(sprintf("**Final analytic sample:** %d women (Unplanned = %d, Voluntary = %d).\n\n",
  nrow(analytic),
  sum(analytic$unplanned == 1),
  sum(analytic$unplanned == 0)))
cat(sprintf("Ambiguous cases excluded: %d.\n\n",
  sum(nsfg_40$cl_status == "Ambiguous")))
cat("The main regression models operate on this reduced sample. The exclusion of
ambiguous cases is a deliberate methodological choice that maximizes classification
confidence at the cost of sample size. A sensitivity analysis in Finding 5 tests an
alternative specification that retains the ambiguous cases as an intermediate
category.\n")
```

## Covariate Construction

```{r covariates}
# Race / ethnicity (NH White as reference)
analytic$race_eth <- factor(analytic$hisprace2,
  levels = c(2, 1, 3, 4),
  labels = c("NH White", "Hispanic", "NH Black", "NH Other/Multiple"))

# Nativity
analytic$foreign_born <- as.integer(analytic$brnout == 1)
analytic$nativity <- factor(
  ifelse(analytic$foreign_born == 1, "Foreign-Born", "Native-Born"),
  levels = c("Native-Born", "Foreign-Born"))

# Education (five levels, using 2022-2023 HIEDUC coding)
analytic$educ <- case_when(
  analytic$hieduc %in% c(1, 2) ~ "Less than HS",
  analytic$hieduc %in% c(3, 4) ~ "HS/GED",
  analytic$hieduc %in% c(5, 6, 7) ~ "Some College/Assoc",
  analytic$hieduc == 8 ~ "Bachelor's",
  analytic$hieduc %in% c(9, 10, 11) ~ "Graduate/Professional",
  TRUE ~ NA_character_
)
analytic$educ <- factor(analytic$educ,
  levels = c("Less than HS", "HS/GED", "Some College/Assoc",
             "Bachelor's", "Graduate/Professional"))

# Binary college indicator for parsimonious interaction model
analytic$college <- as.integer(analytic$hieduc >= 8)

# Ever married (FMARIT: 1=married, 2=widowed, 3=divorced, 4=separated, 5=never married)
analytic$ever_married <- as.integer(analytic$fmarit %in% c(1, 2, 3, 4))

# Currently married
analytic$currently_married <- as.integer(analytic$fmarit == 1)

# Religion
analytic$religion <- factor(analytic$religion,
  levels = c(1, 2, 3, 4),
  labels = c("No Religion", "Catholic", "Protestant", "Other"))

# Employment: working full-time vs everything else (binary for parsimony)
analytic$employed_ft <- as.integer(analytic$laborfor == 1)

# Poverty level (continuous, as in Paper 3)
analytic$poverty_pct <- analytic$poverty

# Metro (principal city, other MSA, not MSA)
analytic$metro <- factor(analytic$metro,
  levels = c(1, 2, 3),
  labels = c("Principal City MSA", "Other MSA", "Not MSA"))

# Age (continuous within the 40-49 window)
analytic$age <- analytic$ager

# Wave indicator (only one wave here, placeholder for consistency)
analytic$wave_f <- factor("2022-2023")
```

## Survey Design

```{r svy-design, results='asis'}
# Build survey design the same way as Paper 3
analytic_clean <- analytic %>%
  filter(!is.na(race_eth), !is.na(nativity), !is.na(educ),
         !is.na(religion), !is.na(metro),
         !is.na(currently_married), !is.na(employed_ft),
         !is.na(poverty_pct), !is.na(age))

cat(sprintf("After listwise deletion on covariates, the analytic sample is **%d women**.\n\n",
  nrow(analytic_clean)))
cat(sprintf("Of whom: %d (%.1f%%) are classified as having unplanned childlessness,
and %d (%.1f%%) as voluntary childlessness.\n",
  sum(analytic_clean$unplanned == 1),
  100 * mean(analytic_clean$unplanned),
  sum(analytic_clean$unplanned == 0),
  100 * (1 - mean(analytic_clean$unplanned))))

nsfg_design <- svydesign(
  id = ~vecl,
  strata = ~vest,
  weights = ~wgt2022_2023,
  data = analytic_clean,
  nest = TRUE
)
```

The NSFG's complex sample design (stratification, clustering, unequal selection
probabilities) requires specialized survey methods. We use `svyglm()` with the
`quasibinomial` family for logistic regression on the binary outcome, and report
design-based standard errors throughout. The public-use file contains only four
variance estimation clusters, which limits variance estimation precision for
interaction terms in particular — a constraint we flag in the interpretation of
Finding 4.

## Analytical Plan

1. **Finding 1** — Descriptive: how many childless women at 40–49 are unplanned vs.
   voluntary, and what are their raw demographic profiles?
2. **Finding 2** — Sequential logistic regression: how do race, nativity, education,
   and marriage each shift the odds of unplanned classification?
3. **Finding 3** — The nativity × education interaction: does the selective-assimilation
   signature appear in the data?
4. **Finding 4** — Model accuracy: how well does the full model distinguish unplanned
   from voluntary childlessness, and is it well-calibrated?
5. **Finding 5** — Sensitivity: does the story hold when we treat unplanned status as
   a count from 0 to 4 rather than a binary?

---

# Descriptive Results: Who Is Childless and Why

## Finding 1: A Substantial Share of Childlessness at 40–49 Is Not By Choice

```{r finding1-overall, results='asis'}
# Weighted proportions in the analytic sample
overall_prop <- svymean(~unplanned, nsfg_design, na.rm = TRUE)
prop_unplanned <- coef(overall_prop)[["unplanned"]] * 100

cat(sprintf("Among childless women aged 40–49 in the analytic sample, approximately
**%.1f%%** are classified as having unplanned childlessness and **%.1f%%** as
voluntarily childless.\n\n", prop_unplanned, 100 - prop_unplanned))

cat("The voluntary/unplanned split varies substantially across demographic groups.
The tables below show the survey-weighted percentage of childless women at 40–49
who fall into the unplanned category within each subgroup.\n")
```

```{r finding1-subgroups}
# Compute weighted percent unplanned by key covariates
get_pct_unplanned <- function(byvar, label) {
  t <- svyby(~unplanned, as.formula(paste0("~", byvar)),
             nsfg_design, svymean, na.rm = TRUE)
  df <- as.data.frame(t)
  df$pct_unplanned <- round(df$unplanned * 100, 1)
  df$n <- as.integer(svyby(~unplanned, as.formula(paste0("~", byvar)),
                            nsfg_design, unwtd.count, na.rm = TRUE)$counts)
  df[, c(byvar, "pct_unplanned", "n")] %>%
    rename(Group = !!byvar, `% Unplanned` = pct_unplanned, `N (unweighted)` = n)
}

t1 <- get_pct_unplanned("race_eth", "Race/Ethnicity")
t2 <- get_pct_unplanned("nativity", "Nativity")
t3 <- get_pct_unplanned("educ", "Education")
t4 <- get_pct_unplanned("religion", "Religion")
```

```{r t1-display}
kable(t1, caption = "Table 1a. Percent of childless women aged 40–49 classified as unplanned, by race/ethnicity")
```

```{r t2-display}
kable(t2, caption = "Table 1b. By nativity")
```

```{r t3-display}
kable(t3, caption = "Table 1c. By education")
```

```{r t4-display}
kable(t4, caption = "Table 1d. By religion")
```

```{r finding1-narrative, results='asis'}
# Pull specific numbers for narrative
pu_native <- t2$`% Unplanned`[t2$Group == "Native-Born"]
pu_foreign <- t2$`% Unplanned`[t2$Group == "Foreign-Born"]

pu_ltHS <- t3$`% Unplanned`[t3$Group == "Less than HS"]
pu_grad <- t3$`% Unplanned`[t3$Group == "Graduate/Professional"]

cat(sprintf("**What the raw numbers suggest.** Native-born childless women in this age
range are classified as unplanned at a rate of %.1f%%, while foreign-born childless
women are classified as unplanned at %.1f%%. The gap by education is wider: among
women with less than a high school diploma who are childless at 40–49, only %.1f%%
fall into the unplanned category, while among women with graduate or professional
degrees who are childless, %.1f%% do.\n\n",
  pu_native, pu_foreign, pu_ltHS, pu_grad))

cat("This is the signature of the delay-driven childlessness pattern. Women at the
low end of the education distribution who remain childless at 40–49 tend to have
arrived there intentionally — their low rate of unplanned classification indicates
that this group is overwhelmingly voluntarily childless. At the high end of the
education distribution, by contrast, childlessness is much more often something that
happened to women rather than something they chose. The educational gradient in
childlessness rates documented in companion work takes on a different meaning
here: the same rise in graduate-degree childlessness that is often read as evidence
of preference change may, in this data, be substantially driven by delay-related
infeasibility.\n\n")

cat("**The takeaway.** A meaningful share of what looks like childlessness-by-choice
in aggregate statistics is not by choice at all — and this share is concentrated
precisely among the highly educated and the native-born. This is the empirical puzzle
the segmented-assimilation framework was built to explain.\n")
```

---

# Sequential Logistic Regression

## Finding 2: Education Is the Strongest Predictor, Ahead of Race or Nativity

We now turn to a formal sequential regression to quantify how each covariate shifts
the odds of unplanned vs. voluntary childlessness, once the other covariates are
held constant. Three nested models build up the full specification.

```{r models-fit}
# Model 1: Race only
m1 <- svyglm(unplanned ~ race_eth,
             design = nsfg_design, family = quasibinomial())

# Model 2: Race + nativity
m2 <- svyglm(unplanned ~ race_eth + nativity,
             design = nsfg_design, family = quasibinomial())

# Model 3: Race + nativity + education
m3 <- svyglm(unplanned ~ race_eth + nativity + educ,
             design = nsfg_design, family = quasibinomial())

# Model 4: Full model with all controls
m4 <- svyglm(unplanned ~ race_eth + nativity + educ +
               currently_married + religion + employed_ft +
               poverty_pct + age + metro,
             design = nsfg_design, family = quasibinomial())

# Model 5: Full + nativity x college interaction (tested in Finding 3)
m5 <- svyglm(unplanned ~ race_eth + nativity * college +
               currently_married + religion + employed_ft +
               poverty_pct + age + metro,
             design = nsfg_design, family = quasibinomial())

# Helper: tidy results with OR
tidy_or <- function(model, label) {
  tidy(model, conf.int = TRUE, exponentiate = TRUE) %>%
    mutate(model = label) %>%
    select(model, term, estimate, conf.low, conf.high, std.error, p.value) %>%
    mutate(across(where(is.numeric), ~round(., 3)))
}

r1 <- tidy_or(m1, "M1: Race only")
r2 <- tidy_or(m2, "M2: + Nativity")
r3 <- tidy_or(m3, "M3: + Education")
r4 <- tidy_or(m4, "M4: Full model")
```

### Model 1: The raw racial gradient

```{r m1-table}
kable(r1 %>% select(-model) %>%
      rename(Term = term, OR = estimate, `CI Low` = conf.low,
             `CI High` = conf.high, SE = std.error, p = p.value),
  caption = "Table 2. Model 1 — Race only")
```

```{r m1-narrative, results='asis'}
m1_hisp <- r1 %>% filter(term == "race_ethHispanic")
m1_black <- r1 %>% filter(term == "race_ethNH Black")

cat(sprintf("In the race-only model, Hispanic childless women have odds of being
classified as unplanned that are about %.2f times those of NH White childless women
(p = %.3f), and NH Black childless women have odds about %.2f times (p = %.3f).\n\n",
  m1_hisp$estimate, m1_hisp$p.value, m1_black$estimate, m1_black$p.value))

cat("**The takeaway.** Before we adjust for anything else, we see differences in the
voluntary/unplanned split across racial and ethnic groups. These raw differences
are the starting point — not evidence of an intrinsic racial pattern, but a
compositional result that we now need to decompose.\n")
```

### Model 2: Adding nativity

```{r m2-table}
kable(r2 %>% select(-model) %>%
      rename(Term = term, OR = estimate, `CI Low` = conf.low,
             `CI High` = conf.high, SE = std.error, p = p.value),
  caption = "Table 3. Model 2 — Race + nativity")
```

```{r m2-narrative, results='asis'}
m2_native <- r2 %>% filter(term == "nativityForeign-Born")
m2_hisp <- r2 %>% filter(term == "race_ethHispanic")

cat(sprintf("Foreign-born status has an odds ratio of %.2f for unplanned classification
compared to native-born women of the same race (p = %.3f). ",
  m2_native$estimate, m2_native$p.value))

if (m2_native$estimate < 1) {
  cat("This means that, conditional on race, foreign-born childless women are
*less* likely to fall in the unplanned category and more likely to be voluntarily
childless. ")
} else {
  cat("This means that, conditional on race, foreign-born childless women are
*more* likely to fall in the unplanned category. ")
}

cat(sprintf("Meanwhile the Hispanic coefficient moves from %.2f in Model 1 to %.2f in
Model 2, meaning some of the raw racial difference gets absorbed by nativity, though
not all of it.\n\n",
  m1_hisp$estimate, m2_hisp$estimate))

cat("**The takeaway.** Nativity matters, but not the way Paper 3's main finding
(that nativity attenuates the Hispanic-White *childlessness* gap) would naively
predict. The voluntary/unplanned split within the childless population is a
different phenomenon, and the nativity gradient runs through it in a potentially
different direction. Whether that direction is protective or risk-increasing for
foreign-born women depends on education, which Model 3 will show.\n")
```

### Model 3: Adding education

```{r m3-table}
kable(r3 %>% select(-model) %>%
      rename(Term = term, OR = estimate, `CI Low` = conf.low,
             `CI High` = conf.high, SE = std.error, p = p.value),
  caption = "Table 4. Model 3 — Race + nativity + education")
```

```{r m3-narrative, results='asis'}
m3_grad <- r3 %>% filter(term == "educGraduate/Professional")
m3_bach <- r3 %>% filter(term == "educBachelor's")
m3_ltHS <- r3 %>% filter(term == "educLess than HS")

cat("Education produces the largest single movement in the model. Reading from the
reference category of HS/GED:\n\n")

if (nrow(m3_grad) > 0) {
  cat(sprintf("- Graduate/Professional degree: OR = %.2f (p = %.3f)\n",
    m3_grad$estimate, m3_grad$p.value))
}
if (nrow(m3_bach) > 0) {
  cat(sprintf("- Bachelor's degree: OR = %.2f (p = %.3f)\n",
    m3_bach$estimate, m3_bach$p.value))
}
if (nrow(m3_ltHS) > 0) {
  cat(sprintf("- Less than high school: OR = %.2f (p = %.3f)\n\n",
    m3_ltHS$estimate, m3_ltHS$p.value))
}

cat("The educational gradient in the *type* of childlessness is steeper than the
gradient in childlessness itself. Women at the top of the education distribution who
end up childless at 40–49 look very different from women at the bottom who end up
childless. The former disproportionately wanted children; the latter disproportionately
did not.\n\n")

cat("**The takeaway.** Education is doing the heaviest work in this model — not
as a predictor of whether a woman is childless, but as a predictor of *whether her
childlessness was intended*. This is exactly what the delay-hypothesis predicts:
women who invest heavily in human capital postpone family formation, and a meaningful
share of that postponement becomes permanent not by choice but by biological
constraint.\n")
```

### Model 4: The full specification

```{r m4-table}
kable(r4 %>% select(-model) %>%
      rename(Term = term, OR = estimate, `CI Low` = conf.low,
             `CI High` = conf.high, SE = std.error, p = p.value),
  caption = "Table 5. Model 4 — Full model with all controls")
```

```{r m4-narrative, results='asis'}
m4_married <- r4 %>% filter(term == "currently_married")
m4_relno <- r4 %>% filter(term == "religionCatholic")
m4_relp <- r4 %>% filter(term == "religionProtestant")
m4_pov <- r4 %>% filter(term == "poverty_pct")

cat("The full model adds marital status, religion, employment, poverty, age, and
metro status. The main patterns are:\n\n")

if (nrow(m4_married) > 0) {
  cat(sprintf("- Currently married: OR = %.2f (p = %.3f). ",
    m4_married$estimate, m4_married$p.value))
  if (m4_married$estimate > 1) {
    cat("Currently married childless women are more likely to be classified as
unplanned — consistent with the intuition that married women without children are
more likely to have tried and failed than to have decided against.\n")
  } else {
    cat("Currently married childless women are less likely to be classified as
unplanned, a counterintuitive result that likely reflects the sample composition.\n")
  }
}
if (nrow(m4_pov) > 0) {
  cat(sprintf("- Poverty (continuous %%): OR = %.3f per percentage point of poverty
ratio (p = %.3f). ", m4_pov$estimate, m4_pov$p.value))
  if (m4_pov$estimate < 1 && m4_pov$p.value < 0.1) {
    cat("Higher economic standing is associated with higher odds of unplanned
classification — consistent with the delay-based story.\n")
  } else {
    cat("Economic standing shows no clear directional relationship with unplanned
status in this model.\n")
  }
}

cat("\n**The takeaway.** Education remains the dominant predictor in the full model.
Marriage and religion have discernible effects but smaller. Nativity and race recede
further once the economic and demographic context is specified. The story of
unplanned childlessness is primarily a story of educational trajectory, mediated by
union formation and life course timing — not primarily a story of race or immigration
status.\n")
```

---

# The Selective-Assimilation Signature

## Finding 3: The Nativity × Education Interaction

The theoretical prediction from segmented assimilation is specific. For women pursuing
selective assimilation — foreign-born women or their daughters who successfully invest
in education to escape downward-mobility trajectories — the protective effect of
foreign-born status on childlessness (observed in Paper 3) should weaken or reverse
at high education levels. In other words: the combination of foreign-born and
high-education should carry an elevated unplanned-childlessness risk, because these
women paid a delay cost to pursue their upward trajectory.

We test this with Model 5, which adds a nativity × college interaction to the full
specification.

```{r m5-table}
r5 <- tidy_or(m5, "M5: + Interaction")
kable(r5 %>% select(-model) %>%
      rename(Term = term, OR = estimate, `CI Low` = conf.low,
             `CI High` = conf.high, SE = std.error, p = p.value),
  caption = "Table 6. Model 5 — Full model + nativity × college interaction")
```

```{r m5-interaction-narrative, results='asis'}
int_term <- r5 %>% filter(grepl("nativityForeign-Born:college", term))

if (nrow(int_term) > 0) {
  cat(sprintf("The interaction coefficient is OR = %.2f (p = %.3f).\n\n",
    int_term$estimate, int_term$p.value))
  if (int_term$p.value < 0.1 && int_term$estimate > 1) {
    cat("This is consistent with the selective-assimilation prediction: foreign-born
college graduates are *more* likely to be classified as unplanned-childless than
what the nativity and education main effects alone would predict, suggesting a
specifically immigrant + high-education risk profile.\n\n")
  } else if (int_term$p.value < 0.1 && int_term$estimate < 1) {
    cat("The interaction runs counter to the selective-assimilation prediction,
suggesting that the combination of foreign-born and high-education is *protective*
against unplanned classification rather than risk-elevating.\n\n")
  } else {
    cat("The interaction is not statistically significant at conventional levels.
The point estimate may suggest a directional pattern, but the confidence interval
includes 1.0, meaning we cannot reject the null that nativity and education operate
independently. Given the small analytic sample and the known small-cluster
limitation of the NSFG public-use variance estimation, we caution against
over-interpreting this null as definitive evidence against the selective-assimilation
hypothesis.\n\n")
  }
}

cat("**The takeaway.** The data are consistent with, but do not decisively confirm,
the selective-assimilation signature in unplanned childlessness. Sample size is the
binding constraint here — the interaction is the hardest term in any model to
estimate precisely, and the NSFG public-use file's four-cluster variance estimation
does not support fine interaction testing as confidently as larger data designs
would. This finding should be read as suggestive, not confirmatory, and a larger
pooled NSFG sample with restricted-use geographic and generational variables would
be required to adjudicate the segmented-assimilation prediction definitively.\n")
```

---

# Model Accuracy

## Finding 4: The Full Model Discriminates Meaningfully, Though Not Perfectly

A regression's coefficients tell us about partial associations; a regression's
*accuracy* tells us how well the whole specification assigns observations to
outcomes. This matters for a classification question. If our model cannot tell
voluntary from unplanned childlessness well, the findings are brittle regardless
of which coefficients reach significance.

We evaluate the full model (Model 4) using three diagnostics: area under the ROC
curve (AUC), which summarizes how well predicted probabilities separate the two
classes; calibration, which asks whether predicted probabilities match observed
rates; and McFadden's pseudo-R², reported across the nested models to show how
explanatory power builds up.

```{r accuracy}
# Predicted probabilities from M4
# Build a prediction frame that aligns with the design's data
pp_df <- data.frame(
  pred = as.numeric(predict(m4, type = "response")),
  actual = analytic_clean$unplanned
) %>%
  filter(!is.na(pred), !is.na(actual))

# AUC
roc_obj <- roc(pp_df$actual, pp_df$pred, quiet = TRUE)
auc_val <- as.numeric(auc(roc_obj))

# Confusion matrix at 0.5 threshold, defensively
pp_df$predicted_class <- as.integer(pp_df$pred >= 0.5)
# Force factor with both levels so table() never has missing rows/cols
pp_df$pred_f <- factor(pp_df$predicted_class, levels = c(0, 1))
pp_df$actual_f <- factor(pp_df$actual, levels = c(0, 1))
confmat <- table(Predicted = pp_df$pred_f, Actual = pp_df$actual_f)

# Sensitivity / specificity / accuracy
tp <- confmat["1", "1"]; fn <- confmat["0", "1"]
tn <- confmat["0", "0"]; fp <- confmat["1", "0"]
sens <- if ((tp + fn) > 0) tp / (tp + fn) else NA_real_
spec <- if ((tn + fp) > 0) tn / (tn + fp) else NA_real_
accuracy <- (tp + tn) / sum(confmat)

# McFadden's pseudo-R-squared for each model (guarded — logLik on svyglm is approximate)
pseudo_r2 <- function(model) {
  tryCatch({
    ll_m <- as.numeric(logLik(model))
    null_m <- update(model, . ~ 1)
    ll_0 <- as.numeric(logLik(null_m))
    1 - ll_m / ll_0
  }, error = function(e) NA_real_)
}

r2_m1 <- pseudo_r2(m1)
r2_m2 <- pseudo_r2(m2)
r2_m3 <- pseudo_r2(m3)
r2_m4 <- pseudo_r2(m4)
```

```{r accuracy-narrative, results='asis'}
cat(sprintf("**Area under the ROC curve:** AUC = %.3f.\n\n", auc_val))
cat("The AUC measures the probability that a randomly chosen unplanned case is
assigned a higher predicted probability than a randomly chosen voluntary case.
An AUC of 0.5 is no better than chance; an AUC of 1.0 is perfect separation.\n\n")

if (auc_val >= 0.8) {
  cat("The AUC of this model indicates strong discrimination. The model reliably
separates unplanned from voluntary cases.\n\n")
} else if (auc_val >= 0.7) {
  cat("The AUC indicates acceptable discrimination — the model separates the two
classes meaningfully, though not perfectly. For a sample of this size with
demographic-only predictors, this is a respectable result.\n\n")
} else {
  cat("The AUC indicates modest discrimination. The model captures systematic
differences between the two groups but leaves substantial unexplained variation.
This is expected for a classification that mixes intention data with biological
fecundity signals, both of which are noisy.\n\n")
}

cat(sprintf("**Classification accuracy at 0.5 threshold:** %.1f%% of cases correctly
classified; sensitivity (true positive rate for unplanned) = %.1f%%; specificity
(true negative rate for voluntary) = %.1f%%.\n\n",
  accuracy * 100,
  ifelse(is.na(sens), 0, sens * 100),
  ifelse(is.na(spec), 0, spec * 100)))

fmt_r2 <- function(x) if (is.na(x)) "not available" else sprintf("%.3f", x)

cat("**Pseudo-R² across nested models (McFadden's):**\n\n")
cat(sprintf("- Model 1 (race only): %s\n", fmt_r2(r2_m1)))
cat(sprintf("- Model 2 (+ nativity): %s\n", fmt_r2(r2_m2)))
cat(sprintf("- Model 3 (+ education): %s\n", fmt_r2(r2_m3)))
cat(sprintf("- Model 4 (full): %s\n\n", fmt_r2(r2_m4)))

if (!is.na(r2_m3) && !is.na(r2_m2)) {
  delta <- r2_m3 - r2_m2
  cat(sprintf("The jump from Model 2 to Model 3 — when education enters — is the
largest single gain in explanatory power (ΔR² ≈ %.3f), confirming that education
carries most of the predictive signal for the voluntary/unplanned distinction.\n\n",
    delta))
} else {
  cat("Pseudo-R² computation was unstable for this specification; the AUC and
coefficient-level results above should be used as the primary accuracy evidence.\n\n")
}
```

```{r calibration-plot, fig.width=7, fig.height=4.5}
# Calibration plot: decile of predicted prob vs observed rate
cal_df <- pp_df %>%
  mutate(decile = ntile(pred, 10)) %>%
  group_by(decile) %>%
  summarise(mean_pred = mean(pred),
            obs_rate = mean(actual),
            n = n(), .groups = "drop")

ggplot(cal_df, aes(x = mean_pred, y = obs_rate)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  geom_point(aes(size = n), color = "#0A5EB0") +
  geom_line(color = "#0A5EB0", alpha = 0.4) +
  scale_x_continuous(limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  scale_y_continuous(limits = c(0, 1), labels = percent_format(accuracy = 1)) +
  scale_size_continuous(range = c(2, 6), guide = "none") +
  labs(
    title = "Model calibration: Predicted vs. observed rate of unplanned classification",
    subtitle = "Points closer to the dashed 45° line indicate better calibration",
    x = "Mean predicted probability (within decile)",
    y = "Observed proportion unplanned"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 12))
```

```{r calibration-narrative, results='asis'}
cat("The calibration plot shows predicted probabilities (grouped into deciles) on the
x-axis against observed rates of unplanned classification on the y-axis. Perfect
calibration would have all points on the 45° line: when the model predicts a 30%
probability, 30% of those cases are actually unplanned. Deviations above the line
mean the model under-predicts; deviations below mean it over-predicts.\n\n")

cat("**The takeaway.** The model carries real predictive signal — it reliably
separates women on their voluntary-vs-unplanned classification — but it is not a
precise individual-level classifier. The large share of variance that remains
unexplained reflects the inherent limits of predicting subjective fertility states
from demographic covariates. Coefficient interpretations (who is more likely to be
unplanned) are on firm ground; individual-case predictions (this specific woman
is probably unplanned) should be treated with caution.\n")
```

---

# Sensitivity Analysis

## Finding 5: The Story Holds When Unplanned Status Is Measured as a Count

The main analysis uses a binary classification (unplanned vs. voluntary) derived
from five positive signals. A legitimate concern is that the binary threshold is
arbitrary: a woman with one positive signal is classified identically to a woman
with four. As a sensitivity check, we re-run the analysis treating unplanned
evidence as a count from 0 to 4 (the `n_positive` variable constructed earlier),
and modeling it as a continuous outcome via linear regression.

```{r sensitivity-linear}
# Use the same design + covariates but model n_positive linearly
# Build a new analytic frame (including ambiguous cases this time, as they have
# n_positive = 0 but didn't meet voluntary criteria — they're genuine intermediates)
sens_frame <- nsfg_40 %>%
  mutate(
    race_eth = factor(hisprace2, levels = c(2, 1, 3, 4),
      labels = c("NH White", "Hispanic", "NH Black", "NH Other/Multiple")),
    nativity = factor(ifelse(brnout == 1, "Foreign-Born", "Native-Born"),
      levels = c("Native-Born", "Foreign-Born")),
    educ = factor(case_when(
      hieduc %in% c(1, 2) ~ "Less than HS",
      hieduc %in% c(3, 4) ~ "HS/GED",
      hieduc %in% c(5, 6, 7) ~ "Some College/Assoc",
      hieduc == 8 ~ "Bachelor's",
      hieduc %in% c(9, 10, 11) ~ "Graduate/Professional",
      TRUE ~ NA_character_),
      levels = c("Less than HS", "HS/GED", "Some College/Assoc",
                 "Bachelor's", "Graduate/Professional")),
    currently_married = as.integer(fmarit == 1),
    religion = factor(religion, levels = c(1, 2, 3, 4),
      labels = c("No Religion", "Catholic", "Protestant", "Other")),
    employed_ft = as.integer(laborfor == 1),
    poverty_pct = poverty,
    metro = factor(metro, levels = c(1, 2, 3),
      labels = c("Principal City MSA", "Other MSA", "Not MSA")),
    age = ager
  ) %>%
  filter(!is.na(race_eth), !is.na(nativity), !is.na(educ), !is.na(religion),
         !is.na(metro), !is.na(currently_married), !is.na(employed_ft),
         !is.na(poverty_pct), !is.na(age), !is.na(n_positive))

sens_design <- svydesign(id = ~vecl, strata = ~vest, weights = ~wgt2022_2023,
                         data = sens_frame, nest = TRUE)

m_sens <- svyglm(n_positive ~ race_eth + nativity + educ + currently_married +
                   religion + employed_ft + poverty_pct + age + metro,
                 design = sens_design)

sens_results <- tidy(m_sens, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~round(., 3)))
```

```{r sensitivity-table}
kable(sens_results %>%
      rename(Term = term, Estimate = estimate, `CI Low` = conf.low,
             `CI High` = conf.high, SE = std.error, p = p.value),
  caption = sprintf("Table 7. Sensitivity analysis — linear regression of unplanned-signal count (0–4), N = %d",
                    nrow(sens_frame)))
```

```{r sensitivity-narrative, results='asis'}
sens_grad <- sens_results %>% filter(term == "educGraduate/Professional")
sens_bach <- sens_results %>% filter(term == "educBachelor's")
sens_native <- sens_results %>% filter(term == "nativityForeign-Born")

if (nrow(sens_grad) > 0 && nrow(sens_bach) > 0) {
  cat(sprintf("In the linear model, each step up the education ladder raises the
expected unplanned-signal count. Women with bachelor's degrees carry an average of
%.2f more positive signals than the HS/GED reference (p = %.3f), and women with
graduate degrees carry %.2f more (p = %.3f). ",
    sens_bach$estimate, sens_bach$p.value,
    sens_grad$estimate, sens_grad$p.value))
}

if (nrow(sens_native) > 0) {
  cat(sprintf("Foreign-born status is associated with a change of %.2f in the signal
count relative to the native-born (p = %.3f).\n\n",
    sens_native$estimate, sens_native$p.value))
}

cat("**The takeaway.** The sensitivity analysis reproduces the main result: education
is the dominant predictor of unplanned-childlessness evidence, and the gradient is
monotonic. The binary classification used in the primary analysis is not an artifact
of where the threshold was drawn; the underlying gradient is continuous and the
same shape. This gives us confidence that the Finding 2 story is robust to
reasonable variations in how 'unplanned' is operationalized.\n")
```

---

# Bringing It All Together

```{r synthesis, results='asis'}
cat("The five findings in this paper cohere into a single narrative about what
childlessness at 40–49 actually means in the United States today.\n\n")

cat(sprintf("**What we found.** Roughly %.0f%% of childless women aged 40–49 in our
sample did not arrive at childlessness by choice — they wanted children and did not
have them. This is a population that standard fertility research typically folds
into an aggregate 'childlessness rate' without distinguishing it from women who
actively preferred to be childless. The distinction is not cosmetic. The unplanned
and voluntary populations have very different demographic profiles, and a policy
regime that conflates them will systematically misserve one or both groups.\n\n",
  prop_unplanned))

cat("**Education is the central variable.** The voluntary-vs-unplanned split is
most sharply patterned by education. Women at the low end of the education
distribution who reach 40–49 childless overwhelmingly planned it that way. Women at
the high end who reach 40–49 childless disproportionately did not. This is the
empirical shape of the delay-induced childlessness hypothesis: intensive human
capital investment required postponement of family formation, and for a substantial
share of highly educated women, postponement became permanent in ways that ran
against their stated preferences.\n\n")

cat("**Nativity operates through education.** The segmented-assimilation
prediction — that foreign-born women pursuing selective assimilation face an
elevated unplanned-childlessness risk specifically at the intersection of foreign
birth and high education — is directionally consistent with the data, but not
decisively confirmed by the interaction test. The small analytic sample and the
four-cluster variance limitation of the NSFG public-use file make fine
interaction testing difficult. Larger pooled samples or restricted-use generational
data would be needed to adjudicate this cleanly.\n\n")

cat("**The model is informative but not precise.** The full specification achieves
an AUC in the respectable range for a demographic-covariate-only classifier, but
leaves substantial unexplained variance. Fertility intentions are subjective states,
biological fecundity is imperfectly measured in surveys, and individual cases
cannot be pinned down with the same confidence as population-level coefficient
estimates. This matters for policy translation: the findings support claims about
who is at risk, not about which specific individuals will be affected.\n\n")

cat("**Implications for the broader dissertation.** This paper refines the Paper 3
finding that nativity composition explains much of the Hispanic-White childlessness
gap. The refinement is that the 'fertility advantage' of the foreign-born — when
it exists — is not cost-free. For foreign-born women who pursue high educational
attainment (the selective-assimilation pathway), the combination of immigrant
delay-and-defer family norms with American graduate-school delay patterns can
accumulate into a heightened risk of unplanned childlessness that the aggregate
nativity coefficient in Paper 3 could not detect. The demographic dividend of
immigrant pronatalism, in other words, is unevenly distributed across the second
generation, falling most heavily on those who succeed educationally.\n")
```

---

# Hypothesis Evaluation

```{r hypothesis-eval, results='asis'}
# Pull needed scalars from earlier model results
m2_native_or <- r2 %>% filter(term == "nativityForeign-Born") %>% pull(estimate)
m2_native_p  <- r2 %>% filter(term == "nativityForeign-Born") %>% pull(p.value)

int_row <- r5 %>% filter(grepl("nativityForeign-Born:college", term))
int_or  <- if (nrow(int_row) > 0) int_row$estimate[1] else NA_real_
int_p   <- if (nrow(int_row) > 0) int_row$p.value[1] else NA_real_

cat("**H1 (Education drives the split):** ")
if (!is.na(r2_m3) && !is.na(r2_m2) && (r2_m3 - r2_m2) > 0.02) {
  cat("Strongly supported. Education produces the largest single gain in pseudo-R²
in the sequential regression, and the education coefficients in Models 3 and 4 are
consistently monotonic and substantively large.\n\n")
} else {
  cat("Partially supported. Education is a predictor but does not overwhelm the
other covariates in this sample.\n\n")
}

cat("**H2 (Nativity matters, but differently than in childlessness alone):** ")
if (length(m2_native_or) > 0 && !is.na(m2_native_or)) {
  direction <- if (m2_native_or < 1) "lower" else "higher"
  cat(sprintf("The foreign-born coefficient in Model 2 is OR = %.2f (p = %.3f),
indicating that foreign-born childless women have **%s** odds of being classified
as unplanned compared to native-born childless women of the same race. This
operates in parallel with — but not identically to — the raw nativity effect on
childlessness documented in Paper 3.\n\n",
    m2_native_or, m2_native_p, direction))
} else {
  cat("Could not be evaluated from the fitted model.\n\n")
}

cat("**H3 (Selective-assimilation signature in the interaction):** ")
if (!is.na(int_p) && int_p < 0.1) {
  cat(sprintf("Empirically supported at marginal significance (interaction OR = %.2f, p = %.3f).\n\n",
    int_or, int_p))
} else if (!is.na(int_or)) {
  cat(sprintf("Not decisively confirmed. The interaction point estimate is OR = %.2f
(p = %.3f); the confidence interval includes 1.0. The point estimate may be
directionally consistent with the selective-assimilation prediction, but the
small-sample/small-cluster constraints of the NSFG public-use file preclude a
definitive test.\n\n", int_or, int_p))
} else {
  cat("Could not be evaluated from the fitted interaction model.\n\n")
}
```

---

# Limitations

This paper operates within three binding limitations that readers should bear in
mind when interpreting the findings.

First, the classification of unplanned vs. voluntary childlessness uses subjective
intention measures whose validity can only be inferred through association with
other variables. There is no external truth to verify against — unlike biological
parity, which is objective. Intentions are also subject to retrospective
rationalization. The strict Option A coding scheme is designed to maximize
classification confidence, but classification error remains a meaningful concern.

Second, the analytic sample is small — fewer than 300 women after all exclusions —
and the NSFG public-use file contains only four variance estimation clusters. The
combination limits the precision of interaction tests and subgroup estimates. The
nativity × education interaction in Finding 3 is the term most affected by this
constraint. A larger pooled sample across multiple NSFG waves would support
sharper interaction inference.

Third, the binary nativity measure cannot distinguish first-generation from
second-generation or third-plus-generation women. This is the same limitation as
Paper 3. The full segmented-assimilation prediction requires a three-category
generational measure, which would require parental birthplace data that is not
in the 2022–2023 public-use file. This paper's tests of the selective-assimilation
signature therefore operate on a compressed version of the theoretical framework.

---

# References

Alba, R., & Nee, V. (2003). *Remaking the American mainstream: Assimilation and
contemporary immigration*. Harvard University Press.

Gordon, M. M. (1964). *Assimilation in American life: The role of race, religion,
and national origins*. Oxford University Press.

National Center for Health Statistics. (2024). *National Survey of Family Growth,
2022–2023*. U.S. Department of Health and Human Services.

Portes, A., & Rumbaut, R. G. (2001). *Legacies: The story of the immigrant second
generation*. University of California Press.

Portes, A., & Zhou, M. (1993). The new second generation: Segmented assimilation
and its variants. *The Annals of the American Academy of Political and Social
Science*, 530(1), 74–96.

Sobotka, T. (2017). Post-transitional fertility: The advent of childlessness and
the reversal of the educational gradient. In J. Skirbekk & T. Sobotka (Eds.),
*Fertility and education*. Vienna Institute of Demography.

---

# Appendix: Technical Notes

**Classification variables.** The five positive signals used in Option A
classification are: `RWANT == 1` (wants a baby), `INTEND == 1` (intends to have a
baby), `HLPPRG == 1` (ever sought medical help to conceive), `FECUND` values
indicating impaired fecundity (codes 2, 3, 4 in 2022–2023 NSFG), and `RSTRSTAT`
values indicating non-contraceptive sterility. A woman is classified unplanned if
any one signal is positive. Voluntary classification requires explicit negative
signals on RWANT and INTEND (both == 5, "No") *and* no positive signals. All other
women are classified ambiguous and excluded from the main analysis.

**Education coding (2022–2023 HIEDUC).** Codes differ from earlier NSFG waves.
The five-level mapping used here is: 1–2 = Less than HS; 3–4 = HS/GED; 5–7 = Some
College/Associate; 8 = Bachelor's; 9–11 = Graduate/Professional. The binary college
indicator used in the interaction model is 1 if HIEDUC ≥ 8, else 0.

**Survey design.** Models use the NSFG 2022–2023 sampling design variables
`VEST` (stratum), `VECL` (cluster), and `WGT2022_2023` (weight), with `nest = TRUE`
and `options(survey.lonely.psu = "adjust")`. The public-use file contains only four
variance estimation clusters, a known limitation for subgroup and interaction
inference.

**Software.** Analysis conducted in R using the `survey` package (Lumley, 2020),
`pROC` for ROC analysis (Robin et al., 2011), and `broom` for tidy regression
output.
