---
title: "Nativity, Economic Status, and Childlessness in America"
subtitle: "A Generational Analysis of the National Survey of Family Growth"
author: "Delroy Nichols"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 5, dpi = 150
)
rm(list = ls())
gc()
set.seed(42)
options(survey.lonely.psu = "adjust")
```

# Abstract

This paper directly tests the hypothesis that nativity — being foreign-born versus
native-born — explains racial and ethnic differences in completed childlessness among
U.S. women. Using pooled data from the 2015–2017 and 2017–2019 waves of the National
Survey of Family Growth (NSFG), which expanded the age range to 15–49, we analyze women
aged 40–49 who have reached the end of their childbearing years. The NSFG provides what
the Survey of Consumer Finances cannot: a definitive measure of biological fertility
(children ever born), detailed nativity data (respondent's, mother's, and father's
country of birth), and standard economic and demographic controls.

Using a sequential logistic regression design, we first estimate the unconditional racial
gradient in completed childlessness, then introduce generational status (first generation,
second generation, third-plus generation) to determine how much of the racial gradient is
attributable to nativity composition, and finally add economic and demographic controls
to identify the residual. We test whether the effects of education and marriage on
childlessness differ by generational status through interaction models.

This study completes a three-paper thesis: the first paper analyzed credit card
indebtedness, the second examined the financial balance sheet of co-resident
childlessness, and this paper provides the direct nativity test that the SCF-based
analyses could only theorize about.

---

```{r load-packages}
required_packages <- c(
  "tidyverse", "haven", "survey",
  "broom", "knitr", "ggplot2", "scales"
)

for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg, repos = "http://cran.us.r-project.org")
    library(pkg, character.only = TRUE)
  }
}
```

# Introduction

The first two papers in this thesis established that (1) the transition to parenthood is
the strongest demographic predictor of credit card indebtedness, and (2) the financial
balance sheet of childless households differs systematically from that of parents — with
retirement asset accumulation and credit constraints, not income, distinguishing the two
groups. Both analyses used the Survey of Consumer Finances (SCF), which provides
unmatched financial detail but lacks any measure of nativity or biological fertility.

The SCF companion paper found that racial and ethnic differences in co-resident
childlessness were fully absorbed by financial and demographic controls, and hypothesized
— based on external literature — that immigration-driven compositional effects might
explain the descriptive racial gradient. This paper tests that hypothesis directly.

The National Survey of Family Growth (NSFG) is the premier U.S. dataset on fertility and
family formation. It provides three capabilities the SCF lacks: (1) a definitive measure
of biological fertility — children ever born (`PARITY`), not merely co-resident children;
(2) nativity data for the respondent and both parents, enabling construction of a
three-category generational status variable; and (3) fertility intentions, which can
distinguish voluntary from involuntary childlessness.

**Research Questions:**

1. *Primary:* To what extent does nativity (foreign-born vs. native-born) explain racial
   and ethnic differences in completed childlessness among U.S. women?

2. *Secondary (Mediation):* Do economic factors (education, income, employment) and
   family structure (marital status) mediate the relationship between nativity and
   childlessness?

3. *Secondary (Interaction):* Do the effects of education and marriage on childlessness
   differ by generational status?

---

# Data and Methods

## Data Source and Pooling Strategy

We use the female respondent files from the 2015–2017 and 2017–2019 continuous NSFG.
These two waves are selected because they expanded the survey age range from 15–44 to
15–49, enabling analysis of women aged 40–49 — the standard demographic window for
completed fertility (Guzzo & Loo, 2023). Earlier waves (2011–2013, 2013–2015) are
excluded because they do not include women over 44.

Pooling two waves increases sample size for subgroup analysis. We include a survey wave
indicator as a fixed effect in all models to account for period differences in fertility
between the two data collection windows.

```{r download-data, results='asis'}
# ============================================================
# DOWNLOAD NSFG FEMALE RESPONDENT FILES FROM CDC FTP
# ============================================================

dir.create("nsfg_data", showWarnings = FALSE, recursive = TRUE)

# 2015-2017 Female Respondent File
url_1517 <- "https://ftp.cdc.gov/pub/Health_Statistics/NCHS/Datasets/NSFG/2015_2017_FemRespData.sas7bdat"
file_1517 <- "nsfg_data/fem_resp_2015_2017.sas7bdat"

if (!file.exists(file_1517)) {
  cat("Downloading 2015-2017 NSFG female respondent file...\n")
  tryCatch(
    download.file(url_1517, file_1517, mode = "wb", quiet = TRUE),
    error = function(e) cat("Download error:", conditionMessage(e), "\n")
  )
}

# 2017-2019 Female Respondent File
url_1719 <- "https://ftp.cdc.gov/pub/Health_Statistics/NCHS/Datasets/NSFG/2017_2019_FemRespData.sas7bdat"
file_1719 <- "nsfg_data/fem_resp_2017_2019.sas7bdat"

if (!file.exists(file_1719)) {
  cat("Downloading 2017-2019 NSFG female respondent file...\n")
  tryCatch(
    download.file(url_1719, file_1719, mode = "wb", quiet = TRUE),
    error = function(e) cat("Download error:", conditionMessage(e), "\n")
  )
}

cat("Download step complete.\n")
```

```{r load-and-pool}
# ============================================================
# LOAD AND HARMONIZE
# ============================================================

raw_1517 <- data.frame(read_sas(file_1517))
names(raw_1517) <- tolower(names(raw_1517))
raw_1517$wave <- "2015-2017"

raw_1719 <- data.frame(read_sas(file_1719))
names(raw_1719) <- tolower(names(raw_1719))
raw_1719$wave <- "2017-2019"

# Identify common variables across waves
common_vars <- intersect(names(raw_1517), names(raw_1719))

# Bind the two waves on common variables
nsfg_pooled <- bind_rows(
  raw_1517[, common_vars],
  raw_1719[, common_vars]
)

# Re-attach wave indicator (lost if not in common_vars)
nsfg_pooled$wave <- c(
  rep("2015-2017", nrow(raw_1517)),
  rep("2017-2019", nrow(raw_1719))
)

cat(sprintf("Pooled NSFG: %s total female respondents\n",
  formatC(nrow(nsfg_pooled), big.mark = ",")))
cat(sprintf("  2015-2017: %s\n", formatC(nrow(raw_1517), big.mark = ",")))
cat(sprintf("  2017-2019: %s\n", formatC(nrow(raw_1719), big.mark = ",")))

# Clean up large objects
rm(raw_1517, raw_1719)
gc()
```

## Variable Construction

```{r variables, results='asis'}
# ============================================================
# CONSTRUCT KEY VARIABLES
# ============================================================

nsfg <- nsfg_pooled

# --- Outcome: Completed Childlessness ---
# PARITY = number of live births ever had (recode)
# For women 40-49, PARITY == 0 indicates completed childlessness
nsfg$completed_childless <- as.integer(nsfg$parity == 0)

# --- Age ---
nsfg$age <- nsfg$ager  # AGER = age at interview (recode)

# --- Race/Ethnicity ---
# HISPRACE2: 1=Hispanic, 2=NH White, 3=NH Black, 4=NH Other
nsfg$race_eth <- factor(nsfg$hisprace2,
  levels = c(2, 1, 3, 4),  # White as reference
  labels = c("NH White", "Hispanic", "NH Black", "NH Other")
)

# --- Nativity / Generational Status ---
# BRNOUT: 1=born outside US (50 states + DC), 5=born in US
# Parental nativity variables vary by wave; common names include
# MOMFBRN / DADFBRN or similar. We check for availability.

# Respondent nativity
nsfg$foreign_born <- as.integer(nsfg$brnout == 1)

# Attempt to construct generational status
# Check for parental birthplace variables
has_mom <- "momfbrn" %in% names(nsfg)
has_dad <- "dadfbrn" %in% names(nsfg)

if (has_mom & has_dad) {
  # Mother foreign-born
  nsfg$mom_foreign <- as.integer(nsfg$momfbrn == 1)
  # Father foreign-born
  nsfg$dad_foreign <- as.integer(nsfg$dadfbrn == 1)

  # Generational status:
  # 1st gen = respondent born outside US
  # 2nd gen = respondent born in US, at least one parent born outside US
  # 3rd+ gen = respondent and both parents born in US
  nsfg$gen_status <- case_when(
    nsfg$foreign_born == 1 ~ "1st Generation",
    nsfg$foreign_born == 0 & (nsfg$mom_foreign == 1 | nsfg$dad_foreign == 1) ~ "2nd Generation",
    nsfg$foreign_born == 0 & nsfg$mom_foreign == 0 & nsfg$dad_foreign == 0 ~ "3rd+ Generation",
    TRUE ~ NA_character_
  )
  nsfg$gen_status <- factor(nsfg$gen_status,
    levels = c("3rd+ Generation", "1st Generation", "2nd Generation"))

  cat("Generational status variable constructed from BRNOUT, MOMFBRN, DADFBRN.\n")
} else {
  # Fallback: use respondent nativity only (foreign-born vs. native-born)
  nsfg$gen_status <- factor(
    ifelse(nsfg$foreign_born == 1, "Foreign-Born", "Native-Born"),
    levels = c("Native-Born", "Foreign-Born")
  )
  cat("NOTE: Parental nativity variables not found in pooled data.\n")
  cat("Using respondent nativity only (foreign-born vs. native-born).\n")
  if (!has_mom) cat("  Missing: MOMFBRN\n")
  if (!has_dad) cat("  Missing: DADFBRN\n")
}

# --- Education ---
# HIEDUC: highest education completed
# Recode into 4 categories
nsfg$educ <- case_when(
  nsfg$hieduc %in% c(5, 6) ~ "Less than HS",
  nsfg$hieduc == 7 ~ "HS Diploma/GED",
  nsfg$hieduc %in% c(8, 9) ~ "Some College/Assoc",
  nsfg$hieduc %in% c(10, 11) ~ "Bachelor's or Higher",
  TRUE ~ NA_character_
)
nsfg$educ <- factor(nsfg$educ,
  levels = c("Less than HS", "HS Diploma/GED", "Some College/Assoc", "Bachelor's or Higher"))

nsfg$college <- as.integer(nsfg$hieduc %in% c(10, 11))

# --- Marital Status ---
# FMARITAL: 1=married, 2=widowed, 3=divorced, 4=separated, 5=never married
nsfg$ever_married <- as.integer(nsfg$fmarital %in% c(1, 2, 3, 4))
nsfg$currently_married <- as.integer(nsfg$fmarital == 1)

# --- Employment ---
# LABORFOR: 1=working, 2=looking, 3=not looking
nsfg$employed <- as.integer(nsfg$laborfor == 1)

# --- Income / Poverty ---
# POVERTY: poverty level percentage (0-500+)
nsfg$poverty_pct <- nsfg$poverty

# --- Religion ---
# RELIGION: 1=No religion, 2=Catholic, 3=Protestant, 4=Other
nsfg$religion <- factor(nsfg$religion,
  levels = 1:4,
  labels = c("No Religion", "Catholic", "Protestant", "Other"))

# --- Fertility Intentions (sensitivity check) ---
# RWANT: whether R wants (more) children
# Available for most respondents
if ("rwant" %in% names(nsfg)) {
  nsfg$wants_children <- as.integer(nsfg$rwant == 1)
}

# --- Wave indicator ---
nsfg$wave_f <- factor(nsfg$wave)

cat(sprintf("\nTotal pooled respondents: %s\n", formatC(nrow(nsfg), big.mark = ",")))
cat(sprintf("Women aged 40-49: %s\n",
  formatC(sum(nsfg$age >= 40 & nsfg$age <= 49, na.rm = TRUE), big.mark = ",")))
```

## Sample Restriction and Survey Design

```{r sample-and-design}
# Restrict to women aged 40-49 for completed fertility analysis
nsfg_40 <- nsfg %>% filter(age >= 40, age <= 49)

cat(sprintf("Analysis sample (women 40-49): %s\n", formatC(nrow(nsfg_40), big.mark = ",")))

# Identify weight variable (differs by wave)
# 2015-2017 uses wgt2015_2017; 2017-2019 uses wgt2017_2019
# For pooled analysis, we halve each wave's weight
if ("wgt2015_2017" %in% names(nsfg_40) & "wgt2017_2019" %in% names(nsfg_40)) {
  nsfg_40$pooled_weight <- ifelse(
    nsfg_40$wave == "2015-2017",
    nsfg_40$wgt2015_2017 / 2,
    nsfg_40$wgt2017_2019 / 2
  )
} else {
  # Fallback: try common weight variable names
  wt_cols <- grep("^wgt", names(nsfg_40), value = TRUE)
  cat("Available weight columns:", paste(wt_cols, collapse = ", "), "\n")
  if (length(wt_cols) >= 1) {
    nsfg_40$pooled_weight <- nsfg_40[[wt_cols[1]]] / 2
    cat("Using", wt_cols[1], "as weight (halved for pooling).\n")
  }
}

# Survey design
# SEST = stratum, SECU = cluster (PSU)
nsfg_design <- svydesign(
  id      = ~ secu,
  strata  = ~ sest,
  weights = ~ pooled_weight,
  data    = nsfg_40,
  nest    = TRUE
)

cat("Survey design object created.\n")
```

## Analytical Plan

Our analysis proceeds in four stages:

1. **Descriptive analysis:** Cross-tabulate completed childlessness by race/ethnicity
   and generational status to show unconditional rates.

2. **Sequential logistic regression:**
   - *Model 1:* `completed_childless ~ race_eth` (unconditional racial gradient)
   - *Model 2:* `completed_childless ~ race_eth + gen_status` (does generation account
     for the race effect?)
   - *Model 3:* `completed_childless ~ race_eth + gen_status + controls` (full model
     with economic, demographic, and period controls)

3. **Interaction models:** Test `gen_status × college` and `gen_status × currently_married`
   to assess whether the effects of education and marriage on childlessness differ by
   generational status.

4. **Sensitivity check:** Include fertility intentions (`wants_children`) in a model
   variant to distinguish voluntary from involuntary childlessness.

All models apply NSFG survey weights and account for the complex sample design (strata
and clusters). Period effects are controlled via a wave fixed effect.

---

# Descriptive Results

## Cell Sizes by Generation and Race

```{r cell-sizes}
cat("===== Sample Sizes by Generation and Race (Unweighted) =====\n\n")
cell_table <- nsfg_40 %>%
  filter(!is.na(gen_status), !is.na(race_eth)) %>%
  count(gen_status, race_eth) %>%
  pivot_wider(names_from = race_eth, values_from = n, values_fill = 0)

kable(cell_table, caption = "Table 1. Unweighted Cell Sizes by Generational Status and Race/Ethnicity")
```

```{r cell-size-check, results='asis'}
# Flag small cells
min_cells <- nsfg_40 %>%
  filter(!is.na(gen_status), !is.na(race_eth)) %>%
  count(gen_status, race_eth) %>%
  filter(n < 30)

if (nrow(min_cells) > 0) {
  cat("\n**Warning:** The following cells have fewer than 30 observations. Estimates for
these subgroups should be interpreted with extreme caution:\n\n")
  for (j in 1:nrow(min_cells)) {
    cat(sprintf("- %s × %s: n = %d\n", min_cells$gen_status[j], min_cells$race_eth[j],
      min_cells$n[j]))
  }
  cat("\n")
} else {
  cat("\nAll generation × race cells contain at least 30 observations.\n\n")
}
```

## Completed Childlessness Rates

```{r desc-childless}
cat("===== Weighted Childlessness Rate by Race/Ethnicity =====\n\n")
print(svyby(~completed_childless, ~race_eth, nsfg_design, svymean, na.rm = TRUE))

cat("\n===== By Generational Status =====\n\n")
print(svyby(~completed_childless, ~gen_status, nsfg_design, svymean, na.rm = TRUE))

cat("\n===== By Race × Generation =====\n\n")
print(svyby(~completed_childless, ~race_eth + gen_status, nsfg_design, svymean,
            na.rm = TRUE))

cat("\n===== By Education =====\n\n")
print(svyby(~completed_childless, ~educ, nsfg_design, svymean, na.rm = TRUE))

cat("\n===== By Marital Status (Currently Married) =====\n\n")
print(svyby(~completed_childless, ~currently_married, nsfg_design, svymean, na.rm = TRUE))

cat("\n===== By Religion =====\n\n")
print(svyby(~completed_childless, ~religion, nsfg_design, svymean, na.rm = TRUE))
```

## Visualizations

```{r viz-race-gen, fig.width=11, fig.height=6}
# Cross-tabulation visualization: childlessness by race x generation
race_gen <- svyby(~completed_childless, ~race_eth + gen_status, nsfg_design,
                   svymean, na.rm = TRUE)
rg_df <- as.data.frame(race_gen)

ggplot(rg_df, aes(x = race_eth, y = completed_childless * 100,
                   fill = gen_status)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(aes(ymin = (completed_childless - 1.96 * se) * 100,
                     ymax = (completed_childless + 1.96 * se) * 100),
                position = position_dodge(width = 0.8), width = 0.2) +
  scale_fill_brewer(palette = "Set2", name = "Generation") +
  labs(
    title = "Completed Childlessness by Race/Ethnicity and Generational Status",
    subtitle = "Women aged 40-49, NSFG 2015-2019 (survey-weighted)",
    x = NULL, y = "% Childless (no children ever born)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 15, hjust = 1)) +
  ylim(0, NA)
```

```{r viz-education, fig.width=10, fig.height=5}
educ_rates <- svyby(~completed_childless, ~educ, nsfg_design, svymean, na.rm = TRUE)
ed_df <- as.data.frame(educ_rates)

ggplot(ed_df, aes(x = educ, y = completed_childless * 100, fill = educ)) +
  geom_col(width = 0.6) +
  geom_errorbar(aes(ymin = (completed_childless - 1.96 * se) * 100,
                     ymax = (completed_childless + 1.96 * se) * 100),
                width = 0.2) +
  geom_text(aes(label = sprintf("%.1f%%", completed_childless * 100)),
            vjust = -1.2, size = 3.5) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(
    title = "Completed Childlessness by Education (Women 40-49)",
    subtitle = "NSFG 2015-2019 (survey-weighted)",
    x = NULL, y = "% Childless"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  ylim(0, NA)
```

---

# Sequential Logistic Regression

## Model 1: Unconditional Racial Gradient

```{r model1}
m1 <- svyglm(completed_childless ~ race_eth,
              design = nsfg_design, family = quasibinomial())
m1_results <- tidy(m1, conf.int = TRUE, exponentiate = TRUE)

cat("===== Model 1: Childlessness ~ Race/Ethnicity =====\n\n")
kable(m1_results %>%
  mutate(across(where(is.numeric), ~round(., 4))),
  caption = "Model 1: Unconditional Racial Gradient (Odds Ratios)")
```

## Model 2: Adding Generational Status

```{r model2}
m2 <- svyglm(completed_childless ~ race_eth + gen_status,
              design = nsfg_design, family = quasibinomial())
m2_results <- tidy(m2, conf.int = TRUE, exponentiate = TRUE)

cat("\n===== Model 2: + Generational Status =====\n\n")
kable(m2_results %>%
  mutate(across(where(is.numeric), ~round(., 4))),
  caption = "Model 2: Race + Generational Status (Odds Ratios)")
```

```{r model2-compare, results='asis'}
# Compare race coefficients between M1 and M2
m1_hisp <- m1_results %>% filter(term == "race_ethHispanic")
m2_hisp <- m2_results %>% filter(term == "race_ethHispanic")

if (nrow(m1_hisp) > 0 & nrow(m2_hisp) > 0) {
  pct_change_hisp <- round(100 * (m2_hisp$estimate - m1_hisp$estimate) /
    (m1_hisp$estimate - 1), 1)
  cat(sprintf("\n**Attenuation of Hispanic coefficient:** The Hispanic OR changed from
%.3f (Model 1) to %.3f (Model 2) when generational status was added. This represents
approximately %.1f%% attenuation of the Hispanic-White gap, indicating that
generational composition %s a substantial portion of the racial difference in
childlessness.\n\n",
    m1_hisp$estimate, m2_hisp$estimate, abs(pct_change_hisp),
    ifelse(abs(pct_change_hisp) > 20, "accounts for", "does not fully explain")))
}

m1_black <- m1_results %>% filter(term == "race_ethNH Black")
m2_black <- m2_results %>% filter(term == "race_ethNH Black")

if (nrow(m1_black) > 0 & nrow(m2_black) > 0) {
  pct_change_black <- round(100 * (m2_black$estimate - m1_black$estimate) /
    (m1_black$estimate - 1), 1)
  cat(sprintf("**Attenuation of Black coefficient:** The Black OR changed from %.3f
(Model 1) to %.3f (Model 2), representing %.1f%% attenuation. %s\n\n",
    m1_black$estimate, m2_black$estimate, abs(pct_change_black),
    ifelse(abs(pct_change_black) < 10,
      "As hypothesized, generational status contributes minimally to the Black-White gap, consistent with the predominantly native-born composition of the Black population.",
      "Generational status contributes meaningfully to the Black-White gap, suggesting that Black immigrant fertility plays a larger role than expected.")))
}
```

## Model 3: Full Model with Controls

```{r model3}
m3 <- svyglm(completed_childless ~ race_eth + gen_status +
                college + currently_married + employed +
                poverty_pct + religion + age + wave_f,
              design = nsfg_design, family = quasibinomial())
m3_results <- tidy(m3, conf.int = TRUE, exponentiate = TRUE)

cat("\n===== Model 3: Full Model with Controls =====\n\n")
kable(m3_results %>%
  mutate(across(where(is.numeric), ~round(., 4))),
  caption = "Model 3: Full Model (Odds Ratios)")
```

```{r model3-fit, results='asis'}
cat(sprintf("\n**Model 3 N:** %s\n", formatC(nobs(m3), big.mark = ",")))
```

## Model Comparison

```{r model-comparison, fig.width=10, fig.height=6}
# Extract race coefficients across models for comparison
race_terms <- c("race_ethHispanic", "race_ethNH Black", "race_ethNH Other")

comp_df <- bind_rows(
  m1_results %>% filter(term %in% race_terms) %>% mutate(model = "M1: Race Only"),
  m2_results %>% filter(term %in% race_terms) %>% mutate(model = "M2: + Generation"),
  m3_results %>% filter(term %in% race_terms) %>% mutate(model = "M3: Full Controls")
)

ggplot(comp_df, aes(x = term, y = estimate, color = model)) +
  geom_point(position = position_dodge(width = 0.5), size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high),
                position = position_dodge(width = 0.5), width = 0.2) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = c("#E74C3C", "#3498DB", "#2ECC71")) +
  labs(
    title = "Attenuation of Race Coefficients Across Models",
    subtitle = "Odds Ratios with 95% CI (reference: NH White)",
    x = NULL, y = "Odds Ratio",
    color = "Model"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 15, hjust = 1))
```

---

# Interaction Models

```{r interactions, results='asis'}
cat("## Generation × Education Interaction\n\n")

m_int_educ <- tryCatch(
  svyglm(completed_childless ~ race_eth + gen_status * college +
           currently_married + employed + poverty_pct + religion + age + wave_f,
         design = nsfg_design, family = quasibinomial()),
  error = function(e) { cat("Model did not converge:", conditionMessage(e), "\n"); NULL }
)

if (!is.null(m_int_educ)) {
  int_educ_results <- tidy(m_int_educ, conf.int = TRUE, exponentiate = TRUE)
  int_terms <- int_educ_results %>% filter(grepl("gen_status.*college|college.*gen_status", term))

  if (nrow(int_terms) > 0) {
    kable(int_terms %>% mutate(across(where(is.numeric), ~round(., 4))),
      caption = "Generation × College Interaction Terms")

    any_sig <- any(int_terms$p.value < 0.05)
    cat(sprintf("\n%s of the generation × education interactions reach significance.\n\n",
      ifelse(any_sig, "One or more", "None")))

    if (any_sig) {
      cat("This suggests that the relationship between college education and childlessness
differs by generational status, consistent with the assimilation hypothesis (H3): the
effect of education on fertility may be stronger among native-born women whose
decisions are more fully embedded in U.S. institutional contexts.\n\n")
    }
  }
}

cat("\n## Generation × Marriage Interaction\n\n")

m_int_mar <- tryCatch(
  svyglm(completed_childless ~ race_eth + gen_status * currently_married +
           college + employed + poverty_pct + religion + age + wave_f,
         design = nsfg_design, family = quasibinomial()),
  error = function(e) { cat("Model did not converge:", conditionMessage(e), "\n"); NULL }
)

if (!is.null(m_int_mar)) {
  int_mar_results <- tidy(m_int_mar, conf.int = TRUE, exponentiate = TRUE)
  int_terms_m <- int_mar_results %>%
    filter(grepl("gen_status.*married|married.*gen_status", term))

  if (nrow(int_terms_m) > 0) {
    kable(int_terms_m %>% mutate(across(where(is.numeric), ~round(., 4))),
      caption = "Generation × Marriage Interaction Terms")

    any_sig_m <- any(int_terms_m$p.value < 0.05)
    cat(sprintf("\n%s of the generation × marriage interactions reach significance.\n",
      ifelse(any_sig_m, "One or more", "None")))
  }
}
```

---

# Sensitivity: Fertility Intentions

```{r sensitivity, results='asis'}
if ("wants_children" %in% names(nsfg_40)) {
  cat("As a sensitivity check, we add fertility intentions (`wants_children`) to Model 3.
This variable allows us to distinguish, at least partially, between women who are
childless by choice and those who desired children but did not have them.\n\n")

  m_intent <- tryCatch(
    svyglm(completed_childless ~ race_eth + gen_status +
             college + currently_married + employed +
             poverty_pct + religion + age + wave_f + wants_children,
           design = nsfg_design, family = quasibinomial()),
    error = function(e) { cat("Model did not converge.\n"); NULL }
  )

  if (!is.null(m_intent)) {
    intent_results <- tidy(m_intent, conf.int = TRUE, exponentiate = TRUE)
    kable(intent_results %>%
      mutate(across(where(is.numeric), ~round(., 4))),
      caption = "Sensitivity Model: Full Model + Fertility Intentions (Odds Ratios)")
  }
} else {
  cat("Fertility intentions variable (`rwant`) not available in the pooled data.
This sensitivity analysis is omitted.\n")
}
```

---

# Discussion and Conclusion

```{r discussion, results='asis'}
cat("This analysis provides the direct nativity test that the SCF-based companion
papers could only hypothesize about. The sequential regression design allows us to
quantify exactly how much of the racial gradient in completed childlessness is
attributable to generational composition.\n\n")

cat("The key findings, evaluated against the three hypotheses:\n\n")

cat("**H1 (Hispanic-White gap and generation):** ")
if (nrow(m2_hisp) > 0) {
  cat(sprintf("The Hispanic coefficient attenuated by approximately %.0f%% when
generational status was added (Model 1 → Model 2). ",
    abs(pct_change_hisp)))
  if (abs(pct_change_hisp) > 20) {
    cat("This supports H1: a substantial portion of the Hispanic-White childlessness
gap is attributable to the higher share of first-generation immigrants in the
Hispanic population.\n\n")
  } else {
    cat("The attenuation was modest, suggesting that generational status alone does not
fully account for the Hispanic-White gap. Other factors — including cultural norms
that persist across generations — may also contribute.\n\n")
  }
}

cat("**H2 (Black-White gap and generation):** ")
if (nrow(m2_black) > 0) {
  cat(sprintf("The Black coefficient changed by approximately %.0f%% between Models 1
and 2. ",
    abs(pct_change_black)))
  if (abs(pct_change_black) < 15) {
    cat("This supports H2: the Black-White childlessness difference is not substantially
explained by generational status, consistent with the predominantly native-born
composition of the Black population.\n\n")
  } else {
    cat("Contrary to H2, generational status contributes meaningfully to the
Black-White gap, suggesting that Black immigrant fertility patterns play a larger
role than the demographic composition of the Black population would predict.\n\n")
  }
}

cat("**H3 (Assimilation and interaction effects):** The interaction models tested
whether education and marriage effects on childlessness differ by generation. ")
if (!is.null(m_int_educ)) {
  int_sig <- int_educ_results %>%
    filter(grepl("gen_status.*college", term), p.value < 0.05)
  if (nrow(int_sig) > 0) {
    cat("At least one education × generation interaction was significant, providing
preliminary support for H3: the fertility-reducing effect of education may be
stronger among later-generation women who are more fully integrated into U.S.
institutional contexts.\n\n")
  } else {
    cat("No significant education × generation interactions were detected, suggesting
that the relationship between education and childlessness is similar across
generations — or that the sample lacks power to detect differential effects.\n\n")
  }
}

cat("**Connection to the thesis.** This paper closes the analytical loop opened by
the SCF-based companion papers. The first paper showed that parenthood drives credit
card debt. The second paper showed that the financial balance sheet — particularly
retirement assets and credit access — shapes co-resident childlessness, with racial
differences fully mediated by financial and demographic controls. This third paper
tests the hypothesis that emerged from that mediation: that nativity composition
underlies the racial gradient. Together, the three papers demonstrate that the
economic context of family formation operates through specific financial mechanisms
(balance sheet composition), is shaped by demographic factors (marriage, education),
and reflects population-level compositional effects (immigration and generational
assimilation) that single-dataset analyses cannot fully capture.\n")
```

---

# Limitations

**Women only.** This analysis is restricted to female respondents. Male fertility is
harder to measure reliably, and the NSFG's male questionnaire has different variable
structures. Extending the analysis to men would require a separate investigation.

**Age range and wave availability.** The 40–49 age window for completed fertility is
only available in the 2015–2017 and 2017–2019 waves (which expanded the upper age limit
from 44 to 49). Earlier waves are excluded, limiting pooled sample size.

**Parental nativity variables.** The availability and coding of parental birthplace
variables may differ across waves. If these variables are not available in the public-use
files, the generational status variable falls back to a simple foreign-born/native-born
dichotomy, which cannot distinguish second from third-plus generation.

**Cross-sectional design.** Childlessness at 40–49 is near-complete but not perfectly
complete fertility. A small number of women will bear children after 49. All findings
are correlational.

**Survey weights for pooled analysis.** We halve each wave's weight for pooling, which
is the standard approach recommended by NCHS. However, this assumes equal representation
across waves and does not account for potential overlap in the continuous interviewing
design.

---

# References

Adserà, A., & Ferrer, A. (2015). Immigrants and demography. In *Handbook of the
Economics of International Migration* (Vol. 1A, pp. 315–374). North-Holland.

Camarota, S. A., & Zeigler, K. (2025). The fertility of immigrants and natives in the
United States, 2023. Center for Immigration Studies.

Guzzo, K. B., & Loo, J. (2023). Number of children to women aged 40–44, 1980–2022.
*Family Profiles* (FP-23-29). National Center for Family & Marriage Research.

Hill, L. E., & Johnson, H. P. (2004). Fertility changes among immigrants. *Social
Science Quarterly*, 85(3), 811–826.

Livingston, G. (2015). Childlessness falls, family size grows among highly educated
women. Pew Research Center.

Osterman, M. J. K., et al. (2023). Births: Final data for 2021. *National Vital
Statistics Reports*, 72(1), 1–53.

Parrado, E. A. (2011). How high is Hispanic/Mexican fertility in the United States?
*Demography*, 48(3), 1059–1080.

Parrado, E. A., & Morgan, S. P. (2008). Intergenerational fertility among Hispanic
women. *Demography*, 45(3), 651–671.

Pew Research Center. (2015). Childlessness falls, family size grows among highly
educated women.

Zill, N. (2025). Four million American women reach ages 40 to 50 without children.
Institute for Family Studies.

---

# Appendix: Technical Notes

**Survey Design.** The NSFG uses a stratified, multistage probability design with
oversampling of Hispanic, Black, and teenage respondents. We use `svydesign()` with
`SEST` (stratum) and `SECU` (cluster) and the appropriate wave-specific weight, halved
for pooling.

**Pooling Weights.** Following NCHS guidance, we divide each wave's weight by the number
of pooled waves (2) to avoid double-counting the target population.

**Generational Status.** Constructed from three variables: respondent's birthplace
(`BRNOUT`), mother's birthplace, and father's birthplace. First generation = respondent
born outside U.S. Second generation = respondent born in U.S. with at least one
foreign-born parent. Third-plus generation = respondent and both parents born in U.S.

**Completed Childlessness.** Defined as `PARITY == 0` among women aged 40–49. `PARITY`
counts live births ever had, not co-resident children, resolving the measurement
limitation of the SCF analysis.

```{r session-info, results='asis'}
cat("\n\n**Session Information**\n\n")
cat(sprintf("R version: %s\n\n", R.version.string))
cat(sprintf("Platform: %s\n\n", R.version$platform))
si <- sessionInfo()
cat(sprintf("Key packages: survey %s, haven %s, tidyverse %s\n",
  packageVersion("survey"), packageVersion("haven"), packageVersion("tidyverse")))
```
