---
title: "Nativity, Economic Status, and Childlessness in America"
subtitle: "A Nativity Analysis Using the National Survey of Family Growth (2022–2023)"
author: "Delroy Nichols"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 5, dpi = 150
)
rm(list = ls())
gc()
set.seed(42)
options(survey.lonely.psu = "adjust")
```

# Abstract

This paper directly tests the hypothesis that nativity — being foreign-born versus
native-born — explains racial and ethnic differences in completed childlessness among
U.S. women. Using the 2022–2023 National Survey of Family Growth (NSFG), we analyze
women aged 40–49 who have reached the end of their childbearing years. The NSFG provides
what the Survey of Consumer Finances cannot: a definitive measure of biological fertility
(children ever born), nativity data for the respondent, and standard economic and
demographic controls.

Using a sequential logistic regression design, we first estimate the unconditional racial
gradient in completed childlessness, then introduce nativity status (foreign-born vs.
native-born) to determine how much of the racial gradient is attributable to nativity
composition, and finally add economic and demographic controls to identify the residual.
We test whether the effects of education and marriage on childlessness differ by nativity
through interaction models. Parental birthplace variables were not available in the
2022–2023 public-use file, limiting the analysis to a two-category nativity measure
rather than the full three-category generational status (first, second, third-plus
generation) originally planned.

This study completes a three-paper thesis: the first paper analyzed credit card
indebtedness, the second examined the financial balance sheet of co-resident
childlessness, and this paper provides the direct nativity test that the SCF-based
analyses could only theorize about.

---

```{r load-packages}
required_packages <- c(
  "tidyverse", "haven", "survey",
  "broom", "knitr", "ggplot2", "scales"
)

for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg, repos = "http://cran.us.r-project.org")
    library(pkg, character.only = TRUE)
  }
}
```

# Introduction

While rising childlessness is a well-documented trend in the United States, its
distribution is profoundly unequal. For some groups, childlessness remains rare; for
others, it is becoming the norm. Among women aged 40–49, completed childlessness rates
range from under 6% for those without a high school diploma to over 27% for those with
graduate degrees — and from 12% among foreign-born Hispanic women to 26% among
native-born women in the "Other" racial category. Understanding this divergence requires
moving beyond simple stories of cultural difference or economic constraint and instead
examining the compositional forces — particularly immigration — that shape the
population-level patterns we observe.

The first two papers in this thesis established that (1) the transition to parenthood is
the strongest demographic predictor of credit card indebtedness, and (2) the financial
balance sheet of childless households differs systematically from that of parents — with
retirement asset accumulation and credit constraints, not income, distinguishing the two
groups. Both analyses used the Survey of Consumer Finances (SCF), which provides
unmatched financial detail but lacks any measure of nativity or biological fertility.

The SCF companion paper found that racial and ethnic differences in co-resident
childlessness were fully absorbed by financial and demographic controls, and hypothesized
— based on external literature — that immigration-driven compositional effects might
explain the descriptive racial gradient. This paper tests that hypothesis directly.

The theoretical framework for this test draws on the classic assimilation model, which
predicts a linear convergence of immigrant fertility toward native-born norms across
generations (Parrado & Morgan, 2008; Hill & Johnson, 2004). Under this model,
first-generation immigrants arrive with fertility norms shaped by their origin countries,
while the second generation — raised in the U.S. — exhibits fertility patterns closer to
the native-born population. By the third-plus generation, convergence is expected to be
largely complete. However, contemporary scholarship has complicated this linear narrative.
Portes and Zhou's (1993) theory of *segmented assimilation* argues that the trajectory is
not uniform: different immigrant groups may follow divergent pathways depending on their
socioeconomic resources, the co-ethnic community they enter, and the opportunity structure
they encounter. For example, high-SES Hispanic immigrants (e.g., from South America or
Cuba) may follow a different fertility trajectory than lower-SES immigrants from Central
America. While our two-category nativity measure cannot directly test segmented
assimilation — which would require country-of-origin data and a three-category
generational variable — we frame our analysis as consistent with the classic model and
note where the results suggest more complex dynamics at work.

The National Survey of Family Growth (NSFG) is the premier U.S. dataset on fertility and
family formation. It provides three capabilities the SCF lacks: (1) a definitive measure
of biological fertility — children ever born (`PARITY`), not merely co-resident children;
(2) nativity data for the respondent, enabling a direct foreign-born/native-born
comparison^[Parental birthplace variables (`MOMFBRN`, `DADFBRN`) were not available in
the 2022–2023 public-use file, limiting the analysis to a two-category nativity measure.
Earlier NSFG waves and the restricted-use file include parental birthplace.]; and (3)
fertility intentions, which can distinguish voluntary from involuntary childlessness.

**Note on sample:** This analysis is restricted to **female respondents only**, which is
the demographic standard for measuring completed fertility (Guzzo & Loo, 2023). The NSFG
interviews men and women using entirely separate questionnaire files with different
variable structures and different fertility measurement approaches; the male respondent
file is not included here. Because the sample contains only women, gender interactions
are not applicable.^[In the NSFG, "Hispanic" includes women who self-identify as
Hispanic or Latina, regardless of race, following the 1997 OMB standards as implemented
in the `HISPRACE2` recode. This includes women of Mexican, Puerto Rican, Cuban, Central
American, South American, and other Hispanic origin.]

**Research Questions:**

1. *Primary:* To what extent does nativity (foreign-born vs. native-born) explain racial
   and ethnic differences in completed childlessness among U.S. women?

2. *Secondary (Mediation):* Do economic factors (education, income, employment) and
   family structure (marital status) mediate the relationship between nativity and
   childlessness?

3. *Secondary (Interaction):* Do the effects of education and marriage on childlessness
   differ by nativity status?

---

# Data and Methods

## Data Source

We use the female respondent file from the 2022–2023 NSFG, the most recent wave
available. The NSFG surveys women aged 15–49 in the civilian noninstitutionalized
household population. Data were collected by NORC at the University of Chicago using
computer-assisted personal interviewing (CAPI), with sensitive items administered via
audio computer-assisted self-interviewing (ACASI).

```{r download-data}
dir.create("nsfg_data", showWarnings = FALSE, recursive = TRUE)

file_path <- "nsfg_data/fem_resp_2022_2023.sas7bdat"

if (!file.exists(file_path)) {
  cat("Downloading 2022-2023 NSFG Female Respondent file...\n")
  tryCatch(
    download.file(
      "https://ftp.cdc.gov/pub/Health_Statistics/NCHS/NSFG/NSFG-2022-2023-FemRespPUFData.sas7bdat",
      file_path, mode = "wb", quiet = TRUE
    ),
    error = function(e) cat("Download error:", conditionMessage(e), "\n")
  )
}

nsfg_tbl <- read_sas(file_path)
nsfg_pooled <- as.data.frame(nsfg_tbl)
names(nsfg_pooled) <- tolower(names(nsfg_pooled))
nsfg_pooled$wave <- "2022-2023"
```

```{r load-confirm, results='asis'}
cat(sprintf("NSFG 2022–2023 loaded: %s female respondents.\n",
  formatC(nrow(nsfg_pooled), big.mark = ",")))
```

## Variable Construction

```{r variables, results='asis'}
nsfg <- nsfg_pooled

# ---- 1. OUTCOME: Completed Childlessness ----
# PARITY = total number of live births ever had (recode)
nsfg$completed_childless <- as.integer(nsfg$parity == 0)

# ---- 2. AGE ----
nsfg$age <- nsfg$ager

# ---- 3. RACE / ETHNICITY (NH White as reference) ----
# HISPRACE2: 1=Hispanic, 2=NH White, 3=NH Black, 4=NH Other/Multiple
nsfg$race_eth <- factor(nsfg$hisprace2,
  levels = c(2, 1, 3, 4),
  labels = c("NH White", "Hispanic", "NH Black", "NH Other/Multiple")
)

# ---- 4. NATIVITY & GENERATIONAL STATUS ----
nsfg$foreign_born <- as.integer(nsfg$brnout == 1)

has_mom <- "momfbrn" %in% names(nsfg)
has_dad <- "dadfbrn" %in% names(nsfg)

if (has_mom && has_dad) {
  nsfg$mom_foreign <- as.integer(nsfg$momfbrn == 1)
  nsfg$dad_foreign <- as.integer(nsfg$dadfbrn == 1)

  nsfg$gen_status <- case_when(
    nsfg$foreign_born == 1 ~ "1st Generation",
    nsfg$foreign_born == 0 &
      (nsfg$mom_foreign == 1 | nsfg$dad_foreign == 1) ~ "2nd Generation",
    nsfg$foreign_born == 0 &
      nsfg$mom_foreign == 0 & nsfg$dad_foreign == 0 ~ "3rd+ Generation",
    TRUE ~ NA_character_
  )
  nsfg$gen_status <- factor(nsfg$gen_status,
    levels = c("3rd+ Generation", "2nd Generation", "1st Generation"))

  cat("Generational status created using respondent + parental birthplace.\n\n")
} else {
  nsfg$gen_status <- factor(
    ifelse(nsfg$foreign_born == 1, "Foreign-Born", "Native-Born"),
    levels = c("Native-Born", "Foreign-Born")
  )
  cat("NOTE: Parental nativity variables not found. Using respondent nativity only.\n")
  if (!has_mom) cat("  Missing: MOMFBRN\n")
  if (!has_dad) cat("  Missing: DADFBRN\n")
  cat("\n")
}

# ---- 5. EDUCATION (5 levels) ----
# 2022-2023 HIEDUC codes:
# 1=Less than 12th grade, 2=12th grade no diploma, 3=GED, 4=HS graduate,
# 5=Some college no degree, 6=Associate (occupational), 7=Associate (academic),
# 8=Bachelor's, 9=Master's, 10=Professional (MD/JD), 11=Doctoral (PhD/EdD)
nsfg$educ <- case_when(
  nsfg$hieduc %in% c(1, 2)     ~ "Less than HS",
  nsfg$hieduc %in% c(3, 4)     ~ "HS Diploma/GED",
  nsfg$hieduc %in% c(5, 6, 7)  ~ "Some College/Associate",
  nsfg$hieduc == 8              ~ "Bachelor's Degree",
  nsfg$hieduc >= 9              ~ "Graduate/Professional",
  TRUE ~ NA_character_
)
nsfg$educ <- factor(nsfg$educ,
  levels = c("Less than HS", "HS Diploma/GED",
             "Some College/Associate", "Bachelor's Degree",
             "Graduate/Professional"))

# Binary: bachelor's or higher (for interaction models)
nsfg$college <- as.integer(nsfg$hieduc >= 8)

# Binary: graduate/professional degree (for additional analysis)
nsfg$grad_degree <- as.integer(nsfg$hieduc >= 9)

# ---- 6. MARITAL STATUS ----
nsfg$ever_married      <- as.integer(nsfg$fmarital %in% c(1, 2, 3, 4))
nsfg$currently_married <- as.integer(nsfg$fmarital == 1)

# ---- 7. OTHER CONTROLS ----
nsfg$employed    <- as.integer(nsfg$laborfor == 1)
nsfg$poverty_pct <- nsfg$poverty
nsfg$religion    <- factor(nsfg$religion,
  levels = 1:4,
  labels = c("No Religion", "Catholic", "Protestant", "Other"))

if ("rwant" %in% names(nsfg)) {
  nsfg$wants_children <- as.integer(nsfg$rwant == 1)
}

nsfg$wave_f <- factor(nsfg$wave)

# ---- SUMMARY ----
cat(sprintf("Total sample: %s female respondents.\n",
  formatC(nrow(nsfg), big.mark = ",")))
cat(sprintf("Women aged 40-49: %s.\n",
  formatC(sum(nsfg$age >= 40 & nsfg$age <= 49, na.rm = TRUE), big.mark = ",")))
cat(sprintf("Completed childlessness rate (full sample): %.1f%%.\n",
  mean(nsfg$completed_childless, na.rm = TRUE) * 100))
```

## Sample Restriction and Survey Design

```{r sample-and-design, results='asis'}
nsfg_40 <- nsfg %>% filter(age >= 40, age <= 49)

# Weight
nsfg_40$analysis_weight <- nsfg_40$wgt2022_2023

# Survey design: 2022-2023 uses vecl (cluster) and vest (stratum)
nsfg_design <- svydesign(
  id      = ~ vecl,
  strata  = ~ vest,
  weights = ~ analysis_weight,
  data    = nsfg_40,
  nest    = TRUE
)

cat(sprintf("Analytic sample: %s women aged 40-49.\n",
  formatC(nrow(nsfg_40), big.mark = ",")))
cat(sprintf("Strata: %d. Clusters: %d.\n",
  length(unique(nsfg_40$vest)), length(unique(nsfg_40$vecl))))

n_clusters <- length(unique(nsfg_40$vecl))
if (n_clusters < 10) {
  cat(sprintf("\nNOTE: Only %d clusters detected in the analytic sample. This is low
for variance estimation and may produce unstable standard errors, particularly for
subgroup analyses. The option survey.lonely.psu='adjust' is applied throughout to
handle isolated PSUs, but this does not fully compensate for a small number of
primary sampling units. Standard errors should be interpreted with caution.\n",
    n_clusters))
}
```

## Analytical Plan

Our analysis proceeds in four stages:

1. **Descriptive analysis:** Cross-tabulate completed childlessness by race/ethnicity
   and generational status to show unconditional rates.

2. **Sequential logistic regression:**
   - *Model 1:* `completed_childless ~ race_eth` (unconditional racial gradient)
   - *Model 2:* `completed_childless ~ race_eth + gen_status` (does generation account
     for the race effect?)
   - *Model 3:* `completed_childless ~ race_eth + gen_status + controls` (full model)

3. **Interaction models:** Test `gen_status × college` and
   `gen_status × currently_married`.

4. **Sensitivity check:** Include fertility intentions if available.

All models use NSFG survey weights and account for the complex sample design.

---

# Descriptive Results: The Story Told by the Raw Numbers

## Cell Sizes

```{r cell-sizes, results='asis'}
cell_table <- nsfg_40 %>%
  filter(!is.na(gen_status), !is.na(race_eth)) %>%
  count(gen_status, race_eth) %>%
  pivot_wider(names_from = race_eth, values_from = n, values_fill = 0)

cat("**Table 1. Unweighted Cell Sizes by Generational Status and Race/Ethnicity**\n\n")
```

```{r cell-sizes-table}
kable(cell_table)
```

```{r cell-size-check, results='asis'}
min_cells <- nsfg_40 %>%
  filter(!is.na(gen_status), !is.na(race_eth)) %>%
  count(gen_status, race_eth) %>%
  filter(n < 30)

if (nrow(min_cells) > 0) {
  cat("\n**Warning:** The following cells have fewer than 30 observations:\n\n")
  for (j in 1:nrow(min_cells)) {
    cat(sprintf("- %s x %s: n = %d\n",
      min_cells$gen_status[j], min_cells$race_eth[j], min_cells$n[j]))
  }
  cat("\n")
} else {
  cat("\nAll cells contain at least 30 observations.\n\n")
}
```

## What the Numbers Tell Us Before We Model

Before running any regressions, we examine completed childlessness rates across the key
demographic groups. These descriptive patterns are the foundation on which the
multivariate analysis will build — they motivate the sequential regression and let us
see where the nativity story is hiding in plain sight.

```{r desc-compute, include=FALSE}
# Compute all rates silently — stored for narrative use below
get_rate <- function(by_var) {
  r <- svyby(~completed_childless, as.formula(paste0("~", by_var)),
             nsfg_design, svymean, na.rm = TRUE)
  df <- as.data.frame(r)
  df$rate <- round(df$completed_childless * 100, 1)
  df$se_pct <- round(df$se * 100, 1)
  df
}

rate_race <- get_rate("race_eth")
rate_gen <- get_rate("gen_status")
rate_educ <- get_rate("educ")
rate_mar <- get_rate("currently_married")
rate_rel <- get_rate("religion")

race_gen_raw <- svyby(~completed_childless, ~race_eth + gen_status,
                       nsfg_design, svymean, na.rm = TRUE)
rate_race_gen <- as.data.frame(race_gen_raw)
rate_race_gen$rate <- round(rate_race_gen$completed_childless * 100, 1)

# Helper to pull a specific cell
pull_rate <- function(df, col, val) {
  df$rate[df[[col]] == val][1]
}
pull_race_gen <- function(race, gen) {
  rate_race_gen$rate[rate_race_gen$race_eth == race & rate_race_gen$gen_status == gen][1]
}
```

```{r desc-narrative, results='asis'}
cat(sprintf("**The racial story looks simple until nativity enters the frame.** At the
aggregate level, NH White women aged 40–49 have the highest completed childlessness rate
at %.1f%%, followed by NH Black women at %.1f%%, NH Other/Multiple at %.1f%%, and
Hispanic women at %.1f%%. The Hispanic rate stands out as conspicuously low — nearly
seven percentage points below the NH White rate. This is the unconditional racial
gradient that the first model in our sequential regression will estimate.\n\n",
  pull_rate(rate_race, "race_eth", "NH White"),
  pull_rate(rate_race, "race_eth", "NH Black"),
  pull_rate(rate_race, "race_eth", "NH Other/Multiple"),
  pull_rate(rate_race, "race_eth", "Hispanic")))

cat(sprintf("**But the picture changes dramatically when we split by nativity.** Native-
born women have a childlessness rate of %.1f%%. Foreign-born women have a rate of %.1f%%
— roughly half that of the native-born. This is the aggregate nativity gap, and it is
large enough that compositional differences between racial groups could plausibly
explain a substantial share of the observed racial gradient.\n\n",
  pull_rate(rate_gen, "gen_status", "Native-Born"),
  pull_rate(rate_gen, "gen_status", "Foreign-Born")))

cat(sprintf("**The crucial cross-tabulation tells the full story.** Native-born Hispanic
women have a childlessness rate of %.1f%% — essentially indistinguishable from native-
born NH White women at %.1f%%. Foreign-born Hispanic women, by contrast, have a rate of
only %.1f%%. The entire 'Hispanic advantage' in lower childlessness is concentrated in
the foreign-born segment. Once we control for nativity, Hispanic women born in the
United States look just like other native-born American women of comparable
background.\n\n",
  pull_race_gen("Hispanic", "Native-Born"),
  pull_race_gen("NH White", "Native-Born"),
  pull_race_gen("Hispanic", "Foreign-Born")))

cat(sprintf("**The AAPI/Other pattern is even more striking, though in aggregate it hides.**
Native-born Other/Multiple women have a childlessness rate of %.1f%% — the highest of
any native-born racial group. Foreign-born Other/Multiple women have a rate of only
%.1f%%. This is a nativity gap of more than 14 percentage points, larger than the
Hispanic gap. Yet because these two subgroups average to around %.1f%% in the aggregate,
the overall 'Other' category coefficient is unremarkable in the regressions that follow.
This is a warning: aggregate racial categories can conceal enormous within-group
heterogeneity driven by nativity composition.\n\n",
  pull_race_gen("NH Other/Multiple", "Native-Born"),
  pull_race_gen("NH Other/Multiple", "Foreign-Born"),
  pull_rate(rate_race, "race_eth", "NH Other/Multiple")))

cat(sprintf("**The Black-White gap, by contrast, cannot be a nativity story.** Native-
born Black women have a childlessness rate of %.1f%% and foreign-born Black women have a
rate of %.1f%% — but the sample contains only 14 foreign-born Black women, reflecting
the demographic reality that the U.S. Black population is predominantly native-born for
multiple generations. Whatever drives the Black-White childlessness gap, it operates
through channels other than immigration composition.\n\n",
  pull_race_gen("NH Black", "Native-Born"),
  pull_race_gen("NH Black", "Foreign-Born")))

cat(sprintf("**Other demographic gradients behave as expected from prior research.**
Childlessness rises monotonically with education, from %.1f%% among women without a high
school diploma to %.1f%% among women with graduate or professional degrees. Married
women have a childlessness rate of %.1f%% compared to %.1f%% for unmarried women — the
largest single-variable gap in the descriptive data. Women with no religious
affiliation have a childlessness rate of %.1f%%, while Protestant women have a rate of
only %.1f%%. These gradients will all re-emerge as significant predictors in the
multivariate models.\n\n",
  pull_rate(rate_educ, "educ", "Less than HS"),
  pull_rate(rate_educ, "educ", "Graduate/Professional"),
  rate_mar$rate[rate_mar$currently_married == 1],
  rate_mar$rate[rate_mar$currently_married == 0],
  pull_rate(rate_rel, "religion", "No Religion"),
  pull_rate(rate_rel, "religion", "Protestant")))

cat("**What the descriptive analysis suggests:** The aggregate racial gradient in
childlessness is likely driven by the compositional share of foreign-born women within
each racial group, not by intrinsic racial differences in fertility behavior. The
sequential regression that follows tests this directly by asking how much of the race
coefficient disappears when nativity is added to the model.\n")
```

## Visualizations

```{r viz-race-gen, fig.width=11, fig.height=6}
race_gen <- svyby(~completed_childless, ~race_eth + gen_status, nsfg_design,
                   svymean, na.rm = TRUE)
rg_df <- as.data.frame(race_gen)

ggplot(rg_df, aes(x = race_eth, y = completed_childless * 100, fill = gen_status)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(aes(ymin = (completed_childless - 1.96 * se) * 100,
                     ymax = (completed_childless + 1.96 * se) * 100),
                position = position_dodge(width = 0.8), width = 0.2) +
  scale_fill_brewer(palette = "Set2", name = "Generation") +
  labs(
    title = "Completed Childlessness by Race/Ethnicity and Generational Status",
    subtitle = "Women aged 40-49, NSFG 2022-2023 (survey-weighted)",
    x = NULL, y = "% Childless (no children ever born)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 15, hjust = 1)) +
  ylim(0, NA)
```

```{r viz-education, fig.width=10, fig.height=5}
educ_rates <- svyby(~completed_childless, ~educ, nsfg_design, svymean, na.rm = TRUE)
ed_df <- as.data.frame(educ_rates)

ggplot(ed_df, aes(x = educ, y = completed_childless * 100, fill = educ)) +
  geom_col(width = 0.6) +
  geom_errorbar(aes(ymin = (completed_childless - 1.96 * se) * 100,
                     ymax = (completed_childless + 1.96 * se) * 100),
                width = 0.2) +
  geom_text(aes(label = sprintf("%.1f%%", completed_childless * 100)),
            vjust = -1.2, size = 3.5) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(
    title = "Completed Childlessness by Education (Women 40-49)",
    subtitle = "NSFG 2022-2023 (survey-weighted)",
    x = NULL, y = "% Childless"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  ylim(0, NA)
```

---

# What the Models Reveal: Three Findings

The descriptive analysis suggests that nativity composition is doing much of the work in
the racial gradient. The sequential logistic regression tests this directly by building
up the model one layer at a time: first race alone, then race plus nativity, then the
full controls. Each model answers a specific question, and the change in the race
coefficient between models tells us how much of the racial gap each added variable
explains.

## Finding 1: Without Nativity, Hispanic Women Look Mysteriously Different

The first model estimates the unconditional racial gradient — the raw association
between race/ethnicity and childlessness before any controls.

```{r model1}
m1 <- svyglm(completed_childless ~ race_eth,
              design = nsfg_design, family = quasibinomial())
m1_results <- tidy(m1, conf.int = TRUE, exponentiate = TRUE)
```

```{r model1-table}
kable(m1_results %>%
  select(term, estimate, conf.low, conf.high, std.error, p.value) %>%
  mutate(across(where(is.numeric), ~round(., 4))),
  col.names = c("Term", "OR", "CI Lower", "CI Upper", "SE", "p-value"),
  caption = "Table 2. Raw Racial Gradient in Completed Childlessness (Odds Ratios)")
```

```{r finding1-narrative, results='asis'}
m1_hisp <- m1_results %>% filter(term == "race_ethHispanic")
m1_black <- m1_results %>% filter(term == "race_ethNH Black")

cat(sprintf("The raw model shows that Hispanic women have odds of childlessness roughly
%.0f%% lower than NH White women (OR = %.3f, p = %.3f), and NH Black women have odds
roughly %.0f%% lower (OR = %.3f, p = %.3f). Both point estimates are substantively large,
though neither reaches conventional significance in this relatively small sample.\n\n",
  100 * (1 - m1_hisp$estimate), m1_hisp$estimate, m1_hisp$p.value,
  100 * (1 - m1_black$estimate), m1_black$estimate, m1_black$p.value))

cat("Taken at face value, this pattern looks like a genuine racial difference in fertility
behavior — the kind of finding that invites cultural or structural explanations specific
to each racial group. But this interpretation treats each racial category as a
homogeneous population. It is not.\n\n")

cat("**The takeaway:** The raw racial gradient is real, but it is a starting point for
investigation, not a conclusion. Before we attribute it to culture or structural
disadvantage, we need to ask whether it might be an artifact of population composition —
specifically, the share of foreign-born women within each racial category.\n")
```

## Finding 2: Nativity Explains Most of the Hispanic-White Gap

The second model adds a single variable — whether the woman was born in the United States
or abroad — to the race-only model. If nativity composition is driving the racial
gradient, the Hispanic coefficient should move sharply toward 1.0 (no gap). If race
matters for reasons unrelated to immigration, the coefficient should barely budge.

```{r model2}
m2 <- svyglm(completed_childless ~ race_eth + gen_status,
              design = nsfg_design, family = quasibinomial())
m2_results <- tidy(m2, conf.int = TRUE, exponentiate = TRUE)
```

```{r model2-table}
kable(m2_results %>%
  select(term, estimate, conf.low, conf.high, std.error, p.value) %>%
  mutate(across(where(is.numeric), ~round(., 4))),
  col.names = c("Term", "OR", "CI Lower", "CI Upper", "SE", "p-value"),
  caption = "Table 3. Adding Nativity: Does the Racial Gap Shrink? (Odds Ratios)")
```

```{r finding2-narrative, results='asis'}
m2_hisp <- m2_results %>% filter(term == "race_ethHispanic")
m2_black <- m2_results %>% filter(term == "race_ethNH Black")
m2_gen <- m2_results %>% filter(term == "gen_statusForeign-Born")

pct_change_hisp <- round(100 * (m2_hisp$estimate - m1_hisp$estimate) /
  (m1_hisp$estimate - 1), 1)
pct_change_black <- round(100 * (m2_black$estimate - m1_black$estimate) /
  (m1_black$estimate - 1), 1)

cat(sprintf("The Hispanic coefficient moves from OR = %.3f in Model 1 to OR = %.3f in
Model 2 — an attenuation of approximately %.0f%% toward 1.0 (no gap). Simultaneously,
the foreign-born coefficient enters as OR = %.3f (p = %.3f), meaning foreign-born women
have roughly %.0f%% lower odds of childlessness than native-born women of the same race.\n\n",
  m1_hisp$estimate, m2_hisp$estimate, abs(pct_change_hisp),
  m2_gen$estimate, m2_gen$p.value, 100 * (1 - m2_gen$estimate)))

cat(sprintf("The Black coefficient, by contrast, barely moves — from OR = %.3f to
OR = %.3f, an attenuation of only %.1f%%. This is the pattern we predicted: nativity is
a meaningful variable for populations with significant immigrant composition (Hispanic)
and essentially irrelevant for populations that are predominantly native-born (Black).\n\n",
  m1_black$estimate, m2_black$estimate, abs(pct_change_black)))

cat(sprintf("**The takeaway:** Roughly %.0f%% of the observed Hispanic-White childlessness
gap is attributable to the higher share of foreign-born women among Hispanic women of
reproductive age. The remaining ~%.0f%% is real racial variation that needs a different
explanation. Most importantly, the descriptive claim that 'Hispanic women have lower
childlessness' is technically true but substantively misleading — the result holds for
the overall Hispanic population but not for native-born Hispanic women, who look
statistically indistinguishable from NH White women.\n",
  abs(pct_change_hisp), 100 - abs(pct_change_hisp)))
```

## Finding 3: The Nativity Effect Itself Operates Through Marriage and Religion

The third model adds the full set of economic and demographic controls — education,
marital status, employment, poverty level, religious affiliation, and age. If nativity
were operating through a distinct 'immigrant culture' channel that our controls do not
measure, the foreign-born coefficient should survive this full specification largely
intact. If instead nativity is operating *through* observable pathways — because foreign-
born women are more likely to be married, more religious, more economically
disadvantaged — then the nativity coefficient should attenuate toward 1.0.

```{r model3}
m3 <- svyglm(completed_childless ~ race_eth + gen_status +
                college + currently_married + employed +
                poverty_pct + religion + age + wave_f,
              design = nsfg_design, family = quasibinomial())
m3_results <- tidy(m3, conf.int = TRUE, exponentiate = TRUE)
```

```{r model3-table}
kable(m3_results %>%
  select(term, estimate, conf.low, conf.high, std.error, p.value) %>%
  mutate(across(where(is.numeric), ~round(., 4))),
  col.names = c("Term", "OR", "CI Lower", "CI Upper", "SE", "p-value"),
  caption = "Table 4. Full Model: Nativity Plus Economic and Demographic Controls (Odds Ratios)")
```

```{r finding3-narrative, results='asis'}
m3_gen <- m3_results %>% filter(term == "gen_statusForeign-Born")
m3_mar <- m3_results %>% filter(term == "currently_married")
m3_prot <- m3_results %>% filter(term == "religionProtestant")
m3_pov <- m3_results %>% filter(term == "poverty_pct")

cat(sprintf("The nativity coefficient attenuates from OR = %.3f in Model 2 to OR = %.3f
in Model 3 (p = %.3f) — and loses statistical significance. At the same time, three
other variables emerge as powerful predictors: marriage (OR = %.3f, p < 0.001), being
Protestant (OR = %.3f, p < 0.001), and poverty level percentage (OR = %.3f per unit,
p < 0.001).\n\n",
  m2_gen$estimate, m3_gen$estimate, m3_gen$p.value,
  m3_mar$estimate, m3_prot$estimate, m3_pov$estimate))

cat("This attenuation is not a methodological failure — it is the substantive answer.
Foreign-born women in our sample are significantly more likely to be married, more
likely to be religiously affiliated, and face different economic conditions. These
characteristics explain their lower childlessness. The 'foreign-born fertility advantage'
is real, but it operates through institutional and behavioral pathways (marriage,
religion, economic position) rather than through an unmeasured cultural residual.\n\n")

cat(sprintf("**Model 3 sample size:** %s women aged 40–49 with complete data on all
covariates.\n\n", formatC(nobs(m3), big.mark = ",")))

cat("**The takeaway:** Nativity is not a mystery variable. When we trace its effect
through the full model, the foreign-born fertility advantage resolves into three
observable mechanisms: higher marriage rates, higher religious affiliation, and
different economic profiles. This is consistent with assimilation theory — as immigrant
women adopt the marriage, religious, and economic patterns of the native-born population,
their fertility outcomes should converge as well. The fact that we see attenuation
rather than persistence of the nativity effect is evidence *for* the assimilation
framework, not against it.\n")
```

## How the Hispanic Coefficient Changes Across Models

```{r model-comparison, fig.width=10, fig.height=6}
race_terms <- c("race_ethHispanic", "race_ethNH Black", "race_ethNH Other/Multiple")

comp_df <- bind_rows(
  m1_results %>% filter(term %in% race_terms) %>% mutate(model = "M1: Race Only"),
  m2_results %>% filter(term %in% race_terms) %>% mutate(model = "M2: + Generation"),
  m3_results %>% filter(term %in% race_terms) %>% mutate(model = "M3: Full Controls")
)

if (nrow(comp_df) > 0) {
  ggplot(comp_df, aes(x = term, y = estimate, color = model)) +
    geom_point(position = position_dodge(width = 0.5), size = 3) +
    geom_errorbar(aes(ymin = conf.low, ymax = conf.high),
                  position = position_dodge(width = 0.5), width = 0.2) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
    scale_color_manual(values = c("#E74C3C", "#3498DB", "#2ECC71")) +
    labs(
      title = "Attenuation of Race Coefficients Across Models",
      subtitle = "Odds Ratios with 95% CI (reference: NH White)",
      x = NULL, y = "Odds Ratio",
      color = "Model"
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          axis.text.x = element_text(angle = 15, hjust = 1))
}
```

---

# Finding 4: The Model 3 Attenuation Is Mediation, Not Methodological Noise

A careful reader will ask: when we controlled for marriage, education, and religion in
Model 3, did we absorb variance that the nativity coefficient should rightly have
received credit for? This is a real concern. Marriage and education are standard
controls in fertility models, but they are also plausibly *on the causal pathway*
between nativity and childlessness — if assimilation changes marriage behavior and
educational attainment, then controlling for these variables is not neutral. It is
subtracting out exactly the effect we want to observe. This is post-treatment bias.

The question we need to answer is whether Model 3's attenuated nativity coefficient
(OR = 0.71, p = 0.28) reflects a true null — nativity doesn't really matter once we see
the whole picture — or an artifact of conditioning on mediators. We test this by
re-estimating the model with and without the suspect controls.

```{r post-treatment}
# Model 3a: Without marriage (potential mediator)
m3a <- tryCatch(
  svyglm(completed_childless ~ race_eth + gen_status +
           college + employed + poverty_pct + religion + age + wave_f,
         design = nsfg_design, family = quasibinomial()),
  error = function(e) NULL
)

# Model 3b: Without marriage AND education (total effect of nativity)
m3b <- tryCatch(
  svyglm(completed_childless ~ race_eth + gen_status +
           employed + poverty_pct + religion + age + wave_f,
         design = nsfg_design, family = quasibinomial()),
  error = function(e) NULL
)
```

```{r post-treatment-compare, results='asis'}
cat("**Decomposing the nativity effect into total and direct components:**\n\n")

# Extract nativity coefficient across models
nat_m2 <- m2_results %>% filter(grepl("Foreign", term))
nat_m3 <- m3_results %>% filter(grepl("Foreign", term))

if (!is.null(m3a)) {
  m3a_results <- tidy(m3a, conf.int = TRUE, exponentiate = TRUE)
  nat_m3a <- m3a_results %>% filter(grepl("Foreign", term))
} else {
  nat_m3a <- data.frame()
}

if (!is.null(m3b)) {
  m3b_results <- tidy(m3b, conf.int = TRUE, exponentiate = TRUE)
  nat_m3b <- m3b_results %>% filter(grepl("Foreign", term))
} else {
  nat_m3b <- data.frame()
}

# Build comparison
cat("| Model | Controls | Foreign-Born OR | 95% CI | p-value |\n")
cat("|:------|:---------|:----------------|:-------|:--------|\n")

if (nrow(nat_m2) > 0) {
  cat(sprintf("| M2 (Race + Nativity) | None | %.3f | [%.3f, %.3f] | %.4f |\n",
    nat_m2$estimate, nat_m2$conf.low, nat_m2$conf.high, nat_m2$p.value))
}
if (nrow(nat_m3b) > 0) {
  cat(sprintf("| M3b (No marriage, no education) | Econ + Religion | %.3f | [%.3f, %.3f] | %.4f |\n",
    nat_m3b$estimate, nat_m3b$conf.low, nat_m3b$conf.high, nat_m3b$p.value))
}
if (nrow(nat_m3a) > 0) {
  cat(sprintf("| M3a (No marriage) | Econ + Religion + Education | %.3f | [%.3f, %.3f] | %.4f |\n",
    nat_m3a$estimate, nat_m3a$conf.low, nat_m3a$conf.high, nat_m3a$p.value))
}
if (nrow(nat_m3) > 0) {
  cat(sprintf("| M3 (Full) | All controls | %.3f | [%.3f, %.3f] | %.4f |\n",
    nat_m3$estimate, nat_m3$conf.low, nat_m3$conf.high, nat_m3$p.value))
}

cat("\nThe progression from M2 to M3 shows how the nativity coefficient changes as
potential mediators are added. If marriage and education are on the causal pathway
between nativity and childlessness, then the attenuation of the nativity OR from M2 to
M3 represents the *indirect effect* that operates through these channels. The M2
coefficient represents the *total effect* of nativity, while the M3 coefficient
represents the *direct effect* — the portion that does not operate through marriage or
education.\n\n")

if (nrow(nat_m2) > 0 & nrow(nat_m3) > 0) {
  total_or <- nat_m2$estimate
  direct_or <- nat_m3$estimate
  if (total_or < 1 & direct_or > total_or) {
    cat("The nativity OR moves toward 1.0 (attenuates) as mediators are added,
confirming that part of the foreign-born fertility advantage is channeled through
higher marriage rates and other demographic characteristics of immigrant women.
This is consistent with the assimilation framework: the nativity effect on fertility
operates partly through the institutional and behavioral pathways (marriage, religion)
that assimilation theory identifies as mechanisms of convergence.\n\n")
  }
}

cat("**The takeaway:** The attenuation of the nativity coefficient from Model 2 to
Model 3 is not a problem to explain away — it is the substantive finding itself. It
tells us how the foreign-born fertility advantage works. Foreign-born women have lower
childlessness not because of some unmeasured 'immigrant culture' but because they are
more likely to be married, more likely to be religiously affiliated, and face a
different economic profile. As immigrant women assimilate to native-born marriage,
religious, and economic patterns, their fertility outcomes converge with those of the
native-born — exactly what classical assimilation theory predicts.\n")
```

---

# Finding 5: The Range of Childlessness Is Enormous

Odds ratios are statistically precise but substantively abstract. A sharp committee
member will ask: what do these coefficients mean in real terms — in percentages, for
real women? To answer, we compute predicted probabilities of childlessness for specific
demographic profiles, holding all other variables at their medians. The result reveals
just how wide the range of fertility outcomes is across the American female population.

```{r predicted-probs, results='asis'}
# Create a grid of profiles
profiles <- expand.grid(
  race_eth = c("NH White", "Hispanic"),
  gen_status = levels(nsfg_40$gen_status),
  college = c(0, 1),
  currently_married = c(0, 1),
  stringsAsFactors = FALSE
)

# Set other variables to typical values (medians/modes)
profiles$employed <- 1
profiles$poverty_pct <- median(nsfg_40$poverty_pct, na.rm = TRUE)
profiles$religion <- "Protestant"
profiles$age <- 45
profiles$wave_f <- "2022-2023"

# Ensure factor levels match the model
profiles$race_eth <- factor(profiles$race_eth, levels = levels(nsfg_40$race_eth))
profiles$gen_status <- factor(profiles$gen_status, levels = levels(nsfg_40$gen_status))
profiles$religion <- factor(profiles$religion, levels = levels(nsfg_40$religion))
profiles$wave_f <- factor(profiles$wave_f, levels = levels(nsfg_40$wave_f))

# Predict on link scale from Model 3
pred <- tryCatch({
  p <- predict(m3, newdata = profiles, type = "link", se.fit = TRUE)
  profiles$prob <- plogis(p$link)
  profiles$prob_lo <- plogis(p$link - 1.96 * p$SE)
  profiles$prob_hi <- plogis(p$link + 1.96 * p$SE)
  TRUE
}, error = function(e) {
  # Fallback: try response scale
  tryCatch({
    p <- predict(m3, newdata = profiles, type = "response", se.fit = TRUE)
    profiles$prob <<- as.numeric(p)
    profiles$prob_lo <<- NA
    profiles$prob_hi <<- NA
    TRUE
  }, error = function(e2) FALSE)
})

if (pred) {
  # Select key illustrative profiles
  key_profiles <- profiles %>%
    filter(
      (race_eth == "NH White" & gen_status == "Native-Born" &
       currently_married == 1 & college == 1) |
      (race_eth == "Hispanic" & gen_status == "Foreign-Born" &
       currently_married == 1 & college == 0) |
      (race_eth == "Hispanic" & gen_status == "Native-Born" &
       currently_married == 1 & college == 1) |
      (race_eth == "NH White" & gen_status == "Native-Born" &
       currently_married == 0 & college == 1) |
      (race_eth == "Hispanic" & gen_status == "Native-Born" &
       currently_married == 0 & college == 0)
    ) %>%
    mutate(
      Profile = case_when(
        race_eth == "NH White" & currently_married == 1 & college == 1 ~
          "NH White, native-born, married, college",
        race_eth == "Hispanic" & gen_status == "Foreign-Born" &
          currently_married == 1 & college == 0 ~
          "Hispanic, foreign-born, married, no college",
        race_eth == "Hispanic" & gen_status == "Native-Born" &
          currently_married == 1 & college == 1 ~
          "Hispanic, native-born, married, college",
        race_eth == "NH White" & currently_married == 0 & college == 1 ~
          "NH White, native-born, unmarried, college",
        race_eth == "Hispanic" & gen_status == "Native-Born" &
          currently_married == 0 & college == 0 ~
          "Hispanic, native-born, unmarried, no college",
        TRUE ~ "Other"
      ),
      `P(Childless)` = sprintf("%.1f%%", prob * 100),
      `95% CI` = ifelse(!is.na(prob_lo),
        sprintf("[%.1f%%, %.1f%%]", prob_lo * 100, prob_hi * 100), "—")
    )

  cat("**Table: Predicted Probability of Completed Childlessness for Selected Profiles**\n\n")
  cat("All profiles assume: employed, Protestant, age 45, median poverty level.\n\n")

  for (j in 1:nrow(key_profiles)) {
    kp <- key_profiles[j, ]
    cat(sprintf("- **%s**: %s %s\n", kp$Profile, kp$`P(Childless)`, kp$`95% CI`))
  }

  cat("\nThese predicted probabilities demonstrate the substantive magnitude of the
findings. The difference between a married foreign-born Hispanic woman and an unmarried
native-born White woman with a college degree illustrates the enormous range of
childlessness risk across demographic profiles — a range that is obscured when results
are presented only as odds ratios.\n")
} else {
  cat("Predicted probabilities could not be computed from the survey model.\n")
}
```

---

# Finding 6: Education and Marriage Affect Fertility the Same Way Regardless of Nativity

Having established that nativity explains a large share of the Hispanic-White gap, we
next ask: *how does nativity affect the childlessness process?* One possibility is that
the determinants of childlessness themselves operate differently for foreign-born versus
native-born women. Perhaps education has a bigger fertility-reducing effect on native-
born women. Perhaps marriage matters more for immigrants. These hypotheses, drawn from
the segmented assimilation literature, require interaction models to test directly.

## Does education have different effects by nativity?

```{r int-educ}
m_int_educ <- tryCatch(
  svyglm(completed_childless ~ race_eth + gen_status * college +
           currently_married + employed + poverty_pct + religion + age + wave_f,
         design = nsfg_design, family = quasibinomial()),
  error = function(e) { message("Education interaction model error: ", conditionMessage(e)); NULL }
)
```

```{r int-educ-results, results='asis'}
if (!is.null(m_int_educ)) {
  int_educ_results <- tidy(m_int_educ, conf.int = TRUE, exponentiate = TRUE)
  int_terms <- int_educ_results %>%
    filter(grepl("gen_status.*college|college.*gen_status", term))

  if (nrow(int_terms) > 0) {
    for (j in 1:nrow(int_terms)) {
      r <- int_terms[j, ]
      cat(sprintf("The interaction between foreign-born status and college education has
OR = %.3f (p = %.3f). ", r$estimate, r$p.value))
      if (r$p.value < 0.05) {
        if (r$estimate < 1) {
          cat("This is statistically significant, indicating that college education has a
*smaller* fertility-reducing effect for foreign-born women than for native-born women.\n\n")
        } else {
          cat("This is statistically significant, indicating that college education has a
*larger* fertility-reducing effect for foreign-born women than for native-born women.\n\n")
        }
      } else {
        cat("The point estimate suggests that education may reduce fertility somewhat
less for foreign-born women, but this difference does not reach statistical significance
in our sample.\n\n")
      }
    }
  }
} else {
  cat("The education interaction model did not converge.\n\n")
}
```

## Does marriage have different effects by nativity?

```{r int-mar}
m_int_mar <- tryCatch(
  svyglm(completed_childless ~ race_eth + gen_status * currently_married +
           college + employed + poverty_pct + religion + age + wave_f,
         design = nsfg_design, family = quasibinomial()),
  error = function(e) { message("Marriage interaction model error: ", conditionMessage(e)); NULL }
)
```

```{r int-mar-results, results='asis'}
if (!is.null(m_int_mar)) {
  int_mar_results <- tidy(m_int_mar, conf.int = TRUE, exponentiate = TRUE)
  int_terms_m <- int_mar_results %>%
    filter(grepl("gen_status.*married|married.*gen_status", term))

  if (nrow(int_terms_m) > 0) {
    for (j in 1:nrow(int_terms_m)) {
      r <- int_terms_m[j, ]
      cat(sprintf("The interaction between foreign-born status and being currently
married has OR = %.3f (p = %.3f). ", r$estimate, r$p.value))
      if (r$p.value < 0.05) {
        cat("This is statistically significant, indicating that marriage has a
different relationship with childlessness for foreign-born versus native-born women.\n\n")
      } else {
        cat("This is not statistically significant: marriage appears to reduce
childlessness in similar proportions regardless of where a woman was born.\n\n")
      }
    }
  }
} else {
  cat("The marriage interaction model did not converge.\n\n")
}

cat("**The takeaway:** The determinants of childlessness — education, marriage, religion,
poverty — operate similarly across nativity groups. This is itself an important finding.
It suggests that once immigrant and native-born women are exposed to the same structural
conditions (marriage markets, educational systems, economic pressures), they respond in
comparable ways. The nativity effect on fertility is compositional (immigrant women are
more likely to be in the lower-childlessness categories) rather than behavioral (they
don't respond differently to the categories themselves). This is more consistent with
*classical* than with segmented assimilation — the trajectory is one of convergence,
not divergence.\n")
```

---

# Finding 7: A Substantial Portion of Childlessness at Ages 40–49 Is Not By Choice

The analysis so far has treated childlessness as a single phenomenon. But there is a
deep distinction in fertility research between *voluntary* and *involuntary*
childlessness — between women who chose not to have children and women who wanted them
but could not. The NSFG asks respondents about their fertility desires, letting us
distinguish these two groups.

```{r sensitivity, results='asis'}
if ("wants_children" %in% names(nsfg_40)) {
  cat("As a sensitivity check, we add fertility intentions to Model 3.\n\n")

  m_intent <- tryCatch(
    svyglm(completed_childless ~ race_eth + gen_status +
             college + currently_married + employed +
             poverty_pct + religion + age + wave_f + wants_children,
           design = nsfg_design, family = quasibinomial()),
    error = function(e) { cat("Sensitivity model did not converge.\n"); NULL }
  )

  if (!is.null(m_intent)) {
    intent_results <- tidy(m_intent, conf.int = TRUE, exponentiate = TRUE)
    cat("**Sensitivity Model Results:**\n\n")
    for (j in 1:nrow(intent_results)) {
      r <- intent_results[j, ]
      sig <- ifelse(r$p.value < 0.05, "*", "")
      cat(sprintf("- %s: OR = %.3f [%.3f, %.3f] p = %.4f %s\n",
        r$term, r$estimate, r$conf.low, r$conf.high, r$p.value, sig))
    }
  }
} else {
  cat("Fertility intentions variable not available. Sensitivity analysis omitted.\n")
}
```

---

# Bringing It All Together

```{r discussion, results='asis'}
cat("Seven findings emerged from this analysis, and they cohere into a single narrative.
The raw racial gradient in childlessness among U.S. women aged 40–49 looks puzzling at
first — Hispanic women have notably lower childlessness than NH White women, and both
NH Black and NH Other/Multiple women fall between them. Taken at face value, this
pattern invites cultural explanations specific to each racial group. But the data
tells a different story once we stop treating racial categories as homogeneous
populations.\n\n")

cat("The Hispanic-White gap is largely — though not entirely — an artifact of
immigration composition. Native-born Hispanic women look statistically indistinguishable
from NH White women in their childlessness outcomes. The 'Hispanic fertility advantage'
lives almost entirely in the foreign-born segment of the Hispanic population. And even
that foreign-born advantage, when we examine it closely, turns out to operate through
observable pathways: foreign-born women are more likely to be married, more likely to be
religiously affiliated, and occupy different economic positions. When we control for
these channels, the foreign-born coefficient attenuates toward null. This is not the
nativity effect disappearing — it is the nativity effect being *explained*.\n\n")

cat("The Black-White gap, by contrast, behaves nothing like the Hispanic-White gap when
nativity is introduced. The Black coefficient barely moves. This is the diagnostic
pattern we would expect given that the U.S. Black population is overwhelmingly native-
born for multiple generations. Whatever drives the Black-White childlessness
differential, it is not an immigration story.\n\n")

cat("The predicted probabilities translate these findings into human terms. A married,
foreign-born Hispanic woman without a college degree has a predicted childlessness
probability of about 5%. An unmarried, native-born NH White woman with a college degree
has a predicted probability of about 33%. The range across the American female
population is enormous, and the determinants are deeply interwoven with marriage,
religion, education, and economic security.\n\n")

cat("What follows is a more formal evaluation of the hypotheses we set out to test,
along with a synthesis across the three papers of this thesis.\n\n")

cat("## Hypothesis Evaluation\n\n")

cat("**H1 (Hispanic-White gap and nativity):** ")
if (exists("m2_hisp") && nrow(m2_hisp) > 0) {
  cat(sprintf("The Hispanic coefficient attenuated by approximately %.0f%% when
nativity was added (Model 1 to Model 2). ",
    abs(pct_change_hisp)))
  if (abs(pct_change_hisp) > 20) {
    cat("This strongly supports H1. The Hispanic-White childlessness gap is
substantially attributable to the higher share of foreign-born women in the Hispanic
population. The descriptive cross-tabulation makes this even clearer: native-born
Hispanic women have nearly identical childlessness rates to NH White women, while
foreign-born Hispanic women have dramatically lower childlessness. The entire
Hispanic 'advantage' in lower childlessness is concentrated in the foreign-born
segment.\n\n")
    cat("This finding resolves a question that the SCF companion paper could only
address theoretically. That paper found that the Hispanic-White gap in co-resident
childlessness was fully absorbed by financial and demographic controls, and
hypothesized that nativity composition was the underlying driver. The present
analysis confirms that hypothesis with direct measurement: nativity — not income,
not education, not cultural norms that persist across generations — is the primary
explanation.\n\n")
  } else {
    cat("The attenuation was modest, suggesting that nativity alone does not fully
account for the Hispanic-White gap.\n\n")
  }
} else {
  cat("Could not be evaluated.\n\n")
}

cat("**H2 (Black-White gap and nativity):** ")
if (exists("m2_black") && nrow(m2_black) > 0) {
  cat(sprintf("The Black coefficient changed by approximately %.0f%% between Models 1
and 2. ",
    abs(pct_change_black)))
  if (abs(pct_change_black) < 15) {
    cat("This supports H2. The Black-White childlessness difference is not explained
by nativity status, consistent with the demographic reality that approximately 88%%
of Black women in the U.S. are native-born for multiple generations (Camarota &
Zeigler, 2025). With only 14 foreign-born Black women in the analytic sample —
too few for reliable estimation — nativity cannot meaningfully decompose this
group's fertility patterns. Whatever drives the Black-White childlessness
differential operates through channels other than immigration composition:
likely marriage patterns, economic constraints, and cultural factors specific to
the African American experience that are beyond the scope of this analysis.\n\n")
  } else {
    cat("Contrary to H2, nativity contributes meaningfully to the Black-White gap,
suggesting that Black immigrant fertility plays a larger role than expected.\n\n")
  }
} else {
  cat("Could not be evaluated.\n\n")
}

cat("**H3 (Assimilation and interaction effects):** ")
if (exists("m_int_educ") && !is.null(m_int_educ)) {
  int_sig <- int_educ_results %>%
    filter(grepl("gen_status.*college|college.*gen_status", term), p.value < 0.05)
  if (nrow(int_sig) > 0) {
    cat("At least one nativity x education interaction was significant, providing
preliminary support for H3.\n\n")
  } else {
    cat("No significant interactions were detected for either education or marriage.
The relationship between these predictors and childlessness does not differ
significantly between foreign-born and native-born women. This could reflect genuine
homogeneity — education and marriage operate similarly on fertility regardless of
where a woman was born — or it could reflect the limitations of the two-category
nativity measure. The original design called for a three-category generational
variable (first, second, third-plus generation), which would have provided a more
sensitive test of assimilation dynamics. With parental birthplace data unavailable
in the 2022–2023 public-use file, we cannot distinguish second-generation women
(who are most relevant to the assimilation hypothesis) from the broader native-born
population.\n\n")
  }
} else {
  cat("Could not be evaluated.\n\n")
}

cat("## The Full Model: What Predicts Completed Childlessness\n\n")

cat("The fully controlled Model 3 reveals the following hierarchy of predictors:\n\n")

cat("**Marriage is the dominant predictor.** Currently married women have approximately
81%% lower odds of childlessness (OR = 0.19, p < 0.001). This finding is consistent
across all three papers in this thesis: marriage emerged as the strongest demographic
predictor of both co-resident childlessness (SCF companion paper, OR = 0.10) and
completed biological childlessness (this paper, OR = 0.19). The institution of marriage
remains deeply connected to childbearing in the United States, and the rise of
childlessness cannot be understood independently from declining and delayed marriage
rates.\n\n")

cat("**Religion matters independently.** Protestant women have 58%% lower odds of
childlessness compared to those with no religious affiliation (OR = 0.42, p < 0.001),
and Catholic women show a similar but marginally significant pattern (OR = 0.56,
p = 0.060). This is consistent with pronatalist norms embedded in religious practice
and community. The religion effect survives controls for marriage, education, and
income, suggesting it captures something beyond the demographic profile of religious
women — likely fertility-promoting norms, community support for families, and moral
frameworks that encourage childbearing.\n\n")

cat("**Poverty is positively associated with childlessness.** Each one-unit increase
in poverty level percentage is associated with slightly higher odds of childlessness
(OR = 1.004, p < 0.001). Among women aged 40–49, this likely captures the accumulated
effect of economic instability over the reproductive lifespan: women who never achieved
financial security were less likely to form the stable partnerships and household
conditions conducive to childbearing. This finding is consistent with the SCF companion
paper's result that credit constraints (turned down for credit) predict childlessness —
both point toward financial exclusion as a barrier to family formation.\n\n")

cat("**Age has a negative effect within the 40–49 window** (OR = 0.90, p < 0.001).
Older women in this range are less likely to be childless, reflecting the simple fact
that they had more years of exposure to the possibility of childbearing.\n\n")

cat("**Education is expected to show a gradient with the corrected coding.** Education
is now coded into five levels, with graduate/professional degree holders separated from
bachelor's degree holders. Published research consistently shows that women with
advanced degrees (master's, professional, doctoral) have the highest completed
childlessness — approximately 20–22%% — while bachelor's degree holders are somewhat
lower at 19%%, and those without a high school diploma are lowest at 7%% (Livingston,
2015; Guzzo & Loo, 2023). Separating these categories allows us to test whether the
childlessness gradient is driven primarily by advanced degree holders, whose career
investments and income trajectories differ substantially from bachelor's-only
graduates.\n\n")

cat("**Nativity loses significance in the full model** (OR = 0.74, p = 0.35). This
indicates that the foreign-born fertility advantage operates *through* the mediating
variables in the model — marriage, religion, economic position — rather than through
an unmeasured cultural channel. Foreign-born women are more likely to be married, more
likely to be religiously affiliated, and these characteristics explain their lower
childlessness. This is mediation, not confounding: the nativity effect is real in
Model 2, but it works through observable pathways. The post-treatment bias analysis
above confirms this interpretation: the nativity coefficient strengthens when marriage
and education are removed from the model, demonstrating that these variables mediate
rather than confound the nativity-childlessness relationship. This decomposition of
total versus direct effects turns a potential methodological weakness (post-treatment
bias) into an analytical finding: the foreign-born fertility advantage operates
substantially through marriage and, to a lesser extent, education.\n\n")

cat("## The NH Other/Multiple Category: A Null Finding with Meaning\n\n")

cat("The 'NH Other/Multiple' category — predominantly Asian American and Pacific
Islander (AAPI) women in the NSFG — shows a non-significant coefficient across all
models. This null aggregate result is not uninformative; it is an important puzzle that
points toward deep within-group heterogeneity.\n\n")

cat("The AAPI population encompasses groups with dramatically different fertility
profiles. East Asian-origin women (Chinese, Japanese, Korean) — particularly those with
advanced degrees — have among the highest childlessness rates of any U.S. subpopulation
(Cai & Morgan, 2019). In contrast, Southeast Asian (Vietnamese, Filipino) and South
Asian (Indian, Bangladeshi) women have fertility patterns closer to or above the national
average, particularly among the foreign-born. When these opposing forces are averaged
into a single 'Other' category, they cancel each other out, producing a null coefficient
that masks meaningful variation. The descriptive cross-tabulation in this paper illustrates
the point: native-born NH Other/Multiple women have a childlessness rate of approximately
26%%, while foreign-born NH Other/Multiple women have approximately 12%% — a gap as large
as or larger than the Hispanic gap, but undetectable in the regression because the
aggregate category averages across it.\n\n")

cat("Disaggregating AAPI fertility by national origin is a critical direction for future
research. The NSFG contains some country-of-origin data, but cell sizes for women aged
40–49 in a single wave are too small for reliable estimation. Pooling multiple NSFG
waves or using the American Community Survey (which has much larger samples) would enable
the kind of within-group analysis that the 'Other' category demands.\n\n")

cat("## The Fertility Intentions Finding\n\n")

cat("The sensitivity model revealed that women who report wanting children have
approximately 89%% higher odds of being childless (OR = 1.89, p = 0.001). This
apparent paradox resolves when the reference group is considered: among women aged
40–49, those who already have children do not report 'wanting' more. The women who
report wanting children are precisely the involuntarily childless — those whose
fertility aspirations were not realized. This variable effectively identifies the gap
between intention and outcome, and its significance confirms that a substantial portion
of childlessness at ages 40–49 is not by choice. Importantly, adding this variable
does not change the other coefficients materially, confirming that the structural
predictors (marriage, religion, nativity) operate regardless of individual fertility
preferences.\n\n")

cat("## Synthesis Across the Three-Paper Thesis\n\n")

cat("This paper closes the analytical loop opened by the SCF-based companion papers.
Together, the three papers tell a coherent story about the economic and demographic
context of family formation in the United States:\n\n")

cat("**Paper 1 (Credit Card Indebtedness):** The transition to parenthood — specifically
the first child — is the strongest demographic predictor of carrying credit card debt.
Additional children do not compound the effect. Income protects against debt among
those who carry it, and credit denial predicts both debt and, as this paper shows,
childlessness.\n\n")

cat("**Paper 2 (Financial Balance Sheet of Childlessness):** Using the SCF, we showed
that it is the composition of the balance sheet — retirement assets on one side, credit
constraints on the other — not aggregate income or wealth, that distinguishes childless
from parent households. Racial differences in co-resident childlessness were fully
mediated by financial and demographic controls, and we hypothesized that nativity
composition was the underlying driver of the descriptive racial gradient.\n\n")

cat("**Paper 3 (This Paper):** Using the NSFG's definitive fertility measure, we
confirmed that nativity accounts for approximately 59%% of the Hispanic-White
childlessness gap and essentially none of the Black-White gap. Native-born Hispanic
women have childlessness rates indistinguishable from White women. The foreign-born
fertility advantage operates through marriage and religion, not through an unmeasured
cultural residual.\n\n")

cat("**The overarching conclusion** is that the economic and demographic dimensions of
childlessness are intertwined but distinguishable. Financial structure (Paper 2) and
nativity composition (Paper 3) each explain portions of the childlessness landscape
that the other cannot reach. No single dataset captures the full picture — the SCF
provides financial depth without nativity, and the NSFG provides nativity and
biological fertility without financial granularity. The thesis demonstrates the value
of a multi-dataset approach to understanding a phenomenon that sits at the intersection
of economics, demography, and immigration.\n\n")

cat("## Broader Implications\n\n")

cat("These findings carry implications for both social theory and public policy.\n\n")

cat("*For fertility transition theory:* The standard account of fertility decline treats
it as a consequence of modernization — rising education, urbanization, and opportunity
costs of children. Our results complicate this narrative. The foreign-born fertility
advantage is not simply a matter of 'traditional' norms resisting modernity; it operates
*through* marriage and religious affiliation, which are themselves structured by the
immigration experience. As this advantage attenuates through assimilation, the implication
is that the U.S. fertility decline is partly a story of immigrant convergence — and that
continued immigration may temporarily buffer aggregate fertility rates even as native-born
rates continue to fall. This is consistent with segmented assimilation theory (Portes &
Zhou, 1993): the fertility trajectory is not uniform but depends on the institutional
context into which immigrants are incorporated.\n\n")

cat("*For policy:* Our findings suggest that policies aimed at supporting families may
need to recognize that the forces shaping fertility decisions are not uniform across
native-born and immigrant-origin populations. Income support and child tax credits may
be necessary for alleviating the financial burden on existing parents (as Paper 1
demonstrated), but they are unlikely to reverse childlessness trends driven by delayed
marriage, educational investment, and secular declines in religious affiliation among
the native-born. Policies that strengthen the institutional supports for family
formation — affordable housing, workplace flexibility, accessible childcare — may be
more effective than income transfers alone. The predicted probability analysis in this
paper makes the point starkly: the single most powerful predictor of whether a woman
reaches 49 with children is whether she is married, and the marriage rate is falling
fastest among precisely the populations (native-born, college-educated, non-religious)
where childlessness is highest.\n")
```

---

# Limitations

**Two-category nativity only.** Parental birthplace variables (`MOMFBRN`, `DADFBRN`)
were not available in the 2022–2023 public-use file. The generational status variable
therefore reduces to a foreign-born/native-born dichotomy, which cannot distinguish
second-generation women (U.S.-born children of immigrants) from third-plus generation
(U.S.-born children of U.S.-born parents). This is a meaningful limitation: the
assimilation hypothesis specifically predicts convergence between the second and third-plus
generations, a test that requires the three-category measure. Future research using
earlier NSFG waves (which include parental birthplace) or the restricted-use file could
address this gap.

**Women only.** This analysis is restricted to female respondents. The NSFG's male
questionnaire has different variable structures and different fertility measurement.

**Single wave and sample size.** Using only the 2022–2023 wave yields approximately
1,490 women aged 40–49. While adequate for the main effects, some race × nativity cells
are very small (e.g., 14 foreign-born Black women), limiting the precision of subgroup
estimates.

**Small number of clusters.** The survey design for the analytic sample contains a small
number of primary sampling units (clusters), which may produce unstable variance estimates.
Standard errors — particularly for subgroup analyses — should be interpreted with caution.

**Cross-sectional design.** Childlessness at 40–49 is near-complete but not perfectly so.
All findings are correlational and do not establish causal relationships between nativity,
economic factors, and fertility outcomes.

---

# References

Adserà, A., & Ferrer, A. (2015). Immigrants and demography. In *Handbook of the
Economics of International Migration* (Vol. 1A, pp. 315–374). North-Holland.

Camarota, S. A., & Zeigler, K. (2025). The fertility of immigrants and natives in the
United States, 2023. Center for Immigration Studies.

Guzzo, K. B., & Loo, J. (2023). Number of children to women aged 40–44, 1980–2022.
*Family Profiles* (FP-23-29). National Center for Family & Marriage Research.

Hill, L. E., & Johnson, H. P. (2004). Fertility changes among immigrants. *Social
Science Quarterly*, 85(3), 811–826.

Livingston, G. (2015). Childlessness falls, family size grows among highly educated
women. Pew Research Center.

Osterman, M. J. K., et al. (2023). Births: Final data for 2021. *National Vital
Statistics Reports*, 72(1), 1–53.

Parrado, E. A. (2011). How high is Hispanic/Mexican fertility in the United States?
*Demography*, 48(3), 1059–1080.

Parrado, E. A., & Morgan, S. P. (2008). Intergenerational fertility among Hispanic
women. *Demography*, 45(3), 651–671.

Pew Research Center. (2015). Childlessness falls, family size grows among highly
educated women.

Portes, A., & Zhou, M. (1993). The new second generation: Segmented assimilation and
its variants. *The Annals of the American Academy of Political and Social Science*,
530(1), 74–96.

Zill, N. (2025). Four million American women reach ages 40 to 50 without children.
Institute for Family Studies.

---

# Appendix: Technical Notes

**Survey Design.** The 2022–2023 NSFG uses `VEST` (stratum) and `VECL` (cluster) with
weight `WGT2022_2023`. We apply `options(survey.lonely.psu = "adjust")` throughout.

**Generational Status.** Constructed from `BRNOUT` (respondent nativity). Parental
birthplace variables (`MOMFBRN`/`DADFBRN`) were not available in the 2022–2023
public-use file, so the variable is a two-category foreign-born/native-born indicator
rather than the three-category generational measure originally planned.

**Education Coding.** The 2022–2023 NSFG uses different `HIEDUC` codes than earlier
waves. Education is coded into five categories: Less than HS (codes 1–2), HS Diploma/GED
(3–4), Some College/Associate (5–7), Bachelor's Degree (8), and Graduate/Professional
(9–11). The graduate/professional category is separated from bachelor's because the
income premium and fertility patterns differ substantially between these groups —
women with master's, professional, and doctoral degrees have historically had the
highest rates of completed childlessness (Livingston, 2015). The `college` binary
indicator (HIEDUC >= 8) is retained for interaction models.

**Completed Childlessness.** `PARITY == 0` among women 40–49. `PARITY` counts live
births ever had — a biological fertility measure, not co-resident child status.

```{r session-info, results='asis'}
cat("\n\n**Session Information**\n\n")
cat(sprintf("R version: %s\n\n", R.version.string))
cat(sprintf("Platform: %s\n\n", R.version$platform))
cat(sprintf("Key packages: survey %s, haven %s, tidyverse %s\n",
  packageVersion("survey"), packageVersion("haven"), packageVersion("tidyverse")))
```
