---
title: "Nativity, Economic Status, and Childlessness in America"
subtitle: "A Generational Analysis of the National Survey of Family Growth (2022–2023)"
author: "Delroy Nichols"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 5, dpi = 150
)
rm(list = ls())
gc()
set.seed(42)
options(survey.lonely.psu = "adjust")
```

# Abstract

This paper directly tests the hypothesis that nativity — being foreign-born versus
native-born — explains racial and ethnic differences in completed childlessness among
U.S. women. Using the 2022–2023 National Survey of Family Growth (NSFG), we analyze
women aged 40–49 who have reached the end of their childbearing years. The NSFG provides
what the Survey of Consumer Finances cannot: a definitive measure of biological fertility
(children ever born), detailed nativity data (respondent's, mother's, and father's
country of birth), and standard economic and demographic controls.

Using a sequential logistic regression design, we first estimate the unconditional racial
gradient in completed childlessness, then introduce generational status (first generation,
second generation, third-plus generation) to determine how much of the racial gradient is
attributable to nativity composition, and finally add economic and demographic controls
to identify the residual. We test whether the effects of education and marriage on
childlessness differ by generational status through interaction models.

This study completes a three-paper thesis: the first paper analyzed credit card
indebtedness, the second examined the financial balance sheet of co-resident
childlessness, and this paper provides the direct nativity test that the SCF-based
analyses could only theorize about.

---

```{r load-packages}
required_packages <- c(
  "tidyverse", "haven", "survey",
  "broom", "knitr", "ggplot2", "scales"
)

for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg, repos = "http://cran.us.r-project.org")
    library(pkg, character.only = TRUE)
  }
}
```

# Introduction

The first two papers in this thesis established that (1) the transition to parenthood is
the strongest demographic predictor of credit card indebtedness, and (2) the financial
balance sheet of childless households differs systematically from that of parents — with
retirement asset accumulation and credit constraints, not income, distinguishing the two
groups. Both analyses used the Survey of Consumer Finances (SCF), which provides
unmatched financial detail but lacks any measure of nativity or biological fertility.

The SCF companion paper found that racial and ethnic differences in co-resident
childlessness were fully absorbed by financial and demographic controls, and hypothesized
— based on external literature — that immigration-driven compositional effects might
explain the descriptive racial gradient. This paper tests that hypothesis directly.

The National Survey of Family Growth (NSFG) is the premier U.S. dataset on fertility and
family formation. It provides three capabilities the SCF lacks: (1) a definitive measure
of biological fertility — children ever born (`PARITY`), not merely co-resident children;
(2) nativity data for the respondent and both parents, enabling construction of a
three-category generational status variable; and (3) fertility intentions, which can
distinguish voluntary from involuntary childlessness.

**Research Questions:**

1. *Primary:* To what extent does nativity (foreign-born vs. native-born) explain racial
   and ethnic differences in completed childlessness among U.S. women?

2. *Secondary (Mediation):* Do economic factors (education, income, employment) and
   family structure (marital status) mediate the relationship between nativity and
   childlessness?

3. *Secondary (Interaction):* Do the effects of education and marriage on childlessness
   differ by generational status?

---

# Data and Methods

## Data Source

We use the female respondent file from the 2022–2023 NSFG, the most recent wave
available. The NSFG surveys women aged 15–49 in the civilian noninstitutionalized
household population. Data were collected by NORC at the University of Chicago using
computer-assisted personal interviewing (CAPI), with sensitive items administered via
audio computer-assisted self-interviewing (ACASI).

```{r download-data}
dir.create("nsfg_data", showWarnings = FALSE, recursive = TRUE)

file_path <- "nsfg_data/fem_resp_2022_2023.sas7bdat"

if (!file.exists(file_path)) {
  cat("Downloading 2022-2023 NSFG Female Respondent file...\n")
  tryCatch(
    download.file(
      "https://ftp.cdc.gov/pub/Health_Statistics/NCHS/NSFG/NSFG-2022-2023-FemRespPUFData.sas7bdat",
      file_path, mode = "wb", quiet = TRUE
    ),
    error = function(e) cat("Download error:", conditionMessage(e), "\n")
  )
}

nsfg_tbl <- read_sas(file_path)
nsfg_pooled <- as.data.frame(nsfg_tbl)
names(nsfg_pooled) <- tolower(names(nsfg_pooled))
nsfg_pooled$wave <- "2022-2023"
```

```{r load-confirm, results='asis'}
cat(sprintf("NSFG 2022–2023 loaded: %s female respondents.\n",
  formatC(nrow(nsfg_pooled), big.mark = ",")))
```

## Variable Construction

```{r variables, results='asis'}
nsfg <- nsfg_pooled

# ---- 1. OUTCOME: Completed Childlessness ----
# PARITY = total number of live births ever had (recode)
nsfg$completed_childless <- as.integer(nsfg$parity == 0)

# ---- 2. AGE ----
nsfg$age <- nsfg$ager

# ---- 3. RACE / ETHNICITY (NH White as reference) ----
# HISPRACE2: 1=Hispanic, 2=NH White, 3=NH Black, 4=NH Other/Multiple
nsfg$race_eth <- factor(nsfg$hisprace2,
  levels = c(2, 1, 3, 4),
  labels = c("NH White", "Hispanic", "NH Black", "NH Other/Multiple")
)

# ---- 4. NATIVITY & GENERATIONAL STATUS ----
nsfg$foreign_born <- as.integer(nsfg$brnout == 1)

has_mom <- "momfbrn" %in% names(nsfg)
has_dad <- "dadfbrn" %in% names(nsfg)

if (has_mom && has_dad) {
  nsfg$mom_foreign <- as.integer(nsfg$momfbrn == 1)
  nsfg$dad_foreign <- as.integer(nsfg$dadfbrn == 1)

  nsfg$gen_status <- case_when(
    nsfg$foreign_born == 1 ~ "1st Generation",
    nsfg$foreign_born == 0 &
      (nsfg$mom_foreign == 1 | nsfg$dad_foreign == 1) ~ "2nd Generation",
    nsfg$foreign_born == 0 &
      nsfg$mom_foreign == 0 & nsfg$dad_foreign == 0 ~ "3rd+ Generation",
    TRUE ~ NA_character_
  )
  nsfg$gen_status <- factor(nsfg$gen_status,
    levels = c("3rd+ Generation", "2nd Generation", "1st Generation"))

  cat("Generational status created using respondent + parental birthplace.\n\n")
} else {
  nsfg$gen_status <- factor(
    ifelse(nsfg$foreign_born == 1, "Foreign-Born", "Native-Born"),
    levels = c("Native-Born", "Foreign-Born")
  )
  cat("NOTE: Parental nativity variables not found. Using respondent nativity only.\n")
  if (!has_mom) cat("  Missing: MOMFBRN\n")
  if (!has_dad) cat("  Missing: DADFBRN\n")
  cat("\n")
}

# ---- 5. EDUCATION ----
nsfg$educ <- case_when(
  nsfg$hieduc %in% c(5, 6)     ~ "Less than HS",
  nsfg$hieduc %in% c(7, 8)     ~ "HS Diploma/GED",
  nsfg$hieduc %in% c(9, 10)    ~ "Some College/Associate",
  nsfg$hieduc >= 11             ~ "Bachelor's or Higher",
  TRUE ~ NA_character_
)
nsfg$educ <- factor(nsfg$educ,
  levels = c("Less than HS", "HS Diploma/GED",
             "Some College/Associate", "Bachelor's or Higher"))

nsfg$college <- as.integer(nsfg$hieduc >= 11)

# ---- 6. MARITAL STATUS ----
nsfg$ever_married      <- as.integer(nsfg$fmarital %in% c(1, 2, 3, 4))
nsfg$currently_married <- as.integer(nsfg$fmarital == 1)

# ---- 7. OTHER CONTROLS ----
nsfg$employed    <- as.integer(nsfg$laborfor == 1)
nsfg$poverty_pct <- nsfg$poverty
nsfg$religion    <- factor(nsfg$religion,
  levels = 1:4,
  labels = c("No Religion", "Catholic", "Protestant", "Other"))

if ("rwant" %in% names(nsfg)) {
  nsfg$wants_children <- as.integer(nsfg$rwant == 1)
}

nsfg$wave_f <- factor(nsfg$wave)

# ---- SUMMARY ----
cat(sprintf("Total sample: %s female respondents.\n",
  formatC(nrow(nsfg), big.mark = ",")))
cat(sprintf("Women aged 40-49: %s.\n",
  formatC(sum(nsfg$age >= 40 & nsfg$age <= 49, na.rm = TRUE), big.mark = ",")))
cat(sprintf("Completed childlessness rate (full sample): %.1f%%.\n",
  mean(nsfg$completed_childless, na.rm = TRUE) * 100))
```

## Sample Restriction and Survey Design

```{r sample-and-design, results='asis'}
nsfg_40 <- nsfg %>% filter(age >= 40, age <= 49)

# Weight
nsfg_40$analysis_weight <- nsfg_40$wgt2022_2023

# Survey design: 2022-2023 uses vecl (cluster) and vest (stratum)
nsfg_design <- svydesign(
  id      = ~ vecl,
  strata  = ~ vest,
  weights = ~ analysis_weight,
  data    = nsfg_40,
  nest    = TRUE
)

cat(sprintf("Analytic sample: %s women aged 40-49.\n",
  formatC(nrow(nsfg_40), big.mark = ",")))
cat(sprintf("Strata: %d. Clusters: %d.\n",
  length(unique(nsfg_40$vest)), length(unique(nsfg_40$vecl))))
```

## Analytical Plan

Our analysis proceeds in four stages:

1. **Descriptive analysis:** Cross-tabulate completed childlessness by race/ethnicity
   and generational status to show unconditional rates.

2. **Sequential logistic regression:**
   - *Model 1:* `completed_childless ~ race_eth` (unconditional racial gradient)
   - *Model 2:* `completed_childless ~ race_eth + gen_status` (does generation account
     for the race effect?)
   - *Model 3:* `completed_childless ~ race_eth + gen_status + controls` (full model)

3. **Interaction models:** Test `gen_status × college` and
   `gen_status × currently_married`.

4. **Sensitivity check:** Include fertility intentions if available.

All models use NSFG survey weights and account for the complex sample design.

---

# Descriptive Results

## Cell Sizes

```{r cell-sizes, results='asis'}
cell_table <- nsfg_40 %>%
  filter(!is.na(gen_status), !is.na(race_eth)) %>%
  count(gen_status, race_eth) %>%
  pivot_wider(names_from = race_eth, values_from = n, values_fill = 0)

cat("**Table 1. Unweighted Cell Sizes by Generational Status and Race/Ethnicity**\n\n")
```

```{r cell-sizes-table}
kable(cell_table)
```

```{r cell-size-check, results='asis'}
min_cells <- nsfg_40 %>%
  filter(!is.na(gen_status), !is.na(race_eth)) %>%
  count(gen_status, race_eth) %>%
  filter(n < 30)

if (nrow(min_cells) > 0) {
  cat("\n**Warning:** The following cells have fewer than 30 observations:\n\n")
  for (j in 1:nrow(min_cells)) {
    cat(sprintf("- %s x %s: n = %d\n",
      min_cells$gen_status[j], min_cells$race_eth[j], min_cells$n[j]))
  }
  cat("\n")
} else {
  cat("\nAll cells contain at least 30 observations.\n\n")
}
```

## Completed Childlessness Rates

```{r desc-childless}
format_svyby <- function(svy_result, group_name) {
  df <- as.data.frame(svy_result)
  df$rate <- round(df$completed_childless * 100, 1)
  df$se_pct <- round(df$se * 100, 1)
  df[, c(group_name, "rate", "se_pct")]
}

cat("===== By Race/Ethnicity =====\n")
print(format_svyby(
  svyby(~completed_childless, ~race_eth, nsfg_design, svymean, na.rm = TRUE),
  "race_eth"), row.names = FALSE)

cat("\n===== By Generational Status =====\n")
print(format_svyby(
  svyby(~completed_childless, ~gen_status, nsfg_design, svymean, na.rm = TRUE),
  "gen_status"), row.names = FALSE)

cat("\n===== By Race x Generation =====\n")
print(format_svyby(
  svyby(~completed_childless, ~race_eth + gen_status, nsfg_design, svymean, na.rm = TRUE),
  c("race_eth", "gen_status")), row.names = FALSE)

cat("\n===== By Education =====\n")
print(format_svyby(
  svyby(~completed_childless, ~educ, nsfg_design, svymean, na.rm = TRUE),
  "educ"), row.names = FALSE)

cat("\n===== By Marital Status =====\n")
mar_rates <- svyby(~completed_childless, ~currently_married, nsfg_design,
                    svymean, na.rm = TRUE)
mar_df <- as.data.frame(mar_rates)
mar_df$status <- ifelse(mar_df$currently_married == 1, "Married", "Not Married")
mar_df$rate <- round(mar_df$completed_childless * 100, 1)
mar_df$se_pct <- round(mar_df$se * 100, 1)
print(mar_df[, c("status", "rate", "se_pct")], row.names = FALSE)

cat("\n===== By Religion =====\n")
print(format_svyby(
  svyby(~completed_childless, ~religion, nsfg_design, svymean, na.rm = TRUE),
  "religion"), row.names = FALSE)
```

## Visualizations

```{r viz-race-gen, fig.width=11, fig.height=6}
race_gen <- svyby(~completed_childless, ~race_eth + gen_status, nsfg_design,
                   svymean, na.rm = TRUE)
rg_df <- as.data.frame(race_gen)

ggplot(rg_df, aes(x = race_eth, y = completed_childless * 100, fill = gen_status)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(aes(ymin = (completed_childless - 1.96 * se) * 100,
                     ymax = (completed_childless + 1.96 * se) * 100),
                position = position_dodge(width = 0.8), width = 0.2) +
  scale_fill_brewer(palette = "Set2", name = "Generation") +
  labs(
    title = "Completed Childlessness by Race/Ethnicity and Generational Status",
    subtitle = "Women aged 40-49, NSFG 2022-2023 (survey-weighted)",
    x = NULL, y = "% Childless (no children ever born)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 15, hjust = 1)) +
  ylim(0, NA)
```

```{r viz-education, fig.width=10, fig.height=5}
educ_rates <- svyby(~completed_childless, ~educ, nsfg_design, svymean, na.rm = TRUE)
ed_df <- as.data.frame(educ_rates)

ggplot(ed_df, aes(x = educ, y = completed_childless * 100, fill = educ)) +
  geom_col(width = 0.6) +
  geom_errorbar(aes(ymin = (completed_childless - 1.96 * se) * 100,
                     ymax = (completed_childless + 1.96 * se) * 100),
                width = 0.2) +
  geom_text(aes(label = sprintf("%.1f%%", completed_childless * 100)),
            vjust = -1.2, size = 3.5) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(
    title = "Completed Childlessness by Education (Women 40-49)",
    subtitle = "NSFG 2022-2023 (survey-weighted)",
    x = NULL, y = "% Childless"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  ylim(0, NA)
```

---

# Sequential Logistic Regression

## Model 1: Unconditional Racial Gradient

```{r model1}
m1 <- svyglm(completed_childless ~ race_eth,
              design = nsfg_design, family = quasibinomial())
m1_results <- tidy(m1, conf.int = TRUE, exponentiate = TRUE)
```

```{r model1-table}
kable(m1_results %>% mutate(across(where(is.numeric), ~round(., 4))),
  caption = "Model 1: Unconditional Racial Gradient (Odds Ratios)")
```

## Model 2: Adding Generational Status

```{r model2}
m2 <- svyglm(completed_childless ~ race_eth + gen_status,
              design = nsfg_design, family = quasibinomial())
m2_results <- tidy(m2, conf.int = TRUE, exponentiate = TRUE)
```

```{r model2-table}
kable(m2_results %>% mutate(across(where(is.numeric), ~round(., 4))),
  caption = "Model 2: Race + Generational Status (Odds Ratios)")
```

```{r model2-compare, results='asis'}
m1_hisp <- m1_results %>% filter(term == "race_ethHispanic")
m2_hisp <- m2_results %>% filter(term == "race_ethHispanic")

if (nrow(m1_hisp) > 0 & nrow(m2_hisp) > 0) {
  pct_change_hisp <- round(100 * (m2_hisp$estimate - m1_hisp$estimate) /
    (m1_hisp$estimate - 1), 1)
  cat(sprintf("**Attenuation of Hispanic coefficient:** The Hispanic OR changed from
%.3f (Model 1) to %.3f (Model 2) when generational status was added, representing
approximately %.1f%% attenuation of the Hispanic-White gap. Generational composition
%s a substantial portion of the racial difference in childlessness.\n\n",
    m1_hisp$estimate, m2_hisp$estimate, abs(pct_change_hisp),
    ifelse(abs(pct_change_hisp) > 20, "accounts for", "does not fully explain")))
}

m1_black <- m1_results %>% filter(term == "race_ethNH Black")
m2_black <- m2_results %>% filter(term == "race_ethNH Black")

if (nrow(m1_black) > 0 & nrow(m2_black) > 0) {
  pct_change_black <- round(100 * (m2_black$estimate - m1_black$estimate) /
    (m1_black$estimate - 1), 1)
  cat(sprintf("**Attenuation of Black coefficient:** The Black OR changed from %.3f
(Model 1) to %.3f (Model 2), representing %.1f%% attenuation. %s\n\n",
    m1_black$estimate, m2_black$estimate, abs(pct_change_black),
    ifelse(abs(pct_change_black) < 10,
      "As hypothesized, generational status contributes minimally to the Black-White gap, consistent with the predominantly native-born composition of the Black population.",
      "Generational status contributes meaningfully to the Black-White gap, suggesting that Black immigrant fertility plays a larger role than expected.")))
}
```

## Model 3: Full Model with Controls

```{r model3}
m3 <- svyglm(completed_childless ~ race_eth + gen_status +
                college + currently_married + employed +
                poverty_pct + religion + age + wave_f,
              design = nsfg_design, family = quasibinomial())
m3_results <- tidy(m3, conf.int = TRUE, exponentiate = TRUE)
```

```{r model3-table}
kable(m3_results %>% mutate(across(where(is.numeric), ~round(., 4))),
  caption = "Model 3: Full Model with Controls (Odds Ratios)")
```

```{r model3-fit, results='asis'}
cat(sprintf("**Model 3 N:** %s\n", formatC(nobs(m3), big.mark = ",")))
```

## Model Comparison

```{r model-comparison, fig.width=10, fig.height=6}
race_terms <- c("race_ethHispanic", "race_ethNH Black", "race_ethNH Other/Multiple")

comp_df <- bind_rows(
  m1_results %>% filter(term %in% race_terms) %>% mutate(model = "M1: Race Only"),
  m2_results %>% filter(term %in% race_terms) %>% mutate(model = "M2: + Generation"),
  m3_results %>% filter(term %in% race_terms) %>% mutate(model = "M3: Full Controls")
)

if (nrow(comp_df) > 0) {
  ggplot(comp_df, aes(x = term, y = estimate, color = model)) +
    geom_point(position = position_dodge(width = 0.5), size = 3) +
    geom_errorbar(aes(ymin = conf.low, ymax = conf.high),
                  position = position_dodge(width = 0.5), width = 0.2) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
    scale_color_manual(values = c("#E74C3C", "#3498DB", "#2ECC71")) +
    labs(
      title = "Attenuation of Race Coefficients Across Models",
      subtitle = "Odds Ratios with 95% CI (reference: NH White)",
      x = NULL, y = "Odds Ratio",
      color = "Model"
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          axis.text.x = element_text(angle = 15, hjust = 1))
}
```

---

# Interaction Models

## Generation x Education

```{r int-educ}
m_int_educ <- tryCatch(
  svyglm(completed_childless ~ race_eth + gen_status * college +
           currently_married + employed + poverty_pct + religion + age + wave_f,
         design = nsfg_design, family = quasibinomial()),
  error = function(e) { message("Education interaction model error: ", conditionMessage(e)); NULL }
)
```

```{r int-educ-results, results='asis'}
if (!is.null(m_int_educ)) {
  int_educ_results <- tidy(m_int_educ, conf.int = TRUE, exponentiate = TRUE)
  int_terms <- int_educ_results %>%
    filter(grepl("gen_status.*college|college.*gen_status", term))

  if (nrow(int_terms) > 0) {
    cat("**Generation x College Interaction Terms:**\n\n")
    for (j in 1:nrow(int_terms)) {
      r <- int_terms[j, ]
      sig <- ifelse(r$p.value < 0.05, "significant", "not significant")
      cat(sprintf("- **%s**: OR = %.3f (p = %.4f, %s)\n",
        r$term, r$estimate, r$p.value, sig))
    }
    any_sig <- any(int_terms$p.value < 0.05)
    cat(sprintf("\n%s of the generation x education interactions reach significance.\n\n",
      ifelse(any_sig, "One or more", "None")))
  } else {
    cat("No interaction terms detected in the model output.\n\n")
  }
} else {
  cat("The education interaction model did not converge.\n\n")
}
```

## Generation x Marriage

```{r int-mar}
m_int_mar <- tryCatch(
  svyglm(completed_childless ~ race_eth + gen_status * currently_married +
           college + employed + poverty_pct + religion + age + wave_f,
         design = nsfg_design, family = quasibinomial()),
  error = function(e) { message("Marriage interaction model error: ", conditionMessage(e)); NULL }
)
```

```{r int-mar-results, results='asis'}
if (!is.null(m_int_mar)) {
  int_mar_results <- tidy(m_int_mar, conf.int = TRUE, exponentiate = TRUE)
  int_terms_m <- int_mar_results %>%
    filter(grepl("gen_status.*married|married.*gen_status", term))

  if (nrow(int_terms_m) > 0) {
    cat("**Generation x Marriage Interaction Terms:**\n\n")
    for (j in 1:nrow(int_terms_m)) {
      r <- int_terms_m[j, ]
      sig <- ifelse(r$p.value < 0.05, "significant", "not significant")
      cat(sprintf("- **%s**: OR = %.3f (p = %.4f, %s)\n",
        r$term, r$estimate, r$p.value, sig))
    }
    any_sig_m <- any(int_terms_m$p.value < 0.05)
    cat(sprintf("\n%s of the generation x marriage interactions reach significance.\n",
      ifelse(any_sig_m, "One or more", "None")))
  } else {
    cat("No interaction terms detected in the model output.\n")
  }
} else {
  cat("The marriage interaction model did not converge.\n")
}
```

---

# Sensitivity: Fertility Intentions

```{r sensitivity, results='asis'}
if ("wants_children" %in% names(nsfg_40)) {
  cat("As a sensitivity check, we add fertility intentions to Model 3.\n\n")

  m_intent <- tryCatch(
    svyglm(completed_childless ~ race_eth + gen_status +
             college + currently_married + employed +
             poverty_pct + religion + age + wave_f + wants_children,
           design = nsfg_design, family = quasibinomial()),
    error = function(e) { cat("Sensitivity model did not converge.\n"); NULL }
  )

  if (!is.null(m_intent)) {
    intent_results <- tidy(m_intent, conf.int = TRUE, exponentiate = TRUE)
    cat("**Sensitivity Model Results:**\n\n")
    for (j in 1:nrow(intent_results)) {
      r <- intent_results[j, ]
      sig <- ifelse(r$p.value < 0.05, "*", "")
      cat(sprintf("- %s: OR = %.3f [%.3f, %.3f] p = %.4f %s\n",
        r$term, r$estimate, r$conf.low, r$conf.high, r$p.value, sig))
    }
  }
} else {
  cat("Fertility intentions variable not available. Sensitivity analysis omitted.\n")
}
```

---

# Discussion and Conclusion

```{r discussion, results='asis'}
cat("This analysis provides the direct nativity test that the SCF-based companion
papers could only hypothesize about. The sequential regression design allows us to
quantify exactly how much of the racial gradient in completed childlessness is
attributable to generational composition.\n\n")

cat("The key findings, evaluated against the three hypotheses:\n\n")

cat("**H1 (Hispanic-White gap and generation):** ")
if (exists("m2_hisp") && nrow(m2_hisp) > 0) {
  cat(sprintf("The Hispanic coefficient attenuated by approximately %.0f%% when
generational status was added (Model 1 to Model 2). ",
    abs(pct_change_hisp)))
  if (abs(pct_change_hisp) > 20) {
    cat("This supports H1: a substantial portion of the Hispanic-White childlessness
gap is attributable to the higher share of first-generation immigrants in the
Hispanic population.\n\n")
  } else {
    cat("The attenuation was modest, suggesting that generational status alone does not
fully account for the Hispanic-White gap.\n\n")
  }
} else {
  cat("Could not be evaluated (coefficient not available).\n\n")
}

cat("**H2 (Black-White gap and generation):** ")
if (exists("m2_black") && nrow(m2_black) > 0) {
  cat(sprintf("The Black coefficient changed by approximately %.0f%% between Models 1
and 2. ",
    abs(pct_change_black)))
  if (abs(pct_change_black) < 15) {
    cat("This supports H2: the Black-White childlessness difference is not substantially
explained by generational status.\n\n")
  } else {
    cat("Contrary to H2, generational status contributes meaningfully to the
Black-White gap.\n\n")
  }
} else {
  cat("Could not be evaluated (coefficient not available).\n\n")
}

cat("**H3 (Assimilation and interaction effects):** ")
if (exists("m_int_educ") && !is.null(m_int_educ)) {
  int_sig <- int_educ_results %>%
    filter(grepl("gen_status.*college", term), p.value < 0.05)
  if (nrow(int_sig) > 0) {
    cat("At least one education x generation interaction was significant, providing
preliminary support for H3.\n\n")
  } else {
    cat("No significant interactions were detected, suggesting the relationship between
education and childlessness is similar across generations.\n\n")
  }
} else {
  cat("Could not be evaluated (interaction model unavailable).\n\n")
}

cat("**Connection to the thesis.** This paper closes the analytical loop opened by
the SCF-based companion papers. The first paper showed that parenthood drives credit
card debt. The second paper showed that the financial balance sheet shapes co-resident
childlessness, with racial differences fully mediated by financial and demographic
controls. This third paper tests the hypothesis that emerged from that mediation:
that nativity composition underlies the racial gradient. Together, the three papers
demonstrate that the economic context of family formation operates through specific
financial mechanisms, is shaped by demographic factors, and reflects population-level
compositional effects that single-dataset analyses cannot fully capture.\n")
```

---

# Limitations

**Women only.** This analysis is restricted to female respondents. The NSFG's male
questionnaire has different variable structures.

**Single wave.** Using only the 2022–2023 wave limits sample size. Pooling with earlier
waves (2015–2017, 2017–2019) would increase power but introduces pooling complexity and
potential period effects.

**Parental nativity.** If parental birthplace variables are not available in the
public-use file, the generational status variable reduces to a foreign-born/native-born
dichotomy, which cannot distinguish second from third-plus generation.

**Cross-sectional design.** Childlessness at 40–49 is near-complete but not perfectly
so. All findings are correlational.

---

# References

Adserà, A., & Ferrer, A. (2015). Immigrants and demography. In *Handbook of the
Economics of International Migration* (Vol. 1A, pp. 315–374). North-Holland.

Camarota, S. A., & Zeigler, K. (2025). The fertility of immigrants and natives in the
United States, 2023. Center for Immigration Studies.

Guzzo, K. B., & Loo, J. (2023). Number of children to women aged 40–44, 1980–2022.
*Family Profiles* (FP-23-29). National Center for Family & Marriage Research.

Hill, L. E., & Johnson, H. P. (2004). Fertility changes among immigrants. *Social
Science Quarterly*, 85(3), 811–826.

Livingston, G. (2015). Childlessness falls, family size grows among highly educated
women. Pew Research Center.

Osterman, M. J. K., et al. (2023). Births: Final data for 2021. *National Vital
Statistics Reports*, 72(1), 1–53.

Parrado, E. A. (2011). How high is Hispanic/Mexican fertility in the United States?
*Demography*, 48(3), 1059–1080.

Parrado, E. A., & Morgan, S. P. (2008). Intergenerational fertility among Hispanic
women. *Demography*, 45(3), 651–671.

Pew Research Center. (2015). Childlessness falls, family size grows among highly
educated women.

Zill, N. (2025). Four million American women reach ages 40 to 50 without children.
Institute for Family Studies.

---

# Appendix: Technical Notes

**Survey Design.** The 2022–2023 NSFG uses `VEST` (stratum) and `VECL` (cluster) with
weight `WGT2022_2023`. We apply `options(survey.lonely.psu = "adjust")` throughout.

**Generational Status.** Constructed from `BRNOUT` (respondent nativity) and, when
available, `MOMFBRN`/`DADFBRN` (parental nativity). First generation = respondent born
outside U.S. Second generation = U.S.-born with at least one foreign-born parent.
Third-plus generation = respondent and both parents U.S.-born.

**Completed Childlessness.** `PARITY == 0` among women 40–49. `PARITY` counts live
births ever had — a biological fertility measure, not co-resident child status.

```{r session-info, results='asis'}
cat("\n\n**Session Information**\n\n")
cat(sprintf("R version: %s\n\n", R.version.string))
cat(sprintf("Platform: %s\n\n", R.version$platform))
cat(sprintf("Key packages: survey %s, haven %s, tidyverse %s\n",
  packageVersion("survey"), packageVersion("haven"), packageVersion("tidyverse")))
```
