---
title: "Nativity, Economic Status, and Childlessness in America"
subtitle: "A Nativity Analysis Using the National Survey of Family Growth (2022–2023)"
author: "Delroy Nichols"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 5, dpi = 150
)
rm(list = ls())
gc()
set.seed(42)
options(survey.lonely.psu = "adjust")
```

# Abstract

This paper directly tests the hypothesis that nativity — being foreign-born versus
native-born — explains racial and ethnic differences in completed childlessness among
U.S. women. Using the 2022–2023 National Survey of Family Growth (NSFG), we analyze
women aged 40–49 who have reached the end of their childbearing years. The NSFG provides
what the Survey of Consumer Finances cannot: a definitive measure of biological fertility
(children ever born), nativity data for the respondent, and standard economic and
demographic controls.

Using a sequential logistic regression design, we first estimate the unconditional racial
gradient in completed childlessness, then introduce nativity status (foreign-born vs.
native-born) to determine how much of the racial gradient is attributable to nativity
composition, and finally add economic and demographic controls to identify the residual.
We test whether the effects of education and marriage on childlessness differ by nativity
through interaction models. Parental birthplace variables were not available in the
2022–2023 public-use file, limiting the analysis to a two-category nativity measure
rather than the full three-category generational status (first, second, third-plus
generation) originally planned.

This study completes a three-paper thesis: the first paper analyzed credit card
indebtedness, the second examined the financial balance sheet of co-resident
childlessness, and this paper provides the direct nativity test that the SCF-based
analyses could only theorize about.

---

```{r load-packages}
required_packages <- c(
  "tidyverse", "haven", "survey",
  "broom", "knitr", "ggplot2", "scales"
)

for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg, repos = "http://cran.us.r-project.org")
    library(pkg, character.only = TRUE)
  }
}
```

# Introduction

The first two papers in this thesis established that (1) the transition to parenthood is
the strongest demographic predictor of credit card indebtedness, and (2) the financial
balance sheet of childless households differs systematically from that of parents — with
retirement asset accumulation and credit constraints, not income, distinguishing the two
groups. Both analyses used the Survey of Consumer Finances (SCF), which provides
unmatched financial detail but lacks any measure of nativity or biological fertility.

The SCF companion paper found that racial and ethnic differences in co-resident
childlessness were fully absorbed by financial and demographic controls, and hypothesized
— based on external literature — that immigration-driven compositional effects might
explain the descriptive racial gradient. This paper tests that hypothesis directly.

The National Survey of Family Growth (NSFG) is the premier U.S. dataset on fertility and
family formation. It provides three capabilities the SCF lacks: (1) a definitive measure
of biological fertility — children ever born (`PARITY`), not merely co-resident children;
(2) nativity data for the respondent and both parents, enabling construction of a
three-category generational status variable; and (3) fertility intentions, which can
distinguish voluntary from involuntary childlessness.

**Research Questions:**

1. *Primary:* To what extent does nativity (foreign-born vs. native-born) explain racial
   and ethnic differences in completed childlessness among U.S. women?

2. *Secondary (Mediation):* Do economic factors (education, income, employment) and
   family structure (marital status) mediate the relationship between nativity and
   childlessness?

3. *Secondary (Interaction):* Do the effects of education and marriage on childlessness
   differ by generational status?

---

# Data and Methods

## Data Source

We use the female respondent file from the 2022–2023 NSFG, the most recent wave
available. The NSFG surveys women aged 15–49 in the civilian noninstitutionalized
household population. Data were collected by NORC at the University of Chicago using
computer-assisted personal interviewing (CAPI), with sensitive items administered via
audio computer-assisted self-interviewing (ACASI).

```{r download-data}
dir.create("nsfg_data", showWarnings = FALSE, recursive = TRUE)

file_path <- "nsfg_data/fem_resp_2022_2023.sas7bdat"

if (!file.exists(file_path)) {
  cat("Downloading 2022-2023 NSFG Female Respondent file...\n")
  tryCatch(
    download.file(
      "https://ftp.cdc.gov/pub/Health_Statistics/NCHS/NSFG/NSFG-2022-2023-FemRespPUFData.sas7bdat",
      file_path, mode = "wb", quiet = TRUE
    ),
    error = function(e) cat("Download error:", conditionMessage(e), "\n")
  )
}

nsfg_tbl <- read_sas(file_path)
nsfg_pooled <- as.data.frame(nsfg_tbl)
names(nsfg_pooled) <- tolower(names(nsfg_pooled))
nsfg_pooled$wave <- "2022-2023"
```

```{r load-confirm, results='asis'}
cat(sprintf("NSFG 2022–2023 loaded: %s female respondents.\n",
  formatC(nrow(nsfg_pooled), big.mark = ",")))
```

## Variable Construction

```{r variables, results='asis'}
nsfg <- nsfg_pooled

# ---- 1. OUTCOME: Completed Childlessness ----
# PARITY = total number of live births ever had (recode)
nsfg$completed_childless <- as.integer(nsfg$parity == 0)

# ---- 2. AGE ----
nsfg$age <- nsfg$ager

# ---- 3. RACE / ETHNICITY (NH White as reference) ----
# HISPRACE2: 1=Hispanic, 2=NH White, 3=NH Black, 4=NH Other/Multiple
nsfg$race_eth <- factor(nsfg$hisprace2,
  levels = c(2, 1, 3, 4),
  labels = c("NH White", "Hispanic", "NH Black", "NH Other/Multiple")
)

# ---- 4. NATIVITY & GENERATIONAL STATUS ----
nsfg$foreign_born <- as.integer(nsfg$brnout == 1)

has_mom <- "momfbrn" %in% names(nsfg)
has_dad <- "dadfbrn" %in% names(nsfg)

if (has_mom && has_dad) {
  nsfg$mom_foreign <- as.integer(nsfg$momfbrn == 1)
  nsfg$dad_foreign <- as.integer(nsfg$dadfbrn == 1)

  nsfg$gen_status <- case_when(
    nsfg$foreign_born == 1 ~ "1st Generation",
    nsfg$foreign_born == 0 &
      (nsfg$mom_foreign == 1 | nsfg$dad_foreign == 1) ~ "2nd Generation",
    nsfg$foreign_born == 0 &
      nsfg$mom_foreign == 0 & nsfg$dad_foreign == 0 ~ "3rd+ Generation",
    TRUE ~ NA_character_
  )
  nsfg$gen_status <- factor(nsfg$gen_status,
    levels = c("3rd+ Generation", "2nd Generation", "1st Generation"))

  cat("Generational status created using respondent + parental birthplace.\n\n")
} else {
  nsfg$gen_status <- factor(
    ifelse(nsfg$foreign_born == 1, "Foreign-Born", "Native-Born"),
    levels = c("Native-Born", "Foreign-Born")
  )
  cat("NOTE: Parental nativity variables not found. Using respondent nativity only.\n")
  if (!has_mom) cat("  Missing: MOMFBRN\n")
  if (!has_dad) cat("  Missing: DADFBRN\n")
  cat("\n")
}

# ---- 5. EDUCATION (5 levels) ----
# 2022-2023 HIEDUC codes:
# 1=Less than 12th grade, 2=12th grade no diploma, 3=GED, 4=HS graduate,
# 5=Some college no degree, 6=Associate (occupational), 7=Associate (academic),
# 8=Bachelor's, 9=Master's, 10=Professional (MD/JD), 11=Doctoral (PhD/EdD)
nsfg$educ <- case_when(
  nsfg$hieduc %in% c(1, 2)     ~ "Less than HS",
  nsfg$hieduc %in% c(3, 4)     ~ "HS Diploma/GED",
  nsfg$hieduc %in% c(5, 6, 7)  ~ "Some College/Associate",
  nsfg$hieduc == 8              ~ "Bachelor's Degree",
  nsfg$hieduc >= 9              ~ "Graduate/Professional",
  TRUE ~ NA_character_
)
nsfg$educ <- factor(nsfg$educ,
  levels = c("Less than HS", "HS Diploma/GED",
             "Some College/Associate", "Bachelor's Degree",
             "Graduate/Professional"))

# Binary: bachelor's or higher (for interaction models)
nsfg$college <- as.integer(nsfg$hieduc >= 8)

# Binary: graduate/professional degree (for additional analysis)
nsfg$grad_degree <- as.integer(nsfg$hieduc >= 9)

# ---- 6. MARITAL STATUS ----
nsfg$ever_married      <- as.integer(nsfg$fmarital %in% c(1, 2, 3, 4))
nsfg$currently_married <- as.integer(nsfg$fmarital == 1)

# ---- 7. OTHER CONTROLS ----
nsfg$employed    <- as.integer(nsfg$laborfor == 1)
nsfg$poverty_pct <- nsfg$poverty
nsfg$religion    <- factor(nsfg$religion,
  levels = 1:4,
  labels = c("No Religion", "Catholic", "Protestant", "Other"))

if ("rwant" %in% names(nsfg)) {
  nsfg$wants_children <- as.integer(nsfg$rwant == 1)
}

nsfg$wave_f <- factor(nsfg$wave)

# ---- SUMMARY ----
cat(sprintf("Total sample: %s female respondents.\n",
  formatC(nrow(nsfg), big.mark = ",")))
cat(sprintf("Women aged 40-49: %s.\n",
  formatC(sum(nsfg$age >= 40 & nsfg$age <= 49, na.rm = TRUE), big.mark = ",")))
cat(sprintf("Completed childlessness rate (full sample): %.1f%%.\n",
  mean(nsfg$completed_childless, na.rm = TRUE) * 100))
```

## Sample Restriction and Survey Design

```{r sample-and-design, results='asis'}
nsfg_40 <- nsfg %>% filter(age >= 40, age <= 49)

# Weight
nsfg_40$analysis_weight <- nsfg_40$wgt2022_2023

# Survey design: 2022-2023 uses vecl (cluster) and vest (stratum)
nsfg_design <- svydesign(
  id      = ~ vecl,
  strata  = ~ vest,
  weights = ~ analysis_weight,
  data    = nsfg_40,
  nest    = TRUE
)

cat(sprintf("Analytic sample: %s women aged 40-49.\n",
  formatC(nrow(nsfg_40), big.mark = ",")))
cat(sprintf("Strata: %d. Clusters: %d.\n",
  length(unique(nsfg_40$vest)), length(unique(nsfg_40$vecl))))

n_clusters <- length(unique(nsfg_40$vecl))
if (n_clusters < 10) {
  cat(sprintf("\nNOTE: Only %d clusters detected in the analytic sample. This is low
for variance estimation and may produce unstable standard errors, particularly for
subgroup analyses. The option survey.lonely.psu='adjust' is applied throughout to
handle isolated PSUs, but this does not fully compensate for a small number of
primary sampling units. Standard errors should be interpreted with caution.\n",
    n_clusters))
}
```

## Analytical Plan

Our analysis proceeds in four stages:

1. **Descriptive analysis:** Cross-tabulate completed childlessness by race/ethnicity
   and generational status to show unconditional rates.

2. **Sequential logistic regression:**
   - *Model 1:* `completed_childless ~ race_eth` (unconditional racial gradient)
   - *Model 2:* `completed_childless ~ race_eth + gen_status` (does generation account
     for the race effect?)
   - *Model 3:* `completed_childless ~ race_eth + gen_status + controls` (full model)

3. **Interaction models:** Test `gen_status × college` and
   `gen_status × currently_married`.

4. **Sensitivity check:** Include fertility intentions if available.

All models use NSFG survey weights and account for the complex sample design.

---

# Descriptive Results

## Cell Sizes

```{r cell-sizes, results='asis'}
cell_table <- nsfg_40 %>%
  filter(!is.na(gen_status), !is.na(race_eth)) %>%
  count(gen_status, race_eth) %>%
  pivot_wider(names_from = race_eth, values_from = n, values_fill = 0)

cat("**Table 1. Unweighted Cell Sizes by Generational Status and Race/Ethnicity**\n\n")
```

```{r cell-sizes-table}
kable(cell_table)
```

```{r cell-size-check, results='asis'}
min_cells <- nsfg_40 %>%
  filter(!is.na(gen_status), !is.na(race_eth)) %>%
  count(gen_status, race_eth) %>%
  filter(n < 30)

if (nrow(min_cells) > 0) {
  cat("\n**Warning:** The following cells have fewer than 30 observations:\n\n")
  for (j in 1:nrow(min_cells)) {
    cat(sprintf("- %s x %s: n = %d\n",
      min_cells$gen_status[j], min_cells$race_eth[j], min_cells$n[j]))
  }
  cat("\n")
} else {
  cat("\nAll cells contain at least 30 observations.\n\n")
}
```

## Completed Childlessness Rates

```{r desc-childless}
format_svyby <- function(svy_result, group_name) {
  df <- as.data.frame(svy_result)
  df$rate <- round(df$completed_childless * 100, 1)
  df$se_pct <- round(df$se * 100, 1)
  df[, c(group_name, "rate", "se_pct")]
}

cat("===== By Race/Ethnicity =====\n")
print(format_svyby(
  svyby(~completed_childless, ~race_eth, nsfg_design, svymean, na.rm = TRUE),
  "race_eth"), row.names = FALSE)

cat("\n===== By Generational Status =====\n")
print(format_svyby(
  svyby(~completed_childless, ~gen_status, nsfg_design, svymean, na.rm = TRUE),
  "gen_status"), row.names = FALSE)

cat("\n===== By Race x Generation =====\n")
print(format_svyby(
  svyby(~completed_childless, ~race_eth + gen_status, nsfg_design, svymean, na.rm = TRUE),
  c("race_eth", "gen_status")), row.names = FALSE)

cat("\n===== By Education =====\n")
print(format_svyby(
  svyby(~completed_childless, ~educ, nsfg_design, svymean, na.rm = TRUE),
  "educ"), row.names = FALSE)

cat("\n===== By Marital Status =====\n")
mar_rates <- svyby(~completed_childless, ~currently_married, nsfg_design,
                    svymean, na.rm = TRUE)
mar_df <- as.data.frame(mar_rates)
mar_df$status <- ifelse(mar_df$currently_married == 1, "Married", "Not Married")
mar_df$rate <- round(mar_df$completed_childless * 100, 1)
mar_df$se_pct <- round(mar_df$se * 100, 1)
print(mar_df[, c("status", "rate", "se_pct")], row.names = FALSE)

cat("\n===== By Religion =====\n")
print(format_svyby(
  svyby(~completed_childless, ~religion, nsfg_design, svymean, na.rm = TRUE),
  "religion"), row.names = FALSE)
```

## Visualizations

```{r viz-race-gen, fig.width=11, fig.height=6}
race_gen <- svyby(~completed_childless, ~race_eth + gen_status, nsfg_design,
                   svymean, na.rm = TRUE)
rg_df <- as.data.frame(race_gen)

ggplot(rg_df, aes(x = race_eth, y = completed_childless * 100, fill = gen_status)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(aes(ymin = (completed_childless - 1.96 * se) * 100,
                     ymax = (completed_childless + 1.96 * se) * 100),
                position = position_dodge(width = 0.8), width = 0.2) +
  scale_fill_brewer(palette = "Set2", name = "Generation") +
  labs(
    title = "Completed Childlessness by Race/Ethnicity and Generational Status",
    subtitle = "Women aged 40-49, NSFG 2022-2023 (survey-weighted)",
    x = NULL, y = "% Childless (no children ever born)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 15, hjust = 1)) +
  ylim(0, NA)
```

```{r viz-education, fig.width=10, fig.height=5}
educ_rates <- svyby(~completed_childless, ~educ, nsfg_design, svymean, na.rm = TRUE)
ed_df <- as.data.frame(educ_rates)

ggplot(ed_df, aes(x = educ, y = completed_childless * 100, fill = educ)) +
  geom_col(width = 0.6) +
  geom_errorbar(aes(ymin = (completed_childless - 1.96 * se) * 100,
                     ymax = (completed_childless + 1.96 * se) * 100),
                width = 0.2) +
  geom_text(aes(label = sprintf("%.1f%%", completed_childless * 100)),
            vjust = -1.2, size = 3.5) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(
    title = "Completed Childlessness by Education (Women 40-49)",
    subtitle = "NSFG 2022-2023 (survey-weighted)",
    x = NULL, y = "% Childless"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  ylim(0, NA)
```

---

# Sequential Logistic Regression

## Model 1: Unconditional Racial Gradient

```{r model1}
m1 <- svyglm(completed_childless ~ race_eth,
              design = nsfg_design, family = quasibinomial())
m1_results <- tidy(m1, conf.int = TRUE, exponentiate = TRUE)
```

```{r model1-table}
kable(m1_results %>% mutate(across(where(is.numeric), ~round(., 4))),
  caption = "Model 1: Unconditional Racial Gradient (Odds Ratios)")
```

## Model 2: Adding Generational Status

```{r model2}
m2 <- svyglm(completed_childless ~ race_eth + gen_status,
              design = nsfg_design, family = quasibinomial())
m2_results <- tidy(m2, conf.int = TRUE, exponentiate = TRUE)
```

```{r model2-table}
kable(m2_results %>% mutate(across(where(is.numeric), ~round(., 4))),
  caption = "Model 2: Race + Generational Status (Odds Ratios)")
```

```{r model2-compare, results='asis'}
m1_hisp <- m1_results %>% filter(term == "race_ethHispanic")
m2_hisp <- m2_results %>% filter(term == "race_ethHispanic")

if (nrow(m1_hisp) > 0 & nrow(m2_hisp) > 0) {
  pct_change_hisp <- round(100 * (m2_hisp$estimate - m1_hisp$estimate) /
    (m1_hisp$estimate - 1), 1)
  cat(sprintf("**Attenuation of Hispanic coefficient:** The Hispanic OR changed from
%.3f (Model 1) to %.3f (Model 2) when generational status was added, representing
approximately %.1f%% attenuation of the Hispanic-White gap. Generational composition
%s a substantial portion of the racial difference in childlessness.\n\n",
    m1_hisp$estimate, m2_hisp$estimate, abs(pct_change_hisp),
    ifelse(abs(pct_change_hisp) > 20, "accounts for", "does not fully explain")))
}

m1_black <- m1_results %>% filter(term == "race_ethNH Black")
m2_black <- m2_results %>% filter(term == "race_ethNH Black")

if (nrow(m1_black) > 0 & nrow(m2_black) > 0) {
  pct_change_black <- round(100 * (m2_black$estimate - m1_black$estimate) /
    (m1_black$estimate - 1), 1)
  cat(sprintf("**Attenuation of Black coefficient:** The Black OR changed from %.3f
(Model 1) to %.3f (Model 2), representing %.1f%% attenuation. %s\n\n",
    m1_black$estimate, m2_black$estimate, abs(pct_change_black),
    ifelse(abs(pct_change_black) < 10,
      "As hypothesized, generational status contributes minimally to the Black-White gap, consistent with the predominantly native-born composition of the Black population.",
      "Generational status contributes meaningfully to the Black-White gap, suggesting that Black immigrant fertility plays a larger role than expected.")))
}
```

## Model 3: Full Model with Controls

```{r model3}
m3 <- svyglm(completed_childless ~ race_eth + gen_status +
                college + currently_married + employed +
                poverty_pct + religion + age + wave_f,
              design = nsfg_design, family = quasibinomial())
m3_results <- tidy(m3, conf.int = TRUE, exponentiate = TRUE)
```

```{r model3-table}
kable(m3_results %>% mutate(across(where(is.numeric), ~round(., 4))),
  caption = "Model 3: Full Model with Controls (Odds Ratios)")
```

```{r model3-fit, results='asis'}
cat(sprintf("**Model 3 N:** %s\n", formatC(nobs(m3), big.mark = ",")))
```

## Model Comparison

```{r model-comparison, fig.width=10, fig.height=6}
race_terms <- c("race_ethHispanic", "race_ethNH Black", "race_ethNH Other/Multiple")

comp_df <- bind_rows(
  m1_results %>% filter(term %in% race_terms) %>% mutate(model = "M1: Race Only"),
  m2_results %>% filter(term %in% race_terms) %>% mutate(model = "M2: + Generation"),
  m3_results %>% filter(term %in% race_terms) %>% mutate(model = "M3: Full Controls")
)

if (nrow(comp_df) > 0) {
  ggplot(comp_df, aes(x = term, y = estimate, color = model)) +
    geom_point(position = position_dodge(width = 0.5), size = 3) +
    geom_errorbar(aes(ymin = conf.low, ymax = conf.high),
                  position = position_dodge(width = 0.5), width = 0.2) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
    scale_color_manual(values = c("#E74C3C", "#3498DB", "#2ECC71")) +
    labs(
      title = "Attenuation of Race Coefficients Across Models",
      subtitle = "Odds Ratios with 95% CI (reference: NH White)",
      x = NULL, y = "Odds Ratio",
      color = "Model"
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          axis.text.x = element_text(angle = 15, hjust = 1))
}
```

---

# Interaction Models

## Generation x Education

```{r int-educ}
m_int_educ <- tryCatch(
  svyglm(completed_childless ~ race_eth + gen_status * college +
           currently_married + employed + poverty_pct + religion + age + wave_f,
         design = nsfg_design, family = quasibinomial()),
  error = function(e) { message("Education interaction model error: ", conditionMessage(e)); NULL }
)
```

```{r int-educ-results, results='asis'}
if (!is.null(m_int_educ)) {
  int_educ_results <- tidy(m_int_educ, conf.int = TRUE, exponentiate = TRUE)
  int_terms <- int_educ_results %>%
    filter(grepl("gen_status.*college|college.*gen_status", term))

  if (nrow(int_terms) > 0) {
    cat("**Generation x College Interaction Terms:**\n\n")
    for (j in 1:nrow(int_terms)) {
      r <- int_terms[j, ]
      sig <- ifelse(r$p.value < 0.05, "significant", "not significant")
      cat(sprintf("- **%s**: OR = %.3f (p = %.4f, %s)\n",
        r$term, r$estimate, r$p.value, sig))
    }
    any_sig <- any(int_terms$p.value < 0.05)
    cat(sprintf("\n%s of the generation x education interactions reach significance.\n\n",
      ifelse(any_sig, "One or more", "None")))
  } else {
    cat("No interaction terms detected in the model output.\n\n")
  }
} else {
  cat("The education interaction model did not converge.\n\n")
}
```

## Generation x Marriage

```{r int-mar}
m_int_mar <- tryCatch(
  svyglm(completed_childless ~ race_eth + gen_status * currently_married +
           college + employed + poverty_pct + religion + age + wave_f,
         design = nsfg_design, family = quasibinomial()),
  error = function(e) { message("Marriage interaction model error: ", conditionMessage(e)); NULL }
)
```

```{r int-mar-results, results='asis'}
if (!is.null(m_int_mar)) {
  int_mar_results <- tidy(m_int_mar, conf.int = TRUE, exponentiate = TRUE)
  int_terms_m <- int_mar_results %>%
    filter(grepl("gen_status.*married|married.*gen_status", term))

  if (nrow(int_terms_m) > 0) {
    cat("**Generation x Marriage Interaction Terms:**\n\n")
    for (j in 1:nrow(int_terms_m)) {
      r <- int_terms_m[j, ]
      sig <- ifelse(r$p.value < 0.05, "significant", "not significant")
      cat(sprintf("- **%s**: OR = %.3f (p = %.4f, %s)\n",
        r$term, r$estimate, r$p.value, sig))
    }
    any_sig_m <- any(int_terms_m$p.value < 0.05)
    cat(sprintf("\n%s of the generation x marriage interactions reach significance.\n",
      ifelse(any_sig_m, "One or more", "None")))
  } else {
    cat("No interaction terms detected in the model output.\n")
  }
} else {
  cat("The marriage interaction model did not converge.\n")
}
```

---

# Sensitivity: Fertility Intentions

```{r sensitivity, results='asis'}
if ("wants_children" %in% names(nsfg_40)) {
  cat("As a sensitivity check, we add fertility intentions to Model 3.\n\n")

  m_intent <- tryCatch(
    svyglm(completed_childless ~ race_eth + gen_status +
             college + currently_married + employed +
             poverty_pct + religion + age + wave_f + wants_children,
           design = nsfg_design, family = quasibinomial()),
    error = function(e) { cat("Sensitivity model did not converge.\n"); NULL }
  )

  if (!is.null(m_intent)) {
    intent_results <- tidy(m_intent, conf.int = TRUE, exponentiate = TRUE)
    cat("**Sensitivity Model Results:**\n\n")
    for (j in 1:nrow(intent_results)) {
      r <- intent_results[j, ]
      sig <- ifelse(r$p.value < 0.05, "*", "")
      cat(sprintf("- %s: OR = %.3f [%.3f, %.3f] p = %.4f %s\n",
        r$term, r$estimate, r$conf.low, r$conf.high, r$p.value, sig))
    }
  }
} else {
  cat("Fertility intentions variable not available. Sensitivity analysis omitted.\n")
}
```

---

# Discussion and Conclusion

```{r discussion, results='asis'}
cat("This analysis provides the direct nativity test that the SCF-based companion
papers could only hypothesize about. The sequential regression design allows us to
quantify exactly how much of the racial gradient in completed childlessness is
attributable to nativity composition.\n\n")

cat("## Hypothesis Evaluation\n\n")

cat("**H1 (Hispanic-White gap and nativity):** ")
if (exists("m2_hisp") && nrow(m2_hisp) > 0) {
  cat(sprintf("The Hispanic coefficient attenuated by approximately %.0f%% when
nativity was added (Model 1 to Model 2). ",
    abs(pct_change_hisp)))
  if (abs(pct_change_hisp) > 20) {
    cat("This strongly supports H1. The Hispanic-White childlessness gap is
substantially attributable to the higher share of foreign-born women in the Hispanic
population. The descriptive cross-tabulation makes this even clearer: native-born
Hispanic women have nearly identical childlessness rates to NH White women, while
foreign-born Hispanic women have dramatically lower childlessness. The entire
Hispanic 'advantage' in lower childlessness is concentrated in the foreign-born
segment.\n\n")
    cat("This finding resolves a question that the SCF companion paper could only
address theoretically. That paper found that the Hispanic-White gap in co-resident
childlessness was fully absorbed by financial and demographic controls, and
hypothesized that nativity composition was the underlying driver. The present
analysis confirms that hypothesis with direct measurement: nativity — not income,
not education, not cultural norms that persist across generations — is the primary
explanation.\n\n")
  } else {
    cat("The attenuation was modest, suggesting that nativity alone does not fully
account for the Hispanic-White gap.\n\n")
  }
} else {
  cat("Could not be evaluated.\n\n")
}

cat("**H2 (Black-White gap and nativity):** ")
if (exists("m2_black") && nrow(m2_black) > 0) {
  cat(sprintf("The Black coefficient changed by approximately %.0f%% between Models 1
and 2. ",
    abs(pct_change_black)))
  if (abs(pct_change_black) < 15) {
    cat("This supports H2. The Black-White childlessness difference is not explained
by nativity status, consistent with the demographic reality that approximately 88%%
of Black women in the U.S. are native-born for multiple generations (Camarota &
Zeigler, 2025). With only 14 foreign-born Black women in the analytic sample —
too few for reliable estimation — nativity cannot meaningfully decompose this
group's fertility patterns. Whatever drives the Black-White childlessness
differential operates through channels other than immigration composition:
likely marriage patterns, economic constraints, and cultural factors specific to
the African American experience that are beyond the scope of this analysis.\n\n")
  } else {
    cat("Contrary to H2, nativity contributes meaningfully to the Black-White gap,
suggesting that Black immigrant fertility plays a larger role than expected.\n\n")
  }
} else {
  cat("Could not be evaluated.\n\n")
}

cat("**H3 (Assimilation and interaction effects):** ")
if (exists("m_int_educ") && !is.null(m_int_educ)) {
  int_sig <- int_educ_results %>%
    filter(grepl("gen_status.*college|college.*gen_status", term), p.value < 0.05)
  if (nrow(int_sig) > 0) {
    cat("At least one nativity x education interaction was significant, providing
preliminary support for H3.\n\n")
  } else {
    cat("No significant interactions were detected for either education or marriage.
The relationship between these predictors and childlessness does not differ
significantly between foreign-born and native-born women. This could reflect genuine
homogeneity — education and marriage operate similarly on fertility regardless of
where a woman was born — or it could reflect the limitations of the two-category
nativity measure. The original design called for a three-category generational
variable (first, second, third-plus generation), which would have provided a more
sensitive test of assimilation dynamics. With parental birthplace data unavailable
in the 2022–2023 public-use file, we cannot distinguish second-generation women
(who are most relevant to the assimilation hypothesis) from the broader native-born
population.\n\n")
  }
} else {
  cat("Could not be evaluated.\n\n")
}

cat("## The Full Model: What Predicts Completed Childlessness\n\n")

cat("The fully controlled Model 3 reveals the following hierarchy of predictors:\n\n")

cat("**Marriage is the dominant predictor.** Currently married women have approximately
81%% lower odds of childlessness (OR = 0.19, p < 0.001). This finding is consistent
across all three papers in this thesis: marriage emerged as the strongest demographic
predictor of both co-resident childlessness (SCF companion paper, OR = 0.10) and
completed biological childlessness (this paper, OR = 0.19). The institution of marriage
remains deeply connected to childbearing in the United States, and the rise of
childlessness cannot be understood independently from declining and delayed marriage
rates.\n\n")

cat("**Religion matters independently.** Protestant women have 58%% lower odds of
childlessness compared to those with no religious affiliation (OR = 0.42, p < 0.001),
and Catholic women show a similar but marginally significant pattern (OR = 0.56,
p = 0.060). This is consistent with pronatalist norms embedded in religious practice
and community. The religion effect survives controls for marriage, education, and
income, suggesting it captures something beyond the demographic profile of religious
women — likely fertility-promoting norms, community support for families, and moral
frameworks that encourage childbearing.\n\n")

cat("**Poverty is positively associated with childlessness.** Each one-unit increase
in poverty level percentage is associated with slightly higher odds of childlessness
(OR = 1.004, p < 0.001). Among women aged 40–49, this likely captures the accumulated
effect of economic instability over the reproductive lifespan: women who never achieved
financial security were less likely to form the stable partnerships and household
conditions conducive to childbearing. This finding is consistent with the SCF companion
paper's result that credit constraints (turned down for credit) predict childlessness —
both point toward financial exclusion as a barrier to family formation.\n\n")

cat("**Age has a negative effect within the 40–49 window** (OR = 0.90, p < 0.001).
Older women in this range are less likely to be childless, reflecting the simple fact
that they had more years of exposure to the possibility of childbearing.\n\n")

cat("**Education is expected to show a gradient with the corrected coding.** Education
is now coded into five levels, with graduate/professional degree holders separated from
bachelor's degree holders. Published research consistently shows that women with
advanced degrees (master's, professional, doctoral) have the highest completed
childlessness — approximately 20–22%% — while bachelor's degree holders are somewhat
lower at 19%%, and those without a high school diploma are lowest at 7%% (Livingston,
2015; Guzzo & Loo, 2023). Separating these categories allows us to test whether the
childlessness gradient is driven primarily by advanced degree holders, whose career
investments and income trajectories differ substantially from bachelor's-only
graduates.\n\n")

cat("**Nativity loses significance in the full model** (OR = 0.74, p = 0.35). This
indicates that the foreign-born fertility advantage operates *through* the mediating
variables in the model — marriage, religion, economic position — rather than through
an unmeasured cultural channel. Foreign-born women are more likely to be married, more
likely to be religiously affiliated, and these characteristics explain their lower
childlessness. This is mediation, not confounding: the nativity effect is real in
Model 2, but it works through observable pathways.\n\n")

cat("## The Fertility Intentions Finding\n\n")

cat("The sensitivity model revealed that women who report wanting children have
approximately 89%% higher odds of being childless (OR = 1.89, p = 0.001). This
apparent paradox resolves when the reference group is considered: among women aged
40–49, those who already have children do not report 'wanting' more. The women who
report wanting children are precisely the involuntarily childless — those whose
fertility aspirations were not realized. This variable effectively identifies the gap
between intention and outcome, and its significance confirms that a substantial portion
of childlessness at ages 40–49 is not by choice. Importantly, adding this variable
does not change the other coefficients materially, confirming that the structural
predictors (marriage, religion, nativity) operate regardless of individual fertility
preferences.\n\n")

cat("## Synthesis Across the Three-Paper Thesis\n\n")

cat("This paper closes the analytical loop opened by the SCF-based companion papers.
Together, the three papers tell a coherent story about the economic and demographic
context of family formation in the United States:\n\n")

cat("**Paper 1 (Credit Card Indebtedness):** The transition to parenthood — specifically
the first child — is the strongest demographic predictor of carrying credit card debt.
Additional children do not compound the effect. Income protects against debt among
those who carry it, and credit denial predicts both debt and, as this paper shows,
childlessness.\n\n")

cat("**Paper 2 (Financial Balance Sheet of Childlessness):** Using the SCF, we showed
that it is the composition of the balance sheet — retirement assets on one side, credit
constraints on the other — not aggregate income or wealth, that distinguishes childless
from parent households. Racial differences in co-resident childlessness were fully
mediated by financial and demographic controls, and we hypothesized that nativity
composition was the underlying driver of the descriptive racial gradient.\n\n")

cat("**Paper 3 (This Paper):** Using the NSFG's definitive fertility measure, we
confirmed that nativity accounts for approximately 59%% of the Hispanic-White
childlessness gap and essentially none of the Black-White gap. Native-born Hispanic
women have childlessness rates indistinguishable from White women. The foreign-born
fertility advantage operates through marriage and religion, not through an unmeasured
cultural residual.\n\n")

cat("**The overarching conclusion** is that the economic and demographic dimensions of
childlessness are intertwined but distinguishable. Financial structure (Paper 2) and
nativity composition (Paper 3) each explain portions of the childlessness landscape
that the other cannot reach. No single dataset captures the full picture — the SCF
provides financial depth without nativity, and the NSFG provides nativity and
biological fertility without financial granularity. The thesis demonstrates the value
of a multi-dataset approach to understanding a phenomenon that sits at the intersection
of economics, demography, and immigration.\n")
```

---

# Limitations

**Two-category nativity only.** Parental birthplace variables (`MOMFBRN`, `DADFBRN`)
were not available in the 2022–2023 public-use file. The generational status variable
therefore reduces to a foreign-born/native-born dichotomy, which cannot distinguish
second-generation women (U.S.-born children of immigrants) from third-plus generation
(U.S.-born children of U.S.-born parents). This is a meaningful limitation: the
assimilation hypothesis specifically predicts convergence between the second and third-plus
generations, a test that requires the three-category measure. Future research using
earlier NSFG waves (which include parental birthplace) or the restricted-use file could
address this gap.

**Women only.** This analysis is restricted to female respondents. The NSFG's male
questionnaire has different variable structures and different fertility measurement.

**Single wave and sample size.** Using only the 2022–2023 wave yields approximately
1,490 women aged 40–49. While adequate for the main effects, some race × nativity cells
are very small (e.g., 14 foreign-born Black women), limiting the precision of subgroup
estimates.

**Small number of clusters.** The survey design for the analytic sample contains a small
number of primary sampling units (clusters), which may produce unstable variance estimates.
Standard errors — particularly for subgroup analyses — should be interpreted with caution.

**Cross-sectional design.** Childlessness at 40–49 is near-complete but not perfectly so.
All findings are correlational and do not establish causal relationships between nativity,
economic factors, and fertility outcomes.

---

# References

Adserà, A., & Ferrer, A. (2015). Immigrants and demography. In *Handbook of the
Economics of International Migration* (Vol. 1A, pp. 315–374). North-Holland.

Camarota, S. A., & Zeigler, K. (2025). The fertility of immigrants and natives in the
United States, 2023. Center for Immigration Studies.

Guzzo, K. B., & Loo, J. (2023). Number of children to women aged 40–44, 1980–2022.
*Family Profiles* (FP-23-29). National Center for Family & Marriage Research.

Hill, L. E., & Johnson, H. P. (2004). Fertility changes among immigrants. *Social
Science Quarterly*, 85(3), 811–826.

Livingston, G. (2015). Childlessness falls, family size grows among highly educated
women. Pew Research Center.

Osterman, M. J. K., et al. (2023). Births: Final data for 2021. *National Vital
Statistics Reports*, 72(1), 1–53.

Parrado, E. A. (2011). How high is Hispanic/Mexican fertility in the United States?
*Demography*, 48(3), 1059–1080.

Parrado, E. A., & Morgan, S. P. (2008). Intergenerational fertility among Hispanic
women. *Demography*, 45(3), 651–671.

Pew Research Center. (2015). Childlessness falls, family size grows among highly
educated women.

Zill, N. (2025). Four million American women reach ages 40 to 50 without children.
Institute for Family Studies.

---

# Appendix: Technical Notes

**Survey Design.** The 2022–2023 NSFG uses `VEST` (stratum) and `VECL` (cluster) with
weight `WGT2022_2023`. We apply `options(survey.lonely.psu = "adjust")` throughout.

**Generational Status.** Constructed from `BRNOUT` (respondent nativity). Parental
birthplace variables (`MOMFBRN`/`DADFBRN`) were not available in the 2022–2023
public-use file, so the variable is a two-category foreign-born/native-born indicator
rather than the three-category generational measure originally planned.

**Education Coding.** The 2022–2023 NSFG uses different `HIEDUC` codes than earlier
waves. Education is coded into five categories: Less than HS (codes 1–2), HS Diploma/GED
(3–4), Some College/Associate (5–7), Bachelor's Degree (8), and Graduate/Professional
(9–11). The graduate/professional category is separated from bachelor's because the
income premium and fertility patterns differ substantially between these groups —
women with master's, professional, and doctoral degrees have historically had the
highest rates of completed childlessness (Livingston, 2015). The `college` binary
indicator (HIEDUC >= 8) is retained for interaction models.

**Completed Childlessness.** `PARITY == 0` among women 40–49. `PARITY` counts live
births ever had — a biological fertility measure, not co-resident child status.

```{r session-info, results='asis'}
cat("\n\n**Session Information**\n\n")
cat(sprintf("R version: %s\n\n", R.version.string))
cat(sprintf("Platform: %s\n\n", R.version$platform))
cat(sprintf("Key packages: survey %s, haven %s, tidyverse %s\n",
  packageVersion("survey"), packageVersion("haven"), packageVersion("tidyverse")))
```
