---
title: "Segmented Assimilation in the American Financial System: Origin-Based Gradients in Immigrant Household Finances"
subtitle: "Evidence from the 2022 Survey of Consumer Finances"
author: "Delroy Nichols"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 5, dpi = 150
)
rm(list = ls())
gc()
set.seed(42)
```

# Origin-Based Gradients in Immigrant Household Finances

## Abstract

Two competing theoretical frameworks make different predictions about the financial
outcomes of foreign-born households in the United States. The human-capital convergence
model, developed in the labor-economics tradition by Borjas and others, predicts that
once individual-level characteristics — age, education, income, time since arrival — are
controlled for, origin region exerts no independent effect on financial outcomes. The
segmented-assimilation framework, developed by Portes, Rumbaut, and Zhou, predicts the
opposite: that origin region channels households into distinct incorporation pathways
(parallel integration, downward assimilation, selective assimilation) whose effects
persist net of human capital and duration of residence. Until 2022, this debate could
not be tested directly in U.S. wealth data because no nationally representative wealth
survey collected respondent national origin.

This paper uses newly added 2022 Survey of Consumer Finances variables on respondent
national origin (X6915) and years lived in the United States (X6906) to test these
competing predictions across seven financial outcomes covering all three theoretical
pathways. Logistic and linear regressions on a foreign-born analytic sample (n=1,103)
fail to reject the human-capital convergence null: origin-region coefficients are
small in magnitude, statistically indistinguishable from zero across all outcomes, and
unchanged when human-capital controls are added. A duration-stratified robustness check
splits the sample into recent arrivals (≤15 years in the U.S.) and long-tenured
immigrants (>15 years) to test whether segmentation effects exist in early-stage
incorporation but converge with time. The results inform discussion of when and where
segmented-assimilation theory is most likely to apply to U.S. wealth outcomes.

**Data Source:** 2022 Survey of Consumer Finances (SCF), Federal Reserve Board. All
estimates are survey-weighted; all models are pooled across the SCF's five multiply-
imputed datasets using Rubin's Rules.

**Note on Scope:** All findings are correlational and limited to the foreign-born
respondent sample. The SCF's cross-sectional design and absence of parental nativity
preclude causal claims and intergenerational analysis.

## The Central Question

Two theoretical frameworks dominate the empirical study of immigrant economic
incorporation in the United States, and they make sharply different predictions about
financial outcomes. The **human-capital convergence model**, rooted in the
labor-economics tradition associated with George Borjas and others, holds that immigrant
economic outcomes are explained by individual-level human capital — education, English
proficiency, age at arrival, time since arrival — and that origin region adds no
independent explanatory power once these characteristics are controlled. Under this
view, "Mexican origin" or "Asian origin" are aggregations of the human capital that
immigrants from those regions happen to bring; controlling for human capital eliminates
origin-region effects.

The **segmented-assimilation framework**, developed by Alejandro Portes, Rubén Rumbaut,
and Min Zhou, makes the opposite prediction. Three pathways are identified: *parallel
integration* into the white middle class through educational achievement, *downward
assimilation* into the racialized urban underclass through structural exclusion, and
*selective assimilation* through retention of ethnic-community ties and resources.
Origin region itself — net of human capital — channels households into these pathways,
because reception contexts (race, neighborhood, labor-market structure) differ
systematically across origin groups. A Mexican-origin household with a college degree
may face different mainstream-finance access than an Asian-origin household with
identical credentials.

Until 2022, this debate could not be tested directly in U.S. wealth data because no
nationally representative wealth survey collected respondent national origin. Earlier
SCF waves identified foreign-born respondents only through indirect indicators and
provided no origin-region detail. The 2022 SCF wave added respondent and spouse
national origin (X6915, X8055-X8057) and years lived in the United States (X6906,
X7796) — the first such measures in any nationally representative U.S. wealth survey.

This paper uses those new variables to test whether financial outcomes among
foreign-born households vary systematically by origin region net of duration and
human capital. The test is genuinely open: a finding of large, statistically robust
origin-region effects would support segmented assimilation; a finding of null or near-
null effects would support human-capital convergence. The two are competing models, not
complementary ones, and the SCF data should now distinguish them.

## Why This Matters: Beyond Race-Only Frameworks

Wealth and credit-access disparities have generally been studied through the lens of
race rather than origin. Studies of Black-white and Hispanic-white wealth gaps treat
the groups they compare as internally homogeneous — but Black households of Caribbean
origin, Mexican-origin Hispanics, and Cuban-origin Hispanics may have very different
financial profiles, shaped by distinct migration histories, reception contexts, and
community resources. Whether they actually do, in 2022 nationally representative data,
is the empirical question this paper addresses.

The 2022 SCF, by adding origin-region detail, lets the field test whether the
segmented-assimilation pathways — parallel integration, downward assimilation, selective
assimilation — show up in wealth and credit data with sufficient magnitude and
statistical precision to differentiate origin groups. If they do, interventions
designed for "Hispanic households" or "Asian households" would need to be redesigned
around within-racial-category origin heterogeneity. If they do not, the existing
race-based analytical frameworks — which are simpler, more reliably measured, and
already supported by a large literature — remain the appropriate analytical lens for
U.S. wealth disparities, even after accounting for the addition of origin-region detail.

The substantive contribution of this paper, then, is not to confirm a predetermined
theoretical position. It is to put two competing models head-to-head with the first
data capable of doing so, and report what the data show.

---

```{r load-packages}
required_packages <- c(
  "tidyverse", "haven", "survey", "labelled",
  "ggplot2", "scales", "scf",
  "broom", "knitr", "moments"
)

for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg, repos = "http://cran.us.r-project.org")
    library(pkg, character.only = TRUE)
  }
}
```

```{r download-data}
# Same download pattern as the credit-card-debt analysis (Paper 1).
# scf_download writes .rds files to the current working directory.

dir.create("scf_data", showWarnings = FALSE, recursive = TRUE)
original_wd <- getwd()

# Move into the data folder
setwd("scf_data")

library(haven)

# Read from your local absolute path
raw_macro_df <- read_dta("C:/Users/delro/OneDrive/Documents/scf2022s/p22i6.dta")

# Save directly to the current working directory (which is already scf_data)
saveRDS(raw_macro_df, "p22i6.rds") 

# Continue with the download check for the bulletin extract
if (!file.exists("scf2022.rds")) {
  tryCatch({
    scf_download(2022)
  }, error = function(e) {
    cat("Error downloading 2022 data:", conditionMessage(e), "\n")
  })
}

# Return to the root project directory
setwd(original_wd)
```

```{r load-data}
scf_2022 <- scf_load(2022, data_directory = "scf_data")
```

# Variable Availability and Diagnostic

Before constructing variables, we check which of the variables having direct exposure without segmented assimilation theory are
actually available in the loaded `scf` package object. The standard SCF Bulletin extract
that the `scf` package builds does not by default include the new national-origin or
years-in-U.S. items (`X6906`, `X7404`, `X6915`); these are raw codebook items rather than
constructed Bulletin variables. If they are missing, we read the raw 2022 SCF `.rds`
file directly and merge them in by household ID.

```{r diagnose-immigration-vars, results='asis'}
imp1 <- scf_2022$mi_design[[1]]$variables

# Variables we need
# X6915 = respondent national origin (the variable for origin region)
# X6906 = years respondent has lived in the U.S. (the YIU variable)
# X7796 = years spouse has lived in the U.S. (auxiliary)
# X7404 = spouse Hispanic/Latino self-id (auxiliary)
# X8049 = spouse race (used in swap correction)
# X8000 = sex/role swap indicator
#
# CORRECTED 2026-04-29: V5 incorrectly used X7437-X7443 as years-in-US.
# Per the official SCF 2022 changes file, X7437-X7443 are MORTGAGE FORBEARANCE
# variables, not immigration variables. The true years-in-US variable is X6906
# for the respondent (and X7796 for the spouse).
imm_vars <- c("X6915", "X6906", "X7796", "X7404", "X8000", "X8049")

present_in_extract <- imm_vars %in% names(imp1)

cat("**Immigration-variable availability in the scf package extract:**\n\n")
diag_df <- data.frame(
  Variable = imm_vars,
  `In extract` = ifelse(present_in_extract, "Yes", "No"),
  check.names = FALSE
)
print(kable(diag_df))
cat("\n\n")

need_raw_merge <- !all(present_in_extract)

if (need_raw_merge) {
  cat("Some required immigration variables are missing from the bulletin extract. ",
      "Attempting to merge them in from the raw 2022 SCF data file...\n\n", sep = "")
} else {
  cat("All required immigration variables are present in the extract. ",
      "No raw merge needed.\n\n", sep = "")
}
```

```{r merge-raw-immigration-vars}
# ============================================================================
# Merge raw immigration variables from the underlying SCF .rds file
# ============================================================================
# The scf package downloads the raw 2022 SCF data into scf_data/ as well as
# building the bulletin extract. The raw file contains all X-variables; we
# read it, select just the immigration-related vars + ID, and merge into
# each implicate's $variables data frame.
#
# Common file names produced by scf_download():
#   - p22i6.rds (raw public-use, 5 implicates stacked, with implicate id)
#   - scf2022.rds (the bulletin extract)
#
# We try a few candidate filenames defensively.

if (need_raw_merge) {

  raw_candidates <- c(
    file.path("scf_data", "p22i6.rds"),
    file.path("scf_data", "scfp2022.rds"),
    file.path("scf_data", "p22.rds"),
    "p22i6.rds",
    "scfp2022.rds"
  )

  raw_path <- raw_candidates[file.exists(raw_candidates)][1]

  if (!is.na(raw_path)) {
    raw_data <- readRDS(raw_path)

    # Standardize variable name casing — the raw file may have either
    # uppercase or lowercase. We try uppercase first since X-variables
    # are conventionally uppercase.
    raw_path <- raw_candidates[file.exists(raw_candidates)][1]

  if (!is.na(raw_path)) {
    raw_data <- readRDS(raw_path)

    # FIX: Force all raw data columns to uppercase immediately.
    # This aligns the Stata output with the canonical SCF uppercase format.
    names(raw_data) <- toupper(names(raw_data))

    raw_names <- names(raw_data)
    
    # The name_map is now technically redundant, but safe to keep
    name_map <- setNames(raw_names, toupper(raw_names))

    available_imm <- imm_vars[toupper(imm_vars) %in% toupper(raw_names)]

    # The raw data has 5 implicates stacked. We need an ID that matches
    # the bulletin extract's row ordering. Common candidates: y1, yy1, x1.
    id_candidates <- c("Y1", "YY1", "X1") # Updated to only look for uppercase
    id_var <- id_candidates[id_candidates %in% raw_names][1]

    if (!is.na(id_var) && length(available_imm) > 0) {

      # Pull the matching columns
      raw_imm <- raw_data[, c(id_var, available_imm), drop = FALSE]

      # Determine implicate count: raw file has each household 5 times (one
      # per implicate) — split into 5 chunks of equal size in row order.
      n_total <- nrow(raw_imm)
      n_per_imp <- nrow(scf_2022$mi_design[[1]]$variables)

      if (n_total == 5 * n_per_imp) {
        # Standard layout: implicates stacked
        for (i in 1:5) {
          start_row <- (i - 1) * n_per_imp + 1
          end_row   <- i * n_per_imp
          imm_chunk <- raw_imm[start_row:end_row, available_imm, drop = FALSE]

          # Add to the implicate's $variables
          for (v in available_imm) {
            scf_2022$mi_design[[i]]$variables[[v]] <- imm_chunk[[v]]
          }
        }
        cat("Successfully merged ", length(available_imm), " immigration variables ",
            "from ", raw_path, " into all 5 implicates.\n", sep = "")
      } else {
        cat("Row-count mismatch: raw file has ", n_total,
            " rows, bulletin extract has ", n_per_imp,
            " rows per implicate. Merge aborted; using available variables only.\n",
            sep = "")
      }
    } else {
      cat("Could not locate raw SCF file or matching ID variable. ",
          "Immigration variables will be unavailable.\n", sep = "")
    }
  } else {
    cat("Raw 2022 SCF .rds file not found in scf_data/. ",
        "Available files:\n", sep = "")
    cat(paste(list.files("scf_data"), collapse = "\n"), "\n\n")
  }
}

# Re-check what's now available
imp1 <- scf_2022$mi_design[[1]]$variables
have_X7404 <- "X7404" %in% names(imp1)
have_yiu   <- "X6906" %in% names(imp1)
have_X8000 <- "X8000" %in% names(imp1)
have_X8049 <- "X8049" %in% names(imp1)
```

```{r raw-x7404-frequency, results='asis'}
# ============================================================================
# Print raw X7404 value frequencies — needed to calibrate the recode
# ============================================================================
# Until we see what values X7404 actually takes in this wave, we cannot
# write the correct recoding rule. This chunk dumps the unweighted frequency
# table so we can adjust the recode.

if (have_X7404) {
  # 1. Extract the column and force it to a standard numeric vector
  # This strips out any problematic 'haven_labelled' metadata from Stata
  x7404_vec <- as.numeric(scf_2022$mi_design[[1]]$variables$X7404)
  
  # 2. Use Base R's table() to get the counts, which avoids package conflicts
  freq_table <- table(x7404_vec, useNA = "ifany")
  
  # 3. Reconstruct it into a clean data frame for kable
  x7404_freq <- data.frame(
    X7404 = names(freq_table),
    n = as.integer(freq_table)
  )
  
  # 4. Sort descending
  x7404_freq <- x7404_freq[order(-x7404_freq$n), ]

  cat("**Raw X7404 frequency (Implicate 1, unweighted):**\n\n")
  # row.names = FALSE keeps the markdown table perfectly clean
  print(kable(x7404_freq, caption = "X7404 = respondent national origin", row.names = FALSE))
  cat("\n\nThis table guides the origin-region recode in the next chunk. ",
      "Code 0 typically indicates 'inapplicable' (i.e., U.S.-born or non-respondent).\n\n",
      sep = "")
}
```

```{r visuals}

library(ggplot2)
library(dplyr)

# ============================================================================
# Visualize X7404 Frequencies: Spouse Hispanic/Latino Culture/Origin
# ============================================================================

# 1. SCF mapping for X7404
spouse_hisp_labels <- c(
  "0" = "Inapplicable (No Spouse/Partner)",
  "1" = "Yes (Spouse is Hispanic/Latino)",
  "5" = "No (Spouse is not Hispanic/Latino)"
)

# 2. Apply labels and generate plot
x7404_freq %>%
  mutate(
    # Map the numeric codes to our accurate text labels
    Spouse_Label = spouse_hisp_labels[as.character(X7404)],
    # Catch any unexpected codes just in case
    Spouse_Label = ifelse(is.na(Spouse_Label), paste("Code", X7404), Spouse_Label)
  ) %>%
  # Order the bars by frequency 
  ggplot(aes(x = reorder(Spouse_Label, n), y = n)) +
  geom_col(fill = "steelblue", color = "black", alpha = 0.8) +
  coord_flip() + # Horizontal layout for readability
  theme_minimal() +
  labs(
    title = "Spouse/Partner Hispanic or Latino Origin",
    subtitle = "Raw frequencies from Implicate 1 (Variable X7404)",
    x = NULL, 
    y = "Number of Respondents"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.y = element_text(size = 11)
  )

```



```{r plot-x6915, echo=FALSE, warning=FALSE, message=FALSE}
# ============================================================================
# Visualize X6915 Frequencies: Reference Person National Origin
# ============================================================================

library(ggplot2)
library(dplyr)

# FIX: Explicitly define have_X6915 so the Knit session knows what it is!
have_X6915 <- "X6915" %in% names(scf_2022$mi_design[[1]]$variables)

if (have_X6915) {
  
  x6915_vec <- as.numeric(scf_2022$mi_design[[1]]$variables$X6915)
  freq_table_6915 <- table(x6915_vec, useNA = "ifany")
  
  x6915_freq <- data.frame(
    X6915 = names(freq_table_6915),
    n = as.integer(freq_table_6915)
  )
  
  origin_labels <- c(
    "0" = "U.S. Born / Inapplicable",
    "100" = "Western European", 
    "200" = "Eastern European",
    "300" = "European",
    "400" = "West Indies",
    "500" = "Central and South American", 
    "600" = "Middle Easterner",
    "700" = "Sub-Saharan African",
    "800" = "South Asian",
    "900" = "East Asian",
    "1000" = "All Other Asians",
    "1100" = "Pacific Islander",
    "1200" = "American"
  )
  
  p_x6915 <- x6915_freq %>%
    mutate(
      Origin_Label = origin_labels[as.character(X6915)],
      Origin_Label = ifelse(is.na(Origin_Label), paste("Code", X6915), Origin_Label)
    ) %>%
    filter(X6915 != "0") %>% 
    ggplot(aes(x = reorder(Origin_Label, n), y = n)) +
    geom_col(fill = "seagreen", color = "black", alpha = 0.8) +
    coord_flip() + 
    theme_minimal() +
    labs(
      title = "Reference Person National Origin (Immigrant Sub-sample)",
      subtitle = "Unweighted frequencies from Implicate 1 (Variable X6915, excluding U.S. Born)",
      x = NULL, 
      y = "Number of Respondents"
    ) +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      axis.text.y = element_text(size = 11)
    )
  
  # Properly print the plot
  print(p_x6915)
  
} else {
  cat("Variable X6915 not found. Please ensure the merge chunk was re-run.")
}
```

```{r table-x6915, echo=FALSE, results='asis'}
# Print the raw counts table
if (have_X6915) {
  cat("\n\n**Raw X6915 frequency (Implicate 1, unweighted):**\n\n")
  tab <- knitr::kable(x6915_freq %>% arrange(desc(n)), row.names = FALSE)
  cat(paste(tab, collapse = "\n"), "\n\n")
}
```


# Variable Engineering

```{r data-engineering-and-subsetting, warning=FALSE, message=FALSE}
# ============================================================================
# Build all analysis variables and create the analytic subset
# ============================================================================
# CRITICAL DESIGN NOTE on foreign-born identification:
#
# X6915 is an ANCESTRY/HERITAGE variable, asked of all respondents — both
# foreign-born and U.S.-born. A respondent reporting "Western European"
# ancestry (code 100) is overwhelmingly U.S.-born (white American with
# European heritage), not an immigrant from Europe. Likewise, "American"
# ancestry (1200) is U.S.-born by definition, and "Sub-Saharan African" (700)
# captures Black Americans reporting African ancestry alongside actual
# African immigrants.
#
# The actual foreign-born signal is in X6906 (years the reference person
# has lived in the United States), which the SCF only asks of foreign-born
# respondents. A non-zero, non-missing years-in-US value confirms foreign-
# born status. (The companion variable X7796 is years-in-US for the spouse
# but is not used in main models, since the unit of analysis is the
# respondent.)
#
# IMPORTANT NOTE: V5 erroneously used X7437-X7443 as the years-in-US
# variables. Per the official SCF 2022 changes file, those are MORTGAGE
# FORBEARANCE variables (which mortgage terms were adjusted after exiting
# forbearance). The correct YIU variable is X6906. This V6 build uses
# X6906 throughout.
#
# Therefore, foreign_born requires BOTH:
#   (a) X6915 indicates a non-American ancestry (i.e., X6915 != 0 AND
#       X6915 != 1200), AND
#   (b) X6906 > 0 (confirming actual years lived in the U.S. as immigrant)
# ============================================================================

# ---- Step 1: Years in U.S. (X6906 = respondent years in US) ----
# X6906 = 0 means inapplicable per SCF data convention (i.e., U.S.-born).
# X6906 > 0 means the respondent reported N years lived in the U.S.,
# which is asked only of foreign-born respondents.
if (have_yiu) {
  scf_2022 <- scf_update(scf_2022,
    years_in_us  = ifelse(as.numeric(X6906) <= 0, NA_real_, as.numeric(X6906)),
    yiu_missing  = as.integer(is.na(years_in_us)),
    yiu_clean    = ifelse(is.na(years_in_us), 0, years_in_us)
  )
} else {
  scf_2022 <- scf_update(scf_2022,
    years_in_us = NA_real_, yiu_missing = 1L, yiu_clean = 0
  )
}

# ---- Step 2: CORRECTED foreign-born status ----
# Requires both ancestry indicating non-US origin AND positive years in US.
# Fall back gracefully if X6915 is unavailable.
if (have_X6915 && have_yiu) {
  scf_2022 <- scf_update(scf_2022,
    foreign_born = as.integer(
      !is.na(years_in_us) &
      years_in_us > 0 &
      as.numeric(X6915) != 0 &
      as.numeric(X6915) != 1200
    )
  )
} else if (have_yiu) {
  # YIU-only fallback if X6915 is unavailable
  scf_2022 <- scf_update(scf_2022,
    foreign_born = as.integer(!is.na(years_in_us) & years_in_us > 0)
  )
} else {
  scf_2022 <- scf_update(scf_2022, foreign_born = 0L)
}

# ---- Step 3: Origin-region recode from X6915 ----
# Used only within the foreign-born sample (filtered by foreign_born flag).
# Reference category: Latin American (the segmented-assimilation canonical
# case in the literature; X6915 == 500 = Central/South American).
if (have_X6915) {
  scf_2022 <- scf_update(scf_2022,
    origin_region = factor(case_when(
      as.numeric(X6915) == 500 ~ "Latin American",
      as.numeric(X6915) == 400 ~ "Caribbean/West Indies",
      as.numeric(X6915) %in% c(800, 900, 1000) ~ "Asian",
      as.numeric(X6915) %in% c(100, 200, 300) ~ "European/Other developed",
      as.numeric(X6915) %in% c(600, 700, 1100) ~ "Other",
      TRUE ~ NA_character_
    ), levels = c("Latin American",
                  "Caribbean/West Indies",
                  "Asian",
                  "European/Other developed",
                  "Other"))
  )
} else {
  scf_2022 <- scf_update(scf_2022, origin_region = factor(NA_character_))
}

# ---- Step 3b: Spouse/Partner Hispanic Culture/Origin from X7404 ----
if (have_X7404) {
  scf_2022 <- scf_update(scf_2022,
    spouse_hisp_origin = factor(case_when(
      X7404 == 1 ~ "Yes (Spouse Hispanic/Latino)",
      X7404 == 5 ~ "No (Spouse not Hispanic/Latino)",
      X7404 == 0 ~ "Unmarried/Inapplicable",
      TRUE ~ NA_character_
    ), levels = c("Yes (Spouse Hispanic/Latino)",
                  "No (Spouse not Hispanic/Latino)",
                  "Unmarried/Inapplicable"))
  )
} else {
  scf_2022 <- scf_update(scf_2022, spouse_hisp_origin = factor(NA_character_))
}

# ---- Step 4: Race/ethnicity ----
scf_2022 <- scf_update(scf_2022,
  race_cat = factor(racecl4, levels = 1:4,
                    labels = c("White", "Black", "Hispanic", "Other"))
)

# ---- Step 5: Standard demographic controls ----
scf_2022 <- scf_update(scf_2022,
  age_centered = age - 50,
  age_sq       = age^2,
  in_age_range = as.integer(age >= 25 & age <= 75),
  educ_cat = factor(case_when(
    educ < 8 ~ "Less than HS",
    educ == 8 ~ "High School",
    educ %in% 9:11 ~ "Some College",
    educ == 12 ~ "College Degree",
    educ %in% 13:14 ~ "Graduate School/PhD",
    TRUE ~ NA_character_
  ), levels = c("Less than HS", "High School", "Some College",
                "College Degree", "Graduate School/PhD")),
  female            = as.integer(hhsex == 2),
  currently_married = as.integer(married == 1),
  has_children      = as.integer(kids > 0),
  is_homeowner      = as.integer(housecl == 1),
  has_pension       = as.integer(anypen == 1),
  log_income        = log(pmax(1, income)),
  asinh_networth    = asinh(networth),
  log_wageinc       = log(pmax(1, wageinc))
)

# ---- Step 6: Outcomes ----
# CRITICAL: nw_pctl is computed on the FULL SCF distribution (before
# subsetting to foreign-born) so it reflects position in the broader
# U.S. wealth distribution, not relative position within the foreign-born
# subsample. This is computed via scf_update on scf_2022 (the full design)
# so all 5 implicates get a consistent percentile based on the full sample.
scf_2022 <- scf_update(scf_2022,
  own_retire    = as.integer(retqliq > 0),
  own_stocks    = as.integer(stocks > 0),
  credit_denied = as.integer(turndown == 1),
  no_checking   = as.integer(checking <= 0),
  own_home      = as.integer(housecl == 1),
  nw_pctl       = 100 * percent_rank(networth)   # full-sample percentile
)

# ---- Step 7: Analytic sample flag ----
scf_2022 <- scf_update(scf_2022,
  in_analytic_sample = as.integer(
    foreign_born == 1 &
    !is.na(origin_region) &
    in_age_range == 1
  )
)

# ---- Step 8: Create the Analytic Sample Subset ----
# Subset each implicate's survey design to foreign-born respondents.
# nw_pctl values are preserved — they were computed on the full distribution
# in Step 6 above, so subsetting just keeps the foreign-born rows with
# their original full-distribution percentile values intact.
scf_fb <- scf_2022
scf_fb$mi_design <- lapply(scf_2022$mi_design, function(des) {
  subset(des, in_analytic_sample == 1)
})

# Diagnostic: report sample size
n_fb <- nrow(scf_fb$mi_design[[1]]$variables)

# Diagnostic: cross-tab of X6915 ancestry × foreign_born flag — useful for
# verifying that the corrected foreign-born identification is sensible.
fb_diagnostic <- scf_2022$mi_design[[1]]$variables %>%
  filter(!is.na(X6915)) %>%
  group_by(X6915) %>%
  summarise(
    n_total       = n(),
    n_foreign     = sum(foreign_born == 1, na.rm = TRUE),
    pct_foreign   = round(100 * mean(foreign_born == 1, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  arrange(desc(n_total))
```

```{r build-foreign-born-design}
# ============================================================================
# Build the foreign-born subset design — same pattern as scf_debtors in Paper 1
# ============================================================================
# NOTE: foreign_born and years_in_us are already defined in the
# `data-engineering-and-subsetting` chunk above, using the X6915-based
# definition (X6915 != 0 = foreign-born). The X6915 definition is preferred
# over X6906 because X6906 has known small missingness (some foreign-born
# missingness, while X6915 is universally coded.
#
# This chunk just rebuilds the scf_fb subset object using the
# in_analytic_sample flag from the prior chunk.

scf_fb <- scf_2022

scf_fb$mi_design <- lapply(scf_2022$mi_design, function(des) {
  subset(des, in_analytic_sample == 1)
})

# Diagnostic: report sample size
n_fb <- nrow(scf_fb$mi_design[[1]]$variables)
```

# Descriptive Findings

## Foreign-Born Identification Diagnostic

The corrected `foreign_born` definition requires both ancestry indicating non-U.S.
origin (X6915 ≠ 0 and X6915 ≠ 1200) AND positive years in the U.S. (X6906).
The table below shows, for each X6915 ancestry code, what percentage of respondents
end up classified as foreign-born. Codes like 100 (Western European), 200 (Eastern
European), and 1200 (American) should be near 0%; codes like 500 (Latin American),
800/900 (Asian), and 600 (Middle Eastern) should be near 100%.

```{r fb-id-diagnostic, results='asis'}
if (exists("fb_diagnostic") && nrow(fb_diagnostic) > 0) {

  origin_label_map <- c(
    "0"    = "U.S. Born / Inapplicable",
    "100"  = "Western European",
    "200"  = "Eastern European",
    "300"  = "European (Other)",
    "400"  = "West Indies / Caribbean",
    "500"  = "Central / South American",
    "600"  = "Middle Eastern",
    "700"  = "Sub-Saharan African",
    "800"  = "South Asian",
    "900"  = "East Asian",
    "1000" = "All Other Asian",
    "1100" = "Pacific Islander",
    "1200" = "American"
  )

  fb_diagnostic_display <- fb_diagnostic %>%
    mutate(
      Ancestry = origin_label_map[as.character(X6915)],
      Ancestry = ifelse(is.na(Ancestry), paste("Code", X6915), Ancestry)
    ) %>%
    select(X6915, Ancestry, n_total, n_foreign, pct_foreign)

  cat("**Table 0. Foreign-born identification by X6915 ancestry code**\n\n")
  print(kable(fb_diagnostic_display,
              col.names = c("X6915", "Ancestry label", "n total",
                            "n foreign-born", "% foreign-born"),
              caption = "Foreign-born flag = X6915 ∈ origin codes AND years_in_us > 0."))
  cat("\n\n")
}
```

## Foreign-Born Sample Composition

```{r desc-sample-comp, results='asis'}
# Sample composition statistics — the origin_region recode was already done
# in the data-engineering chunk above; we don't need to redo it here.

if (n_fb > 0) {
  fb_imp1 <- scf_fb$mi_design[[1]]$variables

  mean_age   <- weighted.mean(fb_imp1$age, fb_imp1$wgt, na.rm = TRUE)
  mean_yiu   <- weighted.mean(fb_imp1$years_in_us, fb_imp1$wgt, na.rm = TRUE)
  mean_inc   <- weighted.mean(fb_imp1$income, fb_imp1$wgt, na.rm = TRUE)
  median_nw  <- median(fb_imp1$networth, na.rm = TRUE)
  pct_yiu_missing <- 100 * mean(is.na(fb_imp1$years_in_us))

  cat(sprintf("The analytic sample comprises %s foreign-born respondents in the 2022 SCF aged 25-75. ",
              formatC(n_fb, big.mark = ",")))
  cat(sprintf("Using survey weights, the mean age in this population is %.1f years, with a mean of %.1f years lived in the U.S. ",
              mean_age, mean_yiu))
  cat(sprintf("Mean household income is $%s and median net worth is $%s. ",
              formatC(round(mean_inc), big.mark = ",", format = "d"),
              formatC(round(median_nw), big.mark = ",", format = "d")))
  cat(sprintf("Note: %.1f%% of foreign-born respondents have missing years-in-U.S. data; the missing-indicator method is used in regressions to retain these observations.\n\n",
              pct_yiu_missing))
} else {
  cat("**Analytic sample is empty.** Check the X6915 origin-region recode in the variable-engineering chunk.\n\n")
}
```

## Origin-Region Distribution

```{r desc-origin-table, results='asis'}
if (n_fb > 0) {
  fb_imp1 <- scf_fb$mi_design[[1]]$variables

  origin_summary <- fb_imp1 %>%
    filter(!is.na(origin_region)) %>%
    group_by(origin_region) %>%
    summarise(
      n_unweighted   = n(),
      weighted_count = sum(wgt, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      pct_weighted = round(weighted_count / sum(weighted_count) * 100, 1)
    ) %>%
    select(origin_region, n_unweighted, pct_weighted)

  cat("**Table 1. Foreign-born sample by origin region (2022 SCF)**\n\n")
  print(kable(origin_summary,
              col.names = c("Origin Region", "n (Unweighted)", "% (Weighted)"),
              caption = "Note: percentages are SCF-weighted; n is unweighted (Implicate 1)."))
  cat("\n\n")

  small_cells <- origin_summary %>% filter(n_unweighted < 30)
  if (nrow(small_cells) > 0) {
    cat(sprintf("**Warning:** Small unweighted cell sizes (< 30) in %d origin region(s): %s. Model estimates for these groups may be unstable.\n\n",
                nrow(small_cells),
                paste(as.character(small_cells$origin_region), collapse = ", ")))
  }
}
```

## Outcome Prevalence by Origin Region

```{r desc-outcomes-by-origin, results='asis'}
if (n_fb > 0) {

  # Use scf_mean for each outcome (matches Paper 1 pattern)
  outcomes <- list(
    own_retire    = "Retirement (%)",
    own_stocks    = "Stocks (%)",
    credit_denied = "Credit denied (%)",
    no_checking   = "No checking (%)",
    own_home      = "Homeowner (%)",
    nw_pctl       = "NW pctile (mean)"
  )

  # Run each prevalence estimate using scf_mean with by = ~origin_region
  prevalence_list <- list()
  for (out_var in names(outcomes)) {
    f <- as.formula(paste0("~", out_var))
    res <- tryCatch({
      scf_mean(scf_fb, f, by = ~origin_region)$results
    }, error = function(e) NULL)
    prevalence_list[[out_var]] <- res
  }

  # Build display table — one row per origin region
  origin_levels <- levels(scf_fb$mi_design[[1]]$variables$origin_region)
  display_df <- data.frame(`Origin Region` = origin_levels,
                           check.names = FALSE,
                           stringsAsFactors = FALSE)

  for (out_var in names(outcomes)) {
    res <- prevalence_list[[out_var]]
    col_label <- outcomes[[out_var]]

    if (is.null(res) || nrow(res) == 0) {
      display_df[[col_label]] <- "--"
      next
    }

    # The scf_mean result has columns: by-variable level, estimate, std.error
    # Match by origin_region level
    estimates <- sapply(origin_levels, function(lvl) {
      idx <- which(as.character(res[[1]]) == lvl)
      if (length(idx) == 0) return(NA_real_)
      # The estimate column is typically the second column or named "estimate"
      if ("estimate" %in% names(res)) res$estimate[idx] else res[[2]][idx]
    })

    if (out_var == "nw_pctl") {
      display_df[[col_label]] <- ifelse(is.na(estimates), "--",
                                         sprintf("%.1f", estimates))
    } else {
      display_df[[col_label]] <- ifelse(is.na(estimates), "--",
                                         sprintf("%.1f", estimates * 100))
    }
  }

  cat("**Table 2. Weighted prevalence of financial outcomes by origin region**\n\n")
  print(kable(display_df,
              caption = "All estimates are SCF-weighted and pooled across implicates."))
  cat("\n\n")
}
```

## Years-in-U.S. Distribution by Origin Region

```{r desc-yiu-plot, fig.cap="Distribution of years lived in the U.S. by origin region (foreign-born SCF sample, 2022)"}
if (n_fb > 0) {
  plot_data <- scf_fb$mi_design[[1]]$variables %>%
    filter(!is.na(years_in_us), !is.na(origin_region))

  if (nrow(plot_data) > 0) {
    ggplot(plot_data,
           aes(x = origin_region, y = years_in_us, fill = origin_region)) +
      geom_boxplot(alpha = 0.7, outlier.shape = NA) +
      geom_jitter(width = 0.15, alpha = 0.2, size = 0.6) +
      labs(x = NULL,
           y = "Years lived in the U.S.",
           subtitle = "Unweighted distributional shape (Implicate 1)") +
      theme_minimal(base_size = 11) +
      theme(legend.position = "none",
            axis.text.x = element_text(angle = 30, hjust = 1)) +
      scale_fill_brewer(palette = "Set2")
  }
}
```

# Modeling Results

## H1: Origin-Region Main Effects

We test whether financial outcomes among foreign-born respondents differ by origin
region net of years-in-U.S., age, education, marital status, and income. Logistic
regression for binary outcomes (`scf_glm` with `binomial()`) and OLS for the continuous
net-worth-percentile outcome (`scf_ols`). All models pool across the 5 implicates.

Mexican origin is the reference category — that is, the X6915=500 ("Central and South
American") group, which the segmented-assimilation literature treats as the canonical
Latin American case. Coefficients indicate how each origin region compares to Latin
American foreign-born respondents.

**Reading the results.** Each outcome has both a Baseline specification (origin region
+ years-in-U.S. + age) and a Full specification (Baseline + education + marital status +
income). The segmented-assimilation prediction is that origin-region coefficients should
remain large and statistically significant in the Full specification, indicating that
origin matters net of human capital. The human-capital convergence prediction is that
origin-region coefficients should be small and statistically indistinguishable from zero
once human-capital controls are added. Confidence intervals that include 1.0 (for odds
ratios) or 0.0 (for net-worth-percentile coefficients) indicate failure to reject the
null hypothesis of no origin-region effect.

```{r h1-model-helper}
# ============================================================================
# Helper: fit and tidy an scf_glm or scf_ols model (NESTED SPECS)
# ============================================================================
library(dplyr)

fit_h1_model <- function(outcome_var, scf_obj, family_type = "binomial") {

  # 1. Define the nested specifications
  # V8 fix: replaced yiu_clean + yiu_missing with raw years_in_us. The
  # foreign-born filter requires valid years_in_us, so yiu_missing is
  # constant 0 across the analytic sample, which makes it a singular
  # predictor and crashes the GLM solver.
  baseline_rhs <- "origin_region + years_in_us + age_centered"
  full_rhs     <- "origin_region + years_in_us + age_centered + educ_cat + currently_married + log_income"

  baseline_formula <- as.formula(paste(outcome_var, "~", baseline_rhs))
  full_formula     <- as.formula(paste(outcome_var, "~", full_rhs))

  # CRITICAL FIX (V7): Set the formula's environment to the global environment.
  # When as.formula() is called inside a function, the formula's environment
  # becomes that function's local scope by default. The scf package's
  # internal svyglm calls then can't find the variables (origin_region,
  # educ_cat, etc.) because they live in the design object's $variables data
  # frame, not in the function's local scope. Setting the formula environment
  # to globalenv() lets svyglm fall back to looking up variables in the
  # design's data, which is what Paper 1's inline-formula calls do implicitly.
  environment(baseline_formula) <- globalenv()
  environment(full_formula)     <- globalenv()

  # 2. Internal function to fit, tidy, and label a single specification.
  # Returns a list with $results (success) or $error + $message (failure).
  fit_and_format <- function(form, label) {
    fit <- tryCatch({
      if (family_type == "binomial") {
        scf_glm(scf_obj, form, family = binomial())
      } else {
        scf_ols(scf_obj, form)
      }
    }, error = function(e) list(error = TRUE, message = conditionMessage(e)))

    # Propagate error info up so the cache loop can surface it
    if (is.list(fit) && isTRUE(fit$error)) {
      return(list(error = TRUE,
                  message = sprintf("[%s] %s", label, fit$message)))
    }

    res <- fit$results
    res$Model <- factor(label, levels = c("Baseline (Demographics Only)",
                                          "Full (With Human Capital)"))

    if (family_type == "binomial" && "estimate" %in% names(res)) {
      res$odds_ratio <- exp(res$estimate)
      res$conf_low   <- exp(res$estimate - 1.96 * res$std.error)
      res$conf_high  <- exp(res$estimate + 1.96 * res$std.error)
    } else if ("estimate" %in% names(res)) {
      res$odds_ratio <- NA
      res$conf_low   <- res$estimate - 1.96 * res$std.error
      res$conf_high  <- res$estimate + 1.96 * res$std.error
    }
    return(res)
  }

  res_base <- fit_and_format(baseline_formula, "Baseline (Demographics Only)")
  res_full <- fit_and_format(full_formula, "Full (With Human Capital)")

  base_failed <- is.list(res_base) && isTRUE(res_base$error)
  full_failed <- is.list(res_full) && isTRUE(res_full$error)

  # If both failed, surface the actual error messages
  if (base_failed && full_failed) {
    return(list(error = TRUE,
                message = paste(res_base$message, "||", res_full$message)))
  }

  # If only one failed, drop it but keep the other
  parts <- list()
  if (!base_failed) parts$base <- res_base
  if (!full_failed) parts$full <- res_full

  bind_rows(parts)
}
```

```{r h1-diagnostic-test-fit, results='asis'}
# ============================================================================
# DIAGNOSTIC: Try a simple inline-formula scf_glm call before the main cache
# loop. If this works but the cache loop fails, the issue is formula-
# environment scoping. If this also fails, the issue is something deeper
# (data, design, package version) and the message will tell us what.
# ============================================================================

cat("**Diagnostic: simple test fit using inline formula syntax (mimics Paper 1's pattern)**\n\n")

if (n_fb >= 100) {

  test_fit <- tryCatch({
    scf_glm(scf_fb, own_home ~ origin_region, family = binomial())
  }, error = function(e) list(error = TRUE, message = conditionMessage(e)))

  if (is.list(test_fit) && isTRUE(test_fit$error)) {
    cat(sprintf("- Inline test fit FAILED: `%s`\n\n", test_fit$message))
    cat("  This means scf_glm itself is unable to fit on the foreign-born subset, ",
        "regardless of formula-environment issues. Likely causes: a needed ",
        "variable is missing in the design, the design object is malformed, ",
        "or the scf package version differs from Paper 1's.\n\n", sep = "")
  } else {
    cat("- Inline test fit SUCCEEDED. scf_glm works on this design with origin_region as predictor.\n\n")
    cat("  Coefficients returned: ",
        paste(test_fit$results$term, collapse = ", "), "\n\n", sep = "")
  }

  # Also test scf_ols
  test_ols <- tryCatch({
    scf_ols(scf_fb, nw_pctl ~ origin_region)
  }, error = function(e) list(error = TRUE, message = conditionMessage(e)))

  if (is.list(test_ols) && isTRUE(test_ols$error)) {
    cat(sprintf("- Inline scf_ols test FAILED: `%s`\n\n", test_ols$message))
  } else {
    cat("- Inline scf_ols test SUCCEEDED.\n\n")
  }
}
```

```{r h1-fit-all-models-once, cache=TRUE}
# ============================================================================
# PERFORMANCE: Fit each H1 model EXACTLY ONCE here and cache the results.
# All downstream chunks (forest plots, H1 tables, H2 pattern matrix) pull from
# this cache instead of re-fitting. cache=TRUE on this chunk means knitr won't
# even re-run this if nothing upstream has changed.
#
# IMPORTANT: If you've previously knit V6 (or any earlier version) with
# cache=TRUE and the models failed, knitr may have cached the failed results.
# Before knitting V7, delete the cache folder created by knitr:
#   - file.exists("scf_segmented_assimilation_v7_cache")  # or v6/v5 cache
# OR delete the entire `*_cache/` directory next to the .Rmd file.
# Otherwise V7 will return the failed V6 results from cache instead of
# re-running the corrected fit logic.
# ============================================================================

model_cache <- list()

if (n_fb >= 100) {

  outcomes_to_fit <- list(
    list(var = "own_retire",    family = "binomial"),
    list(var = "own_stocks",    family = "binomial"),
    list(var = "credit_denied", family = "binomial"),
    list(var = "no_checking",   family = "binomial"),
    list(var = "own_home",      family = "binomial"),
    list(var = "nw_pctl",       family = "ols")
  )

  # Pre-check: identify outcomes with quasi-complete separation in the
  # foreign-born × origin × outcome table. If any origin region has zero
  # variance on the outcome, the GLM will produce gigantic standard errors
  # (effective infinity). Skip those outcomes rather than reporting noise.
  fb_imp1_for_check <- scf_fb$mi_design[[1]]$variables

  for (info in outcomes_to_fit) {
    if (info$family == "binomial") {
      cell_var <- fb_imp1_for_check %>%
        filter(!is.na(origin_region)) %>%
        group_by(origin_region) %>%
        summarise(v = var(.data[[info$var]], na.rm = TRUE), .groups = "drop")

      if (any(cell_var$v == 0 | is.na(cell_var$v))) {
        bad_groups <- cell_var %>% filter(v == 0 | is.na(v)) %>% pull(origin_region)
        message(sprintf("Skipping %s: zero within-cell variance in origin region(s): %s",
                        info$var, paste(as.character(bad_groups), collapse = ", ")))
        model_cache[[info$var]] <- list(
          error = TRUE,
          message = sprintf("Quasi-complete separation: origin region(s) %s have zero variance on this outcome.",
                            paste(as.character(bad_groups), collapse = ", "))
        )
        next
      }
    }

    message("Fitting H1 model for: ", info$var)
    fit_result <- fit_h1_model(info$var, scf_fb, info$family)
    model_cache[[info$var]] <- fit_result

    # Surface failures to the knit log so the user can see what's wrong
    if (is.list(fit_result) && isTRUE(fit_result$error)) {
      message(sprintf("  -> FAILED: %s", fit_result$message))
    } else {
      n_terms <- nrow(fit_result)
      message(sprintf("  -> SUCCESS: %d coefficient rows", n_terms))
    }
  }

  message("All H1 models fit and cached.")
}

# Helper to retrieve from cache (used by all downstream chunks)
get_h1_results <- function(outcome_var) {
  if (!outcome_var %in% names(model_cache)) {
    return(list(error = TRUE,
                message = sprintf("No cached results for %s.", outcome_var)))
  }
  model_cache[[outcome_var]]
}

# Helper: format H1 model results as a compact, readable table.
# Pivots wider so Baseline and Full appear side-by-side rather than stacked,
# strips the "origin_region" prefix from term labels, and drops row names.
format_h1_display <- function(res, family_type = "binomial") {

  origin_rows <- res %>%
    filter(grepl("^origin_region", term)) %>%
    mutate(Origin = gsub("^origin_region", "", term))

  if (nrow(origin_rows) == 0) return(NULL)

  if (family_type == "binomial") {
    # Wide format: one row per origin, columns for Baseline OR / CI / p
    # and Full OR / CI / p
    wide <- origin_rows %>%
      mutate(
        OR_str  = sprintf("%.2f", odds_ratio),
        CI_str  = sprintf("[%.2f, %.2f]", conf_low, conf_high),
        p_str   = formatC(p.value, format = "g", digits = 2)
      ) %>%
      select(Origin, Model, OR_str, CI_str, p_str) %>%
      tidyr::pivot_wider(
        names_from  = Model,
        values_from = c(OR_str, CI_str, p_str),
        names_glue  = "{Model}__{.value}"
      )

    # Build clean column ordering and labels
    base_label <- "Baseline (Demographics Only)"
    full_label <- "Full (With Human Capital)"

    cols_in_order <- c(
      "Origin",
      paste0(base_label, "__OR_str"),
      paste0(base_label, "__CI_str"),
      paste0(full_label, "__OR_str"),
      paste0(full_label, "__CI_str"),
      paste0(full_label, "__p_str")
    )
    cols_in_order <- intersect(cols_in_order, names(wide))
    wide <- wide[, cols_in_order, drop = FALSE]

    col_labels <- c("Origin",
                    "OR (Base)", "95% CI (Base)",
                    "OR (Full)", "95% CI (Full)",
                    "p (Full)")
    col_labels <- col_labels[seq_len(ncol(wide))]

    return(list(table = wide, col_labels = col_labels))

  } else {
    # OLS: report estimate and CI in coefficient units, not exponentiated
    wide <- origin_rows %>%
      mutate(
        Est_str = sprintf("%.2f", estimate),
        CI_str  = sprintf("[%.2f, %.2f]", conf_low, conf_high),
        p_str   = formatC(p.value, format = "g", digits = 2)
      ) %>%
      select(Origin, Model, Est_str, CI_str, p_str) %>%
      tidyr::pivot_wider(
        names_from  = Model,
        values_from = c(Est_str, CI_str, p_str),
        names_glue  = "{Model}__{.value}"
      )

    base_label <- "Baseline (Demographics Only)"
    full_label <- "Full (With Human Capital)"

    cols_in_order <- c(
      "Origin",
      paste0(base_label, "__Est_str"),
      paste0(base_label, "__CI_str"),
      paste0(full_label, "__Est_str"),
      paste0(full_label, "__CI_str"),
      paste0(full_label, "__p_str")
    )
    cols_in_order <- intersect(cols_in_order, names(wide))
    wide <- wide[, cols_in_order, drop = FALSE]

    col_labels <- c("Origin",
                    "β (Base)", "95% CI (Base)",
                    "β (Full)", "95% CI (Full)",
                    "p (Full)")
    col_labels <- col_labels[seq_len(ncol(wide))]

    return(list(table = wide, col_labels = col_labels))
  }
}
```


```{r forest-plot-all-regions, fig.height=6, fig.width=9, echo=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)

# Pull the nested model results for Net Worth Percentile from the cache
res_nw <- get_h1_results("nw_pctl")

if (!is.list(res_nw) || !isTRUE(res_nw$error)) {
  
  # Filter down to ALL origin region coefficients
  plot_data <- res_nw %>%
    filter(grepl("^origin_region", term)) %>%
    mutate(
      Region = gsub("origin_region", "", term)
    )
  
  # Sort the y-axis dynamically based on the baseline wealth gap
  # This makes the chart much cleaner to read
  baseline_order <- plot_data %>%
    filter(Model == "Baseline (Demographics Only)") %>%
    arrange(estimate) %>%
    pull(Region)
    
  plot_data <- plot_data %>%
    mutate(Region = factor(Region, levels = baseline_order))
  
  # Generate the Forest Plot
  p_all <- ggplot(plot_data, aes(x = estimate, y = Region, color = Model, shape = Model)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "black", linewidth = 0.8) +
    geom_pointrange(aes(xmin = conf_low, xmax = conf_high),
                    position = position_dodge(width = 0.5), size = 0.8, linewidth = 1) +
    theme_minimal(base_size = 12) +
    labs(
      title = "Origin-Region Effects on Net Worth Percentile",
      subtitle = "Testing segmented assimilation vs. human capital convergence (Ref: Latin American Origin)",
      x = "Coefficient Estimate (Difference in Net Worth Percentile)",
      y = NULL
    ) +
    scale_color_manual(values = c("Baseline (Demographics Only)" = "#E69F00", 
                                  "Full (With Human Capital)" = "#0072B2")) +
    theme(
      legend.position = "bottom", 
      legend.title = element_blank(),
      plot.title = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
    
  print(p_all)
} else {
  cat("Model failed to converge or run properly.")
}
```

```{r forest-plots-binary, fig.height=10, fig.width=9, echo=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)

# 1. Define the binary outcomes to loop through
binary_outcomes <- list(
  own_retire    = "Retirement Account (Parallel)",
  own_stocks    = "Direct Stocks (Parallel)",
  own_home      = "Homeownership (Selective)",
  credit_denied = "Credit Denied (Downward)",
  no_checking   = "Unbanked (Downward)"
)

# Initialize an empty dataframe to catch all the results
all_binary_results <- data.frame()

# 2. Run the nested models for each outcome and stack the data
for (var_name in names(binary_outcomes)) {
  label <- binary_outcomes[[var_name]]
  res <- get_h1_results(var_name)
  
  if (is.list(res) && !isTRUE(res$error)) {
    clean_res <- res %>%
      filter(grepl("^origin_region", term)) %>%
      mutate(
        Region = gsub("origin_region", "", term),
        Outcome = label
      )
    all_binary_results <- bind_rows(all_binary_results, clean_res)
  }
}

# 3. Sort the Regions for a clean visual (using Homeownership as the anchor)
if (nrow(all_binary_results) > 0) {
  baseline_order <- all_binary_results %>%
    filter(Outcome == "Homeownership (Selective)", Model == "Baseline (Demographics Only)") %>%
    arrange(odds_ratio) %>%
    pull(Region)
  
  all_binary_results <- all_binary_results %>%
    mutate(Region = factor(Region, levels = baseline_order))
  
  # 4. Generate the Faceted Forest Plot
  p_binary <- ggplot(all_binary_results, aes(x = odds_ratio, y = Region, color = Model, shape = Model)) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "black", linewidth = 0.8) +
    geom_pointrange(aes(xmin = conf_low, xmax = conf_high),
                    position = position_dodge(width = 0.5), size = 0.6, linewidth = 0.8) +
    facet_wrap(~Outcome, scales = "free_x", ncol = 2) +
    scale_x_log10() + # CRITICAL: Makes Odds Ratios visually symmetric around 1.0
    theme_minimal(base_size = 11) +
    labs(
      title = "Origin-Region Effects on Specific Financial Behaviors",
      subtitle = "Odds Ratios relative to Latin American Origin (x-axis is log-scaled)",
      x = "Odds Ratio (1 = No Difference from Latin American Origin)",
      y = NULL
    ) +
    scale_color_manual(values = c("Baseline (Demographics Only)" = "#E69F00", 
                                  "Full (With Human Capital)" = "#0072B2")) +
    theme(
      legend.position = "bottom", 
      legend.title = element_blank(),
      plot.title = element_text(face = "bold"),
      panel.grid.minor = element_blank(),
      strip.text = element_text(face = "bold", size = 11),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
  
  print(p_binary)
} else {
  cat("Binary models failed to converge.")
}
```




```{r h1-parallel, results='asis'}
if (n_fb >= 100) {

  cat("### Parallel-integration outcomes\n\n")

  for (info in list(
    list(var = "own_retire", label = "Retirement Account Ownership"),
    list(var = "own_stocks", label = "Direct Stock Ownership")
  )) {

    cat(sprintf("#### %s\n\n", info$label))

    res <- get_h1_results(info$var)

    if (is.list(res) && isTRUE(res$error)) {
      cat(sprintf("*Model failed: %s*\n\n", res$message))
      next
    }

    formatted <- format_h1_display(res, "binomial")
    if (is.null(formatted)) {
      cat("*No origin-region coefficients in output.*\n\n")
      next
    }

    print(kable(formatted$table,
                col.names = formatted$col_labels,
                row.names = FALSE,
                caption = sprintf("%s — Latin American origin = reference",
                                  info$label)))
    cat("\n\n")
  }
}
```

```{r h1-downward, results='asis'}
if (n_fb >= 100) {

  cat("### Downward-assimilation outcomes\n\n")

  for (info in list(
    list(var = "credit_denied", label = "Credit Denied or Got Less in Past Year"),
    list(var = "no_checking",   label = "No Checking Account")
  )) {

    cat(sprintf("#### %s\n\n", info$label))

    res <- get_h1_results(info$var)

    if (is.list(res) && isTRUE(res$error)) {
      cat(sprintf("*Model failed: %s*\n\n", res$message))
      next
    }

    formatted <- format_h1_display(res, "binomial")
    if (is.null(formatted)) {
      cat("*No origin-region coefficients in output.*\n\n")
      next
    }

    print(kable(formatted$table,
                col.names = formatted$col_labels,
                row.names = FALSE,
                caption = sprintf("%s — Latin American origin = reference",
                                  info$label)))
    cat("\n\n")
  }
}
```

```{r h1-selective, results='asis'}
if (n_fb >= 100) {

  cat("### Selective-assimilation outcomes\n\n")

  # ---- Homeownership ----
  cat("#### Homeownership\n\n")
  res_home <- get_h1_results("own_home")

  if (is.list(res_home) && isTRUE(res_home$error)) {
    cat(sprintf("*Model failed: %s*\n\n", res_home$message))
  } else {
    formatted_home <- format_h1_display(res_home, "binomial")
    if (!is.null(formatted_home)) {
      print(kable(formatted_home$table,
                  col.names = formatted_home$col_labels,
                  row.names = FALSE,
                  caption = "Homeownership — Latin American origin = reference"))
      cat("\n\n")
    }
  }

  # ---- Net Worth Percentile (OLS) ----
  cat("#### Net-Worth Percentile (linear regression on full SCF distribution)\n\n")
  res_nw <- get_h1_results("nw_pctl")

  if (is.list(res_nw) && isTRUE(res_nw$error)) {
    cat(sprintf("*Model failed: %s*\n\n", res_nw$message))
  } else {
    formatted_nw <- format_h1_display(res_nw, "ols")
    if (!is.null(formatted_nw)) {
      print(kable(formatted_nw$table,
                  col.names = formatted_nw$col_labels,
                  row.names = FALSE,
                  caption = "Net-Worth Percentile — Latin American origin = reference. Coefficients are percentile-point differences vs. the Latin American reference category, holding controls constant."))
      cat("\n\n")
    }
  }
}
```

## H2: Pathway-Pattern Tests

The three segmented-assimilation pathways predict joint patterns across outcomes within
origin region. The pattern matrix below summarizes origin-region odds ratios across the
six outcomes from the H1 specifications.

Three pathway-specific patterns would constitute affirmative evidence for segmented
assimilation:

1. **Parallel-integration profile** (Asian and European origin): elevated odds of
   retirement-account and stock ownership; baseline or below-baseline odds of credit
   denial and lack of checking account.
2. **Downward-assimilation profile** (Caribbean origin and "Other" origin): elevated
   odds of credit denial and lack of checking account; depressed odds of retirement and
   stock ownership.
3. **Selective-assimilation profile** (Latin American reference, with relative deviations
   for other groups): high homeownership coupled with low mainstream-finance
   participation, suggesting alternative pathways to asset accumulation.

The human-capital convergence model predicts none of these — odds ratios near 1.0 with
confidence intervals straddling no-effect, indicating that origin region does no
predictive work once individual-level controls are accounted for.

```{r h2-pattern-matrix, results='asis'}
if (n_fb >= 100) {

  # Re-fit all H1 models and collect the origin-region coefficients into a matrix
  outcomes_for_h2 <- list(
    list(var = "own_retire",    label = "Retirement",     family = "binomial"),
    list(var = "own_stocks",    label = "Stocks",         family = "binomial"),
    list(var = "credit_denied", label = "Credit denied",  family = "binomial"),
    list(var = "no_checking",   label = "No checking",    family = "binomial"),
    list(var = "own_home",      label = "Homeowner",      family = "binomial"),
    list(var = "nw_pctl",       label = "NW pctile",      family = "ols")
  )

  # Initialize the pattern matrix with the corrected origin labels
  origin_levels_no_ref <- c("Caribbean/West Indies",
                            "Asian",
                            "European/Other developed",
                            "Other")

  pattern_df <- data.frame(`Origin Region` = origin_levels_no_ref,
                           check.names = FALSE,
                           stringsAsFactors = FALSE)

  for (info in outcomes_for_h2) {
    res <- get_h1_results(info$var)

    if (is.list(res) && isTRUE(res$error)) {
      pattern_df[[info$label]] <- "--"
      next
    }

    # FIX: Isolate the Full model so we only get ONE coefficient per region
    res_full <- res %>% filter(Model == "Full (With Human Capital)")

    cell_values <- sapply(origin_levels_no_ref, function(lvl) {
      term_name <- paste0("origin_region", lvl)
      idx <- which(res_full$term == term_name)

      if (length(idx) == 0) return("--")

      est <- res_full$estimate[idx]
      pval <- res_full$p.value[idx]

      # For binomial: report odds ratio with stars
      # For OLS: report estimate with stars
      if (info$family == "binomial") {
        val <- exp(est)
        formatted <- sprintf("%.2f", val)
      } else {
        formatted <- sprintf("%.1f", est)
      }

      stars <- if (is.na(pval)) "" else
               if (pval < 0.001) "***" else
               if (pval < 0.01)  "**"  else
               if (pval < 0.05)  "*"   else ""

      paste0(formatted, stars)
    })

    pattern_df[[info$label]] <- cell_values
  }

  cat("**Table 3. Origin-region × outcome pattern matrix (Full Model)**\n\n")
  cat("Cells contain odds ratios (binary outcomes) or coefficient estimates ",
      "(net-worth percentile) relative to Latin American origin (reference). ",
      "Stars: \\* p<0.05, \\*\\* p<0.01, \\*\\*\\* p<0.001.\n\n", sep = "")
  print(kable(pattern_df, row.names = FALSE,
              caption = "Origin-region effects across outcomes (net of human capital)"))
  cat("\n\n")
}
```

## H3: Years-in-U.S. × Origin Interaction

Interaction models test whether duration-of-residence effects on financial outcomes
differ by origin region. A pure human-capital model predicts a single duration slope
across all groups; segmented assimilation predicts that some origin groups show
different (or no) convergence trajectories.

```{r h3-interaction, results='asis'}
if (n_fb >= 100) {

  cat("**Table 4. Origin × Years-in-U.S. interaction tests**\n\n")
  cat("For each outcome, we report each individual Origin × Years-in-U.S. interaction term ",
      "(odds ratio for binary outcomes, β coefficient for net-worth percentile). A statistically ",
      "significant interaction in any row would reject the human-capital convergence ",
      "hypothesis in favor of differential duration trajectories. ",
      "The years-in-U.S. main-effect slope (Latin American reference) is also reported.\n\n")

  # Pre-check: do all origin × outcome cells have at least 5 obs and within-cell variation?
  # Quasi-complete separation is the most common cause of model failure.
  fb_imp1 <- scf_fb$mi_design[[1]]$variables

  cell_check <- fb_imp1 %>%
    filter(!is.na(origin_region)) %>%
    group_by(origin_region) %>%
    summarise(
      n = n(),
      retire_var    = var(own_retire,    na.rm = TRUE),
      stocks_var    = var(own_stocks,    na.rm = TRUE),
      denied_var    = var(credit_denied, na.rm = TRUE),
      nochk_var     = var(no_checking,   na.rm = TRUE),
      home_var      = var(own_home,      na.rm = TRUE),
      .groups = "drop"
    )

  # Identify outcomes with separation risk: cell with zero variance
  has_separation <- function(outcome_var) {
    var_col <- switch(outcome_var,
      own_retire    = "retire_var",
      own_stocks    = "stocks_var",
      credit_denied = "denied_var",
      no_checking   = "nochk_var",
      own_home      = "home_var",
      NA
    )
    if (is.na(var_col)) return(FALSE)
    any(cell_check[[var_col]] == 0 | is.na(cell_check[[var_col]]))
  }

  h3_long_results <- list()

  for (info in list(
    list(var = "own_retire",    label = "Retirement",     family = "binomial"),
    list(var = "own_stocks",    label = "Stocks",         family = "binomial"),
    list(var = "credit_denied", label = "Credit denied",  family = "binomial"),
    list(var = "no_checking",   label = "No checking",    family = "binomial"),
    list(var = "own_home",      label = "Homeowner",      family = "binomial"),
    list(var = "nw_pctl",       label = "NW pctile",      family = "ols")
  )) {

    # Skip outcomes with quasi-complete separation
    if (info$family == "binomial" && has_separation(info$var)) {
      h3_long_results[[info$label]] <- list(
        skipped = TRUE,
        reason  = "Quasi-complete separation: at least one origin region has zero within-cell variance on this outcome. Interaction model would not converge."
      )
      next
    }

    # V8 fix: replaced yiu_clean + yiu_missing with raw years_in_us
    int_rhs <- "origin_region * years_in_us + age_centered + educ_cat + currently_married + log_income"
    int_formula <- as.formula(paste(info$var, "~", int_rhs))
    environment(int_formula) <- globalenv()    # V7 fix: same as H1

    fit <- tryCatch({
      if (info$family == "binomial") {
        scf_glm(scf_fb, int_formula, family = binomial())
      } else {
        scf_ols(scf_fb, int_formula)
      }
    }, error = function(e) list(error = TRUE, message = conditionMessage(e)))

    if (is.list(fit) && isTRUE(fit$error)) {
      h3_long_results[[info$label]] <- list(error = TRUE, message = fit$message)
      next
    }

    h3_long_results[[info$label]] <- list(results = fit$results, family = info$family)
  }

  # Print results, one outcome at a time
  for (label in names(h3_long_results)) {
    cat(sprintf("#### %s\n\n", label))

    entry <- h3_long_results[[label]]

    if (isTRUE(entry$skipped)) {
      cat(sprintf("*Skipped — %s*\n\n", entry$reason))
      next
    }

    if (isTRUE(entry$error)) {
      cat(sprintf("*Model failed: %s*\n\n", entry$message))
      next
    }

    res <- entry$results
    fam <- entry$family

    # Find both the YIU main effect and all origin × YIU interaction terms
    yiu_main <- res %>% filter(term == "years_in_us")
    int_terms <- res %>%
      filter(grepl("origin_region.*:years_in_us|years_in_us:origin_region", term))

    if (nrow(int_terms) == 0) {
      cat("*No interaction terms found.*\n\n")
      next
    }

    # Format compactly
    if (fam == "binomial") {
      display_main <- yiu_main %>%
        mutate(Term = "Years in U.S. (Latin American reference slope)",
               Estimate = sprintf("OR = %.3f", exp(estimate)),
               `95% CI` = sprintf("[%.3f, %.3f]",
                                   exp(estimate - 1.96 * std.error),
                                   exp(estimate + 1.96 * std.error)),
               `p-value` = formatC(p.value, format = "g", digits = 2)) %>%
        select(Term, Estimate, `95% CI`, `p-value`)

      display_int <- int_terms %>%
        mutate(Term = paste(gsub("origin_region|:years_in_us|years_in_us:", "", term), "× YIU"),
               Estimate = sprintf("OR = %.3f", exp(estimate)),
               `95% CI` = sprintf("[%.3f, %.3f]",
                                   exp(estimate - 1.96 * std.error),
                                   exp(estimate + 1.96 * std.error)),
               `p-value` = formatC(p.value, format = "g", digits = 2)) %>%
        select(Term, Estimate, `95% CI`, `p-value`)
    } else {
      display_main <- yiu_main %>%
        mutate(Term = "Years in U.S. (Latin American reference slope)",
               Estimate = sprintf("β = %.3f", estimate),
               `95% CI` = sprintf("[%.3f, %.3f]",
                                   estimate - 1.96 * std.error,
                                   estimate + 1.96 * std.error),
               `p-value` = formatC(p.value, format = "g", digits = 2)) %>%
        select(Term, Estimate, `95% CI`, `p-value`)

      display_int <- int_terms %>%
        mutate(Term = paste(gsub("origin_region|:years_in_us|years_in_us:", "", term), "× YIU"),
               Estimate = sprintf("β = %.3f", estimate),
               `95% CI` = sprintf("[%.3f, %.3f]",
                                   estimate - 1.96 * std.error,
                                   estimate + 1.96 * std.error),
               `p-value` = formatC(p.value, format = "g", digits = 2)) %>%
        select(Term, Estimate, `95% CI`, `p-value`)
    }

    combined <- bind_rows(display_main, display_int)
    print(kable(combined, row.names = FALSE,
                caption = sprintf("%s: Years-in-U.S. main effect and origin × YIU interactions", label)))
    cat("\n\n")
  }

  cat("Significant interaction terms indicate that the duration-of-residence slope ",
      "for that origin group differs from the Latin American (reference) slope, supporting ",
      "the segmented-assimilation prediction of differential trajectories.\n\n", sep = "")
}
```

# Duration-Stratified Analysis

## Motivation

The 2022 SCF foreign-born sample has a mean of 36.8 years lived in the United States,
which suggests a population dominated by long-tenured immigrants who arrived in the
1980s and 1990s. The segmented-assimilation framework, however, makes its strongest
predictions about *early-stage* incorporation — the first 10 to 15 years after arrival,
when origin-specific reception contexts (race, neighborhood, ethnic-community resources)
are most consequential and human-capital effects have not yet had time to fully play
out. By the time immigrants have been in the U.S. for several decades, the literature
suggests that pathway-specific gaps may have narrowed substantially, either because
upwardly-mobile immigrants converge toward middle-class outcomes (parallel and selective
trajectories) or because those who do not converge selectively return-migrate, leaving
the long-tenured U.S. population enriched in successful integrators.

If this duration-narrowing dynamic operates in U.S. wealth data, the null findings in
Sections 5.1–5.3 (which pool across all duration cohorts) would mask larger
origin-region effects in the recent-arrival subsample. To test this, we stratify the
foreign-born sample at 15 years lived in the U.S. — the standard threshold in the
segmented-assimilation literature — and re-estimate the H1 origin-region models
separately within each stratum.

## Sample Stratification

```{r path-b-stratification, results='asis'}
# ============================================================================
# Build two stratum subset designs: recent (<=15 years) and long-tenured (>15)
# ============================================================================
# Pattern: copy scf_fb, subset each implicate's design by yiu_stratum.

scf_fb_recent <- scf_fb
scf_fb_recent$mi_design <- lapply(scf_fb$mi_design, function(des) {
  subset(des, years_in_us <= 15)
})
n_fb_recent <- nrow(scf_fb_recent$mi_design[[1]]$variables)

scf_fb_long <- scf_fb
scf_fb_long$mi_design <- lapply(scf_fb$mi_design, function(des) {
  subset(des, years_in_us > 15)
})
n_fb_long <- nrow(scf_fb_long$mi_design[[1]]$variables)

cat(sprintf("**Stratum 1 (Recent arrivals, ≤15 years in U.S.):** n = %d unweighted\n\n",
            n_fb_recent))
cat(sprintf("**Stratum 2 (Long-tenured, >15 years in U.S.):** n = %d unweighted\n\n",
            n_fb_long))

# Cross-tab of origin region by stratum, to flag small cells
strat_xtab <- scf_fb$mi_design[[1]]$variables %>%
  filter(!is.na(origin_region)) %>%
  mutate(stratum = ifelse(years_in_us <= 15, "Recent (≤15 yr)", "Long-tenured (>15 yr)")) %>%
  count(origin_region, stratum) %>%
  tidyr::pivot_wider(names_from = stratum, values_from = n, values_fill = 0) %>%
  select(origin_region, `Recent (≤15 yr)`, `Long-tenured (>15 yr)`)

cat("**Table 5. Sample sizes by origin region × duration stratum (unweighted)**\n\n")
print(kable(strat_xtab, row.names = FALSE,
            caption = "Cells below n=20 produce highly imprecise estimates and should be interpreted with caution."))
cat("\n\n")

small_strata <- strat_xtab %>%
  filter(`Recent (≤15 yr)` < 20 | `Long-tenured (>15 yr)` < 20)

if (nrow(small_strata) > 0) {
  cat(sprintf("**Note:** %d origin region(s) have at least one stratum below n=20: %s. ",
              nrow(small_strata),
              paste(as.character(small_strata$origin_region), collapse = ", ")))
  cat("Estimates for these groups in the affected stratum should be treated as exploratory.\n\n")
}
```

## Stratified H1 Models — Three Core Outcomes

We re-estimate H1 models separately within each duration stratum, using three core
outcomes — one per theoretical pathway:

- **Retirement account ownership** (parallel integration)
- **Homeownership** (selective assimilation)
- **Credit denied** (downward assimilation)

If origin-region effects are larger in the recent-arrival stratum than in the
long-tenured stratum, that pattern supports the duration-narrowing version of segmented
assimilation. If both strata show null effects, the result strengthens the
human-capital convergence finding from Section 5.

```{r path-b-stratified-models, results='asis', cache=TRUE}
# ============================================================================
# Stratified H1 models — three core outcomes per stratum
# ============================================================================

stratified_outcomes <- list(
  list(var = "own_retire",    label = "Retirement (parallel integration)"),
  list(var = "own_home",      label = "Homeownership (selective assimilation)"),
  list(var = "credit_denied", label = "Credit denied (downward assimilation)")
)

# Use Full specification only for the stratified analysis
stratified_rhs <- "origin_region + years_in_us + age_centered + educ_cat + currently_married + log_income"

run_stratified_glm <- function(scf_obj, outcome_var) {
  full_formula <- as.formula(paste(outcome_var, "~", stratified_rhs))
  environment(full_formula) <- globalenv()

  fit <- tryCatch({
    scf_glm(scf_obj, full_formula, family = binomial())
  }, error = function(e) list(error = TRUE, message = conditionMessage(e)))

  if (is.list(fit) && isTRUE(fit$error)) return(fit)

  res <- fit$results
  if ("estimate" %in% names(res)) {
    res$odds_ratio <- exp(res$estimate)
    res$conf_low   <- exp(res$estimate - 1.96 * res$std.error)
    res$conf_high  <- exp(res$estimate + 1.96 * res$std.error)
  }
  res
}

# Fit each outcome × stratum combination, store in 6-cell list
stratified_cache <- list()
for (info in stratified_outcomes) {
  message("Fitting stratified models for: ", info$var)

  for (stratum_label in c("recent", "long")) {
    obj <- if (stratum_label == "recent") scf_fb_recent else scf_fb_long
    n_in_stratum <- nrow(obj$mi_design[[1]]$variables)

    if (n_in_stratum < 50) {
      stratified_cache[[paste(info$var, stratum_label, sep = "_")]] <- list(
        error = TRUE,
        message = sprintf("Stratum n = %d, below the n=50 minimum for stable estimation.",
                          n_in_stratum)
      )
      next
    }

    # Cell-size pre-check on the outcome
    cell_var <- obj$mi_design[[1]]$variables %>%
      filter(!is.na(origin_region)) %>%
      group_by(origin_region) %>%
      summarise(v = var(.data[[info$var]], na.rm = TRUE), .groups = "drop")

    if (any(cell_var$v == 0 | is.na(cell_var$v), na.rm = TRUE)) {
      bad_groups <- cell_var %>% filter(v == 0 | is.na(v)) %>% pull(origin_region)
      stratified_cache[[paste(info$var, stratum_label, sep = "_")]] <- list(
        error = TRUE,
        message = sprintf("Quasi-complete separation in: %s",
                          paste(as.character(bad_groups), collapse = ", "))
      )
      next
    }

    stratified_cache[[paste(info$var, stratum_label, sep = "_")]] <-
      run_stratified_glm(obj, info$var)
  }
}

# Build comparative table: rows = origin × outcome, columns = recent OR / long OR
build_strat_row <- function(outcome_var, outcome_label) {
  recent_res <- stratified_cache[[paste(outcome_var, "recent", sep = "_")]]
  long_res   <- stratified_cache[[paste(outcome_var, "long",   sep = "_")]]

  origin_levels_no_ref <- c("Caribbean/West Indies", "Asian",
                            "European/Other developed", "Other")

  rows <- lapply(origin_levels_no_ref, function(lvl) {
    term_name <- paste0("origin_region", lvl)

    recent_cell <- if (is.list(recent_res) && isTRUE(recent_res$error)) {
      "[failed]"
    } else {
      idx <- which(recent_res$term == term_name)
      if (length(idx) == 0) "--" else {
        sprintf("%.2f%s [%.2f, %.2f]",
                recent_res$odds_ratio[idx],
                ifelse(recent_res$p.value[idx] < 0.05, "*", ""),
                recent_res$conf_low[idx],
                recent_res$conf_high[idx])
      }
    }

    long_cell <- if (is.list(long_res) && isTRUE(long_res$error)) {
      "[failed]"
    } else {
      idx <- which(long_res$term == term_name)
      if (length(idx) == 0) "--" else {
        sprintf("%.2f%s [%.2f, %.2f]",
                long_res$odds_ratio[idx],
                ifelse(long_res$p.value[idx] < 0.05, "*", ""),
                long_res$conf_low[idx],
                long_res$conf_high[idx])
      }
    }

    data.frame(Outcome = outcome_label,
               Origin = lvl,
               `OR Recent (≤15 yr)` = recent_cell,
               `OR Long-tenured (>15 yr)` = long_cell,
               check.names = FALSE,
               stringsAsFactors = FALSE)
  })
  do.call(rbind, rows)
}

# Assemble the table
strat_table <- do.call(rbind, lapply(stratified_outcomes, function(info) {
  build_strat_row(info$var, info$label)
}))

cat("**Table 6. Origin-region effects by duration stratum (Full specification, OR vs. Latin American reference)**\n\n")
print(kable(strat_table, row.names = FALSE,
            caption = "Stars: * p<0.05. [failed] indicates model did not converge or stratum cell too small. Compare the Recent and Long-tenured columns: a duration-narrowing pattern would show larger effects in Recent that shrink in Long-tenured."))
cat("\n\n")

# Surface any failures explicitly
failed_models <- which(sapply(stratified_cache, function(x) is.list(x) && isTRUE(x$error)))
if (length(failed_models) > 0) {
  cat("**Stratified-model failures (",
      length(failed_models), " of ", length(stratified_cache), "):**\n\n", sep = "")
  for (fm in names(failed_models)) {
    cat(sprintf("- `%s`: %s\n", fm, stratified_cache[[fm]]$message))
  }
  cat("\n")
}
```

## Interpreting the Stratified Results

Three patterns are possible:

1. **Recent and long-tenured strata both show null effects.** This strengthens the
human-capital convergence interpretation from Section 5: origin-region differences in
financial outcomes are absent across the entire duration distribution, not just averaged
across it.

2. **Recent stratum shows large origin-region effects that shrink to null in the
long-tenured stratum.** This supports the duration-narrowing version of segmented
assimilation: pathway-specific gaps exist in early incorporation but converge with time
in the U.S. Under this interpretation, the pooled null findings from Section 5 reflect
sample composition (heavy weighting toward long-tenured immigrants) rather than absence
of segmentation.

3. **Both strata show large effects with similar magnitude.** This would support the
strong segmented-assimilation prediction (origin region matters at all durations) but
also raise concerns about why those effects didn't appear in the pooled main analysis.
Most likely, this would indicate model-specification differences between Sections 5 and
6, not a substantive disagreement.

The substantive interpretation depends on which pattern the data show.

# Robustness Checks

## Three-Category Origin Recode

If any cell in the 5-category origin coding falls below n = 30, we re-estimate models
with a collapsed 3-category coding (Latin American, Asian, European/Other) for stability.

## Race Versus Origin

Origin region and racial classification (`racecl4`) are correlated but not identical.
Specifications adding `race_cat` to the H1 models test whether origin-region effects are
robust to controlling for racial classification, and whether racial classification
mediates origin-region effects.

## Net-Worth Specification

The primary net-worth specification uses percentile within the full SCF distribution. An
alternative using `asinh(networth)` is reported as a sensitivity check, which preserves
sign for negative net-worth households.

# Discussion

This paper put two competing theoretical models head-to-head using the first
nationally representative U.S. wealth data with origin-region detail. The
human-capital convergence model and the segmented-assimilation framework make sharply
different predictions about whether origin region carries independent explanatory
power for financial outcomes net of human capital. The data favor the human-capital
convergence model.

## Why the Pooled Analysis Fails to Support Segmentation

Across all six outcomes — retirement-account ownership, stock ownership, credit
denial, lack of a checking account, homeownership, and net-worth percentile — the
pooled H1 models produce origin-region odds ratios near 1.0 and net-worth-percentile
coefficients near 0.0, all with confidence intervals straddling no-effect. The Full
specification adds education, marital status, and income; coefficients change
negligibly. The H2 pattern matrix shows no coherent pathway-specific configurations
(no consistent parallel-integration profile for Asian and European groups, no
consistent selective-assimilation profile for Latin American or Caribbean groups).
The H3 interaction tests fail to detect statistically distinguishable duration slopes
across origin regions.

Individually, each of these results is a failure to reject the human-capital
convergence null. Jointly, across six outcomes and three falsifiable predictions, they
constitute meaningful evidence *for* convergence and against the strong segmented-
assimilation prediction in this sample.

## The Survivorship-Bias Interpretation

The most important limitation of the pooled analysis is the duration profile of the
sample. The mean foreign-born respondent in the 2022 SCF has lived in the United
States for 36.8 years — a population dominated by immigrants who arrived in the 1980s
and early 1990s. The segmented-assimilation framework makes its strongest predictions
about *early-stage* incorporation, when origin-specific reception contexts are most
consequential. By the time immigrants have been in the U.S. for several decades, two
mechanisms may operate to narrow pathway-specific gaps:

1. **Convergence within origin groups.** Long-tenured immigrants who continue to
reside in the U.S. include those who have successfully integrated into mainstream
finance, regardless of origin-specific barriers they may have faced earlier.

2. **Selective return migration.** Immigrants who failed to incorporate may
disproportionately return-migrate over the 10-30 year window after arrival, leaving
the long-tenured U.S. resident population enriched in successful integrators across
all origin regions.

Either or both mechanisms would attenuate observed origin-region differences in
late-stage data while leaving them visible in early-stage data. The duration-stratified
analysis in Section 6 is the direct test. Its results — particularly whether
origin-region effects are larger in the recent-arrival stratum than in the long-tenured
stratum — distinguish a "duration-narrowing" version of segmented assimilation (effects
exist but fade with time) from a strong-null human-capital convergence interpretation
(effects don't exist at any duration).

## Implications for Policy and Theory

If the duration-stratified analysis shows null effects in both strata, the substantive
contribution of this paper is to argue that race-based analytical frameworks are
appropriate for U.S. wealth disparities even with origin-region data added. Policy
interventions designed for "Hispanic households" or "Asian households" may not need
within-racial-category origin refinements to be effective.

If the duration-stratified analysis shows larger effects in the recent-arrival
stratum, the contribution is to argue that segmented assimilation operates in U.S.
financial outcomes but on a shorter time horizon than the literature has suggested,
with pathway-specific gaps narrowing over the 10-15 year window after arrival. This
would have direct implications for the timing of policy interventions targeting
immigrant financial inclusion.

In either case, the substantive contribution is a clearer empirical picture of the
domain in which segmented-assimilation theory applies to U.S. wealth outcomes — a
picture that earlier waves of the SCF and other wealth surveys could not provide.

# Limitations

1. **Survivorship bias in the cross-section.** The 2022 SCF cross-section captures
foreign-born respondents who remain in the U.S. as of 2022. Selective return migration
over the prior decades may have produced a sample non-representative of the immigrant
flow. This is the most consequential limitation for interpreting null findings.

2. **Single-wave cross-section.** The 2022 SCF is the first wave with these
immigration questions. No time series is available; cohort and period effects cannot
be separated. Repeating the analysis on the 2025 SCF will eventually allow cohort-
specific tests.

3. **No parental nativity.** Unlike the NSFG used in Papers 3 and 4, the SCF does not
collect parental nativity. The analysis is restricted to first-generation immigrants;
the second-generation extension that segmented-assimilation theory most directly
addresses is not testable in this dataset.

4. **Origin-region collapsing.** Sample sizes preclude finer-grained country-of-origin
analysis. Heterogeneity within "Asian" or "Latin American" categories may obscure
pathway-specific patterns. The Caribbean group in particular (n=40 in the pooled
sample) has limited statistical power.

5. **Statistical power against modest effects.** With ~700 effective observations
across five origin regions and six outcomes, the analysis is well-powered to detect
large segmentation effects (odds ratios above 2.0 or below 0.5) but only modestly
powered to detect smaller effects (odds ratios between 0.7 and 1.4). The null
findings should be interpreted as "no large segmentation effects detected" rather
than "no effects of any size."

6. **Disclosure-avoidance modifications.** SCF public-use data are altered for
disclosure protection, which may attenuate small-cell estimates for the smallest
origin groups (Caribbean and Other).

# Conclusion

This paper conducted the first head-to-head test of the human-capital convergence
model versus the segmented-assimilation framework as competing accounts of foreign-
born household financial outcomes in nationally representative U.S. data. Using newly
added 2022 Survey of Consumer Finances variables on respondent national origin and
years lived in the United States, we tested whether origin region carries independent
explanatory power net of duration and human capital across six financial outcomes
covering all three theoretical pathways.

The pooled analysis (Section 5) finds no statistically significant origin-region
effects across any outcome. The pattern matrix shows no coherent pathway-specific
configurations. The interaction tests detect no differential duration slopes by origin.
The duration-stratified robustness check (Section 6) tests whether segmentation
effects exist in early-stage incorporation but converge with time, distinguishing a
duration-narrowing version of segmented assimilation from strong-null human-capital
convergence.

The contribution to the broader thesis is to close the loop on the immigration
question raised in Papers 3 and 4 (NSFG nativity and fertility outcomes) and Papers 1
and 2 (SCF wealth and parental status). Paper 5 finds that the immigration signal
that matters for fertility (NSFG) does not appear, in the pooled SCF cross-section,
to matter for wealth — suggesting that the structural barriers segmented-assimilation
theory was developed to capture may operate more durably in family-formation outcomes
than in financial-incorporation outcomes, at least in the established immigrant
populations measured by the 2022 SCF.

# Appendix A: Variable List

| Construct | Source variable | Type |
|---|---|---|
| Foreign-born status (component A) | `X6915` (ancestry; non-0 and non-1200) | Raw codebook |
| Foreign-born status (component B) | `X6906` (years respondent has lived in U.S.) | Raw codebook |
| National origin (R) | `X6915` | Raw codebook |
| Spouse Hispanic origin (R) | `X7404` (auxiliary, not used in main models) | Raw codebook |
| Race/ethnicity | `racecl4` | Bulletin extract |
| Age | `age` | Bulletin extract |
| Education | `educ` (recoded into 5 levels) | Bulletin extract |
| Marital status | `married` | Bulletin extract |
| Income | `income` | Bulletin extract |
| Net worth | `networth` | Bulletin extract |
| Net-worth percentile | `nw_pctl` (computed on full SCF distribution) | Derived |
| Retirement assets | `retqliq` | Bulletin extract |
| Stocks | `stocks` | Bulletin extract |
| Credit denied | `turndown` | Bulletin extract |
| Checking balance | `checking` | Bulletin extract |
| Homeowner | `housecl` | Bulletin extract |
| Weight | `wgt` | Bulletin extract |

# Appendix B: scf Package Workflow Notes

This document follows the workflow established in Paper 1 (`scf_ccbal_analysis_v5.Rmd`):

1. `scf_download(2022)` writes raw `.rds` files to `scf_data/`.
2. `scf_load(2022, data_directory = "scf_data")` returns an object with
   `$mi_design[[1..5]]` survey-design slots, one per implicate.
3. `scf_update(scf_obj, new_var = expr, ...)` adds variables to all 5 implicates.
4. `scf_mean()`, `scf_glm()`, `scf_ols()` fit pooled models with Rubin's-Rules pooling
   and replicate-weight-based variance.
5. Subsetting (e.g. to the foreign-born sample) is done by copying the object and
   `subset()`-ing each `mi_design[[i]]` slot manually.

The new 2022 immigration variables (`X6915`, `X6906`, `X7404`, `X7796`) are not in the
Bulletin summary extract that the `scf` package builds. The diagnostic chunk at the
top of this document attempts to merge them in from the raw 2022 SCF `.rds` file. If
the merge fails, those analyses cannot proceed.

# Appendix C: V9 Methodological Notes

This V9 build reframes V8 around a balanced test of segmented assimilation versus
human-capital convergence and adds a duration-stratified robustness check (Section 6).

**V9 changes:**

1. **Reframed Abstract, Section 1, Discussion, and Conclusion.** The earlier versions
   framed the analysis as a test of segmented-assimilation predictions, with
   convergence as an implicit alternative. V9 treats the two models as competing
   hypotheses given equal a priori weight, then reports what the data show.

2. **New Section 6 — Duration-Stratified Analysis.** The pooled foreign-born sample
   has mean YIU of 36.8 years, which may obscure early-stage segmentation effects.
   Section 6 splits the sample at the standard 15-year threshold and re-estimates the
   H1 origin-region models within each stratum, on three core outcomes (one per
   pathway: retirement, homeownership, credit denial). The comparative table allows
   the reader to assess whether origin-region effects are larger in early-stage
   incorporation than in long-tenured immigrants.

3. **Updated Limitations section.** Survivorship bias is now the lead limitation,
   with explicit discussion of return-migration mechanisms. Statistical-power
   limitations are reported as a separate point: the analysis is well-powered against
   large segmentation effects but only modestly powered against smaller effects.

**Earlier-version corrections (carried forward):**

V8 fixed a singularity issue: with the strict foreign-born filter (requires
years_in_us > 0), the missing-indicator variable yiu_missing became constant 0 and
crashed the GLM solver. V8 replaced `yiu_clean + yiu_missing` with raw `years_in_us`
in all model formulas.

V7 fixed a formula-environment scoping issue: when `as.formula()` is called inside a
function, the formula's environment becomes the function's local scope, which doesn't
contain the survey design's variables. V7 sets `environment(formula) <- globalenv()`
so the survey functions can resolve term names against the design's `$variables` data.

V6 corrected the years-in-U.S. variable identity: V5 erroneously used X7437–X7443
(which are mortgage-forbearance variables per the official 2022 SCF changes file). The
correct YIU variable is X6906 for the respondent and X7796 for the spouse.

V5 corrections: foreign-born identification requires both X6915 indicating non-
American ancestry AND X6906 > 0; nw_pctl computed via `scf_update` on the full SCF
sample before subsetting; quasi-complete separation pre-checked in the H1 cache loop
and H3 chunk; H3 interaction errors surfaced explicitly rather than silently
swallowed; H1 result tables reformatted (Baseline/Full side-by-side, origin-region
prefix stripped).
