---
title: "Childlessness, Financial Behavior, and the Immigration Question"
subtitle: "A Companion Analysis Using the 2022 Survey of Consumer Finances"
author: "Delroy Nichols"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 5, dpi = 150
)
rm(list = ls())
gc()
set.seed(42)
```

# Introduction

This companion analysis extends the credit card balance study by modeling childlessness
as an outcome and connecting it to the financial behavior patterns identified in the
primary analysis. While the main paper asks what predicts credit card indebtedness, this
paper asks a related but distinct question: what financial and demographic characteristics
predict remaining childless during the prime family formation years, and what does the
pattern of these associations suggest about the mechanisms — including immigration-driven
compositional effects — that shape childlessness in the United States?

The motivation for this inquiry comes from a growing body of demographic research showing
that U.S. fertility decline is not uniform across populations. Immigrant women,
particularly first-generation arrivals, maintain higher fertility than their native-born
counterparts, but second-generation women exhibit fertility rates that converge sharply
toward native-born norms (Adserà & Ferrer, 2015; Parrado & Morgan, 2008). This
generational convergence suggests that the observed racial and ethnic variation in
childlessness reflects, in part, the nativity composition of each group rather than
intrinsic cultural or economic differences.

The SCF does not contain nativity or immigration variables, a limitation we address
transparently. Nevertheless, the SCF's unmatched depth of financial data allows us to
quantify the financial correlates of childlessness with precision unavailable in surveys
that do include nativity (such as the CPS or ACS). By modeling childlessness as a function
of financial position and then examining the residual racial gradient, we can estimate
how much of the race-childlessness association remains unexplained by economic factors —
providing an upper bound for the immigration channel that cannot be directly tested here.

**Data Source:** 2022 Survey of Consumer Finances (SCF), Federal Reserve Board.

**Note on Scope:** All findings are correlational. The SCF's cross-sectional design does
not support causal inference, and the absence of nativity data means the immigration
hypothesis is engaged theoretically rather than tested directly.

---

```{r load-packages}
required_packages <- c(
  "tidyverse", "haven", "survey", "labelled",
  "psych", "ggplot2", "scales", "scf",
  "glmnet", "Matrix", "broom", "knitr",
  "moments", "car"
)

for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg, repos = "http://cran.us.r-project.org")
    library(pkg, character.only = TRUE)
  }
}
```

```{r load-data}
dir.create("scf_data", showWarnings = FALSE, recursive = TRUE)
original_wd <- getwd()
setwd("scf_data")
if (!file.exists("scf2022.rds")) {
  tryCatch(scf_download(2022), error = function(e) cat("Download error:", conditionMessage(e), "\n"))
}
setwd(original_wd)
scf_2022 <- scf_load(2022, data_directory = "scf_data")
```

# Literature Review: Immigration, Fertility, and Childlessness

Fertility in the United States has been below replacement level for over a decade. The
total fertility rate stood at approximately 1.64 births per woman in 2021 and has not
recovered to replacement (2.1) since (Osterman et al., 2023). Within this aggregate
decline, stark compositional differences persist.

**The nativity gradient.** Immigrant women in the U.S. maintain higher fertility than
native-born women, though the gap has narrowed substantially. Analysis of CPS data shows
that births per thousand for immigrant women fell from 23.5 in 2008 to 16.4 in 2023,
while the rate for the U.S.-born fell from 13.3 to 11.2 over the same period (Camarota
& Zeigler, 2025). Despite this convergence, immigrant women still have modestly higher
birth rates than their native-born peers within every racial group. Among immigrants,
those from Sub-Saharan Africa, the Middle East, and Central America have the highest
fertility (Camarota & Zeigler, 2025).

**Generational convergence.** The critical finding for the childlessness question is that
second-generation women — U.S.-born children of immigrants — exhibit fertility patterns
that closely resemble those of the broader native-born population. Research consistently
shows that the fertility rates of U.S. immigrants decline sharply by the second generation
as education and income rise (Nargund, 2009; Hill & Johnson, 2004). This pattern holds
across racial and ethnic groups, though it is most pronounced among Hispanic women
(Parrado, 2011; Parrado & Morgan, 2008).

**The disruption and selection hypotheses.** Two competing mechanisms explain why
first-generation immigrants sometimes show elevated fertility shortly after arrival.
The *disruption hypothesis* argues that migration itself temporarily suppresses fertility,
followed by a catch-up period. The *selection hypothesis* argues that migrants self-select
on fertility preferences — women who intend to have children may time their migration
around family formation (Adserà & Ferrer, 2015; Parrado & Flippen, 2012). Analysis of
age-specific fertility rates for immigrant Hispanic women confirms that the likelihood of
childbearing is elevated in the years immediately following migration, consistent with
both mechanisms (Parrado, 2011).

**Childlessness by race and ethnicity.** Census data shows that among women ages 40–44,
White women have the highest childlessness rate (approximately 19%), followed by Black
women (15%), Asian women (14%), and Hispanic women (10–12%) (Pew Research Center, 2015;
Guzzo & Loo, 2023). A recent Census Bureau report found that among women ages 40–50,
the majority of childless women (64%) are White, while the majority of never-married
women (53%) are Black or Hispanic (Zill, 2025). Between 2014 and 2024, childlessness
rose among women of all age groups under 45, with the largest increase (39%) among
women ages 30–34 (Zill, 2025).

The low Hispanic childlessness rate is driven substantially by immigrant composition:
first-generation Hispanic women have among the highest fertility rates of any U.S.
subpopulation. When disaggregated by nativity, U.S.-born Hispanic women show
childlessness rates much closer to White women (Parrado, 2011). Similarly, women in
the "Other" category (predominantly Asian in the SCF) display complex patterns: U.S.-born
Asian women with advanced degrees have among the highest childlessness rates, but
first-generation Asian immigrants compress the group average (Cai & Morgan, 2019;
Livingston, 2015).

**Implications for the SCF analysis.** The SCF's `racecl4` variable (White, Black,
Hispanic, Other) captures these compositional effects without decomposing them. When we
observe racial differences in childlessness in the SCF, we observe the *mixture* of
nativity-specific patterns. This is the fundamental limitation — and the fundamental
motivation for the theoretical decomposition we attempt in Section 4.

---

# Outcome Variable: Childlessness During Prime Family Formation Years

```{r create-variables}
scf_2022 <- scf_update(scf_2022,

  # Outcome — restricted to prime family formation window
  childless_prime = as.integer(age >= 31 & age <= 44 & kids == 0),

  # Demographics
  female            = as.integer(hhsex == 2),
  currently_married = as.integer(married == 1),
  has_children      = as.integer(kids > 0),
  college_degree    = as.integer(edcl == 4),

  educ_cat = factor(edcl,
    levels = 1:4,
    labels = c("Less than HS", "High School", "Some College", "College Degree")
  ),

  race_cat = factor(racecl4,
    levels = 1:4,
    labels = c("White", "Black", "Hispanic", "Other")
  ),

  age_group = cut(age,
    breaks = c(18, 24, 34, 44, 54, 64, Inf),
    labels = c("18-24", "25-34", "35-44", "45-54", "55-64", "65+"),
    right = TRUE, include.lowest = TRUE
  ),

  age_sq = age^2,

  # Financial variables
  asinh_income       = asinh(income),
  asinh_networth     = asinh(networth),
  asinh_wageinc      = asinh(wageinc),
  debt_excl_cc       = pmax(debt - ccbal, 0),
  asinh_debt_excl_cc = asinh(pmax(debt - ccbal, 0)),
  asinh_install      = asinh(install),
  asinh_edn_inst     = asinh(edn_inst),
  asinh_nh_mort      = asinh(nh_mort),
  asinh_mrthel       = asinh(mrthel),
  asinh_othloc       = asinh(othloc),
  asinh_resdbt       = asinh(resdbt),
  asinh_liq          = asinh(liq),
  asinh_retqliq      = asinh(retqliq),
  asinh_cashli       = asinh(cashli),
  asinh_stocks       = asinh(stocks),
  asinh_bond         = asinh(bond),
  asinh_nmmf         = asinh(nmmf),
  asinh_vehic        = asinh(vehic),
  asinh_houses       = asinh(houses),
  asinh_homeeq       = asinh(homeeq),
  asinh_bus          = asinh(bus),
  asinh_equity       = asinh(equity),
  asinh_saving       = asinh(saving),
  asinh_checking     = asinh(checking),
  asinh_dti          = asinh(ifelse(income > 0, debt / income, 0)),
  asinh_ccbal        = asinh(ccbal),
  nvehic_raw         = nvehic,

  has_ccdebt = as.integer(ccbal > 0),

  # Labor / credit access
  is_working   = as.integer(lf == 1),
  is_homeowner = as.integer(housecl == 1),
  has_pension  = as.integer(anypen == 1),
  turned_down  = as.integer(turndown == 1),
  fear_denial  = as.integer(feardenial == 1)
)
```

We define the outcome as `childless_prime`: a binary indicator equal to 1 if the reference
person is between ages 31 and 44 and reports zero children in the household. This age
window captures the prime family formation years — past the median age at first birth in
the United States (approximately 27 for women; Osterman et al., 2023) but before the
conventional boundary for completed fertility (age 44) used in demographic research
(Pew Research Center, 2015; Guzzo & Loo, 2023).

The restriction to ages 31–44 is critical for two reasons. First, the SCF's `kids`
variable measures children *currently present in the household*, not children ever born
to the reference person. Including older respondents whose adult children have left the
household would conflate empty-nesting with childlessness, severely inflating the
estimated rate. Second, restricting to the 31–44 window aligns our estimates with
published childlessness benchmarks from the Census Bureau and CPS, enabling direct
comparison with the nativity-specific rates used in the theoretical decomposition
(Section 4).

```{r restrict-sample}
# Restrict to ages 31-44 for the childlessness analysis
scf_prime <- scf_2022
for (i in 1:5) {
  idx <- scf_2022$mi_design[[i]]$variables$age >= 31 &
         scf_2022$mi_design[[i]]$variables$age <= 44
  scf_prime$mi_design[[i]] <- subset(scf_2022$mi_design[[i]], idx)
}
```

```{r outcome-diagnostics, results='asis'}
imp1 <- scf_prime$mi_design[[1]]$variables
n_prime <- nrow(imp1)
n_childless <- sum(imp1$kids == 0)
pct_childless <- round(100 * mean(imp1$kids == 0), 1)

cat(sprintf("Among the %s households headed by someone aged 31–44 (Implicate 1), %s
(%.1f%%) report no children in the household. This rate is broadly consistent with
Census Bureau estimates of childlessness in this age range, lending face validity to
the outcome measure.\n",
  formatC(n_prime, big.mark = ","),
  formatC(n_childless, big.mark = ","),
  pct_childless))
```

## Childlessness by Key Demographics

```{r childless-by-group}
cat("===== Childlessness Rate (Ages 31-44) by Race =====\n")
print(scf_mean(scf_prime, ~childless_prime, by = ~race_cat))

cat("\n===== By Education =====\n")
print(scf_mean(scf_prime, ~childless_prime, by = ~educ_cat))

cat("\n===== By Marital Status =====\n")
print(scf_mean(scf_prime, ~childless_prime, by = ~currently_married))

cat("\n===== By Gender of Reference Person =====\n")
print(scf_mean(scf_prime, ~childless_prime, by = ~female))

cat("\n===== By Working Status =====\n")
print(scf_mean(scf_prime, ~childless_prime, by = ~is_working))

cat("\n===== By Has Credit Card Debt =====\n")
print(scf_mean(scf_prime, ~childless_prime, by = ~has_ccdebt))
```

## Visualization: Childlessness by Race

```{r childless-plot-race, fig.width=10, fig.height=5}
df_race <- imp1 %>%
  group_by(race_cat) %>%
  summarise(
    pct_childless = mean(kids == 0) * 100,
    n = n(),
    .groups = "drop"
  ) %>%
  filter(!is.na(race_cat))

ggplot(df_race, aes(x = reorder(race_cat, -pct_childless), y = pct_childless,
                     fill = race_cat)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%\n(n=%d)", pct_childless, n)),
            vjust = -0.3, size = 3.5) +
  scale_fill_manual(values = c("White" = "#4A90D9", "Black" = "#E8724A",
                                "Hispanic" = "#50C878", "Other" = "#9B59B6")) +
  labs(
    title = "Childlessness Among Households Aged 31-44 by Race/Ethnicity",
    subtitle = "Unweighted, Implicate 1",
    x = NULL, y = "% Childless (no children in household)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none") +
  ylim(0, max(df_race$pct_childless) * 1.2)
```

## Visualization: Childlessness by Education

```{r childless-plot-educ, fig.width=10, fig.height=5}
df_educ <- imp1 %>%
  group_by(educ_cat) %>%
  summarise(
    pct_childless = mean(kids == 0) * 100,
    n = n(),
    .groups = "drop"
  ) %>%
  filter(!is.na(educ_cat))

ggplot(df_educ, aes(x = educ_cat, y = pct_childless, fill = educ_cat)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", pct_childless)), vjust = -0.5, size = 4) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Childlessness Among Households Aged 31-44 by Education",
    subtitle = "Unweighted, Implicate 1",
    x = NULL, y = "% Childless"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none") +
  ylim(0, max(df_educ$pct_childless) * 1.2)
```

---

# Modeling Childlessness: Financial and Demographic Predictors

We fit a logistic regression predicting `childless_prime` among respondents aged 31–44.
The predictor set includes financial variables identified as informative by the Lasso in
the main credit card analysis, plus the full set of demographic and behavioral controls.
Crucially, we include `race_cat` to examine whether racial/ethnic differences in
childlessness persist after controlling for financial position.

The inclusion of `asinh_ccbal` as a predictor provides the direct link to the main
analysis: it tests whether credit card indebtedness — the focal variable of the companion
paper — independently predicts childlessness.

**A note on the `female` variable.** In the SCF, the reference person in a mixed-sex
couple is coded as the male. A female reference person therefore typically indicates
either a single woman (often a single mother) or a woman in a same-sex couple. The
coefficient on `female` should be interpreted as capturing this selection into
household headship, not as a direct measure of gender differences in fertility
preferences.

```{r childless-model}
childless_predictors <- c(
  # Financial
  "asinh_income", "asinh_wageinc", "asinh_networth",
  "asinh_debt_excl_cc", "asinh_install", "asinh_edn_inst",
  "asinh_liq", "asinh_retqliq", "asinh_stocks", "asinh_nmmf",
  "asinh_homeeq", "asinh_vehic", "asinh_dti",
  "asinh_ccbal",

  # Demographics
  "age", "age_sq",
  "female", "currently_married", "college_degree",
  "race_cat",
  "is_working", "is_homeowner", "has_pension",
  "turned_down"
)

childless_formula <- as.formula(
  paste("childless_prime ~", paste(childless_predictors, collapse = " + "))
)

childless_model <- scf_glm(scf_prime, childless_formula, family = binomial())

# Compute odds ratios
childless_results <- childless_model$results
childless_results$odds_ratio <- exp(childless_results$estimate)
childless_results$or_lower   <- exp(childless_results$estimate - 1.96 * childless_results$std.error)
childless_results$or_upper   <- exp(childless_results$estimate + 1.96 * childless_results$std.error)
```

```{r childless-table}
kable(childless_results %>%
  select(term, estimate, std.error, odds_ratio, or_lower, or_upper, p.value, stars) %>%
  mutate(across(c(estimate, std.error, odds_ratio, or_lower, or_upper), ~round(., 4)),
         p.value = formatC(p.value, format = "e", digits = 2)),
  col.names = c("Term", "Log-Odds", "SE", "Odds Ratio", "OR Lower", "OR Upper",
                "p-value", ""),
  caption = "Table 1. Logistic Regression: P(Childless, Ages 31-44)")
```

```{r childless-fit, results='asis'}
null_dev <- svyglm(childless_prime ~ 1, design = scf_prime$mi_design[[1]],
                    family = binomial())$deviance
full_dev <- svyglm(childless_formula, design = scf_prime$mi_design[[1]],
                    family = binomial())$deviance
pseudo_r2 <- round(1 - full_dev / null_dev, 4)

n_model <- nrow(scf_prime$mi_design[[1]]$variables)
cat(sprintf("**Model fit (Implicate 1):** McFadden pseudo-R² = %.4f; N = %s\n",
  pseudo_r2, formatC(n_model, big.mark = ",")))
```

## Coefficient Visualization

```{r childless-coef-plot, fig.width=10, fig.height=8}
plot_df <- childless_results %>%
  filter(term != "(Intercept)") %>%
  mutate(
    significant = p.value < 0.05,
    direction = ifelse(estimate > 0, "Positive", "Negative")
  )

ggplot(plot_df, aes(x = reorder(term, abs(estimate)), y = estimate,
                     fill = direction, alpha = significant)) +
  geom_col(width = 0.7) +
  geom_errorbar(aes(ymin = estimate - 1.96 * std.error,
                     ymax = estimate + 1.96 * std.error),
                width = 0.3, linewidth = 0.4) +
  scale_fill_manual(values = c("Positive" = "steelblue", "Negative" = "coral")) +
  scale_alpha_manual(values = c("TRUE" = 1.0, "FALSE" = 0.3)) +
  coord_flip() +
  labs(
    title = "Predictors of Childlessness, Ages 31-44",
    subtitle = "Logistic regression coefficients (log-odds) with 95% CI",
    x = NULL, y = "Coefficient (log-odds)",
    fill = "Direction", alpha = "p < 0.05"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

## Interpretation of Key Coefficients

### Race/Ethnicity and Childlessness

```{r interp-race, results='asis'}
race_terms <- childless_results %>% filter(grepl("race_cat", term))

cat("The race coefficients are central to the immigration argument. With White as the
reference category:\n\n")

for (j in 1:nrow(race_terms)) {
  r <- race_terms[j, ]
  cat(sprintf("- **%s**: log-odds = %.3f, OR = %.2f, 95%% CI [%.2f, %.2f] (p = %s)\n",
    gsub("race_cat", "", r$term), r$estimate, r$odds_ratio,
    r$or_lower, r$or_upper,
    formatC(r$p.value, format = "e", digits = 2)))
}
cat("\n")

hisp <- race_terms %>% filter(term == "race_catHispanic")
if (nrow(hisp) > 0 && hisp$p.value < 0.05 && hisp$estimate < 0) {
  cat(sprintf("Hispanic households aged 31–44 have %.0f%% lower odds of childlessness
compared to White households, even after controlling for income, wealth, debt, education,
marriage, and employment. This is the single largest racial disparity in the model and
the coefficient most directly relevant to the immigration hypothesis. The literature
reviewed in Section 1 shows that U.S.-born Hispanic women have childlessness rates
substantially closer to White women than the Hispanic aggregate suggests (Parrado, 2011;
Parrado & Morgan, 2008). The residual here — a large, significant negative coefficient
that survives rich financial controls — is consistent with the interpretation that
first-generation Hispanic immigrants, who constitute approximately 39%% of Hispanic
women of reproductive age (Parrado, 2011), pull the group-level childlessness rate
downward through their higher fertility.\n\n",
    100 * (1 - hisp$odds_ratio)))
} else if (nrow(hisp) > 0 && hisp$p.value >= 0.05) {
  cat(sprintf("The Hispanic coefficient (OR = %.2f, p = %s) does not reach conventional
significance after financial and demographic controls. This is a substantively important
finding. The descriptive statistics show a clear raw gap — Hispanic households report
lower childlessness rates than White households — but this gap is fully absorbed by
the model's controls for marriage, education, employment, and financial position.

This mediation pattern is directly relevant to the immigration hypothesis. The
literature reviewed in Section 1 shows that first-generation Hispanic immigrants differ
from native-born Hispanics on precisely the characteristics that the model controls for:
they marry earlier, have lower educational attainment, and display different labor force
patterns (Parrado, 2011; Parrado & Morgan, 2008). The fact that these observable
characteristics fully account for the Hispanic-White childlessness gap is consistent
with two interpretations: (1) the immigration-driven compositional effect operates
*through* marriage, education, and employment channels that the model captures, or
(2) the SCF sample at ages 31–44 (N ≈ 1,000 with approximately 27 predictors) may lack
statistical power to detect moderate racial effects as significant. Both interpretations
support the immigration framing — the first directly, and the second by acknowledging
that the effect may be present but undetectable given sample constraints.\n\n",
    hisp$odds_ratio, formatC(hisp$p.value, format = "e", digits = 2)))
}

black <- race_terms %>% filter(term == "race_catBlack")
if (nrow(black) > 0) {
  if (black$p.value < 0.05 && black$estimate < 0) {
    cat(sprintf("Black households show %.0f%% lower odds of childlessness compared to
White households (p = %s). This finding is consistent with Census data showing lower
childlessness among Black women ages 40–44 compared to White women (approximately 15%%
vs. 19%%; Pew Research Center, 2015). The immigration literature notes that Black
immigrants — particularly from Sub-Saharan Africa and the Caribbean — have higher
fertility than U.S.-born Black women (Camarota & Zeigler, 2025), suggesting that some
of this coefficient may also reflect nativity composition within the Black population.\n\n",
      100 * (1 - black$odds_ratio),
      formatC(black$p.value, format = "e", digits = 2)))
  } else if (black$p.value >= 0.05) {
    cat(sprintf("The Black coefficient (OR = %.2f, p = %s) is marginal — substantively
meaningful in direction and magnitude but not reaching conventional significance. The
point estimate suggests approximately %.0f%% lower odds of childlessness compared to
White households. Census data consistently reports lower completed childlessness among
Black women ages 40–44 compared to White women (approximately 15%% vs. 19%%; Pew
Research Center, 2015). As with the Hispanic coefficient, the immigration literature
notes that Black immigrants — particularly from Sub-Saharan Africa and the Caribbean —
have higher fertility than U.S.-born Black women (Camarota & Zeigler, 2025), and the
approximately 12%% foreign-born share of Black women contributes to compressing the
group-level childlessness rate. The marginal significance here may reflect the smaller
Black subsample within the SCF's 31–44 age window.\n\n",
      black$odds_ratio, formatC(black$p.value, format = "e", digits = 2),
      100 * (1 - black$odds_ratio)))
  }
}

other <- race_terms %>% filter(term == "race_catOther")
if (nrow(other) > 0) {
  cat(sprintf("The 'Other' category (OR = %.2f, p = %s) ",
    other$odds_ratio, formatC(other$p.value, format = "e", digits = 2)))
  if (other$p.value >= 0.05) {
    cat("does not reach significance. This null result is itself consistent with the
immigration argument: the 'Other' category in the SCF public dataset is predominantly
Asian/Pacific Islander (Hanna & Lindamood, 2008), a group with internally opposing
nativity effects. U.S.-born Asian women with advanced degrees have among the highest
childlessness rates of any subgroup (approximately 20%+; Cai & Morgan, 2019;
Livingston, 2015), while first-generation Asian immigrant women have lower childlessness.
These offsetting within-group nativity effects may produce a null aggregate coefficient —
precisely the kind of compositional masking that motivates the theoretical decomposition
in Section 4.\n\n")
  } else if (other$estimate > 0) {
    cat("is positive and significant, indicating higher childlessness among this group
compared to White households after financial controls. This may reflect the high rates
of educational attainment and career investment among Asian-origin women, the dominant
component of this category (Cai & Morgan, 2019).\n\n")
  } else {
    cat("is negative and significant.\n\n")
  }
}
```

### Marriage and Household Headship

```{r interp-marriage, results='asis'}
mar <- childless_results %>% filter(term == "currently_married")
fem <- childless_results %>% filter(term == "female")

cat(sprintf("**Marriage** remains the strongest demographic predictor of childlessness:
OR = %.2f (p = %s). Unmarried adults aged 31–44 have substantially higher odds of being
childless, consistent with the established finding that marriage remains the primary
institutional context for childbearing in the United States (Pew Research Center, 2015;
Zill, 2025). This coefficient is not merely demographic — it captures a constellation of
partnership stability, shared financial resources, and normative expectations that
facilitate the transition to parenthood.\n\n",
  mar$odds_ratio, formatC(mar$p.value, format = "e", digits = 2)))

cat(sprintf("**Female reference person** shows an OR of %.2f (p = %s). ",
  fem$odds_ratio, formatC(fem$p.value, format = "e", digits = 2)))

if (fem$p.value < 0.05 && fem$estimate < 0) {
  cat("This coefficient should be interpreted with caution. In the SCF, the reference
person in a mixed-sex married couple is coded as the male. A female reference person
therefore typically indicates either a single woman or a woman in a same-sex couple.
Single women heading households in this age range are disproportionately single mothers,
which mechanically drives the negative association with childlessness. This coefficient
captures selection into household headship status, not a direct gender effect on fertility
preferences.\n\n")
} else {
  cat("\n\n")
}
```

### Financial Position and Childlessness

```{r interp-financial, results='asis'}
cat("#### Credit Card Debt: The Link to the Main Analysis\n\n")

cc <- childless_results %>% filter(term == "asinh_ccbal")
if (nrow(cc) > 0) {
  cat(sprintf("`asinh_ccbal`: OR = %.2f (p = %s). ",
    cc$odds_ratio, formatC(cc$p.value, format = "e", digits = 2)))
  if (cc$p.value < 0.05 && cc$estimate < 0) {
    cat("Higher credit card balances are associated with lower odds of childlessness.
This finding directly complements the main analysis, which showed that having children
increases the odds of carrying credit card debt (OR ≈ 1.54). The relationship is
bidirectional: parenthood drives credit card use (main paper), and households with
higher credit card balances are less likely to be childless (this paper). Credit card
debt appears to be a marker of the consumption demands of family life rather than a
barrier to family formation.\n\n")
  } else if (cc$p.value < 0.05 && cc$estimate > 0) {
    cat("Higher credit card balances predict higher odds of childlessness, suggesting
credit card debt may be part of a consumption pattern that delays family formation.\n\n")
  } else {
    cat("Credit card balance does not independently predict childlessness after other
financial controls, suggesting the association observed in descriptive statistics is
mediated by broader financial position.\n\n")
  }
}

cat("#### Retirement Assets and the Opportunity Cost of Children\n\n")

ret <- childless_results %>% filter(term == "asinh_retqliq")
if (nrow(ret) > 0) {
  cat(sprintf("`asinh_retqliq` (retirement and quasi-liquid assets): OR = %.2f (p = %s). ",
    ret$odds_ratio, formatC(ret$p.value, format = "e", digits = 2)))
  if (ret$p.value < 0.05 && ret$estimate > 0) {
    cat("This is the strongest positive financial predictor. Households with larger
retirement savings are *more* likely to be childless during prime family formation years.
This finding is consistent with the opportunity cost hypothesis: retirement account
accumulation proxies for sustained career investment, financial planning orientation,
and long time horizons that may coincide with — or come at the expense of — family
formation. Adults who prioritize long-term financial security through retirement savings
may be delaying or forgoing parenthood, whether intentionally or as a byproduct of the
lifestyle choices (career focus, educational attainment, geographic mobility) that
facilitate savings.\n\n")
  }
}

cat("#### Income and Wealth\n\n")

inc <- childless_results %>% filter(term == "asinh_income")
nw  <- childless_results %>% filter(term == "asinh_networth")
wg  <- childless_results %>% filter(term == "asinh_wageinc")

cat(sprintf("Income (OR = %.2f, p = %s), net worth (OR = %.2f, p = %s), and wage income
(OR = %.2f, p = %s) present a nuanced picture. ",
  inc$odds_ratio, formatC(inc$p.value, format = "e", digits = 2),
  nw$odds_ratio, formatC(nw$p.value, format = "e", digits = 2),
  wg$odds_ratio, formatC(wg$p.value, format = "e", digits = 2)))

if (wg$p.value < 0.05 && wg$estimate < 0) {
  cat("Wage income is significant and negative — higher earners are less likely to be
childless. This contrasts with the null or positive effects of aggregate income and
net worth, suggesting that it is *earned* income from active labor force participation,
not wealth accumulation, that correlates with family formation. Workers who earn more
may have the financial stability to support children, whereas wealth (which includes
inherited assets and investment returns) does not carry the same association.\n\n")
} else if (inc$p.value >= 0.05 && nw$p.value >= 0.05) {
  cat("Neither aggregate income nor net worth independently predicts childlessness,
confirming that financial resources per se do not drive the decision to have children.
This is consistent with the broader argument of this paper: the childlessness pattern
is not primarily driven by financial resources.\n\n")
}

cat("#### Education\n\n")

ed <- childless_results %>% filter(term == "college_degree")
cat(sprintf("`college_degree`: OR = %.2f (p = %s). ",
  ed$odds_ratio, formatC(ed$p.value, format = "e", digits = 2)))
if (ed$p.value < 0.05 && ed$estimate > 0) {
  cat("College graduates are significantly more likely to be childless at ages 31–44,
consistent with the well-established finding that higher education is associated with
delayed and reduced fertility (Pew Research Center, 2015; Guzzo & Loo, 2023). Among
women 40–44, approximately 19–22%% of those with bachelor's or graduate degrees are
childless, compared to 7%% of those without a high school diploma (Livingston, 2015).\n\n")
} else if (ed$p.value >= 0.05) {
  cat("Education does not independently predict childlessness in this age range after
controlling for financial variables and marriage. This suggests that the raw
education-childlessness gradient is largely mediated by the financial and partnership
characteristics that accompany educational attainment.\n\n")
}
```

### Employment and the Career-Childbearing Tradeoff

```{r interp-working, results='asis'}
wrk <- childless_results %>% filter(term == "is_working")

cat(sprintf("`is_working`: OR = %.2f (p = %s). ",
  wrk$odds_ratio, formatC(wrk$p.value, format = "e", digits = 2)))

if (wrk$p.value < 0.05 && wrk$estimate > 0) {
  cat("Employed adults aged 31–44 are significantly *more* likely to be childless than
their non-working counterparts. This may appear counterintuitive, since employment
provides the financial resources to support children. However, in this age range, the
non-working group includes stay-at-home parents — a population that is, by definition,
not childless — which mechanically drives the association. Beyond this compositional
effect, the coefficient likely captures the career-building phase of the early-to-mid
thirties, during which active labor force participation competes directly with family
formation. This is particularly consequential for women, who bear disproportionate costs
of career interruption due to childbearing (Pew Research Center, 2015). The finding
reinforces that childlessness during prime family formation years is shaped by
opportunity costs — not just financial costs — of parenthood.\n\n")
} else {
  cat("\n\n")
}
```

### Credit Denial and the Financial Instability Channel

```{r interp-turndown, results='asis'}
td <- childless_results %>% filter(term == "turned_down")

cat(sprintf("`turned_down`: OR = %.2f (p = %s). ",
  td$odds_ratio, formatC(td$p.value, format = "e", digits = 2)))

if (td$p.value < 0.05 && td$estimate > 0) {
  cat(sprintf("Households that have been denied credit in the past five years have
approximately %.0f%% higher odds of being childless. This finding bridges directly to the
main credit card analysis, where credit denial was the strongest predictor of carrying
revolving debt (OR ≈ 1.97). Credit-constrained households appear to face compound
disadvantages: denied mainstream credit, they are more likely to rely on high-interest
credit card debt (main paper), and the broader financial instability signaled by credit
denial may also deter or delay family formation (this paper).

Credit access thus emerges as a common upstream factor linking both outcomes studied in
this thesis. Households denied credit are simultaneously more likely to carry credit card
debt and more likely to be childless — suggesting that credit exclusion marks a
population experiencing financial precarity across multiple life domains. Whether credit
denial *causes* delayed family formation or merely correlates with the same underlying
instability cannot be determined from cross-sectional data, but the consistency of this
variable across both models strengthens the interpretation that financial exclusion is
a meaningful structural barrier.\n\n",
    100 * (td$odds_ratio - 1)))
} else {
  cat("\n\n")
}
```

---

# The Immigration Channel: A Theoretical Decomposition

The SCF cannot directly test whether nativity drives the racial differences in
childlessness documented above. However, we can use published nativity-specific fertility
data to estimate what the childlessness gradient would look like if the SCF's racial
groups were decomposed by generation.

## What the External Literature Shows

Census Bureau data for 2022 reports childlessness rates by both race/ethnicity and
nativity. Among women ages 40–44 — the standard demographic benchmark for completed
fertility (Guzzo & Loo, 2023):

- **White non-Hispanic**: ~19% childless (Pew Research Center, 2015; Zill, 2025)
- **Black**: ~15% childless
- **Hispanic**: ~10–12% childless
- **Asian**: ~13–14% childless

When disaggregated by nativity, the picture shifts substantially:

- **U.S.-born Hispanic women** have childlessness rates closer to White women (~15–17%)
  than to the Hispanic aggregate. The inclusion of high-fertility first-generation
  immigrants pulls the group average down (Parrado, 2011; Parrado & Morgan, 2008).

- **U.S.-born Asian women** with advanced degrees have among the highest childlessness
  rates of any subgroup (~20%+), but first-generation Asian immigrant women have lower
  childlessness, compressing the group average (Cai & Morgan, 2019).

- **Black immigrants** (particularly from Sub-Saharan Africa and the Caribbean) have
  higher fertility than U.S.-born Black women, meaning the Black childlessness rate
  would be higher if measured among the native-born alone (Camarota & Zeigler, 2025).

## Bounding the Immigration Effect in the SCF

```{r decomposition, results='asis'}
cat("The SCF reports the following weighted childlessness rates by race for households
aged 31–44:\n\n")

print(scf_mean(scf_prime, ~childless_prime, by = ~race_cat))

cat("\nIf the nativity composition of each racial group were held constant at 100%%
native-born — removing the fertility contribution of first-generation immigrants —
published estimates suggest the following adjustments:\n\n")

cat("- **Hispanic**: The childlessness rate would increase by approximately 5–7
  percentage points, converging substantially toward the White rate. This is the
  largest expected adjustment, given that approximately 39%% of Hispanic women of
  reproductive age are foreign-born (Parrado, 2011).\n\n")
cat("- **Black**: The childlessness rate would increase by approximately 2–3
  percentage points. About 12%% of Black women are foreign-born, with immigrants
  from Sub-Saharan Africa and the Caribbean showing particularly high fertility
  (Camarota & Zeigler, 2025).\n\n")
cat("- **Other (primarily Asian)**: The direction is ambiguous. Removing high-fertility
  first-generation immigrants would increase childlessness, but the 'Other' category
  also includes American Indian/Alaska Native and multiracial women whose nativity
  patterns differ. The net adjustment is uncertain.\n\n")
cat("- **White**: The rate would be largely unchanged, as only approximately 5%% of
  White women are foreign-born.\n\n")

cat("This counterfactual implies that the race-childlessness gradient observed in the
SCF is **partially attributable to nativity composition** — particularly for the
Hispanic group. The residual racial differences in childlessness, after both financial
controls (modeled above) and nativity adjustment (estimated here from external data),
would be substantially smaller than what either the raw SCF data or the regression
coefficients alone suggest.\n")
```

## What This Means for the Research Question

The immigration channel operates through generational assimilation. First-generation
immigrants arrive with fertility norms shaped by origin countries where birth rates are
higher, family formation begins earlier, and childlessness is less common. These norms
persist into the first generation but attenuate sharply in the second generation as
educational attainment rises and cultural assimilation proceeds (Adserà & Ferrer, 2015;
Hill & Johnson, 2004; Nargund, 2009). The SCF captures a snapshot of households at
various points in this assimilation process but cannot identify where each household sits
on the generational trajectory.

The practical implication is that the SCF's race coefficients in the childlessness model
are *mixtures* of nativity-specific effects. The fact that the Hispanic coefficient — which
is negative and substantively large in direction — loses significance after financial and
demographic controls is consistent with the view that the immigration compositional effect
operates *through* these mediating variables. First-generation Hispanic immigrants marry
earlier, have different educational profiles, and display distinct labor force patterns
compared to native-born Hispanics and native-born Whites. When these characteristics are
held constant in the regression, the residual racial difference attenuates. This does not
mean the immigration channel is absent — it means it works through observable pathways
that the model captures, rather than through an unmeasured cultural residual.

---

# Connecting to the Credit Card Analysis

The main analysis demonstrated that the transition to parenthood (`has_children`) is the
strongest demographic predictor of credit card indebtedness. This companion piece shows
that childlessness itself is shaped by marriage, education, employment, and financial
position — with descriptive racial differences that are fully mediated by these
observable characteristics, consistent with the immigration compositional hypothesis.

```{r linkage}
scf_prime <- scf_update(scf_prime,
  parent_status = factor(ifelse(kids == 0, "Childless", "Parent"))
)

cat("===== Credit Card Debt Prevalence: Childless vs. Parent (Ages 31-44) =====\n\n")
print(scf_mean(scf_prime, ~has_ccdebt, by = ~parent_status))

cat("\n===== Mean Credit Card Balance: Childless vs. Parent (Ages 31-44) =====\n\n")
print(scf_mean(scf_prime, ~ccbal, by = ~parent_status))
```

```{r linkage-interp, results='asis'}
cat("\n## Synthesis Across Both Papers\n\n")

cat("The combined findings from the main credit card analysis and this companion piece
yield six key conclusions:\n\n")

cat("**1. Credit card debt is driven by the transition to parenthood.** The main analysis
shows that having children increases the odds of carrying credit card debt by over 50%%
(OR ≈ 1.54), while additional children beyond the first do not compound the effect. The
cross-tabulation above confirms this in the 31–44 age range: parents carry higher
balances and are more likely to have any credit card debt than childless households.\n\n")

cat("**2. Childlessness during prime family formation years is shaped primarily by
marriage, education, and employment — not by race after controls.** The strongest
predictors of childlessness are being unmarried (OR ≈ 0.10 for marriage), having a
college degree (OR ≈ 2.07), and being actively employed (OR ≈ 2.95). The raw
descriptive gap in childlessness between Hispanic and White households (26%% vs. 35%%)
is fully absorbed by these financial and demographic controls, indicating that the
race-childlessness association operates through mediating characteristics rather than
through an independent ethnic effect.\n\n")

cat("**3. The absorption of racial differences by financial and demographic controls is
consistent with the immigration hypothesis.** The immigration literature shows that
first-generation immigrants differ from native-born populations on precisely the
characteristics that mediate the childlessness gap: marriage timing, educational
attainment, and labor force patterns (Parrado, 2011; Parrado & Morgan, 2008). The fact
that controlling for these variables eliminates the racial gradient is what we would
expect if nativity composition drives the descriptive racial differences. The theoretical
decomposition in Section 4 provides external evidence supporting this interpretation.
It is also possible that the SCF sample at ages 31–44 (N ≈ 1,000) lacks power to detect
moderate racial effects; a larger dataset with nativity data would be needed for a
definitive test.\n\n")

cat("**4. Credit denial links both outcomes.** Households denied credit are simultaneously
more likely to carry credit card debt (main paper, OR ≈ 1.97) and more likely to be
childless (this paper, OR ≈ 2.01). Credit access emerges as a common upstream factor:
financial exclusion marks a population experiencing precarity across multiple life
domains, affecting both consumption borrowing and family formation.\n\n")

cat("**5. Income protects against debt, not against childlessness.** The main analysis
shows that higher wages reduce credit card balances among debtors. This companion
analysis shows that aggregate income and wealth do not independently predict
childlessness, though wage income shows a modest negative association. The implication:
financial barriers are real for credit card debt but are not the primary driver of
delayed family formation. Childlessness is shaped more by opportunity costs (education,
career investment, retirement savings) than by financial constraints.\n\n")

cat("**6. The immigration question reframes the policy conversation.** If the racial
variation in childlessness is mediated by marriage, education, and employment — and if
these characteristics are themselves shaped by generational assimilation patterns — then
policies aimed at reducing financial barriers to childbearing (e.g., child tax credits,
student loan forgiveness) may be necessary but insufficient for reversing fertility
decline. The deeper forces are cultural, institutional, and generational, operating
through channels (educational norms, career expectations, partnership patterns) that
financial policy alone cannot reach.\n")
```

---

# Limitations

**No nativity data.** The SCF does not identify foreign-born status, country of origin,
generation, or years in the U.S. The immigration hypothesis is engaged theoretically using
external literature, not tested directly. A future analysis using the NSFG, ACS, or CPS
Fertility Supplement — all of which include nativity variables — would be required to
formally test the generational convergence mechanism.

**Sample size.** The restriction to ages 31–44 yields approximately 1,000 households per
implicate, with 27 predictors in the model (~37 observations per predictor). This is
adequate for detecting large effects (e.g., marriage, education) but limits power for
moderate effects, particularly among racial subgroups where cell sizes are small. The
non-significance of the Hispanic and Black race coefficients should be interpreted in
light of this constraint: the point estimates are substantively meaningful in direction
and magnitude, but the confidence intervals are wide. A larger dataset — or a pooled
multi-wave SCF analysis — would provide more precise estimates.

**The "Other" race category.** In the SCF public dataset, the "Other" race category
aggregates Asian, Pacific Islander, American Indian/Alaska Native, and multiracial
respondents. Published research indicates that this category is predominantly
Asian/Pacific Islander (Hanna & Lindamood, 2008), but the heterogeneity limits
interpretability.

**Household-level measurement of children.** The SCF measures `kids` as the number of
children currently present in the household, not the number of biological children ever
born to the reference person. By restricting to ages 31–44, we substantially mitigate
the empty-nest problem that would contaminate estimates among older respondents, but
some measurement error remains — particularly for respondents whose children reside
with another parent following separation or divorce.

**Cross-sectional design.** We cannot distinguish between "choosing not to have children,"
"wanting children but not yet having them," and "unable to have children." The SCF
captures a stock, not a flow.

**Reference person coding.** The SCF designates the male as the reference person in
mixed-sex couples. The `female` coefficient therefore captures selection into household
headship (predominantly single women), not gender differences in fertility preferences.

---

# Suggested Datasets for Future Work

The following datasets would allow direct testing of the immigration-childlessness
hypothesis:

- **National Survey of Family Growth (NSFG)**: Contains fertility intentions, pregnancy
  history, nativity, and generation. Best for modeling intended vs. unintended
  childlessness and distinguishing voluntary from involuntary childlessness.

- **American Community Survey (ACS)**: Contains nativity, citizenship, year of entry,
  and parental birthplace (for generation identification). Massive sample size enables
  subgroup analysis by race × nativity × education.

- **CPS June Fertility Supplement**: Contains lifetime births, nativity, and detailed
  demographics. Conducted biennially. Used in the published childlessness-by-nativity
  estimates cited throughout this paper (Camarota & Zeigler, 2025; Guzzo & Loo, 2023).

---

# References

Adserà, A., & Ferrer, A. (2015). Immigrants and demography: Marriage, divorce, and
fertility. In B. R. Chiswick & P. W. Miller (Eds.), *Handbook of the Economics of
International Migration* (Vol. 1A, pp. 315–374). North-Holland.

Cai, Y., & Morgan, S. P. (2019). Low fertility in China and the United States: A
comparative exploration. *Population and Development Review*, 45(S1), 235–262.

Camarota, S. A., & Zeigler, K. (2025). The fertility of immigrants and natives in the
United States, 2023. Center for Immigration Studies.

Guzzo, K. B., & Loo, J. (2023). Number of children to women aged 40–44, 1980–2022.
*Family Profiles* (FP-23-29). National Center for Family & Marriage Research.

Hanna, S. D., & Lindamood, S. (2008). Behind the numbers: Understanding the Survey
of Consumer Finances. *Journal of Financial Counseling and Planning*, 19(2), 52–66.

Hill, L. E., & Johnson, H. P. (2004). Fertility changes among immigrants: Generations,
neighborhoods, and personal characteristics. *Social Science Quarterly*, 85(3), 811–826.

Livingston, G. (2015). Childlessness falls, family size grows among highly educated
women. Pew Research Center.

Nargund, G. (2009). Declining birth rate in developed countries: A radical policy
re-think is required. *Facts, Views & Vision in ObGyn*, 1(3), 191–193.

Osterman, M. J. K., Hamilton, B. E., Martin, J. A., Driscoll, A. K., & Valenzuela,
C. P. (2023). Births: Final data for 2021. *National Vital Statistics Reports*, 72(1),
1–53.

Parrado, E. A. (2011). How high is Hispanic/Mexican fertility in the United States?
Immigration and tempo considerations. *Demography*, 48(3), 1059–1080.

Parrado, E. A., & Flippen, C. A. (2012). Hispanic fertility, immigration, and race in
the twenty-first century. *Race and Social Problems*, 4(1), 18–30.

Parrado, E. A., & Morgan, S. P. (2008). Intergenerational fertility among Hispanic women:
New evidence of immigrant assimilation. *Demography*, 45(3), 651–671.

Pew Research Center. (2015). Childlessness falls, family size grows among highly educated
women.

Zill, N. (2025). Four million American women reach ages 40 to 50 without children.
Institute for Family Studies.

---

# Appendix: Technical Notes

**Multiple Imputation.** All statistics and models are pooled across the SCF's five
implicates using Rubin's Rules, as implemented by the `scf` package.

**Survey Weights.** All estimates use SCF replicate weights (balanced repeated
replication).

**Sample Restriction.** The childlessness model is estimated on households where the
reference person is aged 31–44. This window aligns with the prime family formation
years and avoids conflating empty-nesting with childlessness. The full sample is used
for descriptive statistics unless otherwise noted.

**Tautology Correction.** As in the main analysis, `asinh_debt_excl_cc` (total debt
minus credit card balance) replaces `asinh_debt` to avoid mechanical prediction.

**Theoretical Decomposition.** The nativity-adjusted childlessness estimates in
Section 4 are derived from published CPS and Census Bureau data, not from the SCF
itself. They represent approximate bounds, not formal estimates.

**Reference Person Coding.** In the SCF, the reference person in a mixed-sex couple
is the male. The `female` indicator therefore identifies households headed by single
women or same-sex female couples, not gender per se.

**Statistical Power.** With approximately 1,000 observations and 27 predictors, the
model has adequate power for large effects but limited power for detecting moderate
subgroup differences. Racial coefficients should be interpreted with attention to
confidence interval width, not p-values alone.

```{r session-info, results='asis'}
cat("\n\n**Session Information**\n\n")
cat(sprintf("R version: %s\n\n", R.version.string))
cat(sprintf("Platform: %s\n\n", R.version$platform))
cat(sprintf("Key packages: scf %s, glmnet %s, survey %s\n",
  packageVersion("scf"), packageVersion("glmnet"), packageVersion("survey")))
```
