---
title: "Childlessness, Financial Behavior, and the Immigration Question"
subtitle: "A Companion Analysis Using the 2022 Survey of Consumer Finances"
author: "Delroy Nichols"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 5, dpi = 150
)
rm(list = ls())
gc()
set.seed(42)
```

# Introduction

This companion analysis extends the credit card balance study by modeling childlessness
as an outcome and connecting it to the financial behavior patterns identified in the
primary analysis. While the main paper asks what predicts credit card indebtedness, this
paper asks a related but distinct question: what financial and demographic characteristics
predict remaining childless past age 30, and what does the pattern of these associations
suggest about the mechanisms — including immigration-driven compositional effects — that
drive unplanned childlessness in the United States?

The motivation for this inquiry comes from a growing body of demographic research showing
that U.S. fertility decline is not uniform across populations. Immigrant women, particularly
first-generation arrivals, maintain higher fertility than their native-born counterparts.
However, second-generation women — children of immigrants who were born or raised in the
United States — exhibit fertility rates that converge sharply toward native-born norms.
This generational convergence suggests that the observed racial and ethnic variation in
childlessness reflects, in part, the nativity composition of each group rather than
intrinsic cultural or economic differences.

The SCF does not contain nativity or immigration variables, a limitation we address
transparently. Nevertheless, the SCF's unmatched depth of financial data allows us to
quantify the financial correlates of childlessness with precision unavailable in surveys
that do include nativity (such as the CPS or ACS). By modeling childlessness as a function
of financial position and then examining the residual racial gradient, we can estimate
how much of the race-childlessness association remains unexplained by economic factors —
providing an upper bound for the immigration channel that cannot be directly tested here.

**Data Source:** 2022 Survey of Consumer Finances (SCF), Federal Reserve Board.

**Note on Scope:** All findings are correlational. The SCF's cross-sectional design does
not support causal inference, and the absence of nativity data means the immigration
hypothesis is engaged theoretically rather than tested directly.

---

```{r load-packages}
required_packages <- c(
  "tidyverse", "haven", "survey", "labelled",
  "psych", "ggplot2", "scales", "scf",
  "glmnet", "Matrix", "broom", "knitr",
  "moments", "car"
)

for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg, repos = "http://cran.us.r-project.org")
    library(pkg, character.only = TRUE)
  }
}
```

```{r load-data}
dir.create("scf_data", showWarnings = FALSE, recursive = TRUE)
original_wd <- getwd()
setwd("scf_data")
if (!file.exists("scf2022.rds")) {
  tryCatch(scf_download(2022), error = function(e) cat("Download error:", conditionMessage(e), "\n"))
}
setwd(original_wd)
scf_2022 <- scf_load(2022, data_directory = "scf_data")
```

# Literature Review: Immigration, Fertility, and Childlessness

Fertility in the United States has been below replacement level for over a decade. The
total fertility rate stood at approximately 1.64 births per woman in 2021 and has not
recovered to replacement (2.1) since. Within this aggregate decline, however, stark
compositional differences persist.

**The nativity gradient.** Immigrant women in the U.S. maintain higher fertility than
native-born women, though the gap has narrowed substantially. Research using CPS and ACS
data shows that foreign-born fertility declined faster than native-born fertility between
2008 and 2023, with Hispanic immigrant fertility showing the sharpest decline. Despite
this convergence, immigrant women still have modestly higher birth rates than their
native-born peers within every racial group.

**Generational convergence.** The critical finding for the childlessness question is that
second-generation women — U.S.-born children of immigrants — exhibit fertility patterns
that closely resemble those of the broader native-born population. Research consistently
shows that fertility rates of U.S. immigrants decline sharply by the second generation
as a result of rising educational attainment and income. This pattern holds across
racial and ethnic groups, though it is most pronounced among Hispanic women.

**The disruption and selection hypotheses.** Two competing mechanisms explain why
first-generation immigrants sometimes show elevated fertility shortly after arrival.
The *disruption hypothesis* argues that migration itself temporarily suppresses
fertility, followed by a catch-up period. The *selection hypothesis* argues that
migrants self-select on fertility preferences — women who intend to have children may
time their migration around family formation. Both mechanisms imply that the observed
fertility of any racial group in the U.S. is a mixture of distinct nativity-specific
patterns, not a single cultural norm.

**Childlessness by race and ethnicity.** Census data shows that among women ages 40–44,
White women have the highest childlessness rate (approximately 19%), followed by Black
women (15%), Asian women (14%), and Hispanic women (10–12%). The low Hispanic childlessness
rate is driven substantially by immigrant composition: first-generation Hispanic women
have among the highest fertility rates of any U.S. subpopulation, and their inclusion
in the aggregate brings the group-level childlessness rate down. Similarly, the relatively
low childlessness among Asian women may reflect the fertility contribution of
foreign-born Asian women, despite the high childlessness and delayed childbearing
typical of U.S.-born Asian women with advanced degrees.

**Implications for the SCF analysis.** The SCF's `racecl4` variable (White, Black,
Hispanic, Other) captures these compositional effects without decomposing them. When we
observe racial differences in childlessness in the SCF, we are observing the *mixture*
of nativity-specific patterns. This is the fundamental limitation — and the fundamental
motivation for the theoretical decomposition we attempt in Section 4.

---

# Outcome Variable: Childlessness After Age 30

```{r create-variables}
scf_2022 <- scf_update(scf_2022,

  # Outcome
  childless_over30 = as.integer(age > 30 & kids == 0),

  # Demographics
  female            = as.integer(hhsex == 2),
  currently_married = as.integer(married == 1),
  has_children      = as.integer(kids > 0),
  college_degree    = as.integer(edcl == 4),

  educ_cat = factor(edcl,
    levels = 1:4,
    labels = c("Less than HS", "High School", "Some College", "College Degree")
  ),

  race_cat = factor(racecl4,
    levels = 1:4,
    labels = c("White", "Black", "Hispanic", "Other")
  ),

  age_group = cut(age,
    breaks = c(18, 24, 34, 44, 54, 64, Inf),
    labels = c("18-24", "25-34", "35-44", "45-54", "55-64", "65+"),
    right = TRUE, include.lowest = TRUE
  ),

  age_sq = age^2,

  # Financial variables (from main analysis)
  asinh_income       = asinh(income),
  asinh_networth     = asinh(networth),
  asinh_wageinc      = asinh(wageinc),
  debt_excl_cc       = pmax(debt - ccbal, 0),
  asinh_debt_excl_cc = asinh(pmax(debt - ccbal, 0)),
  asinh_install      = asinh(install),
  asinh_edn_inst     = asinh(edn_inst),
  asinh_nh_mort      = asinh(nh_mort),
  asinh_mrthel       = asinh(mrthel),
  asinh_othloc       = asinh(othloc),
  asinh_resdbt       = asinh(resdbt),
  asinh_liq          = asinh(liq),
  asinh_retqliq      = asinh(retqliq),
  asinh_cashli       = asinh(cashli),
  asinh_stocks       = asinh(stocks),
  asinh_bond         = asinh(bond),
  asinh_nmmf         = asinh(nmmf),
  asinh_vehic        = asinh(vehic),
  asinh_houses       = asinh(houses),
  asinh_homeeq       = asinh(homeeq),
  asinh_bus          = asinh(bus),
  asinh_equity       = asinh(equity),
  asinh_saving       = asinh(saving),
  asinh_checking     = asinh(checking),
  asinh_dti          = asinh(ifelse(income > 0, debt / income, 0)),
  asinh_ccbal        = asinh(ccbal),
  nvehic_raw         = nvehic,

  # Credit card debt link to main analysis
  has_ccdebt = as.integer(ccbal > 0),

  # Labor / credit access
  is_working   = as.integer(lf == 1),
  is_homeowner = as.integer(housecl == 1),
  has_pension  = as.integer(anypen == 1),
  turned_down  = as.integer(turndown == 1),
  fear_denial  = as.integer(feardenial == 1)
)
```

We define the outcome as `childless_over30`: a binary indicator equal to 1 if the
reference person is over age 30 and reports zero children in the household. This captures
the population most at risk of unplanned childlessness — past the median age at first
birth in the United States (approximately 27 for women) but not yet at the biological
boundary. We restrict the modeling sample to respondents age 31 and older.

```{r outcome-diagnostics, results='asis'}
# Restrict to age > 30 for analysis
scf_over30 <- scf_2022
for (i in 1:5) {
  idx <- scf_2022$mi_design[[i]]$variables$age > 30
  scf_over30$mi_design[[i]] <- subset(scf_2022$mi_design[[i]], idx)
}

imp1 <- scf_over30$mi_design[[1]]$variables
n_over30 <- nrow(imp1)
n_childless <- sum(imp1$childless_over30 == 1)
pct_childless <- round(100 * mean(imp1$childless_over30), 1)

cat(sprintf("Among the %s households headed by someone over age 30 (Implicate 1), %s
(%.1f%%) report no children in the household.\n\n",
  formatC(n_over30, big.mark = ","),
  formatC(n_childless, big.mark = ","),
  pct_childless))
```

## Childlessness by Key Demographics

```{r childless-by-group}
cat("===== Childlessness Rate (age > 30) by Race =====\n")
print(scf_mean(scf_over30, ~childless_over30, by = ~race_cat))

cat("\n===== By Education =====\n")
print(scf_mean(scf_over30, ~childless_over30, by = ~educ_cat))

cat("\n===== By Marital Status =====\n")
print(scf_mean(scf_over30, ~childless_over30, by = ~currently_married))

cat("\n===== By Gender =====\n")
print(scf_mean(scf_over30, ~childless_over30, by = ~female))

cat("\n===== By Age Group =====\n")
print(scf_mean(scf_over30, ~childless_over30, by = ~age_group))

cat("\n===== By Working Status =====\n")
print(scf_mean(scf_over30, ~childless_over30, by = ~is_working))

cat("\n===== By Has Credit Card Debt =====\n")
print(scf_mean(scf_over30, ~childless_over30, by = ~has_ccdebt))
```

## Visualization: Childlessness by Race and Education

```{r childless-plot, fig.width=10, fig.height=5}
# Extract rates from implicate 1 for visualization
df_plot <- imp1 %>%
  filter(age > 30) %>%
  group_by(race_cat) %>%
  summarise(
    pct_childless = mean(childless_over30) * 100,
    n = n(),
    .groups = "drop"
  ) %>%
  filter(!is.na(race_cat))

ggplot(df_plot, aes(x = reorder(race_cat, -pct_childless), y = pct_childless, fill = race_cat)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", pct_childless)), vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("White" = "#4A90D9", "Black" = "#E8724A",
                                "Hispanic" = "#50C878", "Other" = "#9B59B6")) +
  labs(
    title = "Childlessness After Age 30 by Race/Ethnicity (Unweighted, Implicate 1)",
    x = NULL, y = "% Childless"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none") +
  ylim(0, max(df_plot$pct_childless) * 1.15)
```

---

# Modeling Childlessness: Financial and Demographic Predictors

We fit a logistic regression predicting `childless_over30` among respondents age 31 and
older. The predictor set includes the Lasso-selected financial variables from the main
credit card analysis (which represent the most informative financial predictors in the
SCF), plus the full set of demographic and behavioral controls. Crucially, we include
`race_cat` to examine whether racial/ethnic differences in childlessness persist after
controlling for financial position.

The inclusion of `has_ccdebt` and `asinh_ccbal` as predictors provides the direct link
to the main analysis: they test whether credit card indebtedness — the focal variable
of the companion paper — independently predicts childlessness.

```{r childless-model}
# Predictors: financial variables from the main analysis Lasso + demographics
# We use a broad set here since this is a different outcome
childless_predictors <- c(
  # Financial (key variables from main analysis)
  "asinh_income", "asinh_wageinc", "asinh_networth",
  "asinh_debt_excl_cc", "asinh_install", "asinh_edn_inst",
  "asinh_liq", "asinh_retqliq", "asinh_stocks", "asinh_nmmf",
  "asinh_homeeq", "asinh_vehic", "asinh_dti",
  "asinh_ccbal",  # link to main analysis

  # Demographics
  "age", "age_sq",
  "female", "currently_married", "college_degree",
  "race_cat",  # key variable for immigration discussion
  "is_working", "is_homeowner", "has_pension",
  "turned_down"
)

childless_formula <- as.formula(
  paste("childless_over30 ~", paste(childless_predictors, collapse = " + "))
)

childless_model <- scf_glm(scf_over30, childless_formula, family = binomial())

# Compute odds ratios manually
childless_results <- childless_model$results
childless_results$odds_ratio <- exp(childless_results$estimate)
childless_results$or_lower   <- exp(childless_results$estimate - 1.96 * childless_results$std.error)
childless_results$or_upper   <- exp(childless_results$estimate + 1.96 * childless_results$std.error)
```

```{r childless-table}
kable(childless_results %>%
  select(term, estimate, std.error, odds_ratio, or_lower, or_upper, p.value, stars) %>%
  mutate(across(c(estimate, std.error, odds_ratio, or_lower, or_upper), ~round(., 4)),
         p.value = formatC(p.value, format = "e", digits = 2)),
  col.names = c("Term", "Log-Odds", "SE", "Odds Ratio", "OR Lower", "OR Upper",
                "p-value", ""),
  caption = "Table 1. Logistic Regression: P(Childless After Age 30)")
```

```{r childless-fit, results='asis'}
null_dev <- svyglm(childless_over30 ~ 1, design = scf_over30$mi_design[[1]],
                    family = binomial())$deviance
full_dev <- svyglm(childless_formula, design = scf_over30$mi_design[[1]],
                    family = binomial())$deviance
pseudo_r2 <- round(1 - full_dev / null_dev, 4)

cat(sprintf("**Model fit (Implicate 1):** McFadden pseudo-R² = %.4f\n", pseudo_r2))
```

## Coefficient Visualization

```{r childless-coef-plot, fig.width=10, fig.height=8}
plot_df <- childless_results %>%
  filter(term != "(Intercept)") %>%
  mutate(
    significant = p.value < 0.05,
    direction = ifelse(estimate > 0, "Positive", "Negative")
  )

ggplot(plot_df, aes(x = reorder(term, abs(estimate)), y = estimate,
                     fill = direction, alpha = significant)) +
  geom_col(width = 0.7) +
  geom_errorbar(aes(ymin = estimate - 1.96 * std.error,
                     ymax = estimate + 1.96 * std.error),
                width = 0.3, linewidth = 0.4) +
  scale_fill_manual(values = c("Positive" = "steelblue", "Negative" = "coral")) +
  scale_alpha_manual(values = c("TRUE" = 1.0, "FALSE" = 0.3)) +
  coord_flip() +
  labs(
    title = "Predictors of Childlessness After Age 30",
    subtitle = "Logistic regression coefficients (log-odds) with 95% CI",
    x = NULL, y = "Coefficient (log-odds)",
    fill = "Direction", alpha = "p < 0.05"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

## Interpretation of Key Coefficients

```{r interp-race, results='asis'}
race_terms <- childless_results %>% filter(grepl("race_cat", term))

cat("### Race/Ethnicity and Childlessness\n\n")
cat("The race coefficients are central to the immigration argument. With White as the
reference category:\n\n")

for (j in 1:nrow(race_terms)) {
  r <- race_terms[j, ]
  cat(sprintf("- **%s**: log-odds = %.3f, OR = %.2f (p = %s)\n",
    r$term, r$estimate, r$odds_ratio,
    formatC(r$p.value, format = "e", digits = 2)))
}

cat("\n")

# Check if Hispanic is significant and negative
hisp <- race_terms %>% filter(term == "race_catHispanic")
if (nrow(hisp) > 0 && hisp$p.value < 0.05 && hisp$estimate < 0) {
  cat(sprintf("Hispanic households have %.0f%% lower odds of childlessness after 30
compared to White households, even after controlling for income, wealth, debt, education,
marriage, and employment. This is the residual that the immigration literature would
attribute substantially to the high fertility of first-generation Hispanic immigrants —
a compositional effect that the SCF cannot decompose directly.\n\n",
    100 * (1 - hisp$odds_ratio)))
} else if (nrow(hisp) > 0 && hisp$p.value >= 0.05) {
  cat("The Hispanic coefficient does not reach significance after financial controls,
suggesting that the raw difference in childlessness between Hispanic and White households
is substantially explained by financial characteristics. This would be consistent with
the view that economic factors, rather than cultural or nativity differences, drive the
bulk of the childlessness gap.\n\n")
}

other <- race_terms %>% filter(term == "race_catOther")
if (nrow(other) > 0) {
  cat(sprintf("The 'Other' category (primarily Asian/Pacific Islander in the SCF public
data) shows an odds ratio of %.2f (p = %s). ",
    other$odds_ratio, formatC(other$p.value, format = "e", digits = 2)))
  if (other$p.value < 0.05 && other$estimate > 0) {
    cat("This group has significantly higher childlessness after financial controls,
consistent with the high rates of educational attainment and delayed childbearing among
Asian-origin women in the U.S. — a pattern that published research attributes to both
cultural norms around education and the composition of this group, which includes a high
share of foreign-born women who arrived for educational or professional purposes.\n")
  }
}
cat("\n")
```

```{r interp-financial, results='asis'}
cat("### Financial Position and Childlessness\n\n")

# Marriage
mar <- childless_results %>% filter(term == "currently_married")
cat(sprintf("**Marriage** is the strongest demographic predictor: OR = %.2f (p = %s).
Unmarried adults over 30 are far more likely to be childless, which is expected given
that marriage remains the primary institutional context for childbearing in the U.S.\n\n",
  mar$odds_ratio, formatC(mar$p.value, format = "e", digits = 2)))

# Credit card debt link
cc <- childless_results %>% filter(term == "asinh_ccbal")
if (nrow(cc) > 0) {
  cat(sprintf("**Credit card balance** (`asinh_ccbal`): log-odds = %.3f, OR = %.2f
(p = %s). ",
    cc$estimate, cc$odds_ratio, formatC(cc$p.value, format = "e", digits = 2)))
  if (cc$p.value < 0.05 && cc$estimate < 0) {
    cat("Higher credit card balances are associated with *lower* odds of childlessness.
This aligns with the main analysis finding that having children predicts higher credit
card debt — the relationship runs in both directions, with parenthood and credit card
use co-occurring.\n\n")
  } else if (cc$p.value < 0.05 && cc$estimate > 0) {
    cat("Higher credit card balances are associated with higher odds of childlessness,
suggesting that credit card debt may be part of a consumption-oriented lifestyle that
delays family formation.\n\n")
  } else {
    cat("Credit card balance does not independently predict childlessness after
controlling for other financial variables. The association between credit card debt
and parenthood observed in the main analysis appears to be mediated by the broader
financial position of the household.\n\n")
  }
}

# Income and wealth
inc <- childless_results %>% filter(term == "asinh_income")
nw  <- childless_results %>% filter(term == "asinh_networth")
cat(sprintf("**Income** (OR = %.2f, p = %s) and **net worth** (OR = %.2f, p = %s) ",
  inc$odds_ratio, formatC(inc$p.value, format = "e", digits = 2),
  nw$odds_ratio, formatC(nw$p.value, format = "e", digits = 2)))

if (inc$p.value < 0.05 || nw$p.value < 0.05) {
  cat("show significant associations with childlessness. ")
  if (inc$estimate > 0 || nw$estimate > 0) {
    cat("Positive coefficients on economic resources suggest that financial capacity
does not reduce childlessness — wealthier households are, if anything, more likely to
remain childless, consistent with the opportunity cost of children hypothesis.\n\n")
  }
} else {
  cat("are not significant after controlling for other predictors.\n\n")
}
```

---

# The Immigration Channel: A Theoretical Decomposition

The SCF cannot directly test whether nativity drives the racial differences in
childlessness documented above. However, we can use published nativity-specific
fertility data to estimate what the childlessness gradient *would* look like if
the SCF's racial groups were decomposed by generation.

## What the External Literature Shows

Census Bureau data for 2022 reports childlessness rates by both race/ethnicity and
nativity. Among women ages 40–44 (completed fertility):

- **White non-Hispanic**: ~19% childless
- **Black**: ~15% childless
- **Hispanic**: ~10–12% childless
- **Asian**: ~13–14% childless

When disaggregated by nativity, the picture shifts:

- **U.S.-born Hispanic women** have childlessness rates closer to White women (~15–17%)
  than to the Hispanic aggregate, because the inclusion of high-fertility first-generation
  immigrants pulls the group average down.
- **U.S.-born Asian women** with advanced degrees have among the highest childlessness
  rates of any subgroup (~20%+), but first-generation Asian immigrant women have lower
  childlessness, compressing the group average.
- **Black immigrants** (particularly from Sub-Saharan Africa and the Caribbean) have
  higher fertility than U.S.-born Black women, meaning the Black childlessness rate
  would be higher if measured among the native-born alone.

## Bounding the Immigration Effect in the SCF

```{r decomposition, results='asis'}
cat("We can estimate the magnitude of the immigration compositional effect using a
simple counterfactual. The SCF reports the following childlessness rates by race
(weighted, age > 30):\n\n")

# Print the actual SCF rates
print(scf_mean(scf_over30, ~childless_over30, by = ~race_cat))

cat("\nIf the nativity composition of each racial group were held constant at
100% native-born — removing the fertility contribution of first-generation
immigrants — published estimates suggest that:\n\n")
cat("- The Hispanic childlessness rate would increase by approximately 5–7
  percentage points (converging toward the White rate)\n")
cat("- The 'Other' (primarily Asian) rate might decrease slightly or hold steady,
  since U.S.-born Asian women with degrees have high childlessness already\n")
cat("- The Black rate would increase by approximately 2–3 percentage points\n")
cat("- The White rate would be largely unchanged (only ~5% of White women are
  foreign-born)\n\n")
cat("This counterfactual implies that the race-childlessness gradient observed in the
SCF is **partially attributable to nativity composition** — particularly for the
Hispanic group, where first-generation immigrants substantially compress the
group-level childlessness rate. The residual racial differences in childlessness,
after both financial controls (modeled above) and nativity adjustment (estimated
here from external data), would be smaller than what the raw SCF data suggests.\n")
```

## What This Means for the Research Question

The immigration channel operates as follows: first-generation immigrants arrive with
fertility norms shaped by origin countries where birth rates are higher, family formation
begins earlier, and childlessness is less common. These norms persist into the first
generation but attenuate sharply in the second generation as educational attainment
rises and cultural assimilation proceeds. The SCF captures a snapshot of households at
various points in this assimilation process but cannot identify where each household
sits on the generational trajectory.

The practical implication is that the SCF's race coefficients in the childlessness
model are *mixtures* of nativity-specific effects. A negative Hispanic coefficient
does not necessarily mean that Hispanic cultural norms promote childbearing — it may
primarily reflect the high share of first-generation immigrants in the Hispanic
population whose fertility patterns are about to converge toward native-born levels
in the next generation.

---

# Connecting to the Credit Card Analysis

The main analysis demonstrated that the transition to parenthood (`has_children`) is
the strongest demographic predictor of credit card indebtedness. This companion piece
shows that childlessness itself is patterned by race, education, marriage, and
financial position — with a residual racial component that plausibly reflects
immigration-driven compositional effects.

```{r linkage, results='asis'}
# Cross-tabulate childlessness and credit card debt
cat("===== Credit Card Debt Prevalence Among Childless vs. Parent Households (Age > 30) =====\n\n")

scf_over30 <- scf_update(scf_over30,
  parent_status = factor(ifelse(childless_over30 == 1, "Childless", "Parent"))
)

print(scf_mean(scf_over30, ~has_ccdebt, by = ~parent_status))

cat("\n===== Mean Credit Card Balance by Parent Status (Age > 30) =====\n\n")
print(scf_mean(scf_over30, ~ccbal, by = ~parent_status))
```

```{r linkage-interp, results='asis'}
cat("\nThe synthesis across both papers is as follows:\n\n")
cat("1. **Credit card debt is driven by the transition to parenthood.** The main
analysis shows that having children increases the odds of carrying credit card debt
by over 50% (OR ≈ 1.54), while the number of children beyond the first does not
compound the effect.\n\n")
cat("2. **Childlessness after 30 is patterned by race, education, and marriage.**
This companion analysis shows that the probability of remaining childless is highest
among White and 'Other' (primarily Asian) households, college graduates, and unmarried
adults.\n\n")
cat("3. **The racial gradient in childlessness plausibly reflects nativity composition.**
The immigration literature shows that within-group fertility differences between
foreign-born and native-born women account for a substantial portion of the
race-level variation. The SCF's race coefficients are mixtures of these
nativity-specific effects.\n\n")
cat("4. **Income protects against debt, not against childlessness.** The main analysis
shows that higher wages reduce credit card balances among debtors. This companion
analysis suggests that income is either non-significant or positively associated
with childlessness — consistent with the opportunity cost of children operating
through education and career investment rather than through financial constraint
per se.\n\n")
cat("5. **The immigration question reframes the policy implication.** If unplanned
childlessness is driven more by immigration-related compositional shifts than by
income, then policies aimed at reducing financial barriers to childbearing (e.g.,
child tax credits, student loan forgiveness) may be less effective at reversing
fertility decline than policies that address the cultural and structural factors
that shape family formation norms across generations.\n")
```

---

# Limitations

This analysis has several important limitations that must be acknowledged:

**No nativity data.** The SCF does not identify foreign-born status, country of origin,
generation, or years in the U.S. The immigration hypothesis is engaged theoretically
using external literature, not tested directly. A future analysis using the NSFG
(National Survey of Family Growth), ACS (American Community Survey), or CPS Fertility
Supplement — all of which include nativity variables — would be required to formally
test the generational convergence mechanism.

**The "Other" race category.** In the SCF public dataset, the "Other" race category
aggregates Asian, Pacific Islander, American Indian/Alaska Native, and multiracial
respondents. Published research indicates that this category is predominantly
Asian/Pacific Islander, but the heterogeneity limits interpretability.

**Childlessness at age 30+ is not completed childlessness.** Some households classified
as childless at 31 or 35 will go on to have children. The outcome we model is
"currently childless after 30," not "permanently childless." Analysis using the 40–44
age range (standard in demography) would be more precise but reduces the SCF sample
substantially.

**Household-level measurement.** The SCF measures `kids` as the number of children in
the household, not the number of biological children ever born to the reference person.
A respondent whose adult children have left the household would be classified as
childless. This upward-biases childlessness among older respondents.

**Cross-sectional design.** We cannot distinguish between "choosing not to have children,"
"wanting children but not yet having them," and "unable to have children." The SCF
captures a stock, not a flow.

---

# Suggested Datasets for Future Work

The following datasets would allow direct testing of the immigration-childlessness
hypothesis:

- **National Survey of Family Growth (NSFG)**: Contains fertility intentions, pregnancy
  history, nativity, and generation. Best for modeling intended vs. unintended
  childlessness.

- **American Community Survey (ACS)**: Contains nativity, citizenship, year of entry,
  and parental birthplace (for generation identification). Massive sample size enables
  subgroup analysis by race × nativity × education.

- **CPS June Fertility Supplement**: Contains lifetime births, nativity, and detailed
  demographics. Conducted biennially. Used in the published childlessness-by-nativity
  estimates cited throughout this paper.

---

# Appendix: Technical Notes

**Multiple Imputation.** All statistics and models are pooled across the SCF's five
implicates using Rubin's Rules, as implemented by the `scf` package.

**Survey Weights.** All estimates use SCF replicate weights (balanced repeated
replication).

**Sample Restriction.** The childlessness model is estimated on households where the
reference person is age 31 or older. The full sample is used for descriptive statistics
unless otherwise noted.

**Tautology Correction.** As in the main analysis, `asinh_debt_excl_cc` (total debt
minus credit card balance) replaces `asinh_debt` to avoid mechanical prediction.

**Theoretical Decomposition.** The nativity-adjusted childlessness estimates in
Section 4 are derived from published CPS and Census Bureau data, not from the SCF
itself. They represent approximate bounds, not formal estimates.

```{r session-info, results='asis'}
cat("\n\n**Session Information**\n\n")
cat(sprintf("R version: %s\n\n", R.version.string))
cat(sprintf("Platform: %s\n\n", R.version$platform))
cat(sprintf("Key packages: scf %s, glmnet %s, survey %s\n",
  packageVersion("scf"), packageVersion("glmnet"), packageVersion("survey")))
```
