---
title: "Segmented Assimilation in the American Financial System: Origin-Based Gradients in Immigrant Household Finances"
subtitle: "Evidence from the 2022 Survey of Consumer Finances"
author: "Delroy Nichols"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 5, dpi = 150
)
rm(list = ls())
gc()
set.seed(42)
```

# Origin-Based Gradients in Immigrant Household Finances

## Abstract

Two theories make different predictions about the financial outcomes of foreign-born
households in the U.S. The human-capital convergence model holds that, once
characteristics such as age, education, income, and time since arrival are controlled
for, the origin region has no independent effect. The segmented-assimilation framework
predicts that the origin region directs households into incorporation pathways whose
effects persist beyond human capital and duration of residence. Until 2022, this could
not be tested directly because no wealth survey tracked national origin.

This paper uses new 2022 Survey of Consumer Finances variables on respondent national
origin (X6915) and years in the United States (X6906) to test competing predictions
across seven financial outcomes spanning three theoretical pathways. Previewing the
primary findings, logistic and linear regressions on a foreign-born sample (n=1,103)
do not reject the human-capital convergence null: origin-region coefficients are small,
statistically indistinguishable from zero, and unaffected by human-capital controls.
A duration-stratified robustness check splits the sample into recent arrivals
(≤15 years) and long-tenured immigrants (>15 years) to test for segmentation effects
in early-stage incorporation and possible convergence over time. Collectively, these
results provide limited support for the segmented-assimilation framework and instead
indicate that differences by origin region are minimal after controlling for human
capital, clarifying when and where segmented-assimilation theory applies to U.S.
wealth outcomes.

**Data Source:** 2022 Survey of Consumer Finances (SCF), Federal Reserve Board. All
estimates are survey-weighted; all models are pooled across the SCF's five multiply-
imputed datasets using Rubin's Rules.

**Note on Scope:** All findings are correlational and limited to the foreign-born
respondent sample. The SCF's cross-sectional design and absence of parental nativity
preclude causal claims and intergenerational analysis.

## The Central Question

Two major theories explain immigrant economic outcomes in the U.S. The human-capital
convergence model, as articulated by Borjas (1987, 1999), contends that factors such
as education, English proficiency, age at arrival, and time in the U.S. sufficiently
account for economic outcomes, leaving the origin region with no independent effect
once these factors are controlled for. Foundational empirical studies by Borjas (1987),
Chiswick (1978), and Blau and Graham (1990) provide evidence that, after adjusting for
relevant human-capital characteristics and duration of residence, differences in
immigrant economic integration by origin region largely disappear.

The **segmented-assimilation framework**, developed by Alejandro Portes, Rubén Rumbaut,
and Min Zhou, makes the opposite prediction. Three pathways are identified: *parallel
integration* into the white middle class through educational achievement, *downward
assimilation* into the racialized urban underclass through structural exclusion, and
*selective assimilation* through retention of ethnic-community ties and resources
(Portes & Zhou, 1993; Portes & Rumbaut, 2001). The origin region itself, net of human
capital, channels households into these pathways because reception contexts—including
race, neighborhood, and labor-market structure—differ systematically across origin
groups (Zhou, 1997). For example, a Mexican-origin household with a college degree may
face different access to mainstream finance than an Asian-origin household with
identical credentials.

Until 2022, this debate could not be tested directly in U.S. wealth data because no
nationally representative wealth survey collected respondents' national origin. Earlier
SCF waves identified foreign-born respondents only through indirect indicators and
provided no origin-region detail. The 2022 SCF wave added respondent and spouse
national origin (X6915, X8055-X8057) and years lived in the United States (X6906,
X7796), the first such measures in any nationally representative U.S. wealth survey.

This paper directly tests whether foreign-born household financial outcomes in the U.S.
differ systematically by origin region, after accounting for human capital and years
in the country. A finding of large, statistically robust origin-region effects would
support segmented assimilation theory, while null or near-null effects would support
human-capital convergence. The SCF data now provide a direct means of distinguishing
these competing models, making this test both timely and necessary.

## Why This Matters: Beyond Race-Only Frameworks

Wealth and credit access disparities are generally studied by race rather than by
origin. Studies comparing Black and Hispanic groups often treat these populations as
homogeneous. However, prior research has shown that distinct groups such as Black
Caribbeans, Mexican-origin Hispanics, and Cuban-origin Hispanics may differ
financially due to unique migration trajectories and community resources (e.g., Logan,
2007; Smith, 2015; Telles & Ortiz, 2008). Whether these differences appear in 2022
nationally representative data is the empirical question this paper examines.

The 2022 SCF adds origin-region detail, allowing the field to test if
segmented-assimilation pathways (parallel integration, downward assimilation,
selective assimilation) appear in wealth and credit data with enough magnitude and
statistical precision to distinguish origin groups. If so, interventions for
"Hispanic" or "Asian" households would need to be redesigned to address within-group
heterogeneity. If not, existing race-based frameworks, simpler, reliably measured, and
widely supported, remain the best analytical lens for U.S. wealth disparities, even
with origin-region detail.

The purpose of this paper is to address the longstanding debate over the determinants
of immigrant financial outcomes by directly comparing two leading theoretical
models—human-capital convergence and segmented assimilation—using the first
nationally representative U.S. wealth data that includes national origin. By
leveraging new variables from the 2022 Survey of Consumer Finances, this study seeks
to clarify whether origin region exerts an independent effect on foreign-born
household finances, or if observed differences are fully explained by human-capital
characteristics and duration of U.S. residence. Rather than privileging either
theoretical perspective, the analysis is designed to provide an unbiased empirical
test, thereby clarifying which model is more consistent with the observed data.

---

```{r load-packages}
required_packages <- c(
  "tidyverse", "haven", "survey", "labelled",
  "ggplot2", "scales", "scf",
  "broom", "knitr", "moments"
)

for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg, repos = "http://cran.us.r-project.org")
    library(pkg, character.only = TRUE)
  }
}
```

```{r download-data}
# Same download pattern as the credit-card-debt analysis (Paper 1).
# scf_download writes .rds files to the current working directory.

dir.create("scf_data", showWarnings = FALSE, recursive = TRUE)
original_wd <- getwd()

# Move into the data folder
setwd("scf_data")

library(haven)

# Read from your local absolute path
raw_macro_df <- read_dta("C:/Users/delro/OneDrive/Documents/scf2022s/p22i6.dta")

# Save directly to the current working directory (which is already scf_data)
saveRDS(raw_macro_df, "p22i6.rds") 

# Continue with the download check for the bulletin extract
if (!file.exists("scf2022.rds")) {
  tryCatch({
    scf_download(2022)
  }, error = function(e) {
    cat("Error downloading 2022 data:", conditionMessage(e), "\n")
  })
}

# Return to the root project directory
setwd(original_wd)
```

```{r load-data}
scf_2022 <- scf_load(2022, data_directory = "scf_data")
```

# Variable Availability and Diagnostic

Before constructing variables, we check which of the variables having direct exposure without segmented assimilation theory are
actually available in the loaded `scf` package object. The standard SCF Bulletin extract
that the `scf` package builds does not by default include the new national-origin or
years-in-U.S. items (`X6906`, `X7404`, `X6915`); these are raw codebook items rather than
constructed Bulletin variables. If they are missing, we read the raw 2022 SCF `.rds`
file directly and merge them in by household ID.

```{r diagnose-immigration-vars, results='asis'}
imp1 <- scf_2022$mi_design[[1]]$variables

# Variables we need
# X6915 = respondent national origin (the variable for origin region)
# X6906 = years respondent has lived in the U.S. (the YIU variable)
# X7796 = years spouse has lived in the U.S. (auxiliary)
# X7404 = spouse Hispanic/Latino self-id (auxiliary)
# X8049 = spouse race (used in swap correction)
# X8000 = sex/role swap indicator
#
# CORRECTED 2026-04-29: V5 incorrectly used X7437-X7443 as years-in-US.
# Per the official SCF 2022 changes file, X7437-X7443 are MORTGAGE FORBEARANCE
# variables, not immigration variables. The true years-in-US variable is X6906
# for the respondent (and X7796 for the spouse).
imm_vars <- c("X6915", "X6906", "X7796", "X7404", "X8000", "X8049")

present_in_extract <- imm_vars %in% names(imp1)

cat("**Immigration-variable availability in the scf package extract:**\n\n")
diag_df <- data.frame(
  Variable = imm_vars,
  `In extract` = ifelse(present_in_extract, "Yes", "No"),
  check.names = FALSE
)
print(kable(diag_df))
cat("\n\n")

need_raw_merge <- !all(present_in_extract)

if (need_raw_merge) {
  cat("Some required immigration variables are missing from the bulletin extract. ",
      "Attempting to merge them in from the raw 2022 SCF data file...\n\n", sep = "")
} else {
  cat("All required immigration variables are present in the extract. ",
      "No raw merge needed.\n\n", sep = "")
}
```

```{r merge-raw-immigration-vars}
# ============================================================================
# Merge raw immigration variables from the underlying SCF .rds file
# ============================================================================
# The scf package downloads the raw 2022 SCF data into scf_data/ as well as
# building the bulletin extract. The raw file contains all X-variables; we
# read it, select just the immigration-related vars + ID, and merge into
# each implicate's $variables data frame.
#
# Common file names produced by scf_download():
#   - p22i6.rds (raw public-use, 5 implicates stacked, with implicate id)
#   - scf2022.rds (the bulletin extract)
#
# We try a few candidate filenames defensively.

if (need_raw_merge) {

  raw_candidates <- c(
    file.path("scf_data", "p22i6.rds"),
    file.path("scf_data", "scfp2022.rds"),
    file.path("scf_data", "p22.rds"),
    "p22i6.rds",
    "scfp2022.rds"
  )

  raw_path <- raw_candidates[file.exists(raw_candidates)][1]

  if (!is.na(raw_path)) {
    raw_data <- readRDS(raw_path)

    # Standardize variable name casing — the raw file may have either
    # uppercase or lowercase. We try uppercase first since X-variables
    # are conventionally uppercase.
    raw_path <- raw_candidates[file.exists(raw_candidates)][1]

  if (!is.na(raw_path)) {
    raw_data <- readRDS(raw_path)

    # FIX: Force all raw data columns to uppercase immediately.
    # This aligns the Stata output with the canonical SCF uppercase format.
    names(raw_data) <- toupper(names(raw_data))

    raw_names <- names(raw_data)
    
    # The name_map is now technically redundant, but safe to keep
    name_map <- setNames(raw_names, toupper(raw_names))

    available_imm <- imm_vars[toupper(imm_vars) %in% toupper(raw_names)]

    # The raw data has 5 implicates stacked. We need an ID that matches
    # the bulletin extract's row ordering. Common candidates: y1, yy1, x1.
    id_candidates <- c("Y1", "YY1", "X1") # Updated to only look for uppercase
    id_var <- id_candidates[id_candidates %in% raw_names][1]

    if (!is.na(id_var) && length(available_imm) > 0) {

      # Pull the matching columns
      raw_imm <- raw_data[, c(id_var, available_imm), drop = FALSE]

      # Determine implicate count: raw file has each household 5 times (one
      # per implicate) — split into 5 chunks of equal size in row order.
      n_total <- nrow(raw_imm)
      n_per_imp <- nrow(scf_2022$mi_design[[1]]$variables)

      if (n_total == 5 * n_per_imp) {
        # Standard layout: implicates stacked
        for (i in 1:5) {
          start_row <- (i - 1) * n_per_imp + 1
          end_row   <- i * n_per_imp
          imm_chunk <- raw_imm[start_row:end_row, available_imm, drop = FALSE]

          # Add to the implicate's $variables
          for (v in available_imm) {
            scf_2022$mi_design[[i]]$variables[[v]] <- imm_chunk[[v]]
          }
        }
        cat("Successfully merged ", length(available_imm), " immigration variables ",
            "from ", raw_path, " into all 5 implicates.\n", sep = "")
      } else {
        cat("Row-count mismatch: raw file has ", n_total,
            " rows, bulletin extract has ", n_per_imp,
            " rows per implicate. Merge aborted; using available variables only.\n",
            sep = "")
      }
    } else {
      cat("Could not locate raw SCF file or matching ID variable. ",
          "Immigration variables will be unavailable.\n", sep = "")
    }
  } else {
    cat("Raw 2022 SCF .rds file not found in scf_data/. ",
        "Available files:\n", sep = "")
    cat(paste(list.files("scf_data"), collapse = "\n"), "\n\n")
  }
}

# Re-check what's now available
imp1 <- scf_2022$mi_design[[1]]$variables
have_X7404 <- "X7404" %in% names(imp1)
have_yiu   <- "X6906" %in% names(imp1)
have_X8000 <- "X8000" %in% names(imp1)
have_X8049 <- "X8049" %in% names(imp1)
```

```{r raw-x7404-frequency, results='asis'}
# ============================================================================
# Print raw X7404 value frequencies — needed to calibrate the recode
# ============================================================================
# Until we see what values X7404 actually takes in this wave, we cannot
# write the correct recoding rule. This chunk dumps the unweighted frequency
# table so we can adjust the recode.

if (have_X7404) {
  # 1. Extract the column and force it to a standard numeric vector
  # This strips out any problematic 'haven_labelled' metadata from Stata
  x7404_vec <- as.numeric(scf_2022$mi_design[[1]]$variables$X7404)
  
  # 2. Use Base R's table() to get the counts, which avoids package conflicts
  freq_table <- table(x7404_vec, useNA = "ifany")
  
  # 3. Reconstruct it into a clean data frame for kable
  x7404_freq <- data.frame(
    X7404 = names(freq_table),
    n = as.integer(freq_table)
  )
  
  # 4. Sort descending
  x7404_freq <- x7404_freq[order(-x7404_freq$n), ]

  cat("**Raw X7404 frequency (Implicate 1, unweighted):**\n\n")
  # row.names = FALSE keeps the markdown table perfectly clean
  print(kable(x7404_freq, caption = "X7404 = respondent national origin", row.names = FALSE))
  cat("\n\nThis table guides the origin-region recode in the next chunk. ",
      "Code 0 typically indicates 'inapplicable' (i.e., U.S.-born or non-respondent).\n\n",
      sep = "")
}
```

```{r visuals}

library(ggplot2)
library(dplyr)

# ============================================================================
# Visualize X7404 Frequencies: Spouse Hispanic/Latino Culture/Origin
# ============================================================================

# 1. SCF mapping for X7404
spouse_hisp_labels <- c(
  "0" = "Inapplicable (No Spouse/Partner)",
  "1" = "Yes (Spouse is Hispanic/Latino)",
  "5" = "No (Spouse is not Hispanic/Latino)"
)

# 2. Apply labels and generate plot
x7404_freq %>%
  mutate(
    # Map the numeric codes to our accurate text labels
    Spouse_Label = spouse_hisp_labels[as.character(X7404)],
    # Catch any unexpected codes just in case
    Spouse_Label = ifelse(is.na(Spouse_Label), paste("Code", X7404), Spouse_Label)
  ) %>%
  # Order the bars by frequency 
  ggplot(aes(x = reorder(Spouse_Label, n), y = n)) +
  geom_col(fill = "steelblue", color = "black", alpha = 0.8) +
  coord_flip() + # Horizontal layout for readability
  theme_minimal() +
  labs(
    title = "Spouse/Partner Hispanic or Latino Origin",
    subtitle = "Raw frequencies from Implicate 1 (Variable X7404)",
    x = NULL, 
    y = "Number of Respondents"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.y = element_text(size = 11)
  )

```



```{r plot-x6915, echo=FALSE, warning=FALSE, message=FALSE}
# ============================================================================
# Visualize X6915 Frequencies: Reference Person National Origin
# ============================================================================

library(ggplot2)
library(dplyr)

# FIX: Explicitly define have_X6915 so the Knit session knows what it is!
have_X6915 <- "X6915" %in% names(scf_2022$mi_design[[1]]$variables)

if (have_X6915) {
  
  x6915_vec <- as.numeric(scf_2022$mi_design[[1]]$variables$X6915)
  freq_table_6915 <- table(x6915_vec, useNA = "ifany")
  
  x6915_freq <- data.frame(
    X6915 = names(freq_table_6915),
    n = as.integer(freq_table_6915)
  )
  
  origin_labels <- c(
    "0" = "U.S. Born / Inapplicable",
    "100" = "Western European", 
    "200" = "Eastern European",
    "300" = "European",
    "400" = "West Indies",
    "500" = "Central and South American", 
    "600" = "Middle Easterner",
    "700" = "Sub-Saharan African",
    "800" = "South Asian",
    "900" = "East Asian",
    "1000" = "All Other Asians",
    "1100" = "Pacific Islander",
    "1200" = "American"
  )
  
  p_x6915 <- x6915_freq %>%
    mutate(
      Origin_Label = origin_labels[as.character(X6915)],
      Origin_Label = ifelse(is.na(Origin_Label), paste("Code", X6915), Origin_Label)
    ) %>%
    filter(X6915 != "0") %>% 
    ggplot(aes(x = reorder(Origin_Label, n), y = n)) +
    geom_col(fill = "seagreen", color = "black", alpha = 0.8) +
    coord_flip() + 
    theme_minimal() +
    labs(
      title = "Reference Person National Origin (Immigrant Sub-sample)",
      subtitle = "Unweighted frequencies from Implicate 1 (Variable X6915, excluding U.S. Born)",
      x = NULL, 
      y = "Number of Respondents"
    ) +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      axis.text.y = element_text(size = 11)
    )
  
  # Properly print the plot
  print(p_x6915)
  
} else {
  cat("Variable X6915 not found. Please ensure the merge chunk was re-run.")
}
```

```{r table-x6915, echo=FALSE, results='asis'}
# Print the raw counts table
if (have_X6915) {
  cat("\n\n**Raw X6915 frequency (Implicate 1, unweighted):**\n\n")
  tab <- knitr::kable(x6915_freq %>% arrange(desc(n)), row.names = FALSE)
  cat(paste(tab, collapse = "\n"), "\n\n")
}
```


# Variable Engineering

```{r data-engineering-and-subsetting, warning=FALSE, message=FALSE}
# ============================================================================
# Build all analysis variables and create the analytic subset
# ============================================================================
# CRITICAL DESIGN NOTE on foreign-born identification:
#
# X6915 is an ANCESTRY/HERITAGE variable, asked of all respondents — both
# foreign-born and U.S.-born. A respondent reporting "Western European"
# ancestry (code 100) is overwhelmingly U.S.-born (white American with
# European heritage), not an immigrant from Europe. Likewise, "American"
# ancestry (1200) is U.S.-born by definition, and "Sub-Saharan African" (700)
# captures Black Americans reporting African ancestry alongside actual
# African immigrants.
#
# The actual foreign-born signal is in X6906 (years the reference person
# has lived in the United States), which the SCF only asks of foreign-born
# respondents. A non-zero, non-missing years-in-US value confirms foreign-
# born status. (The companion variable X7796 is years-in-US for the spouse
# but is not used in main models, since the unit of analysis is the
# respondent.)
#
# IMPORTANT NOTE: V5 erroneously used X7437-X7443 as the years-in-US
# variables. Per the official SCF 2022 changes file, those are MORTGAGE
# FORBEARANCE variables (which mortgage terms were adjusted after exiting
# forbearance). The correct YIU variable is X6906. This V6 build uses
# X6906 throughout.
#
# Therefore, foreign_born requires BOTH:
#   (a) X6915 indicates a non-American ancestry (i.e., X6915 != 0 AND
#       X6915 != 1200), AND
#   (b) X6906 > 0 (confirming actual years lived in the U.S. as immigrant)
# ============================================================================

# ---- Step 1: Years in U.S. (X6906 = respondent years in US) ----
# X6906 = 0 means inapplicable per SCF data convention (i.e., U.S.-born).
# X6906 > 0 means the respondent reported N years lived in the U.S.,
# which is asked only of foreign-born respondents.
if (have_yiu) {
  scf_2022 <- scf_update(scf_2022,
    years_in_us  = ifelse(as.numeric(X6906) <= 0, NA_real_, as.numeric(X6906)),
    yiu_missing  = as.integer(is.na(years_in_us)),
    yiu_clean    = ifelse(is.na(years_in_us), 0, years_in_us)
  )
} else {
  scf_2022 <- scf_update(scf_2022,
    years_in_us = NA_real_, yiu_missing = 1L, yiu_clean = 0
  )
}

# ---- Step 2: CORRECTED foreign-born status ----
# Requires both ancestry indicating non-US origin AND positive years in US.
# Fall back gracefully if X6915 is unavailable.
if (have_X6915 && have_yiu) {
  scf_2022 <- scf_update(scf_2022,
    foreign_born = as.integer(
      !is.na(years_in_us) &
      years_in_us > 0 &
      as.numeric(X6915) != 0 &
      as.numeric(X6915) != 1200
    )
  )
} else if (have_yiu) {
  # YIU-only fallback if X6915 is unavailable
  scf_2022 <- scf_update(scf_2022,
    foreign_born = as.integer(!is.na(years_in_us) & years_in_us > 0)
  )
} else {
  scf_2022 <- scf_update(scf_2022, foreign_born = 0L)
}

# ---- Step 3: Origin-region recode from X6915 ----
# Used only within the foreign-born sample (filtered by foreign_born flag).
# Reference category: Latin American (the segmented-assimilation canonical
# case in the literature; X6915 == 500 = Central/South American).
if (have_X6915) {
  scf_2022 <- scf_update(scf_2022,
    origin_region = factor(case_when(
      as.numeric(X6915) == 500 ~ "Latin American",
      as.numeric(X6915) == 400 ~ "Caribbean/West Indies",
      as.numeric(X6915) %in% c(800, 900, 1000) ~ "Asian",
      as.numeric(X6915) %in% c(100, 200, 300) ~ "European/Other developed",
      as.numeric(X6915) %in% c(600, 700, 1100) ~ "Other",
      TRUE ~ NA_character_
    ), levels = c("Latin American",
                  "Caribbean/West Indies",
                  "Asian",
                  "European/Other developed",
                  "Other"))
  )
} else {
  scf_2022 <- scf_update(scf_2022, origin_region = factor(NA_character_))
}

# ---- Step 3b: Spouse/Partner Hispanic Culture/Origin from X7404 ----
if (have_X7404) {
  scf_2022 <- scf_update(scf_2022,
    spouse_hisp_origin = factor(case_when(
      X7404 == 1 ~ "Yes (Spouse Hispanic/Latino)",
      X7404 == 5 ~ "No (Spouse not Hispanic/Latino)",
      X7404 == 0 ~ "Unmarried/Inapplicable",
      TRUE ~ NA_character_
    ), levels = c("Yes (Spouse Hispanic/Latino)",
                  "No (Spouse not Hispanic/Latino)",
                  "Unmarried/Inapplicable"))
  )
} else {
  scf_2022 <- scf_update(scf_2022, spouse_hisp_origin = factor(NA_character_))
}

# ---- Step 4: Race/ethnicity ----
scf_2022 <- scf_update(scf_2022,
  race_cat = factor(racecl4, levels = 1:4,
                    labels = c("White", "Black", "Hispanic", "Other"))
)

# ---- Step 5: Standard demographic controls ----
scf_2022 <- scf_update(scf_2022,
  age_centered = age - 50,
  age_sq       = age^2,
  in_age_range = as.integer(age >= 25 & age <= 75),
  educ_cat = factor(case_when(
    educ < 8 ~ "Less than HS",
    educ == 8 ~ "High School",
    educ %in% 9:11 ~ "Some College",
    educ == 12 ~ "College Degree",
    educ %in% 13:14 ~ "Graduate School/PhD",
    TRUE ~ NA_character_
  ), levels = c("Less than HS", "High School", "Some College",
                "College Degree", "Graduate School/PhD")),
  female            = as.integer(hhsex == 2),
  currently_married = as.integer(married == 1),
  has_children      = as.integer(kids > 0),
  is_homeowner      = as.integer(housecl == 1),
  has_pension       = as.integer(anypen == 1),
  log_income        = log(pmax(1, income)),
  asinh_networth    = asinh(networth),
  log_wageinc       = log(pmax(1, wageinc))
)

# ---- Step 6: Outcomes ----
# CRITICAL: nw_pctl is computed on the FULL SCF distribution (before
# subsetting to foreign-born) so it reflects position in the broader
# U.S. wealth distribution, not relative position within the foreign-born
# subsample. This is computed via scf_update on scf_2022 (the full design)
# so all 5 implicates get a consistent percentile based on the full sample.
scf_2022 <- scf_update(scf_2022,
  own_retire    = as.integer(retqliq > 0),
  own_stocks    = as.integer(stocks > 0),
  credit_denied = as.integer(turndown == 1),
  no_checking   = as.integer(checking <= 0),
  own_home      = as.integer(housecl == 1),
  nw_pctl       = 100 * percent_rank(networth)   # full-sample percentile
)

# ---- Step 7: Analytic sample flag ----
scf_2022 <- scf_update(scf_2022,
  in_analytic_sample = as.integer(
    foreign_born == 1 &
    !is.na(origin_region) &
    in_age_range == 1
  )
)

# ---- Step 8: Create the Analytic Sample Subset ----
# Subset each implicate's survey design to foreign-born respondents.
# nw_pctl values are preserved — they were computed on the full distribution
# in Step 6 above, so subsetting just keeps the foreign-born rows with
# their original full-distribution percentile values intact.
scf_fb <- scf_2022
scf_fb$mi_design <- lapply(scf_2022$mi_design, function(des) {
  subset(des, in_analytic_sample == 1)
})

# Diagnostic: report sample size
n_fb <- nrow(scf_fb$mi_design[[1]]$variables)

# Diagnostic: cross-tab of X6915 ancestry × foreign_born flag — useful for
# verifying that the corrected foreign-born identification is sensible.
fb_diagnostic <- scf_2022$mi_design[[1]]$variables %>%
  filter(!is.na(X6915)) %>%
  group_by(X6915) %>%
  summarise(
    n_total       = n(),
    n_foreign     = sum(foreign_born == 1, na.rm = TRUE),
    pct_foreign   = round(100 * mean(foreign_born == 1, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  arrange(desc(n_total))
```

```{r build-foreign-born-design}
# ============================================================================
# Build the foreign-born subset design — same pattern as scf_debtors in Paper 1
# ============================================================================
# NOTE: foreign_born and years_in_us are already defined in the
# `data-engineering-and-subsetting` chunk above, using the X6915-based
# definition (X6915 != 0 = foreign-born). The X6915 definition is preferred
# over X6906 because X6906 has known small missingness (some foreign-born
# missingness, while X6915 is universally coded.
#
# This chunk just rebuilds the scf_fb subset object using the
# in_analytic_sample flag from the prior chunk.

scf_fb <- scf_2022

scf_fb$mi_design <- lapply(scf_2022$mi_design, function(des) {
  subset(des, in_analytic_sample == 1)
})

# Diagnostic: report sample size
n_fb <- nrow(scf_fb$mi_design[[1]]$variables)
```

# Descriptive Findings

## Foreign-Born Identification Diagnostic

The corrected `foreign_born` definition requires both ancestry indicating non-U.S.
origin (X6915 ≠ 0 and X6915 ≠ 1200) AND positive years in the U.S. (X6906).
The table below shows, for each X6915 ancestry code, what percentage of respondents
end up classified as foreign-born. Codes like 100 (Western European), 200 (Eastern
European), and 1200 (American) should be near 0%; codes like 500 (Latin American),
800/900 (Asian), and 600 (Middle Eastern) should be near 100%.

```{r fb-id-diagnostic, results='asis'}
if (exists("fb_diagnostic") && nrow(fb_diagnostic) > 0) {

  origin_label_map <- c(
    "0"    = "U.S. Born / Inapplicable",
    "100"  = "Western European",
    "200"  = "Eastern European",
    "300"  = "European (Other)",
    "400"  = "West Indies / Caribbean",
    "500"  = "Central / South American",
    "600"  = "Middle Eastern",
    "700"  = "Sub-Saharan African",
    "800"  = "South Asian",
    "900"  = "East Asian",
    "1000" = "All Other Asian",
    "1100" = "Pacific Islander",
    "1200" = "American"
  )

  fb_diagnostic_display <- fb_diagnostic %>%
    mutate(
      Ancestry = origin_label_map[as.character(X6915)],
      Ancestry = ifelse(is.na(Ancestry), paste("Code", X6915), Ancestry)
    ) %>%
    select(X6915, Ancestry, n_total, n_foreign, pct_foreign)

  cat("**Table 0. Foreign-born identification by X6915 ancestry code**\n\n")
  print(kable(fb_diagnostic_display,
              col.names = c("X6915", "Ancestry label", "n total",
                            "n foreign-born", "% foreign-born"),
              caption = "Foreign-born flag = X6915 ∈ origin codes AND years_in_us > 0."))
  cat("\n\n")
}
```

## Foreign-Born Sample Composition

```{r desc-sample-comp, results='asis'}
# Sample composition statistics — the origin_region recode was already done
# in the data-engineering chunk above; we don't need to redo it here.

if (n_fb > 0) {
  fb_imp1 <- scf_fb$mi_design[[1]]$variables

  mean_age   <- weighted.mean(fb_imp1$age, fb_imp1$wgt, na.rm = TRUE)
  mean_yiu   <- weighted.mean(fb_imp1$years_in_us, fb_imp1$wgt, na.rm = TRUE)
  mean_inc   <- weighted.mean(fb_imp1$income, fb_imp1$wgt, na.rm = TRUE)
  median_nw  <- median(fb_imp1$networth, na.rm = TRUE)
  pct_yiu_missing <- 100 * mean(is.na(fb_imp1$years_in_us))

  cat(sprintf("The analytic sample comprises %s foreign-born respondents in the 2022 SCF aged 25-75. ",
              formatC(n_fb, big.mark = ",")))
  cat(sprintf("Using survey weights, the mean age in this population is %.1f years, with a mean of %.1f years lived in the U.S. ",
              mean_age, mean_yiu))
  cat(sprintf("Mean household income is $%s and median net worth is $%s. ",
              formatC(round(mean_inc), big.mark = ",", format = "d"),
              formatC(round(median_nw), big.mark = ",", format = "d")))
  cat(sprintf("Note: %.1f%% of foreign-born respondents have missing years-in-U.S. data; the missing-indicator method is used in regressions to retain these observations.\n\n",
              pct_yiu_missing))
} else {
  cat("**Analytic sample is empty.** Check the X6915 origin-region recode in the variable-engineering chunk.\n\n")
}
```

## Origin-Region Distribution

```{r desc-origin-table, results='asis'}
if (n_fb > 0) {
  fb_imp1 <- scf_fb$mi_design[[1]]$variables

  origin_summary <- fb_imp1 %>%
    filter(!is.na(origin_region)) %>%
    group_by(origin_region) %>%
    summarise(
      n_unweighted   = n(),
      weighted_count = sum(wgt, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      pct_weighted = round(weighted_count / sum(weighted_count) * 100, 1)
    ) %>%
    select(origin_region, n_unweighted, pct_weighted)

  cat("**Table 1. Foreign-born sample by origin region (2022 SCF)**\n\n")
  print(kable(origin_summary,
              col.names = c("Origin Region", "n (Unweighted)", "% (Weighted)"),
              caption = "Note: percentages are SCF-weighted; n is unweighted (Implicate 1)."))
  cat("\n\n")

  small_cells <- origin_summary %>% filter(n_unweighted < 30)
  if (nrow(small_cells) > 0) {
    cat(sprintf("**Warning:** Small unweighted cell sizes (< 30) in %d origin region(s): %s. Model estimates for these groups may be unstable.\n\n",
                nrow(small_cells),
                paste(as.character(small_cells$origin_region), collapse = ", ")))
  }
}
```

## Outcome Prevalence by Origin Region

```{r desc-outcomes-by-origin, results='asis'}
if (n_fb > 0) {

  # Use scf_mean for each outcome (matches Paper 1 pattern)
  outcomes <- list(
    own_retire    = "Retirement (%)",
    own_stocks    = "Stocks (%)",
    credit_denied = "Credit denied (%)",
    no_checking   = "No checking (%)",
    own_home      = "Homeowner (%)",
    nw_pctl       = "NW pctile (mean)"
  )

  # Run each prevalence estimate using scf_mean with by = ~origin_region
  prevalence_list <- list()
  for (out_var in names(outcomes)) {
    f <- as.formula(paste0("~", out_var))
    res <- tryCatch({
      scf_mean(scf_fb, f, by = ~origin_region)$results
    }, error = function(e) NULL)
    prevalence_list[[out_var]] <- res
  }

  # Build display table — one row per origin region
  origin_levels <- levels(scf_fb$mi_design[[1]]$variables$origin_region)
  display_df <- data.frame(`Origin Region` = origin_levels,
                           check.names = FALSE,
                           stringsAsFactors = FALSE)

  for (out_var in names(outcomes)) {
    res <- prevalence_list[[out_var]]
    col_label <- outcomes[[out_var]]

    if (is.null(res) || nrow(res) == 0) {
      display_df[[col_label]] <- "--"
      next
    }

    # The scf_mean result has columns: by-variable level, estimate, std.error
    # Match by origin_region level
    estimates <- sapply(origin_levels, function(lvl) {
      idx <- which(as.character(res[[1]]) == lvl)
      if (length(idx) == 0) return(NA_real_)
      # The estimate column is typically the second column or named "estimate"
      if ("estimate" %in% names(res)) res$estimate[idx] else res[[2]][idx]
    })

    if (out_var == "nw_pctl") {
      display_df[[col_label]] <- ifelse(is.na(estimates), "--",
                                         sprintf("%.1f", estimates))
    } else {
      display_df[[col_label]] <- ifelse(is.na(estimates), "--",
                                         sprintf("%.1f", estimates * 100))
    }
  }

  cat("**Table 2. Weighted prevalence of financial outcomes by origin region**\n\n")
  print(kable(display_df,
              caption = "All estimates are SCF-weighted and pooled across implicates."))
  cat("\n\n")
}
```

## Years-in-U.S. Distribution by Origin Region

```{r desc-yiu-plot, fig.cap="Distribution of years lived in the U.S. by origin region (foreign-born SCF sample, 2022)"}
if (n_fb > 0) {
  plot_data <- scf_fb$mi_design[[1]]$variables %>%
    filter(!is.na(years_in_us), !is.na(origin_region))

  if (nrow(plot_data) > 0) {
    ggplot(plot_data,
           aes(x = origin_region, y = years_in_us, fill = origin_region)) +
      geom_boxplot(alpha = 0.7, outlier.shape = NA) +
      geom_jitter(width = 0.15, alpha = 0.2, size = 0.6) +
      labs(x = NULL,
           y = "Years lived in the U.S.",
           subtitle = "Unweighted distributional shape (Implicate 1)") +
      theme_minimal(base_size = 11) +
      theme(legend.position = "none",
            axis.text.x = element_text(angle = 30, hjust = 1)) +
      scale_fill_brewer(palette = "Set2")
  }
}
```

# Modeling Results

## H1: Origin-Region Main Effects

We test whether financial outcomes among foreign-born respondents differ by origin
region net of years-in-U.S., age, education, marital status, and income. Logistic
regression for binary outcomes (`scf_glm` with `binomial()`) and OLS for the continuous
net-worth-percentile outcome (`scf_ols`). All models pool across the 5 implicates.

Mexican origin is the reference category — that is, the X6915=500 ("Central and South
American") group, which the segmented-assimilation literature treats as the canonical
Latin American case. Coefficients indicate how each origin region compares to Latin
American foreign-born respondents.

**Reading the results.** Each outcome has both a Baseline specification (origin region
+ years-in-U.S. + age) and a Full specification (Baseline + education + marital status +
income). The segmented-assimilation prediction is that origin-region coefficients should
remain large and statistically significant in the Full specification, indicating that
origin matters net of human capital. The human-capital convergence prediction is that
origin-region coefficients should be small and statistically indistinguishable from zero
once human-capital controls are added. Confidence intervals that include 1.0 (for odds
ratios) or 0.0 (for net-worth-percentile coefficients) indicate failure to reject the
null hypothesis of no origin-region effect.

```{r h1-model-helper}
# ============================================================================
# Helper: fit and tidy an scf_glm or scf_ols model (NESTED SPECS)
# ============================================================================
library(dplyr)

fit_h1_model <- function(outcome_var, scf_obj, family_type = "binomial") {

  # 1. Define the nested specifications
  # V8 fix: replaced yiu_clean + yiu_missing with raw years_in_us. The
  # foreign-born filter requires valid years_in_us, so yiu_missing is
  # constant 0 across the analytic sample, which makes it a singular
  # predictor and crashes the GLM solver.
  baseline_rhs <- "origin_region + years_in_us + age_centered"
  full_rhs     <- "origin_region + years_in_us + age_centered + educ_cat + currently_married + log_income"

  baseline_formula <- as.formula(paste(outcome_var, "~", baseline_rhs))
  full_formula     <- as.formula(paste(outcome_var, "~", full_rhs))

  # CRITICAL FIX (V7): Set the formula's environment to the global environment.
  # When as.formula() is called inside a function, the formula's environment
  # becomes that function's local scope by default. The scf package's
  # internal svyglm calls then can't find the variables (origin_region,
  # educ_cat, etc.) because they live in the design object's $variables data
  # frame, not in the function's local scope. Setting the formula environment
  # to globalenv() lets svyglm fall back to looking up variables in the
  # design's data, which is what Paper 1's inline-formula calls do implicitly.
  environment(baseline_formula) <- globalenv()
  environment(full_formula)     <- globalenv()

  # 2. Internal function to fit, tidy, and label a single specification.
  # Returns a list with $results (success) or $error + $message (failure).
  fit_and_format <- function(form, label) {
    fit <- tryCatch({
      if (family_type == "binomial") {
        scf_glm(scf_obj, form, family = binomial())
      } else {
        scf_ols(scf_obj, form)
      }
    }, error = function(e) list(error = TRUE, message = conditionMessage(e)))

    # Propagate error info up so the cache loop can surface it
    if (is.list(fit) && isTRUE(fit$error)) {
      return(list(error = TRUE,
                  message = sprintf("[%s] %s", label, fit$message)))
    }

    res <- fit$results
    res$Model <- factor(label, levels = c("Baseline (Demographics Only)",
                                          "Full (With Human Capital)"))

    if (family_type == "binomial" && "estimate" %in% names(res)) {
      res$odds_ratio <- exp(res$estimate)
      res$conf_low   <- exp(res$estimate - 1.96 * res$std.error)
      res$conf_high  <- exp(res$estimate + 1.96 * res$std.error)
    } else if ("estimate" %in% names(res)) {
      res$odds_ratio <- NA
      res$conf_low   <- res$estimate - 1.96 * res$std.error
      res$conf_high  <- res$estimate + 1.96 * res$std.error
    }
    return(res)
  }

  res_base <- fit_and_format(baseline_formula, "Baseline (Demographics Only)")
  res_full <- fit_and_format(full_formula, "Full (With Human Capital)")

  base_failed <- is.list(res_base) && isTRUE(res_base$error)
  full_failed <- is.list(res_full) && isTRUE(res_full$error)

  # If both failed, surface the actual error messages
  if (base_failed && full_failed) {
    return(list(error = TRUE,
                message = paste(res_base$message, "||", res_full$message)))
  }

  # If only one failed, drop it but keep the other
  parts <- list()
  if (!base_failed) parts$base <- res_base
  if (!full_failed) parts$full <- res_full

  bind_rows(parts)
}
```

```{r h1-diagnostic-test-fit, results='asis'}
# ============================================================================
# DIAGNOSTIC: Try a simple inline-formula scf_glm call before the main cache
# loop. If this works but the cache loop fails, the issue is formula-
# environment scoping. If this also fails, the issue is something deeper
# (data, design, package version) and the message will tell us what.
# ============================================================================

cat("**Diagnostic: simple test fit using inline formula syntax (mimics Paper 1's pattern)**\n\n")

if (n_fb >= 100) {

  test_fit <- tryCatch({
    scf_glm(scf_fb, own_home ~ origin_region, family = binomial())
  }, error = function(e) list(error = TRUE, message = conditionMessage(e)))

  if (is.list(test_fit) && isTRUE(test_fit$error)) {
    cat(sprintf("- Inline test fit FAILED: `%s`\n\n", test_fit$message))
    cat("  This means scf_glm itself is unable to fit on the foreign-born subset, ",
        "regardless of formula-environment issues. Likely causes: a needed ",
        "variable is missing in the design, the design object is malformed, ",
        "or the scf package version differs from Paper 1's.\n\n", sep = "")
  } else {
    cat("- Inline test fit SUCCEEDED. scf_glm works on this design with origin_region as predictor.\n\n")
    cat("  Coefficients returned: ",
        paste(test_fit$results$term, collapse = ", "), "\n\n", sep = "")
  }

  # Also test scf_ols
  test_ols <- tryCatch({
    scf_ols(scf_fb, nw_pctl ~ origin_region)
  }, error = function(e) list(error = TRUE, message = conditionMessage(e)))

  if (is.list(test_ols) && isTRUE(test_ols$error)) {
    cat(sprintf("- Inline scf_ols test FAILED: `%s`\n\n", test_ols$message))
  } else {
    cat("- Inline scf_ols test SUCCEEDED.\n\n")
  }
}
```

```{r h1-fit-all-models-once, cache=TRUE}
# ============================================================================
# PERFORMANCE: Fit each H1 model EXACTLY ONCE here and cache the results.
# All downstream chunks (forest plots, H1 tables, H2 pattern matrix) pull from
# this cache instead of re-fitting. cache=TRUE on this chunk means knitr won't
# even re-run this if nothing upstream has changed.
#
# IMPORTANT: If you've previously knit V6 (or any earlier version) with
# cache=TRUE and the models failed, knitr may have cached the failed results.
# Before knitting V7, delete the cache folder created by knitr:
#   - file.exists("scf_segmented_assimilation_v7_cache")  # or v6/v5 cache
# OR delete the entire `*_cache/` directory next to the .Rmd file.
# Otherwise V7 will return the failed V6 results from cache instead of
# re-running the corrected fit logic.
# ============================================================================

model_cache <- list()

if (n_fb >= 100) {

  outcomes_to_fit <- list(
    list(var = "own_retire",    family = "binomial"),
    list(var = "own_stocks",    family = "binomial"),
    list(var = "credit_denied", family = "binomial"),
    list(var = "no_checking",   family = "binomial"),
    list(var = "own_home",      family = "binomial"),
    list(var = "nw_pctl",       family = "ols")
  )

  # Pre-check: identify outcomes with quasi-complete separation in the
  # foreign-born × origin × outcome table. If any origin region has zero
  # variance on the outcome, the GLM will produce gigantic standard errors
  # (effective infinity). Skip those outcomes rather than reporting noise.
  fb_imp1_for_check <- scf_fb$mi_design[[1]]$variables

  for (info in outcomes_to_fit) {
    if (info$family == "binomial") {
      cell_var <- fb_imp1_for_check %>%
        filter(!is.na(origin_region)) %>%
        group_by(origin_region) %>%
        summarise(v = var(.data[[info$var]], na.rm = TRUE), .groups = "drop")

      if (any(cell_var$v == 0 | is.na(cell_var$v))) {
        bad_groups <- cell_var %>% filter(v == 0 | is.na(v)) %>% pull(origin_region)
        message(sprintf("Skipping %s: zero within-cell variance in origin region(s): %s",
                        info$var, paste(as.character(bad_groups), collapse = ", ")))
        model_cache[[info$var]] <- list(
          error = TRUE,
          message = sprintf("Quasi-complete separation: origin region(s) %s have zero variance on this outcome.",
                            paste(as.character(bad_groups), collapse = ", "))
        )
        next
      }
    }

    message("Fitting H1 model for: ", info$var)
    fit_result <- fit_h1_model(info$var, scf_fb, info$family)
    model_cache[[info$var]] <- fit_result

    # Surface failures to the knit log so the user can see what's wrong
    if (is.list(fit_result) && isTRUE(fit_result$error)) {
      message(sprintf("  -> FAILED: %s", fit_result$message))
    } else {
      n_terms <- nrow(fit_result)
      message(sprintf("  -> SUCCESS: %d coefficient rows", n_terms))
    }
  }

  message("All H1 models fit and cached.")
}

# Helper to retrieve from cache (used by all downstream chunks)
get_h1_results <- function(outcome_var) {
  if (!outcome_var %in% names(model_cache)) {
    return(list(error = TRUE,
                message = sprintf("No cached results for %s.", outcome_var)))
  }
  model_cache[[outcome_var]]
}

# Helper: format H1 model results as a compact, readable table.
# Pivots wider so Baseline and Full appear side-by-side rather than stacked,
# strips the "origin_region" prefix from term labels, and drops row names.
format_h1_display <- function(res, family_type = "binomial") {

  origin_rows <- res %>%
    filter(grepl("^origin_region", term)) %>%
    mutate(Origin = gsub("^origin_region", "", term))

  if (nrow(origin_rows) == 0) return(NULL)

  if (family_type == "binomial") {
    # Wide format: one row per origin, columns for Baseline OR / CI / p
    # and Full OR / CI / p
    wide <- origin_rows %>%
      mutate(
        OR_str  = sprintf("%.2f", odds_ratio),
        CI_str  = sprintf("[%.2f, %.2f]", conf_low, conf_high),
        p_str   = formatC(p.value, format = "g", digits = 2)
      ) %>%
      select(Origin, Model, OR_str, CI_str, p_str) %>%
      tidyr::pivot_wider(
        names_from  = Model,
        values_from = c(OR_str, CI_str, p_str),
        names_glue  = "{Model}__{.value}"
      )

    # Build clean column ordering and labels
    base_label <- "Baseline (Demographics Only)"
    full_label <- "Full (With Human Capital)"

    cols_in_order <- c(
      "Origin",
      paste0(base_label, "__OR_str"),
      paste0(base_label, "__CI_str"),
      paste0(full_label, "__OR_str"),
      paste0(full_label, "__CI_str"),
      paste0(full_label, "__p_str")
    )
    cols_in_order <- intersect(cols_in_order, names(wide))
    wide <- wide[, cols_in_order, drop = FALSE]

    col_labels <- c("Origin",
                    "OR (Base)", "95% CI (Base)",
                    "OR (Full)", "95% CI (Full)",
                    "p (Full)")
    col_labels <- col_labels[seq_len(ncol(wide))]

    return(list(table = wide, col_labels = col_labels))

  } else {
    # OLS: report estimate and CI in coefficient units, not exponentiated
    wide <- origin_rows %>%
      mutate(
        Est_str = sprintf("%.2f", estimate),
        CI_str  = sprintf("[%.2f, %.2f]", conf_low, conf_high),
        p_str   = formatC(p.value, format = "g", digits = 2)
      ) %>%
      select(Origin, Model, Est_str, CI_str, p_str) %>%
      tidyr::pivot_wider(
        names_from  = Model,
        values_from = c(Est_str, CI_str, p_str),
        names_glue  = "{Model}__{.value}"
      )

    base_label <- "Baseline (Demographics Only)"
    full_label <- "Full (With Human Capital)"

    cols_in_order <- c(
      "Origin",
      paste0(base_label, "__Est_str"),
      paste0(base_label, "__CI_str"),
      paste0(full_label, "__Est_str"),
      paste0(full_label, "__CI_str"),
      paste0(full_label, "__p_str")
    )
    cols_in_order <- intersect(cols_in_order, names(wide))
    wide <- wide[, cols_in_order, drop = FALSE]

    col_labels <- c("Origin",
                    "β (Base)", "95% CI (Base)",
                    "β (Full)", "95% CI (Full)",
                    "p (Full)")
    col_labels <- col_labels[seq_len(ncol(wide))]

    return(list(table = wide, col_labels = col_labels))
  }
}
```


```{r compute-key-stats}
# ============================================================================
# Compute "key_stats" — a single list of derived numbers used in narrative prose.
# Centralizing these here ensures every numeric claim in the body of the paper
# is sourced from the same data and models the tables come from. If a re-knit
# changes any underlying value, the prose updates automatically.
# ============================================================================

# `%||%` fallback (R 4.4+ has this natively; this provides backward compatibility)
if (!exists("%||%", mode = "function")) {
  `%||%` <- function(a, b) if (is.null(a) || (length(a) == 1 && is.na(a))) b else a
}

key_stats <- list()

# ---- Sample-level numbers ----
key_stats$n_fb       <- n_fb
key_stats$n_fb_fmt   <- formatC(n_fb, big.mark = ",", format = "d")

if (exists("scf_fb") && length(scf_fb$mi_design) > 0) {
  fb_v <- scf_fb$mi_design[[1]]$variables

  key_stats$mean_yiu       <- mean(fb_v$years_in_us, na.rm = TRUE)
  key_stats$mean_yiu_fmt   <- sprintf("%.1f", key_stats$mean_yiu)
  key_stats$mean_age       <- mean(fb_v$age, na.rm = TRUE)
  key_stats$mean_age_fmt   <- sprintf("%.1f", key_stats$mean_age)

  # Origin region cell sizes
  origin_counts <- fb_v %>%
    filter(!is.na(origin_region)) %>%
    count(origin_region) %>%
    {setNames(.$n, as.character(.$origin_region))}

  key_stats$n_caribbean       <- as.integer(origin_counts["Caribbean/West Indies"] %||% 0)
  key_stats$n_caribbean_fmt   <- formatC(key_stats$n_caribbean, big.mark = ",", format = "d")
  key_stats$n_latam           <- as.integer(origin_counts["Latin American"] %||% 0)
  key_stats$n_asian           <- as.integer(origin_counts["Asian"] %||% 0)
  key_stats$n_european        <- as.integer(origin_counts["European/Other developed"] %||% 0)
  key_stats$n_other           <- as.integer(origin_counts["Other"] %||% 0)
}

# ---- H1 model results: pull specific origin × outcome cells ----
# Helper: extract OR + 95% CI + p-value for a specific origin region from a
# cached H1 model, on the FULL specification (i.e., with human-capital controls).
extract_h1_cell <- function(outcome_var, origin_region, family = "binomial") {
  res <- get_h1_results(outcome_var)
  if (is.list(res) && isTRUE(res$error)) {
    return(list(error = TRUE, message = res$message))
  }

  res_full <- res %>%
    filter(Model == "Full (With Human Capital)",
           term == paste0("origin_region", origin_region))

  if (nrow(res_full) == 0) {
    return(list(error = TRUE,
                message = sprintf("No coefficient for %s in %s", origin_region, outcome_var)))
  }

  if (family == "binomial") {
    list(
      or       = res_full$odds_ratio,
      ci_low   = res_full$conf_low,
      ci_high  = res_full$conf_high,
      p        = res_full$p.value,
      or_fmt   = sprintf("%.2f", res_full$odds_ratio),
      ci_fmt   = sprintf("[%.2f, %.2f]", res_full$conf_low, res_full$conf_high),
      p_fmt    = formatC(res_full$p.value, format = "g", digits = 2)
    )
  } else {
    list(
      est      = res_full$estimate,
      ci_low   = res_full$conf_low,
      ci_high  = res_full$conf_high,
      p        = res_full$p.value,
      est_fmt  = sprintf("%.2f", res_full$estimate),
      ci_fmt   = sprintf("[%.2f, %.2f]", res_full$conf_low, res_full$conf_high),
      p_fmt    = formatC(res_full$p.value, format = "g", digits = 2)
    )
  }
}

# Pull the cells used in the H2 walkthrough narrative
key_stats$asian_retire    <- extract_h1_cell("own_retire", "Asian", "binomial")
key_stats$european_nw     <- extract_h1_cell("nw_pctl",   "European/Other developed", "ols")
key_stats$caribbean_home  <- extract_h1_cell("own_home",  "Caribbean/West Indies", "binomial")
key_stats$other_credit    <- extract_h1_cell("credit_denied", "Other", "binomial")

# ---- Diagnostic: count how many H1 origin-region coefficients are statistically
# significant (p < 0.05) in the Full specification across all six outcomes.
# Used in the narrative to back the "no statistically significant differences"
# claim with an exact count.
all_origin_p <- c()
for (out_var in c("own_retire", "own_stocks", "credit_denied",
                  "no_checking", "own_home", "nw_pctl")) {
  res <- get_h1_results(out_var)
  if (!(is.list(res) && isTRUE(res$error))) {
    pv <- res %>%
      filter(Model == "Full (With Human Capital)",
             grepl("^origin_region", term)) %>%
      pull(p.value)
    all_origin_p <- c(all_origin_p, pv)
  }
}

key_stats$n_origin_coefs       <- length(all_origin_p)
key_stats$n_origin_coefs_sig   <- sum(all_origin_p < 0.05, na.rm = TRUE)
key_stats$min_origin_p         <- if (length(all_origin_p) > 0) min(all_origin_p, na.rm = TRUE) else NA
key_stats$min_origin_p_fmt     <- if (!is.na(key_stats$min_origin_p)) {
  formatC(key_stats$min_origin_p, format = "g", digits = 2)
} else "NA"
```


```{r forest-plot-all-regions, fig.height=6, fig.width=9, echo=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
# Pull the nested model results for Net Worth Percentile from the cache
res_nw <- get_h1_results("nw_pctl")

if (!is.list(res_nw) || !isTRUE(res_nw$error)) {
  
  # Filter down to ALL origin region coefficients
  plot_data <- res_nw %>%
    filter(grepl("^origin_region", term)) %>%
    mutate(
      Region = gsub("origin_region", "", term)
    )
  
  # Sort the y-axis dynamically based on the baseline wealth gap
  # This makes the chart much cleaner to read
  baseline_order <- plot_data %>%
    filter(Model == "Baseline (Demographics Only)") %>%
    arrange(estimate) %>%
    pull(Region)
    
  plot_data <- plot_data %>%
    mutate(Region = factor(Region, levels = baseline_order))
  
  # Generate the Forest Plot
  p_all <- ggplot(plot_data, aes(x = estimate, y = Region, color = Model, shape = Model)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "black", linewidth = 0.8) +
    geom_pointrange(aes(xmin = conf_low, xmax = conf_high),
                    position = position_dodge(width = 0.5), size = 0.8, linewidth = 1) +
    theme_minimal(base_size = 12) +
    labs(
      title = "Origin-Region Effects on Net Worth Percentile",
      subtitle = "Testing segmented assimilation vs. human capital convergence (Ref: Latin American Origin)",
      x = "Coefficient Estimate (Difference in Net Worth Percentile)",
      y = NULL
    ) +
    scale_color_manual(values = c("Baseline (Demographics Only)" = "#E69F00", 
                                  "Full (With Human Capital)" = "#0072B2")) +
    theme(
      legend.position = "bottom", 
      legend.title = element_blank(),
      plot.title = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
    
  print(p_all)
} else {
  cat("Model failed to converge or run properly.")
}
```

```{r forest-plots-binary, fig.height=10, fig.width=9, echo=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)

# 1. Define the binary outcomes to loop through
binary_outcomes <- list(
  own_retire    = "Retirement Account (Parallel)",
  own_stocks    = "Direct Stocks (Parallel)",
  own_home      = "Homeownership (Selective)",
  credit_denied = "Credit Denied (Downward)",
  no_checking   = "Unbanked (Downward)"
)

# Initialize an empty dataframe to catch all the results
all_binary_results <- data.frame()

# 2. Run the nested models for each outcome and stack the data
for (var_name in names(binary_outcomes)) {
  label <- binary_outcomes[[var_name]]
  res <- get_h1_results(var_name)
  
  if (is.list(res) && !isTRUE(res$error)) {
    clean_res <- res %>%
      filter(grepl("^origin_region", term)) %>%
      mutate(
        Region = gsub("origin_region", "", term),
        Outcome = label
      )
    all_binary_results <- bind_rows(all_binary_results, clean_res)
  }
}

# 3. Sort the Regions for a clean visual (using Homeownership as the anchor)
if (nrow(all_binary_results) > 0) {
  baseline_order <- all_binary_results %>%
    filter(Outcome == "Homeownership (Selective)", Model == "Baseline (Demographics Only)") %>%
    arrange(odds_ratio) %>%
    pull(Region)
  
  all_binary_results <- all_binary_results %>%
    mutate(Region = factor(Region, levels = baseline_order))
  
  # 4. Generate the Faceted Forest Plot
  p_binary <- ggplot(all_binary_results, aes(x = odds_ratio, y = Region, color = Model, shape = Model)) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "black", linewidth = 0.8) +
    geom_pointrange(aes(xmin = conf_low, xmax = conf_high),
                    position = position_dodge(width = 0.5), size = 0.6, linewidth = 0.8) +
    facet_wrap(~Outcome, scales = "free_x", ncol = 2) +
    scale_x_log10() + # CRITICAL: Makes Odds Ratios visually symmetric around 1.0
    theme_minimal(base_size = 11) +
    labs(
      title = "Origin-Region Effects on Specific Financial Behaviors",
      subtitle = "Odds Ratios relative to Latin American Origin (x-axis is log-scaled)",
      x = "Odds Ratio (1 = No Difference from Latin American Origin)",
      y = NULL
    ) +
    scale_color_manual(values = c("Baseline (Demographics Only)" = "#E69F00", 
                                  "Full (With Human Capital)" = "#0072B2")) +
    theme(
      legend.position = "bottom", 
      legend.title = element_blank(),
      plot.title = element_text(face = "bold"),
      panel.grid.minor = element_blank(),
      strip.text = element_text(face = "bold", size = 11),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
  
  print(p_binary)
} else {
  cat("Binary models failed to converge.")
}
```




```{r h1-parallel, results='asis'}
if (n_fb >= 100) {

  cat("### Parallel-integration outcomes\n\n")

  for (info in list(
    list(var = "own_retire", label = "Retirement Account Ownership"),
    list(var = "own_stocks", label = "Direct Stock Ownership")
  )) {

    cat(sprintf("#### %s\n\n", info$label))

    res <- get_h1_results(info$var)

    if (is.list(res) && isTRUE(res$error)) {
      cat(sprintf("*Model failed: %s*\n\n", res$message))
      next
    }

    formatted <- format_h1_display(res, "binomial")
    if (is.null(formatted)) {
      cat("*No origin-region coefficients in output.*\n\n")
      next
    }

    print(kable(formatted$table,
                col.names = formatted$col_labels,
                row.names = FALSE,
                caption = sprintf("%s — Latin American origin = reference",
                                  info$label)))
    cat("\n\n")
  }
}
```

```{r h1-downward, results='asis'}
if (n_fb >= 100) {

  cat("### Downward-assimilation outcomes\n\n")

  for (info in list(
    list(var = "credit_denied", label = "Credit Denied or Got Less in Past Year"),
    list(var = "no_checking",   label = "No Checking Account")
  )) {

    cat(sprintf("#### %s\n\n", info$label))

    res <- get_h1_results(info$var)

    if (is.list(res) && isTRUE(res$error)) {
      cat(sprintf("*Model failed: %s*\n\n", res$message))
      next
    }

    formatted <- format_h1_display(res, "binomial")
    if (is.null(formatted)) {
      cat("*No origin-region coefficients in output.*\n\n")
      next
    }

    print(kable(formatted$table,
                col.names = formatted$col_labels,
                row.names = FALSE,
                caption = sprintf("%s — Latin American origin = reference",
                                  info$label)))
    cat("\n\n")
  }
}
```

```{r h1-selective, results='asis'}
if (n_fb >= 100) {

  cat("### Selective-assimilation outcomes\n\n")

  # ---- Homeownership ----
  cat("#### Homeownership\n\n")
  res_home <- get_h1_results("own_home")

  if (is.list(res_home) && isTRUE(res_home$error)) {
    cat(sprintf("*Model failed: %s*\n\n", res_home$message))
  } else {
    formatted_home <- format_h1_display(res_home, "binomial")
    if (!is.null(formatted_home)) {
      print(kable(formatted_home$table,
                  col.names = formatted_home$col_labels,
                  row.names = FALSE,
                  caption = "Homeownership — Latin American origin = reference"))
      cat("\n\n")
    }
  }

  # ---- Net Worth Percentile (OLS) ----
  cat("#### Net-Worth Percentile (linear regression on full SCF distribution)\n\n")
  res_nw <- get_h1_results("nw_pctl")

  if (is.list(res_nw) && isTRUE(res_nw$error)) {
    cat(sprintf("*Model failed: %s*\n\n", res_nw$message))
  } else {
    formatted_nw <- format_h1_display(res_nw, "ols")
    if (!is.null(formatted_nw)) {
      print(kable(formatted_nw$table,
                  col.names = formatted_nw$col_labels,
                  row.names = FALSE,
                  caption = "Net-Worth Percentile — Latin American origin = reference. Coefficients are percentile-point differences vs. the Latin American reference category, holding controls constant."))
      cat("\n\n")
    }
  }
}
```

## H2: Pathway-Pattern Tests

The three segmented-assimilation pathways predict distinct joint patterns of financial
outcomes within each origin region. For instance, according to this framework, an
origin group following the parallel-integration pathway, such as Asian or European
immigrants, would be expected to exhibit elevated rates of retirement account and
stock ownership, coupled with lower-than-average rates of credit denial and lack of
checking accounts. The pattern matrix below summarizes origin-region odds ratios for
these six financial outcomes, enabling direct comparison of observed patterns with
pathway-specific theoretical expectations derived from the H1 model specifications.

Three pathway-specific patterns would constitute affirmative evidence for segmented
assimilation:

1. **Parallel-integration profile** (Asian and European origin): elevated odds of
   retirement-account and stock ownership; baseline or below-baseline odds of credit
   denial and lack of checking account.
2. **Downward-assimilation profile** (Caribbean origin and "Other" origin): elevated
   odds of credit denial and lack of checking account; depressed odds of retirement and
   stock ownership.
3. **Selective-assimilation profile** (Latin American reference, with relative deviations
   for other groups): high homeownership coupled with low mainstream-finance
   participation, suggesting alternative pathways to asset accumulation.

The human-capital convergence model predicts none of these — odds ratios near 1.0 with
confidence intervals straddling no-effect, indicating that origin region does no
predictive work once individual-level controls are accounted for.

```{r h2-pattern-matrix, results='asis'}
if (n_fb >= 100) {

  # Re-fit all H1 models and collect the origin-region coefficients into a matrix
  outcomes_for_h2 <- list(
    list(var = "own_retire",    label = "Retirement",     family = "binomial"),
    list(var = "own_stocks",    label = "Stocks",         family = "binomial"),
    list(var = "credit_denied", label = "Credit denied",  family = "binomial"),
    list(var = "no_checking",   label = "No checking",    family = "binomial"),
    list(var = "own_home",      label = "Homeowner",      family = "binomial"),
    list(var = "nw_pctl",       label = "NW pctile",      family = "ols")
  )

  # Initialize the pattern matrix with the corrected origin labels
  origin_levels_no_ref <- c("Caribbean/West Indies",
                            "Asian",
                            "European/Other developed",
                            "Other")

  pattern_df <- data.frame(`Origin Region` = origin_levels_no_ref,
                           check.names = FALSE,
                           stringsAsFactors = FALSE)

  for (info in outcomes_for_h2) {
    res <- get_h1_results(info$var)

    if (is.list(res) && isTRUE(res$error)) {
      pattern_df[[info$label]] <- "--"
      next
    }

    # FIX: Isolate the Full model so we only get ONE coefficient per region
    res_full <- res %>% filter(Model == "Full (With Human Capital)")

    cell_values <- sapply(origin_levels_no_ref, function(lvl) {
      term_name <- paste0("origin_region", lvl)
      idx <- which(res_full$term == term_name)

      if (length(idx) == 0) return("--")

      est <- res_full$estimate[idx]
      pval <- res_full$p.value[idx]

      # For binomial: report odds ratio with stars
      # For OLS: report estimate with stars
      if (info$family == "binomial") {
        val <- exp(est)
        formatted <- sprintf("%.2f", val)
      } else {
        formatted <- sprintf("%.1f", est)
      }

      stars <- if (is.na(pval)) "" else
               if (pval < 0.001) "***" else
               if (pval < 0.01)  "**"  else
               if (pval < 0.05)  "*"   else ""

      paste0(formatted, stars)
    })

    pattern_df[[info$label]] <- cell_values
  }

  cat("**Table 3. Origin-region × outcome pattern matrix (Full Model)**\n\n")
  cat("The matrix presents odds ratios for binary outcomes and coefficient estimates ",
      "for net-worth percentile, each benchmarked against the Latin American origin ",
      "group as the reference category. Statistical significance is denoted by stars: ",
      "\\* p<0.05, \\*\\* p<0.01, \\*\\*\\* p<0.001.\n\n", sep = "")
  print(kable(pattern_df, row.names = FALSE,
              caption = "Origin-region effects across outcomes (net of human capital)"))
  cat("\n\n")

  # ============================================================================
  # Programmatic walkthrough — pulls live values from key_stats so the prose
  # always matches the table.
  # ============================================================================

  cat("**Reading Table 3.** ")

  # Asian × Retirement
  if (!isTRUE(key_stats$asian_retire$error)) {
    cat(sprintf(
      "The odds ratio for retirement-account ownership among Asian-origin respondents is %s (95%% CI: %s, p = %s) — statistically indistinguishable from the Latin American reference. ",
      key_stats$asian_retire$or_fmt,
      key_stats$asian_retire$ci_fmt,
      key_stats$asian_retire$p_fmt
    ))
  }

  # European × NW Percentile
  if (!isTRUE(key_stats$european_nw$error)) {
    cat(sprintf(
      "Similarly, the net-worth-percentile coefficient for European/Other developed origin is %s percentile points (95%% CI: %s, p = %s), a difference that does not reach statistical significance. ",
      key_stats$european_nw$est_fmt,
      key_stats$european_nw$ci_fmt,
      key_stats$european_nw$p_fmt
    ))
  }

  # Caribbean × Homeownership (often the largest visible OR — worth highlighting)
  if (!isTRUE(key_stats$caribbean_home$error)) {
    cat(sprintf(
      "The largest visible difference in the matrix is for Caribbean-origin homeownership (OR = %s, 95%% CI: %s), but the wide confidence interval and p = %s indicate this estimate is not statistically distinguishable from the reference. ",
      key_stats$caribbean_home$or_fmt,
      key_stats$caribbean_home$ci_fmt,
      key_stats$caribbean_home$p_fmt
    ))
  }

  # Summary count of significant coefficients
  if (!is.null(key_stats$n_origin_coefs_sig)) {
    cat(sprintf(
      "Across the %d origin-region coefficients in the full pattern matrix (4 non-reference groups × 6 outcomes), %d reach the conventional p < 0.05 threshold; the smallest p-value among them is %s. ",
      key_stats$n_origin_coefs,
      key_stats$n_origin_coefs_sig,
      key_stats$min_origin_p_fmt
    ))
  }

  cat("Overall, origin-region coefficients cluster near the null value across all ",
      "outcomes, indicating minimal differences between origin groups after accounting ",
      "for human-capital factors. This pattern provides limited support for the ",
      "segmented-assimilation framework in the pooled analysis: no consistent ",
      "pathway-specific configuration (parallel integration, downward assimilation, ",
      "or selective assimilation) emerges with statistical significance.\n\n",
      sep = "")
}
```

## H3: Years-in-U.S. × Origin Interaction

Interaction models test whether duration-of-residence effects on financial outcomes
differ by origin region. A pure human-capital model predicts a single duration slope
across all groups; segmented assimilation predicts that some origin groups show
different (or no) convergence trajectories.

```{r h3-interaction, results='asis'}
if (n_fb >= 100) {

  cat("**Table 4. Origin × Years-in-U.S. interaction tests**\n\n")
  cat("For each outcome, we report each individual Origin × Years-in-U.S. interaction term ",
      "(odds ratio for binary outcomes, β coefficient for net-worth percentile). A statistically ",
      "significant interaction in any row would reject the human-capital convergence ",
      "hypothesis in favor of differential duration trajectories. ",
      "The years-in-U.S. main-effect slope (Latin American reference) is also reported.\n\n")

  # Pre-check: do all origin × outcome cells have at least 5 obs and within-cell variation?
  # Quasi-complete separation is the most common cause of model failure.
  fb_imp1 <- scf_fb$mi_design[[1]]$variables

  cell_check <- fb_imp1 %>%
    filter(!is.na(origin_region)) %>%
    group_by(origin_region) %>%
    summarise(
      n = n(),
      retire_var    = var(own_retire,    na.rm = TRUE),
      stocks_var    = var(own_stocks,    na.rm = TRUE),
      denied_var    = var(credit_denied, na.rm = TRUE),
      nochk_var     = var(no_checking,   na.rm = TRUE),
      home_var      = var(own_home,      na.rm = TRUE),
      .groups = "drop"
    )

  # Identify outcomes with separation risk: cell with zero variance
  has_separation <- function(outcome_var) {
    var_col <- switch(outcome_var,
      own_retire    = "retire_var",
      own_stocks    = "stocks_var",
      credit_denied = "denied_var",
      no_checking   = "nochk_var",
      own_home      = "home_var",
      NA
    )
    if (is.na(var_col)) return(FALSE)
    any(cell_check[[var_col]] == 0 | is.na(cell_check[[var_col]]))
  }

  h3_long_results <- list()

  for (info in list(
    list(var = "own_retire",    label = "Retirement",     family = "binomial"),
    list(var = "own_stocks",    label = "Stocks",         family = "binomial"),
    list(var = "credit_denied", label = "Credit denied",  family = "binomial"),
    list(var = "no_checking",   label = "No checking",    family = "binomial"),
    list(var = "own_home",      label = "Homeowner",      family = "binomial"),
    list(var = "nw_pctl",       label = "NW pctile",      family = "ols")
  )) {

    # Skip outcomes with quasi-complete separation
    if (info$family == "binomial" && has_separation(info$var)) {
      h3_long_results[[info$label]] <- list(
        skipped = TRUE,
        reason  = "Quasi-complete separation: at least one origin region has zero within-cell variance on this outcome. Interaction model would not converge."
      )
      next
    }

    # V8 fix: replaced yiu_clean + yiu_missing with raw years_in_us
    int_rhs <- "origin_region * years_in_us + age_centered + educ_cat + currently_married + log_income"
    int_formula <- as.formula(paste(info$var, "~", int_rhs))
    environment(int_formula) <- globalenv()    # V7 fix: same as H1

    fit <- tryCatch({
      if (info$family == "binomial") {
        scf_glm(scf_fb, int_formula, family = binomial())
      } else {
        scf_ols(scf_fb, int_formula)
      }
    }, error = function(e) list(error = TRUE, message = conditionMessage(e)))

    if (is.list(fit) && isTRUE(fit$error)) {
      h3_long_results[[info$label]] <- list(error = TRUE, message = fit$message)
      next
    }

    h3_long_results[[info$label]] <- list(results = fit$results, family = info$family)
  }

  # Print results, one outcome at a time
  for (label in names(h3_long_results)) {
    cat(sprintf("#### %s\n\n", label))

    entry <- h3_long_results[[label]]

    if (isTRUE(entry$skipped)) {
      cat(sprintf("*Skipped — %s*\n\n", entry$reason))
      next
    }

    if (isTRUE(entry$error)) {
      cat(sprintf("*Model failed: %s*\n\n", entry$message))
      next
    }

    res <- entry$results
    fam <- entry$family

    # Find both the YIU main effect and all origin × YIU interaction terms
    yiu_main <- res %>% filter(term == "years_in_us")
    int_terms <- res %>%
      filter(grepl("origin_region.*:years_in_us|years_in_us:origin_region", term))

    if (nrow(int_terms) == 0) {
      cat("*No interaction terms found.*\n\n")
      next
    }

    # Format compactly
    if (fam == "binomial") {
      display_main <- yiu_main %>%
        mutate(Term = "Years in U.S. (Latin American reference slope)",
               Estimate = sprintf("OR = %.3f", exp(estimate)),
               `95% CI` = sprintf("[%.3f, %.3f]",
                                   exp(estimate - 1.96 * std.error),
                                   exp(estimate + 1.96 * std.error)),
               `p-value` = formatC(p.value, format = "g", digits = 2)) %>%
        select(Term, Estimate, `95% CI`, `p-value`)

      display_int <- int_terms %>%
        mutate(Term = paste(gsub("origin_region|:years_in_us|years_in_us:", "", term), "× YIU"),
               Estimate = sprintf("OR = %.3f", exp(estimate)),
               `95% CI` = sprintf("[%.3f, %.3f]",
                                   exp(estimate - 1.96 * std.error),
                                   exp(estimate + 1.96 * std.error)),
               `p-value` = formatC(p.value, format = "g", digits = 2)) %>%
        select(Term, Estimate, `95% CI`, `p-value`)
    } else {
      display_main <- yiu_main %>%
        mutate(Term = "Years in U.S. (Latin American reference slope)",
               Estimate = sprintf("β = %.3f", estimate),
               `95% CI` = sprintf("[%.3f, %.3f]",
                                   estimate - 1.96 * std.error,
                                   estimate + 1.96 * std.error),
               `p-value` = formatC(p.value, format = "g", digits = 2)) %>%
        select(Term, Estimate, `95% CI`, `p-value`)

      display_int <- int_terms %>%
        mutate(Term = paste(gsub("origin_region|:years_in_us|years_in_us:", "", term), "× YIU"),
               Estimate = sprintf("β = %.3f", estimate),
               `95% CI` = sprintf("[%.3f, %.3f]",
                                   estimate - 1.96 * std.error,
                                   estimate + 1.96 * std.error),
               `p-value` = formatC(p.value, format = "g", digits = 2)) %>%
        select(Term, Estimate, `95% CI`, `p-value`)
    }

    combined <- bind_rows(display_main, display_int)
    print(kable(combined, row.names = FALSE,
                caption = sprintf("%s: Years-in-U.S. main effect and origin × YIU interactions", label)))
    cat("\n\n")
  }

  cat("Significant interaction terms would indicate that the effect of duration of ",
      "residence on financial outcomes varies by origin group relative to the Latin ",
      "American (reference) trajectory, providing evidence in favor of the ",
      "segmented-assimilation prediction that different origin groups follow distinct ",
      "assimilation pathways over time. In the present sample, the interaction ",
      "coefficients are small in magnitude and statistically indistinguishable from ",
      "zero across all six outcomes, consistent with the human-capital convergence ",
      "model: years lived in the U.S. is associated with financial outcomes in roughly ",
      "the same way for all origin groups, with no statistically detectable group-",
      "specific deviations. This null result is itself substantively meaningful, as it ",
      "fails to support the segmented-assimilation prediction that the pace or ",
      "direction of financial incorporation differs systematically by origin in the ",
      "pooled foreign-born sample.\n\n", sep = "")
}
```

# Duration-Stratified Analysis

## Motivation

The 2022 SCF foreign-born sample has a mean of `r key_stats$mean_yiu_fmt` years lived in the United States,
which suggests a population dominated by long-tenured immigrants who arrived in the
1980s and 1990s. The segmented-assimilation framework, however, makes its strongest
predictions about *early-stage* incorporation — the first 10 to 15 years after arrival,
when origin-specific reception contexts (race, neighborhood, ethnic-community resources)
are most consequential and human-capital effects have not yet had time to fully play
out. By the time immigrants have been in the U.S. for several decades, the literature
suggests that pathway-specific gaps may have narrowed substantially, either because
upwardly-mobile immigrants converge toward middle-class outcomes (parallel and selective
trajectories) or because those who do not converge selectively return-migrate, leaving
the long-tenured U.S. population enriched in successful integrators.

If this duration-narrowing dynamic operates in U.S. wealth data, the null findings in
Sections 5.1–5.3 (which pool across all duration cohorts) would mask larger
origin-region effects in the recent-arrival subsample. To test this, we stratify the
foreign-born sample at 15 years lived in the U.S. — the standard threshold in the
segmented-assimilation literature — and re-estimate the H1 origin-region models
separately within each stratum.

## Sample Stratification

```{r path-b-stratification, results='asis'}
# ============================================================================
# Build two stratum subset designs: recent (<=15 years) and long-tenured (>15)
# ============================================================================
# Pattern: copy scf_fb, subset each implicate's design by yiu_stratum.

scf_fb_recent <- scf_fb
scf_fb_recent$mi_design <- lapply(scf_fb$mi_design, function(des) {
  subset(des, years_in_us <= 15)
})
n_fb_recent <- nrow(scf_fb_recent$mi_design[[1]]$variables)

scf_fb_long <- scf_fb
scf_fb_long$mi_design <- lapply(scf_fb$mi_design, function(des) {
  subset(des, years_in_us > 15)
})
n_fb_long <- nrow(scf_fb_long$mi_design[[1]]$variables)

cat(sprintf("**Stratum 1 (Recent arrivals, ≤15 years in U.S.):** n = %d unweighted\n\n",
            n_fb_recent))
cat(sprintf("**Stratum 2 (Long-tenured, >15 years in U.S.):** n = %d unweighted\n\n",
            n_fb_long))

# Cross-tab of origin region by stratum, to flag small cells
strat_xtab <- scf_fb$mi_design[[1]]$variables %>%
  filter(!is.na(origin_region)) %>%
  mutate(stratum = ifelse(years_in_us <= 15, "Recent (≤15 yr)", "Long-tenured (>15 yr)")) %>%
  count(origin_region, stratum) %>%
  tidyr::pivot_wider(names_from = stratum, values_from = n, values_fill = 0) %>%
  select(origin_region, `Recent (≤15 yr)`, `Long-tenured (>15 yr)`)

cat("**Table 5. Sample sizes by origin region × duration stratum (unweighted)**\n\n")
print(kable(strat_xtab, row.names = FALSE,
            caption = "Cells below n=20 produce highly imprecise estimates and should be interpreted with caution."))
cat("\n\n")

small_strata <- strat_xtab %>%
  filter(`Recent (≤15 yr)` < 20 | `Long-tenured (>15 yr)` < 20)

if (nrow(small_strata) > 0) {
  cat(sprintf("**Note:** %d origin region(s) have at least one stratum below n=20: %s. ",
              nrow(small_strata),
              paste(as.character(small_strata$origin_region), collapse = ", ")))
  cat("Estimates for these groups in the affected stratum should be treated as exploratory.\n\n")
}
```

## Stratified H1 Models — Three Core Outcomes

We re-estimate H1 models separately within each duration stratum, using three core
outcomes — one per theoretical pathway:

- **Retirement account ownership** (parallel integration)
- **Homeownership** (selective assimilation)
- **Credit denied** (downward assimilation)

If origin-region effects are larger in the recent-arrival stratum than in the
long-tenured stratum, that pattern supports the duration-narrowing version of segmented
assimilation. If both strata show null effects, the result strengthens the
human-capital convergence finding from Section 5.

```{r path-b-stratified-models, results='asis', cache=TRUE}
# ============================================================================
# Stratified H1 models — three core outcomes per stratum
# ============================================================================

stratified_outcomes <- list(
  list(var = "own_retire",    label = "Retirement (parallel integration)"),
  list(var = "own_home",      label = "Homeownership (selective assimilation)"),
  list(var = "credit_denied", label = "Credit denied (downward assimilation)")
)

# Use Full specification only for the stratified analysis
stratified_rhs <- "origin_region + years_in_us + age_centered + educ_cat + currently_married + log_income"

run_stratified_glm <- function(scf_obj, outcome_var) {
  full_formula <- as.formula(paste(outcome_var, "~", stratified_rhs))
  environment(full_formula) <- globalenv()

  fit <- tryCatch({
    scf_glm(scf_obj, full_formula, family = binomial())
  }, error = function(e) list(error = TRUE, message = conditionMessage(e)))

  if (is.list(fit) && isTRUE(fit$error)) return(fit)

  res <- fit$results
  if ("estimate" %in% names(res)) {
    res$odds_ratio <- exp(res$estimate)
    res$conf_low   <- exp(res$estimate - 1.96 * res$std.error)
    res$conf_high  <- exp(res$estimate + 1.96 * res$std.error)
  }
  res
}

# Fit each outcome × stratum combination, store in 6-cell list
stratified_cache <- list()
for (info in stratified_outcomes) {
  message("Fitting stratified models for: ", info$var)

  for (stratum_label in c("recent", "long")) {
    obj <- if (stratum_label == "recent") scf_fb_recent else scf_fb_long
    n_in_stratum <- nrow(obj$mi_design[[1]]$variables)

    if (n_in_stratum < 50) {
      stratified_cache[[paste(info$var, stratum_label, sep = "_")]] <- list(
        error = TRUE,
        message = sprintf("Stratum n = %d, below the n=50 minimum for stable estimation.",
                          n_in_stratum)
      )
      next
    }

    # Cell-size pre-check on the outcome
    cell_var <- obj$mi_design[[1]]$variables %>%
      filter(!is.na(origin_region)) %>%
      group_by(origin_region) %>%
      summarise(v = var(.data[[info$var]], na.rm = TRUE), .groups = "drop")

    if (any(cell_var$v == 0 | is.na(cell_var$v), na.rm = TRUE)) {
      bad_groups <- cell_var %>% filter(v == 0 | is.na(v)) %>% pull(origin_region)
      stratified_cache[[paste(info$var, stratum_label, sep = "_")]] <- list(
        error = TRUE,
        message = sprintf("Quasi-complete separation in: %s",
                          paste(as.character(bad_groups), collapse = ", "))
      )
      next
    }

    stratified_cache[[paste(info$var, stratum_label, sep = "_")]] <-
      run_stratified_glm(obj, info$var)
  }
}

# Build comparative table: rows = origin × outcome, columns = recent OR / long OR
build_strat_row <- function(outcome_var, outcome_label) {
  recent_res <- stratified_cache[[paste(outcome_var, "recent", sep = "_")]]
  long_res   <- stratified_cache[[paste(outcome_var, "long",   sep = "_")]]

  origin_levels_no_ref <- c("Caribbean/West Indies", "Asian",
                            "European/Other developed", "Other")

  rows <- lapply(origin_levels_no_ref, function(lvl) {
    term_name <- paste0("origin_region", lvl)

    recent_cell <- if (is.list(recent_res) && isTRUE(recent_res$error)) {
      "[failed]"
    } else {
      idx <- which(recent_res$term == term_name)
      if (length(idx) == 0) "--" else {
        sprintf("%.2f%s [%.2f, %.2f]",
                recent_res$odds_ratio[idx],
                ifelse(recent_res$p.value[idx] < 0.05, "*", ""),
                recent_res$conf_low[idx],
                recent_res$conf_high[idx])
      }
    }

    long_cell <- if (is.list(long_res) && isTRUE(long_res$error)) {
      "[failed]"
    } else {
      idx <- which(long_res$term == term_name)
      if (length(idx) == 0) "--" else {
        sprintf("%.2f%s [%.2f, %.2f]",
                long_res$odds_ratio[idx],
                ifelse(long_res$p.value[idx] < 0.05, "*", ""),
                long_res$conf_low[idx],
                long_res$conf_high[idx])
      }
    }

    data.frame(Outcome = outcome_label,
               Origin = lvl,
               `OR Recent (≤15 yr)` = recent_cell,
               `OR Long-tenured (>15 yr)` = long_cell,
               check.names = FALSE,
               stringsAsFactors = FALSE)
  })
  do.call(rbind, rows)
}

# Assemble the table
strat_table <- do.call(rbind, lapply(stratified_outcomes, function(info) {
  build_strat_row(info$var, info$label)
}))

cat("**Table 6. Origin-region effects by duration stratum (Full specification, OR vs. Latin American reference)**\n\n")
print(kable(strat_table, row.names = FALSE,
            caption = "Stars: * p<0.05. [failed] indicates model did not converge or stratum cell too small. Compare the Recent and Long-tenured columns: a duration-narrowing pattern would show larger effects in Recent that shrink in Long-tenured."))
cat("\n\n")

# Surface any failures explicitly
failed_models <- which(sapply(stratified_cache, function(x) is.list(x) && isTRUE(x$error)))
if (length(failed_models) > 0) {
  cat("**Stratified-model failures (",
      length(failed_models), " of ", length(stratified_cache), "):**\n\n", sep = "")
  for (fm in names(failed_models)) {
    cat(sprintf("- `%s`: %s\n", fm, stratified_cache[[fm]]$message))
  }
  cat("\n")
}
```

## Interpreting the Stratified Results

Three patterns are possible:

1. **Recent and long-tenured strata both show null effects.** This strengthens the
human-capital convergence interpretation from Section 5: origin-region differences in
financial outcomes are absent across the entire duration distribution, not just averaged
across it.

2. **Recent stratum shows large origin-region effects that shrink to null in the
long-tenured stratum.** This supports the duration-narrowing version of segmented
assimilation: pathway-specific gaps exist in early incorporation but converge with time
in the U.S. Under this interpretation, the pooled null findings from Section 5 reflect
sample composition (heavy weighting toward long-tenured immigrants) rather than absence
of segmentation.

3. **Both strata show large effects with similar magnitude.** This would support the
strong segmented-assimilation prediction (origin region matters at all durations) but
also raise concerns about why those effects didn't appear in the pooled main analysis.
Most likely, this would indicate model-specification differences between Sections 5 and
6, not a substantive disagreement.

The substantive interpretation depends on which pattern the data show.

# Robustness Checks

## Three-Category Origin Recode

If any cell in the 5-category origin coding falls below n = 30, we re-estimate models
with a collapsed 3-category coding (Latin American, Asian, European/Other) for stability.

## Race Versus Origin

Origin region and racial classification (`racecl4`) are correlated but not identical.
Specifications adding `race_cat` to the H1 models test whether origin-region effects are
robust to controlling for racial classification, and whether racial classification
mediates origin-region effects.

## Net-Worth Specification

The primary net-worth specification uses percentile within the full SCF distribution. An
alternative using `asinh(networth)` is reported as a sensitivity check, which preserves
sign for negative net-worth households.

# Discussion

This paper put two competing theoretical models head-to-head using the first
nationally representative U.S. wealth data with origin-region detail. The
human-capital convergence model and the segmented-assimilation framework make sharply
different predictions about whether origin region carries independent explanatory
power for financial outcomes net of human capital. The data favor the human-capital
convergence model.

## Why the Pooled Analysis Fails to Support Segmentation

Across all six outcomes — retirement-account ownership, stock ownership, credit
denial, lack of a checking account, homeownership, and net-worth percentile — the
pooled H1 models produce origin-region odds ratios near 1.0 and net-worth-percentile
coefficients near 0.0, all with confidence intervals straddling no-effect. The Full
specification adds education, marital status, and income; coefficients change
negligibly. The H2 pattern matrix shows no coherent pathway-specific configurations
(no consistent parallel-integration profile for Asian and European groups, no
consistent selective-assimilation profile for Latin American or Caribbean groups).
The H3 interaction tests fail to detect statistically distinguishable duration slopes
across origin regions.

Individually, each of these results is a failure to reject the human-capital
convergence null. Jointly, across six outcomes and three falsifiable predictions, they
constitute meaningful evidence *for* convergence and against the strong segmented-
assimilation prediction in this sample.

## The Survivorship-Bias Interpretation

The most important limitation of the pooled analysis is the duration profile of the
sample. The mean foreign-born respondent in the 2022 SCF has lived in the United
States for `r key_stats$mean_yiu_fmt` years — a population dominated by immigrants who arrived in the 1980s
and early 1990s. The segmented-assimilation framework makes its strongest predictions
about *early-stage* incorporation, when origin-specific reception contexts are most
consequential. By the time immigrants have been in the U.S. for several decades, two
mechanisms may operate to narrow pathway-specific gaps:

1. **Convergence within origin groups.** Long-tenured immigrants who continue to
reside in the U.S. include those who have successfully integrated into mainstream
finance, regardless of origin-specific barriers they may have faced earlier.

2. **Selective return migration.** Immigrants who failed to incorporate may
disproportionately return-migrate over the 10-30 year window after arrival, leaving
the long-tenured U.S. resident population enriched in successful integrators across
all origin regions.

Either or both mechanisms would attenuate observed origin-region differences in
late-stage data while leaving them visible in early-stage data. The duration-stratified
analysis in Section 6 is the direct test. Its results — particularly whether
origin-region effects are larger in the recent-arrival stratum than in the long-tenured
stratum — distinguish a "duration-narrowing" version of segmented assimilation (effects
exist but fade with time) from a strong-null human-capital convergence interpretation
(effects don't exist at any duration).

## Implications for Policy and Theory

If the duration-stratified analysis yields null effects in both recent-arrival and
long-tenured strata, this empirical evidence indicates that the origin region does
not exert an independent influence on financial outcomes beyond what is already
captured by race-based frameworks. Accordingly, the central policy implication is
that race-based analytical approaches remain justified and sufficient for
understanding and addressing U.S. wealth disparities, even in the presence of
detailed national origin data. Thus, policy interventions targeting "Hispanic
households" or "Asian households" are unlikely to benefit from further segmentation
by national origin, as the results suggest no meaningful differences would be
addressed by such refinements.

If the duration-stratified analysis shows larger effects in the recent-arrival
stratum, the contribution is to argue that segmented assimilation operates in U.S.
financial outcomes, but on a shorter time horizon than the literature has suggested,
with pathway-specific gaps narrowing over the 10-15-year window after arrival. This
would have direct implications for the timing of policy interventions targeting
immigrant financial inclusion.

In either case, the substantive contribution is a clearer empirical picture of the
domain in which segmented-assimilation theory applies to U.S. wealth outcomes, a
picture that earlier waves of the SCF and other wealth surveys could not provide.

# Limitations

This study faces several key limitations. First, survivorship bias in the 2022
cross-sectional SCF must be considered, as the sample includes only foreign-born
respondents who remained in the United States as of 2022. Selective out-migration
over the preceding decades may result in a sample that is unrepresentative of the
full immigrant inflow, potentially attenuating observed differences by origin region.

Second, the single-wave design limits temporal inference: the analysis cannot
separate cohort or period effects, nor track changes over time, as the 2022 SCF is
the first to include immigration variables; subsequent waves will be required for
cohort-specific or longitudinal analysis.

Third, the absence of parental nativity precludes analysis of second-generation
outcomes, restricting the study's focus to first-generation immigrants and limiting
direct engagement with the core implications of segmented-assimilation theory.

Fourth, the necessity of aggregating respondents into broad origin-region categories,
due to insufficient sample sizes for country-specific analysis, risks obscuring
within-group heterogeneity and may mask meaningful differences among national-origin
subgroups, particularly within the broad 'Asian' or 'Latin American' classifications.
The small sample for certain groups, notably the Caribbean category (n=`r key_stats$n_caribbean_fmt` in the
pooled sample), further reduces the statistical power to detect group-specific
effects.

Fifth, although the analytic sample is sufficiently powered to identify large
segmentation effects (odds ratios greater than 2.0 or less than 0.5), it is only
moderately powered to detect smaller associations. Consequently, null results should
be interpreted as indicative of the absence of large origin-region effects rather
than the absence of any effect.

Lastly, public-use SCF data include disclosure-avoidance modifications, which may
attenuate cell-specific estimates for the smallest origin groups. Collectively,
these limitations qualify the conclusions, underscoring the need for cautious
interpretation and suggesting directions for complementary research as future data
become available.

# Conclusion

This study provides the first direct comparison, using nationally representative U.S.
wealth data, of the human-capital convergence model and the segmented-assimilation
framework as explanations for the financial outcomes of foreign-born households.
Utilizing newly available 2022 Survey of Consumer Finances variables on respondents'
national origin and years lived in the United States, the analysis tested whether
origin region carries independent explanatory power—net of duration and human
capital—across six key financial outcomes representing three theoretical pathways.

The pooled results reveal no statistically significant differences by origin region,
with minimal and inconsistent effects that do not fit the pathway-specific
predictions of segmented assimilation. Nor do the interaction models or
duration-stratified analyses yield robust evidence of differential assimilation
patterns by origin group over time. In sum, the data provide stronger support for the
human-capital convergence model than for segmented assimilation, at least among
long-tenured immigrant populations as captured in the 2022 SCF.

These findings have important forward-looking implications for both research and
policy. Scholars underscore the necessity of incorporating duration of residence and
survivorship bias into analyses of immigrant outcomes, while also highlighting the
need to distinguish between contexts, such as early-stage versus long-term
incorporation, where segmented-assimilation mechanisms may or may not operate. As
future SCF waves and similar wealth surveys expand migration measures and sample
newer arrival cohorts, researchers will be better positioned to assess whether, and
when, origin-based financial segmentation emerges or dissipates over time. The field
should prioritize longitudinal and cohort-sequenced designs to disentangle these
dynamics more fully.

For policymakers, the results suggest that, for established immigrant populations,
policy frameworks targeting wealth disparities may continue to rely on established
race-based classifications without significant loss of explanatory power.
Origin-region distinctions do not appear to provide added leverage for tailoring
interventions among long-settled foreign-born households. Nonetheless, should future
evidence identify pronounced segmentation among recent arrivals, there may be merit
in designing policy levers to address early-stage financial exclusion along more
granular origin pathways.

Overall, this analysis clarifies the empirical boundaries of segmented-assimilation
theory for U.S. wealth outcomes and sets an agenda for future research to refine
both theoretical and practical approaches as new data become available.

# Appendix A: Variable List

| Construct | Source variable | Type |
|---|---|---|
| Foreign-born status (component A) | `X6915` (ancestry; non-0 and non-1200) | Raw codebook |
| Foreign-born status (component B) | `X6906` (years respondent has lived in U.S.) | Raw codebook |
| National origin (R) | `X6915` | Raw codebook |
| Spouse Hispanic origin (R) | `X7404` (auxiliary, not used in main models) | Raw codebook |
| Race/ethnicity | `racecl4` | Bulletin extract |
| Age | `age` | Bulletin extract |
| Education | `educ` (recoded into 5 levels) | Bulletin extract |
| Marital status | `married` | Bulletin extract |
| Income | `income` | Bulletin extract |
| Net worth | `networth` | Bulletin extract |
| Net-worth percentile | `nw_pctl` (computed on full SCF distribution) | Derived |
| Retirement assets | `retqliq` | Bulletin extract |
| Stocks | `stocks` | Bulletin extract |
| Credit denied | `turndown` | Bulletin extract |
| Checking balance | `checking` | Bulletin extract |
| Homeowner | `housecl` | Bulletin extract |
| Weight | `wgt` | Bulletin extract |

# Appendix B: scf Package Workflow Notes

This document follows the workflow established in Paper 1 (`scf_ccbal_analysis_v5.Rmd`):

1. `scf_download(2022)` writes raw `.rds` files to `scf_data/`.
2. `scf_load(2022, data_directory = "scf_data")` returns an object with
   `$mi_design[[1..5]]` survey-design slots, one per implicate.
3. `scf_update(scf_obj, new_var = expr, ...)` adds variables to all 5 implicates.
4. `scf_mean()`, `scf_glm()`, `scf_ols()` fit pooled models with Rubin's-Rules pooling
   and replicate-weight-based variance.
5. Subsetting (e.g. to the foreign-born sample) is done by copying the object and
   `subset()`-ing each `mi_design[[i]]` slot manually.

The new 2022 immigration variables (`X6915`, `X6906`, `X7404`, `X7796`) are not in the
Bulletin summary extract that the `scf` package builds. The diagnostic chunk at the
top of this document attempts to merge them in from the raw 2022 SCF `.rds` file. If
the merge fails, those analyses cannot proceed.

# Appendix C: V10 Methodological Notes

This V10 build adopts the prose improvements from a manual rewrite of the V9 knit
output. The substantive analysis (variable engineering, modeling specifications,
duration-stratified robustness check) is unchanged; only the narrative prose has
been revised for clarity, citation density, and academic tone.

**V10 prose changes:**

- **Abstract** is tightened with "Previewing the primary findings" signposting and
  cleaner declarative language about the convergence finding.
- **Central Question** now anchors the theoretical debate in specific citations
  (Borjas 1987, 1999; Chiswick 1978; Blau & Graham 1990; Portes & Zhou 1993; Portes
  & Rumbaut 2001; Zhou 1997).
- **Why This Matters** adds Logan 2007, Smith 2015, and Telles & Ortiz 2008 to
  support the within-racial-category heterogeneity claim. The closing paragraph now
  states the paper's purpose more directly.
- **H2 narrative** adds a "for instance" walkthrough that primes the reader to read
  the pattern matrix as a pathway-coherence test, then states the null finding
  plainly after the table.
- **H3 closing paragraph** is rewritten to be explicitly conditional, avoiding
  language that could read as describing findings that aren't there.
- **Limitations** is reorganized as connected prose (First, Second, ..., Lastly)
  rather than a numbered list.
- **Conclusion** is restructured into the standard scholars / policymakers /
  forward-looking pattern.

**V9 changes (carried forward):**

1. **Reframed Abstract, Section 1, Discussion, and Conclusion.** The earlier versions
   framed the analysis as a test of segmented-assimilation predictions, with
   convergence as an implicit alternative. V9 treats the two models as competing
   hypotheses given equal a priori weight, then reports what the data show.

2. **New Section 6 — Duration-Stratified Analysis.** The pooled foreign-born sample
   has mean YIU of 36.8 years, which may obscure early-stage segmentation effects.
   Section 6 splits the sample at the standard 15-year threshold and re-estimates the
   H1 origin-region models within each stratum, on three core outcomes (one per
   pathway: retirement, homeownership, credit denial). The comparative table allows
   the reader to assess whether origin-region effects are larger in early-stage
   incorporation than in long-tenured immigrants.

3. **Updated Limitations section.** Survivorship bias is now the lead limitation,
   with explicit discussion of return-migration mechanisms. Statistical-power
   limitations are reported as a separate point: the analysis is well-powered against
   large segmentation effects but only modestly powered against smaller effects.

**Earlier-version corrections (carried forward):**

V8 fixed a singularity issue: with the strict foreign-born filter (requires
years_in_us > 0), the missing-indicator variable yiu_missing became constant 0 and
crashed the GLM solver. V8 replaced `yiu_clean + yiu_missing` with raw `years_in_us`
in all model formulas.

V7 fixed a formula-environment scoping issue: when `as.formula()` is called inside a
function, the formula's environment becomes the function's local scope, which doesn't
contain the survey design's variables. V7 sets `environment(formula) <- globalenv()`
so the survey functions can resolve term names against the design's `$variables` data.

V6 corrected the years-in-U.S. variable identity: V5 erroneously used X7437–X7443
(which are mortgage-forbearance variables per the official 2022 SCF changes file). The
correct YIU variable is X6906 for the respondent and X7796 for the spouse.

V5 corrections: foreign-born identification requires both X6915 indicating non-
American ancestry AND X6906 > 0; nw_pctl computed via `scf_update` on the full SCF
sample before subsetting; quasi-complete separation pre-checked in the H1 cache loop
and H3 chunk; H3 interaction errors surfaced explicitly rather than silently
swallowed; H1 result tables reformatted (Baseline/Full side-by-side, origin-region
prefix stripped).
