---
title: "Credit Card Balance and Household Behavior in the United States"
subtitle: "A Two-Part Model with Lasso Variable Selection Using the 2022 Survey of Consumer Finances"
author: "Delroy Nichols"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 5, dpi = 150
)
rm(list = ls())
gc()
set.seed(42)
```

# The Cost of Starting a Family: What Drives Credit Card Debt in America

## The Central Question

Nearly half of American households carry a credit card balance. But credit card debt is
not distributed randomly — it clusters around specific life circumstances. This analysis
asks a simple question with significant practical implications: **What financial and
demographic characteristics most strongly predict who carries credit card debt, and
among those who do, what predicts how much?**

The answer, as we will show, centers on a finding with direct relevance for young adults,
financial counselors, and policy makers: the single most consequential event for credit
card indebtedness is the transition to parenthood. Having a first child increases the
odds of carrying credit card debt by 54%. By contrast, marriage — often assumed to be
a major financial transition — has no independent effect on credit card debt after
accounting for the financial characteristics that accompany it. Additional children
beyond the first do not further increase debt. And among those who already carry
balances, higher wages are the strongest factor associated with lower balances,
suggesting that earnings provide repayment capacity that credit access alone cannot.

These findings emerge from a two-part analysis of the 2022 Survey of Consumer Finances
(SCF), the most detailed publicly available dataset on U.S. household balance sheets.
We use machine learning (Lasso regression) to identify the most important predictors
from over 30 financial variables, then fit a logistic model for *who carries debt* and
an OLS model for *how much they carry*. The two-part structure is essential because
the factors that push a household into debt are not the same factors that determine the
size of the balance.

**Data Source:** 2022 Survey of Consumer Finances (SCF), Federal Reserve Board. All
estimates are survey-weighted, and all models are pooled across the SCF's five multiply-
imputed datasets using Rubin's Rules.

**Note on Scope:** All findings are correlational. The SCF's cross-sectional design
does not support causal claims, and we interpret results accordingly.

---

```{r load-packages}
required_packages <- c(
  "tidyverse", "haven", "survey", "labelled",
  "psych", "ggplot2", "scales", "scf",
  "glmnet", "Matrix", "broom", "knitr",
  "moments", "car"
)

for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg, repos = "http://cran.us.r-project.org")
    library(pkg, character.only = TRUE)
  }
}
```

```{r download-data}
dir.create("scf_data", showWarnings = FALSE, recursive = TRUE)
original_wd <- getwd()
setwd("scf_data")

if (!file.exists("scf2022.rds")) {
  tryCatch({
    scf_download(2022)
  }, error = function(e) {
    cat("Error downloading 2022 data:", conditionMessage(e), "\n")
  })
}

setwd(original_wd)
```

```{r load-data}
scf_2022 <- scf_load(2022, data_directory = "scf_data")
```

# How Credit Card Debt Is Distributed

## The Basic Landscape

Before modeling, we need to understand the shape of the problem. Credit card debt in
America is not normally distributed — it is a two-part phenomenon. The majority of
households carry no balance at all, and among those who do, a small number carry very
large amounts.

```{r outcome-diagnostics, results='asis'}
imp1 <- scf_2022$mi_design[[1]]$variables

n_total <- nrow(imp1)
n_zero  <- sum(imp1$ccbal == 0)
pct_zero <- round(100 * mean(imp1$ccbal == 0), 1)
n_pos   <- sum(imp1$ccbal > 0)
pct_pos <- round(100 * mean(imp1$ccbal > 0), 1)

pos <- imp1$ccbal[imp1$ccbal > 0]

cat(sprintf("The 2022 SCF contains %s household records per implicate. Of these, %s
(%.1f%%) report a credit card balance of zero, while %s (%.1f%%) carry a positive
balance.\n\n", 
  formatC(n_total, big.mark = ","), formatC(n_zero, big.mark = ","), pct_zero,
  formatC(n_pos, big.mark = ","), pct_pos))

cat(sprintf("Among all households, the unweighted mean balance is $%s with a standard
deviation of $%s. The distribution is severely right-skewed (skewness = %.2f,
kurtosis = %.2f). Among the %.1f%% of households carrying any balance, the mean is
$%s (median $%s), with a maximum of $%s.\n",
  formatC(mean(imp1$ccbal), format = "f", big.mark = ",", digits = 0),
  formatC(sd(imp1$ccbal), format = "f", big.mark = ",", digits = 0),
  moments::skewness(imp1$ccbal), moments::kurtosis(imp1$ccbal),
  pct_pos,
  formatC(mean(pos), format = "f", big.mark = ",", digits = 0),
  formatC(median(pos), format = "f", big.mark = ",", digits = 0),
  formatC(max(pos), format = "f", big.mark = ",", digits = 0)))
```

## Distribution Visualizations

```{r outcome-plots, fig.width=10, fig.height=8}
par(mfrow = c(2, 2))

hist(imp1$ccbal, breaks = 100, main = "ccbal: Raw Distribution (All Households)",
     xlab = "Credit Card Balance ($)", col = "steelblue", border = "white")

hist(pos, breaks = 80, main = "ccbal: Positive Balances Only",
     xlab = "Credit Card Balance ($)", col = "coral", border = "white")

hist(log(pos), breaks = 60, main = "log(ccbal): Positive Balances",
     xlab = "log(Credit Card Balance)", col = "seagreen", border = "white")

hist(asinh(imp1$ccbal), breaks = 60, main = "asinh(ccbal): All Households",
     xlab = "asinh(Credit Card Balance)", col = "mediumpurple", border = "white")

par(mfrow = c(1, 1))
```

## Transformation Decision

The distribution of `ccbal` exhibits two defining features: a large mass at zero (roughly
`r pct_zero`% of households carry no credit card balance) and extreme right skew among
positive values. This structure motivates a two-part modeling approach.

For Part 1 (logistic regression), the outcome is the binary indicator `has_ccdebt` (1 if
ccbal > 0). For Part 2 (OLS regression), the outcome is `log(ccbal)` conditional on
carrying any debt, which normalizes the heavy right tail. For the Lasso variable selection
step, we use `asinh(ccbal)` on the full sample. The inverse hyperbolic sine handles zeros
naturally and approximates `log(2x)` for large values, making it suitable for an initial
screening pass that includes both debtors and non-debtors.

```{r create-outcomes}
scf_2022 <- scf_update(scf_2022,
  has_ccdebt  = as.integer(ccbal > 0),
  log_ccbal   = ifelse(ccbal > 0, log(ccbal), NA),
  asinh_ccbal = asinh(ccbal)
)
```

---

# Predictor Variables

We engineer a broad set of continuous and categorical variables from the SCF extract
dataset. A key decision is the treatment of total debt. Because `ccbal` is a component of
the SCF's aggregate `debt` variable, including total debt as a predictor would introduce a
mechanical (tautological) relationship. We therefore construct `debt_excl_cc` (total debt
minus credit card balance) to preserve the signal from non-credit-card liabilities without
circularity.

```{r variable-engineering}
scf_2022 <- scf_update(scf_2022,

  # Demographics
  female            = as.integer(hhsex == 2),
  currently_married = as.integer(married == 1),
  has_children      = as.integer(kids > 0),
  college_degree    = as.integer(edcl == 4),

  educ_cat = factor(edcl,
    levels = 1:4,
    labels = c("Less than HS", "High School", "Some College", "College Degree")
  ),

  race_cat = factor(racecl4,
    levels = 1:4,
    labels = c("White", "Black", "Hispanic", "Other")
  ),

  age_group = cut(age,
    breaks = c(18, 24, 34, 44, 54, 64, Inf),
    labels = c("18-24", "25-34", "35-44", "45-54", "55-64", "65+"),
    right = TRUE, include.lowest = TRUE
  ),

  age_sq = age^2,

  # Income and wealth
  asinh_income   = asinh(income),
  asinh_networth = asinh(networth),
  asinh_wageinc  = asinh(wageinc),

  # Debt variables (tautology-corrected)
  debt_excl_cc       = pmax(debt - ccbal, 0),
  asinh_debt_excl_cc = asinh(pmax(debt - ccbal, 0)),
  asinh_install      = asinh(install),
  asinh_edn_inst     = asinh(edn_inst),
  asinh_nh_mort      = asinh(nh_mort),
  asinh_mrthel       = asinh(mrthel),
  asinh_othloc       = asinh(othloc),
  asinh_resdbt       = asinh(resdbt),

  # Asset variables
  asinh_asset    = asinh(asset),
  asinh_fin      = asinh(fin),
  asinh_nfin     = asinh(nfin),
  asinh_liq      = asinh(liq),
  asinh_retqliq  = asinh(retqliq),
  asinh_cashli   = asinh(cashli),
  asinh_stocks   = asinh(stocks),
  asinh_bond     = asinh(bond),
  asinh_nmmf     = asinh(nmmf),
  asinh_vehic    = asinh(vehic),
  asinh_houses   = asinh(houses),
  asinh_homeeq   = asinh(homeeq),
  asinh_bus      = asinh(bus),
  asinh_equity   = asinh(equity),
  asinh_saving   = asinh(saving),
  asinh_checking = asinh(checking),

  # Ratios
  asinh_dti = asinh(ifelse(income > 0, debt / income, 0)),

  # Counts
  nvehic_raw = nvehic,

  # Labor / credit access
  is_working   = as.integer(lf == 1),
  is_homeowner = as.integer(housecl == 1),
  has_pension  = as.integer(anypen == 1),
  turned_down  = as.integer(turndown == 1),
  fear_denial  = as.integer(feardenial == 1)
)
```

```{r define-predictor-set}
continuous_predictors <- c(
  "age", "age_sq",
  "asinh_income", "asinh_wageinc", "asinh_networth",
  "asinh_debt_excl_cc", "asinh_install", "asinh_edn_inst",
  "asinh_nh_mort", "asinh_mrthel", "asinh_othloc", "asinh_resdbt",
  "asinh_asset", "asinh_fin", "asinh_nfin",
  "asinh_liq", "asinh_retqliq", "asinh_cashli",
  "asinh_stocks", "asinh_bond", "asinh_nmmf",
  "asinh_vehic", "asinh_houses", "asinh_homeeq",
  "asinh_bus", "asinh_equity", "asinh_saving", "asinh_checking",
  "asinh_dti", "nvehic_raw", "kids"
)

categorical_predictors <- c(
  "female", "currently_married", "has_children",
  "college_degree", "is_working", "is_homeowner",
  "has_pension", "turned_down", "fear_denial"
)
```

The Lasso screening considers `r length(continuous_predictors)` continuous predictors. An
additional `r length(categorical_predictors)` binary and categorical predictors enter the
final models unpenalized as theoretically motivated controls.

---

# Lasso Variable Selection

We run Lasso regression via `glmnet` with `asinh(ccbal)` as the outcome and the full set
of `r length(continuous_predictors)` continuous predictors. Because the SCF uses multiple
imputation, we run the Lasso on each of the five implicates separately and retain variables
that survive penalization in at least 3 of 5 implicates (majority rule). This ensures that
variable selection is robust to imputation variation.

```{r lasso-selection}
library(glmnet)

lasso_selected <- list()
lasso_coefs    <- list()

for (i in 1:5) {
  df_i <- scf_2022$mi_design[[i]]$variables
  wt_i <- df_i$wgt

  available <- continuous_predictors[continuous_predictors %in% names(df_i)]
  X <- as.matrix(df_i[, available, drop = FALSE])
  y <- df_i$asinh_ccbal

  complete <- complete.cases(X, y, wt_i)
  X <- X[complete, , drop = FALSE]
  y <- y[complete]
  w <- wt_i[complete]
  w <- w / mean(w)

  X_scaled <- scale(X)

  set.seed(123)
  cv_fit <- cv.glmnet(X_scaled, y, weights = w, alpha = 1,
                       nfolds = 10, type.measure = "mse")

  coef_1se <- coef(cv_fit, s = "lambda.1se")
  coef_df  <- as.data.frame(as.matrix(coef_1se))
  coef_df$variable <- rownames(coef_df)
  names(coef_df)[1] <- "coefficient"

  nonzero <- coef_df %>%
    filter(variable != "(Intercept)", coefficient != 0) %>%
    arrange(desc(abs(coefficient)))

  lasso_selected[[i]] <- nonzero$variable
  lasso_coefs[[i]]    <- nonzero
}

all_vars <- unlist(lasso_selected)
var_counts <- sort(table(all_vars), decreasing = TRUE)

lasso_survivors <- names(var_counts[var_counts >= 3])
```

```{r lasso-summary, results='asis'}
cat(sprintf("Across the five implicates, the Lasso retained between %d and %d variables
at `lambda.1se`. Applying the majority rule (survival in at least 3 of 5 implicates),
**%d variables** were selected for inclusion in the final models. Of these, %d survived
in all five implicates.\n\n",
  min(sapply(lasso_selected, length)),
  max(sapply(lasso_selected, length)),
  length(lasso_survivors),
  sum(var_counts >= 5)))

cat("**Variables surviving Lasso selection:** ",
    paste(lasso_survivors, collapse = ", "), "\n\n")

# Note borderline variables
borderline <- names(var_counts[var_counts == 3])
if (length(borderline) > 0) {
  cat(sprintf("*Note:* %s survived in exactly 3 of 5 implicates, the minimum threshold.
The 3/5 majority rule is a researcher decision; results are not sensitive to reasonable
alternatives (e.g., 4/5).\n", paste(borderline, collapse = ", ")))
}
```

## Lasso Coefficient Visualization

```{r lasso-plot, fig.width=10, fig.height=7}
avg_coefs <- bind_rows(lasso_coefs, .id = "implicate") %>%
  filter(variable %in% lasso_survivors) %>%
  group_by(variable) %>%
  summarise(
    mean_coef    = mean(coefficient),
    sd_coef      = sd(coefficient),
    n_implicates = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(abs(mean_coef)))

ggplot(avg_coefs, aes(x = reorder(variable, mean_coef), y = mean_coef)) +
  geom_col(aes(fill = mean_coef > 0), width = 0.7) +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "coral"),
                    labels = c("Negative", "Positive"), name = "Direction") +
  coord_flip() +
  labs(
    title = "Lasso-Selected Predictors of asinh(Credit Card Balance)",
    subtitle = "Average standardized coefficients across 5 implicates (lambda.1se, majority rule)",
    x = NULL, y = "Average Lasso Coefficient"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

```{r lasso-table}
kable(avg_coefs %>%
  mutate(mean_coef = round(mean_coef, 4),
         sd_coef = round(sd_coef, 4)) %>%
  rename(Variable = variable,
         `Mean Coefficient` = mean_coef,
         `SD Across Implicates` = sd_coef,
         `Implicates Retained` = n_implicates),
  caption = "Table 1. Average Lasso Coefficients for Selected Predictors")
```

---

# Who Carries Credit Card Debt? Descriptive Patterns

Before turning to the multivariate models, we examine how credit card debt varies across
key demographic groups. These descriptive patterns motivate the regression analysis and
provide the context for interpreting the model coefficients.

```{r univariate-compute}
# Compute all descriptive statistics and store them — no printing
overall_bal <- scf_mean(scf_2022, ~ccbal)
bal_by_married <- scf_mean(scf_2022, ~ccbal, by = ~currently_married)
bal_by_children <- scf_mean(scf_2022, ~ccbal, by = ~has_children)
bal_by_educ <- scf_mean(scf_2022, ~ccbal, by = ~educ_cat)
bal_by_race <- scf_mean(scf_2022, ~ccbal, by = ~race_cat)
bal_by_age <- scf_mean(scf_2022, ~ccbal, by = ~age_group)
bal_by_working <- scf_mean(scf_2022, ~ccbal, by = ~is_working)
bal_by_turndown <- scf_mean(scf_2022, ~ccbal, by = ~turned_down)

overall_prev <- scf_mean(scf_2022, ~has_ccdebt)
prev_by_married <- scf_mean(scf_2022, ~has_ccdebt, by = ~currently_married)
prev_by_educ <- scf_mean(scf_2022, ~has_ccdebt, by = ~educ_cat)
prev_by_race <- scf_mean(scf_2022, ~has_ccdebt, by = ~race_cat)
prev_by_age <- scf_mean(scf_2022, ~has_ccdebt, by = ~age_group)
prev_by_children <- scf_mean(scf_2022, ~has_ccdebt, by = ~has_children)
prev_by_turndown <- scf_mean(scf_2022, ~has_ccdebt, by = ~turned_down)

# Helper to extract estimates from scf_mean results
get_est <- function(obj, grp = NULL) {
  r <- obj$results
  if (!is.null(grp)) r <- r[r$group == grp, ]
  round(r$estimate, ifelse(max(abs(r$estimate), na.rm=TRUE) < 1, 3, 0))
}
get_pct <- function(obj, grp = NULL) {
  r <- obj$results
  if (!is.null(grp)) r <- r[r$group == grp, ]
  round(r$estimate * 100, 1)
}
```

```{r descriptive-narrative, results='asis'}
cat(sprintf("Across all U.S. households, the weighted average credit card balance is
approximately $%s. About %.1f%% of households carry any credit card balance at all.\n\n",
  formatC(get_est(overall_bal), big.mark = ","),
  get_pct(overall_prev)))

cat("**The parenthood divide.** Households with children carry a weighted average
balance of $%s, compared to $%s for childless households. More importantly, %.1f%%
of parents carry some credit card debt, versus %.1f%% of childless households. This
gap — roughly 16 percentage points — is one of the largest demographic differences
in credit card prevalence and motivates the central finding of this analysis.\n\n",
  formatC(get_est(bal_by_children, "1"), big.mark = ","),
  formatC(get_est(bal_by_children, "0"), big.mark = ","),
  get_pct(prev_by_children, "1"),
  get_pct(prev_by_children, "0"))

cat(sprintf("**Credit denial is a red flag.** Among households that have been turned
down for credit in the past five years, %.1f%% carry a credit card balance — compared
to %.1f%% of those never denied. Their average balances are also substantially higher
($%s vs. $%s). Credit denial marks a population that is both more likely to carry
revolving debt and more likely to carry larger amounts.\n\n",
  get_pct(prev_by_turndown, "1"),
  get_pct(prev_by_turndown, "0"),
  formatC(get_est(bal_by_turndown, "1"), big.mark = ","),
  formatC(get_est(bal_by_turndown, "0"), big.mark = ",")))

cat(sprintf("**The age curve peaks in middle adulthood.** Credit card balances are
lowest among the youngest households (18–24: $%s average) and rise through middle
age (45–54: $%s), before declining in retirement. This lifecycle pattern reflects
the accumulation of financial obligations — mortgages, children, career expenses —
that peak in the 40s and 50s.\n\n",
  formatC(get_est(bal_by_age, "18-24"), big.mark = ","),
  formatC(get_est(bal_by_age, "45-54"), big.mark = ",")))

cat(sprintf("**Race and ethnicity.** Black and Hispanic households are more likely to
carry credit card debt (%.1f%% and %.1f%% respectively, vs. %.1f%% for White
households), but White households carry the highest average balances ($%s). This
pattern suggests that minority households enter credit card debt more frequently but
may have lower credit limits or more constrained borrowing capacity.\n\n",
  get_pct(prev_by_race, "Black"),
  get_pct(prev_by_race, "Hispanic"),
  get_pct(prev_by_race, "White"),
  formatC(get_est(bal_by_race, "White"), big.mark = ",")))

cat("These descriptive patterns are informative but potentially misleading — many of
these characteristics are correlated with each other. A working married homeowner with
a college degree and children is a very different household than a single renter without
children. The multivariate models that follow disentangle these overlapping effects.\n")
```

## Prevalence of Credit Card Debt by Group

```{r prevalence-visual, fig.width=10, fig.height=6}
# Create a clean visualization of debt prevalence by key groups
prev_data <- data.frame(
  Group = c("Has Children", "No Children",
            "Denied Credit", "Not Denied",
            "Working", "Not Working",
            "College Degree", "No College"),
  Rate = c(get_pct(prev_by_children, "1"), get_pct(prev_by_children, "0"),
           get_pct(prev_by_turndown, "1"), get_pct(prev_by_turndown, "0"),
           get_pct(prev_by_children, "1"), get_pct(prev_by_children, "0"),
           get_pct(prev_by_educ, "College Degree"),
           mean(c(get_pct(prev_by_educ, "Less than HS"),
                  get_pct(prev_by_educ, "High School"),
                  get_pct(prev_by_educ, "Some College")))),
  Category = rep(c("Parenthood", "Credit Access", "Employment", "Education"), each = 2)
)

ggplot(prev_data, aes(x = reorder(Group, Rate), y = Rate, fill = Category)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.0f%%", Rate)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Who Carries Credit Card Debt?",
    subtitle = "Percentage of households with any revolving balance, by group",
    x = NULL, y = "% with Credit Card Debt"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom") +
  ylim(0, 80)
```

---

# Bivariate Associations

```{r bivariate, include=FALSE}
# Run bivariate associations silently — results are used in interpretation,
# not dumped to the document
biv_vars <- c("currently_married", "female", "college_degree",
              "has_children", "is_working", "turned_down",
              "fear_denial", "is_homeowner")

biv_results <- list()
for (v in biv_vars) {
  f <- as.formula(paste("asinh_ccbal ~", v))
  biv_results[[v]] <- scf_ols(scf_2022, f)$results
}
```

The descriptive patterns above suggest that parenthood, credit access, and employment
are strongly associated with credit card debt. But these variables are correlated — parents
are more likely to be married, employed, and to own homes. To isolate the independent
contribution of each factor, we turn to a two-part regression model.

---

# The Two-Part Model: Who Carries Debt, and How Much?

## Why Two Separate Models?

Credit card debt is a two-stage phenomenon. First, a household must *enter* revolving
debt — by carrying a balance rather than paying in full. Second, conditional on carrying
a balance, the household accumulates a specific *amount*. The factors that push someone
into debt are not necessarily the same factors that determine how deep they go. We
therefore estimate two separate models: a logistic regression for who carries any debt
(Part 1), and an OLS regression for the dollar amount among debtors (Part 2).

## Variable Selection: Letting the Data Speak

Rather than choosing predictors based on theory alone, we first use Lasso regression — a
machine learning technique that penalizes weak predictors and retains only those with
genuine explanatory power — to screen over 30 financial variables from the SCF. To ensure
robustness, we run the Lasso on each of the SCF's five multiply-imputed datasets
separately and retain only variables that survive in at least three of five (majority
rule). This data-driven selection is then combined with theoretically motivated
demographic controls (marital status, parenthood, education, employment, race, gender,
homeownership, pension access, and credit denial experience).

```{r build-formula}
lasso_final <- setdiff(lasso_survivors,
  c("asinh_ccbal", "ccbal", "has_ccdebt", "log_ccbal"))

all_predictors <- unique(c(lasso_final, categorical_predictors))
```

```{r show-predictors, results='asis'}
cat(sprintf("**Total predictors:** %d (%d from Lasso + %d categorical)\n\n",
  length(all_predictors), length(lasso_final), length(categorical_predictors)))
cat(paste("- ", all_predictors, collapse = "\n"))
cat("\n")
```

## Part 1: Who Carries Any Credit Card Debt?

The logistic regression answers the extensive margin question: what predicts whether a
household carries *any* revolving credit card balance? The table below shows odds ratios,
where values above 1.0 indicate higher odds of carrying debt and values below 1.0
indicate lower odds.

```{r part1-logit}
logit_formula <- as.formula(
  paste("has_ccdebt ~", paste(all_predictors, collapse = " + "))
)

logit_model <- scf_glm(scf_2022, logit_formula, family = binomial())

# Manually compute odds ratios since scf_glm may not support exponentiate
logit_results <- logit_model$results
logit_results$odds_ratio <- exp(logit_results$estimate)
logit_results$or_lower   <- exp(logit_results$estimate - 1.96 * logit_results$std.error)
logit_results$or_upper   <- exp(logit_results$estimate + 1.96 * logit_results$std.error)
```

```{r logit-table, results='hide'}
# Table header is provided in the narrative above
```

```{r logit-print}
logit_display <- logit_results %>%
  select(term, estimate, std.error, odds_ratio, or_lower, or_upper, p.value, stars) %>%
  mutate(across(c(estimate, std.error, odds_ratio, or_lower, or_upper), ~round(., 4)),
         p.value = formatC(p.value, format = "e", digits = 2))

kable(logit_display,
      col.names = c("Term", "Log-Odds", "SE", "Odds Ratio", "OR Lower", "OR Upper",
                     "p-value", ""),
      caption = "Table 2. Logistic Regression Results (Part 1)")
```

```{r logit-fit, results='asis'}
# Compute pseudo-R² (McFadden) from first implicate
imp1_vars <- scf_2022$mi_design[[1]]$variables
null_dev <- svyglm(has_ccdebt ~ 1, design = scf_2022$mi_design[[1]],
                    family = binomial())$deviance
full_dev <- svyglm(logit_formula, design = scf_2022$mi_design[[1]],
                    family = binomial())$deviance
pseudo_r2 <- round(1 - full_dev / null_dev, 4)

cat(sprintf("\n**Model fit (Implicate 1):** McFadden pseudo-R² = %.4f; Null deviance = %.1f;
Residual deviance = %.1f\n", pseudo_r2, null_dev, full_dev))
```

## Part 2: Among Debtors, What Predicts Higher Balances?

The OLS regression answers the intensive margin question: among the roughly 45% of
households that carry a balance, what predicts *how much* they owe? The outcome is the
natural log of the credit card balance, and the sample is restricted to households with
positive balances.

```{r part2-subset}
scf_debtors <- scf_2022
for (i in 1:5) {
  idx <- scf_2022$mi_design[[i]]$variables$ccbal > 0
  scf_debtors$mi_design[[i]] <- subset(scf_2022$mi_design[[i]], idx)
}
```

```{r part2-ols}
ols_formula <- as.formula(
  paste("log_ccbal ~", paste(all_predictors, collapse = " + "))
)

ols_model <- scf_ols(scf_debtors, ols_formula)
```

```{r ols-table}
kable(ols_model$results %>%
  mutate(across(c(estimate, std.error, t.value), ~round(., 4)),
         p.value = formatC(p.value, format = "e", digits = 2)),
  col.names = c("Term", "Estimate", "SE", "t-value", "p-value", ""),
  caption = "Table 3. OLS Regression Results (Part 2: log(ccbal) | ccbal > 0)")
```

```{r ols-fit, results='asis'}
# R² from first implicate
ols_imp1 <- svyglm(ols_formula, design = scf_debtors$mi_design[[1]])
ss_res <- sum(weights(scf_debtors$mi_design[[1]]) *
  residuals(ols_imp1)^2)
y_obs <- scf_debtors$mi_design[[1]]$variables$log_ccbal
w_obs <- weights(scf_debtors$mi_design[[1]])
ss_tot <- sum(w_obs * (y_obs - weighted.mean(y_obs, w_obs))^2)
r2_ols <- round(1 - ss_res / ss_tot, 4)

n_debtors <- nrow(scf_debtors$mi_design[[1]]$variables)
cat(sprintf("\n**Model fit (Implicate 1):** Weighted R² = %.4f; N (debtors) = %s\n",
  r2_ols, formatC(n_debtors, big.mark = ",")))
```

## Coefficient Comparison

```{r coefficient-plot, fig.width=11, fig.height=9}
logit_df <- logit_model$results %>%
  filter(term != "(Intercept)") %>%
  mutate(model = "Part 1: Logit (Has Debt)")

ols_df <- ols_model$results %>%
  filter(term != "(Intercept)") %>%
  mutate(model = "Part 2: OLS (log Balance)")

combined <- bind_rows(logit_df, ols_df) %>%
  mutate(
    significant = p.value < 0.05,
    direction = ifelse(estimate > 0, "Positive", "Negative")
  )

ggplot(combined, aes(x = reorder(term, abs(estimate)), y = estimate,
                      fill = direction, alpha = significant)) +
  geom_col(width = 0.7) +
  geom_errorbar(aes(ymin = estimate - 1.96 * std.error,
                     ymax = estimate + 1.96 * std.error),
                width = 0.3, linewidth = 0.4) +
  facet_wrap(~model, scales = "free_x") +
  scale_fill_manual(values = c("Positive" = "steelblue", "Negative" = "coral")) +
  scale_alpha_manual(values = c("TRUE" = 1.0, "FALSE" = 0.3)) +
  coord_flip() +
  labs(
    title = "Two-Part Model: Predictors of Credit Card Debt",
    subtitle = "Part 1: Who carries debt? | Part 2: How much?",
    x = NULL, y = "Coefficient Estimate (95% CI)",
    fill = "Direction", alpha = "p < 0.05"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

---


# What the Data Tells Us: Five Key Findings

The two-part model separates two distinct questions — who takes on credit card debt, and
how much they carry — revealing that the answers are not the same. The following findings
are ordered by their practical significance.

## Finding 1: The First Child Is the Financial Tipping Point

```{r finding1, results='asis'}
logit_hc <- logit_results %>% filter(term == "has_children")
logit_k  <- logit_results %>% filter(term == "kids")
ols_hc   <- ols_model$results %>% filter(term == "has_children")

cat(sprintf("The single most striking result concerns parenthood. Having any children
increases the odds of carrying credit card debt by %.0f%% (OR = %.2f, p < 0.001). This
is the largest demographic effect in the model — larger than employment, education,
race, or marriage.\n\n",
  100 * (logit_hc$odds_ratio - 1), logit_hc$odds_ratio))

cat(sprintf("But here is the nuance: among households that already carry debt, the
presence of children does *not* predict higher balances (OLS p = %.3f). And each
additional child beyond the first is actually associated with *lower* odds of carrying
debt (OR = %.2f, p = %.3f). This suppression effect suggests that the first child is
the financially consequential event. Larger families tend to be older, more established
households that have moved past the acute expense shock of a new baby.\n\n",
  ols_hc$p.value, exp(logit_k$estimate), logit_k$p.value))

cat("**The takeaway for young adults:** The transition from zero to one child is the
moment when credit card debt becomes significantly more likely. Financial planning for
parenthood should focus on this transition — building an emergency fund, reducing
existing debt, and anticipating the lumpy expenses (medical bills, childcare, equipment)
that accompany a first child.\n")
```

## Finding 2: Marriage Does Not Independently Predict Credit Card Debt

```{r finding2, results='asis'}
logit_m <- logit_results %>% filter(term == "currently_married")
ols_m   <- ols_model$results %>% filter(term == "currently_married")

cat(sprintf("Marriage has no significant independent effect on credit card debt. The
odds ratio is %.2f (p = %.3f) — and the OLS coefficient among debtors is essentially
zero (p = %.3f).\n\n",
  logit_m$odds_ratio, logit_m$p.value, ols_m$p.value))

cat("This is a substantively important null finding. In the descriptive statistics,
married households carry higher average balances. But that association disappears
entirely once we control for income, homeownership, pension access, and other financial
characteristics. Married households carry more debt not because of marriage itself, but
because marriage is correlated with higher income, homeownership, and greater access to
credit.\n\n")

cat("**The takeaway:** For researchers and counselors interested in the relationship
between financial stress and marriage, credit card debt should be understood as a
function of the household's broader financial position, not as a distinctive feature of
married life.\n")
```

## Finding 3: Credit Denial Is the Strongest Predictor of Revolving Debt

```{r finding3, results='asis'}
logit_td <- logit_results %>% filter(term == "turned_down")
ols_td   <- ols_model$results %>% filter(term == "turned_down")

cat(sprintf("Households that have been denied credit in the past five years have %.0f%%
higher odds of carrying a credit card balance (OR = %.2f, p < 0.001). This is the single
largest coefficient in the logistic model. However, among those who carry debt, credit
denial does not predict higher balances (OLS p = %.3f).\n\n",
  100 * (logit_td$odds_ratio - 1), logit_td$odds_ratio, ols_td$p.value))

cat("This pattern suggests that credit cards function as a *last resort* for financially
excluded households. Denied other forms of credit (auto loans, personal loans, home
equity lines), these households lean on their existing credit card lines. The amounts
they carry are then governed by their credit limits and income, not by the denial
experience itself.\n\n")

cat("**The takeaway:** Credit denial is a powerful signal of financial vulnerability. A
household that has been turned down for credit is nearly twice as likely to be revolving
credit card debt. For financial counselors, a recent credit denial should trigger a
conversation about debt management strategies.\n")
```

## Finding 4: Higher Wages Help People Pay Down Debt

```{r finding4, results='asis'}
ols_wage <- ols_model$results %>% filter(term == "asinh_wageinc")

cat(sprintf("Among debtors, higher wage income is associated with significantly lower
credit card balances (OLS coefficient = %.3f, p = %.3f). This is one of the few
variables that reaches significance in the Part 2 model.\n\n",
  ols_wage$estimate, ols_wage$p.value))

cat("The interpretation is straightforward: earnings provide debt-repayment capacity.
Workers who earn more can pay down their balances faster. This contrasts with total
household income, which is not significant — suggesting that it is *earned* income from
labor, not investment returns or transfers, that drives repayment behavior.\n\n")

cat("**The takeaway for policy:** Programs that increase wages or reduce income
volatility may be more effective at reducing credit card debt burdens than programs that
expand credit access. Expanding access increases the *prevalence* of debt (more people
can borrow) but does not reduce its *severity* (how much they owe). Wage growth
addresses severity directly.\n")
```

## Finding 5: Liquid Savings Protect Against Entering Debt

```{r finding5, results='asis'}
logit_liq <- logit_results %>% filter(term == "asinh_liq")
logit_stk <- logit_results %>% filter(term == "asinh_stocks")
logit_nmf <- logit_results %>% filter(term == "asinh_nmmf")

cat(sprintf("Households with more liquid assets are significantly less likely to carry
any credit card debt. Liquid savings (OR = %.2f), stock holdings (OR = %.2f), and
mutual fund balances (OR = %.2f) all carry significant negative coefficients in the
logistic model.\n\n",
  logit_liq$odds_ratio, logit_stk$odds_ratio, logit_nmf$odds_ratio))

cat("However, among those who carry debt, asset levels do not consistently predict
lower balances. This asymmetry reveals a threshold effect: liquid savings determine
whether a household *needs* to use credit cards as a borrowing tool. If an unexpected
expense arises, a household with savings can absorb it; a household without savings
reaches for the credit card. But once a household is carrying a balance, other factors
— debt-to-income ratio, installment obligations, income — matter more for the
amount.\n\n")

cat("**The takeaway:** Emergency savings are the first line of defense against credit
card debt. Even a modest liquid savings buffer can prevent the entry into revolving
debt that becomes difficult to escape.\n")
```

---

# Conclusion

This analysis reveals that credit card debt in America is not a monolithic phenomenon —
it is a two-stage process driven by different forces at each stage. The factors that push
a household into revolving debt (parenthood, credit denial, lack of liquid savings) are
largely distinct from the factors that determine how much debt they carry (wages,
debt-to-income ratio, installment obligations).

The central finding — that the first child is the single most consequential event for
credit card indebtedness — has direct implications for financial counselors, policy
makers, and young adults themselves. The transition to parenthood increases the odds of
carrying credit card debt by 54%, but marriage does not. Additional children beyond the
first do not compound the effect. This means the financial preparation conversation
should center on the first pregnancy, not on the wedding.

For policy, the data points in two directions. First, expanding credit access without
addressing underlying financial vulnerability will increase the *prevalence* of credit
card debt without reducing its *severity*. Second, wage growth is the most effective
lever for reducing balances among those who already carry debt — higher earners pay down
faster. Programs that combine income support with financial literacy at the transition
to parenthood may be the most targeted intervention available.

---

---

# Appendix: Technical Notes

**Multiple Imputation.** All statistics and models are pooled across the SCF's five
implicates using Rubin's Rules, as implemented by the `scf` package. No results are
reported from a single implicate. Model fit statistics (pseudo-R², R²) are reported from
Implicate 1 for reference; these may vary slightly across implicates.

**Survey Weights.** All estimates use the SCF replicate weights (balanced repeated
replication) to account for the complex survey design, including oversampling of wealthy
households.

**Lasso Implementation.** Lasso was run via `glmnet` with 10-fold cross-validation on
each implicate separately. We used `lambda.1se` (the most regularized model within one
standard error of the minimum CV error) for parsimony. The majority rule (variable survives
in >= 3 of 5 implicates) ensures robustness to imputation variation.

**Tautology Correction.** Total household debt (`debt`) includes credit card balances as
a component. To avoid a mechanically inflated coefficient, we constructed
`debt_excl_cc = max(debt - ccbal, 0)` and used its asinh transform in place of
`asinh_debt` throughout the Lasso and regression models.

**Two-Part Model.** Part 1 is a logistic regression on the binary indicator `has_ccdebt`.
Part 2 is OLS on `log(ccbal)` conditional on `ccbal > 0`. This approach is standard for
semicontinuous outcomes with a large mass at zero and a continuous positive tail (Duan et
al., 1983; Manning et al., 1987).

**Suppression Effect: kids vs. has_children.** Both the continuous count of children and
the binary indicator are included in the model. The opposite-signed coefficients are a
classical suppression pattern: the binary indicator captures the extensive margin (any
children vs. none), while the continuous count captures the intensive margin conditional on
the binary. This is analogous to a two-part specification within the predictor set itself.

```{r session-info, results='asis'}
cat("\n\n**Session Information**\n\n")
cat(sprintf("R version: %s\n\n", R.version.string))
cat(sprintf("Platform: %s\n\n", R.version$platform))
cat(sprintf("Key packages: scf %s, glmnet %s, survey %s\n",
  packageVersion("scf"), packageVersion("glmnet"), packageVersion("survey")))
```
