---
title: "Credit Card Balance and Household Behavior in the United States"
subtitle: "A Two-Part Model with Lasso Variable Selection Using the 2022 Survey of Consumer Finances"
author: "Delroy Nichols"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 5, dpi = 150
)
rm(list = ls())
gc()
set.seed(42)
```

# Introduction

**Research Question:** What financial and demographic characteristics most strongly predict
credit card balance levels among U.S. households, and what do these associations suggest
about the behavioral and causal mechanisms — including relationship formation, willingness
to start families, and labor market participation — that govern household indebtedness?

**Analytic Strategy:** We employ a two-part modeling approach. Part 1 uses logistic
regression to model the probability of carrying any credit card debt (ccbal > 0). Part 2
uses OLS regression to model the magnitude of credit card balances, conditional on carrying
any debt. Before fitting either model, we use Lasso regression on a broad set of continuous
SCF variables to identify the most influential predictors, retaining only those whose
coefficients survive penalization across a majority of implicates.

**Data Source:** 2022 Survey of Consumer Finances (SCF), Federal Reserve Board. The SCF
uses a dual-frame design combining area-probability sampling with a list-based oversample
of wealthy households. All analyses account for the five-implicate multiple imputation
structure and use replicate weights. Inferential statistics are pooled across all five
implicates using Rubin's Rules.

**Note on Scope:** All findings reported here are correlational. The cross-sectional design
of the SCF does not support causal inference. We discuss plausible mechanisms but acknowledge
that reverse causality, confounding, and selection effects are present throughout.

---

# Package Loading

```{r load-packages}
required_packages <- c(
  "tidyverse", "haven", "survey", "labelled",
  "psych", "ggplot2", "scales", "scf",
  "glmnet", "Matrix", "broom", "knitr",
  "moments", "car"
)

for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg, repos = "http://cran.us.r-project.org")
    library(pkg, character.only = TRUE)
  }
}

cat("All packages loaded.\n")
```

---

# Data Download and Load

```{r download-data, message=TRUE}
dir.create("scf_data", showWarnings = FALSE, recursive = TRUE)
original_wd <- getwd()
setwd("scf_data")

if (!file.exists("scf2022.rds")) {
  cat("Downloading 2022 SCF data (this may take a few minutes)...\n")
  tryCatch({
    scf_download(2022)
    cat("2022 download complete.\n")
  }, error = function(e) {
    cat("Error downloading 2022 data:", conditionMessage(e), "\n")
  })
} else {
  cat("scf2022.rds already exists — skipping download.\n")
}

setwd(original_wd)
cat("Download step complete.\n")
```

```{r load-data}
scf_2022 <- scf_load(2022, data_directory = "scf_data")
cat("2022 SCF loaded successfully.\n")
```

---

# Outcome Variable: Credit Card Balance (ccbal)

We examine the distribution of `ccbal` to determine the most appropriate modeling approach.

## Raw Distribution Diagnostics

```{r outcome-diagnostics}
imp1 <- scf_2022$mi_design[[1]]$variables

cat("===== ccbal: Raw Distribution Summary (Implicate 1) =====\n")
cat("N total:       ", nrow(imp1), "\n")
cat("N zero:        ", sum(imp1$ccbal == 0), "\n")
cat("Pct zero:      ", round(100 * mean(imp1$ccbal == 0), 1), "%\n")
cat("N positive:    ", sum(imp1$ccbal > 0), "\n")
cat("Pct positive:  ", round(100 * mean(imp1$ccbal > 0), 1), "%\n\n")

cat("--- Among ALL households ---\n")
cat("Mean:     $", formatC(mean(imp1$ccbal), format = "f", big.mark = ",", digits = 0), "\n")
cat("Median:   $", formatC(median(imp1$ccbal), format = "f", big.mark = ",", digits = 0), "\n")
cat("SD:       $", formatC(sd(imp1$ccbal), format = "f", big.mark = ",", digits = 0), "\n")
cat("Skewness: ", round(moments::skewness(imp1$ccbal), 2), "\n")
cat("Kurtosis: ", round(moments::kurtosis(imp1$ccbal), 2), "\n\n")

pos <- imp1$ccbal[imp1$ccbal > 0]
cat("--- Among households WITH credit card debt ---\n")
cat("Mean:     $", formatC(mean(pos), format = "f", big.mark = ",", digits = 0), "\n")
cat("Median:   $", formatC(median(pos), format = "f", big.mark = ",", digits = 0), "\n")
cat("SD:       $", formatC(sd(pos), format = "f", big.mark = ",", digits = 0), "\n")
cat("Skewness: ", round(moments::skewness(pos), 2), "\n")
cat("Kurtosis: ", round(moments::kurtosis(pos), 2), "\n")
cat("Max:      $", formatC(max(pos), format = "f", big.mark = ",", digits = 0), "\n")
```

## Distribution Visualizations

```{r outcome-plots, fig.width=10, fig.height=8}
par(mfrow = c(2, 2))

hist(imp1$ccbal, breaks = 100, main = "ccbal: Raw Distribution (All)",
     xlab = "Credit Card Balance ($)", col = "steelblue", border = "white")

hist(pos, breaks = 80, main = "ccbal: Positive Balances Only",
     xlab = "Credit Card Balance ($)", col = "coral", border = "white")

hist(log(pos), breaks = 60, main = "log(ccbal): Positive Balances",
     xlab = "log(Credit Card Balance)", col = "seagreen", border = "white")

hist(asinh(imp1$ccbal), breaks = 60, main = "asinh(ccbal): All Households",
     xlab = "asinh(Credit Card Balance)", col = "mediumpurple", border = "white")

par(mfrow = c(1, 1))
```

## Transformation Decision

The distribution of `ccbal` exhibits two defining features: a large mass at zero (roughly
62% of households carry no credit card balance) and extreme right skew among positive
values (skewness > 4, kurtosis > 30). This structure motivates a two-part modeling approach.

For Part 1 (logistic regression), the outcome is the binary indicator `has_ccdebt` (1 if
ccbal > 0). For Part 2 (OLS regression), the outcome is `log(ccbal)` conditional on
carrying any debt, which normalizes the heavy right tail. For the Lasso variable selection
step, we use `asinh(ccbal)` on the full sample. The inverse hyperbolic sine handles zeros
naturally and approximates `log(2x)` for large values, making it suitable for an initial
screening pass that includes both debtors and non-debtors.

## Create Outcome Variables

```{r create-outcomes}
scf_2022 <- scf_update(scf_2022,
  has_ccdebt  = as.integer(ccbal > 0),
  log_ccbal   = ifelse(ccbal > 0, log(ccbal), NA),
  asinh_ccbal = asinh(ccbal)
)

cat("Outcome variables created: has_ccdebt, log_ccbal, asinh_ccbal\n")
```

---

# Predictor Variables: Wide Net and Wrangling

We engineer a broad set of continuous and categorical variables from the SCF extract
dataset. A key decision here is the treatment of total debt (`debt`). Because `ccbal` is
a component of `debt`, including `asinh_debt` as a predictor would introduce a mechanical
(tautological) relationship — total debt predicts credit card debt partly because it
contains credit card debt. We therefore construct `debt_excl_cc` (total debt minus credit
card balance) to preserve the signal from non-credit-card liabilities without the
circularity.

## Variable Engineering

```{r variable-engineering}
scf_2022 <- scf_update(scf_2022,

  # ---- Demographics ----
  female            = as.integer(hhsex == 2),
  currently_married = as.integer(married == 1),
  has_children      = as.integer(kids > 0),
  college_degree    = as.integer(edcl == 4),

  educ_cat = factor(edcl,
    levels = 1:4,
    labels = c("Less than HS", "High School", "Some College", "College Degree")
  ),

  race_cat = factor(racecl4,
    levels = 1:4,
    labels = c("White", "Black", "Hispanic", "Other")
  ),

  age_group = cut(age,
    breaks = c(18, 24, 34, 44, 54, 64, Inf),
    labels = c("18-24", "25-34", "35-44", "45-54", "55-64", "65+"),
    right = TRUE, include.lowest = TRUE
  ),

  age_sq = age^2,

  # ---- Income and wealth (asinh for zeros/negatives) ----
  asinh_income   = asinh(income),
  asinh_networth = asinh(networth),
  asinh_wageinc  = asinh(wageinc),

  # ---- Debt variables (asinh) ----
  # debt_excl_cc removes the tautological component
  debt_excl_cc       = pmax(debt - ccbal, 0),
  asinh_debt_excl_cc = asinh(pmax(debt - ccbal, 0)),
  asinh_install      = asinh(install),
  asinh_edn_inst     = asinh(edn_inst),
  asinh_nh_mort      = asinh(nh_mort),
  asinh_mrthel       = asinh(mrthel),
  asinh_othloc       = asinh(othloc),
  asinh_resdbt       = asinh(resdbt),

  # ---- Asset variables (asinh) ----
  asinh_asset    = asinh(asset),
  asinh_fin      = asinh(fin),
  asinh_nfin     = asinh(nfin),
  asinh_liq      = asinh(liq),
  asinh_retqliq  = asinh(retqliq),
  asinh_cashli   = asinh(cashli),
  asinh_stocks   = asinh(stocks),
  asinh_bond     = asinh(bond),
  asinh_nmmf     = asinh(nmmf),
  asinh_vehic    = asinh(vehic),
  asinh_houses   = asinh(houses),
  asinh_homeeq   = asinh(homeeq),
  asinh_bus      = asinh(bus),
  asinh_equity   = asinh(equity),
  asinh_saving   = asinh(saving),
  asinh_checking = asinh(checking),

  # ---- Ratios ----
  asinh_dti = asinh(ifelse(income > 0, debt / income, 0)),

  # ---- Counts ----
  nvehic_raw = nvehic,

  # ---- Labor / credit access indicators ----
  is_working  = as.integer(lf == 1),
  is_homeowner = as.integer(housecl == 1),
  has_pension  = as.integer(anypen == 1),
  turned_down  = as.integer(turndown == 1),
  fear_denial  = as.integer(feardenial == 1)
)

cat("All predictor variables engineered across implicates.\n")
```

## Define Predictor Sets

```{r define-predictor-set}
# Full continuous predictor set for Lasso (wide net)
# NOTE: asinh_debt_excl_cc replaces asinh_debt to avoid tautology
continuous_predictors <- c(
  "age", "age_sq",
  "asinh_income", "asinh_wageinc", "asinh_networth",
  "asinh_debt_excl_cc", "asinh_install", "asinh_edn_inst",
  "asinh_nh_mort", "asinh_mrthel", "asinh_othloc", "asinh_resdbt",
  "asinh_asset", "asinh_fin", "asinh_nfin",
  "asinh_liq", "asinh_retqliq", "asinh_cashli",
  "asinh_stocks", "asinh_bond", "asinh_nmmf",
  "asinh_vehic", "asinh_houses", "asinh_homeeq",
  "asinh_bus", "asinh_equity", "asinh_saving", "asinh_checking",
  "asinh_dti", "nvehic_raw", "kids"
)

# Categorical / binary predictors (enter final models directly, not penalized by Lasso)
categorical_predictors <- c(
  "female", "currently_married", "has_children",
  "college_degree", "is_working", "is_homeowner",
  "has_pension", "turned_down", "fear_denial"
)

cat("Continuous predictors for Lasso: ", length(continuous_predictors), "\n")
cat("Binary/categorical predictors:   ", length(categorical_predictors), "\n")
```

---

# Lasso Variable Selection

We run Lasso regression via `glmnet` with `asinh(ccbal)` as the outcome and the full set
of continuous predictors. The Lasso shrinks uninformative coefficients to zero, retaining
only variables with genuine predictive signal.

Because the SCF uses multiple imputation, we run the Lasso on each of the five implicates
separately and retain variables that survive penalization in at least 3 of 5 implicates
(majority rule). This ensures that variable selection is robust to imputation variation.

```{r lasso-selection, fig.width=10, fig.height=7}
library(glmnet)

lasso_selected <- list()
lasso_coefs    <- list()

for (i in 1:5) {
  cat("--- Implicate", i, "---\n")

  df_i <- scf_2022$mi_design[[i]]$variables
  wt_i <- df_i$wgt

  # Build predictor matrix (continuous only)
  available <- continuous_predictors[continuous_predictors %in% names(df_i)]
  X <- as.matrix(df_i[, available, drop = FALSE])
  y <- df_i$asinh_ccbal

  # Remove rows with NA
  complete <- complete.cases(X, y, wt_i)
  X <- X[complete, , drop = FALSE]
  y <- y[complete]
  w <- wt_i[complete]
  w <- w / mean(w)

  # Standardize
  X_scaled <- scale(X)

  # Cross-validated Lasso
  set.seed(123)
  cv_fit <- cv.glmnet(X_scaled, y, weights = w, alpha = 1,
                       nfolds = 10, type.measure = "mse")

  # Coefficients at lambda.1se (parsimonious)
  coef_1se <- coef(cv_fit, s = "lambda.1se")
  coef_df  <- as.data.frame(as.matrix(coef_1se))
  coef_df$variable <- rownames(coef_df)
  names(coef_df)[1] <- "coefficient"

  nonzero <- coef_df %>%
    filter(variable != "(Intercept)", coefficient != 0) %>%
    arrange(desc(abs(coefficient)))

  lasso_selected[[i]] <- nonzero$variable
  lasso_coefs[[i]]    <- nonzero

  cat("  lambda.1se =", round(cv_fit$lambda.1se, 4), "\n")
  cat("  Variables retained:", nrow(nonzero), "of", length(available), "\n\n")
}

# Majority rule: keep variables surviving in >= 3 of 5 implicates
all_vars <- unlist(lasso_selected)
var_counts <- sort(table(all_vars), decreasing = TRUE)

cat("===== Lasso Survival Across Implicates =====\n")
print(var_counts)

lasso_survivors <- names(var_counts[var_counts >= 3])
cat("\n===== Variables Surviving Majority Rule (>= 3/5) =====\n")
cat(paste(lasso_survivors, collapse = ", "), "\n")
cat("Total retained:", length(lasso_survivors), "\n")
```

## Lasso Coefficient Visualization

```{r lasso-plot, fig.width=10, fig.height=7}
avg_coefs <- bind_rows(lasso_coefs, .id = "implicate") %>%
  filter(variable %in% lasso_survivors) %>%
  group_by(variable) %>%
  summarise(
    mean_coef    = mean(coefficient),
    sd_coef      = sd(coefficient),
    n_implicates = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(abs(mean_coef)))

cat("\n===== Average Lasso Coefficients (Standardized, Across Implicates) =====\n")
print(avg_coefs, n = 50)

ggplot(avg_coefs, aes(x = reorder(variable, mean_coef), y = mean_coef)) +
  geom_col(aes(fill = mean_coef > 0), width = 0.7) +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "coral"),
                    labels = c("Negative", "Positive"), name = "Direction") +
  coord_flip() +
  labs(
    title = "Lasso-Selected Predictors of asinh(Credit Card Balance)",
    subtitle = "Average standardized coefficients across 5 implicates (lambda.1se, majority rule)",
    x = NULL, y = "Average Lasso Coefficient"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

The Lasso winnowed `r length(continuous_predictors)` continuous predictors down to
`r length(lasso_survivors)` variables that carry genuine predictive signal for credit card
balances. Variables shrunk to zero across most implicates are considered noise or redundant.
We carry the surviving variables forward into the two-part model alongside the categorical
predictors.

---

# Univariate Statistics

Before modeling, we examine the marginal distributions of `ccbal` using proper
survey-weighted estimates pooled across all five implicates.

## Credit Card Balance: Weighted Estimates

```{r univariate-ccbal}
cat("===== Weighted Mean: ccbal (Overall) =====\n")
print(scf_mean(scf_2022, ~ccbal))

cat("\n===== By Marital Status =====\n")
print(scf_mean(scf_2022, ~ccbal, by = ~currently_married))

cat("\n===== By Gender =====\n")
print(scf_mean(scf_2022, ~ccbal, by = ~female))

cat("\n===== By Education =====\n")
print(scf_mean(scf_2022, ~ccbal, by = ~educ_cat))

cat("\n===== By Race =====\n")
print(scf_mean(scf_2022, ~ccbal, by = ~race_cat))

cat("\n===== By Age Group =====\n")
print(scf_mean(scf_2022, ~ccbal, by = ~age_group))

cat("\n===== By Has Children =====\n")
print(scf_mean(scf_2022, ~ccbal, by = ~has_children))

cat("\n===== By Working Status =====\n")
print(scf_mean(scf_2022, ~ccbal, by = ~is_working))

cat("\n===== By Turned Down for Credit =====\n")
print(scf_mean(scf_2022, ~ccbal, by = ~turned_down))
```

## Prevalence of Credit Card Debt

```{r univariate-prevalence}
cat("===== Weighted Prevalence: Carries Any CC Debt =====\n")
print(scf_mean(scf_2022, ~has_ccdebt))

cat("\n===== By Marital Status =====\n")
print(scf_mean(scf_2022, ~has_ccdebt, by = ~currently_married))

cat("\n===== By Education =====\n")
print(scf_mean(scf_2022, ~has_ccdebt, by = ~educ_cat))

cat("\n===== By Race =====\n")
print(scf_mean(scf_2022, ~has_ccdebt, by = ~race_cat))

cat("\n===== By Age Group =====\n")
print(scf_mean(scf_2022, ~has_ccdebt, by = ~age_group))

cat("\n===== By Has Children =====\n")
print(scf_mean(scf_2022, ~has_ccdebt, by = ~has_children))

cat("\n===== By Turned Down for Credit =====\n")
print(scf_mean(scf_2022, ~has_ccdebt, by = ~turned_down))
```

---

# Bivariate Associations

We estimate simple bivariate regressions of `asinh(ccbal)` on each key grouping variable.
Each coefficient represents a survey-weighted, MI-pooled association.

```{r bivariate}
cat("===== asinh_ccbal ~ currently_married =====\n")
print(scf_ols(scf_2022, asinh_ccbal ~ currently_married)$results)

cat("\n===== asinh_ccbal ~ female =====\n")
print(scf_ols(scf_2022, asinh_ccbal ~ female)$results)

cat("\n===== asinh_ccbal ~ college_degree =====\n")
print(scf_ols(scf_2022, asinh_ccbal ~ college_degree)$results)

cat("\n===== asinh_ccbal ~ has_children =====\n")
print(scf_ols(scf_2022, asinh_ccbal ~ has_children)$results)

cat("\n===== asinh_ccbal ~ is_working =====\n")
print(scf_ols(scf_2022, asinh_ccbal ~ is_working)$results)

cat("\n===== asinh_ccbal ~ turned_down =====\n")
print(scf_ols(scf_2022, asinh_ccbal ~ turned_down)$results)

cat("\n===== asinh_ccbal ~ fear_denial =====\n")
print(scf_ols(scf_2022, asinh_ccbal ~ fear_denial)$results)

cat("\n===== asinh_ccbal ~ is_homeowner =====\n")
print(scf_ols(scf_2022, asinh_ccbal ~ is_homeowner)$results)
```

---

# Two-Part Regression Models

## Build Final Predictor Set

The final predictor set combines the Lasso-selected continuous variables with the full
set of categorical/binary predictors. The categorical variables enter the model unpenalized
because they represent theoretically motivated controls (demographics, labor force status,
credit access experience) that should be included regardless of Lasso selection.

```{r build-formula}
lasso_final <- setdiff(lasso_survivors,
  c("asinh_ccbal", "ccbal", "has_ccdebt", "log_ccbal"))

all_predictors <- unique(c(lasso_final, categorical_predictors))

cat("Final predictor set for both models:\n")
cat(paste(all_predictors, collapse = "\n"), "\n")
cat("\nTotal predictors:", length(all_predictors), "\n")
```

## Part 1: Logistic Regression — P(Carries Any Credit Card Debt)

```{r part1-logit}
logit_formula <- as.formula(
  paste("has_ccdebt ~", paste(all_predictors, collapse = " + "))
)

cat("===== Part 1: Logistic Regression (Log-Odds) =====\n")
cat("Outcome: has_ccdebt (1 = carries any credit card balance)\n\n")

logit_model <- scf_glm(scf_2022, logit_formula, exponentiate = FALSE)
print(logit_model$results)
```

```{r logit-odds-ratios}
cat("\n===== Part 1: Odds Ratios (Exponentiated) =====\n")
logit_or <- scf_glm(scf_2022, logit_formula, exponentiate = TRUE)
print(logit_or$results)
```

## Part 2: OLS Regression — log(ccbal), Conditional on ccbal > 0

```{r part2-subset}
scf_debtors <- scf_2022
for (i in 1:5) {
  idx <- scf_2022$mi_design[[i]]$variables$ccbal > 0
  scf_debtors$mi_design[[i]] <- subset(scf_2022$mi_design[[i]], idx)
}

cat("Subsetting to households with ccbal > 0:\n")
for (i in 1:5) {
  cat("  Implicate", i, ":", nrow(scf_debtors$mi_design[[i]]$variables), "debtors\n")
}
```

```{r part2-ols}
ols_formula <- as.formula(
  paste("log_ccbal ~", paste(all_predictors, collapse = " + "))
)

cat("\n===== Part 2: OLS Regression =====\n")
cat("Outcome: log(ccbal), conditional on ccbal > 0\n\n")

ols_model <- scf_ols(scf_debtors, ols_formula)
print(ols_model$results)
```

## Coefficient Comparison Visualization

```{r coefficient-plot, fig.width=11, fig.height=9}
logit_df <- logit_model$results %>%
  filter(term != "(Intercept)") %>%
  mutate(model = "Part 1: Logit (Has Debt)")

ols_df <- ols_model$results %>%
  filter(term != "(Intercept)") %>%
  mutate(model = "Part 2: OLS (log Balance)")

combined <- bind_rows(logit_df, ols_df) %>%
  mutate(
    significant = p.value < 0.05,
    direction = ifelse(estimate > 0, "Positive", "Negative")
  )

ggplot(combined, aes(x = reorder(term, abs(estimate)), y = estimate,
                      fill = direction, alpha = significant)) +
  geom_col(width = 0.7) +
  geom_errorbar(aes(ymin = estimate - 1.96 * std.error,
                     ymax = estimate + 1.96 * std.error),
                width = 0.3, linewidth = 0.4) +
  facet_wrap(~model, scales = "free_x") +
  scale_fill_manual(values = c("Positive" = "steelblue", "Negative" = "coral")) +
  scale_alpha_manual(values = c("TRUE" = 1.0, "FALSE" = 0.3)) +
  coord_flip() +
  labs(
    title = "Two-Part Model: Predictors of Credit Card Debt",
    subtitle = "Part 1: Who carries debt? | Part 2: How much?",
    x = NULL, y = "Coefficient Estimate (95% CI)",
    fill = "Direction", alpha = "p < 0.05"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

---

# Interpretation: What the Data Suggests

The two-part model separates two distinct behavioral questions about credit card debt in
the United States. Part 1 asks who takes on any credit card debt at all, and Part 2 asks,
among those who carry a balance, what predicts higher or lower amounts. The following
interpretation is grounded in the actual coefficients estimated above and connects them to
plausible behavioral mechanisms.

## Marriage, Partnership, and Credit Card Debt

```{r interp-marriage, echo=FALSE, results='asis'}
# Extract the marriage coefficient from both models
logit_married <- logit_model$results %>% filter(term == "currently_married")
ols_married   <- ols_model$results %>% filter(term == "currently_married")

cat(sprintf(
  "In the logistic model, the coefficient on `currently_married` is %.3f (p = %.4f). ",
  logit_married$estimate, logit_married$p.value))

if (logit_married$p.value < 0.05) {
  if (logit_married$estimate > 0) {
    cat("This indicates that married households have significantly higher odds of carrying
any credit card debt compared to unmarried households, holding other predictors constant.
This likely reflects dual-earner households having greater access to credit lines, higher
aggregate consumption needs, and the financial demands of maintaining a joint household.\n\n")
  } else {
    cat("This indicates that married households have significantly lower odds of carrying
any credit card debt. Pooled resources and the financial discipline associated with
partnership may provide a buffer against revolving debt.\n\n")
  }
} else {
  cat("This coefficient is not statistically significant, suggesting that marriage status
does not independently predict the probability of carrying credit card debt once other
financial and demographic characteristics are accounted for.\n\n")
}

cat(sprintf(
  "In the OLS model (Part 2), the coefficient on `currently_married` is %.3f (p = %.4f). ",
  ols_married$estimate, ols_married$p.value))

if (ols_married$p.value < 0.05) {
  if (ols_married$estimate > 0) {
    cat("Among debtors, married households carry approximately %.0f%% higher balances.
This could reflect joint spending patterns and lifestyle inflation associated with
dual-income households.",
    100 * (exp(ols_married$estimate) - 1))
  } else {
    cat("Among debtors, married households carry approximately %.0f%% lower balances,
suggesting that pooled resources help manage debt levels even when a balance exists.",
    100 * (1 - exp(ols_married$estimate)))
  }
} else {
  cat("This coefficient is not statistically significant, indicating that marriage does
not predict the magnitude of credit card balances among those who carry debt.")
}

cat("\n\nThese patterns matter for relationship health. Financial stress — and credit card
debt in particular, given its high interest rates and discretionary nature — is among the
strongest predictors of relationship conflict and divorce. The coefficients here quantify
the nationally representative association between partnership and this specific form of
indebtedness.\n")
```

## Children, Family Formation, and Debt

```{r interp-children, echo=FALSE, results='asis'}
logit_kids <- logit_model$results %>% filter(term == "has_children")
ols_kids   <- ols_model$results %>% filter(term == "has_children")

cat(sprintf(
  "The logit coefficient on `has_children` is %.3f (p = %.4f). ",
  logit_kids$estimate, logit_kids$p.value))

if (logit_kids$p.value < 0.05 & logit_kids$estimate > 0) {
  cat("Households with children are significantly more likely to carry credit card debt.
This is consistent with the well-documented cost-of-children channel: childcare,
education, healthcare, clothing, and food expenses can push household spending beyond
what income covers, leading to revolving credit use.\n\n")
} else if (logit_kids$p.value < 0.05 & logit_kids$estimate < 0) {
  cat("Households with children are less likely to carry credit card debt, which may
reflect selection effects — financially prepared households may be more likely to have
children.\n\n")
} else {
  cat("This coefficient is not statistically significant.\n\n")
}

cat(sprintf(
  "In Part 2, the coefficient is %.3f (p = %.4f). ",
  ols_kids$estimate, ols_kids$p.value))

if (ols_kids$p.value < 0.05) {
  cat(sprintf("Among debtors, the presence of children is associated with a %.0f%% %s
in credit card balances.",
    abs(100 * (exp(ols_kids$estimate) - 1)),
    ifelse(ols_kids$estimate > 0, "increase", "decrease")))
} else {
  cat("Among debtors, the presence of children does not significantly predict the
magnitude of credit card balances.")
}

cat("\n\nThis has direct implications for willingness to start families. If young households
observe that parenthood is associated with credit card debt accumulation, this may
contribute to the well-documented trend of delayed family formation, particularly among
college-educated women.\n")
```

## Labor Force Participation and Earnings

```{r interp-labor, echo=FALSE, results='asis'}
logit_work <- logit_model$results %>% filter(term == "is_working")
ols_work   <- ols_model$results %>% filter(term == "is_working")

# Check if wage income survived Lasso
has_wage <- "asinh_wageinc" %in% all_predictors
if (has_wage) {
  logit_wage <- logit_model$results %>% filter(term == "asinh_wageinc")
  ols_wage   <- ols_model$results %>% filter(term == "asinh_wageinc")
}

cat(sprintf(
  "The coefficient on `is_working` in the logit is %.3f (p = %.4f) and in the OLS is
%.3f (p = %.4f). ",
  logit_work$estimate, logit_work$p.value,
  ols_work$estimate, ols_work$p.value))

if (logit_work$p.value < 0.05 & logit_work$estimate > 0 &
    (ols_work$p.value >= 0.05 | ols_work$estimate < 0)) {
  cat("This pattern — employment increasing the probability of having debt but not
increasing (or reducing) balance levels — suggests that employment primarily operates
through the credit access channel. Working households qualify for credit cards and use
them, but their income also provides capacity to manage balances.\n\n")
} else {
  cat("\n\n")
}

if (has_wage) {
  cat(sprintf(
    "Wage income (`asinh_wageinc`) has a logit coefficient of %.3f (p = %.4f) and an OLS
coefficient of %.3f (p = %.4f). ",
    logit_wage$estimate, logit_wage$p.value,
    ols_wage$estimate, ols_wage$p.value))

  if (ols_wage$p.value < 0.05 & ols_wage$estimate < 0) {
    cat("The negative OLS coefficient is important: among debtors, higher wage income
is associated with lower credit card balances. This supports the interpretation that
earnings provide capacity to pay down revolving debt rather than fuel additional spending.")
  }
}
cat("\n")
```

## Credit Access and Financial Exclusion

```{r interp-credit-access, echo=FALSE, results='asis'}
logit_td <- logit_model$results %>% filter(term == "turned_down")
ols_td   <- ols_model$results %>% filter(term == "turned_down")
logit_fd <- logit_model$results %>% filter(term == "fear_denial")
ols_fd   <- ols_model$results %>% filter(term == "fear_denial")

cat(sprintf(
  "The coefficient on `turned_down` is %.3f (p = %.4f) in the logit and %.3f (p = %.4f)
in the OLS. ",
  logit_td$estimate, logit_td$p.value,
  ols_td$estimate, ols_td$p.value))

if (logit_td$p.value < 0.05 & logit_td$estimate > 0) {
  cat("Households that have been turned down for credit in the past five years are
significantly more likely to carry credit card debt. This suggests that credit cards
serve as a last-resort borrowing channel for credit-constrained households — precisely
the population least equipped to manage high-interest revolving debt.\n\n")
}

cat(sprintf(
  "The coefficient on `fear_denial` is %.3f (p = %.4f) in the logit and %.3f (p = %.4f)
in the OLS. ",
  logit_fd$estimate, logit_fd$p.value,
  ols_fd$estimate, ols_fd$p.value))

if (logit_fd$p.value < 0.05) {
  cat("Fear of credit denial — a psychological rather than experiential variable —
independently predicts credit card debt. This captures anticipatory financial anxiety
that may shape spending and borrowing decisions even absent actual rejection.\n")
}
cat("\n")
```

## Wealth, Assets, and the Role of Financial Cushions

```{r interp-assets, echo=FALSE, results='asis'}
asset_vars <- c("asinh_liq", "asinh_saving", "asinh_checking",
                "asinh_retqliq", "asinh_homeeq", "asinh_fin",
                "asinh_stocks", "asinh_nmmf", "asinh_cashli")
present_assets <- intersect(asset_vars, all_predictors)

if (length(present_assets) > 0) {
  cat("The Lasso retained several asset-side variables, allowing us to examine whether
financial cushions reduce reliance on credit card borrowing.\n\n")

  for (v in present_assets) {
    logit_v <- logit_model$results %>% filter(term == v)
    ols_v   <- ols_model$results %>% filter(term == v)
    if (nrow(logit_v) > 0 & nrow(ols_v) > 0) {
      sig_l <- ifelse(logit_v$p.value < 0.05, "*", "ns")
      sig_o <- ifelse(ols_v$p.value < 0.05, "*", "ns")
      cat(sprintf("- **%s**: Logit = %.3f (%s), OLS = %.3f (%s)\n",
                  v, logit_v$estimate, sig_l, ols_v$estimate, sig_o))
    }
  }

  cat("\nNegative coefficients on liquid asset variables in the logit suggest that
households with savings buffers are less likely to carry credit card debt — savings
substitute for revolving credit. Negative coefficients in the OLS indicate that even
among debtors, those with more savings carry lower balances. Positive coefficients on
illiquid assets (e.g., home equity) may reflect a leveraging pattern where homeowners
have more credit access generally.\n")
} else {
  cat("No asset variables survived Lasso selection.\n")
}
```

## Connecting to Broader Trends

The coefficients estimated in this analysis provide a quantified, nationally representative
snapshot of the financial behaviors that connect to broader sociological trends. Delayed
marriage, declining birth rates, rising household debt, and labor market polarization are
not independent phenomena. Credit card balances sit at the intersection of these forces:
they reflect consumption aspirations bumping against income constraints, credit access
interacting with financial literacy, and household formation decisions shaped by the
perceived or actual cost of partnership and children.

The two-part structure reveals that the predictors of *having* credit card debt are not
identical to the predictors of *how much* debt one carries. This distinction matters for
policy and for understanding household decision-making. Programs that expand credit access
(e.g., credit-building initiatives) may increase the prevalence of revolving debt without
necessarily increasing balance levels. Conversely, income support or savings incentives
may reduce balances among existing debtors without changing the prevalence of debt.

---

# Appendix: Technical Notes

**Multiple Imputation.** All statistics and models are pooled across the SCF's five
implicates using Rubin's Rules, as implemented by the `scf` package. No results are
reported from a single implicate.

**Survey Weights.** All estimates use the SCF replicate weights (balanced repeated
replication) to account for the complex survey design, including oversampling of wealthy
households.

**Lasso Implementation.** Lasso was run via `glmnet` with 10-fold cross-validation on
each implicate separately. We used `lambda.1se` (the most regularized model within one
standard error of the minimum CV error) for parsimony. The majority rule (variable survives
in >= 3 of 5 implicates) ensures robustness to imputation variation.

**Tautology Correction.** Total household debt (`debt`) includes credit card balances as
a component. To avoid a mechanically inflated coefficient, we constructed
`debt_excl_cc = max(debt - ccbal, 0)` and used its asinh transform in place of
`asinh_debt` throughout the Lasso and regression models.

**Two-Part Model.** Part 1 is a logistic regression on the binary indicator `has_ccdebt`.
Part 2 is OLS on `log(ccbal)` conditional on `ccbal > 0`. This approach is standard for
semicontinuous outcomes with a large mass at zero and a continuous positive tail (Duan et
al., 1983; Manning et al., 1987).

```{r session-info}
sessionInfo()
```
