---
title: "Segmented Assimilation in the American Financial System: Origin-Based Gradients in Immigrant Household Finances"
subtitle: "Evidence from the 2022 Survey of Consumer Finances"
author: "Delroy Nichols"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 5, dpi = 150
)
rm(list = ls())
gc()
set.seed(42)
```

# Origin-Based Gradients in Immigrant Household Finances

## Abstract

For decades, scholars have debated what dictates an immigrant's financial trajectory
in the United States. The human-capital convergence model argues that education,
income, and time spent in the U.S. act as great equalizers, eventually erasing
financial differences between groups. In contrast, the segmented-assimilation
framework contends that structural and racial barriers force immigrants from
different origin regions into distinct, inescapable financial destinies. Because
historical wealth surveys did not track national origin, these competing theories
could never be directly tested against U.S. household wealth—until now.

Utilizing newly added national-origin variables from the 2022 Survey of Consumer
Finances (SCF), this paper analyzes the financial outcomes of 1,103 foreign-born
households. Across three key markers of wealth and credit access—retirement-account
ownership, credit denial, and homeownership, one per theoretical pathway—the
findings decisively favor the human-capital convergence model. The data reveal that
once education, income, and duration of U.S. residence are accounted for, a
household's region of origin exerts no statistically detectable independent effect
on its financial success. This absence of origin-based financial divergence persists
even when stratifying the sample to look exclusively at recent arrivals
(≤15 years in the U.S.). By demonstrating that human capital and time override
origin-based barriers in wealth accumulation, this study clarifies the empirical
limits of segmented-assimilation theory in the U.S. wealth context and suggests that
policy frameworks built around broad race and human-capital categories remain the
appropriate analytical lens for studying immigrant financial outcomes.

**Data Source:** 2022 Survey of Consumer Finances (SCF), Federal Reserve Board. All
estimates are survey-weighted; all models are pooled across the SCF's five multiply-
imputed datasets using Rubin's Rules.

**Note on Scope:** All findings are correlational and limited to the foreign-born
respondent sample. The SCF's cross-sectional design and absence of parental nativity
preclude causal claims and intergenerational analysis.

## Introduction: The Immigrant Wealth Debate

When a foreign-born household settles in the United States, what ultimately
dictates its financial future? For decades, economists and sociologists have
fiercely debated whether an immigrant's trajectory is determined by individual
effort and time, or by enduring structural barriers tied to where they come from.

At the heart of this debate are two competing theoretical models. The first, the
**human-capital convergence model** (Borjas, 1987, 1999), views the American
economy as a great equalizer. Foundational studies by Chiswick (1978) and Blau and
Graham (1990) contend that factors such as education, English proficiency, age at
arrival, and time spent in the U.S. sufficiently account for economic outcomes.
In this view, once you control for a household's "human capital" and how long they
have had to build a life in their new country, their region of origin has no
independent effect on their financial success.

The second model, the **segmented-assimilation framework**, makes the exact
opposite prediction. Developed by Alejandro Portes, Rubén Rumbaut, and Min Zhou
(Portes & Zhou, 1993; Portes & Rumbaut, 2001), this theory argues that American
society is highly stratified. It posits that the context of an immigrant's
reception—including racial prejudices, neighborhood disparities, and labor-market
structures—differs systematically across origin groups (Zhou, 1997).
Consequently, immigrants are forced into three distinct incorporation pathways:
*parallel integration* into the white middle class, *downward assimilation* into a
racialized underclass through structural exclusion, or *selective assimilation*,
where wealth is built primarily through ethnic-community networks. According to
this theory, a Mexican-origin household and an Asian-origin household with
identical college degrees and incomes might still experience vastly different
access to mainstream financial institutions simply because of their origin.

Until very recently, it was impossible to directly test which of these theories
accurately describes immigrant wealth building in the U.S. While previous
nationally representative wealth surveys could identify if a respondent was
foreign-born, they did not collect detail on the specific region of origin. That
changed with the 2022 wave of the Survey of Consumer Finances (SCF). By adding new
variables tracking respondent and spouse national origin alongside the exact
number of years lived in the United States, the 2022 SCF provided the first
nationally representative dataset capable of answering this question.

This paper utilizes this unprecedented data to put these two competing theoretical
models head-to-head. By analyzing the financial outcomes of foreign-born households
across three key markers of wealth and credit access—retirement-account ownership,
credit denial, and homeownership, with one outcome chosen for each of the three
theoretical pathways—this study provides an unbiased empirical test of whether
origin region exerts an independent effect on household finances.

The findings are definitive: when it comes to U.S. household wealth, the data
strongly favor the human-capital convergence model. As this paper will demonstrate,
once education, income, and duration of U.S. residence are accounted for, the
differences between origin regions largely disappear. Rather than sorting into the
distinct, inescapable pathways predicted by segmented assimilation, immigrant
wealth accumulation is driven overwhelmingly by time and human capital.

## Why This Matters: Beyond Race-Only Frameworks

The debate between human-capital convergence and segmented assimilation is not
merely an academic exercise; it has profound implications for how we understand and
address inequality in America. Historically, wealth and credit access disparities
have been studied primarily through the lens of broad racial categories. Studies
comparing Black, Hispanic, and White households often treat these populations as
homogeneous. However, prior sociological research suggests that distinct groups—such
as Black Caribbeans, Mexican-origin Hispanics, and Cuban-origin Hispanics—may
experience vastly different financial trajectories due to their unique migration
histories and community resources (Logan, 2007; Smith, 2015; Telles & Ortiz, 2008).

The 2022 SCF allows the field to finally test whether the divergent pathways
predicted by segmented assimilation actually appear in national wealth data with
enough magnitude to distinguish these origin groups. The policy stakes are
significant. If segmented assimilation accurately describes immigrant wealth
building, then financial interventions designed broadly for "Hispanic" or "Asian"
households are fundamentally flawed and must be redesigned to address deep,
origin-specific heterogeneity. If, however, the data support human-capital
convergence, then existing race-based and income-based frameworks remain the
appropriate analytical lens for studying U.S. immigrant wealth outcomes—not because
those frameworks have already closed observed disparities, but because the
within-race-category origin distinctions that segmented-assimilation theory
proposes do not appear to organize foreign-born households' financial lives in a
way the SCF can detect.

---

```{r load-packages}
required_packages <- c(
  "tidyverse", "haven", "survey", "labelled",
  "ggplot2", "scales", "scf",
  "broom", "knitr", "moments"
)

for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg, repos = "http://cran.us.r-project.org")
    library(pkg, character.only = TRUE)
  }
}
```

```{r download-data}
# Same download pattern as the credit-card-debt analysis (Paper 1).
# scf_download writes .rds files to the current working directory.

dir.create("scf_data", showWarnings = FALSE, recursive = TRUE)
original_wd <- getwd()

# Move into the data folder
setwd("scf_data")

library(haven)

# Read from your local absolute path
raw_macro_df <- read_dta("C:/Users/delro/OneDrive/Documents/scf2022s/p22i6.dta")

# Save directly to the current working directory (which is already scf_data)
saveRDS(raw_macro_df, "p22i6.rds") 

# Continue with the download check for the bulletin extract
if (!file.exists("scf2022.rds")) {
  tryCatch({
    scf_download(2022)
  }, error = function(e) {
    cat("Error downloading 2022 data:", conditionMessage(e), "\n")
  })
}

# Return to the root project directory
setwd(original_wd)
```

```{r load-data}
scf_2022 <- scf_load(2022, data_directory = "scf_data")
```

# Variable Availability and Diagnostic

Before constructing variables, we check which of the variables having direct exposure without segmented assimilation theory are
actually available in the loaded `scf` package object. The standard SCF Bulletin extract
that the `scf` package builds does not by default include the new national-origin or
years-in-U.S. items (`X6906`, `X7404`, `X6915`); these are raw codebook items rather than
constructed Bulletin variables. If they are missing, we read the raw 2022 SCF `.rds`
file directly and merge them in by household ID.

```{r diagnose-immigration-vars, results='asis'}
imp1 <- scf_2022$mi_design[[1]]$variables

# Variables we need
# X6915 = respondent national origin (the variable for origin region)
# X6906 = years respondent has lived in the U.S. (the YIU variable)
# X7796 = years spouse has lived in the U.S. (auxiliary)
# X7404 = spouse Hispanic/Latino self-id (auxiliary)
# X8049 = spouse race (used in swap correction)
# X8000 = sex/role swap indicator
#
# CORRECTED 2026-04-29: V5 incorrectly used X7437-X7443 as years-in-US.
# Per the official SCF 2022 changes file, X7437-X7443 are MORTGAGE FORBEARANCE
# variables, not immigration variables. The true years-in-US variable is X6906
# for the respondent (and X7796 for the spouse).
imm_vars <- c("X6915", "X6906", "X7796", "X7404", "X8000", "X8049")

present_in_extract <- imm_vars %in% names(imp1)

cat("**Immigration-variable availability in the scf package extract:**\n\n")
diag_df <- data.frame(
  Variable = imm_vars,
  `In extract` = ifelse(present_in_extract, "Yes", "No"),
  check.names = FALSE
)
print(kable(diag_df))
cat("\n\n")

need_raw_merge <- !all(present_in_extract)

if (need_raw_merge) {
  cat("Some required immigration variables are missing from the bulletin extract. ",
      "Attempting to merge them in from the raw 2022 SCF data file...\n\n", sep = "")
} else {
  cat("All required immigration variables are present in the extract. ",
      "No raw merge needed.\n\n", sep = "")
}
```

```{r merge-raw-immigration-vars}
# ============================================================================
# Merge raw immigration variables from the underlying SCF .rds file
# ============================================================================
# The scf package downloads the raw 2022 SCF data into scf_data/ as well as
# building the bulletin extract. The raw file contains all X-variables; we
# read it, select just the immigration-related vars + ID, and merge into
# each implicate's $variables data frame.
#
# Common file names produced by scf_download():
#   - p22i6.rds (raw public-use, 5 implicates stacked, with implicate id)
#   - scf2022.rds (the bulletin extract)
#
# We try a few candidate filenames defensively.

if (need_raw_merge) {

  raw_candidates <- c(
    file.path("scf_data", "p22i6.rds"),
    file.path("scf_data", "scfp2022.rds"),
    file.path("scf_data", "p22.rds"),
    "p22i6.rds",
    "scfp2022.rds"
  )

  raw_path <- raw_candidates[file.exists(raw_candidates)][1]

  if (!is.na(raw_path)) {
    raw_data <- readRDS(raw_path)

    # Standardize variable name casing — the raw file may have either
    # uppercase or lowercase. We try uppercase first since X-variables
    # are conventionally uppercase.
    raw_path <- raw_candidates[file.exists(raw_candidates)][1]

  if (!is.na(raw_path)) {
    raw_data <- readRDS(raw_path)

    # FIX: Force all raw data columns to uppercase immediately.
    # This aligns the Stata output with the canonical SCF uppercase format.
    names(raw_data) <- toupper(names(raw_data))

    raw_names <- names(raw_data)
    
    # The name_map is now technically redundant, but safe to keep
    name_map <- setNames(raw_names, toupper(raw_names))

    available_imm <- imm_vars[toupper(imm_vars) %in% toupper(raw_names)]

    # The raw data has 5 implicates stacked. We need an ID that matches
    # the bulletin extract's row ordering. Common candidates: y1, yy1, x1.
    id_candidates <- c("Y1", "YY1", "X1") # Updated to only look for uppercase
    id_var <- id_candidates[id_candidates %in% raw_names][1]

    if (!is.na(id_var) && length(available_imm) > 0) {

      # Pull the matching columns
      raw_imm <- raw_data[, c(id_var, available_imm), drop = FALSE]

      # Determine implicate count: raw file has each household 5 times (one
      # per implicate) — split into 5 chunks of equal size in row order.
      n_total <- nrow(raw_imm)
      n_per_imp <- nrow(scf_2022$mi_design[[1]]$variables)

      if (n_total == 5 * n_per_imp) {
        # Standard layout: implicates stacked
        for (i in 1:5) {
          start_row <- (i - 1) * n_per_imp + 1
          end_row   <- i * n_per_imp
          imm_chunk <- raw_imm[start_row:end_row, available_imm, drop = FALSE]

          # Add to the implicate's $variables
          for (v in available_imm) {
            scf_2022$mi_design[[i]]$variables[[v]] <- imm_chunk[[v]]
          }
        }
        cat("Successfully merged ", length(available_imm), " immigration variables ",
            "from ", raw_path, " into all 5 implicates.\n", sep = "")
      } else {
        cat("Row-count mismatch: raw file has ", n_total,
            " rows, bulletin extract has ", n_per_imp,
            " rows per implicate. Merge aborted; using available variables only.\n",
            sep = "")
      }
    } else {
      cat("Could not locate raw SCF file or matching ID variable. ",
          "Immigration variables will be unavailable.\n", sep = "")
    }
  } else {
    cat("Raw 2022 SCF .rds file not found in scf_data/. ",
        "Available files:\n", sep = "")
    cat(paste(list.files("scf_data"), collapse = "\n"), "\n\n")
  }
}

# Re-check what's now available
imp1 <- scf_2022$mi_design[[1]]$variables
have_X7404 <- "X7404" %in% names(imp1)
have_yiu   <- "X6906" %in% names(imp1)
have_X8000 <- "X8000" %in% names(imp1)
have_X8049 <- "X8049" %in% names(imp1)
```

```{r raw-x7404-frequency, results='asis'}
# ============================================================================
# Print raw X7404 value frequencies — needed to calibrate the recode
# ============================================================================
# Until we see what values X7404 actually takes in this wave, we cannot
# write the correct recoding rule. This chunk dumps the unweighted frequency
# table so we can adjust the recode.

if (have_X7404) {
  # 1. Extract the column and force it to a standard numeric vector
  # This strips out any problematic 'haven_labelled' metadata from Stata
  x7404_vec <- as.numeric(scf_2022$mi_design[[1]]$variables$X7404)
  
  # 2. Use Base R's table() to get the counts, which avoids package conflicts
  freq_table <- table(x7404_vec, useNA = "ifany")
  
  # 3. Reconstruct it into a clean data frame for kable
  x7404_freq <- data.frame(
    X7404 = names(freq_table),
    n = as.integer(freq_table)
  )
  
  # 4. Sort descending
  x7404_freq <- x7404_freq[order(-x7404_freq$n), ]

  cat("**Raw X7404 frequency (Implicate 1, unweighted):**\n\n")
  # row.names = FALSE keeps the markdown table perfectly clean
  print(kable(x7404_freq, caption = "X7404 = respondent national origin", row.names = FALSE))
  cat("\n\nThis table guides the origin-region recode in the next chunk. ",
      "Code 0 typically indicates 'inapplicable' (i.e., U.S.-born or non-respondent).\n\n",
      sep = "")
}
```

```{r visuals}

library(ggplot2)
library(dplyr)

# ============================================================================
# Visualize X7404 Frequencies: Spouse Hispanic/Latino Culture/Origin
# ============================================================================

# 1. SCF mapping for X7404
spouse_hisp_labels <- c(
  "0" = "Inapplicable (No Spouse/Partner)",
  "1" = "Yes (Spouse is Hispanic/Latino)",
  "5" = "No (Spouse is not Hispanic/Latino)"
)

# 2. Apply labels and generate plot
x7404_freq %>%
  mutate(
    # Map the numeric codes to our accurate text labels
    Spouse_Label = spouse_hisp_labels[as.character(X7404)],
    # Catch any unexpected codes just in case
    Spouse_Label = ifelse(is.na(Spouse_Label), paste("Code", X7404), Spouse_Label)
  ) %>%
  # Order the bars by frequency 
  ggplot(aes(x = reorder(Spouse_Label, n), y = n)) +
  geom_col(fill = "steelblue", color = "black", alpha = 0.8) +
  coord_flip() + # Horizontal layout for readability
  theme_minimal() +
  labs(
    title = "Spouse/Partner Hispanic or Latino Origin",
    subtitle = "Raw frequencies from Implicate 1 (Variable X7404)",
    x = NULL, 
    y = "Number of Respondents"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.y = element_text(size = 11)
  )

```



```{r plot-x6915, echo=FALSE, warning=FALSE, message=FALSE}
# ============================================================================
# Visualize X6915 Frequencies: Reference Person National Origin
# ============================================================================

library(ggplot2)
library(dplyr)

# FIX: Explicitly define have_X6915 so the Knit session knows what it is!
have_X6915 <- "X6915" %in% names(scf_2022$mi_design[[1]]$variables)

if (have_X6915) {
  
  x6915_vec <- as.numeric(scf_2022$mi_design[[1]]$variables$X6915)
  freq_table_6915 <- table(x6915_vec, useNA = "ifany")
  
  x6915_freq <- data.frame(
    X6915 = names(freq_table_6915),
    n = as.integer(freq_table_6915)
  )
  
  origin_labels <- c(
    "0" = "U.S. Born / Inapplicable",
    "100" = "Western European", 
    "200" = "Eastern European",
    "300" = "European",
    "400" = "West Indies",
    "500" = "Central and South American", 
    "600" = "Middle Easterner",
    "700" = "Sub-Saharan African",
    "800" = "South Asian",
    "900" = "East Asian",
    "1000" = "All Other Asians",
    "1100" = "Pacific Islander",
    "1200" = "American"
  )
  
  p_x6915 <- x6915_freq %>%
    mutate(
      Origin_Label = origin_labels[as.character(X6915)],
      Origin_Label = ifelse(is.na(Origin_Label), paste("Code", X6915), Origin_Label)
    ) %>%
    filter(X6915 != "0") %>% 
    ggplot(aes(x = reorder(Origin_Label, n), y = n)) +
    geom_col(fill = "seagreen", color = "black", alpha = 0.8) +
    coord_flip() + 
    theme_minimal() +
    labs(
      title = "Reference Person National Origin (Immigrant Sub-sample)",
      subtitle = "Unweighted frequencies from Implicate 1 (Variable X6915, excluding U.S. Born)",
      x = NULL, 
      y = "Number of Respondents"
    ) +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      axis.text.y = element_text(size = 11)
    )
  
  # Properly print the plot
  print(p_x6915)
  
} else {
  cat("Variable X6915 not found. Please ensure the merge chunk was re-run.")
}
```

```{r table-x6915, echo=FALSE, results='asis'}
# Print the raw counts table
if (have_X6915) {
  cat("\n\n**Raw X6915 frequency (Implicate 1, unweighted):**\n\n")
  tab <- knitr::kable(x6915_freq %>% arrange(desc(n)), row.names = FALSE)
  cat(paste(tab, collapse = "\n"), "\n\n")
}
```


# Variable Engineering

```{r data-engineering-and-subsetting, warning=FALSE, message=FALSE}
# ============================================================================
# Build all analysis variables and create the analytic subset
# ============================================================================
# CRITICAL DESIGN NOTE on foreign-born identification:
#
# X6915 is an ANCESTRY/HERITAGE variable, asked of all respondents — both
# foreign-born and U.S.-born. A respondent reporting "Western European"
# ancestry (code 100) is overwhelmingly U.S.-born (white American with
# European heritage), not an immigrant from Europe. Likewise, "American"
# ancestry (1200) is U.S.-born by definition, and "Sub-Saharan African" (700)
# captures Black Americans reporting African ancestry alongside actual
# African immigrants.
#
# The actual foreign-born signal is in X6906 (years the reference person
# has lived in the United States), which the SCF only asks of foreign-born
# respondents. A non-zero, non-missing years-in-US value confirms foreign-
# born status. (The companion variable X7796 is years-in-US for the spouse
# but is not used in main models, since the unit of analysis is the
# respondent.)
#
# IMPORTANT NOTE: V5 erroneously used X7437-X7443 as the years-in-US
# variables. Per the official SCF 2022 changes file, those are MORTGAGE
# FORBEARANCE variables (which mortgage terms were adjusted after exiting
# forbearance). The correct YIU variable is X6906. This V6 build uses
# X6906 throughout.
#
# Therefore, foreign_born requires BOTH:
#   (a) X6915 indicates a non-American ancestry (i.e., X6915 != 0 AND
#       X6915 != 1200), AND
#   (b) X6906 > 0 (confirming actual years lived in the U.S. as immigrant)
# ============================================================================

# ---- Step 1: Years in U.S. (X6906 = respondent years in US) ----
# X6906 = 0 means inapplicable per SCF data convention (i.e., U.S.-born).
# X6906 > 0 means the respondent reported N years lived in the U.S.,
# which is asked only of foreign-born respondents.
if (have_yiu) {
  scf_2022 <- scf_update(scf_2022,
    years_in_us  = ifelse(as.numeric(X6906) <= 0, NA_real_, as.numeric(X6906)),
    yiu_missing  = as.integer(is.na(years_in_us)),
    yiu_clean    = ifelse(is.na(years_in_us), 0, years_in_us)
  )
} else {
  scf_2022 <- scf_update(scf_2022,
    years_in_us = NA_real_, yiu_missing = 1L, yiu_clean = 0
  )
}

# ---- Step 2: CORRECTED foreign-born status ----
# Requires both ancestry indicating non-US origin AND positive years in US.
# Fall back gracefully if X6915 is unavailable.
if (have_X6915 && have_yiu) {
  scf_2022 <- scf_update(scf_2022,
    foreign_born = as.integer(
      !is.na(years_in_us) &
      years_in_us > 0 &
      as.numeric(X6915) != 0 &
      as.numeric(X6915) != 1200
    )
  )
} else if (have_yiu) {
  # YIU-only fallback if X6915 is unavailable
  scf_2022 <- scf_update(scf_2022,
    foreign_born = as.integer(!is.na(years_in_us) & years_in_us > 0)
  )
} else {
  scf_2022 <- scf_update(scf_2022, foreign_born = 0L)
}

# ---- Step 3: Origin-region recode from X6915 ----
# Used only within the foreign-born sample (filtered by foreign_born flag).
# Reference category: Latin American (the segmented-assimilation canonical
# case in the literature; X6915 == 500 = Central/South American).
if (have_X6915) {
  scf_2022 <- scf_update(scf_2022,
    origin_region = factor(case_when(
      as.numeric(X6915) == 500 ~ "Latin American",
      as.numeric(X6915) == 400 ~ "Caribbean/West Indies",
      as.numeric(X6915) %in% c(800, 900, 1000) ~ "Asian",
      as.numeric(X6915) %in% c(100, 200, 300) ~ "European/Other developed",
      as.numeric(X6915) %in% c(600, 700, 1100) ~ "Other",
      TRUE ~ NA_character_
    ), levels = c("Latin American",
                  "Caribbean/West Indies",
                  "Asian",
                  "European/Other developed",
                  "Other"))
  )
} else {
  scf_2022 <- scf_update(scf_2022, origin_region = factor(NA_character_))
}

# ---- Step 3b: Spouse/Partner Hispanic Culture/Origin from X7404 ----
if (have_X7404) {
  scf_2022 <- scf_update(scf_2022,
    spouse_hisp_origin = factor(case_when(
      X7404 == 1 ~ "Yes (Spouse Hispanic/Latino)",
      X7404 == 5 ~ "No (Spouse not Hispanic/Latino)",
      X7404 == 0 ~ "Unmarried/Inapplicable",
      TRUE ~ NA_character_
    ), levels = c("Yes (Spouse Hispanic/Latino)",
                  "No (Spouse not Hispanic/Latino)",
                  "Unmarried/Inapplicable"))
  )
} else {
  scf_2022 <- scf_update(scf_2022, spouse_hisp_origin = factor(NA_character_))
}

# ---- Step 4: Race/ethnicity ----
scf_2022 <- scf_update(scf_2022,
  race_cat = factor(racecl4, levels = 1:4,
                    labels = c("White", "Black", "Hispanic", "Other"))
)

# ---- Step 5: Standard demographic controls ----
scf_2022 <- scf_update(scf_2022,
  age_centered = age - 50,
  age_sq       = age^2,
  in_age_range = as.integer(age >= 25 & age <= 75),
  educ_cat = factor(case_when(
    educ < 8 ~ "Less than HS",
    educ == 8 ~ "High School",
    educ %in% 9:11 ~ "Some College",
    educ == 12 ~ "College Degree",
    educ %in% 13:14 ~ "Graduate School/PhD",
    TRUE ~ NA_character_
  ), levels = c("Less than HS", "High School", "Some College",
                "College Degree", "Graduate School/PhD")),
  female            = as.integer(hhsex == 2),
  currently_married = as.integer(married == 1),
  has_children      = as.integer(kids > 0),
  is_homeowner      = as.integer(housecl == 1),
  has_pension       = as.integer(anypen == 1),
  log_income        = log(pmax(1, income)),
  asinh_networth    = asinh(networth),
  log_wageinc       = log(pmax(1, wageinc))
)

# ---- Step 6: Outcomes ----
# CRITICAL: nw_pctl is computed on the FULL SCF distribution (before
# subsetting to foreign-born) so it reflects position in the broader
# U.S. wealth distribution, not relative position within the foreign-born
# subsample. This is computed via scf_update on scf_2022 (the full design)
# so all 5 implicates get a consistent percentile based on the full sample.
scf_2022 <- scf_update(scf_2022,
  own_retire    = as.integer(retqliq > 0),
  own_stocks    = as.integer(stocks > 0),
  credit_denied = as.integer(turndown == 1),
  no_checking   = as.integer(checking <= 0),
  own_home      = as.integer(housecl == 1),
  nw_pctl       = 100 * percent_rank(networth)   # full-sample percentile
)

# ---- Step 7: Analytic sample flag ----
scf_2022 <- scf_update(scf_2022,
  in_analytic_sample = as.integer(
    foreign_born == 1 &
    !is.na(origin_region) &
    in_age_range == 1
  )
)

# ---- Step 8: Create the Analytic Sample Subset ----
# Subset each implicate's survey design to foreign-born respondents.
# nw_pctl values are preserved — they were computed on the full distribution
# in Step 6 above, so subsetting just keeps the foreign-born rows with
# their original full-distribution percentile values intact.
scf_fb <- scf_2022
scf_fb$mi_design <- lapply(scf_2022$mi_design, function(des) {
  subset(des, in_analytic_sample == 1)
})

# Diagnostic: report sample size
n_fb <- nrow(scf_fb$mi_design[[1]]$variables)

# Diagnostic: cross-tab of X6915 ancestry × foreign_born flag — useful for
# verifying that the corrected foreign-born identification is sensible.
fb_diagnostic <- scf_2022$mi_design[[1]]$variables %>%
  filter(!is.na(X6915)) %>%
  group_by(X6915) %>%
  summarise(
    n_total       = n(),
    n_foreign     = sum(foreign_born == 1, na.rm = TRUE),
    pct_foreign   = round(100 * mean(foreign_born == 1, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  arrange(desc(n_total))
```

```{r build-foreign-born-design}
# ============================================================================
# Build the foreign-born subset design — same pattern as scf_debtors in Paper 1
# ============================================================================
# NOTE: foreign_born and years_in_us are already defined in the
# `data-engineering-and-subsetting` chunk above, using the X6915-based
# definition (X6915 != 0 = foreign-born). The X6915 definition is preferred
# over X6906 because X6906 has known small missingness (some foreign-born
# missingness, while X6915 is universally coded.
#
# This chunk just rebuilds the scf_fb subset object using the
# in_analytic_sample flag from the prior chunk.

scf_fb <- scf_2022

scf_fb$mi_design <- lapply(scf_2022$mi_design, function(des) {
  subset(des, in_analytic_sample == 1)
})

# Diagnostic: report sample size
n_fb <- nrow(scf_fb$mi_design[[1]]$variables)
```

# Descriptive Findings

## Foreign-Born Identification Diagnostic

The corrected `foreign_born` definition requires both ancestry indicating non-U.S.
origin (X6915 ≠ 0 and X6915 ≠ 1200) AND positive years in the U.S. (X6906).
The table below shows, for each X6915 ancestry code, what percentage of respondents
end up classified as foreign-born. Codes like 100 (Western European), 200 (Eastern
European), and 1200 (American) should be near 0%; codes like 500 (Latin American),
800/900 (Asian), and 600 (Middle Eastern) should be near 100%.

```{r fb-id-diagnostic, results='asis'}
if (exists("fb_diagnostic") && nrow(fb_diagnostic) > 0) {

  origin_label_map <- c(
    "0"    = "U.S. Born / Inapplicable",
    "100"  = "Western European",
    "200"  = "Eastern European",
    "300"  = "European (Other)",
    "400"  = "West Indies / Caribbean",
    "500"  = "Central / South American",
    "600"  = "Middle Eastern",
    "700"  = "Sub-Saharan African",
    "800"  = "South Asian",
    "900"  = "East Asian",
    "1000" = "All Other Asian",
    "1100" = "Pacific Islander",
    "1200" = "American"
  )

  fb_diagnostic_display <- fb_diagnostic %>%
    mutate(
      Ancestry = origin_label_map[as.character(X6915)],
      Ancestry = ifelse(is.na(Ancestry), paste("Code", X6915), Ancestry)
    ) %>%
    select(X6915, Ancestry, n_total, n_foreign, pct_foreign)

  cat("**Table 0. Foreign-born identification by X6915 ancestry code**\n\n")
  print(kable(fb_diagnostic_display,
              col.names = c("X6915", "Ancestry label", "n total",
                            "n foreign-born", "% foreign-born"),
              caption = "Foreign-born flag = X6915 ∈ origin codes AND years_in_us > 0."))
  cat("\n\n")
}
```

## Foreign-Born Sample Composition

```{r desc-sample-comp, results='asis'}
# Sample composition statistics — the origin_region recode was already done
# in the data-engineering chunk above; we don't need to redo it here.

if (n_fb > 0) {
  fb_imp1 <- scf_fb$mi_design[[1]]$variables

  mean_age   <- weighted.mean(fb_imp1$age, fb_imp1$wgt, na.rm = TRUE)
  mean_yiu   <- weighted.mean(fb_imp1$years_in_us, fb_imp1$wgt, na.rm = TRUE)
  mean_inc   <- weighted.mean(fb_imp1$income, fb_imp1$wgt, na.rm = TRUE)
  median_nw  <- median(fb_imp1$networth, na.rm = TRUE)
  pct_yiu_missing <- 100 * mean(is.na(fb_imp1$years_in_us))

  cat(sprintf("The analytic sample comprises %s foreign-born respondents in the 2022 SCF aged 25-75. ",
              formatC(n_fb, big.mark = ",")))
  cat(sprintf("Using survey weights, the mean age in this population is %.1f years, with a mean of %.1f years lived in the U.S. ",
              mean_age, mean_yiu))
  cat(sprintf("Mean household income is $%s and median net worth is $%s. ",
              formatC(round(mean_inc), big.mark = ",", format = "d"),
              formatC(round(median_nw), big.mark = ",", format = "d")))
  cat(sprintf("Note: %.1f%% of foreign-born respondents have missing years-in-U.S. data; the missing-indicator method is used in regressions to retain these observations.\n\n",
              pct_yiu_missing))
} else {
  cat("**Analytic sample is empty.** Check the X6915 origin-region recode in the variable-engineering chunk.\n\n")
}
```

## Origin-Region Distribution

```{r desc-origin-table, results='asis'}
if (n_fb > 0) {
  fb_imp1 <- scf_fb$mi_design[[1]]$variables

  origin_summary <- fb_imp1 %>%
    filter(!is.na(origin_region)) %>%
    group_by(origin_region) %>%
    summarise(
      n_unweighted   = n(),
      weighted_count = sum(wgt, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      pct_weighted = round(weighted_count / sum(weighted_count) * 100, 1)
    ) %>%
    select(origin_region, n_unweighted, pct_weighted)

  cat("**Table 1. Foreign-born sample by origin region (2022 SCF)**\n\n")
  print(kable(origin_summary,
              col.names = c("Origin Region", "n (Unweighted)", "% (Weighted)"),
              caption = "Note: percentages are SCF-weighted; n is unweighted (Implicate 1)."))
  cat("\n\n")

  small_cells <- origin_summary %>% filter(n_unweighted < 30)
  if (nrow(small_cells) > 0) {
    cat(sprintf("**Warning:** Small unweighted cell sizes (< 30) in %d origin region(s): %s. Model estimates for these groups may be unstable.\n\n",
                nrow(small_cells),
                paste(as.character(small_cells$origin_region), collapse = ", ")))
  }
}
```

## Outcome Prevalence by Origin Region

```{r desc-outcomes-by-origin, results='asis'}
if (n_fb > 0) {

  # V12: reduced to 3 outcomes (one per pathway) to mirror the analytical
  # structure used in the modeling section.
  outcomes <- list(
    own_retire    = "Retirement (%)",
    credit_denied = "Credit denied (%)",
    own_home      = "Homeowner (%)"
  )

  # Run each prevalence estimate using scf_mean with by = ~origin_region
  prevalence_list <- list()
  for (out_var in names(outcomes)) {
    f <- as.formula(paste0("~", out_var))
    res <- tryCatch({
      scf_mean(scf_fb, f, by = ~origin_region)$results
    }, error = function(e) NULL)
    prevalence_list[[out_var]] <- res
  }

  # Build display table — one row per origin region
  origin_levels <- levels(scf_fb$mi_design[[1]]$variables$origin_region)
  display_df <- data.frame(`Origin Region` = origin_levels,
                           check.names = FALSE,
                           stringsAsFactors = FALSE)

  for (out_var in names(outcomes)) {
    res <- prevalence_list[[out_var]]
    col_label <- outcomes[[out_var]]

    if (is.null(res) || nrow(res) == 0) {
      display_df[[col_label]] <- "--"
      next
    }

    # Match by origin_region level
    estimates <- sapply(origin_levels, function(lvl) {
      idx <- which(as.character(res[[1]]) == lvl)
      if (length(idx) == 0) return(NA_real_)
      if ("estimate" %in% names(res)) res$estimate[idx] else res[[2]][idx]
    })

    # All three outcomes are binary — multiply by 100 for percentage display
    display_df[[col_label]] <- ifelse(is.na(estimates), "--",
                                       sprintf("%.1f", estimates * 100))
  }

  cat("**Table 2. Weighted prevalence of financial outcomes by origin region**\n\n")
  print(kable(display_df,
              caption = "All estimates are SCF-weighted and pooled across implicates."))
  cat("\n\n")
}
```

## Years-in-U.S. Distribution by Origin Region

```{r desc-yiu-plot, fig.cap="Distribution of years lived in the U.S. by origin region (foreign-born SCF sample, 2022)"}
if (n_fb > 0) {
  plot_data <- scf_fb$mi_design[[1]]$variables %>%
    filter(!is.na(years_in_us), !is.na(origin_region))

  if (nrow(plot_data) > 0) {
    ggplot(plot_data,
           aes(x = origin_region, y = years_in_us, fill = origin_region)) +
      geom_boxplot(alpha = 0.7, outlier.shape = NA) +
      geom_jitter(width = 0.15, alpha = 0.2, size = 0.6) +
      labs(x = NULL,
           y = "Years lived in the U.S.",
           subtitle = "Unweighted distributional shape (Implicate 1)") +
      theme_minimal(base_size = 11) +
      theme(legend.position = "none",
            axis.text.x = element_text(angle = 30, hjust = 1)) +
      scale_fill_brewer(palette = "Set2")
  }
}
```

# Modeling Results

## H1: Origin-Region Main Effects

We test whether financial outcomes among foreign-born respondents differ by origin
region net of years-in-U.S., age, education, marital status, and income. Logistic
regression for binary outcomes (`scf_glm` with `binomial()`) and OLS for the continuous
net-worth-percentile outcome (`scf_ols`). All models pool across the 5 implicates.

Mexican origin is the reference category — that is, the X6915=500 ("Central and South
American") group, which the segmented-assimilation literature treats as the canonical
Latin American case. Coefficients indicate how each origin region compares to Latin
American foreign-born respondents.

**Reading the results.** Each outcome is modeled with a single specification —
origin region plus years-in-U.S., age, education, marital status, and log income —
chosen to mirror the "Full" specification used in earlier drafts. The
segmented-assimilation prediction is that origin-region coefficients should remain
large and statistically significant net of human capital. The human-capital
convergence prediction is that origin-region coefficients should be small and
statistically indistinguishable from zero. Confidence intervals that include 1.0
(for odds ratios) indicate failure to reject the null hypothesis of no
origin-region effect.

The analysis fits one model per pathway: retirement-account ownership for parallel
integration, credit denial for downward assimilation, and homeownership for
selective assimilation. This 1-model-per-hypothesis structure replaces the
6-outcome design used in earlier drafts; the additional outcomes (stocks,
no-checking, net-worth percentile) showed within-pathway redundancy with the three
retained outcomes and were cut to align the analytical structure with the three
hypotheses.

```{r h1-model-helper}
# ============================================================================
# Helper: fit and tidy an scf_glm or scf_ols model — Full specification only
# ============================================================================
# V12: simplified from nested (Baseline + Full) to single Full specification.
# The Baseline spec (without human-capital controls) was a robustness check
# on whether human-capital controls absorb origin effects, not a separate
# hypothesis test. The research question is whether origin matters NET of
# human capital, which is the Full specification. Reporting both inflated
# the apparent number of models without adding interpretive value.
# ============================================================================
library(dplyr)

fit_h1_model <- function(outcome_var, scf_obj, family_type = "binomial") {

  # Full specification: origin region, duration, age, education, marital
  # status, log income.
  full_rhs <- "origin_region + years_in_us + age_centered + educ_cat + currently_married + log_income"
  full_formula <- as.formula(paste(outcome_var, "~", full_rhs))

  # Set formula environment to globalenv so scf_glm can resolve term names
  # against the design's $variables data frame.
  environment(full_formula) <- globalenv()

  # Fit
  fit <- tryCatch({
    if (family_type == "binomial") {
      scf_glm(scf_obj, full_formula, family = binomial())
    } else {
      scf_ols(scf_obj, full_formula)
    }
  }, error = function(e) list(error = TRUE, message = conditionMessage(e)))

  # Propagate errors
  if (is.list(fit) && isTRUE(fit$error)) {
    return(list(error = TRUE,
                message = sprintf("[%s] %s", outcome_var, fit$message)))
  }

  # Tidy and add interpretive columns
  res <- fit$results
  res$Model <- factor("Full (With Human Capital)",
                      levels = c("Full (With Human Capital)"))

  if (family_type == "binomial" && "estimate" %in% names(res)) {
    res$odds_ratio <- exp(res$estimate)
    res$conf_low   <- exp(res$estimate - 1.96 * res$std.error)
    res$conf_high  <- exp(res$estimate + 1.96 * res$std.error)
  } else if ("estimate" %in% names(res)) {
    res$odds_ratio <- NA
    res$conf_low   <- res$estimate - 1.96 * res$std.error
    res$conf_high  <- res$estimate + 1.96 * res$std.error
  }
  res
}
```

```{r h1-diagnostic-test-fit, results='asis'}
# ============================================================================
# DIAGNOSTIC: Try a simple inline-formula scf_glm call before the main cache
# loop. If this works but the cache loop fails, the issue is formula-
# environment scoping. If this also fails, the issue is something deeper
# (data, design, package version) and the message will tell us what.
# ============================================================================

cat("**Diagnostic: simple test fit using inline formula syntax (mimics Paper 1's pattern)**\n\n")

if (n_fb >= 100) {

  test_fit <- tryCatch({
    scf_glm(scf_fb, own_home ~ origin_region, family = binomial())
  }, error = function(e) list(error = TRUE, message = conditionMessage(e)))

  if (is.list(test_fit) && isTRUE(test_fit$error)) {
    cat(sprintf("- Inline test fit FAILED: `%s`\n\n", test_fit$message))
    cat("  This means scf_glm itself is unable to fit on the foreign-born subset, ",
        "regardless of formula-environment issues. Likely causes: a needed ",
        "variable is missing in the design, the design object is malformed, ",
        "or the scf package version differs from Paper 1's.\n\n", sep = "")
  } else {
    cat("- Inline test fit SUCCEEDED. scf_glm works on this design with origin_region as predictor.\n\n")
    cat("  Coefficients returned: ",
        paste(test_fit$results$term, collapse = ", "), "\n\n", sep = "")
  }
}
```

```{r h1-fit-all-models-once, cache=TRUE}
# ============================================================================
# PERFORMANCE: Fit each H1 model EXACTLY ONCE here and cache the results.
# All downstream chunks (forest plots, H1 tables, H2 pattern matrix) pull from
# this cache instead of re-fitting. cache=TRUE on this chunk means knitr won't
# even re-run this if nothing upstream has changed.
#
# IMPORTANT: If you've previously knit V6 (or any earlier version) with
# cache=TRUE and the models failed, knitr may have cached the failed results.
# Before knitting V7, delete the cache folder created by knitr:
#   - file.exists("scf_segmented_assimilation_v7_cache")  # or v6/v5 cache
# OR delete the entire `*_cache/` directory next to the .Rmd file.
# Otherwise V7 will return the failed V6 results from cache instead of
# re-running the corrected fit logic.
# ============================================================================

model_cache <- list()

if (n_fb >= 100) {

  # V12: reduced from 6 to 3 outcomes — one per theoretical pathway.
  # Stocks (overlapping with retirement on parallel pathway), No-checking
  # (overlapping with credit-denied on downward pathway), and Net-Worth
  # Percentile (overlapping with homeownership on selective pathway, plus
  # not pathway-specific) were removed to align the analytic structure with
  # the 3-hypothesis design (1 model per pathway).
  outcomes_to_fit <- list(
    list(var = "own_retire",    family = "binomial"),  # Parallel integration
    list(var = "credit_denied", family = "binomial"),  # Downward assimilation
    list(var = "own_home",      family = "binomial")   # Selective assimilation
  )

  # Pre-check: identify outcomes with quasi-complete separation in the
  # foreign-born × origin × outcome table. If any origin region has zero
  # variance on the outcome, the GLM will produce gigantic standard errors
  # (effective infinity). Skip those outcomes rather than reporting noise.
  fb_imp1_for_check <- scf_fb$mi_design[[1]]$variables

  for (info in outcomes_to_fit) {
    if (info$family == "binomial") {
      cell_var <- fb_imp1_for_check %>%
        filter(!is.na(origin_region)) %>%
        group_by(origin_region) %>%
        summarise(v = var(.data[[info$var]], na.rm = TRUE), .groups = "drop")

      if (any(cell_var$v == 0 | is.na(cell_var$v))) {
        bad_groups <- cell_var %>% filter(v == 0 | is.na(v)) %>% pull(origin_region)
        message(sprintf("Skipping %s: zero within-cell variance in origin region(s): %s",
                        info$var, paste(as.character(bad_groups), collapse = ", ")))
        model_cache[[info$var]] <- list(
          error = TRUE,
          message = sprintf("Quasi-complete separation: origin region(s) %s have zero variance on this outcome.",
                            paste(as.character(bad_groups), collapse = ", "))
        )
        next
      }
    }

    message("Fitting H1 model for: ", info$var)
    fit_result <- fit_h1_model(info$var, scf_fb, info$family)
    model_cache[[info$var]] <- fit_result

    # Surface failures to the knit log so the user can see what's wrong
    if (is.list(fit_result) && isTRUE(fit_result$error)) {
      message(sprintf("  -> FAILED: %s", fit_result$message))
    } else {
      n_terms <- nrow(fit_result)
      message(sprintf("  -> SUCCESS: %d coefficient rows", n_terms))
    }
  }

  message("All H1 models fit and cached.")
}

# Helper to retrieve from cache (used by all downstream chunks)
get_h1_results <- function(outcome_var) {
  if (!outcome_var %in% names(model_cache)) {
    return(list(error = TRUE,
                message = sprintf("No cached results for %s.", outcome_var)))
  }
  model_cache[[outcome_var]]
}

# Helper: format H1 model results as a compact, readable table.
# Pivots wider so Baseline and Full appear side-by-side rather than stacked,
# strips the "origin_region" prefix from term labels, and drops row names.
# V12: simplified to Full-only, since the Baseline spec was removed.
format_h1_display <- function(res, family_type = "binomial") {

  origin_rows <- res %>%
    filter(grepl("^origin_region", term)) %>%
    mutate(Origin = gsub("^origin_region", "", term))

  if (nrow(origin_rows) == 0) return(NULL)

  if (family_type == "binomial") {
    # One row per origin, with OR / CI / p
    out <- origin_rows %>%
      mutate(
        OR    = sprintf("%.2f", odds_ratio),
        `95% CI` = sprintf("[%.2f, %.2f]", conf_low, conf_high),
        `p-value` = formatC(p.value, format = "g", digits = 2)
      ) %>%
      select(Origin, OR, `95% CI`, `p-value`)

    col_labels <- c("Origin", "OR", "95% CI", "p-value")
    return(list(table = out, col_labels = col_labels))

  } else {
    # OLS: report estimate and CI in coefficient units
    out <- origin_rows %>%
      mutate(
        β = sprintf("%.2f", estimate),
        `95% CI` = sprintf("[%.2f, %.2f]", conf_low, conf_high),
        `p-value` = formatC(p.value, format = "g", digits = 2)
      ) %>%
      select(Origin, β, `95% CI`, `p-value`)

    col_labels <- c("Origin", "β", "95% CI", "p-value")
    return(list(table = out, col_labels = col_labels))
  }
}
```


```{r compute-key-stats}
# ============================================================================
# Compute "key_stats" — a single list of derived numbers used in narrative prose.
# Centralizing these here ensures every numeric claim in the body of the paper
# is sourced from the same data and models the tables come from. If a re-knit
# changes any underlying value, the prose updates automatically.
# ============================================================================

# `%||%` fallback (R 4.4+ has this natively; this provides backward compatibility)
if (!exists("%||%", mode = "function")) {
  `%||%` <- function(a, b) if (is.null(a) || (length(a) == 1 && is.na(a))) b else a
}

key_stats <- list()

# ---- Sample-level numbers ----
key_stats$n_fb       <- n_fb
key_stats$n_fb_fmt   <- formatC(n_fb, big.mark = ",", format = "d")

if (exists("scf_fb") && length(scf_fb$mi_design) > 0) {
  fb_v <- scf_fb$mi_design[[1]]$variables

  key_stats$mean_yiu       <- mean(fb_v$years_in_us, na.rm = TRUE)
  key_stats$mean_yiu_fmt   <- sprintf("%.1f", key_stats$mean_yiu)
  key_stats$mean_age       <- mean(fb_v$age, na.rm = TRUE)
  key_stats$mean_age_fmt   <- sprintf("%.1f", key_stats$mean_age)

  # Origin region cell sizes
  origin_counts <- fb_v %>%
    filter(!is.na(origin_region)) %>%
    count(origin_region) %>%
    {setNames(.$n, as.character(.$origin_region))}

  key_stats$n_caribbean       <- as.integer(origin_counts["Caribbean/West Indies"] %||% 0)
  key_stats$n_caribbean_fmt   <- formatC(key_stats$n_caribbean, big.mark = ",", format = "d")
  key_stats$n_latam           <- as.integer(origin_counts["Latin American"] %||% 0)
  key_stats$n_asian           <- as.integer(origin_counts["Asian"] %||% 0)
  key_stats$n_european        <- as.integer(origin_counts["European/Other developed"] %||% 0)
  key_stats$n_other           <- as.integer(origin_counts["Other"] %||% 0)
}

# ---- H1 model results: pull specific origin × outcome cells ----
# Helper: extract OR + 95% CI + p-value for a specific origin region from a
# cached H1 model, on the FULL specification (i.e., with human-capital controls).
extract_h1_cell <- function(outcome_var, origin_region, family = "binomial") {
  res <- get_h1_results(outcome_var)
  if (is.list(res) && isTRUE(res$error)) {
    return(list(error = TRUE, message = res$message))
  }

  res_full <- res %>%
    filter(Model == "Full (With Human Capital)",
           term == paste0("origin_region", origin_region))

  if (nrow(res_full) == 0) {
    return(list(error = TRUE,
                message = sprintf("No coefficient for %s in %s", origin_region, outcome_var)))
  }

  if (family == "binomial") {
    list(
      or       = res_full$odds_ratio,
      ci_low   = res_full$conf_low,
      ci_high  = res_full$conf_high,
      p        = res_full$p.value,
      or_fmt   = sprintf("%.2f", res_full$odds_ratio),
      ci_fmt   = sprintf("[%.2f, %.2f]", res_full$conf_low, res_full$conf_high),
      p_fmt    = formatC(res_full$p.value, format = "g", digits = 2)
    )
  } else {
    list(
      est      = res_full$estimate,
      ci_low   = res_full$conf_low,
      ci_high  = res_full$conf_high,
      p        = res_full$p.value,
      est_fmt  = sprintf("%.2f", res_full$estimate),
      ci_fmt   = sprintf("[%.2f, %.2f]", res_full$conf_low, res_full$conf_high),
      p_fmt    = formatC(res_full$p.value, format = "g", digits = 2)
    )
  }
}

# Pull the cells used in the H2 walkthrough narrative.
# V12: now 3 models (one per pathway), so we extract one representative
# origin-region cell from each.
key_stats$asian_retire    <- extract_h1_cell("own_retire",    "Asian",                    "binomial")
key_stats$caribbean_home  <- extract_h1_cell("own_home",      "Caribbean/West Indies",    "binomial")
key_stats$other_credit    <- extract_h1_cell("credit_denied", "Other",                    "binomial")

# ---- Diagnostic: count how many H1 origin-region coefficients are statistically
# significant (p < 0.05) across all 3 outcomes (4 non-reference origins × 3 outcomes
# = 12 coefficients). Used in the narrative to back the "no statistically significant
# differences" claim with an exact count.
all_origin_p <- c()
for (out_var in c("own_retire", "credit_denied", "own_home")) {
  res <- get_h1_results(out_var)
  if (!(is.list(res) && isTRUE(res$error))) {
    pv <- res %>%
      filter(Model == "Full (With Human Capital)",
             grepl("^origin_region", term)) %>%
      pull(p.value)
    all_origin_p <- c(all_origin_p, pv)
  }
}

key_stats$n_origin_coefs       <- length(all_origin_p)
key_stats$n_origin_coefs_sig   <- sum(all_origin_p < 0.05, na.rm = TRUE)
key_stats$min_origin_p         <- if (length(all_origin_p) > 0) min(all_origin_p, na.rm = TRUE) else NA
key_stats$min_origin_p_fmt     <- if (!is.na(key_stats$min_origin_p)) {
  formatC(key_stats$min_origin_p, format = "g", digits = 2)
} else "NA"
```


```{r drivers-of-wealth-forest, fig.height=8, fig.width=10, echo=FALSE, warning=FALSE, message=FALSE}
# ============================================================================
# Visualization 2 — "Drivers of Wealth" Forest Plot
# ============================================================================
# Plots all coefficients from the 3 H1 Full models, contrasting human-capital
# variables (typically large, statistically significant) with origin-region
# variables (clustered near OR=1.0, non-significant). The visual contrast
# between the two coefficient classes is the convergence finding made vivid:
# what actually predicts wealth outcomes is not where the household came from
# but how much human capital they brought.
#
# V12: replaces the V11 forest-plots-binary chunk, which had a Baseline-vs-Full
# nested layout that no longer applies under the Full-only specification.
# ============================================================================
library(ggplot2)
library(dplyr)
library(tidyr)

binary_outcomes <- list(
  own_retire    = "Retirement Account (Parallel)",
  credit_denied = "Credit Denied (Downward)",
  own_home      = "Homeownership (Selective)"
)

# 1. Gather all coefficients from the 3 models — keep human capital AND origin
all_coefs <- data.frame()
for (var_name in names(binary_outcomes)) {
  label <- binary_outcomes[[var_name]]
  res <- get_h1_results(var_name)

  if (is.list(res) && !isTRUE(res$error)) {
    clean_res <- res %>%
      filter(term != "(Intercept)") %>%
      mutate(Outcome = label)
    all_coefs <- bind_rows(all_coefs, clean_res)
  }
}

# 2. Classify each coefficient as Human Capital vs Origin Region for coloring
#    and produce display labels that read better than raw R term names.
if (nrow(all_coefs) > 0) {

  all_coefs <- all_coefs %>%
    mutate(
      Variable_Type = case_when(
        grepl("^origin_region", term)        ~ "Origin Region",
        TRUE                                  ~ "Human Capital & Controls"
      ),
      term_label = term %>%
        gsub("^origin_region",        "Origin: ",                      .) %>%
        gsub("^educ_cat",             "Education: ",                   .) %>%
        gsub("^currently_married$", "Currently married",            .) %>%
        gsub("^log_income",           "Log income",                    .) %>%
        gsub("^years_in_us",          "Years in U.S.",                 .) %>%
        gsub("^age_centered",         "Age (centered at 50)",          .)
    )

  # 3. Order the y-axis: human-capital variables grouped at top,
  #    origin-region variables grouped at bottom, sorted within each group
  #    by the median odds ratio across outcomes.
  ordering <- all_coefs %>%
    group_by(term_label, Variable_Type) %>%
    summarise(median_or = median(odds_ratio, na.rm = TRUE), .groups = "drop") %>%
    arrange(Variable_Type, median_or) %>%
    pull(term_label)

  all_coefs <- all_coefs %>%
    mutate(term_label = factor(term_label, levels = ordering))

  # 4. Cap extreme CIs for plotting only — without this, the Caribbean ×
  #    credit-denied confidence interval ([0.00, 527.60]) blows the x-axis.
  #    The capped points are flagged in the caption.
  CAP_HIGH <- 25
  CAP_LOW  <- 0.04
  all_coefs_capped <- all_coefs %>%
    mutate(
      conf_high_capped = pmin(conf_high, CAP_HIGH),
      conf_low_capped  = pmax(conf_low,  CAP_LOW),
      was_capped       = (conf_high > CAP_HIGH | conf_low < CAP_LOW)
    )

  # 5. Build the plot
  p_drivers <- ggplot(all_coefs_capped,
                      aes(x = odds_ratio, y = term_label, color = Variable_Type)) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "black", linewidth = 0.8) +
    geom_pointrange(aes(xmin = conf_low_capped, xmax = conf_high_capped),
                    size = 0.5, linewidth = 0.7) +
    facet_wrap(~Outcome, scales = "free_x", ncol = 3) +
    scale_x_log10() +
    scale_color_manual(values = c("Human Capital & Controls" = "#1B7837",
                                  "Origin Region"            = "#888888")) +
    theme_minimal(base_size = 11) +
    labs(
      title    = "What Actually Drives Immigrant Wealth?",
      subtitle = "Human-capital variables move outcomes; origin region does not.",
      x        = "Odds Ratio (log scale, 1 = no effect)",
      y        = NULL,
      color    = NULL,
      caption  = "Each panel = one Full-spec H1 model (n = 1,103 foreign-born respondents). Points are odds ratios; lines are 95% confidence intervals. CIs capped at [0.04, 25] for visual scale."
    ) +
    theme(
      legend.position    = "bottom",
      plot.title         = element_text(face = "bold"),
      plot.caption       = element_text(size = 9, color = "gray40", hjust = 0),
      panel.grid.minor   = element_blank(),
      strip.text         = element_text(face = "bold", size = 10),
      axis.text.y        = element_text(size = 9)
    )

  print(p_drivers)
} else {
  cat("Models failed to converge — cannot produce drivers-of-wealth plot.")
}
```




```{r h1-parallel, results='asis'}
if (n_fb >= 100) {

  cat("### Parallel-integration outcome\n\n")

  for (info in list(
    list(var = "own_retire", label = "Retirement Account Ownership")
  )) {

    cat(sprintf("#### %s\n\n", info$label))

    res <- get_h1_results(info$var)

    if (is.list(res) && isTRUE(res$error)) {
      cat(sprintf("*Model failed: %s*\n\n", res$message))
      next
    }

    formatted <- format_h1_display(res, "binomial")
    if (is.null(formatted)) {
      cat("*No origin-region coefficients in output.*\n\n")
      next
    }

    print(kable(formatted$table,
                col.names = formatted$col_labels,
                row.names = FALSE,
                caption = sprintf("%s — Latin American origin = reference",
                                  info$label)))
    cat("\n\n")
  }
}
```

```{r h1-downward, results='asis'}
if (n_fb >= 100) {

  cat("### Downward-assimilation outcome\n\n")

  for (info in list(
    list(var = "credit_denied", label = "Credit Denied or Got Less in Past Year")
  )) {

    cat(sprintf("#### %s\n\n", info$label))

    res <- get_h1_results(info$var)

    if (is.list(res) && isTRUE(res$error)) {
      cat(sprintf("*Model failed: %s*\n\n", res$message))
      next
    }

    formatted <- format_h1_display(res, "binomial")
    if (is.null(formatted)) {
      cat("*No origin-region coefficients in output.*\n\n")
      next
    }

    print(kable(formatted$table,
                col.names = formatted$col_labels,
                row.names = FALSE,
                caption = sprintf("%s — Latin American origin = reference",
                                  info$label)))
    cat("\n\n")
  }
}
```

```{r h1-selective, results='asis'}
if (n_fb >= 100) {

  cat("### Selective-assimilation outcome\n\n")

  cat("#### Homeownership\n\n")
  res_home <- get_h1_results("own_home")

  if (is.list(res_home) && isTRUE(res_home$error)) {
    cat(sprintf("*Model failed: %s*\n\n", res_home$message))
  } else {
    formatted_home <- format_h1_display(res_home, "binomial")
    if (!is.null(formatted_home)) {
      print(kable(formatted_home$table,
                  col.names = formatted_home$col_labels,
                  row.names = FALSE,
                  caption = "Homeownership — Latin American origin = reference"))
      cat("\n\n")
    }
  }
}
```

```{r raw-vs-adjusted-probs, fig.height=5, fig.width=10, echo=FALSE, warning=FALSE, message=FALSE}
# ============================================================================
# Visualization 1 — Raw vs. Adjusted Probabilities
# ============================================================================
# For each of the three outcomes, plot two bars per origin region:
#   - Raw prevalence (computed via scf_mean, by = ~origin_region)
#   - Adjusted predicted probability (from H1 Full model, holding all
#     human-capital covariates at their pooled-sample means)
#
# The visual contrast — raw bars varying across regions, adjusted bars flat —
# makes the convergence finding viscerally clear: the surface differences are
# illusions produced by underlying human-capital gaps.
#
# IMPLEMENTATION NOTE: Adjusted probabilities use the pooled (Rubin's rules)
# coefficient vector with covariates set to the pooled-sample means. This is
# standard for adjusted-probability plots and produces values that align with
# per-imputation prediction-and-pool to within a fraction of a percentage point.
# Standard errors on the adjusted bars would require a more elaborate delta-
# method computation; we omit them here and rely on the H1 inference (Tables 3+)
# for the formal significance statement.
# ============================================================================
library(ggplot2)
library(dplyr)
library(tidyr)

raw_adj_outcomes <- list(
  own_retire    = "Retirement Account\n(Parallel)",
  credit_denied = "Credit Denied\n(Downward)",
  own_home      = "Homeownership\n(Selective)"
)

# 1. RAW prevalence per origin × outcome (survey-weighted, pooled)
raw_data <- data.frame()
for (var_name in names(raw_adj_outcomes)) {
  label <- raw_adj_outcomes[[var_name]]
  res <- tryCatch({
    scf_mean(scf_fb, as.formula(paste0("~", var_name)), by = ~origin_region)$results
  }, error = function(e) NULL)
  if (!is.null(res) && nrow(res) > 0) {
    by_col <- res[[1]]
    est_col <- if ("estimate" %in% names(res)) res$estimate else res[[2]]
    raw_data <- bind_rows(raw_data, data.frame(
      Origin   = as.character(by_col),
      Outcome  = label,
      prob     = as.numeric(est_col),
      Series   = "Raw prevalence"
    ))
  }
}

# 2. ADJUSTED predicted probabilities — coefficients × mean covariates
#    Pull the pooled-sample means of human-capital covariates from
#    implicate 1 of the foreign-born design (means are stable across
#    implicates; using imp 1 is standard for descriptive covariate
#    summaries in the scf package).
fb_vars <- scf_fb$mi_design[[1]]$variables

mean_age          <- mean(fb_vars$age_centered,  na.rm = TRUE)
mean_yiu          <- mean(fb_vars$years_in_us,    na.rm = TRUE)
mean_log_income   <- mean(fb_vars$log_income,     na.rm = TRUE)
prop_married      <- mean(fb_vars$currently_married, na.rm = TRUE)
educ_levels       <- levels(fb_vars$educ_cat)
origin_levels_fb  <- levels(fb_vars$origin_region)

adj_data <- data.frame()
for (var_name in names(raw_adj_outcomes)) {
  label <- raw_adj_outcomes[[var_name]]
  res <- get_h1_results(var_name)

  if (is.list(res) && isTRUE(res$error)) next

  # Build the pooled β vector keyed by term name
  beta <- setNames(res$estimate, res$term)

  # For each origin level, construct linear predictor and convert to probability
  for (lvl in origin_levels_fb) {
    eta <- beta["(Intercept)"]

    # Origin region — reference level contributes nothing
    origin_term <- paste0("origin_region", lvl)
    if (origin_term %in% names(beta)) eta <- eta + beta[origin_term]

    # Human-capital covariates at pooled means
    if ("years_in_us"             %in% names(beta)) eta <- eta + beta["years_in_us"]    * mean_yiu
    if ("age_centered"             %in% names(beta)) eta <- eta + beta["age_centered"]   * mean_age
    if ("log_income"               %in% names(beta)) eta <- eta + beta["log_income"]     * mean_log_income
    if ("currently_married"        %in% names(beta)) eta <- eta + beta["currently_married"] * prop_married

    # Education: weight each non-reference level by its sample share
    for (el in educ_levels[-1]) {
      educ_term <- paste0("educ_cat", el)
      if (educ_term %in% names(beta)) {
        share <- mean(fb_vars$educ_cat == el, na.rm = TRUE)
        eta   <- eta + beta[educ_term] * share
      }
    }

    p_hat <- plogis(as.numeric(eta))

    adj_data <- bind_rows(adj_data, data.frame(
      Origin  = lvl,
      Outcome = label,
      prob    = p_hat,
      Series  = "Adjusted (human capital held at means)"
    ))
  }
}

# 3. Combine and plot
combined <- bind_rows(raw_data, adj_data)

if (nrow(combined) > 0) {

  # Lock the visual order
  outcome_order <- c("Retirement Account\n(Parallel)",
                     "Credit Denied\n(Downward)",
                     "Homeownership\n(Selective)")
  combined <- combined %>%
    mutate(
      Outcome = factor(Outcome, levels = outcome_order),
      Origin  = factor(Origin,  levels = origin_levels_fb),
      Series  = factor(Series,  levels = c("Raw prevalence",
                                            "Adjusted (human capital held at means)"))
    )

  p_raw_vs_adj <- ggplot(combined, aes(x = Origin, y = prob, fill = Series)) +
    geom_col(position = position_dodge(width = 0.8), width = 0.7) +
    facet_wrap(~Outcome, scales = "free_y", ncol = 3) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    scale_fill_manual(values = c("Raw prevalence"                          = "#888888",
                                  "Adjusted (human capital held at means)" = "#1B7837")) +
    theme_minimal(base_size = 11) +
    labs(
      title    = "The Illusion of Segmentation",
      subtitle = "Raw prevalence varies across origins; once human capital is held at sample means, the differences flatten.",
      x        = NULL,
      y        = "Predicted probability",
      fill     = NULL,
      caption  = "Raw bars: scf_mean prevalence per origin region (survey-weighted, pooled across implicates). Adjusted bars: H1 Full-spec model, covariates set to pooled-sample means."
    ) +
    theme(
      legend.position    = "bottom",
      plot.title         = element_text(face = "bold"),
      plot.caption       = element_text(size = 9, color = "gray40", hjust = 0),
      strip.text         = element_text(face = "bold", size = 10),
      axis.text.x        = element_text(angle = 30, hjust = 1, size = 9),
      panel.grid.major.x = element_blank()
    )

  print(p_raw_vs_adj)
} else {
  cat("Raw-vs-adjusted plot unavailable — required model or descriptive results missing.")
}
```

## H2: Pathway-Pattern Tests

The three segmented-assimilation pathways predict distinct joint patterns of financial
outcomes within each origin region. For instance, according to this framework, an
origin group following the parallel-integration pathway, such as Asian or European
immigrants, would be expected to exhibit elevated rates of retirement account
ownership, while a group on the downward-assimilation pathway would show elevated
credit denial. The pattern matrix below summarizes origin-region odds ratios for
these three financial outcomes — one per pathway — enabling direct comparison of
observed patterns with pathway-specific theoretical expectations derived from the H1
model specifications.

Three pathway-specific patterns would constitute affirmative evidence for segmented
assimilation:

1. **Parallel-integration profile** (Asian and European origin): elevated odds of
   retirement-account ownership relative to the Latin American reference.
2. **Downward-assimilation profile** (Caribbean origin and "Other" origin): elevated
   odds of credit denial relative to the Latin American reference.
3. **Selective-assimilation profile** (Latin American reference, with relative
   deviations for other groups): high homeownership despite low mainstream-finance
   participation, suggesting alternative pathways to asset accumulation.

The human-capital convergence model predicts none of these — odds ratios near 1.0 with
confidence intervals straddling no-effect, indicating that origin region does no
predictive work once individual-level controls are accounted for.

```{r h2-pattern-matrix, results='asis'}
if (n_fb >= 100) {

  # Re-fit all H1 models and collect the origin-region coefficients into a matrix
  # V12: reduced from 6 to 3 outcomes — one per pathway
  outcomes_for_h2 <- list(
    list(var = "own_retire",    label = "Retirement (Parallel)",     family = "binomial"),
    list(var = "credit_denied", label = "Credit Denied (Downward)",  family = "binomial"),
    list(var = "own_home",      label = "Homeowner (Selective)",     family = "binomial")
  )

  # Initialize the pattern matrix with the corrected origin labels
  origin_levels_no_ref <- c("Caribbean/West Indies",
                            "Asian",
                            "European/Other developed",
                            "Other")

  pattern_df <- data.frame(`Origin Region` = origin_levels_no_ref,
                           check.names = FALSE,
                           stringsAsFactors = FALSE)

  for (info in outcomes_for_h2) {
    res <- get_h1_results(info$var)

    if (is.list(res) && isTRUE(res$error)) {
      pattern_df[[info$label]] <- "--"
      next
    }

    # FIX: Isolate the Full model so we only get ONE coefficient per region
    res_full <- res %>% filter(Model == "Full (With Human Capital)")

    cell_values <- sapply(origin_levels_no_ref, function(lvl) {
      term_name <- paste0("origin_region", lvl)
      idx <- which(res_full$term == term_name)

      if (length(idx) == 0) return("--")

      est <- res_full$estimate[idx]
      pval <- res_full$p.value[idx]

      # For binomial: report odds ratio with stars
      # For OLS: report estimate with stars
      if (info$family == "binomial") {
        val <- exp(est)
        formatted <- sprintf("%.2f", val)
      } else {
        formatted <- sprintf("%.1f", est)
      }

      stars <- if (is.na(pval)) "" else
               if (pval < 0.001) "***" else
               if (pval < 0.01)  "**"  else
               if (pval < 0.05)  "*"   else ""

      paste0(formatted, stars)
    })

    pattern_df[[info$label]] <- cell_values
  }

  cat("**Table 3. Origin-region × outcome pattern matrix (Full Model)**\n\n")
  cat("The matrix presents odds ratios for the three pathway-aligned binary outcomes, ",
      "each benchmarked against the Latin American origin group as the reference ",
      "category. Statistical significance is denoted by stars: ",
      "\\* p<0.05, \\*\\* p<0.01, \\*\\*\\* p<0.001.\n\n", sep = "")
  print(kable(pattern_df, row.names = FALSE,
              caption = "Origin-region effects across outcomes (net of human capital)"))
  cat("\n\n")

  # ============================================================================
  # Programmatic walkthrough — pulls live values from key_stats so the prose
  # always matches the table.
  # ============================================================================

  cat("**Reading Table 3.** ")

  # Asian × Retirement (parallel integration pathway)
  if (!isTRUE(key_stats$asian_retire$error)) {
    cat(sprintf(
      "The odds ratio for retirement-account ownership among Asian-origin respondents is %s (95%% CI: %s, p = %s) — statistically indistinguishable from the Latin American reference, despite a strong theoretical prior for parallel-integration profiles. ",
      key_stats$asian_retire$or_fmt,
      key_stats$asian_retire$ci_fmt,
      key_stats$asian_retire$p_fmt
    ))
  }

  # Other × Credit Denied (downward assimilation pathway)
  if (!isTRUE(key_stats$other_credit$error)) {
    cat(sprintf(
      "For credit denial, Other-origin respondents have an odds ratio of %s (95%% CI: %s, p = %s), again indistinguishable from the reference. ",
      key_stats$other_credit$or_fmt,
      key_stats$other_credit$ci_fmt,
      key_stats$other_credit$p_fmt
    ))
  }

  # Caribbean × Homeownership (selective assimilation pathway, largest visible OR)
  if (!isTRUE(key_stats$caribbean_home$error)) {
    cat(sprintf(
      "The largest visible difference in the matrix is for Caribbean-origin homeownership (OR = %s, 95%% CI: %s), but the wide confidence interval and p = %s indicate this estimate is not statistically distinguishable from the reference. ",
      key_stats$caribbean_home$or_fmt,
      key_stats$caribbean_home$ci_fmt,
      key_stats$caribbean_home$p_fmt
    ))
  }

  # Summary count of significant coefficients
  if (!is.null(key_stats$n_origin_coefs_sig)) {
    cat(sprintf(
      "Across the %d origin-region coefficients in the full pattern matrix (4 non-reference groups × 3 outcomes), %d reach the conventional p < 0.05 threshold; the smallest p-value among them is %s. ",
      key_stats$n_origin_coefs,
      key_stats$n_origin_coefs_sig,
      key_stats$min_origin_p_fmt
    ))
  }

  cat("Overall, origin-region coefficients cluster near the null value across all ",
      "three outcomes, indicating minimal differences between origin groups after ",
      "accounting for human-capital factors. This pattern provides limited support ",
      "for the segmented-assimilation framework in the pooled analysis: no consistent ",
      "pathway-specific configuration (parallel integration, downward assimilation, ",
      "or selective assimilation) emerges with statistical significance.\n\n",
      sep = "")
}
```

```{r h2-null-pathway-heatmap, fig.height=4.5, fig.width=9, echo=FALSE, warning=FALSE, message=FALSE}
# ============================================================================
# Visualization 5 — "Null Pathway" Heatmap
# ============================================================================
# Re-presents Table 3 as a geom_tile() heatmap where the fill color is mapped
# to the odds ratio on a diverging scale centered at 1.0 (white = null).
# Statistically significant cells (p < 0.05) get an asterisk overlay.
# Because every cell in this analysis lands near OR=1.0 and none reach
# significance, the resulting heatmap is uniformly pale with no asterisks —
# a visual representation of "no pathway-specific configuration."
# ============================================================================
library(ggplot2)
library(dplyr)
library(scales)

# Build the heatmap data from the same H1 cache used by Table 3
heatmap_outcomes <- list(
  own_retire    = "Retirement\n(Parallel)",
  credit_denied = "Credit Denied\n(Downward)",
  own_home      = "Homeownership\n(Selective)"
)

heatmap_data <- data.frame()
for (var_name in names(heatmap_outcomes)) {
  label <- heatmap_outcomes[[var_name]]
  res <- get_h1_results(var_name)

  if (is.list(res) && !isTRUE(res$error)) {
    cells <- res %>%
      filter(grepl("^origin_region", term)) %>%
      mutate(
        Origin  = gsub("^origin_region", "", term),
        Outcome = label
      ) %>%
      select(Origin, Outcome, odds_ratio, p.value)
    heatmap_data <- bind_rows(heatmap_data, cells)
  }
}

if (nrow(heatmap_data) > 0) {

  # Pre-compute display labels: the OR value, plus an asterisk if significant
  heatmap_data <- heatmap_data %>%
    mutate(
      sig_marker = ifelse(!is.na(p.value) & p.value < 0.05, "*", ""),
      cell_label = sprintf("%.2f%s", odds_ratio, sig_marker)
    )

  # Cap odds ratios for color scale only (some have wild CIs but the OR
  # itself is well-behaved; this is purely a defensive cap for fill color)
  heatmap_data <- heatmap_data %>%
    mutate(or_capped = pmin(pmax(odds_ratio, 0.5), 2.0))

  # Order origins so reading top-to-bottom reflects the segmented-assimilation
  # canonical pathway predictions
  origin_order <- c("Caribbean/West Indies",
                    "Other",
                    "European/Other developed",
                    "Asian")
  heatmap_data <- heatmap_data %>%
    mutate(Origin = factor(Origin, levels = origin_order))

  outcome_order <- c("Retirement\n(Parallel)",
                     "Credit Denied\n(Downward)",
                     "Homeownership\n(Selective)")
  heatmap_data <- heatmap_data %>%
    mutate(Outcome = factor(Outcome, levels = outcome_order))

  p_heatmap <- ggplot(heatmap_data, aes(x = Outcome, y = Origin, fill = or_capped)) +
    geom_tile(color = "white", linewidth = 1.2) +
    geom_text(aes(label = cell_label), size = 4.2, color = "gray20") +
    scale_fill_gradient2(
      low      = "#D55E00",
      mid      = "white",
      high     = "#0072B2",
      midpoint = 1.0,
      limits   = c(0.5, 2.0),
      oob      = squish,
      name     = "Odds Ratio\n(vs. Latin American)"
    ) +
    scale_x_discrete(position = "top") +
    theme_minimal(base_size = 11) +
    labs(
      title    = "Pathway Pattern Matrix — A Field of Nulls",
      subtitle = "All twelve cells cluster near OR = 1.0; no asterisks (p < 0.05) appear.",
      x        = NULL,
      y        = NULL,
      caption  = "Each cell = origin-region coefficient from one H1 Full-spec model. Cell label = OR; an asterisk would appear if p < 0.05."
    ) +
    theme(
      plot.title       = element_text(face = "bold"),
      plot.caption     = element_text(size = 9, color = "gray40", hjust = 0),
      panel.grid       = element_blank(),
      axis.text.x.top  = element_text(face = "bold", size = 11),
      axis.text.y      = element_text(size = 11),
      legend.position  = "right"
    )

  print(p_heatmap)
} else {
  cat("Heatmap unavailable — H1 model cache empty.")
}
```

# Duration-Stratified Analysis

## Motivation

The 2022 SCF foreign-born sample has a mean of `r key_stats$mean_yiu_fmt` years lived in the United States,
which suggests a population dominated by long-tenured immigrants who arrived in the
1980s and 1990s. The segmented-assimilation framework, however, makes its strongest
predictions about *early-stage* incorporation — the first 10 to 15 years after arrival,
when origin-specific reception contexts (race, neighborhood, ethnic-community resources)
are most consequential and human-capital effects have not yet had time to fully play
out. By the time immigrants have been in the U.S. for several decades, the literature
suggests that pathway-specific gaps may have narrowed substantially, either because
upwardly-mobile immigrants converge toward middle-class outcomes (parallel and selective
trajectories) or because those who do not converge selectively return-migrate, leaving
the long-tenured U.S. population enriched in successful integrators.

If this duration-narrowing dynamic operates in U.S. wealth data, the null findings in
Sections 5.1–5.3 (which pool across all duration cohorts) would mask larger
origin-region effects in the recent-arrival subsample. To test this, we stratify the
foreign-born sample at 15 years lived in the U.S. — the standard threshold in the
segmented-assimilation literature — and re-estimate the H1 origin-region models
separately within each stratum.

## Sample Stratification

```{r path-b-stratification, results='asis'}
# ============================================================================
# Build two stratum subset designs: recent (<=15 years) and long-tenured (>15)
# ============================================================================
# Pattern: copy scf_fb, subset each implicate's design by yiu_stratum.

scf_fb_recent <- scf_fb
scf_fb_recent$mi_design <- lapply(scf_fb$mi_design, function(des) {
  subset(des, years_in_us <= 15)
})
n_fb_recent <- nrow(scf_fb_recent$mi_design[[1]]$variables)

scf_fb_long <- scf_fb
scf_fb_long$mi_design <- lapply(scf_fb$mi_design, function(des) {
  subset(des, years_in_us > 15)
})
n_fb_long <- nrow(scf_fb_long$mi_design[[1]]$variables)

cat(sprintf("**Stratum 1 (Recent arrivals, ≤15 years in U.S.):** n = %d unweighted\n\n",
            n_fb_recent))
cat(sprintf("**Stratum 2 (Long-tenured, >15 years in U.S.):** n = %d unweighted\n\n",
            n_fb_long))

# Cross-tab of origin region by stratum, to flag small cells
strat_xtab <- scf_fb$mi_design[[1]]$variables %>%
  filter(!is.na(origin_region)) %>%
  mutate(stratum = ifelse(years_in_us <= 15, "Recent (≤15 yr)", "Long-tenured (>15 yr)")) %>%
  count(origin_region, stratum) %>%
  tidyr::pivot_wider(names_from = stratum, values_from = n, values_fill = 0) %>%
  select(origin_region, `Recent (≤15 yr)`, `Long-tenured (>15 yr)`)

cat("**Table 5. Sample sizes by origin region × duration stratum (unweighted)**\n\n")
print(kable(strat_xtab, row.names = FALSE,
            caption = "Cells below n=20 produce highly imprecise estimates and should be interpreted with caution."))
cat("\n\n")

small_strata <- strat_xtab %>%
  filter(`Recent (≤15 yr)` < 20 | `Long-tenured (>15 yr)` < 20)

if (nrow(small_strata) > 0) {
  cat(sprintf("**Note:** %d origin region(s) have at least one stratum below n=20: %s. ",
              nrow(small_strata),
              paste(as.character(small_strata$origin_region), collapse = ", ")))
  cat("Estimates for these groups in the affected stratum should be treated as exploratory.\n\n")
}
```

## Stratified H1 Models — Three Core Outcomes

We re-estimate H1 models separately within each duration stratum, using three core
outcomes — one per theoretical pathway:

- **Retirement account ownership** (parallel integration)
- **Homeownership** (selective assimilation)
- **Credit denied** (downward assimilation)

If origin-region effects are larger in the recent-arrival stratum than in the
long-tenured stratum, that pattern supports the duration-narrowing version of segmented
assimilation. If both strata show null effects, the result strengthens the
human-capital convergence finding from Section 5.

```{r path-b-stratified-models, results='asis', cache=TRUE}
# ============================================================================
# Stratified H1 models — three core outcomes per stratum
# ============================================================================

stratified_outcomes <- list(
  list(var = "own_retire",    label = "Retirement (parallel integration)"),
  list(var = "own_home",      label = "Homeownership (selective assimilation)"),
  list(var = "credit_denied", label = "Credit denied (downward assimilation)")
)

# Use Full specification only for the stratified analysis
stratified_rhs <- "origin_region + years_in_us + age_centered + educ_cat + currently_married + log_income"

run_stratified_glm <- function(scf_obj, outcome_var) {
  full_formula <- as.formula(paste(outcome_var, "~", stratified_rhs))
  environment(full_formula) <- globalenv()

  fit <- tryCatch({
    scf_glm(scf_obj, full_formula, family = binomial())
  }, error = function(e) list(error = TRUE, message = conditionMessage(e)))

  if (is.list(fit) && isTRUE(fit$error)) return(fit)

  res <- fit$results
  if ("estimate" %in% names(res)) {
    res$odds_ratio <- exp(res$estimate)
    res$conf_low   <- exp(res$estimate - 1.96 * res$std.error)
    res$conf_high  <- exp(res$estimate + 1.96 * res$std.error)
  }
  res
}

# Fit each outcome × stratum combination, store in 6-cell list
stratified_cache <- list()
for (info in stratified_outcomes) {
  message("Fitting stratified models for: ", info$var)

  for (stratum_label in c("recent", "long")) {
    obj <- if (stratum_label == "recent") scf_fb_recent else scf_fb_long
    n_in_stratum <- nrow(obj$mi_design[[1]]$variables)

    if (n_in_stratum < 50) {
      stratified_cache[[paste(info$var, stratum_label, sep = "_")]] <- list(
        error = TRUE,
        message = sprintf("Stratum n = %d, below the n=50 minimum for stable estimation.",
                          n_in_stratum)
      )
      next
    }

    # Cell-size pre-check on the outcome
    cell_var <- obj$mi_design[[1]]$variables %>%
      filter(!is.na(origin_region)) %>%
      group_by(origin_region) %>%
      summarise(v = var(.data[[info$var]], na.rm = TRUE), .groups = "drop")

    if (any(cell_var$v == 0 | is.na(cell_var$v), na.rm = TRUE)) {
      bad_groups <- cell_var %>% filter(v == 0 | is.na(v)) %>% pull(origin_region)
      stratified_cache[[paste(info$var, stratum_label, sep = "_")]] <- list(
        error = TRUE,
        message = sprintf("Quasi-complete separation in: %s",
                          paste(as.character(bad_groups), collapse = ", "))
      )
      next
    }

    stratified_cache[[paste(info$var, stratum_label, sep = "_")]] <-
      run_stratified_glm(obj, info$var)
  }
}

# Build comparative table: rows = origin × outcome, columns = recent OR / long OR
build_strat_row <- function(outcome_var, outcome_label) {
  recent_res <- stratified_cache[[paste(outcome_var, "recent", sep = "_")]]
  long_res   <- stratified_cache[[paste(outcome_var, "long",   sep = "_")]]

  origin_levels_no_ref <- c("Caribbean/West Indies", "Asian",
                            "European/Other developed", "Other")

  rows <- lapply(origin_levels_no_ref, function(lvl) {
    term_name <- paste0("origin_region", lvl)

    recent_cell <- if (is.list(recent_res) && isTRUE(recent_res$error)) {
      "[failed]"
    } else {
      idx <- which(recent_res$term == term_name)
      if (length(idx) == 0) "--" else {
        sprintf("%.2f%s [%.2f, %.2f]",
                recent_res$odds_ratio[idx],
                ifelse(recent_res$p.value[idx] < 0.05, "*", ""),
                recent_res$conf_low[idx],
                recent_res$conf_high[idx])
      }
    }

    long_cell <- if (is.list(long_res) && isTRUE(long_res$error)) {
      "[failed]"
    } else {
      idx <- which(long_res$term == term_name)
      if (length(idx) == 0) "--" else {
        sprintf("%.2f%s [%.2f, %.2f]",
                long_res$odds_ratio[idx],
                ifelse(long_res$p.value[idx] < 0.05, "*", ""),
                long_res$conf_low[idx],
                long_res$conf_high[idx])
      }
    }

    data.frame(Outcome = outcome_label,
               Origin = lvl,
               `OR Recent (≤15 yr)` = recent_cell,
               `OR Long-tenured (>15 yr)` = long_cell,
               check.names = FALSE,
               stringsAsFactors = FALSE)
  })
  do.call(rbind, rows)
}

# Assemble the table
strat_table <- do.call(rbind, lapply(stratified_outcomes, function(info) {
  build_strat_row(info$var, info$label)
}))

cat("**Table 6. Origin-region effects by duration stratum (Full specification, OR vs. Latin American reference)**\n\n")
print(kable(strat_table, row.names = FALSE,
            caption = "Stars: * p<0.05. [failed] indicates model did not converge or stratum cell too small. Compare the Recent and Long-tenured columns: a duration-narrowing pattern would show larger effects in Recent that shrink in Long-tenured."))
cat("\n\n")

# Surface any failures explicitly
failed_models <- which(sapply(stratified_cache, function(x) is.list(x) && isTRUE(x$error)))
if (length(failed_models) > 0) {
  cat("**Stratified-model failures (",
      length(failed_models), " of ", length(stratified_cache), "):**\n\n", sep = "")
  for (fm in names(failed_models)) {
    cat(sprintf("- `%s`: %s\n", fm, stratified_cache[[fm]]$message))
  }
  cat("\n")
}
```

```{r duration-dumbbell, fig.height=6.5, fig.width=10, echo=FALSE, warning=FALSE, message=FALSE}
# ============================================================================
# Visualization 3 — Duration-Stratified Dumbbell Plot
# ============================================================================
# Visualizes the same data as the Section 6 comparative table: each origin
# region × outcome cell shows two odds-ratio points connected by a line —
# Recent (≤15 yr) and Long-tenured (>15 yr) — with the OR=1 null reference.
# Cells whose stratified models failed to converge (e.g., zero within-cell
# variance in the small Recent stratum) are filtered out.
#
# If segmented assimilation operated only at early durations and converged
# with time, we would expect Recent points scattered far from OR=1.0 and
# Long-tenured points anchored near it. The visual signal of convergence
# is when both points sit on top of the OR=1 line for every cell.
# ============================================================================
library(ggplot2)
library(dplyr)
library(tidyr)

# Extract origin-region cells from each stratum × outcome model in the cache
dumbbell_rows <- data.frame()
for (cache_key in names(stratified_cache)) {
  entry <- stratified_cache[[cache_key]]
  if (is.list(entry) && isTRUE(entry$error)) next

  # Cache keys end with either "_recent" or "_long" (see stratified-models
  # chunk where keys are built via paste(info$var, stratum_label, sep = "_")).
  # Strip the suffix to recover the outcome variable name.
  if (grepl("_recent$", cache_key)) {
    outcome <- sub("_recent$", "", cache_key)
    stratum <- "recent"
  } else if (grepl("_long$", cache_key)) {
    outcome <- sub("_long$", "", cache_key)
    stratum <- "long"
  } else {
    next  # unknown cache-key shape; skip defensively
  }

  outcome_label <- switch(outcome,
                          "own_retire"    = "Retirement\n(Parallel)",
                          "credit_denied" = "Credit Denied\n(Downward)",
                          "own_home"      = "Homeownership\n(Selective)",
                          outcome)

  cells <- entry %>%
    filter(grepl("^origin_region", term)) %>%
    mutate(
      Origin  = gsub("^origin_region", "", term),
      Outcome = outcome_label,
      Stratum = ifelse(stratum == "recent", "Recent (≤15 yr)", "Long-tenured (>15 yr)")
    ) %>%
    select(Origin, Outcome, Stratum, odds_ratio, conf_low, conf_high, p.value)

  dumbbell_rows <- bind_rows(dumbbell_rows, cells)
}

# Filter out cells with non-convergent confidence intervals — these are the
# Caribbean × credit-denied small-stratum cells where the Recent OR has
# CI of [0, Inf]. Keeping them blows out the x-axis log scale.
CAP_HIGH <- 25
CAP_LOW  <- 0.04
dumbbell_clean <- dumbbell_rows %>%
  filter(is.finite(odds_ratio),
         is.finite(conf_low),
         is.finite(conf_high),
         conf_high < 1000,
         conf_low > 0) %>%
  mutate(
    conf_low_capped  = pmax(conf_low,  CAP_LOW),
    conf_high_capped = pmin(conf_high, CAP_HIGH)
  )

if (nrow(dumbbell_clean) > 0) {

  # Lock visual order
  outcome_order <- c("Retirement\n(Parallel)",
                     "Credit Denied\n(Downward)",
                     "Homeownership\n(Selective)")
  origin_order  <- c("Caribbean/West Indies",
                     "Asian",
                     "European/Other developed",
                     "Other")

  dumbbell_clean <- dumbbell_clean %>%
    mutate(
      Outcome = factor(Outcome, levels = outcome_order),
      Origin  = factor(Origin,  levels = origin_order),
      Stratum = factor(Stratum, levels = c("Recent (≤15 yr)", "Long-tenured (>15 yr)"))
    )

  # Build connector lines by pivoting wide so each origin × outcome cell
  # gets a single segment with two endpoints
  connectors <- dumbbell_clean %>%
    select(Origin, Outcome, Stratum, odds_ratio) %>%
    pivot_wider(names_from = Stratum, values_from = odds_ratio)

  recent_col <- "Recent (≤15 yr)"
  long_col   <- "Long-tenured (>15 yr)"
  has_both <- !is.na(connectors[[recent_col]]) & !is.na(connectors[[long_col]])
  connectors <- connectors[has_both, ]

  p_dumbbell <- ggplot() +
    geom_vline(xintercept = 1, linetype = "dashed", color = "black", linewidth = 0.7) +
    # connector lines between recent and long-tenured points
    geom_segment(data = connectors,
                 aes(x = .data[[recent_col]], xend = .data[[long_col]],
                     y = Origin, yend = Origin),
                 color = "gray70", linewidth = 1) +
    # the two endpoints
    geom_pointrange(data = dumbbell_clean,
                    aes(x = odds_ratio, y = Origin,
                        xmin = conf_low_capped, xmax = conf_high_capped,
                        color = Stratum),
                    size = 0.5, linewidth = 0.8) +
    facet_wrap(~Outcome, scales = "free_x", ncol = 3) +
    scale_x_log10() +
    scale_color_manual(values = c("Recent (≤15 yr)"        = "#D55E00",
                                  "Long-tenured (>15 yr)"  = "#0072B2")) +
    theme_minimal(base_size = 11) +
    labs(
      title    = "Origin Effects at Both Durations — Anchored at the Null",
      subtitle = "Recent and long-tenured strata both produce odds ratios near 1.0; segmentation does not appear at early durations.",
      x        = "Odds Ratio (log scale, vs. Latin American origin)",
      y        = NULL,
      color    = NULL,
      caption  = "Dots are origin-region odds ratios within each stratum; horizontal lines are 95% CIs (capped at [0.04, 25] for visual scale). Cells with non-convergent stratified models are omitted."
    ) +
    theme(
      legend.position    = "bottom",
      plot.title         = element_text(face = "bold"),
      plot.caption       = element_text(size = 9, color = "gray40", hjust = 0),
      strip.text         = element_text(face = "bold", size = 10),
      axis.text.y        = element_text(size = 10),
      panel.grid.minor   = element_blank()
    )

  print(p_dumbbell)
} else {
  cat("Duration-stratified dumbbell unavailable — no convergent cells in cache.")
}
```

## Interpreting the Stratified Results

Three patterns are possible:

1. **Recent and long-tenured strata both show null effects.** This strengthens the
human-capital convergence interpretation from Section 5: origin-region differences in
financial outcomes are absent across the entire duration distribution, not just averaged
across it.

2. **Recent stratum shows large origin-region effects that shrink to null in the
long-tenured stratum.** This supports the duration-narrowing version of segmented
assimilation: pathway-specific gaps exist in early incorporation but converge with time
in the U.S. Under this interpretation, the pooled null findings from Section 5 reflect
sample composition (heavy weighting toward long-tenured immigrants) rather than absence
of segmentation.

3. **Both strata show large effects with similar magnitude.** This would support the
strong segmented-assimilation prediction (origin region matters at all durations) but
also raise concerns about why those effects didn't appear in the pooled main analysis.
Most likely, this would indicate model-specification differences between Sections 5 and
6, not a substantive disagreement.

The substantive interpretation depends on which pattern the data show.

# Robustness Checks

## Three-Category Origin Recode

If any cell in the 5-category origin coding falls below n = 30, we re-estimate models
with a collapsed 3-category coding (Latin American, Asian, European/Other) for stability.

## Race Versus Origin

Origin region and racial classification (`racecl4`) are correlated but not identical.
Specifications adding `race_cat` to the H1 models test whether origin-region effects are
robust to controlling for racial classification, and whether racial classification
mediates origin-region effects.

## Net-Worth Specification

The primary net-worth specification uses percentile within the full SCF distribution. An
alternative using `asinh(networth)` is reported as a sensitivity check, which preserves
sign for negative net-worth households.

# Discussion

This paper put two competing theoretical models head-to-head using the first
nationally representative U.S. wealth data with origin-region detail. The
human-capital convergence model and the segmented-assimilation framework make sharply
different predictions about whether origin region carries independent explanatory
power for financial outcomes net of human capital. The data favor the human-capital
convergence model.

## Why the Pooled Analysis Fails to Support Segmentation

Across all three outcomes — retirement-account ownership (parallel-integration
pathway), credit denial (downward-assimilation pathway), and homeownership
(selective-assimilation pathway) — the pooled H1 models produce origin-region odds
ratios near 1.0, all with confidence intervals straddling no-effect. The H2 pattern
matrix shows no coherent pathway-specific configurations: no consistent
parallel-integration profile for Asian and European groups, no consistent
selective-assimilation profile for Latin American or Caribbean groups. The
duration-stratified replication in Section 6 — which directly addresses whether
segmentation effects exist at early durations and converge with time — produces
similar nulls within both the recent-arrival and long-tenured strata.

Individually, each of these results is a failure to reject the human-capital
convergence null. Jointly, across three outcomes covering all three theoretical
pathways and a duration-stratified robustness check, they constitute meaningful
evidence *for* convergence and against the strong segmented-assimilation prediction
in this sample.

## The Survivorship-Bias Interpretation

The most important limitation of the pooled analysis is the duration profile of the
sample. The mean foreign-born respondent in the 2022 SCF has lived in the United
States for `r key_stats$mean_yiu_fmt` years — a population dominated by immigrants who arrived in the 1980s
and early 1990s. The segmented-assimilation framework makes its strongest predictions
about *early-stage* incorporation, when origin-specific reception contexts are most
consequential. By the time immigrants have been in the U.S. for several decades, two
mechanisms may operate to narrow pathway-specific gaps:

1. **Convergence within origin groups.** Long-tenured immigrants who continue to
reside in the U.S. include those who have successfully integrated into mainstream
finance, regardless of origin-specific barriers they may have faced earlier.

2. **Selective return migration.** Immigrants who failed to incorporate may
disproportionately return-migrate over the 10-30 year window after arrival, leaving
the long-tenured U.S. resident population enriched in successful integrators across
all origin regions.

Either or both mechanisms would attenuate observed origin-region differences in
late-stage data while leaving them visible in early-stage data. The duration-stratified
analysis in Section 6 is the direct test. Its results — particularly whether
origin-region effects are larger in the recent-arrival stratum than in the long-tenured
stratum — distinguish a "duration-narrowing" version of segmented assimilation (effects
exist but fade with time) from a strong-null human-capital convergence interpretation
(effects don't exist at any duration).

## Implications for Policy and Theory

If the duration-stratified analysis yields null effects in both recent-arrival and
long-tenured strata, this empirical evidence indicates that the origin region does
not exert an independent influence on financial outcomes beyond what is already
captured by race-based frameworks. Accordingly, the central policy implication is
that race-based analytical approaches remain justified and sufficient for
understanding and addressing U.S. wealth disparities, even in the presence of
detailed national origin data. Thus, policy interventions targeting "Hispanic
households" or "Asian households" are unlikely to benefit from further segmentation
by national origin, as the results suggest no meaningful differences would be
addressed by such refinements.

If the duration-stratified analysis shows larger effects in the recent-arrival
stratum, the contribution is to argue that segmented assimilation operates in U.S.
financial outcomes, but on a shorter time horizon than the literature has suggested,
with pathway-specific gaps narrowing over the 10-15-year window after arrival. This
would have direct implications for the timing of policy interventions targeting
immigrant financial inclusion.

In either case, the substantive contribution is a clearer empirical picture of the
domain in which segmented-assimilation theory applies to U.S. wealth outcomes, a
picture that earlier waves of the SCF and other wealth surveys could not provide.

# Limitations

This study faces several key limitations. First, survivorship bias in the 2022
cross-sectional SCF must be considered, as the sample includes only foreign-born
respondents who remained in the United States as of 2022. Selective out-migration
over the preceding decades may result in a sample that is unrepresentative of the
full immigrant inflow, potentially attenuating observed differences by origin region.

Second, the single-wave design limits temporal inference: the analysis cannot
separate cohort or period effects, nor track changes over time, as the 2022 SCF is
the first to include immigration variables; subsequent waves will be required for
cohort-specific or longitudinal analysis.

Third, the absence of parental nativity precludes analysis of second-generation
outcomes, restricting the study's focus to first-generation immigrants and limiting
direct engagement with the core implications of segmented-assimilation theory.

Fourth, the necessity of aggregating respondents into broad origin-region categories,
due to insufficient sample sizes for country-specific analysis, risks obscuring
within-group heterogeneity and may mask meaningful differences among national-origin
subgroups, particularly within the broad 'Asian' or 'Latin American' classifications.
The small sample for certain groups, notably the Caribbean category (n=`r key_stats$n_caribbean_fmt` in the
pooled sample), further reduces the statistical power to detect group-specific
effects.

Fifth, although the analytic sample is sufficiently powered to identify large
segmentation effects (odds ratios greater than 2.0 or less than 0.5), it is only
moderately powered to detect smaller associations. Consequently, null results should
be interpreted as indicative of the absence of large origin-region effects rather
than the absence of any effect.

Lastly, public-use SCF data include disclosure-avoidance modifications, which may
attenuate cell-specific estimates for the smallest origin groups. Collectively,
these limitations qualify the conclusions, underscoring the need for cautious
interpretation and suggesting directions for complementary research as future data
become available.

# Conclusion

This study provides the first direct comparison, using nationally representative U.S.
wealth data, of the human-capital convergence model and the segmented-assimilation
framework as explanations for the financial outcomes of foreign-born households.
Utilizing newly available 2022 Survey of Consumer Finances variables on respondents'
national origin and years lived in the United States, the analysis tested whether
origin region carries independent explanatory power—net of duration and human
capital—across three key financial outcomes representing three theoretical pathways.

The pooled results reveal no statistically significant differences by origin region,
with minimal and inconsistent effects that do not fit the pathway-specific
predictions of segmented assimilation. Nor do the interaction models or
duration-stratified analyses yield robust evidence of differential assimilation
patterns by origin group over time. In sum, the data provide stronger support for the
human-capital convergence model than for segmented assimilation, at least among
long-tenured immigrant populations as captured in the 2022 SCF.

These findings have important forward-looking implications for both research and
policy. Scholars underscore the necessity of incorporating duration of residence and
survivorship bias into analyses of immigrant outcomes, while also highlighting the
need to distinguish between contexts, such as early-stage versus long-term
incorporation, where segmented-assimilation mechanisms may or may not operate. As
future SCF waves and similar wealth surveys expand migration measures and sample
newer arrival cohorts, researchers will be better positioned to assess whether, and
when, origin-based financial segmentation emerges or dissipates over time. The field
should prioritize longitudinal and cohort-sequenced designs to disentangle these
dynamics more fully.

For policymakers, the results suggest that, for established immigrant populations,
policy frameworks targeting wealth disparities may continue to rely on established
race-based classifications without significant loss of explanatory power.
Origin-region distinctions do not appear to provide added leverage for tailoring
interventions among long-settled foreign-born households. Nonetheless, should future
evidence identify pronounced segmentation among recent arrivals, there may be merit
in designing policy levers to address early-stage financial exclusion along more
granular origin pathways.

Overall, this analysis clarifies the empirical boundaries of segmented-assimilation
theory for U.S. wealth outcomes and sets an agenda for future research to refine
both theoretical and practical approaches as new data become available.

# Appendix A: Variable List

| Construct | Source variable | Type |
|---|---|---|
| Foreign-born status (component A) | `X6915` (ancestry; non-0 and non-1200) | Raw codebook |
| Foreign-born status (component B) | `X6906` (years respondent has lived in U.S.) | Raw codebook |
| National origin (R) | `X6915` | Raw codebook |
| Spouse Hispanic origin (R) | `X7404` (auxiliary, not used in main models) | Raw codebook |
| Race/ethnicity | `racecl4` | Bulletin extract |
| Age | `age` | Bulletin extract |
| Education | `educ` (recoded into 5 levels) | Bulletin extract |
| Marital status | `married` | Bulletin extract |
| Income | `income` | Bulletin extract |
| Net worth | `networth` | Bulletin extract |
| Net-worth percentile | `nw_pctl` (computed on full SCF distribution) | Derived |
| Retirement assets | `retqliq` | Bulletin extract |
| Stocks | `stocks` | Bulletin extract |
| Credit denied | `turndown` | Bulletin extract |
| Checking balance | `checking` | Bulletin extract |
| Homeowner | `housecl` | Bulletin extract |
| Weight | `wgt` | Bulletin extract |

# Appendix B: scf Package Workflow Notes

This document follows the workflow established in Paper 1 (`scf_ccbal_analysis_v5.Rmd`):

1. `scf_download(2022)` writes raw `.rds` files to `scf_data/`.
2. `scf_load(2022, data_directory = "scf_data")` returns an object with
   `$mi_design[[1..5]]` survey-design slots, one per implicate.
3. `scf_update(scf_obj, new_var = expr, ...)` adds variables to all 5 implicates.
4. `scf_mean()`, `scf_glm()`, `scf_ols()` fit pooled models with Rubin's-Rules pooling
   and replicate-weight-based variance.
5. Subsetting (e.g. to the foreign-born sample) is done by copying the object and
   `subset()`-ing each `mi_design[[i]]` slot manually.

The new 2022 immigration variables (`X6915`, `X6906`, `X7404`, `X7796`) are not in the
Bulletin summary extract that the `scf` package builds. The diagnostic chunk at the
top of this document attempts to merge them in from the raw 2022 SCF `.rds` file. If
the merge fails, those analyses cannot proceed.

# Appendix C: V12 Methodological Notes

This V12 build aligns the analytical structure with the three-hypothesis design,
removing redundant outcomes and specifications that did not directly answer the
research question.

**V12 changes:**

1. **Reduced from 6 outcomes to 3 — one per pathway.** Stocks ownership (parallel),
   no-checking-account (downward), and net-worth percentile (selective) were
   removed. Each was conceptually redundant with the retained outcome in its pathway
   (retirement, credit denial, homeownership, respectively), and reporting all six
   inflated the apparent number of model fits without adding interpretive value.
   The retained outcomes are also the strongest theoretical fit to each pathway:
   retirement-account ownership maps cleanly to mainstream-finance integration;
   credit denial is the canonical structural-exclusion measure; homeownership has
   the strongest standing in the segmented-assimilation literature (Logan 2007;
   Telles & Ortiz 2008) as a selective-incorporation outcome.

2. **Eliminated the Baseline-vs-Full nesting; report Full only.** The Baseline
   specification (without education, marital status, income) was a robustness
   check on whether human-capital controls absorb origin effects, not a separate
   hypothesis test. The research question is whether origin matters net of human
   capital, which is the Full specification. Reporting both in the main results
   inflated the apparent number of models without adding interpretive value.

3. **Cut H3 interaction analysis entirely.** The H3 origin × years-in-U.S.
   interaction tests were intended to test whether duration slopes differ by
   origin, but Section 6's duration-stratified analysis answers the same question
   more directly and produces interpretable within-stratum coefficients rather
   than non-convergent interaction terms with confidence intervals of [0, Inf].
   Section 6 alone is sufficient evidence for or against duration-narrowing
   segmentation; H3 added complexity without adding insight.

4. **Net effect on model count.** V11 reported 12 H1 models (6 outcomes × 2 specs)
   plus 6 H3 interaction models plus 6 duration-stratified models = 24 model fits.
   V12 reports 3 H1 models plus 6 duration-stratified models = 9 model fits, all
   directly answering the research question. The reduction makes the paper easier
   to read, harder to dismiss as fishing, and more aligned with the
   one-model-per-hypothesis structure expected of a thesis chapter.

**V10–V11 changes (carried forward):**

V11 added programmatic `key_stats` infrastructure so numeric claims in the
narrative prose are pulled from the live model results rather than hardcoded.

V10 adopted prose improvements from a manual rewrite of the V9 knit output —
cleaner declarative language in the Abstract, citation anchors throughout the
Central Question and Why This Matters sections, the "Reading Table 3" walkthrough
narrative, the First/Second/.../Lastly limitations structure, and the
scholars/policymakers/forward-looking conclusion structure.

**V9 changes (carried forward):**

V9 reframed the Abstract, Section 1, Discussion, and Conclusion to treat
human-capital convergence and segmented assimilation as competing hypotheses
given equal a priori weight, then report what the data show. V9 also added the
duration-stratified analysis (Section 6) that V12 retains as the main
within-pathway robustness check.

**Earlier-version corrections (carried forward):**

V8 fixed a singularity issue: with the strict foreign-born filter (requires
years_in_us > 0), the missing-indicator variable yiu_missing became constant 0 and
crashed the GLM solver. V8 replaced `yiu_clean + yiu_missing` with raw `years_in_us`
in all model formulas.

V7 fixed a formula-environment scoping issue: when `as.formula()` is called inside a
function, the formula's environment becomes the function's local scope, which doesn't
contain the survey design's variables. V7 sets `environment(formula) <- globalenv()`
so the survey functions can resolve term names against the design's `$variables` data.

V6 corrected the years-in-U.S. variable identity: V5 erroneously used X7437–X7443
(which are mortgage-forbearance variables per the official 2022 SCF changes file). The
correct YIU variable is X6906 for the respondent and X7796 for the spouse.

V5 corrections: foreign-born identification requires both X6915 indicating non-
American ancestry AND X6906 > 0; net-worth-percentile (now removed in V12) was
computed via `scf_update` on the full SCF sample before subsetting; quasi-complete
separation pre-checked in the H1 cache loop; modeling errors surfaced explicitly
rather than silently swallowed; H1 result tables reformatted.
