---
title: "The Financial Balance Sheet of Childlessness"
subtitle: "An Analysis of Household Wealth, Debt, and Credit Constraints in the 2022 Survey of Consumer Finances"
author: "Delroy Nichols"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 5, dpi = 150
)
rm(list = ls())
gc()
set.seed(42)
```

# Abstract

This paper examines the financial correlates of childlessness among prime-age U.S.
households. While demographic research has extensively studied the social and cultural
drivers of fertility decline, the detailed financial circumstances of childless
households — particularly their assets, debts, and access to credit — remain less well
understood. Using replicate-weighted, multiply-imputed data from the 2022 Survey of
Consumer Finances (SCF), we analyze households headed by individuals aged 31–44 and
define childlessness as having no co-resident children, a measure that captures a
household's current living arrangement rather than biological fertility history.

Our logistic regression models reveal that, beyond powerful demographic predictors like
marital status and education, specific financial variables are strongly associated with
childlessness. Holding higher levels of retirement assets is positively associated with
childlessness, consistent with an opportunity cost framework in which long-term savings
and family formation compete for household resources. Conversely, households that have
experienced credit denial are more likely to be childless, suggesting that financial
exclusion and the instability it signals may constitute a barrier to family formation.
Income and net worth, by contrast, are not significant predictors after other financial
characteristics are controlled.

This study repositions the economic dimension of childlessness beyond simple income
measures, demonstrating that the composition of the household balance sheet — what a
household owns, what it owes, and whether it can access credit — matters more than
aggregate financial resources for understanding the economic context of family formation.

---

```{r load-packages}
required_packages <- c(
  "tidyverse", "haven", "survey", "labelled",
  "psych", "ggplot2", "scales", "scf",
  "glmnet", "Matrix", "broom", "knitr",
  "moments", "car"
)

for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg, repos = "http://cran.us.r-project.org")
    library(pkg, character.only = TRUE)
  }
}
```

```{r load-data}
dir.create("scf_data", showWarnings = FALSE, recursive = TRUE)
original_wd <- getwd()
setwd("scf_data")
if (!file.exists("scf2022.rds")) {
  tryCatch(scf_download(2022), error = function(e)
    cat("Download error:", conditionMessage(e), "\n"))
}
setwd(original_wd)
scf_2022 <- scf_load(2022, data_directory = "scf_data")
```

# Introduction

The rise of childlessness is a defining demographic trend of the 21st century. The U.S.
total fertility rate has fallen below replacement and continued to decline through 2023
(Osterman et al., 2023; Camarota & Zeigler, 2025). A vast body of work has explored the
demographic and social drivers of this trend, from educational attainment and marital
patterns to shifting cultural norms and, more recently, immigration-related compositional
effects (Parrado & Morgan, 2008; Adserà & Ferrer, 2015; Zill, 2025). Yet the economic
dimension is often simplified to measures of income or employment status, collapsing the
richness of household finances into a single axis.

This paper argues for a more comprehensive financial perspective. A household's
decision-making, opportunities, and constraints are shaped not just by income flows but
by its entire balance sheet: what it owns (assets), what it owes (debts), and whether it
can access credit. A household with $80,000 in retirement savings, $15,000 in student
loans, and a recent credit denial inhabits a fundamentally different economic reality than
one with $80,000 in income — even though both might appear similar on a single financial
measure.

Using the unparalleled financial detail of the Survey of Consumer Finances (SCF), this
paper asks: **What are the financial correlates of co-resident childlessness among
prime-age U.S. households?** We move beyond income to investigate how different forms of
wealth (retirement vs. liquid assets), debt (installment loans vs. credit card balances),
and credit access are associated with the likelihood of a household having no children
present. By centering the analysis on the SCF's core strength — granular financial data —
we provide a descriptive but deeply nuanced portrait of the economic trade-offs and
circumstances linked to household structure during the prime family formation years.

This paper serves as a companion to a primary analysis of credit card indebtedness using
the same dataset, in which the transition to parenthood was identified as the strongest
demographic predictor of carrying revolving credit card debt. The present analysis
inverts that question: given what we know about the financial burdens of parenthood, what
does the financial profile of childless households look like?

---

# Data and Methods

## Data Source

We use the 2022 Survey of Consumer Finances (SCF), a triennial cross-sectional survey of
U.S. households conducted by the Federal Reserve Board in cooperation with NORC at the
University of Chicago. The SCF employs a dual-frame design combining area-probability
sampling with a list-based oversample of wealthy households drawn from IRS records. The
resulting dataset is the most detailed publicly available source of information on
household balance sheets, income, pensions, and credit experiences in the United States
(Bricker et al., 2023).

The SCF addresses item nonresponse through multiple imputation, providing five complete
implicates of the dataset. All analyses account for this structure using Rubin's Rules,
as implemented by the `scf` R package (Cohen, 2025). Survey replicate weights (balanced
repeated replication) are used throughout to produce design-consistent variance estimates.

## Sample

We restrict the analysis to households where the reference person is aged 31–44. This
window captures the prime family formation years — past the median age at first birth in
the United States (approximately 27; Osterman et al., 2023) but before the conventional
boundary for completed fertility (age 44) used in demographic research (Guzzo & Loo,
2023). The restriction yields approximately 1,000 households per implicate.

## Outcome Variable

The primary outcome, `childless_prime`, is an indicator equal to 1 if the reference
person is aged 31–44 and reports zero children currently residing in the household.

**It is critical to underscore the precise nature of this variable.** It is a measure of
co-resident child status, not biological fertility. This measure will systematically
misclassify non-custodial parents — primarily fathers whose children reside with another
parent — as childless. It may also misclassify parents whose children live with
grandparents or other relatives. This misclassification is non-trivial and is likely
correlated with key predictors including race, marital status, and income.

However, this measure is the only indicator of child status available in the SCF. The
primary justification for using the SCF despite this limitation is its status as the gold
standard for household financial data. No other survey provides comparable granularity on
assets, liabilities, income sources, and credit experiences. The trade-off is explicit:
we sacrifice precision in the fertility measure to gain depth in the financial measures,
and we interpret all findings accordingly as pertaining to co-resident childlessness, not
biological childlessness.

```{r create-variables}
scf_2022 <- scf_update(scf_2022,

  # Outcome
  childless_prime = as.integer(age >= 31 & age <= 44 & kids == 0),

  # Demographics
  female            = as.integer(hhsex == 2),
  currently_married = as.integer(married == 1),
  has_children      = as.integer(kids > 0),
  college_degree    = as.integer(edcl == 4),

  educ_cat = factor(edcl,
    levels = 1:4,
    labels = c("Less than HS", "High School", "Some College", "College Degree")
  ),

  race_cat = factor(racecl4,
    levels = 1:4,
    labels = c("White", "Black", "Hispanic", "Other")
  ),

  age_sq = age^2,

  # Financial variables (asinh-transformed)
  asinh_income       = asinh(income),
  asinh_networth     = asinh(networth),
  asinh_wageinc      = asinh(wageinc),
  asinh_debt_excl_cc = asinh(pmax(debt - ccbal, 0)),
  asinh_install      = asinh(install),
  asinh_edn_inst     = asinh(edn_inst),
  asinh_liq          = asinh(liq),
  asinh_retqliq      = asinh(retqliq),
  asinh_stocks       = asinh(stocks),
  asinh_nmmf         = asinh(nmmf),
  asinh_homeeq       = asinh(homeeq),
  asinh_vehic        = asinh(vehic),
  asinh_dti          = asinh(ifelse(income > 0, debt / income, 0)),
  asinh_ccbal        = asinh(ccbal),

  has_ccdebt = as.integer(ccbal > 0),

  # Labor / credit access
  is_working   = as.integer(lf == 1),
  is_homeowner = as.integer(housecl == 1),
  has_pension  = as.integer(anypen == 1),
  turned_down  = as.integer(turndown == 1)
)
```

## Predictor Variables

Our key financial predictors, summarized below, are grouped by balance sheet category.
All continuous financial variables are transformed using the inverse hyperbolic sine
(asinh), which accommodates zeros and negative values while approximating a log
transformation for large values.

**Assets:** retirement and quasi-liquid assets (`asinh_retqliq`), liquid transaction
accounts (`asinh_liq`), directly held stocks (`asinh_stocks`), pooled investment funds
(`asinh_nmmf`), home equity (`asinh_homeeq`), vehicle value (`asinh_vehic`).

**Debts:** installment debt (`asinh_install`), education debt (`asinh_edn_inst`),
non-credit-card debt (`asinh_debt_excl_cc`), credit card balance (`asinh_ccbal`),
debt-to-income ratio (`asinh_dti`).

**Credit access:** a binary indicator for having been turned down for credit in the past
five years (`turned_down`).

**Income:** total household income (`asinh_income`), wage income (`asinh_wageinc`), net
worth (`asinh_networth`).

**Demographic controls:** age and age-squared (to capture nonlinear lifecycle effects),
marital status, college degree, race/ethnicity, gender of reference person, employment
status, homeownership, and pension access.

**A note on the `female` variable.** In the SCF, the reference person in a mixed-sex
couple is coded as the male. A female reference person therefore typically indicates a
single woman (often a single mother) or a woman in a same-sex couple. This variable
functions as a control for household structure, not a direct measure of gender differences
in fertility.

**A note on the "Other" race category.** The SCF public dataset aggregates Asian, Pacific
Islander, American Indian/Alaska Native, and multiracial respondents into a single
"Other" category. Published research indicates this group is predominantly Asian/Pacific
Islander (Hanna & Lindamood, 2008), but the internal heterogeneity is substantial. We
retain this category in the model for completeness but urge caution in interpreting its
coefficient. The opposing fertility patterns of U.S.-born Asian women (high
childlessness) and first-generation Asian immigrants (lower childlessness) may produce
attenuated or null aggregate effects that mask meaningful within-group variation (Cai &
Morgan, 2019).

```{r restrict-sample}
scf_prime <- scf_2022
for (i in 1:5) {
  idx <- scf_2022$mi_design[[i]]$variables$age >= 31 &
         scf_2022$mi_design[[i]]$variables$age <= 44
  scf_prime$mi_design[[i]] <- subset(scf_2022$mi_design[[i]], idx)
}
```

---

# Descriptive Results

```{r descriptives-sample, results='asis'}
imp1 <- scf_prime$mi_design[[1]]$variables
n_prime <- nrow(imp1)
n_childless <- sum(imp1$kids == 0)
pct_childless <- round(100 * mean(imp1$kids == 0), 1)

cat(sprintf("The analysis sample comprises approximately %s households per implicate.
Among these, %.1f%% report no co-resident children. This rate is broadly consistent with
Census Bureau estimates of childlessness for women in this age range (Guzzo & Loo, 2023),
though it likely overstates biological childlessness due to the non-custodial parent
misclassification discussed above.\n\n",
  formatC(n_prime, big.mark = ","), pct_childless))
```

## Co-Resident Childlessness by Demographics

```{r desc-demographics}
cat("===== By Race/Ethnicity =====\n")
print(scf_mean(scf_prime, ~childless_prime, by = ~race_cat))

cat("\n===== By Education =====\n")
print(scf_mean(scf_prime, ~childless_prime, by = ~educ_cat))

cat("\n===== By Marital Status =====\n")
print(scf_mean(scf_prime, ~childless_prime, by = ~currently_married))

cat("\n===== By Credit Denial Experience =====\n")
print(scf_mean(scf_prime, ~childless_prime, by = ~turned_down))
```

## Visualizations

```{r viz-race, fig.width=10, fig.height=5}
# Use weighted estimates from all implicates via scf_mean
race_rates <- scf_mean(scf_prime, ~childless_prime, by = ~race_cat)
race_df <- as.data.frame(race_rates$results)

ggplot(race_df, aes(x = reorder(group, -estimate), y = estimate * 100, fill = group)) +
  geom_col(width = 0.6) +
  geom_errorbar(aes(ymin = (estimate - 1.96 * se) * 100,
                     ymax = (estimate + 1.96 * se) * 100),
                width = 0.2) +
  geom_text(aes(label = sprintf("%.1f%%", estimate * 100)), vjust = -1.2, size = 4) +
  scale_fill_manual(values = c("White" = "#4A90D9", "Black" = "#E8724A",
                                "Hispanic" = "#50C878", "Other" = "#9B59B6")) +
  labs(
    title = "Co-Resident Childlessness by Race/Ethnicity (Ages 31-44)",
    subtitle = "Survey-weighted, multiply-imputed estimates with 95% CI",
    x = NULL, y = "% Childless"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none") +
  ylim(0, 60)
```

```{r viz-education, fig.width=10, fig.height=5}
educ_rates <- scf_mean(scf_prime, ~childless_prime, by = ~educ_cat)
educ_df <- as.data.frame(educ_rates$results)

ggplot(educ_df, aes(x = group, y = estimate * 100, fill = group)) +
  geom_col(width = 0.6) +
  geom_errorbar(aes(ymin = (estimate - 1.96 * se) * 100,
                     ymax = (estimate + 1.96 * se) * 100),
                width = 0.2) +
  geom_text(aes(label = sprintf("%.1f%%", estimate * 100)), vjust = -1.2, size = 4) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Co-Resident Childlessness by Education (Ages 31-44)",
    subtitle = "Survey-weighted, multiply-imputed estimates with 95% CI",
    x = NULL, y = "% Childless"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none") +
  ylim(0, 60)
```

---

# Regression Results

```{r model}
model_predictors <- c(
  # Balance sheet: Assets
  "asinh_retqliq", "asinh_liq", "asinh_stocks", "asinh_nmmf",
  "asinh_homeeq", "asinh_vehic",
  # Balance sheet: Debts
  "asinh_install", "asinh_edn_inst", "asinh_debt_excl_cc",
  "asinh_ccbal", "asinh_dti",
  # Income and wealth
  "asinh_income", "asinh_wageinc", "asinh_networth",
  # Credit access
  "turned_down",
  # Demographics
  "age", "age_sq",
  "female", "currently_married", "college_degree",
  "race_cat",
  "is_working", "is_homeowner", "has_pension"
)

model_formula <- as.formula(
  paste("childless_prime ~", paste(model_predictors, collapse = " + "))
)

model_fit <- scf_glm(scf_prime, model_formula, family = binomial())

# Compute odds ratios
results <- model_fit$results
results$odds_ratio <- exp(results$estimate)
results$or_lower   <- exp(results$estimate - 1.96 * results$std.error)
results$or_upper   <- exp(results$estimate + 1.96 * results$std.error)
```

```{r model-table}
kable(results %>%
  select(term, estimate, std.error, odds_ratio, or_lower, or_upper, p.value, stars) %>%
  mutate(across(c(estimate, std.error, odds_ratio, or_lower, or_upper), ~round(., 4)),
         p.value = formatC(p.value, format = "e", digits = 2)),
  col.names = c("Term", "Log-Odds", "SE", "Odds Ratio", "OR 2.5%", "OR 97.5%",
                "p-value", ""),
  caption = "Table 1. Logistic Regression: P(Co-Resident Childless, Ages 31-44)")
```

```{r model-fit, results='asis'}
null_dev <- svyglm(childless_prime ~ 1, design = scf_prime$mi_design[[1]],
                    family = binomial())$deviance
full_dev <- svyglm(model_formula, design = scf_prime$mi_design[[1]],
                    family = binomial())$deviance
pseudo_r2 <- round(1 - full_dev / null_dev, 4)
n_model <- nrow(scf_prime$mi_design[[1]]$variables)

cat(sprintf("**Model fit (Implicate 1):** McFadden pseudo-R² = %.4f; N = %s;
Predictors = %d\n",
  pseudo_r2, formatC(n_model, big.mark = ","), length(model_predictors)))
```

## Coefficient Visualization

```{r coef-plot, fig.width=11, fig.height=9}
plot_df <- results %>%
  filter(term != "(Intercept)") %>%
  mutate(
    significant = p.value < 0.05,
    direction = ifelse(estimate > 0, "Positive", "Negative"),
    # Group variables for visual clarity
    category = case_when(
      term %in% c("asinh_retqliq", "asinh_liq", "asinh_stocks", "asinh_nmmf",
                   "asinh_homeeq", "asinh_vehic") ~ "Assets",
      term %in% c("asinh_install", "asinh_edn_inst", "asinh_debt_excl_cc",
                   "asinh_ccbal", "asinh_dti") ~ "Debts",
      term %in% c("asinh_income", "asinh_wageinc", "asinh_networth") ~ "Income/Wealth",
      term == "turned_down" ~ "Credit Access",
      TRUE ~ "Demographics"
    )
  )

ggplot(plot_df, aes(x = reorder(term, abs(estimate)), y = estimate,
                     fill = category, alpha = significant)) +
  geom_col(width = 0.7) +
  geom_errorbar(aes(ymin = estimate - 1.96 * std.error,
                     ymax = estimate + 1.96 * std.error),
                width = 0.3, linewidth = 0.4) +
  scale_fill_manual(values = c("Assets" = "#2E86C1", "Debts" = "#E74C3C",
                                "Income/Wealth" = "#27AE60", "Credit Access" = "#F39C12",
                                "Demographics" = "#8E44AD")) +
  scale_alpha_manual(values = c("TRUE" = 1.0, "FALSE" = 0.3)) +
  coord_flip() +
  labs(
    title = "Predictors of Co-Resident Childlessness (Ages 31-44)",
    subtitle = "Log-odds coefficients with 95% CI, colored by balance sheet category",
    x = NULL, y = "Coefficient (log-odds)",
    fill = "Category", alpha = "p < 0.05"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

---

# Interpretation

## The Balance Sheet Findings

```{r interp-retirement, results='asis'}
ret <- results %>% filter(term == "asinh_retqliq")

cat("### Retirement Assets and the Opportunity Cost of Children\n\n")
cat(sprintf("The value of retirement and quasi-liquid assets (`asinh_retqliq`) is the
strongest financial predictor of childlessness: OR = %.2f (p = %s). Households with
larger retirement savings are significantly more likely to be childless during the prime
family formation years.\n\n",
  ret$odds_ratio, formatC(ret$p.value, format = "e", digits = 2)))

cat("This finding lends clear support to an opportunity cost framework. Retirement
account accumulation reflects sustained career investment, long time horizons, and a
financial planning orientation that may coincide with — or come at the expense of —
family formation. The cost of raising a child through age 18 is estimated at $237,000
to $310,000 in 2022 dollars (Lino et al., 2020), and the career interruptions associated
with parenthood (particularly for women) reduce lifetime earnings and retirement savings.
Households that have prioritized or been able to accumulate more retirement wealth are,
on this evidence, less likely to have children present — suggesting a genuine trade-off
between investing in financial assets for the future and bearing the immense cost of
raising children in the present.\n\n")

cat("Importantly, this is not a simple income effect. As we show below, aggregate income
and net worth are not significant predictors of childlessness. The specificity of the
retirement asset finding — that it is the *composition* of wealth, not its level, that
matters — is the core contribution of this analysis.\n")
```

```{r interp-credit, results='asis'}
td <- results %>% filter(term == "turned_down")

cat("\n### Credit Constraints and Financial Exclusion\n\n")
cat(sprintf("Having been turned down for credit in the past five years (`turned_down`)
is significantly associated with childlessness: OR = %.2f (p = %s). Households that
have experienced credit denial are approximately %.0f%% more likely to be childless.\n\n",
  td$odds_ratio, formatC(td$p.value, format = "e", digits = 2),
  100 * (td$odds_ratio - 1)))

cat("The interpretation of this coefficient requires care regarding causal direction.
Two plausible mechanisms are consistent with this association:\n\n")

cat("*Financial exclusion as a barrier to family formation.* Credit-constrained
households may face material barriers to the consumption demands of parenthood — housing,
vehicles, childcare — that require access to credit. If these households cannot borrow
to smooth the lumpy expenses associated with having a child, they may delay or forgo
parenthood. On this reading, credit denial signals a broader financial precarity that
deters family formation.\n\n")

cat("*Parenthood as a strain on creditworthiness.* Alternatively, the presence of
children strains a household's liquidity and makes it more vulnerable to financial shocks,
which may damage creditworthiness and lead to credit denial. On this reading, parents are
more likely to have been turned down *because* child-related expenses exhausted their
financial buffer.\n\n")

cat("The cross-sectional design of the SCF cannot distinguish between these mechanisms.
However, this finding bridges directly to the companion analysis of credit card debt,
where credit denial was the strongest predictor of carrying revolving balances (OR ≈ 1.97).
Credit access thus emerges as a common upstream factor linking both childlessness and
credit card indebtedness — marking a population experiencing financial precarity across
multiple life domains.\n")
```

```{r interp-income, results='asis'}
inc <- results %>% filter(term == "asinh_income")
nw  <- results %>% filter(term == "asinh_networth")
wg  <- results %>% filter(term == "asinh_wageinc")

cat("\n### Income, Wealth, and the Null Finding\n\n")
cat(sprintf("Total household income (OR = %.2f, p = %s) and net worth (OR = %.2f,
p = %s) are not significant predictors of childlessness after controlling for the
composition of the balance sheet. ",
  inc$odds_ratio, formatC(inc$p.value, format = "e", digits = 2),
  nw$odds_ratio, formatC(nw$p.value, format = "e", digits = 2)))

cat("This null finding is itself a substantive result. It indicates that the economic
context of childlessness is not about how much a household has in aggregate, but about
how that wealth is allocated and whether the household faces credit constraints. A
high-income household that has channeled resources into retirement accounts and a
low-income household that has been denied credit occupy opposite ends of the financial
spectrum, yet both are more likely to be childless than their demographic peers.\n\n")

if (wg$p.value < 0.05 && wg$estimate < 0) {
  cat(sprintf("Wage income (`asinh_wageinc`) is the one income-adjacent variable that
reaches significance (OR = %.2f, p = %s), with higher wages associated with *lower*
childlessness. This may reflect financial stability enabling family formation, or it may
capture the mechanical relationship between non-employment and parenthood (discussed
below under structural controls). The distinction between wage income (significant) and
total income (non-significant) suggests that it is earned income from active labor force
participation, not investment returns or transfers, that correlates with having children
present.\n",
    wg$odds_ratio, formatC(wg$p.value, format = "e", digits = 2)))
}
```

```{r interp-ccbal, results='asis'}
cc <- results %>% filter(term == "asinh_ccbal")

cat("\n### Credit Card Balances: The Link to the Companion Analysis\n\n")
cat(sprintf("Credit card balance (`asinh_ccbal`): OR = %.2f (p = %s). ",
  cc$odds_ratio, formatC(cc$p.value, format = "e", digits = 2)))

if (cc$p.value >= 0.05) {
  cat("Credit card balances do not independently predict childlessness after other
financial variables are controlled. The companion analysis showed that the transition
to parenthood is the strongest demographic predictor of carrying credit card debt
(OR ≈ 1.54). The present analysis confirms the reverse direction: credit card debt does
not appear to *cause* childlessness — rather, parenthood drives credit card use. Once
we control for the broader financial structure of the household, the credit card
balance carries no additional predictive power for childlessness.\n")
} else if (cc$estimate < 0) {
  cat("Higher credit card balances are associated with lower odds of childlessness,
consistent with the companion analysis finding that parenthood drives credit card use.
The relationship appears bidirectional: parents carry more revolving debt, and households
carrying more revolving debt are less likely to be childless.\n")
}
```

## Demographic Controls

```{r interp-demographics, results='asis'}
mar <- results %>% filter(term == "currently_married")
ed  <- results %>% filter(term == "college_degree")
fem <- results %>% filter(term == "female")
wrk <- results %>% filter(term == "is_working")

cat("### Expected Demographic Patterns\n\n")

cat(sprintf("**Marriage** is overwhelmingly the strongest predictor: OR = %.2f
(p = %s). Unmarried adults aged 31–44 have dramatically higher odds of co-resident
childlessness. Marriage remains the primary institutional context for childbearing in
the United States, and this coefficient captures the combined effect of partnership
stability, shared financial resources, and normative expectations (Pew Research Center,
2015; Zill, 2025).\n\n",
  mar$odds_ratio, formatC(mar$p.value, format = "e", digits = 2)))

cat(sprintf("**College degree**: OR = %.2f (p = %s). ",
  ed$odds_ratio, formatC(ed$p.value, format = "e", digits = 2)))
if (ed$p.value < 0.05 && ed$estimate > 0) {
  cat("College graduates are significantly more likely to be childless, consistent with
extensive demographic research showing that higher education delays and reduces fertility
(Livingston, 2015; Guzzo & Loo, 2023). Among women aged 40–44, approximately 19–22%% of
degree holders are childless compared to 7%% of those without a high school diploma
(Livingston, 2015).\n\n")
} else {
  cat("\n\n")
}

cat("### Structural Controls\n\n")

cat(sprintf("**Female reference person** (OR = %.2f, p = %s) and **employment**
(OR = %.2f, p = %s) are both significant but should be interpreted as structural
controls rather than substantive predictors of fertility preferences. ",
  fem$odds_ratio, formatC(fem$p.value, format = "e", digits = 2),
  wrk$odds_ratio, formatC(wrk$p.value, format = "e", digits = 2)))

cat("The female coefficient is driven by the SCF's reference person coding: female-headed
households in this age range are disproportionately single mothers, producing a mechanical
negative association with childlessness. The employment coefficient captures the
compositional fact that the non-working population aged 31–44 includes stay-at-home
parents, who are by definition not childless. These variables are included to control for
household structure; they should not be read as evidence of gender or employment effects
on fertility decisions.\n")
```

## Racial and Ethnic Categories

```{r interp-race, results='asis'}
race_terms <- results %>% filter(grepl("race_cat", term))

cat("With White households as the reference category, the coefficients for racial and
ethnic groups are as follows:\n\n")

for (j in 1:nrow(race_terms)) {
  r <- race_terms[j, ]
  cat(sprintf("- **%s**: OR = %.2f, 95%% CI [%.2f, %.2f] (p = %s)\n",
    gsub("race_cat", "", r$term), r$odds_ratio,
    r$or_lower, r$or_upper,
    formatC(r$p.value, format = "e", digits = 2)))
}

cat("\nIn the fully controlled model, none of the racial coefficients reach conventional
statistical significance. Given the moderate sample size relative to the number of
predictors (N ≈ %s with %d predictors, yielding approximately %d observations per
predictor), this model is likely underpowered to precisely estimate racial differences.
The confidence intervals are wide enough to encompass both substantively meaningful
effects and true nulls.\n\n",
  formatC(n_model, big.mark = ","), length(model_predictors),
  round(n_model / length(model_predictors)))

cat("**We do not interpret these non-significant coefficients as evidence for or against
any specific theory** — including theories of immigrant assimilation or cultural
differences in fertility norms. The financial and demographic controls in the model
account for a substantial portion of the unconditional racial variance in co-resident
childlessness, but whether the remaining variance reflects nativity composition,
unmeasured cultural factors, differential non-custodial parenthood rates, or simply
sampling variability cannot be determined from these data. Unpacking the source of
racial and ethnic differences in childlessness requires a dataset with direct measures
of nativity and biological fertility history, which the SCF lacks.\n\n")

# Caveat the Other category specifically
other <- race_terms %>% filter(term == "race_catOther")
if (nrow(other) > 0) {
  cat(sprintf("The 'Other' category merits particular caution (OR = %.2f, p = %s).
As noted in the Methods section, this group aggregates Asian, Pacific Islander, American
Indian/Alaska Native, and multiracial respondents. The internal heterogeneity of this
category — particularly the opposing nativity-specific fertility patterns within the
Asian subpopulation (Cai & Morgan, 2019) — means that its aggregate coefficient is
difficult to interpret substantively. We report it for completeness but do not draw
inferences from it.\n",
    other$odds_ratio, formatC(other$p.value, format = "e", digits = 2)))
}
```

---

# Connection to the Companion Analysis

The primary analysis in this thesis examined credit card indebtedness, finding that the
transition to parenthood is the strongest demographic predictor of carrying revolving
credit card debt. This companion piece inverts the question and finds complementary
results.

```{r linkage}
scf_prime <- scf_update(scf_prime,
  parent_status = factor(ifelse(kids == 0, "Childless", "Parent"))
)

cat("===== Credit Card Debt Prevalence: Childless vs. Parent (Ages 31-44) =====\n\n")
print(scf_mean(scf_prime, ~has_ccdebt, by = ~parent_status))

cat("\n===== Mean Credit Card Balance: Childless vs. Parent (Ages 31-44) =====\n\n")
print(scf_mean(scf_prime, ~ccbal, by = ~parent_status))
```

```{r linkage-interp, results='asis'}
cat("\nThe cross-tabulation confirms the expected pattern: parents aged 31–44 carry
higher credit card balances and are more likely to have any revolving debt than childless
households. Combined with the regression findings from both papers, the synthesis is:\n\n")

cat("**Parenthood drives credit card debt; the balance sheet shapes childlessness.**
The main paper shows that having children increases the odds of carrying credit card
debt by over 50%%, while additional children beyond the first do not compound the effect.
This companion paper shows that childlessness is predicted not by income or total wealth,
but by the composition of the balance sheet — retirement assets on one side, credit
constraints on the other. The economic story of family formation is richer than a single
income number can capture.\n\n")

cat("**Credit denial links both outcomes.** Households denied credit are simultaneously
more likely to carry revolving credit card debt (main paper) and more likely to be
childless (this paper). Financial exclusion marks a population experiencing precarity
across multiple domains.\n\n")

cat("**Income protects against debt, not against childlessness.** Higher wages reduce
credit card balances among debtors (main paper) but do not reduce childlessness (this
paper). The implication for policy is that income support may alleviate the debt burden
of existing parents but is unlikely, on its own, to shift fertility patterns.\n")
```

---

# Robustness: Interaction Effects

```{r interactions, results='asis'}
cat("As a robustness check, we test whether the association between education and
childlessness, and between marriage and childlessness, varies by race/ethnicity. If
the financial and demographic predictors operate differently across racial groups —
as immigration-related assimilation theories would predict — we would expect significant
interaction terms.\n\n")
```

```{r interaction-models}
# Race x Education interaction
int_educ_formula <- update(model_formula, . ~ . + race_cat:college_degree)
int_educ_fit <- tryCatch(
  scf_glm(scf_prime, int_educ_formula, family = binomial()),
  error = function(e) NULL
)

# Race x Marriage interaction
int_mar_formula <- update(model_formula, . ~ . + race_cat:currently_married)
int_mar_fit <- tryCatch(
  scf_glm(scf_prime, int_mar_formula, family = binomial()),
  error = function(e) NULL
)
```

```{r interaction-results, results='asis'}
if (!is.null(int_educ_fit)) {
  int_educ_terms <- int_educ_fit$results %>%
    filter(grepl("race_cat.*college_degree|college_degree.*race_cat", term))

  cat("### Race × Education Interactions\n\n")
  if (nrow(int_educ_terms) > 0) {
    for (j in 1:nrow(int_educ_terms)) {
      r <- int_educ_terms[j, ]
      sig <- ifelse(r$p.value < 0.05, "significant", "not significant")
      cat(sprintf("- **%s**: log-odds = %.3f (p = %s, %s)\n",
        r$term, r$estimate, formatC(r$p.value, format = "e", digits = 2), sig))
    }
    any_sig_educ <- any(int_educ_terms$p.value < 0.05)
    cat(sprintf("\n%s of the race × education interactions reach significance at the
0.05 level.\n\n",
      ifelse(any_sig_educ, "One or more", "None")))
  }
} else {
  cat("### Race × Education Interactions\n\n")
  cat("The interaction model did not converge, likely due to small cell sizes in the
race × education cross-tabulation within the 31–44 age window.\n\n")
}

if (!is.null(int_mar_fit)) {
  int_mar_terms <- int_mar_fit$results %>%
    filter(grepl("race_cat.*currently_married|currently_married.*race_cat", term))

  cat("### Race × Marriage Interactions\n\n")
  if (nrow(int_mar_terms) > 0) {
    for (j in 1:nrow(int_mar_terms)) {
      r <- int_mar_terms[j, ]
      sig <- ifelse(r$p.value < 0.05, "significant", "not significant")
      cat(sprintf("- **%s**: log-odds = %.3f (p = %s, %s)\n",
        r$term, r$estimate, formatC(r$p.value, format = "e", digits = 2), sig))
    }
    any_sig_mar <- any(int_mar_terms$p.value < 0.05)
    cat(sprintf("\n%s of the race × marriage interactions reach significance.\n\n",
      ifelse(any_sig_mar, "One or more", "None")))
  }
} else {
  cat("### Race × Marriage Interactions\n\n")
  cat("The interaction model did not converge.\n\n")
}

cat("These interaction tests should be interpreted with caution given the sample size.
The 31–44 age restriction yields approximately 1,000 observations, and adding interaction
terms further reduces the effective sample size within each cell. Non-significant
interactions may reflect insufficient power rather than true homogeneity of effects
across racial groups.\n")
```

---

# Discussion and Conclusion

This analysis demonstrates that a household's financial balance sheet is intricately
linked to its co-resident childlessness status during the prime family formation years.
The key contribution is that the economic context of childlessness is richer than simple
measures of income or wealth can capture.

The strong positive association between retirement assets and childlessness, combined with
the positive association between credit denial and childlessness, paints a picture of
competing financial pressures operating at opposite ends of the economic spectrum. At the
upper end, households investing in long-term financial security face opportunity costs
that may delay or preclude family formation. At the lower end, households denied access
to credit face material barriers and financial instability that may produce the same
outcome through a different mechanism. In between, aggregate income and wealth do not
distinguish childless households from parents — it is the *structure* of the balance
sheet, not its total, that matters.

The non-significant racial coefficients in the fully controlled model are notable but
should be interpreted conservatively. The descriptive statistics show clear differences
in co-resident childlessness rates by race and ethnicity, with Hispanic households
showing the lowest rates and "Other" (primarily Asian) households showing the highest.
These differences could reflect nativity composition, differential rates of non-custodial
parenthood, cultural variation in family formation norms, or other factors that the SCF
cannot isolate. Other scholars have argued persuasively that immigration-driven
compositional effects play a substantial role in racial fertility differentials (Parrado
& Morgan, 2008; Adserà & Ferrer, 2015; Camarota & Zeigler, 2025). Testing this
hypothesis directly requires a dataset that combines detailed fertility history with
nativity data — a task for which the National Survey of Family Growth (NSFG), American
Community Survey (ACS), or CPS Fertility Supplement would be better suited.

Nonetheless, this paper makes a valuable contribution by centering the role of wealth
composition, debt structure, and credit access in the study of household formation. The
findings suggest that policies aimed at easing the financial burden of child-rearing
should consider not only income support but also mechanisms to improve household
liquidity, access to credit, and the ability to save for both retirement and
child-related expenses simultaneously. The framing of children as a "financial trade-off"
against retirement savings is not merely metaphorical — the data confirm it is a
measurable feature of the American household balance sheet.

---

# Limitations

**Co-resident child status, not biological fertility.** The SCF's `kids` variable
measures children currently present in the household. Non-custodial parents (primarily
fathers), parents whose children live with other relatives, and other scenarios of
parent-child separation are misclassified as childless. This error is likely correlated
with race, income, and marital status, and could bias coefficients in unpredictable
ways. For example, if non-custodial fatherhood is more prevalent among Black or
lower-income households, their childlessness rate will be artificially inflated.

**Sample size and statistical power.** With approximately 1,000 observations and
`r length(model_predictors)` predictors (~`r round(n_model / length(model_predictors))`
observations per predictor), the model has adequate power for large effects (e.g.,
marriage, education) but limited power for moderate effects, particularly among racial
subgroups with smaller cell sizes.

**No nativity data.** The SCF does not identify foreign-born status, country of origin,
generation, or years in the U.S. The role of immigration in shaping racial fertility
differentials cannot be tested directly.

**Cross-sectional design.** We cannot distinguish between voluntary childlessness,
involuntary childlessness, and intended-but-not-yet-realized parenthood. All findings
are correlational.

**The "Other" race category.** This group aggregates Asian, Pacific Islander, American
Indian/Alaska Native, and multiracial respondents. Its internal heterogeneity severely
limits interpretability.

**Reference person coding.** The SCF codes the male as reference person in mixed-sex
couples. The `female` coefficient captures household structure, not gender.

---

# References

Adserà, A., & Ferrer, A. (2015). Immigrants and demography: Marriage, divorce, and
fertility. In B. R. Chiswick & P. W. Miller (Eds.), *Handbook of the Economics of
International Migration* (Vol. 1A, pp. 315–374). North-Holland.

Bricker, J., et al. (2023). Changes in U.S. family finances from 2019 to 2022: Evidence
from the Survey of Consumer Finances. *Federal Reserve Bulletin*.

Cai, Y., & Morgan, S. P. (2019). Low fertility in China and the United States.
*Population and Development Review*, 45(S1), 235–262.

Camarota, S. A., & Zeigler, K. (2025). The fertility of immigrants and natives in the
United States, 2023. Center for Immigration Studies.

Cohen, G. (2025). scf: Analyzing the Survey of Consumer Finances. R package version 1.0.5.

Guzzo, K. B., & Loo, J. (2023). Number of children to women aged 40–44, 1980–2022.
*Family Profiles* (FP-23-29). National Center for Family & Marriage Research.

Hanna, S. D., & Lindamood, S. (2008). Behind the numbers: Understanding the Survey of
Consumer Finances. *Journal of Financial Counseling and Planning*, 19(2), 52–66.

Lino, M., Kuczynski, K., Rodriguez, N., & Schap, T. (2020). Expenditures on children
by families, 2015. U.S. Department of Agriculture, Miscellaneous Report No. 1528-2015.

Livingston, G. (2015). Childlessness falls, family size grows among highly educated
women. Pew Research Center.

Osterman, M. J. K., Hamilton, B. E., Martin, J. A., Driscoll, A. K., & Valenzuela,
C. P. (2023). Births: Final data for 2021. *National Vital Statistics Reports*, 72(1),
1–53.

Parrado, E. A., & Morgan, S. P. (2008). Intergenerational fertility among Hispanic
women: New evidence of immigrant assimilation. *Demography*, 45(3), 651–671.

Pew Research Center. (2015). Childlessness falls, family size grows among highly
educated women.

Zill, N. (2025). Four million American women reach ages 40 to 50 without children.
Institute for Family Studies.

---

# Appendix: Technical Notes

**Multiple Imputation.** All statistics and models are pooled across the SCF's five
implicates using Rubin's Rules.

**Survey Weights.** All estimates — including figures — use SCF replicate weights
(balanced repeated replication). No unweighted estimates are reported.

**Sample Restriction.** Ages 31–44. This window aligns with the prime family formation
years and avoids conflating empty-nesting with childlessness.

**Tautology Correction.** `asinh_debt_excl_cc` (total debt minus credit card balance)
replaces `asinh_debt` to avoid mechanical prediction, consistent with the companion
analysis.

**Statistical Power.** With ~1,000 observations and `r length(model_predictors)`
predictors, the model is adequately powered for large effects but may miss moderate
subgroup differences. Confidence intervals, not p-values alone, should guide
interpretation of non-significant coefficients.

```{r session-info, results='asis'}
cat("\n\n**Session Information**\n\n")
cat(sprintf("R version: %s\n\n", R.version.string))
cat(sprintf("Platform: %s\n\n", R.version$platform))
cat(sprintf("Key packages: scf %s, glmnet %s, survey %s\n",
  packageVersion("scf"), packageVersion("glmnet"), packageVersion("survey")))
```
