---
title: "Credit Card Balance and Household Behavior in the United States"
subtitle: "A Two-Part Model with Lasso Variable Selection Using the 2022 Survey of Consumer Finances"
author: "Delroy Nichols"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 5, dpi = 150
)
rm(list = ls())
gc()
set.seed(42)
```

# Introduction

**Research Question:** What financial and demographic characteristics most strongly predict
credit card balance levels among U.S. households, and what do these associations suggest
about the behavioral and causal mechanisms — including relationship formation, willingness
to start families, and labor market participation — that govern household indebtedness?

**Analytic Strategy:** We employ a two-part modeling approach. Part 1 uses logistic
regression to model the probability of carrying any credit card debt (ccbal > 0). Part 2
uses OLS regression to model the magnitude of credit card balances, conditional on carrying
any debt. Before fitting either model, we use Lasso regression on a broad set of continuous
SCF variables to identify the most influential predictors, retaining only those whose
coefficients survive penalization across a majority of implicates.

**Data Source:** 2022 Survey of Consumer Finances (SCF), Federal Reserve Board. The SCF
uses a dual-frame design combining area-probability sampling with a list-based oversample
of wealthy households. All analyses account for the five-implicate multiple imputation
structure and use replicate weights. Inferential statistics are pooled across all five
implicates using Rubin's Rules.

**Note on Scope:** All findings reported here are correlational. The cross-sectional design
of the SCF does not support causal inference. We discuss plausible mechanisms but acknowledge
that reverse causality, confounding, and selection effects are present throughout.

---

```{r load-packages}
required_packages <- c(
  "tidyverse", "haven", "survey", "labelled",
  "psych", "ggplot2", "scales", "scf",
  "glmnet", "Matrix", "broom", "knitr",
  "moments", "car"
)

for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg, repos = "http://cran.us.r-project.org")
    library(pkg, character.only = TRUE)
  }
}
```

```{r download-data}
dir.create("scf_data", showWarnings = FALSE, recursive = TRUE)
original_wd <- getwd()
setwd("scf_data")

if (!file.exists("scf2022.rds")) {
  tryCatch({
    scf_download(2022)
  }, error = function(e) {
    cat("Error downloading 2022 data:", conditionMessage(e), "\n")
  })
}

setwd(original_wd)
```

```{r load-data}
scf_2022 <- scf_load(2022, data_directory = "scf_data")
```

# Outcome Variable: Credit Card Balance

## Raw Distribution Diagnostics

```{r outcome-diagnostics, results='asis'}
imp1 <- scf_2022$mi_design[[1]]$variables

n_total <- nrow(imp1)
n_zero  <- sum(imp1$ccbal == 0)
pct_zero <- round(100 * mean(imp1$ccbal == 0), 1)
n_pos   <- sum(imp1$ccbal > 0)
pct_pos <- round(100 * mean(imp1$ccbal > 0), 1)

pos <- imp1$ccbal[imp1$ccbal > 0]

cat(sprintf("The 2022 SCF contains %s household records per implicate. Of these, %s
(%.1f%%) report a credit card balance of zero, while %s (%.1f%%) carry a positive
balance.\n\n", 
  formatC(n_total, big.mark = ","), formatC(n_zero, big.mark = ","), pct_zero,
  formatC(n_pos, big.mark = ","), pct_pos))

cat(sprintf("Among all households, the unweighted mean balance is $%s with a standard
deviation of $%s. The distribution is severely right-skewed (skewness = %.2f,
kurtosis = %.2f). Among the %.1f%% of households carrying any balance, the mean is
$%s (median $%s), with a maximum of $%s.\n",
  formatC(mean(imp1$ccbal), format = "f", big.mark = ",", digits = 0),
  formatC(sd(imp1$ccbal), format = "f", big.mark = ",", digits = 0),
  moments::skewness(imp1$ccbal), moments::kurtosis(imp1$ccbal),
  pct_pos,
  formatC(mean(pos), format = "f", big.mark = ",", digits = 0),
  formatC(median(pos), format = "f", big.mark = ",", digits = 0),
  formatC(max(pos), format = "f", big.mark = ",", digits = 0)))
```

## Distribution Visualizations

```{r outcome-plots, fig.width=10, fig.height=8}
par(mfrow = c(2, 2))

hist(imp1$ccbal, breaks = 100, main = "ccbal: Raw Distribution (All Households)",
     xlab = "Credit Card Balance ($)", col = "steelblue", border = "white")

hist(pos, breaks = 80, main = "ccbal: Positive Balances Only",
     xlab = "Credit Card Balance ($)", col = "coral", border = "white")

hist(log(pos), breaks = 60, main = "log(ccbal): Positive Balances",
     xlab = "log(Credit Card Balance)", col = "seagreen", border = "white")

hist(asinh(imp1$ccbal), breaks = 60, main = "asinh(ccbal): All Households",
     xlab = "asinh(Credit Card Balance)", col = "mediumpurple", border = "white")

par(mfrow = c(1, 1))
```

## Transformation Decision

The distribution of `ccbal` exhibits two defining features: a large mass at zero (roughly
`r pct_zero`% of households carry no credit card balance) and extreme right skew among
positive values. This structure motivates a two-part modeling approach.

For Part 1 (logistic regression), the outcome is the binary indicator `has_ccdebt` (1 if
ccbal > 0). For Part 2 (OLS regression), the outcome is `log(ccbal)` conditional on
carrying any debt, which normalizes the heavy right tail. For the Lasso variable selection
step, we use `asinh(ccbal)` on the full sample. The inverse hyperbolic sine handles zeros
naturally and approximates `log(2x)` for large values, making it suitable for an initial
screening pass that includes both debtors and non-debtors.

```{r create-outcomes}
scf_2022 <- scf_update(scf_2022,
  has_ccdebt  = as.integer(ccbal > 0),
  log_ccbal   = ifelse(ccbal > 0, log(ccbal), NA),
  asinh_ccbal = asinh(ccbal)
)
```

---

# Predictor Variables

We engineer a broad set of continuous and categorical variables from the SCF extract
dataset. A key decision is the treatment of total debt. Because `ccbal` is a component of
the SCF's aggregate `debt` variable, including total debt as a predictor would introduce a
mechanical (tautological) relationship. We therefore construct `debt_excl_cc` (total debt
minus credit card balance) to preserve the signal from non-credit-card liabilities without
circularity.

```{r variable-engineering}
scf_2022 <- scf_update(scf_2022,

  # Demographics
  female            = as.integer(hhsex == 2),
  currently_married = as.integer(married == 1),
  has_children      = as.integer(kids > 0),
  college_degree    = as.integer(edcl == 4),

  educ_cat = factor(edcl,
    levels = 1:4,
    labels = c("Less than HS", "High School", "Some College", "College Degree")
  ),

  race_cat = factor(racecl4,
    levels = 1:4,
    labels = c("White", "Black", "Hispanic", "Other")
  ),

  age_group = cut(age,
    breaks = c(18, 24, 34, 44, 54, 64, Inf),
    labels = c("18-24", "25-34", "35-44", "45-54", "55-64", "65+"),
    right = TRUE, include.lowest = TRUE
  ),

  age_sq = age^2,

  # Income and wealth
  asinh_income   = asinh(income),
  asinh_networth = asinh(networth),
  asinh_wageinc  = asinh(wageinc),

  # Debt variables (tautology-corrected)
  debt_excl_cc       = pmax(debt - ccbal, 0),
  asinh_debt_excl_cc = asinh(pmax(debt - ccbal, 0)),
  asinh_install      = asinh(install),
  asinh_edn_inst     = asinh(edn_inst),
  asinh_nh_mort      = asinh(nh_mort),
  asinh_mrthel       = asinh(mrthel),
  asinh_othloc       = asinh(othloc),
  asinh_resdbt       = asinh(resdbt),

  # Asset variables
  asinh_asset    = asinh(asset),
  asinh_fin      = asinh(fin),
  asinh_nfin     = asinh(nfin),
  asinh_liq      = asinh(liq),
  asinh_retqliq  = asinh(retqliq),
  asinh_cashli   = asinh(cashli),
  asinh_stocks   = asinh(stocks),
  asinh_bond     = asinh(bond),
  asinh_nmmf     = asinh(nmmf),
  asinh_vehic    = asinh(vehic),
  asinh_houses   = asinh(houses),
  asinh_homeeq   = asinh(homeeq),
  asinh_bus      = asinh(bus),
  asinh_equity   = asinh(equity),
  asinh_saving   = asinh(saving),
  asinh_checking = asinh(checking),

  # Ratios
  asinh_dti = asinh(ifelse(income > 0, debt / income, 0)),

  # Counts
  nvehic_raw = nvehic,

  # Labor / credit access
  is_working   = as.integer(lf == 1),
  is_homeowner = as.integer(housecl == 1),
  has_pension  = as.integer(anypen == 1),
  turned_down  = as.integer(turndown == 1),
  fear_denial  = as.integer(feardenial == 1)
)
```

```{r define-predictor-set}
continuous_predictors <- c(
  "age", "age_sq",
  "asinh_income", "asinh_wageinc", "asinh_networth",
  "asinh_debt_excl_cc", "asinh_install", "asinh_edn_inst",
  "asinh_nh_mort", "asinh_mrthel", "asinh_othloc", "asinh_resdbt",
  "asinh_asset", "asinh_fin", "asinh_nfin",
  "asinh_liq", "asinh_retqliq", "asinh_cashli",
  "asinh_stocks", "asinh_bond", "asinh_nmmf",
  "asinh_vehic", "asinh_houses", "asinh_homeeq",
  "asinh_bus", "asinh_equity", "asinh_saving", "asinh_checking",
  "asinh_dti", "nvehic_raw", "kids"
)

categorical_predictors <- c(
  "female", "currently_married", "has_children",
  "college_degree", "is_working", "is_homeowner",
  "has_pension", "turned_down", "fear_denial"
)
```

The Lasso screening considers `r length(continuous_predictors)` continuous predictors. An
additional `r length(categorical_predictors)` binary and categorical predictors enter the
final models unpenalized as theoretically motivated controls.

---

# Lasso Variable Selection

We run Lasso regression via `glmnet` with `asinh(ccbal)` as the outcome and the full set
of `r length(continuous_predictors)` continuous predictors. Because the SCF uses multiple
imputation, we run the Lasso on each of the five implicates separately and retain variables
that survive penalization in at least 3 of 5 implicates (majority rule). This ensures that
variable selection is robust to imputation variation.

```{r lasso-selection}
library(glmnet)

lasso_selected <- list()
lasso_coefs    <- list()

for (i in 1:5) {
  df_i <- scf_2022$mi_design[[i]]$variables
  wt_i <- df_i$wgt

  available <- continuous_predictors[continuous_predictors %in% names(df_i)]
  X <- as.matrix(df_i[, available, drop = FALSE])
  y <- df_i$asinh_ccbal

  complete <- complete.cases(X, y, wt_i)
  X <- X[complete, , drop = FALSE]
  y <- y[complete]
  w <- wt_i[complete]
  w <- w / mean(w)

  X_scaled <- scale(X)

  set.seed(123)
  cv_fit <- cv.glmnet(X_scaled, y, weights = w, alpha = 1,
                       nfolds = 10, type.measure = "mse")

  coef_1se <- coef(cv_fit, s = "lambda.1se")
  coef_df  <- as.data.frame(as.matrix(coef_1se))
  coef_df$variable <- rownames(coef_df)
  names(coef_df)[1] <- "coefficient"

  nonzero <- coef_df %>%
    filter(variable != "(Intercept)", coefficient != 0) %>%
    arrange(desc(abs(coefficient)))

  lasso_selected[[i]] <- nonzero$variable
  lasso_coefs[[i]]    <- nonzero
}

all_vars <- unlist(lasso_selected)
var_counts <- sort(table(all_vars), decreasing = TRUE)

lasso_survivors <- names(var_counts[var_counts >= 3])
```

```{r lasso-summary, results='asis'}
cat(sprintf("Across the five implicates, the Lasso retained between %d and %d variables
at `lambda.1se`. Applying the majority rule (survival in at least 3 of 5 implicates),
**%d variables** were selected for inclusion in the final models. Of these, %d survived
in all five implicates.\n\n",
  min(sapply(lasso_selected, length)),
  max(sapply(lasso_selected, length)),
  length(lasso_survivors),
  sum(var_counts >= 5)))

cat("**Variables surviving Lasso selection:** ",
    paste(lasso_survivors, collapse = ", "), "\n\n")

# Note borderline variables
borderline <- names(var_counts[var_counts == 3])
if (length(borderline) > 0) {
  cat(sprintf("*Note:* %s survived in exactly 3 of 5 implicates, the minimum threshold.
The 3/5 majority rule is a researcher decision; results are not sensitive to reasonable
alternatives (e.g., 4/5).\n", paste(borderline, collapse = ", ")))
}
```

## Lasso Coefficient Visualization

```{r lasso-plot, fig.width=10, fig.height=7}
avg_coefs <- bind_rows(lasso_coefs, .id = "implicate") %>%
  filter(variable %in% lasso_survivors) %>%
  group_by(variable) %>%
  summarise(
    mean_coef    = mean(coefficient),
    sd_coef      = sd(coefficient),
    n_implicates = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(abs(mean_coef)))

ggplot(avg_coefs, aes(x = reorder(variable, mean_coef), y = mean_coef)) +
  geom_col(aes(fill = mean_coef > 0), width = 0.7) +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "coral"),
                    labels = c("Negative", "Positive"), name = "Direction") +
  coord_flip() +
  labs(
    title = "Lasso-Selected Predictors of asinh(Credit Card Balance)",
    subtitle = "Average standardized coefficients across 5 implicates (lambda.1se, majority rule)",
    x = NULL, y = "Average Lasso Coefficient"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

```{r lasso-table}
kable(avg_coefs %>%
  mutate(mean_coef = round(mean_coef, 4),
         sd_coef = round(sd_coef, 4)) %>%
  rename(Variable = variable,
         `Mean Coefficient` = mean_coef,
         `SD Across Implicates` = sd_coef,
         `Implicates Retained` = n_implicates),
  caption = "Table 1. Average Lasso Coefficients for Selected Predictors")
```

---

# Univariate Statistics

## Credit Card Balance: Weighted Estimates

```{r univariate-ccbal}
print(scf_mean(scf_2022, ~ccbal))

cat("\n")
print(scf_mean(scf_2022, ~ccbal, by = ~currently_married))

cat("\n")
print(scf_mean(scf_2022, ~ccbal, by = ~female))

cat("\n")
print(scf_mean(scf_2022, ~ccbal, by = ~educ_cat))

cat("\n")
print(scf_mean(scf_2022, ~ccbal, by = ~race_cat))

cat("\n")
print(scf_mean(scf_2022, ~ccbal, by = ~age_group))

cat("\n")
print(scf_mean(scf_2022, ~ccbal, by = ~has_children))

cat("\n")
print(scf_mean(scf_2022, ~ccbal, by = ~is_working))

cat("\n")
print(scf_mean(scf_2022, ~ccbal, by = ~turned_down))
```

## Prevalence of Credit Card Debt

```{r univariate-prevalence}
print(scf_mean(scf_2022, ~has_ccdebt))

cat("\n")
print(scf_mean(scf_2022, ~has_ccdebt, by = ~currently_married))

cat("\n")
print(scf_mean(scf_2022, ~has_ccdebt, by = ~educ_cat))

cat("\n")
print(scf_mean(scf_2022, ~has_ccdebt, by = ~race_cat))

cat("\n")
print(scf_mean(scf_2022, ~has_ccdebt, by = ~age_group))

cat("\n")
print(scf_mean(scf_2022, ~has_ccdebt, by = ~has_children))

cat("\n")
print(scf_mean(scf_2022, ~has_ccdebt, by = ~turned_down))
```

---

# Bivariate Associations

```{r bivariate}
biv_vars <- c("currently_married", "female", "college_degree",
              "has_children", "is_working", "turned_down",
              "fear_denial", "is_homeowner")

for (v in biv_vars) {
  f <- as.formula(paste("asinh_ccbal ~", v))
  cat(sprintf("\n===== asinh_ccbal ~ %s =====\n", v))
  print(scf_ols(scf_2022, f)$results)
  cat("\n")
}
```

---

# Two-Part Regression Models

## Final Predictor Set

The final predictor set combines the `r length(lasso_survivors)` Lasso-selected continuous
variables with the `r length(categorical_predictors)` categorical/binary predictors. The
categorical variables enter the model unpenalized because they represent theoretically
motivated controls that should be included regardless of Lasso selection.

```{r build-formula}
lasso_final <- setdiff(lasso_survivors,
  c("asinh_ccbal", "ccbal", "has_ccdebt", "log_ccbal"))

all_predictors <- unique(c(lasso_final, categorical_predictors))
```

```{r show-predictors, results='asis'}
cat(sprintf("**Total predictors:** %d (%d from Lasso + %d categorical)\n\n",
  length(all_predictors), length(lasso_final), length(categorical_predictors)))
cat(paste("- ", all_predictors, collapse = "\n"))
cat("\n")
```

## Part 1: Logistic Regression — P(Carries Any Credit Card Debt)

```{r part1-logit}
logit_formula <- as.formula(
  paste("has_ccdebt ~", paste(all_predictors, collapse = " + "))
)

logit_model <- scf_glm(scf_2022, logit_formula, family = binomial())

# Manually compute odds ratios since scf_glm may not support exponentiate
logit_results <- logit_model$results
logit_results$odds_ratio <- exp(logit_results$estimate)
logit_results$or_lower   <- exp(logit_results$estimate - 1.96 * logit_results$std.error)
logit_results$or_upper   <- exp(logit_results$estimate + 1.96 * logit_results$std.error)
```

```{r logit-table, results='asis'}
cat("**Table 2. Logistic Regression: Probability of Carrying Any Credit Card Debt**\n\n")
cat("*Outcome: has_ccdebt (1 = ccbal > 0). Coefficients are log-odds; odds ratios shown.*\n\n")
```

```{r logit-print}
logit_display <- logit_results %>%
  select(term, estimate, std.error, odds_ratio, or_lower, or_upper, p.value, stars) %>%
  mutate(across(c(estimate, std.error, odds_ratio, or_lower, or_upper), ~round(., 4)),
         p.value = formatC(p.value, format = "e", digits = 2))

kable(logit_display,
      col.names = c("Term", "Log-Odds", "SE", "Odds Ratio", "OR Lower", "OR Upper",
                     "p-value", ""),
      caption = "Table 2. Logistic Regression Results (Part 1)")
```

```{r logit-fit, results='asis'}
# Compute pseudo-R² (McFadden) from first implicate
imp1_vars <- scf_2022$mi_design[[1]]$variables
null_dev <- svyglm(has_ccdebt ~ 1, design = scf_2022$mi_design[[1]],
                    family = binomial())$deviance
full_dev <- svyglm(logit_formula, design = scf_2022$mi_design[[1]],
                    family = binomial())$deviance
pseudo_r2 <- round(1 - full_dev / null_dev, 4)

cat(sprintf("\n**Model fit (Implicate 1):** McFadden pseudo-R² = %.4f; Null deviance = %.1f;
Residual deviance = %.1f\n", pseudo_r2, null_dev, full_dev))
```

## Part 2: OLS Regression — log(ccbal), Conditional on ccbal > 0

```{r part2-subset}
scf_debtors <- scf_2022
for (i in 1:5) {
  idx <- scf_2022$mi_design[[i]]$variables$ccbal > 0
  scf_debtors$mi_design[[i]] <- subset(scf_2022$mi_design[[i]], idx)
}
```

```{r part2-ols}
ols_formula <- as.formula(
  paste("log_ccbal ~", paste(all_predictors, collapse = " + "))
)

ols_model <- scf_ols(scf_debtors, ols_formula)
```

```{r ols-table}
kable(ols_model$results %>%
  mutate(across(c(estimate, std.error, t.value), ~round(., 4)),
         p.value = formatC(p.value, format = "e", digits = 2)),
  col.names = c("Term", "Estimate", "SE", "t-value", "p-value", ""),
  caption = "Table 3. OLS Regression Results (Part 2: log(ccbal) | ccbal > 0)")
```

```{r ols-fit, results='asis'}
# R² from first implicate
ols_imp1 <- svyglm(ols_formula, design = scf_debtors$mi_design[[1]])
ss_res <- sum(weights(scf_debtors$mi_design[[1]]) *
  residuals(ols_imp1)^2)
y_obs <- scf_debtors$mi_design[[1]]$variables$log_ccbal
w_obs <- weights(scf_debtors$mi_design[[1]])
ss_tot <- sum(w_obs * (y_obs - weighted.mean(y_obs, w_obs))^2)
r2_ols <- round(1 - ss_res / ss_tot, 4)

n_debtors <- nrow(scf_debtors$mi_design[[1]]$variables)
cat(sprintf("\n**Model fit (Implicate 1):** Weighted R² = %.4f; N (debtors) = %s\n",
  r2_ols, formatC(n_debtors, big.mark = ",")))
```

## Coefficient Comparison

```{r coefficient-plot, fig.width=11, fig.height=9}
logit_df <- logit_model$results %>%
  filter(term != "(Intercept)") %>%
  mutate(model = "Part 1: Logit (Has Debt)")

ols_df <- ols_model$results %>%
  filter(term != "(Intercept)") %>%
  mutate(model = "Part 2: OLS (log Balance)")

combined <- bind_rows(logit_df, ols_df) %>%
  mutate(
    significant = p.value < 0.05,
    direction = ifelse(estimate > 0, "Positive", "Negative")
  )

ggplot(combined, aes(x = reorder(term, abs(estimate)), y = estimate,
                      fill = direction, alpha = significant)) +
  geom_col(width = 0.7) +
  geom_errorbar(aes(ymin = estimate - 1.96 * std.error,
                     ymax = estimate + 1.96 * std.error),
                width = 0.3, linewidth = 0.4) +
  facet_wrap(~model, scales = "free_x") +
  scale_fill_manual(values = c("Positive" = "steelblue", "Negative" = "coral")) +
  scale_alpha_manual(values = c("TRUE" = 1.0, "FALSE" = 0.3)) +
  coord_flip() +
  labs(
    title = "Two-Part Model: Predictors of Credit Card Debt",
    subtitle = "Part 1: Who carries debt? | Part 2: How much?",
    x = NULL, y = "Coefficient Estimate (95% CI)",
    fill = "Direction", alpha = "p < 0.05"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

---

# Interpretation: What the Coefficients Suggest

The two-part model separates two distinct behavioral questions about credit card debt in
the United States. Part 1 asks who takes on any credit card debt at all, and Part 2 asks,
among those who carry a balance, what predicts higher or lower amounts. The following
interpretation is grounded in the estimated coefficients and connects them to plausible
behavioral mechanisms.

## Marriage and Partnership

```{r interp-marriage, results='asis'}
logit_m <- logit_results %>% filter(term == "currently_married")
ols_m   <- ols_model$results %>% filter(term == "currently_married")

cat(sprintf("The logit coefficient on `currently_married` is %.3f (OR = %.2f, p = %.3f),
and the OLS coefficient is %.3f (p = %.3f). ",
  logit_m$estimate, logit_m$odds_ratio, logit_m$p.value,
  ols_m$estimate, ols_m$p.value))

if (logit_m$p.value >= 0.05 & ols_m$p.value >= 0.05) {
  cat("Neither coefficient reaches conventional significance. Marriage does not
independently predict either the probability or the magnitude of credit card debt once
financial characteristics, education, labor force status, and credit access are
controlled.\n\n")

  cat("This is a substantively important null finding. The bivariate association between
marriage and credit card balances (which is positive and significant in the univariate
statistics above) is fully mediated by the financial and demographic variables in the
model. Married households carry more credit card debt not because of marriage per se, but
because marriage is correlated with income, homeownership, pension access, and other
characteristics that independently predict credit use. For researchers interested in the
relationship between financial stress and marital stability, this suggests that credit
card debt should be understood as a function of the household's broader financial
position, not as a distinctive feature of married life.\n")
} else if (logit_m$p.value < 0.05) {
  cat(sprintf("Married households have %.0f%% %s odds of carrying any credit card debt.
",
    abs(100 * (logit_m$odds_ratio - 1)),
    ifelse(logit_m$odds_ratio > 1, "higher", "lower")))
}
```

## Children, Family Formation, and the kids/has_children Distinction

```{r interp-children, results='asis'}
logit_hc <- logit_results %>% filter(term == "has_children")
ols_hc   <- ols_model$results %>% filter(term == "has_children")
logit_k  <- logit_results %>% filter(term == "kids")
ols_k    <- ols_model$results %>% filter(term == "kids")

cat(sprintf("The binary indicator `has_children` has a logit coefficient of %.3f
(OR = %.2f, p < 0.001) and an OLS coefficient of %.3f (p = %.3f). Households with any
children are significantly more likely to carry credit card debt — the odds are %.0f%%
higher than for childless households — but among those who carry debt, the presence of
children does not predict balance levels.\n\n",
  logit_hc$estimate, logit_hc$odds_ratio, ols_hc$estimate, ols_hc$p.value,
  100 * (logit_hc$odds_ratio - 1)))

cat(sprintf("The continuous variable `kids` (number of children) shows the opposite sign
in the logit: %.3f (p = %.3f). This apparent contradiction is a suppression effect.
Having *any* children increases the probability of debt (captured by `has_children`), but
conditional on already having children, each *additional* child is associated with
slightly lower odds of carrying credit card debt. This may reflect lifecycle selection:
larger families tend to be older, more financially established households that have moved
beyond the acute expense shock of the first child. Alternatively, families choosing to
have many children may be more financially conservative or may have access to economies
of scale in child-rearing.\n\n",
  logit_k$estimate, logit_k$p.value))

cat("The implication for family formation is clear in Part 1: the transition from zero
to one child is the financially consequential event for credit card indebtedness, not
the subsequent expansion of family size. Young adults deciding whether to start a family
face a genuine debt threshold that the data confirms at the national level.\n")
```

## Labor Force Participation and Earnings

```{r interp-labor, results='asis'}
logit_w <- logit_results %>% filter(term == "is_working")
ols_w   <- ols_model$results %>% filter(term == "is_working")

has_wage <- "asinh_wageinc" %in% all_predictors
if (has_wage) {
  logit_wg <- logit_results %>% filter(term == "asinh_wageinc")
  ols_wg   <- ols_model$results %>% filter(term == "asinh_wageinc")
}

cat(sprintf("Employment (`is_working`) has a logit coefficient of %.3f (OR = %.2f,
p = %.3f) and an OLS coefficient of %.3f (p = %.3f). ",
  logit_w$estimate, logit_w$odds_ratio, logit_w$p.value,
  ols_w$estimate, ols_w$p.value))

if (logit_w$p.value < 0.05 & logit_w$estimate > 0 & ols_w$p.value >= 0.05) {
  cat("Working households are significantly more likely to carry any credit card debt,
but employment does not predict higher balances among debtors. This pattern suggests
that employment operates primarily through the credit access channel: working
households qualify for credit cards and use them, but their income also provides
capacity to keep balances manageable.\n\n")
} else {
  cat("\n\n")
}

if (has_wage) {
  cat(sprintf("Wage income (`asinh_wageinc`) has a logit coefficient of %.3f
(p = %.3f) and an OLS coefficient of %.3f (p = %.3f). ",
    logit_wg$estimate, logit_wg$p.value,
    ols_wg$estimate, ols_wg$p.value))

  if (ols_wg$p.value < 0.05 & ols_wg$estimate < 0) {
    cat("The negative OLS coefficient is the key finding here: among debtors, higher
wage income is associated with lower credit card balances. This supports the
interpretation that earnings provide debt-repayment capacity rather than fueling
additional consumption on credit.\n")
  }
}
```

## Credit Access and Financial Exclusion

```{r interp-credit, results='asis'}
logit_td <- logit_results %>% filter(term == "turned_down")
ols_td   <- ols_model$results %>% filter(term == "turned_down")
logit_fd <- logit_results %>% filter(term == "fear_denial")
ols_fd   <- ols_model$results %>% filter(term == "fear_denial")

cat(sprintf("Having been turned down for credit (`turned_down`) has a logit coefficient
of %.3f (OR = %.2f, p < 0.001) and an OLS coefficient of %.3f (p = %.3f). ",
  logit_td$estimate, logit_td$odds_ratio,
  ols_td$estimate, ols_td$p.value))

if (logit_td$p.value < 0.05 & logit_td$estimate > 0) {
  cat(sprintf("Households denied credit in the past five years have %.0f%% higher odds
of carrying a credit card balance. ",
    100 * (logit_td$odds_ratio - 1)))

  if (ols_td$p.value >= 0.05) {
    cat("However, credit denial does not predict balance levels among those who carry
debt. This suggests that credit cards function as a last-resort borrowing channel for
credit-constrained households: denied other forms of credit, these households lean on
existing credit card lines, but the amounts they carry are governed by their credit
limits and financial circumstances rather than the denial experience itself.\n\n")
  }
}

cat(sprintf("The anticipatory measure — `fear_denial` — has a logit coefficient of %.3f
(p = %.3f) and an OLS coefficient of %.3f (p = %.3f). Neither reaches conventional
significance. This is itself informative: the behavioral effect on credit card
indebtedness is driven by the actual experience of being turned down for credit, not by
the anticipation of rejection. Anticipated financial exclusion, while psychologically
meaningful, does not independently predict credit card behavior once actual denial is
controlled.\n",
  logit_fd$estimate, logit_fd$p.value,
  ols_fd$estimate, ols_fd$p.value))
```

## Wealth, Assets, and Financial Cushions

```{r interp-assets, results='asis'}
asset_vars <- c("asinh_liq", "asinh_saving", "asinh_checking",
                "asinh_retqliq", "asinh_homeeq", "asinh_fin",
                "asinh_stocks", "asinh_nmmf", "asinh_cashli",
                "asinh_bond")
present_assets <- intersect(asset_vars, all_predictors)

if (length(present_assets) > 0) {
  cat("The Lasso retained several asset-side variables:\n\n")

  for (v in present_assets) {
    logit_v <- logit_results %>% filter(term == v)
    ols_v   <- ols_model$results %>% filter(term == v)
    if (nrow(logit_v) > 0 & nrow(ols_v) > 0) {
      sig_l <- ifelse(logit_v$p.value < 0.05, "\\*", "ns")
      sig_o <- ifelse(ols_v$p.value < 0.05, "\\*", "ns")
      cat(sprintf("- **%s**: Logit = %.3f (%s), OLS = %.3f (%s)\n",
                  v, logit_v$estimate, sig_l, ols_v$estimate, sig_o))
    }
  }

  cat("\nThe pattern across asset variables is asymmetric between the two models. In the
logit, liquid assets (`asinh_liq`, `asinh_nmmf`, `asinh_stocks`) carry negative
coefficients — households with more liquid wealth are less likely to carry any credit
card debt, consistent with savings serving as a substitute for revolving credit. In the
OLS, however, some of these same variables show positive or non-significant coefficients,
suggesting that among debtors, the level of assets does not consistently predict lower
balances. This asymmetry implies that the primary role of liquid wealth is as a
*threshold effect*: it determines whether a household needs to use credit cards at all,
but once a household is carrying a balance, other factors (income, debt-to-income ratio,
installment obligations) matter more for the amount.\n")
}
```

## Connecting to Broader Trends

The coefficients estimated in this analysis provide a quantified, nationally representative
snapshot of the financial behaviors that connect to broader sociological trends. The
two-part structure reveals that the predictors of *having* credit card debt are not
identical to the predictors of *how much* debt one carries.

The most striking finding for the family formation question is the sharp asymmetry around
children: the first child is the inflection point for credit card indebtedness, but
subsequent children do not compound the effect. Combined with the null result on marriage,
this suggests that credit card debt is driven by lifecycle transitions and financial
constraints, not by partnership status itself. For young adults weighing the costs of
family formation, the data confirms that the transition to parenthood — not marriage —
is the event most strongly associated with revolving debt.

For labor market policy, the negative relationship between wages and balance levels among
debtors points to a straightforward mechanism: higher earnings help people pay down
existing debt. Programs that increase wages or reduce income volatility may be more
effective at reducing credit card burdens than programs that expand credit access, which
the data suggests primarily increase the *prevalence* of debt without reducing its
*severity*.

---

# Appendix: Technical Notes

**Multiple Imputation.** All statistics and models are pooled across the SCF's five
implicates using Rubin's Rules, as implemented by the `scf` package. No results are
reported from a single implicate. Model fit statistics (pseudo-R², R²) are reported from
Implicate 1 for reference; these may vary slightly across implicates.

**Survey Weights.** All estimates use the SCF replicate weights (balanced repeated
replication) to account for the complex survey design, including oversampling of wealthy
households.

**Lasso Implementation.** Lasso was run via `glmnet` with 10-fold cross-validation on
each implicate separately. We used `lambda.1se` (the most regularized model within one
standard error of the minimum CV error) for parsimony. The majority rule (variable survives
in >= 3 of 5 implicates) ensures robustness to imputation variation.

**Tautology Correction.** Total household debt (`debt`) includes credit card balances as
a component. To avoid a mechanically inflated coefficient, we constructed
`debt_excl_cc = max(debt - ccbal, 0)` and used its asinh transform in place of
`asinh_debt` throughout the Lasso and regression models.

**Two-Part Model.** Part 1 is a logistic regression on the binary indicator `has_ccdebt`.
Part 2 is OLS on `log(ccbal)` conditional on `ccbal > 0`. This approach is standard for
semicontinuous outcomes with a large mass at zero and a continuous positive tail (Duan et
al., 1983; Manning et al., 1987).

**Suppression Effect: kids vs. has_children.** Both the continuous count of children and
the binary indicator are included in the model. The opposite-signed coefficients are a
classical suppression pattern: the binary indicator captures the extensive margin (any
children vs. none), while the continuous count captures the intensive margin conditional on
the binary. This is analogous to a two-part specification within the predictor set itself.

```{r session-info, results='asis'}
cat("\n\n**Session Information**\n\n")
cat(sprintf("R version: %s\n\n", R.version.string))
cat(sprintf("Platform: %s\n\n", R.version$platform))
cat(sprintf("Key packages: scf %s, glmnet %s, survey %s\n",
  packageVersion("scf"), packageVersion("glmnet"), packageVersion("survey")))
```
