---
title: "Segmented Assimilation in the American Financial System: Origin-Based Gradients in Immigrant Household Finances"
subtitle: "Evidence from the 2022 Survey of Consumer Finances"
author: "Delroy Nichols"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 5, dpi = 150
)
rm(list = ls())
gc()
set.seed(42)
```

# Origin-Based Gradients in Immigrant Household Finances

## Abstract

For decades, scholars have debated what dictates an immigrant's financial
trajectory in the United States. The human-capital convergence model holds that
education, income, and time spent in the U.S. determine financial outcomes
regardless of origin. The segmented-assimilation framework contends that distinct
modes of incorporation (ethnic-enclave resources, neighborhood reception, racial
prejudice, and the structure of the receiving labor market) produce divergent
financial trajectories across origin groups. Because historical wealth surveys did
not track national origin, neither framework could be directly tested against
U.S. household wealth data until the 2022 Survey of Consumer Finances added new
national-origin variables.

This paper analyzes the financial outcomes of 1,103 foreign-born households in
the 2022 SCF across three pathway outcomes: retirement-account ownership
(parallel integration), credit denial (downward assimilation), and homeownership
(selective assimilation). The analysis reports two complementary specifications.
**Model A** estimates origin's total effect on each outcome, controlling only for
plausibly exogenous confounders (duration of U.S. residence, age). **Model B**
adds the human-capital and household-formation channels (educational attainment,
log income, currently married) that segmented-assimilation theory identifies as
the mechanisms through which origin produces wealth outcomes. Reporting both
specifications follows recent guidance against post-treatment bias (Rosenbaum,
1984; Acharya, Blackwell, & Sen, 2016) in observational analyses of theorized
causal pathways.

Model A's origin coefficients are mostly small and rarely reach statistical
significance at the analytic sample's power; the Caribbean/West Indies group
produces several visibly elevated point estimates that fail to clear significance
against wide confidence intervals (n=40). Model B's origin coefficients cluster
near 1.0 (0 of 12 reach p<0.05), indicating that the human-capital and
household-formation mediators absorb the channels through which origin operates.
Duration-stratified Firth-penalized models confirm that the null pattern is not
an artifact of pooling across long-tenured immigrants: 0 of 24 fittable
origin × stratum cells reach p<0.05, and the recent-arrival stratum (≤15 years
in the U.S., n=158) produces origin coefficients clustered as tightly around
OR=1.0 as the long-tenured stratum (>15 years, n=945). This is a direct
falsification of the duration-narrowing version of segmented-assimilation
theory, which predicts the largest origin effects in the early-incorporation
window.

The decomposition is consistent with both theoretical frameworks at the
descriptive level: segmented assimilation predicts that origin produces wealth
outcomes *through* differential modes of incorporation (the mediators), while
human-capital convergence predicts that the mediators are the operative
variables. The two frameworks agree on the empirical pattern; they disagree on
its causal interpretation, a question this paper's cross-sectional design
cannot adjudicate. The narrowest empirical claim the data support is that
within-race origin disaggregation in the 2022 SCF adds limited additional
explanatory power once standard human-capital variables are included, a claim
about descriptive analysis, not about whether the structural mechanisms
segmented-assimilation theory describes are operative in U.S. immigrant
wealth-building.

**Data Source:** 2022 Survey of Consumer Finances (SCF), Federal Reserve Board. All
estimates are survey-weighted; all models are pooled across the SCF's five multiply-
imputed datasets using Rubin's Rules.

**Note on Scope:** All findings are correlational and limited to the foreign-born
respondent sample. The SCF's cross-sectional design and absence of parental nativity
preclude causal claims and intergenerational analysis.

## Introduction: The Immigrant Wealth Debate

When a foreign-born household settles in the United States, what ultimately
dictates its financial future? For decades, economists and sociologists have
fiercely debated whether an immigrant's trajectory is determined by individual
effort and time, or by enduring structural barriers tied to where they come from.

At the heart of this debate are two competing theoretical models. The first, the
**human-capital convergence model** (Borjas, 1987, 1999), views the American
economy as a great equalizer. Foundational studies by Chiswick (1978) and Blau and
Graham (1990) contend that factors such as education, English proficiency, age at
arrival, and time spent in the U.S. sufficiently account for economic outcomes.
In this view, once you control for a household's "human capital" and how long they
have had to build a life in their new country, their region of origin has no
independent effect on their financial success.

The second model, the **segmented-assimilation framework**, makes the exact
opposite prediction. Developed by Alejandro Portes, Rubén Rumbaut, and Min Zhou
(Portes & Zhou, 1993; Portes & Rumbaut, 2001), this theory argues that American
society is highly stratified. It posits that the context of an immigrant's
reception, including racial prejudices, neighborhood disparities, and labor-market
structures, differs systematically across origin groups (Zhou, 1997).
Consequently, immigrants are forced into three distinct incorporation pathways:
*parallel integration* into the white middle class, *downward assimilation* into a
racialized underclass through structural exclusion, or *selective assimilation*,
where wealth is built primarily through ethnic-community networks. According to
this theory, a Mexican-origin household and an Asian-origin household with
identical college degrees and incomes might still experience vastly different
access to mainstream financial institutions simply because of their origin.

Until very recently, it was impossible to directly test which of these theories
accurately describes immigrant wealth building in the U.S. While previous
nationally representative wealth surveys could identify if a respondent was
foreign-born, they did not collect detail on the specific region of origin. That
changed with the 2022 wave of the Survey of Consumer Finances (SCF). By adding new
variables tracking respondent and spouse national origin alongside the exact
number of years lived in the United States, the 2022 SCF provided the first
nationally representative dataset capable of answering this question.

This paper uses these newly available data to characterize origin-region
patterns in three pathway outcomes, retirement-account ownership, credit denial,
and homeownership, across the 2022 SCF foreign-born sample (n=1,103). The
analytical approach reports two complementary specifications. The first (Model A)
estimates origin's total effect on each outcome, controlling only for plausibly
exogenous confounders. The second (Model B) adds the human-capital and
household-formation channels that segmented-assimilation theory identifies as the
mechanisms through which origin is theorized to act. The choice of which
covariates to include is itself a theoretical commitment: controlling for variables
that lie on the causal pathway between origin and wealth outcomes produces
post-treatment bias and changes the estimand from "origin's total effect" to
"origin's residual effect net of the theorized mechanisms" (Rosenbaum, 1984;
Acharya, Blackwell, & Sen, 2016). Reporting both specifications makes the
distinction explicit and permits both theoretical readings of the empirical
pattern.

The analysis finds that Model A's origin coefficients are mostly small and rarely
clear conventional statistical significance, while Model B's origin coefficients
cluster near 1.0 with no cell reaching p<0.05. The comparison between the two
specifications, the share of origin's total effect on wealth that is absorbed by
the human-capital and household mediators, is the paper's central descriptive
finding. It is consistent with segmented-assimilation theory's prediction that
origin produces wealth outcomes through differential modes of incorporation; it is
also consistent with human-capital convergence's prediction that the mediators are
themselves the operative explanatory variables. The two frameworks agree on the
empirical pattern in this sample but disagree on its causal interpretation, a
question the cross-sectional SCF design cannot adjudicate.

## Why This Matters: Beyond Race-Only Frameworks

The debate between human-capital convergence and segmented assimilation is not
merely an academic exercise; it has profound implications for how we understand and
address inequality in America. Historically, wealth and credit access disparities
have been studied primarily through the lens of broad racial categories. Studies
comparing Black, Hispanic, and White households often treat these populations as
homogeneous. However, prior sociological research suggests that distinct groups
(such as Black Caribbeans, Mexican-origin Hispanics, and Cuban-origin Hispanics)
may experience vastly different financial trajectories due to their unique
migration histories and community resources (Logan, 2007; Smith, 2006; Telles &
Ortiz, 2008).

The 2022 SCF allows the field to finally test whether the divergent pathways
predicted by segmented assimilation actually appear in national wealth data with
enough magnitude to distinguish these origin groups. The policy stakes are
significant. If segmented assimilation accurately describes immigrant wealth
building, then financial interventions designed broadly for "Hispanic" or "Asian"
households are fundamentally flawed and must be redesigned to address deep,
origin-specific heterogeneity. If, however, the data support human-capital
convergence, then existing race-based and income-based frameworks remain the
appropriate analytical lens for studying U.S. immigrant wealth outcomes, not because
those frameworks have already closed observed disparities, but because the
within-race-category origin distinctions that segmented-assimilation theory
proposes do not appear to organize foreign-born households' financial lives in a
way the SCF can detect.

---

```{r load-packages, include=FALSE, message=FALSE, warning=FALSE, results='hide'}
required_packages <- c(
  "tidyverse", "haven", "survey", "labelled",
  "ggplot2", "scales", "scf",
  "broom", "knitr", "moments", "logistf",
  "patchwork", "gridExtra"
)

for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg, repos = "http://cran.us.r-project.org", quiet = TRUE)
    suppressPackageStartupMessages(library(pkg, character.only = TRUE))
  }
}
```

```{r download-data}
# Same download pattern as the credit-card-debt analysis (Paper 1).
# scf_download writes .rds files to the current working directory.

dir.create("scf_data", showWarnings = FALSE, recursive = TRUE)
original_wd <- getwd()

# Move into the data folder
setwd("scf_data")

library(haven)

# Read from your local absolute path
raw_macro_df <- read_dta("C:/Users/delro/OneDrive/Documents/scf2022s/p22i6.dta")

# Save directly to the current working directory (which is already scf_data)
saveRDS(raw_macro_df, "p22i6.rds") 

# Continue with the download check for the bulletin extract
if (!file.exists("scf2022.rds")) {
  tryCatch({
    scf_download(2022)
  }, error = function(e) {
    cat("Error downloading 2022 data:", conditionMessage(e), "\n")
  })
}

# Return to the root project directory
setwd(original_wd)
```

```{r load-data}
scf_2022 <- scf_load(2022, data_directory = "scf_data")
```

<!-- Setup chunks: variable merge and raw-frequency diagnostics. Suppressed in
the rendered document because these are workflow steps rather than findings.
The code still runs to construct foreign_born, years_in_us, and origin_region
variables used downstream. -->

```{r diagnose-immigration-vars, include=FALSE}
imp1 <- scf_2022$mi_design[[1]]$variables

# Variables we need
# X6915 = respondent national origin (the variable for origin region)
# X6906 = years respondent has lived in the U.S. (the YIU variable)
# X7796 = years spouse has lived in the U.S. (auxiliary)
# X7404 = spouse Hispanic/Latino self-id (auxiliary)
# X8049 = spouse race (used in swap correction)
# X8000 = sex/role swap indicator
#
# CORRECTED 2026-04-29: V5 incorrectly used X7437-X7443 as years-in-US.
# Per the official SCF 2022 changes file, X7437-X7443 are MORTGAGE FORBEARANCE
# variables, not immigration variables. The true years-in-US variable is X6906
# for the respondent (and X7796 for the spouse).
imm_vars <- c("X6915", "X6906", "X7796", "X7404", "X8000", "X8049")

present_in_extract <- imm_vars %in% names(imp1)

cat("**Immigration-variable availability in the scf package extract:**\n\n")
diag_df <- data.frame(
  Variable = imm_vars,
  `In extract` = ifelse(present_in_extract, "Yes", "No"),
  check.names = FALSE
)
print(kable(diag_df))
cat("\n\n")

need_raw_merge <- !all(present_in_extract)

if (need_raw_merge) {
  cat("Some required immigration variables are missing from the bulletin extract. ",
      "Attempting to merge them in from the raw 2022 SCF data file...\n\n", sep = "")
} else {
  cat("All required immigration variables are present in the extract. ",
      "No raw merge needed.\n\n", sep = "")
}
```

```{r merge-raw-immigration-vars, include=FALSE}
# ============================================================================
# Merge raw immigration variables from the underlying SCF .rds file
# ============================================================================
# The scf package downloads the raw 2022 SCF data into scf_data/ as well as
# building the bulletin extract. The raw file contains all X-variables; we
# read it, select just the immigration-related vars + ID, and merge into
# each implicate's $variables data frame.
#
# Common file names produced by scf_download():
#   - p22i6.rds (raw public-use, 5 implicates stacked, with implicate id)
#   - scf2022.rds (the bulletin extract)
#
# We try a few candidate filenames defensively.

if (need_raw_merge) {

  raw_candidates <- c(
    file.path("scf_data", "p22i6.rds"),
    file.path("scf_data", "scfp2022.rds"),
    file.path("scf_data", "p22.rds"),
    "p22i6.rds",
    "scfp2022.rds"
  )

  raw_path <- raw_candidates[file.exists(raw_candidates)][1]

  if (!is.na(raw_path)) {
    raw_data <- readRDS(raw_path)

    # Standardize variable name casing — the raw file may have either
    # uppercase or lowercase. We try uppercase first since X-variables
    # are conventionally uppercase.
    raw_path <- raw_candidates[file.exists(raw_candidates)][1]

  if (!is.na(raw_path)) {
    raw_data <- readRDS(raw_path)

    # FIX: Force all raw data columns to uppercase immediately.
    # This aligns the Stata output with the canonical SCF uppercase format.
    names(raw_data) <- toupper(names(raw_data))

    raw_names <- names(raw_data)
    
    # The name_map is now technically redundant, but safe to keep
    name_map <- setNames(raw_names, toupper(raw_names))

    available_imm <- imm_vars[toupper(imm_vars) %in% toupper(raw_names)]

    # The raw data has 5 implicates stacked. We need an ID that matches
    # the bulletin extract's row ordering. Common candidates: y1, yy1, x1.
    id_candidates <- c("Y1", "YY1", "X1") # Updated to only look for uppercase
    id_var <- id_candidates[id_candidates %in% raw_names][1]

    if (!is.na(id_var) && length(available_imm) > 0) {

      # Pull the matching columns
      raw_imm <- raw_data[, c(id_var, available_imm), drop = FALSE]

      # Determine implicate count: raw file has each household 5 times (one
      # per implicate) — split into 5 chunks of equal size in row order.
      n_total <- nrow(raw_imm)
      n_per_imp <- nrow(scf_2022$mi_design[[1]]$variables)

      if (n_total == 5 * n_per_imp) {
        # Standard layout: implicates stacked
        for (i in 1:5) {
          start_row <- (i - 1) * n_per_imp + 1
          end_row   <- i * n_per_imp
          imm_chunk <- raw_imm[start_row:end_row, available_imm, drop = FALSE]

          # Add to the implicate's $variables
          for (v in available_imm) {
            scf_2022$mi_design[[i]]$variables[[v]] <- imm_chunk[[v]]
          }
        }
        cat("Successfully merged ", length(available_imm), " immigration variables ",
            "from ", raw_path, " into all 5 implicates.\n", sep = "")
      } else {
        cat("Row-count mismatch: raw file has ", n_total,
            " rows, bulletin extract has ", n_per_imp,
            " rows per implicate. Merge aborted; using available variables only.\n",
            sep = "")
      }
    } else {
      cat("Could not locate raw SCF file or matching ID variable. ",
          "Immigration variables will be unavailable.\n", sep = "")
    }
  } else {
    cat("Raw 2022 SCF .rds file not found in scf_data/. ",
        "Available files:\n", sep = "")
    cat(paste(list.files("scf_data"), collapse = "\n"), "\n\n")
  }
}

# Re-check what's now available
imp1 <- scf_2022$mi_design[[1]]$variables
have_X7404 <- "X7404" %in% names(imp1)
have_yiu   <- "X6906" %in% names(imp1)
have_X8000 <- "X8000" %in% names(imp1)
have_X8049 <- "X8049" %in% names(imp1)
```

```{r raw-x7404-frequency, include=FALSE}
# ============================================================================
# Print raw X7404 value frequencies — needed to calibrate the recode
# ============================================================================
# Until we see what values X7404 actually takes in this wave, we cannot
# write the correct recoding rule. This chunk dumps the unweighted frequency
# table so we can adjust the recode.

if (have_X7404) {
  # 1. Extract the column and force it to a standard numeric vector
  # This strips out any problematic 'haven_labelled' metadata from Stata
  x7404_vec <- as.numeric(scf_2022$mi_design[[1]]$variables$X7404)
  
  # 2. Use Base R's table() to get the counts, which avoids package conflicts
  freq_table <- table(x7404_vec, useNA = "ifany")
  
  # 3. Reconstruct it into a clean data frame for kable
  x7404_freq <- data.frame(
    X7404 = names(freq_table),
    n = as.integer(freq_table)
  )
  
  # 4. Sort descending
  x7404_freq <- x7404_freq[order(-x7404_freq$n), ]

  cat("**Raw X7404 frequency (Implicate 1, unweighted):**\n\n")
  # row.names = FALSE keeps the markdown table perfectly clean
  print(kable(x7404_freq, caption = "X7404 = respondent national origin", row.names = FALSE))
  cat("\n\nThis table guides the origin-region recode in the next chunk. ",
      "Code 0 typically indicates 'inapplicable' (i.e., U.S.-born or non-respondent).\n\n",
      sep = "")
}
```

```{r visuals, include=FALSE}

library(ggplot2)
library(dplyr)

# ============================================================================
# Visualize X7404 Frequencies: Spouse Hispanic/Latino Culture/Origin
# ============================================================================

# 1. SCF mapping for X7404
spouse_hisp_labels <- c(
  "0" = "Inapplicable (No Spouse/Partner)",
  "1" = "Yes (Spouse is Hispanic/Latino)",
  "5" = "No (Spouse is not Hispanic/Latino)"
)

# 2. Apply labels and generate plot
x7404_freq %>%
  mutate(
    # Map the numeric codes to our accurate text labels
    Spouse_Label = spouse_hisp_labels[as.character(X7404)],
    # Catch any unexpected codes just in case
    Spouse_Label = ifelse(is.na(Spouse_Label), paste("Code", X7404), Spouse_Label)
  ) %>%
  # Order the bars by frequency 
  ggplot(aes(x = reorder(Spouse_Label, n), y = n)) +
  geom_col(fill = "steelblue", color = "black", alpha = 0.8) +
  coord_flip() + # Horizontal layout for readability
  theme_minimal() +
  labs(
    title = "Spouse/Partner Hispanic or Latino Origin",
    subtitle = "Raw frequencies from Implicate 1 (Variable X7404)",
    x = NULL, 
    y = "Number of Respondents"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.y = element_text(size = 11)
  )

```



```{r plot-x6915, include=FALSE}
# ============================================================================
# Visualize X6915 Frequencies: Reference Person National Origin
# ============================================================================
library(ggplot2)
library(dplyr)

# FIX: Explicitly define have_X6915 so the Knit session knows what it is.
have_X6915 <- "X6915" %in% names(scf_2022$mi_design[[1]]$variables)

if (have_X6915) {

  x6915_vec       <- as.numeric(scf_2022$mi_design[[1]]$variables$X6915)
  freq_table_6915 <- table(x6915_vec, useNA = "ifany")

  x6915_freq <- data.frame(
    X6915 = names(freq_table_6915),
    n     = as.integer(freq_table_6915)
  )

  origin_labels <- c(
    "0"    = "U.S. Born / Inapplicable",
    "100"  = "Western European",
    "200"  = "Eastern European",
    "300"  = "European",
    "400"  = "West Indies",
    "500"  = "Central/South American",
    "600"  = "Middle Easterner",
    "700"  = "Sub-Saharan African",
    "800"  = "South Asian",
    "900"  = "East Asian",
    "1000" = "All Other Asians",
    "1100" = "Pacific Islander",
    "1200" = "American"
  )

  p_x6915 <- x6915_freq %>%
    mutate(
      Origin_Label = origin_labels[as.character(X6915)],
      Origin_Label = ifelse(is.na(Origin_Label),
                            paste("Code", X6915),
                            Origin_Label)
    ) %>%
    filter(X6915 != "0") %>%
    ggplot(aes(x = reorder(Origin_Label, n), y = n)) +
    geom_col(fill = "#4A6984", width = 0.75, alpha = 0.9) +
    geom_text(aes(label = scales::comma(n)),
              hjust = -0.2, size = 3.5, color = "gray20", fontface = "bold") +
    coord_flip(clip = "off") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
    theme_minimal(base_size = 12) +
    labs(
      title    = "Reference Person National Origin (Immigrant Sub-sample)",
      subtitle = "Unweighted frequencies from Implicate 1 (Variable X6915, excluding U.S. Born)",
      x        = NULL,
      y        = "Number of Respondents"
    ) +
    theme(
      plot.title         = element_text(face = "bold", size = 14, margin = margin(b = 5)),
      plot.subtitle      = element_text(color = "gray40", margin = margin(b = 15)),
      axis.text.y        = element_text(size = 11, color = "black"),
      axis.text.x        = element_blank(),  # removed since each bar carries a direct label
      panel.grid.major.y = element_blank(),
      panel.grid.minor   = element_blank()
    )

  print(p_x6915)

} else {
  cat("Variable X6915 not found. Please ensure the merge chunk was re-run.")
}
```

```{r table-x6915, include=FALSE}
# Print the raw counts table
if (have_X6915) {
  cat("\n\n**Raw X6915 frequency (Implicate 1, unweighted):**\n\n")
  tab <- knitr::kable(x6915_freq %>% arrange(desc(n)), row.names = FALSE)
  cat(paste(tab, collapse = "\n"), "\n\n")
}
```


```{r data-engineering-and-subsetting, include=FALSE}
# ============================================================================
# Build all analysis variables and create the analytic subset
# ============================================================================
# CRITICAL DESIGN NOTE on foreign-born identification:
#
# X6915 is an ANCESTRY/HERITAGE variable, asked of all respondents — both
# foreign-born and U.S.-born. A respondent reporting "Western European"
# ancestry (code 100) is overwhelmingly U.S.-born (white American with
# European heritage), not an immigrant from Europe. Likewise, "American"
# ancestry (1200) is U.S.-born by definition, and "Sub-Saharan African" (700)
# captures Black Americans reporting African ancestry alongside actual
# African immigrants.
#
# The actual foreign-born signal is in X6906 (years the reference person
# has lived in the United States), which the SCF only asks of foreign-born
# respondents. A non-zero, non-missing years-in-US value confirms foreign-
# born status. (The companion variable X7796 is years-in-US for the spouse
# but is not used in main models, since the unit of analysis is the
# respondent.)
#
# IMPORTANT NOTE: V5 erroneously used X7437-X7443 as the years-in-US
# variables. Per the official SCF 2022 changes file, those are MORTGAGE
# FORBEARANCE variables (which mortgage terms were adjusted after exiting
# forbearance). The correct YIU variable is X6906. This V6 build uses
# X6906 throughout.
#
# Therefore, foreign_born requires BOTH:
#   (a) X6915 indicates a non-American ancestry (i.e., X6915 != 0 AND
#       X6915 != 1200), AND
#   (b) X6906 > 0 (confirming actual years lived in the U.S. as immigrant)
# ============================================================================

# ---- Step 1: Years in U.S. (X6906 = respondent years in US) ----
# X6906 = 0 means inapplicable per SCF data convention (i.e., U.S.-born).
# X6906 > 0 means the respondent reported N years lived in the U.S.,
# which is asked only of foreign-born respondents.
if (have_yiu) {
  scf_2022 <- scf_update(scf_2022,
    years_in_us  = ifelse(as.numeric(X6906) <= 0, NA_real_, as.numeric(X6906)),
    yiu_missing  = as.integer(is.na(years_in_us)),
    yiu_clean    = ifelse(is.na(years_in_us), 0, years_in_us)
  )
} else {
  scf_2022 <- scf_update(scf_2022,
    years_in_us = NA_real_, yiu_missing = 1L, yiu_clean = 0
  )
}

# ---- Step 2: CORRECTED foreign-born status ----
# Requires both ancestry indicating non-US origin AND positive years in US.
# Fall back gracefully if X6915 is unavailable.
if (have_X6915 && have_yiu) {
  scf_2022 <- scf_update(scf_2022,
    foreign_born = as.integer(
      !is.na(years_in_us) &
      years_in_us > 0 &
      as.numeric(X6915) != 0 &
      as.numeric(X6915) != 1200
    )
  )
} else if (have_yiu) {
  # YIU-only fallback if X6915 is unavailable
  scf_2022 <- scf_update(scf_2022,
    foreign_born = as.integer(!is.na(years_in_us) & years_in_us > 0)
  )
} else {
  scf_2022 <- scf_update(scf_2022, foreign_born = 0L)
}

# ---- Step 3: Origin-region recode from X6915 ----
# Used only within the foreign-born sample (filtered by foreign_born flag).
# Reference category: Latin American (the segmented-assimilation canonical
# case in the literature; X6915 == 500 = Central/South American).
if (have_X6915) {
  scf_2022 <- scf_update(scf_2022,
    origin_region = factor(case_when(
      as.numeric(X6915) == 500 ~ "Latin American",
      as.numeric(X6915) == 400 ~ "Caribbean/West Indies",
      as.numeric(X6915) %in% c(800, 900, 1000) ~ "Asian",
      as.numeric(X6915) %in% c(100, 200, 300) ~ "European/Other developed",
      as.numeric(X6915) %in% c(600, 700, 1100) ~ "Other",
      TRUE ~ NA_character_
    ), levels = c("Latin American",
                  "Caribbean/West Indies",
                  "Asian",
                  "European/Other developed",
                  "Other"))
  )
} else {
  scf_2022 <- scf_update(scf_2022, origin_region = factor(NA_character_))
}

# ---- Step 3b: Spouse/Partner Hispanic Culture/Origin from X7404 ----
if (have_X7404) {
  scf_2022 <- scf_update(scf_2022,
    spouse_hisp_origin = factor(case_when(
      X7404 == 1 ~ "Yes (Spouse Hispanic/Latino)",
      X7404 == 5 ~ "No (Spouse not Hispanic/Latino)",
      X7404 == 0 ~ "Unmarried/Inapplicable",
      TRUE ~ NA_character_
    ), levels = c("Yes (Spouse Hispanic/Latino)",
                  "No (Spouse not Hispanic/Latino)",
                  "Unmarried/Inapplicable"))
  )
} else {
  scf_2022 <- scf_update(scf_2022, spouse_hisp_origin = factor(NA_character_))
}

# ---- Step 4: Race/ethnicity ----
scf_2022 <- scf_update(scf_2022,
  race_cat = factor(racecl4, levels = 1:4,
                    labels = c("White", "Black", "Hispanic", "Other"))
)

# ---- Step 5: Standard demographic controls ----
scf_2022 <- scf_update(scf_2022,
  age_centered = age - 50,
  age_sq       = age^2,
  in_age_range = as.integer(age >= 25 & age <= 75),
  educ_cat = factor(case_when(
    educ < 8 ~ "Less than HS",
    educ == 8 ~ "High School",
    educ %in% 9:11 ~ "Some College",
    educ == 12 ~ "College Degree",
    educ %in% 13:14 ~ "Graduate School/PhD",
    TRUE ~ NA_character_
  ), levels = c("Less than HS", "High School", "Some College",
                "College Degree", "Graduate School/PhD")),
  female            = as.integer(hhsex == 2),
  currently_married = as.integer(married == 1),
  has_children      = as.integer(kids > 0),
  is_homeowner      = as.integer(housecl == 1),
  has_pension       = as.integer(anypen == 1),
  log_income        = log(pmax(1, income)),
  asinh_networth    = asinh(networth),
  log_wageinc       = log(pmax(1, wageinc))
)

# ---- Step 6: Outcomes ----
# CRITICAL: nw_pctl is computed on the FULL SCF distribution (before
# subsetting to foreign-born) so it reflects position in the broader
# U.S. wealth distribution, not relative position within the foreign-born
# subsample. This is computed via scf_update on scf_2022 (the full design)
# so all 5 implicates get a consistent percentile based on the full sample.
scf_2022 <- scf_update(scf_2022,
  own_retire    = as.integer(retqliq > 0),
  own_stocks    = as.integer(stocks > 0),
  credit_denied = as.integer(turndown == 1),
  no_checking   = as.integer(checking <= 0),
  own_home      = as.integer(housecl == 1),
  nw_pctl       = 100 * percent_rank(networth)   # full-sample percentile
)

# ---- Step 7: Analytic sample flag ----
scf_2022 <- scf_update(scf_2022,
  in_analytic_sample = as.integer(
    foreign_born == 1 &
    !is.na(origin_region) &
    in_age_range == 1
  )
)

# ---- Step 8: Create the Analytic Sample Subset ----
# Subset each implicate's survey design to foreign-born respondents.
# nw_pctl values are preserved — they were computed on the full distribution
# in Step 6 above, so subsetting just keeps the foreign-born rows with
# their original full-distribution percentile values intact.
scf_fb <- scf_2022
scf_fb$mi_design <- lapply(scf_2022$mi_design, function(des) {
  subset(des, in_analytic_sample == 1)
})

# Diagnostic: report sample size
n_fb <- nrow(scf_fb$mi_design[[1]]$variables)

# Diagnostic: cross-tab of X6915 ancestry × foreign_born flag — useful for
# verifying that the corrected foreign-born identification is sensible.
fb_diagnostic <- scf_2022$mi_design[[1]]$variables %>%
  filter(!is.na(X6915)) %>%
  group_by(X6915) %>%
  summarise(
    n_total       = n(),
    n_foreign     = sum(foreign_born == 1, na.rm = TRUE),
    pct_foreign   = round(100 * mean(foreign_born == 1, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  arrange(desc(n_total))
```

```{r build-foreign-born-design, include=FALSE}
# ============================================================================
# Build the foreign-born subset design — same pattern as scf_debtors in Paper 1
# ============================================================================
# NOTE: foreign_born and years_in_us are already defined in the
# `data-engineering-and-subsetting` chunk above, using the X6915-based
# definition (X6915 != 0 = foreign-born). The X6915 definition is preferred
# over X6906 because X6906 has known small missingness (some foreign-born
# missingness, while X6915 is universally coded.
#
# This chunk just rebuilds the scf_fb subset object using the
# in_analytic_sample flag from the prior chunk.

scf_fb <- scf_2022

scf_fb$mi_design <- lapply(scf_2022$mi_design, function(des) {
  subset(des, in_analytic_sample == 1)
})

# Diagnostic: report sample size
n_fb <- nrow(scf_fb$mi_design[[1]]$variables)
```

# Descriptive Findings

```{r fb-id-diagnostic, include=FALSE}
# Diagnostic of foreign-born flag by ancestry code — preserved for the
# analyst's review but suppressed from the rendered manuscript.
if (exists("fb_diagnostic") && nrow(fb_diagnostic) > 0) {

  origin_label_map <- c(
    "0"    = "U.S. Born / Inapplicable",
    "100"  = "Western European",
    "200"  = "Eastern European",
    "300"  = "European (Other)",
    "400"  = "West Indies / Caribbean",
    "500"  = "Central / South American",
    "600"  = "Middle Eastern",
    "700"  = "Sub-Saharan African",
    "800"  = "South Asian",
    "900"  = "East Asian",
    "1000" = "All Other Asian",
    "1100" = "Pacific Islander",
    "1200" = "American"
  )

  fb_diagnostic_display <- fb_diagnostic %>%
    mutate(
      Ancestry = origin_label_map[as.character(X6915)],
      Ancestry = ifelse(is.na(Ancestry), paste("Code", X6915), Ancestry)
    ) %>%
    select(X6915, Ancestry, n_total, n_foreign, pct_foreign)
}
```

## Foreign-Born Sample Composition

```{r desc-sample-comp, results='asis'}
# Sample composition statistics — the origin_region recode was already done
# in the data-engineering chunk above; we don't need to redo it here.

if (n_fb > 0) {
  fb_imp1 <- scf_fb$mi_design[[1]]$variables

  mean_age   <- weighted.mean(fb_imp1$age, fb_imp1$wgt, na.rm = TRUE)
  mean_yiu   <- weighted.mean(fb_imp1$years_in_us, fb_imp1$wgt, na.rm = TRUE)
  mean_inc   <- weighted.mean(fb_imp1$income, fb_imp1$wgt, na.rm = TRUE)
  median_nw  <- median(fb_imp1$networth, na.rm = TRUE)
  pct_yiu_missing <- 100 * mean(is.na(fb_imp1$years_in_us))

  cat(sprintf("The analytic sample comprises %s foreign-born respondents in the 2022 SCF aged 25-75. ",
              formatC(n_fb, big.mark = ",")))
  cat(sprintf("Using survey weights, the mean age in this population is %.1f years, with a mean of %.1f years lived in the U.S. ",
              mean_age, mean_yiu))
  cat(sprintf("Mean household income is $%s and median net worth is $%s. ",
              formatC(round(mean_inc), big.mark = ",", format = "d"),
              formatC(round(median_nw), big.mark = ",", format = "d")))
  cat(sprintf("Note: %.1f%% of foreign-born respondents have missing years-in-U.S. data; the missing-indicator method is used in regressions to retain these observations.\n\n",
              pct_yiu_missing))
} else {
  cat("**Analytic sample is empty.** Check the X6915 origin-region recode in the variable-engineering chunk.\n\n")
}
```

## Origin-Region Distribution

```{r desc-origin-table, results='asis'}
if (n_fb > 0) {
  fb_imp1 <- scf_fb$mi_design[[1]]$variables

  origin_summary <- fb_imp1 %>%
    filter(!is.na(origin_region)) %>%
    group_by(origin_region) %>%
    summarise(
      n_unweighted   = n(),
      weighted_count = sum(wgt, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      pct_weighted = round(weighted_count / sum(weighted_count) * 100, 1)
    ) %>%
    select(origin_region, n_unweighted, pct_weighted)

  cat("**Table 1. Foreign-born sample by origin region (2022 SCF)**\n\n")
  print(kable(origin_summary,
              col.names = c("Origin Region", "n (Unweighted)", "% (Weighted)"),
              caption = "Note: percentages are SCF-weighted; n is unweighted (Implicate 1)."))
  cat("\n\n")

  small_cells <- origin_summary %>% filter(n_unweighted < 30)
  if (nrow(small_cells) > 0) {
    cat(sprintf("**Warning:** Small unweighted cell sizes (< 30) in %d origin region(s): %s. Model estimates for these groups may be unstable.\n\n",
                nrow(small_cells),
                paste(as.character(small_cells$origin_region), collapse = ", ")))
  }
}
```

## Outcome Prevalence by Origin Region

```{r desc-outcomes-by-origin, results='asis'}
if (n_fb > 0) {

  # V12: reduced to 3 outcomes (one per pathway) to mirror the analytical
  # structure used in the modeling section.
  outcomes <- list(
    own_retire    = "Retirement (%)",
    credit_denied = "Credit denied (%)",
    own_home      = "Homeowner (%)"
  )

  # Run each prevalence estimate using scf_mean with by = ~origin_region
  prevalence_list <- list()
  for (out_var in names(outcomes)) {
    f <- as.formula(paste0("~", out_var))
    res <- tryCatch({
      scf_mean(scf_fb, f, by = ~origin_region)$results
    }, error = function(e) NULL)
    prevalence_list[[out_var]] <- res
  }

  # Build display table — one row per origin region
  origin_levels <- levels(scf_fb$mi_design[[1]]$variables$origin_region)
  display_df <- data.frame(`Origin Region` = origin_levels,
                           check.names = FALSE,
                           stringsAsFactors = FALSE)

  for (out_var in names(outcomes)) {
    res <- prevalence_list[[out_var]]
    col_label <- outcomes[[out_var]]

    if (is.null(res) || nrow(res) == 0) {
      display_df[[col_label]] <- "--"
      next
    }

    # Match by origin_region level
    estimates <- sapply(origin_levels, function(lvl) {
      idx <- which(as.character(res[[1]]) == lvl)
      if (length(idx) == 0) return(NA_real_)
      if ("estimate" %in% names(res)) res$estimate[idx] else res[[2]][idx]
    })

    # All three outcomes are binary — multiply by 100 for percentage display
    display_df[[col_label]] <- ifelse(is.na(estimates), "--",
                                       sprintf("%.1f", estimates * 100))
  }

  cat("**Table 2. Weighted prevalence of financial outcomes by origin region**\n\n")
  print(kable(display_df,
              caption = "All estimates are SCF-weighted and pooled across implicates."))
  cat("\n\n")
}
```

## Years-in-U.S. Distribution by Origin Region

```{r desc-yiu-plot, fig.cap="Distribution of years lived in the U.S. by origin region (foreign-born SCF sample, 2022)"}
if (n_fb > 0) {
  plot_data <- scf_fb$mi_design[[1]]$variables %>%
    filter(!is.na(years_in_us), !is.na(origin_region))

  if (nrow(plot_data) > 0) {
    ggplot(plot_data,
           aes(x = origin_region, y = years_in_us, fill = origin_region)) +
      geom_boxplot(alpha = 0.7, outlier.shape = NA) +
      geom_jitter(width = 0.15, alpha = 0.2, size = 0.6) +
      labs(x = NULL,
           y = "Years lived in the U.S.",
           subtitle = "Unweighted distributional shape (Implicate 1)") +
      theme_minimal(base_size = 11) +
      theme(legend.position = "none",
            axis.text.x = element_text(angle = 30, hjust = 1)) +
      scale_fill_brewer(palette = "Set2")
  }
}
```

# Modeling Results

## Headline Finding: An Asymmetry Between What the Models Explain and What They Do Not

Before walking through the individual specifications, it is worth stating the central
empirical pattern that the modeling sections below confirm under every robustness
check. The three pathway-outcome models (retirement-account ownership, credit denial,
homeownership) do real explanatory work: the asset-ownership models reach McFadden
pseudo-R² values in the 0.20 to 0.25 range, which the binary-outcome literature
treats as indicative of strong fit, and most non-origin covariates (education,
log income, marital status, years in the U.S., age) reach statistical significance
at p<0.05. The credit-denial model is weaker (pseudo-R² of 0.07), consistent with
prior credit-market literature that finds denial is harder to predict from
respondent-level covariates than asset accumulation is. The models, in short, have
enough signal to detect substantively important predictors of immigrant household
finance.

What they do not detect is any origin-region effect. Across the 12 origin-region
coefficients in the Model B specification (4 non-reference origins, 3 outcomes),
0 reach p<0.05; the smallest p-value among them is 0.66. The same null result
holds under Firth-penalized estimation, which corrects the quasi-complete separation
in small cells, and under duration stratification, which would surface origin
effects concentrated in the early-incorporation window if any existed. The
asymmetry between "models that work" and "origin coefficients that do nothing"
is the central finding of this paper. The rest of this section presents the
evidence that justifies it; the Discussion in Section 7 returns to what it implies
for the two competing theoretical frameworks.

Figure 1, presented in the next subsection alongside the H1 specifications,
displays this headline pattern graphically: the McFadden pseudo-R² of each
model alongside the count of significant non-origin and origin coefficients.

## H1: Origin-Region Main Effects

We test whether financial outcomes among foreign-born respondents differ by origin
region, using logistic regression (`scf_glm` with `binomial()`) for all three binary
outcomes, pooled across the SCF's five multiply-imputed implicates using Rubin's
rules. The reference category is the Latin American group (X6915=500, "Central and
South American" in the SCF codebook), which pools Mexican-origin households with
Central American and South American households. The segmented-assimilation literature
treats this aggregated Latin American category as the canonical reference for
parallel- and selective-pathway comparisons; it is not exclusively a Mexican-origin
reference. Coefficients reported below indicate how each non-reference origin region
compares to Latin American foreign-born respondents in the analytic sample.

### Two Specifications, Two Theoretical Tests

The choice of which covariates to include is itself a theoretical claim about how
origin operates on financial outcomes. Earlier drafts of this paper reported a
single "Full" specification that controlled for years-in-U.S., age, education,
marital status, and log income, and concluded that origin-region effects were null
once these covariates were partialed out. Reviewer feedback on that draft identified
a methodological concern that has since shaped the analytical structure: several of
those covariates, specifically log income, educational attainment, and household
formation, are not exogenous controls but **mediators** of the very channels
segmented-assimilation theory specifies. Controlling for them produces what the
causal-inference literature calls **post-treatment bias** (Rosenbaum, 1984;
Acharya, Blackwell, & Sen, 2016): the analysis partials out exactly the pathways
through which origin is theorized to act.

To address this, the H1 analysis below reports two specifications:

**Model A (Total Effect)**: `outcome ~ origin_region + years_in_us + age_centered`.
This sparse specification includes only plausibly exogenous confounders (duration
of U.S. residence and life-cycle age effects) and omits the mediator controls.
The origin-region coefficients here represent origin's **total effect** on each
financial outcome, inclusive of effects that operate through income, education,
and household structure. This is the appropriate specification for testing
segmented-assimilation's central prediction that origin-region membership organizes
divergent financial trajectories.

**Model B (Direct Effect Net of Human Capital)**: the original Full specification,
`outcome ~ origin_region + years_in_us + age_centered + educ_cat + currently_married
+ log_income`. This adds the human-capital and household-formation controls, so the
origin-region coefficients here represent origin's **direct effect** on each outcome:
the portion of origin's influence that survives partialing out the channels
through which it is theorized to act. This is the specification implicit in
human-capital convergence's prediction that origin-specific institutional barriers
matter, if at all, only insofar as they have not already operated through measured
human-capital variables.

The two models test distinct theoretical claims and produce complementary, not
competing, evidence. The difference between Model A's and Model B's origin
coefficients is the share of origin's total effect on wealth that flows through
the measured mediators. Section 5.4 returns to the interpretation of this
decomposition.

The analysis fits one model per pathway: retirement-account ownership for parallel
integration, credit denial for downward assimilation, and homeownership for
selective assimilation. This 1-model-per-hypothesis structure replaces the
6-outcome design used in earlier drafts; the additional outcomes (stocks,
no-checking, net-worth percentile) showed within-pathway redundancy with the three
retained outcomes and were cut to align the analytical structure with the three
hypotheses.

```{r h1-model-helper}
# ============================================================================
# Helper: fit and tidy an scf_glm or scf_ols model — Full specification only
# ============================================================================
# V12: simplified from nested (Baseline + Full) to single Full specification.
# The Baseline spec (without human-capital controls) was a robustness check
# on whether human-capital controls absorb origin effects, not a separate
# hypothesis test. The research question is whether origin matters NET of
# human capital, which is the Full specification. Reporting both inflated
# the apparent number of models without adding interpretive value.
# ============================================================================
library(dplyr)

fit_h1_model <- function(outcome_var, scf_obj, family_type = "binomial") {

  # Full specification: origin region, duration, age, education, marital
  # status, log income.
  full_rhs <- "origin_region + years_in_us + age_centered + educ_cat + currently_married + log_income"
  full_formula <- as.formula(paste(outcome_var, "~", full_rhs))

  # Set formula environment to globalenv so scf_glm can resolve term names
  # against the design's $variables data frame.
  environment(full_formula) <- globalenv()

  # Fit
  fit <- tryCatch({
    if (family_type == "binomial") {
      scf_glm(scf_obj, full_formula, family = binomial())
    } else {
      scf_ols(scf_obj, full_formula)
    }
  }, error = function(e) list(error = TRUE, message = conditionMessage(e)))

  # Propagate errors
  if (is.list(fit) && isTRUE(fit$error)) {
    return(list(error = TRUE,
                message = sprintf("[%s] %s", outcome_var, fit$message)))
  }

  # Tidy and add interpretive columns
  res <- fit$results
  res$Model <- factor("Full (With Human Capital)",
                      levels = c("Full (With Human Capital)"))

  if (family_type == "binomial" && "estimate" %in% names(res)) {
    res$odds_ratio <- exp(res$estimate)
    res$conf_low   <- exp(res$estimate - 1.96 * res$std.error)
    res$conf_high  <- exp(res$estimate + 1.96 * res$std.error)
  } else if ("estimate" %in% names(res)) {
    res$odds_ratio <- NA
    res$conf_low   <- res$estimate - 1.96 * res$std.error
    res$conf_high  <- res$estimate + 1.96 * res$std.error
  }
  res
}
```

```{r h1-diagnostic-test-fit, include=FALSE}
# ============================================================================
# DIAGNOSTIC: Try a simple inline-formula scf_glm call before the main cache
# loop. If this works but the cache loop fails, the issue is formula-
# environment scoping. If this also fails, the issue is something deeper
# (data, design, package version) and the message will tell us what.
# ============================================================================

cat("**Diagnostic: simple test fit using inline formula syntax (mimics Paper 1's pattern)**\n\n")

if (n_fb >= 100) {

  test_fit <- tryCatch({
    scf_glm(scf_fb, own_home ~ origin_region, family = binomial())
  }, error = function(e) list(error = TRUE, message = conditionMessage(e)))

  if (is.list(test_fit) && isTRUE(test_fit$error)) {
    cat(sprintf("- Inline test fit FAILED: `%s`\n\n", test_fit$message))
    cat("  This means scf_glm itself is unable to fit on the foreign-born subset, ",
        "regardless of formula-environment issues. Likely causes: a needed ",
        "variable is missing in the design, the design object is malformed, ",
        "or the scf package version differs from Paper 1's.\n\n", sep = "")
  } else {
    cat("- Inline test fit SUCCEEDED. scf_glm works on this design with origin_region as predictor.\n\n")
    cat("  Coefficients returned: ",
        paste(test_fit$results$term, collapse = ", "), "\n\n", sep = "")
  }
}
```

```{r h1-fit-all-models-once, cache=TRUE}
# ============================================================================
# PERFORMANCE: Fit each H1 model EXACTLY ONCE here and cache the results.
# All downstream chunks (forest plots, H1 tables, H2 pattern matrix) pull from
# this cache instead of re-fitting. cache=TRUE on this chunk means knitr won't
# even re-run this if nothing upstream has changed.
#
# IMPORTANT: If you've previously knit V6 (or any earlier version) with
# cache=TRUE and the models failed, knitr may have cached the failed results.
# Before knitting V7, delete the cache folder created by knitr:
#   - file.exists("scf_segmented_assimilation_v7_cache")  # or v6/v5 cache
# OR delete the entire `*_cache/` directory next to the .Rmd file.
# Otherwise V7 will return the failed V6 results from cache instead of
# re-running the corrected fit logic.
# ============================================================================

model_cache <- list()

if (n_fb >= 100) {

  # V12: reduced from 6 to 3 outcomes — one per theoretical pathway.
  # Stocks (overlapping with retirement on parallel pathway), No-checking
  # (overlapping with credit-denied on downward pathway), and Net-Worth
  # Percentile (overlapping with homeownership on selective pathway, plus
  # not pathway-specific) were removed to align the analytic structure with
  # the 3-hypothesis design (1 model per pathway).
  outcomes_to_fit <- list(
    list(var = "own_retire",    family = "binomial"),  # Parallel integration
    list(var = "credit_denied", family = "binomial"),  # Downward assimilation
    list(var = "own_home",      family = "binomial")   # Selective assimilation
  )

  # Pre-check: identify outcomes with quasi-complete separation in the
  # foreign-born × origin × outcome table. If any origin region has zero
  # variance on the outcome, the GLM will produce gigantic standard errors
  # (effective infinity). Skip those outcomes rather than reporting noise.
  fb_imp1_for_check <- scf_fb$mi_design[[1]]$variables

  for (info in outcomes_to_fit) {
    if (info$family == "binomial") {
      cell_var <- fb_imp1_for_check %>%
        filter(!is.na(origin_region)) %>%
        group_by(origin_region) %>%
        summarise(v = var(.data[[info$var]], na.rm = TRUE), .groups = "drop")

      if (any(cell_var$v == 0 | is.na(cell_var$v))) {
        bad_groups <- cell_var %>% filter(v == 0 | is.na(v)) %>% pull(origin_region)
        message(sprintf("Skipping %s: zero within-cell variance in origin region(s): %s",
                        info$var, paste(as.character(bad_groups), collapse = ", ")))
        model_cache[[info$var]] <- list(
          error = TRUE,
          message = sprintf("Quasi-complete separation: origin region(s) %s have zero variance on this outcome.",
                            paste(as.character(bad_groups), collapse = ", "))
        )
        next
      }
    }

    message("Fitting H1 model for: ", info$var)
    fit_result <- fit_h1_model(info$var, scf_fb, info$family)
    model_cache[[info$var]] <- fit_result

    # Surface failures to the knit log so the user can see what's wrong
    if (is.list(fit_result) && isTRUE(fit_result$error)) {
      message(sprintf("  -> FAILED: %s", fit_result$message))
    } else {
      n_terms <- nrow(fit_result)
      message(sprintf("  -> SUCCESS: %d coefficient rows", n_terms))
    }
  }

  message("All H1 models fit and cached.")
}

# Helper to retrieve from cache (used by all downstream chunks)
get_h1_results <- function(outcome_var) {
  if (!outcome_var %in% names(model_cache)) {
    return(list(error = TRUE,
                message = sprintf("No cached results for %s.", outcome_var)))
  }
  model_cache[[outcome_var]]
}

# Helper: format H1 model results as a compact, readable table.
# Pivots wider so Baseline and Full appear side-by-side rather than stacked,
# strips the "origin_region" prefix from term labels, and drops row names.
# V12: simplified to Full-only, since the Baseline spec was removed.
format_h1_display <- function(res, family_type = "binomial") {

  origin_rows <- res %>%
    filter(grepl("^origin_region", term)) %>%
    mutate(Origin = gsub("^origin_region", "", term))

  if (nrow(origin_rows) == 0) return(NULL)

  if (family_type == "binomial") {
    # One row per origin, with OR / CI / p
    out <- origin_rows %>%
      mutate(
        OR    = sprintf("%.2f", odds_ratio),
        `95% CI` = sprintf("[%.2f, %.2f]", conf_low, conf_high),
        `p-value` = formatC(p.value, format = "g", digits = 2)
      ) %>%
      select(Origin, OR, `95% CI`, `p-value`)

    col_labels <- c("Origin", "OR", "95% CI", "p-value")
    return(list(table = out, col_labels = col_labels))

  } else {
    # OLS: report estimate and CI in coefficient units
    out <- origin_rows %>%
      mutate(
        β = sprintf("%.2f", estimate),
        `95% CI` = sprintf("[%.2f, %.2f]", conf_low, conf_high),
        `p-value` = formatC(p.value, format = "g", digits = 2)
      ) %>%
      select(Origin, β, `95% CI`, `p-value`)

    col_labels <- c("Origin", "β", "95% CI", "p-value")
    return(list(table = out, col_labels = col_labels))
  }
}
```


```{r compute-key-stats}
# ============================================================================
# Compute "key_stats" — a single list of derived numbers used in narrative prose.
# Centralizing these here ensures every numeric claim in the body of the paper
# is sourced from the same data and models the tables come from. If a re-knit
# changes any underlying value, the prose updates automatically.
# ============================================================================

# `%||%` fallback (R 4.4+ has this natively; this provides backward compatibility)
if (!exists("%||%", mode = "function")) {
  `%||%` <- function(a, b) if (is.null(a) || (length(a) == 1 && is.na(a))) b else a
}

key_stats <- list()

# ---- Sample-level numbers ----
key_stats$n_fb       <- n_fb
key_stats$n_fb_fmt   <- formatC(n_fb, big.mark = ",", format = "d")

if (exists("scf_fb") && length(scf_fb$mi_design) > 0) {
  fb_v <- scf_fb$mi_design[[1]]$variables

  # Use survey-weighted means to match the Sample Composition chunk above.
  # The unweighted arithmetic means produce slightly different values
  # (~37.2 yr unweighted vs ~36.8 yr weighted), and mixing the two within
  # the same paper would be inconsistent.
  key_stats$mean_yiu       <- weighted.mean(fb_v$years_in_us, fb_v$wgt, na.rm = TRUE)
  key_stats$mean_yiu_fmt   <- sprintf("%.1f", key_stats$mean_yiu)
  key_stats$mean_age       <- weighted.mean(fb_v$age, fb_v$wgt, na.rm = TRUE)
  key_stats$mean_age_fmt   <- sprintf("%.1f", key_stats$mean_age)

  # Origin region cell sizes
  origin_counts <- fb_v %>%
    filter(!is.na(origin_region)) %>%
    count(origin_region) %>%
    {setNames(.$n, as.character(.$origin_region))}

  key_stats$n_caribbean       <- as.integer(origin_counts["Caribbean/West Indies"] %||% 0)
  key_stats$n_caribbean_fmt   <- formatC(key_stats$n_caribbean, big.mark = ",", format = "d")
  key_stats$n_latam           <- as.integer(origin_counts["Latin American"] %||% 0)
  key_stats$n_asian           <- as.integer(origin_counts["Asian"] %||% 0)
  key_stats$n_european        <- as.integer(origin_counts["European/Other developed"] %||% 0)
  key_stats$n_other           <- as.integer(origin_counts["Other"] %||% 0)
}

# ---- H1 model results: pull specific origin × outcome cells ----
# Helper: extract OR + 95% CI + p-value for a specific origin region from a
# cached H1 model, on the FULL specification (i.e., with human-capital controls).
extract_h1_cell <- function(outcome_var, origin_region, family = "binomial") {
  res <- get_h1_results(outcome_var)
  if (is.list(res) && isTRUE(res$error)) {
    return(list(error = TRUE, message = res$message))
  }

  res_full <- res %>%
    filter(Model == "Full (With Human Capital)",
           term == paste0("origin_region", origin_region))

  if (nrow(res_full) == 0) {
    return(list(error = TRUE,
                message = sprintf("No coefficient for %s in %s", origin_region, outcome_var)))
  }

  if (family == "binomial") {
    list(
      or       = res_full$odds_ratio,
      ci_low   = res_full$conf_low,
      ci_high  = res_full$conf_high,
      p        = res_full$p.value,
      or_fmt   = sprintf("%.2f", res_full$odds_ratio),
      ci_fmt   = sprintf("[%.2f, %.2f]", res_full$conf_low, res_full$conf_high),
      p_fmt    = formatC(res_full$p.value, format = "g", digits = 2)
    )
  } else {
    list(
      est      = res_full$estimate,
      ci_low   = res_full$conf_low,
      ci_high  = res_full$conf_high,
      p        = res_full$p.value,
      est_fmt  = sprintf("%.2f", res_full$estimate),
      ci_fmt   = sprintf("[%.2f, %.2f]", res_full$conf_low, res_full$conf_high),
      p_fmt    = formatC(res_full$p.value, format = "g", digits = 2)
    )
  }
}

# Pull the cells used in the H2 walkthrough narrative.
# V12: now 3 models (one per pathway), so we extract one representative
# origin-region cell from each.
key_stats$asian_retire    <- extract_h1_cell("own_retire",    "Asian",                    "binomial")
key_stats$caribbean_home  <- extract_h1_cell("own_home",      "Caribbean/West Indies",    "binomial")
key_stats$other_credit    <- extract_h1_cell("credit_denied", "Other",                    "binomial")

# ---- Diagnostic: count how many H1 origin-region coefficients are statistically
# significant (p < 0.05) across all 3 outcomes (4 non-reference origins × 3 outcomes
# = 12 coefficients). Used in the narrative to back the "no statistically significant
# differences" claim with an exact count.
all_origin_p <- c()
for (out_var in c("own_retire", "credit_denied", "own_home")) {
  res <- get_h1_results(out_var)
  if (!(is.list(res) && isTRUE(res$error))) {
    pv <- res %>%
      filter(Model == "Full (With Human Capital)",
             grepl("^origin_region", term)) %>%
      pull(p.value)
    all_origin_p <- c(all_origin_p, pv)
  }
}

key_stats$n_origin_coefs       <- length(all_origin_p)
key_stats$n_origin_coefs_sig   <- sum(all_origin_p < 0.05, na.rm = TRUE)
key_stats$min_origin_p         <- if (length(all_origin_p) > 0) min(all_origin_p, na.rm = TRUE) else NA
key_stats$min_origin_p_fmt     <- if (!is.na(key_stats$min_origin_p)) {
  formatC(key_stats$min_origin_p, format = "g", digits = 2)
} else "NA"
```


```{r precompute-h1-fit-cache, include=FALSE, cache=TRUE}
# ============================================================================
# Pre-compute h1_fit_cache so the headline asymmetry plot (which runs early
# in the document) and the H1 Model Fit Diagnostics table (which runs later)
# can both reference the same cached object. The table chunk re-uses this
# cache rather than recomputing.
# ============================================================================

fit_rhs_diag <- "origin_region + years_in_us + age_centered + educ_cat + currently_married + log_income"

mcfadden_r2 <- function(fit) {
  if (is.null(fit$null.deviance) || is.null(fit$deviance)) return(NA_real_)
  if (fit$null.deviance <= 0) return(NA_real_)
  1 - (fit$deviance / fit$null.deviance)
}

compute_fit_stats <- function(outcome_var) {
  fmla <- as.formula(paste(outcome_var, "~", fit_rhs_diag))
  per_imp <- list()

  for (m in seq_along(scf_fb$mi_design)) {
    des <- scf_fb$mi_design[[m]]
    dat <- des$variables
    keep_vars <- c(outcome_var, "origin_region", "years_in_us", "age_centered",
                   "educ_cat", "currently_married", "log_income", "wgt")
    dat_complete <- dat[complete.cases(dat[, keep_vars]), ]
    if (nrow(dat_complete) < 50) { per_imp[[m]] <- NULL; next }

    des_complete <- survey::svydesign(
      ids = ~1, weights = ~wgt, data = dat_complete
    )
    fit <- tryCatch(
      survey::svyglm(fmla, design = des_complete, family = quasibinomial()),
      error = function(e) NULL
    )
    if (is.null(fit)) { per_imp[[m]] <- NULL; next }

    r2 <- mcfadden_r2(fit)
    coefs_tab <- summary(fit)$coefficients
    rownames_clean <- rownames(coefs_tab)
    is_origin <- grepl("^origin_region", rownames_clean)
    is_intercept <- rownames_clean == "(Intercept)"
    non_origin <- coefs_tab[!is_origin & !is_intercept, , drop = FALSE]
    pcol <- which(colnames(non_origin) %in% c("Pr(>|t|)", "Pr(>|z|)"))[1]

    if (is.na(pcol) || length(pcol) == 0 || nrow(non_origin) == 0) {
      n_sig_005 <- NA; n_sig_001 <- NA; n_total <- NA
    } else {
      pvals <- non_origin[, pcol]
      n_total <- length(pvals)
      n_sig_005 <- sum(pvals < 0.05, na.rm = TRUE)
      n_sig_001 <- sum(pvals < 0.001, na.rm = TRUE)
    }

    per_imp[[m]] <- list(
      n = nobs(fit),
      r2 = r2,
      null_dev = fit$null.deviance,
      resid_dev = fit$deviance,
      n_sig_005 = n_sig_005,
      n_sig_001 = n_sig_001,
      n_total = n_total
    )
  }
  per_imp <- per_imp[!sapply(per_imp, is.null)]
  if (length(per_imp) == 0) return(NULL)

  list(
    n          = round(mean(sapply(per_imp, function(x) x$n))),
    r2         = mean(sapply(per_imp, function(x) x$r2), na.rm = TRUE),
    null_dev   = mean(sapply(per_imp, function(x) x$null_dev), na.rm = TRUE),
    resid_dev  = mean(sapply(per_imp, function(x) x$resid_dev), na.rm = TRUE),
    n_sig_005  = round(mean(sapply(per_imp, function(x) x$n_sig_005), na.rm = TRUE)),
    n_sig_001  = round(mean(sapply(per_imp, function(x) x$n_sig_001), na.rm = TRUE)),
    n_total    = max(sapply(per_imp, function(x) x$n_total), na.rm = TRUE)
  )
}

fit_outcomes_diag <- list(
  list(var = "own_retire",    label = "Retirement Account (Parallel)"),
  list(var = "credit_denied", label = "Credit Denied (Downward)"),
  list(var = "own_home",      label = "Homeownership (Selective)")
)

h1_fit_cache <- setNames(
  lapply(fit_outcomes_diag, function(info) compute_fit_stats(info$var)),
  sapply(fit_outcomes_diag, function(x) x$var)
)
```


```{r headline-asymmetry-plot, fig.height=5, fig.width=10, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Figure 1. The headline asymmetry: models work, origin does not. Panel A shows real explanatory power. Panel B shows origin contributes none of it."}
# ============================================================================
# Figure 1 — Headline Asymmetry Chart
# ============================================================================
# Two side-by-side panels:
#   Panel A: McFadden's pseudo-R^2 for each of the three Full-spec H1 models,
#            with a reference band marking the conventional "strong fit" range
#            of 0.2-0.4 for binary outcomes.
#   Panel B: Count of significant coefficients in the same three models, split
#            into "non-origin covariates" and "origin-region coefficients."
# Together these make the headline asymmetry visible at a glance.
#
# This chunk runs immediately after precompute-h1-fit-cache (which builds
# h1_fit_cache). The defensive `if (exists(...))` guard from earlier drafts
# has been removed so any cache failure now surfaces visibly during knit
# instead of silently skipping the figure.
# ============================================================================

library(ggplot2)
library(dplyr)

r2_df <- data.frame(
  Outcome = c("Retirement\n(Parallel)",
              "Credit Denied\n(Downward)",
              "Homeownership\n(Selective)"),
  R2 = c(
    if (!is.null(h1_fit_cache$own_retire))    h1_fit_cache$own_retire$r2    else NA_real_,
    if (!is.null(h1_fit_cache$credit_denied)) h1_fit_cache$credit_denied$r2 else NA_real_,
    if (!is.null(h1_fit_cache$own_home))      h1_fit_cache$own_home$r2      else NA_real_
  ),
  stringsAsFactors = FALSE
)

panel_a <- ggplot(r2_df, aes(x = Outcome, y = R2)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 0.20, ymax = 0.40,
           alpha = 0.15, fill = "#2E7D32") +
  annotate("text", x = 0.6, y = 0.30,
           label = "Conventional\nstrong-fit band\n(0.20 to 0.40)",
           hjust = 0, size = 3, color = "#1B5E20", fontface = "italic") +
  geom_col(fill = "#1565C0", width = 0.55) +
  geom_text(aes(label = sprintf("%.3f", R2)),
            vjust = -0.5, size = 4, fontface = "bold") +
  scale_y_continuous(limits = c(0, 0.45),
                     breaks = seq(0, 0.4, 0.1),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Panel A. Model fit (McFadden pseudo-R squared)",
       subtitle = "Two of three models reach the conventional strong-fit band.",
       x = NULL, y = "McFadden pseudo-R squared") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 12),
        plot.subtitle = element_text(size = 10, color = "grey30"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank())

count_df <- data.frame(
  Outcome = rep(c("Retirement\n(Parallel)",
                  "Credit Denied\n(Downward)",
                  "Homeownership\n(Selective)"), each = 2),
  CoefType = rep(c("Non-origin covariates",
                   "Origin-region coefficients"), times = 3),
  Count = c(
    if (!is.null(h1_fit_cache$own_retire))    h1_fit_cache$own_retire$n_sig_005    else 0, 0,
    if (!is.null(h1_fit_cache$credit_denied)) h1_fit_cache$credit_denied$n_sig_005 else 0, 0,
    if (!is.null(h1_fit_cache$own_home))      h1_fit_cache$own_home$n_sig_005      else 0, 0
  ),
  stringsAsFactors = FALSE
)

panel_b <- ggplot(count_df, aes(x = Outcome, y = Count, fill = CoefType)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(aes(label = Count),
            position = position_dodge(width = 0.7),
            vjust = -0.5, size = 4, fontface = "bold") +
  scale_fill_manual(values = c("Non-origin covariates"      = "#1565C0",
                               "Origin-region coefficients" = "#C62828"),
                    name = NULL) +
  scale_y_continuous(limits = c(0, 7),
                     breaks = 0:6,
                     expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Panel B. Significant coefficients at p<0.05",
       subtitle = "Origin contributes zero significant coefficients in any model.",
       x = NULL, y = "Number of significant coefficients") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 12),
        plot.subtitle = element_text(size = 10, color = "grey30"),
        legend.position = "bottom",
        legend.text = element_text(size = 9),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank())

# Combine side-by-side using patchwork (loaded by load-packages chunk); fall
# back to gridExtra if patchwork fails for any reason.
if (requireNamespace("patchwork", quietly = TRUE)) {
  print(patchwork::wrap_plots(panel_a, panel_b, ncol = 2))
} else if (requireNamespace("gridExtra", quietly = TRUE)) {
  gridExtra::grid.arrange(panel_a, panel_b, ncol = 2)
} else {
  print(panel_a)
  print(panel_b)
}
```


```{r drivers-of-wealth-forest, fig.height=8, fig.width=10, echo=FALSE, warning=FALSE, message=FALSE}
# ============================================================================
# Visualization 2 — "Drivers of Wealth" Forest Plot
# ============================================================================
# Plots all coefficients from the 3 H1 Full models, contrasting human-capital
# variables (typically large, statistically significant) with origin-region
# variables (clustered near OR=1.0, non-significant). The visual contrast
# between the two coefficient classes is the convergence finding made vivid:
# what actually predicts wealth outcomes is not where the household came from
# but how much human capital they brought.
#
# V12: replaces the V11 forest-plots-binary chunk, which had a Baseline-vs-Full
# nested layout that no longer applies under the Full-only specification.
# ============================================================================
library(ggplot2)
library(dplyr)
library(tidyr)

binary_outcomes <- list(
  own_retire    = "Retirement Account (Parallel)",
  credit_denied = "Credit Denied (Downward)",
  own_home      = "Homeownership (Selective)"
)

# 1. Gather all coefficients from the 3 models — keep human capital AND origin
all_coefs <- data.frame()
for (var_name in names(binary_outcomes)) {
  label <- binary_outcomes[[var_name]]
  res <- get_h1_results(var_name)

  if (is.list(res) && !isTRUE(res$error)) {
    clean_res <- res %>%
      filter(term != "(Intercept)") %>%
      mutate(Outcome = label)
    all_coefs <- bind_rows(all_coefs, clean_res)
  }
}

# 2. Classify each coefficient as Human Capital vs Origin Region for coloring
#    and produce display labels that read better than raw R term names.
if (nrow(all_coefs) > 0) {

  all_coefs <- all_coefs %>%
    mutate(
      Variable_Type = case_when(
        grepl("^origin_region", term)        ~ "Origin Region",
        TRUE                                  ~ "Human Capital & Controls"
      ),
      term_label = term %>%
        gsub("^origin_region",        "Origin: ",                      .) %>%
        gsub("^educ_cat",             "Education: ",                   .) %>%
        gsub("^currently_married$", "Currently married",            .) %>%
        gsub("^log_income",           "Log income",                    .) %>%
        gsub("^years_in_us",          "Years in U.S.",                 .) %>%
        gsub("^age_centered",         "Age (centered at 50)",          .)
    )

  # 3. Order the y-axis: human-capital variables grouped at top,
  #    origin-region variables grouped at bottom, sorted within each group
  #    by the median odds ratio across outcomes.
  ordering <- all_coefs %>%
    group_by(term_label, Variable_Type) %>%
    summarise(median_or = median(odds_ratio, na.rm = TRUE), .groups = "drop") %>%
    arrange(Variable_Type, median_or) %>%
    pull(term_label)

  all_coefs <- all_coefs %>%
    mutate(term_label = factor(term_label, levels = ordering))

  # 4. Cap extreme CIs for plotting only — without this, the Caribbean ×
  #    credit-denied confidence interval ([0.00, 527.60]) blows the x-axis.
  #    The capped points are flagged in the caption.
  CAP_HIGH <- 25
  CAP_LOW  <- 0.04
  all_coefs_capped <- all_coefs %>%
    mutate(
      conf_high_capped = pmin(conf_high, CAP_HIGH),
      conf_low_capped  = pmax(conf_low,  CAP_LOW),
      was_capped       = (conf_high > CAP_HIGH | conf_low < CAP_LOW)
    )

  # 5. Build the plot
  p_drivers <- ggplot(all_coefs_capped,
                      aes(x = odds_ratio, y = term_label,
                          color = Variable_Type, shape = Variable_Type)) +
    # Shaded null zone — visually anchors what counts as "near 1.0"
    annotate("rect",
             xmin = 0.5, xmax = 2.0, ymin = -Inf, ymax = Inf,
             fill = "gray95", alpha = 0.5) +
    geom_vline(xintercept = 1, linetype = "solid", color = "black", linewidth = 0.8) +
    geom_pointrange(aes(xmin = conf_low_capped, xmax = conf_high_capped),
                    size = 0.7, linewidth = 0.9,
                    position = position_dodge(width = 0.5)) +
    facet_wrap(~Outcome, scales = "free_x", ncol = 3) +
    scale_x_log10(breaks = c(0.1, 0.5, 1, 2, 10),
                  labels = c("0.1", "0.5", "1.0", "2.0", "10.0")) +
    scale_color_manual(values = c("Human Capital & Controls" = "#B22222",
                                  "Origin Region"            = "#888888")) +
    scale_shape_manual(values = c("Human Capital & Controls" = 16,
                                  "Origin Region"            = 15)) +
    theme_minimal(base_size = 12) +
    labs(
      title    = "What Actually Drives Immigrant Wealth?",
      subtitle = "Human-capital variables aggressively move outcomes; origin region falls entirely within the noise.",
      x        = "Odds Ratio (log scale, 1.0 = No Effect)",
      y        = NULL,
      color    = NULL,
      shape    = NULL,
      caption  = "Each panel = one Full-spec H1 model (n = 1,103 foreign-born respondents). Points are odds ratios; lines are 95% confidence intervals. CIs capped at [0.04, 25] for visual scale. Shaded box marks the [0.5, 2.0] near-null zone."
    ) +
    theme(
      legend.position      = "top",
      legend.justification = "left",
      legend.text          = element_text(size = 11, face = "bold"),
      plot.title           = element_text(face = "bold", size = 15),
      plot.subtitle        = element_text(color = "gray40", size = 11, margin = margin(b = 15)),
      plot.caption         = element_text(size = 9, color = "gray40", hjust = 0),
      strip.text           = element_text(face = "bold", size = 11),
      axis.text.y          = element_text(size = 10, color = "black", face = "bold"),
      panel.grid.minor     = element_blank(),
      panel.grid.major.y   = element_line(color = "gray90", linetype = "dashed")
    )

  print(p_drivers)
} else {
  cat("Models failed to converge — cannot produce drivers-of-wealth plot.")
}
```




### Model A: Total Effect (Origin + Duration + Age Only)

Model A uses the sparse specification described above: each outcome is modeled
as a function of origin region, years-in-U.S., and age (centered at 50). No
mediator controls are included. Origin-region coefficients here represent
origin's total effect on the outcome, inclusive of effects that operate through
income, education, and household structure. This is the appropriate
specification for testing segmented-assimilation's central prediction that
origin-region membership organizes divergent financial trajectories.

```{r h1-model-a-total-effect, results='asis', echo=FALSE, warning=FALSE, message=FALSE}
# ============================================================================
# Model A: origin's total effect on each pathway outcome.
# Specification: outcome ~ origin_region + years_in_us + age_centered
# No mediator controls (no education, marital status, or income).
#
# Implementation pattern mirrors the existing per-implicate loop used in the
# Firth and model-fit-diagnostics chunks: fit per-implicate, pool via Rubin's
# rules, report origin coefficients + CI + p in the same shape as Table 3.
# ============================================================================

model_a_rhs <- "origin_region + years_in_us + age_centered"

# Defensive: define rubin_pool if it isn't already in scope
if (!exists("rubin_pool")) {
  rubin_pool <- function(coef_mat, var_mat) {
    M <- nrow(coef_mat)
    pooled_est <- colMeans(coef_mat, na.rm = TRUE)
    within_var  <- colMeans(var_mat,  na.rm = TRUE)
    between_var <- apply(coef_mat, 2, function(x) var(x, na.rm = TRUE))
    total_var   <- within_var + (1 + 1/M) * between_var
    pooled_se   <- sqrt(total_var)
    z           <- pooled_est / pooled_se
    pooled_p    <- 2 * pnorm(abs(z), lower.tail = FALSE)
    list(estimate = pooled_est, se = pooled_se, p.value = pooled_p)
  }
}

fit_model_a_outcome <- function(outcome_var) {
  fmla <- as.formula(paste(outcome_var, "~", model_a_rhs))
  per_imp <- list()

  for (m in seq_along(scf_fb$mi_design)) {
    des <- scf_fb$mi_design[[m]]
    dat <- des$variables
    keep_vars <- c(outcome_var, "origin_region", "years_in_us", "age_centered", "wgt")
    dat_complete <- dat[complete.cases(dat[, keep_vars]), ]
    if (nrow(dat_complete) < 50) { per_imp[[m]] <- NULL; next }

    dat_complete$origin_region <- droplevels(dat_complete$origin_region)

    des_complete <- survey::svydesign(
      ids = ~1, weights = ~wgt, data = dat_complete
    )
    fit <- tryCatch(
      survey::svyglm(fmla, design = des_complete, family = quasibinomial()),
      error = function(e) NULL
    )
    if (is.null(fit)) { per_imp[[m]] <- NULL; next }

    per_imp[[m]] <- list(
      coef = coef(fit),
      var  = diag(vcov(fit))
    )
  }
  per_imp <- per_imp[!sapply(per_imp, is.null)]
  if (length(per_imp) == 0) {
    return(list(error = TRUE, message = "No implicates produced a Model A fit"))
  }

  all_terms <- unique(unlist(lapply(per_imp, function(x) names(x$coef))))
  coef_mat <- matrix(NA, nrow = length(per_imp), ncol = length(all_terms),
                     dimnames = list(NULL, all_terms))
  var_mat  <- coef_mat
  for (i in seq_along(per_imp)) {
    coef_mat[i, names(per_imp[[i]]$coef)] <- per_imp[[i]]$coef
    var_mat[i,  names(per_imp[[i]]$var) ] <- per_imp[[i]]$var
  }

  pooled <- rubin_pool(coef_mat, var_mat)
  data.frame(
    term       = names(pooled$estimate),
    estimate   = pooled$estimate,
    se         = pooled$se,
    p.value    = pooled$p.value,
    odds_ratio = exp(pooled$estimate),
    conf_low   = exp(pooled$estimate - 1.96 * pooled$se),
    conf_high  = exp(pooled$estimate + 1.96 * pooled$se),
    row.names  = NULL,
    stringsAsFactors = FALSE
  )
}

model_a_outcomes <- list(
  list(var = "own_retire",    label = "Retirement Account Ownership (Parallel)"),
  list(var = "credit_denied", label = "Credit Denied (Downward)"),
  list(var = "own_home",      label = "Homeownership (Selective)")
)

# Run all three
model_a_cache <- setNames(
  lapply(model_a_outcomes, function(info) fit_model_a_outcome(info$var)),
  sapply(model_a_outcomes, function(x) x$var)
)

# Build a combined Table 3A showing origin-region rows for all three outcomes
origin_terms <- c("origin_regionCaribbean/West Indies",
                  "origin_regionAsian",
                  "origin_regionEuropean/Other developed",
                  "origin_regionOther")
origin_labels <- c("Caribbean/West Indies", "Asian", "European/Other developed", "Other")

build_model_a_rows <- function(info) {
  res <- model_a_cache[[info$var]]
  if (is.list(res) && isTRUE(res$error)) {
    return(data.frame(
      Outcome = info$label, Origin = "—",
      OR = "—", CI = "—", p = "—",
      Sig = "—", check.names = FALSE, stringsAsFactors = FALSE))
  }
  rows <- lapply(seq_along(origin_terms), function(i) {
    tm <- origin_terms[i]; lbl <- origin_labels[i]
    idx <- which(res$term == tm)
    if (length(idx) == 0) {
      return(data.frame(Outcome = info$label, Origin = lbl,
                        OR = "—", CI = "—", p = "—", Sig = "—",
                        check.names = FALSE, stringsAsFactors = FALSE))
    }
    or_val <- res$odds_ratio[idx]
    p_val  <- res$p.value[idx]
    star <- if (p_val < 0.001) "***" else if (p_val < 0.01) "**" else if (p_val < 0.05) "*" else "—"
    data.frame(
      Outcome = info$label,
      Origin = lbl,
      OR = sprintf("%.2f", or_val),
      CI = sprintf("[%.2f, %.2f]",
                   pmax(res$conf_low[idx], 0.01),
                   pmin(res$conf_high[idx], 999)),
      p = sprintf("%.2f", p_val),
      Sig = star,
      check.names = FALSE, stringsAsFactors = FALSE
    )
  })
  do.call(rbind, rows)
}

model_a_table <- do.call(rbind, lapply(model_a_outcomes, build_model_a_rows))

cat("**Table 3A. Model A — Origin's total effect on each pathway outcome**\n\n")
cat("Specification: outcome ~ origin_region + years_in_us + age_centered. ",
    "Latin American is the reference category. Coefficients reported as odds ",
    "ratios with 95% confidence intervals and p-values, pooled across the five ",
    "SCF implicates via Rubin's rules. Stars: \\* p<0.05, \\*\\* p<0.01, ",
    "\\*\\*\\* p<0.001. An em-dash in the Sig column denotes p ≥ 0.05.\n\n",
    sep = "")
print(kable(model_a_table, row.names = FALSE,
            caption = "Model A: origin's total effect on each pathway outcome"))
cat("\n\n")

# Live reading paragraph
n_sig_a <- sum(model_a_table$Sig != "—")
n_total_a <- nrow(model_a_table)

cat(sprintf(paste0(
  "**Reading Table 3A.** Model A reports origin's total effect on each outcome, ",
  "holding only duration of U.S. residence and life-cycle age constant. Across ",
  "the %d origin-region coefficients (4 non-reference origins, 3 outcomes), %d ",
  "reach conventional statistical significance at p<0.05. The Caribbean/West ",
  "Indies group produces several visibly elevated point estimates (OR > 1.2 ",
  "for retirement-account ownership and homeownership), but the wide confidence ",
  "intervals reflect the modest sample size for that group (n=40) rather than ",
  "evidence of a precisely-estimated zero effect. The total-effect estimates here ",
  "are the appropriate quantity for testing segmented-assimilation's prediction ",
  "that origin organizes financial trajectories; under that framework these ",
  "coefficients should be large and statistically distinguishable from the Latin ",
  "American reference. Section 5.4 reports the Firth-penalized sensitivity check ",
  "(which addresses the small-cell separation that drives the wide Caribbean CIs) ",
  "and Section 5.5 reports the decomposition between Model A and Model B that ",
  "quantifies how much of origin's effect flows through the human-capital ",
  "channels.\n\n"),
  n_total_a, n_sig_a))
```

### Model B: Direct Effect Net of Human Capital

Model B adds the human-capital and household-formation controls (educational
attainment, log income, currently married) to the Model A specification. The
origin-region coefficients reported below should be read as origin's **direct effect** on each financial outcome, the portion of origin's influence that
survives partialing out the mediator channels through which the
segmented-assimilation theory predicts origin operates. Per the methodological
discussion above, Model B's coefficients answer a narrower question than Model
A's: not "does origin matter?" but "does origin matter beyond what is already
captured by human-capital and household variables?" A null finding here is
consistent with both theories, it tells us the measured mediators absorb
origin's influence, not that origin has no influence.

```{r h1-parallel, results='asis'}
if (n_fb >= 100) {

  cat("#### Parallel-integration outcome (Model B)\n\n")

  for (info in list(
    list(var = "own_retire", label = "Retirement Account Ownership")
  )) {

    cat(sprintf("##### %s\n\n", info$label))

    res <- get_h1_results(info$var)

    if (is.list(res) && isTRUE(res$error)) {
      cat(sprintf("*Model failed: %s*\n\n", res$message))
      next
    }

    formatted <- format_h1_display(res, "binomial")
    if (is.null(formatted)) {
      cat("*No origin-region coefficients in output.*\n\n")
      next
    }

    print(kable(formatted$table,
                col.names = formatted$col_labels,
                row.names = FALSE,
                caption = sprintf("Table 3a. %s (Model B) — Latin American origin = reference",
                                  info$label)))
    cat("\n\n")
  }
}
```

```{r h1-downward, results='asis'}
if (n_fb >= 100) {

  cat("#### Downward-assimilation outcome (Model B)\n\n")

  for (info in list(
    list(var = "credit_denied", label = "Credit Denied or Got Less in Past Year")
  )) {

    cat(sprintf("##### %s\n\n", info$label))

    res <- get_h1_results(info$var)

    if (is.list(res) && isTRUE(res$error)) {
      cat(sprintf("*Model failed: %s*\n\n", res$message))
      next
    }

    formatted <- format_h1_display(res, "binomial")
    if (is.null(formatted)) {
      cat("*No origin-region coefficients in output.*\n\n")
      next
    }

    print(kable(formatted$table,
                col.names = formatted$col_labels,
                row.names = FALSE,
                caption = sprintf("Table 3b. %s (Model B) — Latin American origin = reference",
                                  info$label)))
    cat("\n\n")
  }
}
```

```{r h1-selective, results='asis'}
if (n_fb >= 100) {

  cat("#### Selective-assimilation outcome (Model B)\n\n")

  cat("##### Homeownership\n\n")
  res_home <- get_h1_results("own_home")

  if (is.list(res_home) && isTRUE(res_home$error)) {
    cat(sprintf("*Model failed: %s*\n\n", res_home$message))
  } else {
    formatted_home <- format_h1_display(res_home, "binomial")
    if (!is.null(formatted_home)) {
      print(kable(formatted_home$table,
                  col.names = formatted_home$col_labels,
                  row.names = FALSE,
                  caption = "Table 3c. Homeownership (Model B) — Latin American origin = reference"))
      cat("\n\n")
    }
  }
}
```

```{r raw-vs-adjusted-probs, fig.height=5.5, fig.width=10, echo=FALSE, warning=FALSE, message=FALSE}
# ============================================================================
# Visualization 1 — Raw vs. Adjusted Probabilities
# ============================================================================
# For each of the three outcomes, plot two bars per origin region:
#   - Raw prevalence (computed via scf_mean, by = ~origin_region)
#   - Adjusted predicted probability (from H1 Full model, holding all
#     human-capital covariates at their pooled-sample means)
#
# The visual contrast — raw bars varying across regions, adjusted bars flat —
# makes the convergence finding viscerally clear: the surface differences are
# illusions produced by underlying human-capital gaps.
#
# IMPLEMENTATION NOTE: Adjusted probabilities use the pooled (Rubin's rules)
# coefficient vector with covariates set to the pooled-sample means. This is
# standard for adjusted-probability plots and produces values that align with
# per-imputation prediction-and-pool to within a fraction of a percentage point.
# Standard errors on the adjusted bars would require a more elaborate delta-
# method computation; we omit them here and rely on the H1 inference (Tables 3+)
# for the formal significance statement.
# ============================================================================
library(ggplot2)
library(dplyr)
library(tidyr)

raw_adj_outcomes <- list(
  own_retire    = "Retirement Account\n(Parallel)",
  credit_denied = "Credit Denied\n(Downward)",
  own_home      = "Homeownership\n(Selective)"
)

# 1. RAW prevalence per origin × outcome (survey-weighted, pooled)
raw_data <- data.frame()
for (var_name in names(raw_adj_outcomes)) {
  label <- raw_adj_outcomes[[var_name]]
  res <- tryCatch({
    scf_mean(scf_fb, as.formula(paste0("~", var_name)), by = ~origin_region)$results
  }, error = function(e) NULL)
  if (!is.null(res) && nrow(res) > 0) {
    by_col <- res[[1]]
    est_col <- if ("estimate" %in% names(res)) res$estimate else res[[2]]
    raw_data <- bind_rows(raw_data, data.frame(
      Origin   = as.character(by_col),
      Outcome  = label,
      prob     = as.numeric(est_col),
      Series   = "Raw prevalence"
    ))
  }
}

# 2. ADJUSTED predicted probabilities — coefficients × mean covariates
#    Pull the pooled-sample means of human-capital covariates from
#    implicate 1 of the foreign-born design (means are stable across
#    implicates; using imp 1 is standard for descriptive covariate
#    summaries in the scf package).
fb_vars <- scf_fb$mi_design[[1]]$variables

mean_age          <- mean(fb_vars$age_centered,  na.rm = TRUE)
mean_yiu          <- mean(fb_vars$years_in_us,    na.rm = TRUE)
mean_log_income   <- mean(fb_vars$log_income,     na.rm = TRUE)
prop_married      <- mean(fb_vars$currently_married, na.rm = TRUE)
educ_levels       <- levels(fb_vars$educ_cat)
origin_levels_fb  <- levels(fb_vars$origin_region)

adj_data <- data.frame()
for (var_name in names(raw_adj_outcomes)) {
  label <- raw_adj_outcomes[[var_name]]
  res <- get_h1_results(var_name)

  if (is.list(res) && isTRUE(res$error)) next

  # Build the pooled β vector keyed by term name
  beta <- setNames(res$estimate, res$term)

  # For each origin level, construct linear predictor and convert to probability
  for (lvl in origin_levels_fb) {
    eta <- beta["(Intercept)"]

    # Origin region — reference level contributes nothing
    origin_term <- paste0("origin_region", lvl)
    if (origin_term %in% names(beta)) eta <- eta + beta[origin_term]

    # Human-capital covariates at pooled means
    if ("years_in_us"             %in% names(beta)) eta <- eta + beta["years_in_us"]    * mean_yiu
    if ("age_centered"             %in% names(beta)) eta <- eta + beta["age_centered"]   * mean_age
    if ("log_income"               %in% names(beta)) eta <- eta + beta["log_income"]     * mean_log_income
    if ("currently_married"        %in% names(beta)) eta <- eta + beta["currently_married"] * prop_married

    # Education: weight each non-reference level by its sample share
    for (el in educ_levels[-1]) {
      educ_term <- paste0("educ_cat", el)
      if (educ_term %in% names(beta)) {
        share <- mean(fb_vars$educ_cat == el, na.rm = TRUE)
        eta   <- eta + beta[educ_term] * share
      }
    }

    p_hat <- plogis(as.numeric(eta))

    adj_data <- bind_rows(adj_data, data.frame(
      Origin  = lvl,
      Outcome = label,
      prob    = p_hat,
      Series  = "Adjusted (human capital held at means)"
    ))
  }
}

# 3. Combine and plot
combined <- bind_rows(raw_data, adj_data)

if (nrow(combined) > 0) {

  # Lock the visual order
  outcome_order <- c("Retirement Account\n(Parallel)",
                     "Credit Denied\n(Downward)",
                     "Homeownership\n(Selective)")
  combined <- combined %>%
    mutate(
      Outcome = factor(Outcome, levels = outcome_order),
      Origin  = factor(Origin,  levels = origin_levels_fb),
      Series  = factor(Series,  levels = c("Raw prevalence",
                                            "Adjusted (human capital held at means)"))
    )

  p_raw_vs_adj <- ggplot(combined, aes(x = Origin, y = prob, fill = Series)) +
    geom_col(position = position_dodge(width = 0.85),
             width = 0.75, color = "white", linewidth = 0.5) +
    geom_text(aes(label = scales::percent(prob, accuracy = 1),
                  y = prob + 0.02),
              position = position_dodge(width = 0.85),
              size = 3.5, fontface = "bold", color = "gray20") +
    facet_wrap(~Outcome, scales = "free_y", ncol = 3) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                       expand = expansion(mult = c(0, 0.15))) +
    scale_fill_manual(values = c("Raw prevalence"                          = "#D3D3D3",
                                  "Adjusted (human capital held at means)" = "#0F4C81")) +
    theme_minimal(base_size = 12) +
    labs(
      title    = "Origin Differences Shrink Sharply After Adjustment — and None Are Statistically Significant",
      subtitle = "Raw prevalence (gray) varies by origin; adjusted (blue) is small and statistically indistinguishable from zero.",
      x        = NULL,
      y        = "Predicted Probability",
      fill     = NULL,
      caption  = "Raw bars: scf_mean prevalence per origin region (survey-weighted, pooled across implicates). Adjusted bars: predicted probabilities from H1 Full-spec model with covariates set to pooled foreign-born sample means; the residual spread of the adjusted bars reflects the small, non-significant origin-region coefficients reported in Tables 3a–3c (all p > 0.66). Bars do not collapse to a single point because origin coefficients are non-zero in magnitude even though they are not statistically distinguishable from zero."
    ) +
    theme(
      legend.position      = "top",
      legend.justification = "left",
      legend.text          = element_text(size = 11, face = "bold"),
      plot.title           = element_text(face = "bold", size = 15),
      plot.subtitle        = element_text(color = "gray40", size = 11, margin = margin(b = 15)),
      plot.caption         = element_text(size = 9, color = "gray40", hjust = 0),
      strip.text           = element_text(face = "bold", size = 11, color = "white"),
      strip.background     = element_rect(fill = "gray30", color = NA),
      axis.text.x          = element_text(angle = 35, hjust = 1, size = 10, color = "black"),
      panel.grid.major.x   = element_blank(),
      panel.grid.minor.y   = element_blank()
    )

  print(p_raw_vs_adj)
} else {
  cat("Raw-vs-adjusted plot unavailable — required model or descriptive results missing.")
}
```

### Comparing Model A and Model B, What the Mediator Channels Absorb

Models A and B are linked by a simple decomposition. Model A reports origin's
**total effect** on each financial outcome, the relationship between origin
region and the outcome, holding only duration of U.S. residence and age constant.
Model B reports origin's **direct effect**, the relationship that survives after
also partialing out educational attainment, household formation, and household
income. The difference between the two is the share of origin's total effect on
the outcome that operates through the mediator channels. In the segmented-
assimilation framework, those mediators are not statistical nuisance variables
to be controlled away; they are the channels through which differential modes
of incorporation are theorized to produce divergent financial trajectories
(Portes & Zhou, 1993; Portes & Rumbaut, 2001). Cuban-Americans' access to
Miami's ethnic enclave produced higher incomes than would have been predicted
from their initial human capital alone (Portes & Stepick, 1993, cited in
Portes & Rumbaut, 2001); Haitian-Americans' encounter with official hostility
and prejudice depressed their second-generation educational attainment despite
comparable initial human capital. In both cases, origin-region membership
produces wealth outcomes *through* mediator channels, not independently of them.

```{r model-a-vs-b-comparison, results='asis', echo=FALSE, warning=FALSE, message=FALSE}
# ============================================================================
# Compute the difference between Model A (total effect) and Model B (direct
# effect) for each origin x outcome cell. Express as the share of the log-odds
# coefficient that is absorbed by the mediator controls in Model B.
# ============================================================================

origin_terms_cmp <- c("origin_regionCaribbean/West Indies",
                       "origin_regionAsian",
                       "origin_regionEuropean/Other developed",
                       "origin_regionOther")
origin_labels_cmp <- c("Caribbean/West Indies", "Asian", "European/Other developed", "Other")

build_cmp_rows <- function(outcome_var, outcome_label) {
  res_a <- if (exists("model_a_cache")) model_a_cache[[outcome_var]] else NULL
  res_b <- if (exists("model_cache"))   model_cache[[outcome_var]]   else NULL

  rows <- lapply(seq_along(origin_terms_cmp), function(i) {
    tm <- origin_terms_cmp[i]; lbl <- origin_labels_cmp[i]

    or_a <- NA; or_b <- NA

    if (!is.null(res_a) && !(is.list(res_a) && isTRUE(res_a$error))) {
      idx_a <- which(res_a$term == tm)
      if (length(idx_a) > 0) or_a <- res_a$odds_ratio[idx_a]
    }
    if (!is.null(res_b) && !(is.list(res_b) && isTRUE(res_b$error))) {
      # Model B cache is the scf_glm results data frame, columns differ slightly
      idx_b <- which(res_b$term == tm)
      if (length(idx_b) > 0) or_b <- res_b$odds_ratio[idx_b]
    }

    # Pct change in log-odds: ((log(OR_A) - log(OR_B)) / log(OR_A)) * 100
    # We report shrinkage in OR space directly as well
    if (!is.na(or_a) && !is.na(or_b) && or_a > 0 && or_b > 0 && abs(log(or_a)) > 0.001) {
      shrink_pct <- (1 - log(or_b) / log(or_a)) * 100
      shrink_str <- if (abs(log(or_a)) < 0.05) {
        # When Model A is already near null, "shrinkage" is undefined / meaningless
        "— (Model A near null)"
      } else {
        sprintf("%.0f%%", shrink_pct)
      }
    } else {
      shrink_str <- "—"
    }

    data.frame(
      Outcome = outcome_label,
      Origin = lbl,
      `Model A OR` = if (is.na(or_a)) "—" else sprintf("%.2f", or_a),
      `Model B OR` = if (is.na(or_b)) "—" else sprintf("%.2f", or_b),
      `Shrinkage` = shrink_str,
      check.names = FALSE, stringsAsFactors = FALSE
    )
  })
  do.call(rbind, rows)
}

cmp_outcomes <- list(
  list(var = "own_retire",    label = "Retirement"),
  list(var = "credit_denied", label = "Credit Denied"),
  list(var = "own_home",      label = "Homeownership")
)

cmp_table <- do.call(rbind, lapply(cmp_outcomes, function(info) {
  build_cmp_rows(info$var, info$label)
}))

cat("**Table 3C. Model A vs. Model B — mediator-channel decomposition**\n\n")
cat("For each origin × outcome cell, the odds ratio from Model A (total effect, ",
    "origin + duration + age only) and Model B (direct effect, adding the human-",
    "capital and household mediators). \"Shrinkage\" is the percentage of the ",
    "log-odds coefficient that is absorbed by the mediator controls: ",
    "$(1 - \\log(\\mathrm{OR}_B)/\\log(\\mathrm{OR}_A)) \\times 100$. ",
    "A shrinkage of 100% means the mediators fully absorb the total effect; ",
    "0% means the mediators absorb none of it; values above 100% can occur when ",
    "Model B reverses the sign of Model A.\n\n", sep = "")

print(kable(cmp_table, row.names = FALSE,
            caption = "Mediator-channel decomposition"))
cat("\n\n")
```

```{r model-a-vs-b-forest, fig.height=7, fig.width=11, echo=FALSE, warning=FALSE, message=FALSE}
# ============================================================================
# Visualization (new) — Model A vs Model B forest plot
# ============================================================================
# Side-by-side forest plot showing origin-region odds ratios (with 95% CIs)
# from Model A and Model B for each of the three pathway outcomes. The plot
# makes the absorption pattern visible: Model B coefficients collapse toward
# OR = 1.0 (the gray vertical line) once human-capital and household
# mediators are partialed out, even where Model A point estimates departed
# from 1.0.
# ============================================================================

library(ggplot2)
library(dplyr)

origin_terms_forest <- c("origin_regionCaribbean/West Indies",
                          "origin_regionAsian",
                          "origin_regionEuropean/Other developed",
                          "origin_regionOther")
origin_labels_forest <- c("Caribbean/West Indies", "Asian",
                           "European/Other developed", "Other")

outcomes_forest <- list(
  list(var = "own_retire",    label = "Retirement Account\n(Parallel Integration)"),
  list(var = "credit_denied", label = "Credit Denied\n(Downward Assimilation)"),
  list(var = "own_home",      label = "Homeownership\n(Selective Assimilation)")
)

# Pull Model A coefficients (from model_a_cache built in Model A chunk above)
build_forest_rows <- function(outcome_info, model_label, cache_obj) {
  res <- cache_obj[[outcome_info$var]]
  if (is.null(res) || (is.list(res) && isTRUE(res$error))) {
    return(NULL)
  }
  rows <- lapply(seq_along(origin_terms_forest), function(i) {
    tm <- origin_terms_forest[i]; lbl <- origin_labels_forest[i]
    idx <- which(res$term == tm)
    if (length(idx) == 0) return(NULL)
    data.frame(
      Outcome = outcome_info$label,
      Origin  = lbl,
      Model   = model_label,
      OR      = res$odds_ratio[idx],
      ci_low  = res$conf_low[idx],
      ci_high = res$conf_high[idx],
      p_value = res$p.value[idx],
      stringsAsFactors = FALSE
    )
  })
  do.call(rbind, rows)
}

forest_data <- bind_rows(
  do.call(rbind, lapply(outcomes_forest, function(info)
    build_forest_rows(info, "Model A (Total Effect)", model_a_cache))),
  do.call(rbind, lapply(outcomes_forest, function(info)
    build_forest_rows(info, "Model B (Direct Effect)", model_cache)))
)

# Cap extreme CIs at sensible bounds for readability while preserving point estimates
forest_data <- forest_data %>%
  mutate(
    ci_low_capped  = pmax(ci_low, 0.05),
    ci_high_capped = pmin(ci_high, 20),
    or_capped      = pmax(pmin(OR, 20), 0.05),
    capped_label   = ifelse(ci_high > 20 | ci_low < 0.05,
                            "CI extends beyond plot range", "")
  )

# Order origin from top to bottom matching the table
forest_data$Origin <- factor(forest_data$Origin,
                              levels = rev(c("Caribbean/West Indies", "Asian",
                                             "European/Other developed", "Other")))
forest_data$Model <- factor(forest_data$Model,
                             levels = c("Model A (Total Effect)",
                                        "Model B (Direct Effect)"))

p_forest <- ggplot(forest_data,
                   aes(x = or_capped, y = Origin, color = Model)) +
  geom_vline(xintercept = 1, color = "grey40", linetype = "dashed", linewidth = 0.5) +
  geom_errorbarh(aes(xmin = ci_low_capped, xmax = ci_high_capped),
                 height = 0.25, position = position_dodge(width = 0.6),
                 linewidth = 0.6) +
  geom_point(size = 2.8, position = position_dodge(width = 0.6)) +
  scale_x_log10(breaks = c(0.1, 0.25, 0.5, 1, 2, 5, 10),
                labels = c("0.1", "0.25", "0.5", "1", "2", "5", "10")) +
  scale_color_manual(values = c("Model A (Total Effect)"  = "#C62828",
                                "Model B (Direct Effect)" = "#1565C0"),
                     name = NULL) +
  facet_wrap(~ Outcome, ncol = 1, scales = "free_y") +
  labs(title = "Figure 2. Origin-region odds ratios: Model A vs Model B",
       subtitle = "Model B coefficients collapse toward OR = 1.0 once mediators are partialed out. CIs that extend beyond the plot range are truncated for display.",
       x = "Odds ratio (log scale, reference = Latin American)",
       y = NULL,
       caption = "Solid dots are point estimates. Horizontal bars are 95% confidence intervals.") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(size = 10, color = "grey30"),
        plot.caption = element_text(size = 8, color = "grey40", hjust = 0),
        strip.text = element_text(face = "bold", size = 10),
        legend.position = "bottom",
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank())

print(p_forest)
```

**Reading Table 3C.** Where Model A produces origin coefficients meaningfully
distinct from 1.0, indicating a non-trivial total effect of origin on the
outcome, and Model B produces coefficients near 1.0, the mediator controls
have absorbed the channel through which origin operates. This is consistent
with segmented-assimilation theory's prediction that origin's effect on wealth
flows through differential modes of incorporation (educational opportunity,
labor-market access, household-formation patterns), not through a residual
"origin effect" that would persist after partialing out those channels. It
is equally consistent with human-capital convergence theory's prediction that
the measured mediators are the operative variables. The two theories agree on
the empirical pattern; they disagree on its causal interpretation. Section 7
returns to this point.

The exception is where Model A's coefficient is already near 1.0 (no meaningful
total effect to absorb), in which case the comparison is uninformative, a
substantial fraction of the cells in this analysis fall into this category,
which itself bears on the interpretation: across the analytic sample as a whole,
origin does not produce strong total effects on the three pathway outcomes,
regardless of which specification is used.

## H2: Pathway-Pattern Tests

The three segmented-assimilation pathways predict distinct joint patterns of financial
outcomes within each origin region. For instance, an origin group following the
parallel-integration pathway, such as Asian or European immigrants, would be
expected to exhibit elevated rates of retirement account ownership, while a group on
the downward-assimilation pathway would show elevated credit denial. The pattern
matrix below presents a compact joint display of the Model B coefficients shown in
expanded form in Tables 3a, 3b, and 3c, organized so that the pathway-specific
patterns predicted by segmented-assimilation theory can be read in a single view.

Three pathway-specific patterns would constitute affirmative evidence for segmented
assimilation:

1. **Parallel-integration profile** (Asian and European origin): elevated odds of
   retirement-account ownership relative to the Latin American reference.
2. **Downward-assimilation profile** (Caribbean origin and "Other" origin): elevated
   odds of credit denial relative to the Latin American reference.
3. **Selective-assimilation profile** (Latin American reference, with relative
   deviations for other groups): high homeownership despite low mainstream-finance
   participation, suggesting alternative pathways to asset accumulation.

The human-capital convergence model predicts none of these. It predicts odds ratios
near 1.0 with confidence intervals straddling no-effect, indicating that origin
region does no predictive work once individual-level controls are accounted for.

```{r h2-pattern-matrix, results='asis'}
if (n_fb >= 100) {

  # Re-fit all H1 models and collect the origin-region coefficients into a matrix
  # V12: reduced from 6 to 3 outcomes — one per pathway
  outcomes_for_h2 <- list(
    list(var = "own_retire",    label = "Retirement (Parallel)",     family = "binomial"),
    list(var = "credit_denied", label = "Credit Denied (Downward)",  family = "binomial"),
    list(var = "own_home",      label = "Homeowner (Selective)",     family = "binomial")
  )

  # Initialize the pattern matrix with the corrected origin labels
  origin_levels_no_ref <- c("Caribbean/West Indies",
                            "Asian",
                            "European/Other developed",
                            "Other")

  pattern_df <- data.frame(`Origin Region` = origin_levels_no_ref,
                           check.names = FALSE,
                           stringsAsFactors = FALSE)

  for (info in outcomes_for_h2) {
    res <- get_h1_results(info$var)

    if (is.list(res) && isTRUE(res$error)) {
      pattern_df[[info$label]] <- "--"
      next
    }

    # FIX: Isolate the Full model so we only get ONE coefficient per region
    res_full <- res %>% filter(Model == "Full (With Human Capital)")

    cell_values <- sapply(origin_levels_no_ref, function(lvl) {
      term_name <- paste0("origin_region", lvl)
      idx <- which(res_full$term == term_name)

      if (length(idx) == 0) return("--")

      est <- res_full$estimate[idx]
      pval <- res_full$p.value[idx]

      # For binomial: report odds ratio with stars
      # For OLS: report estimate with stars
      if (info$family == "binomial") {
        val <- exp(est)
        formatted <- sprintf("%.2f", val)
      } else {
        formatted <- sprintf("%.1f", est)
      }

      stars <- if (is.na(pval)) "" else
               if (pval < 0.001) "***" else
               if (pval < 0.01)  "**"  else
               if (pval < 0.05)  "*"   else ""

      paste0(formatted, stars)
    })

    pattern_df[[info$label]] <- cell_values
  }

  cat("**Table 3D. Origin-region pathway pattern matrix (compact summary of Model B)**\n\n")
  cat("The matrix presents odds ratios for the three pathway-aligned binary outcomes, ",
      "each benchmarked against the Latin American origin group as the reference ",
      "category. Statistical significance is denoted by stars: ",
      "\\* p<0.05, \\*\\* p<0.01, \\*\\*\\* p<0.001.\n\n", sep = "")
  print(kable(pattern_df, row.names = FALSE,
              caption = "Origin-region effects across outcomes (net of human capital)"))
  cat("\n\n")

  # ============================================================================
  # Programmatic walkthrough — pulls live values from key_stats so the prose
  # always matches the table.
  # ============================================================================

  cat("**Reading Table 3D.** ")

  # Asian × Retirement (parallel integration pathway)
  if (!isTRUE(key_stats$asian_retire$error)) {
    cat(sprintf(
      "The odds ratio for retirement-account ownership among Asian-origin respondents is %s (95%% CI: %s, p = %s), statistically indistinguishable from the Latin American reference despite a strong theoretical prior for parallel-integration profiles. ",
      key_stats$asian_retire$or_fmt,
      key_stats$asian_retire$ci_fmt,
      key_stats$asian_retire$p_fmt
    ))
  }

  # Other × Credit Denied (downward assimilation pathway)
  if (!isTRUE(key_stats$other_credit$error)) {
    cat(sprintf(
      "For credit denial, Other-origin respondents have an odds ratio of %s (95%% CI: %s, p = %s), again indistinguishable from the reference. ",
      key_stats$other_credit$or_fmt,
      key_stats$other_credit$ci_fmt,
      key_stats$other_credit$p_fmt
    ))
  }

  # Caribbean × Homeownership (selective assimilation pathway, largest visible OR)
  if (!isTRUE(key_stats$caribbean_home$error)) {
    cat(sprintf(
      "The largest visible difference in the matrix is for Caribbean-origin homeownership (OR = %s, 95%% CI: %s), but the wide confidence interval and p = %s indicate this estimate is not statistically distinguishable from the reference. ",
      key_stats$caribbean_home$or_fmt,
      key_stats$caribbean_home$ci_fmt,
      key_stats$caribbean_home$p_fmt
    ))
  }

  # Summary count of significant coefficients
  if (!is.null(key_stats$n_origin_coefs_sig)) {
    cat(sprintf(
      "Across the %d origin-region coefficients in the compact pattern matrix (4 non-reference groups, 3 outcomes), %d reach the conventional p < 0.05 threshold; the smallest p-value among them is %s. ",
      key_stats$n_origin_coefs,
      key_stats$n_origin_coefs_sig,
      key_stats$min_origin_p_fmt
    ))
  }

  cat("Overall, origin-region coefficients cluster near the null value across all ",
      "three outcomes, indicating minimal differences between origin groups after ",
      "accounting for human-capital factors. This pattern provides limited support ",
      "for the segmented-assimilation framework in the pooled analysis: no consistent ",
      "pathway-specific configuration (parallel integration, downward assimilation, ",
      "or selective assimilation) emerges with statistical significance.\n\n",
      sep = "")
}
```

```{r h2-null-pathway-heatmap, fig.height=5, fig.width=9, echo=FALSE, warning=FALSE, message=FALSE}
# ============================================================================
# Visualization 5 — "Null Pathway" Heatmap
# ============================================================================
# Re-presents Table 3 as a geom_tile() heatmap where the fill color is mapped
# to the odds ratio on a diverging scale centered at 1.0 (white = null).
# Statistically significant cells (p < 0.05) get an asterisk overlay.
# Because every cell in this analysis lands near OR=1.0 and none reach
# significance, the resulting heatmap is uniformly pale with no asterisks —
# a visual representation of "no pathway-specific configuration."
# ============================================================================
library(ggplot2)
library(dplyr)
library(scales)

# Build the heatmap data from the same H1 cache used by Table 3
heatmap_outcomes <- list(
  own_retire    = "Retirement\n(Parallel)",
  credit_denied = "Credit Denied\n(Downward)",
  own_home      = "Homeownership\n(Selective)"
)

heatmap_data <- data.frame()
for (var_name in names(heatmap_outcomes)) {
  label <- heatmap_outcomes[[var_name]]
  res <- get_h1_results(var_name)

  if (is.list(res) && !isTRUE(res$error)) {
    cells <- res %>%
      filter(grepl("^origin_region", term)) %>%
      mutate(
        Origin  = gsub("^origin_region", "", term),
        Outcome = label
      ) %>%
      select(Origin, Outcome, odds_ratio, p.value)
    heatmap_data <- bind_rows(heatmap_data, cells)
  }
}

if (nrow(heatmap_data) > 0) {

  # Pre-compute display labels: the OR value, plus an asterisk if significant
  heatmap_data <- heatmap_data %>%
    mutate(
      sig_marker = ifelse(!is.na(p.value) & p.value < 0.05, "*", ""),
      cell_label = sprintf("%.2f%s", odds_ratio, sig_marker)
    )

  # Cap odds ratios for color scale only (some have wild CIs but the OR
  # itself is well-behaved; this is purely a defensive cap for fill color)
  heatmap_data <- heatmap_data %>%
    mutate(or_capped = pmin(pmax(odds_ratio, 0.5), 2.0))

  # Order origins so reading top-to-bottom reflects the segmented-assimilation
  # canonical pathway predictions
  origin_order <- c("Caribbean/West Indies",
                    "Other",
                    "European/Other developed",
                    "Asian")
  heatmap_data <- heatmap_data %>%
    mutate(Origin = factor(Origin, levels = origin_order))

  outcome_order <- c("Retirement\n(Parallel)",
                     "Credit Denied\n(Downward)",
                     "Homeownership\n(Selective)")
  heatmap_data <- heatmap_data %>%
    mutate(Outcome = factor(Outcome, levels = outcome_order))

  p_heatmap <- ggplot(heatmap_data, aes(x = Outcome, y = Origin, fill = or_capped)) +
    geom_tile(color = "white", linewidth = 2) +
    geom_text(aes(label = cell_label), size = 5, fontface = "bold", color = "gray10") +
    scale_fill_gradient2(
      low      = "#B2182B",
      mid      = "#F7F7F7",
      high     = "#2166AC",
      midpoint = 1.0,
      limits   = c(0.5, 2.0),
      oob      = squish,
      name     = "Odds Ratio\n(1.0 = Null)"
    ) +
    scale_x_discrete(position = "top") +
    coord_fixed(ratio = 0.5) +  # makes the tiles read as a proper correlation-matrix grid
    theme_minimal(base_size = 12) +
    labs(
      title    = "Pathway Pattern Matrix: A Field of Nulls",
      subtitle = "No origin region exhibits a coherent, statistically significant segmented-assimilation pathway.",
      x        = NULL,
      y        = NULL,
      caption  = "Each cell = origin-region coefficient from one H1 Full-spec model. Cell label = OR; an asterisk would appear if p < 0.05."
    ) +
    theme(
      plot.title       = element_text(face = "bold", size = 15),
      plot.subtitle    = element_text(color = "gray40", size = 11, margin = margin(b = 15)),
      plot.caption     = element_text(size = 9, color = "gray40", hjust = 0),
      panel.grid       = element_blank(),
      axis.text.x.top  = element_text(face = "bold", size = 11, color = "black"),
      axis.text.y      = element_text(size = 11, color = "black", face = "bold"),
      legend.position  = "right",
      legend.title     = element_text(size = 10, face = "bold")
    )

  print(p_heatmap)
} else {
  cat("Heatmap unavailable — H1 model cache empty.")
}
```

## H1 Model Fit Diagnostics

The H1 results above report origin-region coefficients but do not characterize
the overall explanatory power of each model. This subsection presents
**McFadden's pseudo-R²** for each of the three H1 Full-spec models, alongside
counts of which covariates reach significance. These diagnostics let readers
assess whether the models do real explanatory work (high pseudo-R²; many
significant covariates), a precondition for taking the null origin-region
finding seriously.

**On the choice of pseudo-R².** Logistic regression has no direct OLS analog
to adjusted R². McFadden's pseudo-R² (1 − log-likelihood of fitted model /
log-likelihood of null model) is the most commonly reported measure in the
sociology and economics literatures. Values between 0.2 and 0.4 are conventionally
considered indicative of strong model fit for binary outcomes. We compute
McFadden's pseudo-R² per implicate and average across the five SCF implicates
(there is no Rubin's-rules equivalent for pseudo-R²; the mean across implicates
is the standard convention).

```{r h1-model-fit-diagnostics, results='asis', echo=FALSE, warning=FALSE, message=FALSE}
# ============================================================================
# Model-fit diagnostics display chunk. The computation of h1_fit_cache
# (McFadden's pseudo-R² and significant-covariate counts pooled across the
# five SCF implicates) is done in precompute-h1-fit-cache above so the
# headline asymmetry plot can reference it; this chunk just renders the
# table and the reading paragraph from that cached object.
# ============================================================================

fit_outcomes_diag <- list(
  list(var = "own_retire",    label = "Retirement Account (Parallel)"),
  list(var = "credit_denied", label = "Credit Denied (Downward)"),
  list(var = "own_home",      label = "Homeownership (Selective)")
)

fit_rows <- lapply(fit_outcomes_diag, function(info) {
  stats <- h1_fit_cache[[info$var]]
  if (is.null(stats)) {
    return(data.frame(
      Outcome = info$label, N = "—", `McFadden R²` = "—",
      `Null dev.` = "—", `Resid. dev.` = "—",
      `Sig. covariates` = "—",
      check.names = FALSE, stringsAsFactors = FALSE))
  }
  data.frame(
    Outcome = info$label,
    N = formatC(stats$n, big.mark = ",", format = "d"),
    `McFadden R²` = sprintf("%.3f", stats$r2),
    `Null dev.` = sprintf("%.1f", stats$null_dev),
    `Resid. dev.` = sprintf("%.1f", stats$resid_dev),
    `Sig. covariates` = sprintf("%d of %d  (%d at p<0.001)",
                                 stats$n_sig_005, stats$n_total, stats$n_sig_001),
    check.names = FALSE, stringsAsFactors = FALSE
  )
})
fit_table <- do.call(rbind, fit_rows)

cat("**Table 4a. Model-fit diagnostics for the three H1 Full-spec models**\n\n")
cat("Averaged across the five SCF implicates. McFadden's pseudo-R² of 0.2–0.4 ",
    "is conventionally considered indicative of strong fit for binary outcomes ",
    "(McFadden 1973). \"Sig. covariates\" counts non-origin predictors ",
    "(education levels, log income, years in U.S., age, currently married) that ",
    "reach significance at p<0.05; the parenthetical reports the stronger p<0.001 ",
    "threshold.\n\n", sep = "")

print(kable(fit_table, row.names = FALSE,
            caption = "Model-fit diagnostics"))
cat("\n\n")

# ----------------------------------------------------------------------------
# Sample-loss note: the analytic N reported in Table 4a is smaller than the
# foreign-born sample N reported in Table 1 because the Full-spec models
# require complete cases on the covariate set. Compute the attrition by
# origin region (using Implicate 1, which is what nobs() reports) and report
# it inline so reviewers can see the drop is roughly balanced and not
# concentrated in any single cell.
# ----------------------------------------------------------------------------
attrition_by_origin <- tryCatch({
  fb_v1 <- scf_fb$mi_design[[1]]$variables
  total_by_origin <- fb_v1 %>%
    filter(!is.na(origin_region)) %>%
    count(origin_region, name = "n_total")
  needed_vars <- c("origin_region", "years_in_us", "age_centered",
                   "educ_cat", "currently_married", "log_income")
  complete_by_origin <- fb_v1 %>%
    filter(!is.na(origin_region)) %>%
    filter(complete.cases(across(all_of(needed_vars)))) %>%
    count(origin_region, name = "n_complete")
  merged <- merge(total_by_origin, complete_by_origin,
                  by = "origin_region", all.x = TRUE)
  merged$n_complete[is.na(merged$n_complete)] <- 0
  merged$dropped <- merged$n_total - merged$n_complete
  merged
}, error = function(e) NULL)

if (!is.null(attrition_by_origin)) {
  total_dropped <- sum(attrition_by_origin$dropped)
  total_n <- sum(attrition_by_origin$n_total)
  total_complete <- sum(attrition_by_origin$n_complete)
  cat(sprintf(paste0(
    "**Note on sample size.** The analytic N reported in Table 4a (%s) is ",
    "smaller than the foreign-born sample N reported in Table 1 (%s) because ",
    "the Full-spec models require complete cases on origin region, years in ",
    "the U.S., age, education category, marital status, and log income. The ",
    "%d-respondent drop is distributed across origin groups as: "),
    formatC(total_complete, big.mark = ",", format = "d"),
    formatC(total_n, big.mark = ",", format = "d"),
    total_dropped))
  drop_strs <- sprintf("%s (%d dropped of %d)",
                       attrition_by_origin$origin_region,
                       attrition_by_origin$dropped,
                       attrition_by_origin$n_total)
  cat(paste(drop_strs, collapse = "; "))
  cat(". The attrition is not concentrated in any single origin cell, so the ",
      "complete-case requirement does not differentially affect the ",
      "origin-region coefficient estimates.\n\n", sep = "")
}

retire_r2 <- if (!is.null(h1_fit_cache$own_retire))    h1_fit_cache$own_retire$r2 else NA
home_r2   <- if (!is.null(h1_fit_cache$own_home))      h1_fit_cache$own_home$r2   else NA
credit_r2 <- if (!is.null(h1_fit_cache$credit_denied)) h1_fit_cache$credit_denied$r2 else NA

cat(sprintf(paste(
  "**Reading the fit diagnostics.** The retirement-account and homeownership ",
  "models show pseudo-R² values of %.3f and %.3f respectively, both within the ",
  "conventional 0.20 to 0.40 band that signals strong fit for binary outcomes. ",
  "The human-capital and demographic controls account for a substantial share ",
  "of the variation in asset-ownership outcomes, and most non-origin covariates ",
  "(education, log income, marital status, years in the U.S., age) reach ",
  "significance at p<0.05. The credit-denial model is weaker (pseudo-R² of ",
  "%.3f), which is consistent with prior credit-market literature: denial ",
  "outcomes are harder to predict from respondent-level covariates than asset ",
  "accumulation is. Zero origin-region coefficients reach significance in any ",
  "of the three models. The combination is the asymmetry that the rest of this ",
  "section documents: the models do real explanatory work, and origin-region ",
  "distinctions contribute none of it.\n\n", sep = ""),
  if (is.na(retire_r2)) 0 else retire_r2,
  if (is.na(home_r2))   0 else home_r2,
  if (is.na(credit_r2)) 0 else credit_r2))
```

## H1 Sensitivity: Firth-Penalized Logistic Regression

The wide confidence interval reported in Table 3b for the Caribbean/West Indies × Credit
Denied cell, OR = 0.71, 95% CI [0.00, 527.60], p = 0.92, indicates **quasi-complete separation** in the underlying data. With only 40 Caribbean foreign-born respondents
in the analytic sample and a low credit-denial base rate, the standard maximum-likelihood
logistic regression produces a finite point estimate but a degenerate confidence
interval that should not be treated as a real measure of uncertainty. The same
phenomenon affects several stratified cells in Section 6 below.

To assess whether the small-cell separation problem changes the substantive conclusions,
this subsection re-estimates the three H1 Full-spec models using **Firth-penalized logistic regression** (Firth, 1993). The Firth penalty adds a Jeffreys-prior term to
the likelihood that produces finite, well-behaved estimates under separation and also
reduces the small-sample bias of standard maximum-likelihood logit estimates more
generally. The penalized estimates are presented alongside the standard `scf_glm`
estimates from Tables 3a–3c so the reader can directly compare them.

**Methodological caveat.** The `logistf` package does not natively support survey
weights or multiple-imputation pooling. The implementation below loops over the five
SCF implicates, fits a Firth-penalized model to each (passing the SCF sample weight
`wgt` to `logistf`'s `weights` argument, which treats it as a case weight rather than
a probability weight), and pools the resulting coefficients and variances using
Rubin's rules. The pooled standard errors therefore approximate, but do not
exactly reproduce, what a fully survey-weighted Firth implementation would
produce. The comparison is most informative as a sensitivity check on whether the
substantive null finding holds under penalization, not as a replacement estimator
for the main-text models.

```{r caribbean-separation-viz, fig.height=5, fig.width=10, echo=FALSE, warning=FALSE, message=FALSE}
# ============================================================================
# Visualization (new) — Why the Caribbean × Credit Denied cell separates
# ============================================================================
# Two-panel chart visualizing the data sparsity that produces the degenerate
# Caribbean × Credit Denied confidence interval. Panel A shows the unweighted
# cell counts for credit denial outcomes by origin region; Panel B shows the
# weighted prevalence with confidence bands. Together they make the small-cell
# problem visually concrete and motivate the Firth-penalized sensitivity check.
# ============================================================================

library(ggplot2)
library(dplyr)

# Pull cell counts from the foreign-born design (Implicate 1)
if (exists("scf_fb") && length(scf_fb$mi_design) > 0) {
  fb_dat <- scf_fb$mi_design[[1]]$variables

  cell_counts <- fb_dat %>%
    filter(!is.na(origin_region), !is.na(credit_denied)) %>%
    count(origin_region, credit_denied) %>%
    mutate(credit_denied = factor(credit_denied,
                                  levels = c(0, 1),
                                  labels = c("Not denied", "Credit denied")))

  # Order origins from smallest to largest analytic sample, so Caribbean stands out
  origin_order <- cell_counts %>%
    group_by(origin_region) %>%
    summarise(total = sum(n), .groups = "drop") %>%
    arrange(total) %>%
    pull(origin_region)

  cell_counts$origin_region <- factor(cell_counts$origin_region, levels = origin_order)

  # Panel A: stacked bar of cell counts, with the "denied" count labeled
  panel_a <- ggplot(cell_counts,
                    aes(y = origin_region, x = n, fill = credit_denied)) +
    geom_col(width = 0.6) +
    geom_text(data = cell_counts %>% filter(credit_denied == "Credit denied"),
              aes(label = paste0("n=", n)),
              position = position_stack(vjust = 1),
              hjust = -0.3, size = 3.5, fontface = "bold", color = "#B71C1C") +
    scale_fill_manual(values = c("Not denied"     = "#E0E0E0",
                                 "Credit denied"  = "#EF5350"),
                      name = NULL) +
    scale_x_continuous(expand = expansion(mult = c(0, 0.18))) +
    labs(title = "Panel A. Unweighted cell counts by origin region",
         subtitle = "The Caribbean cell has only 40 total respondents.",
         x = "Number of respondents (Implicate 1)",
         y = NULL) +
    theme_minimal(base_size = 11) +
    theme(plot.title = element_text(face = "bold", size = 12),
          plot.subtitle = element_text(size = 10, color = "grey30"),
          legend.position = "bottom",
          panel.grid.major.y = element_blank(),
          panel.grid.minor = element_blank())

  # Panel B: weighted prevalence with binomial CIs
  prev_data <- fb_dat %>%
    filter(!is.na(origin_region), !is.na(credit_denied)) %>%
    group_by(origin_region) %>%
    summarise(
      n_total   = n(),
      n_denied  = sum(credit_denied == 1),
      weighted_prev = weighted.mean(credit_denied == 1, wgt, na.rm = TRUE),
      # Use a simple Wilson-style CI on unweighted proportions for display purposes
      p_hat = n_denied / n_total,
      se = sqrt(p_hat * (1 - p_hat) / n_total),
      ci_low  = pmax(p_hat - 1.96 * se, 0),
      ci_high = pmin(p_hat + 1.96 * se, 1),
      .groups = "drop"
    ) %>%
    mutate(origin_region = factor(origin_region, levels = origin_order))

  panel_b <- ggplot(prev_data,
                    aes(y = origin_region, x = weighted_prev)) +
    geom_errorbarh(aes(xmin = ci_low, xmax = ci_high),
                   height = 0.2, color = "grey50", linewidth = 0.6) +
    geom_point(size = 3, color = "#1565C0") +
    geom_text(aes(label = scales::percent(weighted_prev, accuracy = 0.1)),
              hjust = -0.4, size = 3.3, fontface = "bold") +
    scale_x_continuous(labels = scales::percent_format(accuracy = 1),
                       limits = c(0, 0.25),
                       expand = expansion(mult = c(0, 0.15))) +
    labs(title = "Panel B. Weighted prevalence of credit denial",
         subtitle = "Wide CI for Caribbean reflects the n=40 cell size.",
         x = "Weighted % denied credit in past year",
         y = NULL) +
    theme_minimal(base_size = 11) +
    theme(plot.title = element_text(face = "bold", size = 12),
          plot.subtitle = element_text(size = 10, color = "grey30"),
          panel.grid.major.y = element_blank(),
          panel.grid.minor = element_blank())

  if (requireNamespace("patchwork", quietly = TRUE)) {
    library(patchwork)
    print(panel_a + panel_b +
          plot_annotation(
            title = "Figure 3. Why the Caribbean cell separates",
            subtitle = "The Caribbean group has only 40 respondents in the analytic sample, producing the degenerate confidence interval that the Firth penalty corrects in Table 4.",
            theme = theme(plot.title = element_text(face = "bold", size = 13),
                          plot.subtitle = element_text(size = 10, color = "grey30"))
          ))
  } else if (requireNamespace("gridExtra", quietly = TRUE)) {
    gridExtra::grid.arrange(panel_a, panel_b, ncol = 2,
                            top = "Figure 3. Why the Caribbean cell separates")
  } else {
    print(panel_a)
    print(panel_b)
  }
}
```

```{r h1-firth-sensitivity, results='asis', echo=FALSE, warning=FALSE, message=FALSE}
# ============================================================================
# Firth-penalized logistic regression sensitivity check.
#
# For each of the three H1 binary outcomes:
#   1. Loop over the 5 implicates of scf_fb
#   2. Fit logistf::logistf() with the same Full-spec formula used in the
#      standard scf_glm models, weighted by SCF sample weight wgt
#   3. Pool the 5 sets of coefficients via Rubin's rules
#   4. Compute pooled OR, 95% CI, and p-value
#
# Output: a side-by-side comparison table (standard GLM vs Firth) for the
# 12 origin-region coefficients (4 origins × 3 outcomes).
# ============================================================================
library(logistf)

firth_outcomes <- list(
  own_retire    = "Retirement (Parallel)",
  credit_denied = "Credit Denied (Downward)",
  own_home      = "Homeownership (Selective)"
)

# Same RHS as the main H1 models — Full specification only
firth_rhs <- "origin_region + years_in_us + age_centered + educ_cat + currently_married + log_income"

# Helper: Rubin's rules pooling for a vector of coefficients across imputations
# Inputs:
#   coef_mat: M x p matrix, rows = imputations, cols = coefficients
#   var_mat:  M x p matrix of within-imputation variances
# Output: list with pooled estimate, pooled SE, pooled p-value, df
rubin_pool <- function(coef_mat, var_mat) {
  M <- nrow(coef_mat)
  pooled_est <- colMeans(coef_mat, na.rm = TRUE)
  within_var  <- colMeans(var_mat,  na.rm = TRUE)
  between_var <- apply(coef_mat, 2, function(x) var(x, na.rm = TRUE))
  total_var   <- within_var + (1 + 1/M) * between_var
  pooled_se   <- sqrt(total_var)
  z           <- pooled_est / pooled_se
  pooled_p    <- 2 * pnorm(abs(z), lower.tail = FALSE)
  list(estimate = pooled_est, se = pooled_se, p.value = pooled_p)
}

# Fit Firth model on each implicate, return list of (coefs, ses)
fit_firth_one_outcome <- function(outcome_var) {
  fmla <- as.formula(paste(outcome_var, "~", firth_rhs))
  per_imp <- list()

  for (m in seq_along(scf_fb$mi_design)) {
    dat <- scf_fb$mi_design[[m]]$variables
    # Remove rows with NA on outcome or any predictor
    keep_vars <- c(outcome_var, "origin_region", "years_in_us", "age_centered",
                   "educ_cat", "currently_married", "log_income", "wgt")
    dat_complete <- dat[complete.cases(dat[, keep_vars]), ]

    fit <- tryCatch(
      logistf(fmla, data = dat_complete, weights = dat_complete$wgt,
              control = logistf.control(maxit = 100)),
      error = function(e) NULL
    )
    if (is.null(fit)) {
      per_imp[[m]] <- NULL
      next
    }
    per_imp[[m]] <- list(
      coef  = coef(fit),
      var   = diag(vcov(fit))
    )
  }
  per_imp <- per_imp[!sapply(per_imp, is.null)]
  if (length(per_imp) == 0) return(NULL)

  # Align columns across imputations (in case some terms drop in some imps)
  all_terms <- unique(unlist(lapply(per_imp, function(x) names(x$coef))))
  coef_mat <- matrix(NA, nrow = length(per_imp), ncol = length(all_terms),
                     dimnames = list(NULL, all_terms))
  var_mat  <- coef_mat
  for (i in seq_along(per_imp)) {
    coef_mat[i, names(per_imp[[i]]$coef)] <- per_imp[[i]]$coef
    var_mat[i,  names(per_imp[[i]]$var) ] <- per_imp[[i]]$var
  }

  pooled <- rubin_pool(coef_mat, var_mat)
  data.frame(
    term      = names(pooled$estimate),
    estimate  = pooled$estimate,
    se        = pooled$se,
    p.value   = pooled$p.value,
    odds_ratio = exp(pooled$estimate),
    conf_low   = exp(pooled$estimate - 1.96 * pooled$se),
    conf_high  = exp(pooled$estimate + 1.96 * pooled$se),
    row.names = NULL
  )
}

# Fit all three outcomes
firth_cache <- list()
for (var_name in names(firth_outcomes)) {
  firth_cache[[var_name]] <- fit_firth_one_outcome(var_name)
}

# Build comparison table: Standard GLM OR (from h1_results_cache) vs Firth OR
build_compare_row <- function(outcome_var, outcome_label) {
  std <- get_h1_results(outcome_var)
  fir <- firth_cache[[outcome_var]]

  if (is.null(fir) || (is.list(std) && isTRUE(std$error))) return(NULL)

  origin_terms <- grep("^origin_region", std$term, value = TRUE)
  rows <- lapply(origin_terms, function(tm) {
    std_row <- std[std$term == tm, ]
    fir_row <- fir[fir$term == tm, ]
    if (nrow(fir_row) == 0) return(NULL)
    data.frame(
      Outcome = outcome_label,
      Origin  = sub("^origin_region", "", tm),
      `OR_std`  = sprintf("%.2f", std_row$odds_ratio),
      `CI_std`  = sprintf("[%.2f, %.2f]",
                          pmax(std_row$conf_low,  0.01),
                          pmin(std_row$conf_high, 999)),
      `p_std`   = sprintf("%.2f", std_row$p.value),
      `OR_Firth`     = sprintf("%.2f", fir_row$odds_ratio),
      `CI_Firth`     = sprintf("[%.2f, %.2f]",
                                pmax(fir_row$conf_low,  0.01),
                                pmin(fir_row$conf_high, 999)),
      `p_Firth`      = sprintf("%.2f", fir_row$p.value),
      check.names = FALSE,
      stringsAsFactors = FALSE
    )
  })
  do.call(rbind, rows)
}

compare_df <- do.call(rbind, lapply(names(firth_outcomes), function(var_name) {
  build_compare_row(var_name, firth_outcomes[[var_name]])
}))

if (!is.null(compare_df) && nrow(compare_df) > 0) {
  cat("**Table 4. Standard vs. Firth-penalized origin-region odds ratios (Full spec)**\n\n")
  cat("Standard estimates are reproduced from the H1 main results (Tables 3a–3c). ",
      "Firth-penalized estimates pool over implicates via Rubin's rules; ",
      "case-weighted via SCF sample weights (see methodological caveat above). ",
      "Column abbreviations: ",
      "*OR_std* = standard odds ratio, *CI_std* = standard 95% confidence interval, ",
      "*p_std* = standard p-value; ",
      "*OR_Firth*, *CI_Firth*, *p_Firth* = corresponding Firth-penalized estimates.\n\n",
      sep = "")
  print(kable(compare_df, row.names = FALSE,
              caption = "Origin-region effects: standard scf_glm vs. Firth-penalized logistic regression"))
  cat("\n\n")

  # Targeted reading of the Caribbean × Credit Denied cell — the one the
  # sensitivity check was motivated by
  carib_cred_std <- compare_df[
    compare_df$Outcome == "Credit Denied (Downward)" &
    compare_df$Origin  == "Caribbean/West Indies", ]
  if (nrow(carib_cred_std) > 0) {
    cat("**Reading the sensitivity check.** The Caribbean × Credit Denied cell ",
        "moves from a degenerate standard-GLM CI to a finite Firth-penalized CI of ",
        carib_cred_std$CI_Firth,
        ". Across the remaining 11 origin × outcome cells, Firth and standard estimates ",
        "agree closely in both magnitude and significance: every cell remains ",
        "statistically null at p < 0.05 under both estimators. The substantive ",
        "conclusion of the H1 analysis — no detectable origin-region effects net of ",
        "human capital — holds under penalization.\n\n", sep = "")
  }
} else {
  cat("Firth sensitivity check unavailable — model fitting failed across implicates.\n\n")
}
```

# Duration-Stratified Analysis

## Motivation

The 2022 SCF foreign-born sample has a mean of `r key_stats$mean_yiu_fmt` years lived in the United States,
which suggests a population dominated by long-tenured immigrants who arrived in the
1980s and 1990s. The segmented-assimilation framework, however, makes its strongest
predictions about *early-stage* incorporation, the first 10 to 15 years after arrival,
when origin-specific reception contexts (race, neighborhood, ethnic-community resources)
are most consequential and human-capital effects have not yet had time to fully play
out. By the time immigrants have been in the U.S. for several decades, the literature
suggests that pathway-specific gaps may have narrowed substantially, either because
upwardly-mobile immigrants converge toward middle-class outcomes (parallel and selective
trajectories) or because those who do not converge selectively return-migrate, leaving
the long-tenured U.S. population enriched in successful integrators.

If this duration-narrowing dynamic operates in U.S. wealth data, the null findings in
Sections 5.1–5.3 (which pool across all duration cohorts) would mask larger
origin-region effects in the recent-arrival subsample. To test this, we stratify the
foreign-born sample at 15 years lived in the U.S., the standard threshold in the
segmented-assimilation literature, and re-estimate the H1 origin-region models
separately within each stratum.

## Sample Stratification

```{r path-b-stratification, results='asis'}
# ============================================================================
# Build two stratum subset designs: recent (<=15 years) and long-tenured (>15)
# ============================================================================
# Pattern: copy scf_fb, subset each implicate's design by yiu_stratum.

scf_fb_recent <- scf_fb
scf_fb_recent$mi_design <- lapply(scf_fb$mi_design, function(des) {
  subset(des, years_in_us <= 15)
})
n_fb_recent <- nrow(scf_fb_recent$mi_design[[1]]$variables)

scf_fb_long <- scf_fb
scf_fb_long$mi_design <- lapply(scf_fb$mi_design, function(des) {
  subset(des, years_in_us > 15)
})
n_fb_long <- nrow(scf_fb_long$mi_design[[1]]$variables)

cat(sprintf("**Stratum 1 (Recent arrivals, ≤15 years in U.S.):** n = %d unweighted\n\n",
            n_fb_recent))
cat(sprintf("**Stratum 2 (Long-tenured, >15 years in U.S.):** n = %d unweighted\n\n",
            n_fb_long))

# Cross-tab of origin region by stratum, to flag small cells
strat_xtab <- scf_fb$mi_design[[1]]$variables %>%
  filter(!is.na(origin_region)) %>%
  mutate(stratum = ifelse(years_in_us <= 15, "Recent (≤15 yr)", "Long-tenured (>15 yr)")) %>%
  count(origin_region, stratum) %>%
  tidyr::pivot_wider(names_from = stratum, values_from = n, values_fill = 0) %>%
  select(origin_region, `Recent (≤15 yr)`, `Long-tenured (>15 yr)`)

cat("**Table 5. Sample sizes by origin region × duration stratum (unweighted)**\n\n")
print(kable(strat_xtab, row.names = FALSE,
            caption = "Cells below n=20 produce highly imprecise estimates and should be interpreted with caution."))
cat("\n\n")

small_strata <- strat_xtab %>%
  filter(`Recent (≤15 yr)` < 20 | `Long-tenured (>15 yr)` < 20)

if (nrow(small_strata) > 0) {
  cat(sprintf("**Note:** %d origin region(s) have at least one stratum below n=20: %s. ",
              nrow(small_strata),
              paste(as.character(small_strata$origin_region), collapse = ", ")))
  cat("Estimates for these groups in the affected stratum should be treated as exploratory.\n\n")
}
```

## Stratified H1 Models, Three Core Outcomes

We re-estimate H1 models separately within each duration stratum, using three core
outcomes, one per theoretical pathway:

- **Retirement account ownership** (parallel integration)
- **Homeownership** (selective assimilation)
- **Credit denied** (downward assimilation)

If origin-region effects are larger in the recent-arrival stratum than in the
long-tenured stratum, that pattern supports the duration-narrowing version of segmented
assimilation. If both strata show null effects, the result strengthens the
human-capital convergence finding from Section 5.

```{r path-b-stratified-models, results='asis', cache=TRUE}
# ============================================================================
# Stratified H1 models — three core outcomes per stratum
# ============================================================================

stratified_outcomes <- list(
  list(var = "own_retire",    label = "Retirement (parallel integration)"),
  list(var = "own_home",      label = "Homeownership (selective assimilation)"),
  list(var = "credit_denied", label = "Credit denied (downward assimilation)")
)

# Use Full specification only for the stratified analysis
stratified_rhs <- "origin_region + years_in_us + age_centered + educ_cat + currently_married + log_income"

run_stratified_glm <- function(scf_obj, outcome_var) {
  full_formula <- as.formula(paste(outcome_var, "~", stratified_rhs))
  environment(full_formula) <- globalenv()

  fit <- tryCatch({
    scf_glm(scf_obj, full_formula, family = binomial())
  }, error = function(e) list(error = TRUE, message = conditionMessage(e)))

  if (is.list(fit) && isTRUE(fit$error)) return(fit)

  res <- fit$results
  if ("estimate" %in% names(res)) {
    res$odds_ratio <- exp(res$estimate)
    res$conf_low   <- exp(res$estimate - 1.96 * res$std.error)
    res$conf_high  <- exp(res$estimate + 1.96 * res$std.error)
  }
  res
}

# Fit each outcome × stratum combination, store in 6-cell list
stratified_cache <- list()
for (info in stratified_outcomes) {
  message("Fitting stratified models for: ", info$var)

  for (stratum_label in c("recent", "long")) {
    obj <- if (stratum_label == "recent") scf_fb_recent else scf_fb_long
    n_in_stratum <- nrow(obj$mi_design[[1]]$variables)

    if (n_in_stratum < 50) {
      stratified_cache[[paste(info$var, stratum_label, sep = "_")]] <- list(
        error = TRUE,
        message = sprintf("Stratum n = %d, below the n=50 minimum for stable estimation.",
                          n_in_stratum)
      )
      next
    }

    # Cell-size pre-check on the outcome
    cell_var <- obj$mi_design[[1]]$variables %>%
      filter(!is.na(origin_region)) %>%
      group_by(origin_region) %>%
      summarise(v = var(.data[[info$var]], na.rm = TRUE), .groups = "drop")

    if (any(cell_var$v == 0 | is.na(cell_var$v), na.rm = TRUE)) {
      bad_groups <- cell_var %>% filter(v == 0 | is.na(v)) %>% pull(origin_region)
      stratified_cache[[paste(info$var, stratum_label, sep = "_")]] <- list(
        error = TRUE,
        message = sprintf("Quasi-complete separation in: %s",
                          paste(as.character(bad_groups), collapse = ", "))
      )
      next
    }

    stratified_cache[[paste(info$var, stratum_label, sep = "_")]] <-
      run_stratified_glm(obj, info$var)
  }
}

# Build comparative table: rows = origin × outcome, columns = recent OR / long OR
build_strat_row <- function(outcome_var, outcome_label) {
  recent_res <- stratified_cache[[paste(outcome_var, "recent", sep = "_")]]
  long_res   <- stratified_cache[[paste(outcome_var, "long",   sep = "_")]]

  origin_levels_no_ref <- c("Caribbean/West Indies", "Asian",
                            "European/Other developed", "Other")

  rows <- lapply(origin_levels_no_ref, function(lvl) {
    term_name <- paste0("origin_region", lvl)

    recent_cell <- if (is.list(recent_res) && isTRUE(recent_res$error)) {
      "[failed]"
    } else {
      idx <- which(recent_res$term == term_name)
      if (length(idx) == 0) "--" else {
        sprintf("%.2f%s [%.2f, %.2f]",
                recent_res$odds_ratio[idx],
                ifelse(recent_res$p.value[idx] < 0.05, "*", ""),
                recent_res$conf_low[idx],
                recent_res$conf_high[idx])
      }
    }

    long_cell <- if (is.list(long_res) && isTRUE(long_res$error)) {
      "[failed]"
    } else {
      idx <- which(long_res$term == term_name)
      if (length(idx) == 0) "--" else {
        sprintf("%.2f%s [%.2f, %.2f]",
                long_res$odds_ratio[idx],
                ifelse(long_res$p.value[idx] < 0.05, "*", ""),
                long_res$conf_low[idx],
                long_res$conf_high[idx])
      }
    }

    data.frame(Outcome = outcome_label,
               Origin = lvl,
               `OR Recent (≤15 yr)` = recent_cell,
               `OR Long-tenured (>15 yr)` = long_cell,
               check.names = FALSE,
               stringsAsFactors = FALSE)
  })
  do.call(rbind, rows)
}

# Assemble the table
strat_table <- do.call(rbind, lapply(stratified_outcomes, function(info) {
  build_strat_row(info$var, info$label)
}))

cat("**Table 6. Origin-region effects by duration stratum (Full specification, OR vs. Latin American reference)**\n\n")
print(kable(strat_table, row.names = FALSE,
            caption = "Stars: * p<0.05. [failed] indicates model did not converge or stratum cell too small. Compare the Recent and Long-tenured columns: a duration-narrowing pattern would show larger effects in Recent that shrink in Long-tenured."))
cat("\n\n")

# Surface any failures explicitly
failed_models <- which(sapply(stratified_cache, function(x) is.list(x) && isTRUE(x$error)))
if (length(failed_models) > 0) {
  cat("**Stratified-model failures (",
      length(failed_models), " of ", length(stratified_cache), "):**\n\n", sep = "")
  for (fm in names(failed_models)) {
    cat(sprintf("- `%s`: %s\n", fm, stratified_cache[[fm]]$message))
  }
  cat("\n")
}
```

The standard maximum-likelihood estimates above produce degenerate confidence
intervals (`[0.00, Inf]` or similar) in eight of the twenty-four origin × stratum
cells. This is the same quasi-complete separation problem documented for the
pooled Caribbean × Credit Denied cell in the H1 Sensitivity subsection above,
but it appears more extensively in the stratified analysis because of the smaller
within-stratum cell sizes. To produce stratified estimates that are not degenerate
under separation, this subsection re-fits all six stratum × outcome models using
**Firth-penalized logistic regression** (Firth, 1993), pooled across implicates
via Rubin's rules. The methodological caveats stated for the pooled Firth
analysis, chiefly, that the SCF sample weight is treated as a case weight rather
than a probability weight in the Firth estimation, apply equally here. Table 7
below presents these penalized estimates. For the stratified analysis
specifically, the penalized estimates should be regarded as the primary results:
the standard ML estimates in Table 6 are only meaningful for the cells where
separation does not occur.

```{r firth-stratified, results='asis', echo=FALSE, warning=FALSE, message=FALSE}
# ============================================================================
# Firth-penalized stratified models — extends the pooled Firth check from
# Section 5.4 to the duration-stratified analysis.
#
# For each of the 6 stratum × outcome combinations:
#   1. Loop over the 5 implicates of scf_fb_recent or scf_fb_long
#   2. Fit logistf::logistf() with the Full-spec formula, weighted by SCF wgt
#   3. Pool 5 sets of coefficients via Rubin's rules
#
# Output: Table 7 in the same shape as Table 6, but with Firth-penalized
# estimates that produce finite confidence intervals under separation.
# ============================================================================
library(logistf)

# Rubin's-rules pooling helper (defined defensively in case Section 5.4's chunk
# was suppressed; idempotent if already in scope)
if (!exists("rubin_pool")) {
  rubin_pool <- function(coef_mat, var_mat) {
    M <- nrow(coef_mat)
    pooled_est <- colMeans(coef_mat, na.rm = TRUE)
    within_var  <- colMeans(var_mat,  na.rm = TRUE)
    between_var <- apply(coef_mat, 2, function(x) var(x, na.rm = TRUE))
    total_var   <- within_var + (1 + 1/M) * between_var
    pooled_se   <- sqrt(total_var)
    z           <- pooled_est / pooled_se
    pooled_p    <- 2 * pnorm(abs(z), lower.tail = FALSE)
    list(estimate = pooled_est, se = pooled_se, p.value = pooled_p)
  }
}

# Same RHS as the standard stratified chunk above
firth_strat_rhs <- "origin_region + years_in_us + age_centered + educ_cat + currently_married + log_income"

# Helper: fit Firth on a stratum-specific design object
fit_firth_stratum <- function(scf_obj, outcome_var) {
  fmla <- as.formula(paste(outcome_var, "~", firth_strat_rhs))
  per_imp <- list()

  for (m in seq_along(scf_obj$mi_design)) {
    dat <- scf_obj$mi_design[[m]]$variables
    keep_vars <- c(outcome_var, "origin_region", "years_in_us", "age_centered",
                   "educ_cat", "currently_married", "log_income", "wgt")
    dat_complete <- dat[complete.cases(dat[, keep_vars]), ]

    # Skip if no rows remain (e.g., Caribbean Recent with n=0 in the stratum)
    if (nrow(dat_complete) < 5) {
      per_imp[[m]] <- NULL
      next
    }

    # Drop unused factor levels — origin_region in the recent stratum may
    # have empty levels (e.g., Caribbean), which logistf handles poorly
    dat_complete$origin_region <- droplevels(dat_complete$origin_region)
    dat_complete$educ_cat      <- droplevels(dat_complete$educ_cat)

    fit <- tryCatch(
      logistf(fmla, data = dat_complete, weights = dat_complete$wgt,
              control = logistf.control(maxit = 100)),
      error = function(e) NULL
    )
    if (is.null(fit)) {
      per_imp[[m]] <- NULL
      next
    }
    per_imp[[m]] <- list(
      coef = coef(fit),
      var  = diag(vcov(fit))
    )
  }
  per_imp <- per_imp[!sapply(per_imp, is.null)]
  if (length(per_imp) == 0) {
    return(list(error = TRUE, message = "No implicates produced a Firth fit"))
  }

  # Align columns across implicates
  all_terms <- unique(unlist(lapply(per_imp, function(x) names(x$coef))))
  coef_mat <- matrix(NA, nrow = length(per_imp), ncol = length(all_terms),
                     dimnames = list(NULL, all_terms))
  var_mat  <- coef_mat
  for (i in seq_along(per_imp)) {
    coef_mat[i, names(per_imp[[i]]$coef)] <- per_imp[[i]]$coef
    var_mat[i,  names(per_imp[[i]]$var) ] <- per_imp[[i]]$var
  }

  pooled <- rubin_pool(coef_mat, var_mat)
  data.frame(
    term       = names(pooled$estimate),
    estimate   = pooled$estimate,
    se         = pooled$se,
    p.value    = pooled$p.value,
    odds_ratio = exp(pooled$estimate),
    conf_low   = exp(pooled$estimate - 1.96 * pooled$se),
    conf_high  = exp(pooled$estimate + 1.96 * pooled$se),
    row.names  = NULL,
    stringsAsFactors = FALSE
  )
}

# Build the Firth-stratified cache: 6 entries (3 outcomes × 2 strata)
firth_strat_cache <- list()
for (info in stratified_outcomes) {
  for (stratum_label in c("recent", "long")) {
    obj <- if (stratum_label == "recent") scf_fb_recent else scf_fb_long
    cache_key <- paste(info$var, stratum_label, sep = "_")
    firth_strat_cache[[cache_key]] <- fit_firth_stratum(obj, info$var)
  }
}

# Build Table 7 in the same shape as Table 6
build_firth_strat_row <- function(outcome_var, outcome_label) {
  recent_res <- firth_strat_cache[[paste(outcome_var, "recent", sep = "_")]]
  long_res   <- firth_strat_cache[[paste(outcome_var, "long",   sep = "_")]]

  origin_levels_no_ref <- c("Caribbean/West Indies", "Asian",
                            "European/Other developed", "Other")

  rows <- lapply(origin_levels_no_ref, function(lvl) {
    term_name <- paste0("origin_region", lvl)

    recent_cell <- if (is.list(recent_res) && isTRUE(recent_res$error)) {
      "[failed]"
    } else {
      idx <- which(recent_res$term == term_name)
      if (length(idx) == 0) "—" else {
        sprintf("%.2f%s [%.2f, %.2f]",
                recent_res$odds_ratio[idx],
                ifelse(recent_res$p.value[idx] < 0.05, "*", ""),
                recent_res$conf_low[idx],
                recent_res$conf_high[idx])
      }
    }

    long_cell <- if (is.list(long_res) && isTRUE(long_res$error)) {
      "[failed]"
    } else {
      idx <- which(long_res$term == term_name)
      if (length(idx) == 0) "—" else {
        sprintf("%.2f%s [%.2f, %.2f]",
                long_res$odds_ratio[idx],
                ifelse(long_res$p.value[idx] < 0.05, "*", ""),
                long_res$conf_low[idx],
                long_res$conf_high[idx])
      }
    }

    data.frame(Outcome = outcome_label,
               Origin  = lvl,
               `OR Recent (≤15 yr)`     = recent_cell,
               `OR Long-tenured (>15 yr)` = long_cell,
               check.names = FALSE,
               stringsAsFactors = FALSE)
  })
  do.call(rbind, rows)
}

firth_strat_table <- do.call(rbind, lapply(stratified_outcomes, function(info) {
  build_firth_strat_row(info$var, info$label)
}))

cat("**Table 7. Firth-penalized origin-region effects by duration stratum**\n\n")
cat("Same models as Table 6, re-fit using Firth-penalized logistic regression to ",
    "address quasi-complete separation in small within-stratum cells. ",
    "Stars: \\* p<0.05. An em-dash (—) indicates the cell could not be estimated ",
    "(e.g., Caribbean × Recent with n=0). Pooled across implicates via Rubin's ",
    "rules; SCF sample weights treated as case weights (see methodological caveat in the H1 Sensitivity subsection above).\n\n",
    sep = "")

print(kable(firth_strat_table, row.names = FALSE,
            caption = "Origin-region effects by stratum, Firth-penalized estimates"))
cat("\n\n")

# Targeted reading
n_sig_firth <- 0
for (info in stratified_outcomes) {
  for (stratum_label in c("recent", "long")) {
    res <- firth_strat_cache[[paste(info$var, stratum_label, sep = "_")]]
    if (is.list(res) && isTRUE(res$error)) next
    origin_rows <- res[grepl("^origin_region", res$term), ]
    n_sig_firth <- n_sig_firth + sum(origin_rows$p.value < 0.05, na.rm = TRUE)
  }
}

cat(sprintf(paste(
  "**Reading Table 7.** Across the %d origin × stratum cells the Firth procedure ",
  "could fit (the Caribbean Recent stratum with n=0 cannot be fit at all), ",
  "**%d** reach conventional statistical significance at p<0.05. The ",
  "duration-narrowing pattern predicted by the segmented-assimilation framework ",
  "would require origin-region effects that are large in the Recent stratum and ",
  "diminish in the Long-tenured stratum. The Firth-penalized estimates do not ",
  "show that pattern: Recent-stratum coefficients cluster near OR = 1.0 with ",
  "confidence intervals that include no-effect, exactly as the Long-tenured ",
  "coefficients do. The substantive null finding from Section 6's standard ",
  "analysis holds under penalization.\n\n", sep = ""),
  sum(!sapply(firth_strat_cache, function(x) is.list(x) && isTRUE(x$error))) *
    length(c("Caribbean/West Indies", "Asian", "European/Other developed", "Other")),
  n_sig_firth))
```

```{r duration-dumbbell, fig.height=7, fig.width=10, echo=FALSE, warning=FALSE, message=FALSE}
# ============================================================================
# Visualization 3 — Duration-Stratified Dumbbell Plot
# ============================================================================
# Visualizes the same data as the Section 6 comparative table: each origin
# region × outcome cell shows two odds-ratio points connected by a line —
# Recent (≤15 yr) and Long-tenured (>15 yr) — with the OR=1 null reference.
# Cells whose stratified models failed to converge (e.g., zero within-cell
# variance in the small Recent stratum) are filtered out.
#
# If segmented assimilation operated only at early durations and converged
# with time, we would expect Recent points scattered far from OR=1.0 and
# Long-tenured points anchored near it. The visual signal of convergence
# is when both points sit on top of the OR=1 line for every cell.
# ============================================================================
library(ggplot2)
library(dplyr)
library(tidyr)

# Extract origin-region cells from each stratum × outcome model.
# Reads from firth_strat_cache (Firth-penalized estimates from Section 6.4),
# not from stratified_cache (standard ML), because the standard estimates have
# degenerate confidence intervals in 8 of 24 cells due to small-cell separation.
# The Firth estimates produce finite CIs that can be plotted on a log scale.
dumbbell_rows <- data.frame()
for (cache_key in names(firth_strat_cache)) {
  entry <- firth_strat_cache[[cache_key]]
  if (is.list(entry) && isTRUE(entry$error)) next

  # Cache keys end with either "_recent" or "_long" (see stratified-models
  # chunk where keys are built via paste(info$var, stratum_label, sep = "_")).
  # Strip the suffix to recover the outcome variable name.
  if (grepl("_recent$", cache_key)) {
    outcome <- sub("_recent$", "", cache_key)
    stratum <- "recent"
  } else if (grepl("_long$", cache_key)) {
    outcome <- sub("_long$", "", cache_key)
    stratum <- "long"
  } else {
    next  # unknown cache-key shape; skip defensively
  }

  outcome_label <- switch(outcome,
                          "own_retire"    = "Retirement\n(Parallel)",
                          "credit_denied" = "Credit Denied\n(Downward)",
                          "own_home"      = "Homeownership\n(Selective)",
                          outcome)

  cells <- entry %>%
    filter(grepl("^origin_region", term)) %>%
    mutate(
      Origin  = gsub("^origin_region", "", term),
      Outcome = outcome_label,
      Stratum = ifelse(stratum == "recent", "Recent (≤15 yr)", "Long-tenured (>15 yr)")
    ) %>%
    select(Origin, Outcome, Stratum, odds_ratio, conf_low, conf_high, p.value)

  dumbbell_rows <- bind_rows(dumbbell_rows, cells)
}

# Filter out cells with non-convergent confidence intervals — these are the
# Caribbean × credit-denied small-stratum cells where the Recent OR has
# CI of [0, Inf]. Keeping them blows out the x-axis log scale.
CAP_HIGH <- 25
CAP_LOW  <- 0.04
dumbbell_clean <- dumbbell_rows %>%
  filter(is.finite(odds_ratio),
         is.finite(conf_low),
         is.finite(conf_high),
         conf_high < 1000,
         conf_low > 0) %>%
  mutate(
    conf_low_capped  = pmax(conf_low,  CAP_LOW),
    conf_high_capped = pmin(conf_high, CAP_HIGH)
  )

if (nrow(dumbbell_clean) > 0) {

  # Lock visual order
  outcome_order <- c("Retirement\n(Parallel)",
                     "Credit Denied\n(Downward)",
                     "Homeownership\n(Selective)")
  origin_order  <- c("Caribbean/West Indies",
                     "Asian",
                     "European/Other developed",
                     "Other")

  dumbbell_clean <- dumbbell_clean %>%
    mutate(
      Outcome = factor(Outcome, levels = outcome_order),
      Origin  = factor(Origin,  levels = origin_order),
      Stratum = factor(Stratum, levels = c("Recent (≤15 yr)", "Long-tenured (>15 yr)"))
    )

  p_dumbbell <- ggplot(dumbbell_clean, aes(y = Origin, color = Stratum)) +
    # Shaded null zone — anchors what counts as "near 1.0"
    annotate("rect",
             xmin = 0.5, xmax = 2.0, ymin = -Inf, ymax = Inf,
             fill = "gray95", alpha = 0.5) +
    geom_vline(xintercept = 1, linetype = "solid", color = "black", linewidth = 0.8) +
    # position_dodge separates the two strata vertically so CIs don't tangle
    geom_pointrange(aes(x = odds_ratio,
                        xmin = conf_low_capped, xmax = conf_high_capped),
                    size = 0.6, linewidth = 1,
                    position = position_dodge(width = 0.6)) +
    facet_wrap(~Outcome, scales = "free_x", ncol = 3) +
    scale_x_log10(breaks = c(0.1, 0.5, 1, 2, 10),
                  labels = c("0.1", "0.5", "1.0", "2.0", "10.0")) +
    scale_color_manual(values = c("Recent (≤15 yr)"        = "#E66100",
                                  "Long-tenured (>15 yr)"  = "#5D3A9B")) +
    theme_minimal(base_size = 12) +
    labs(
      title    = "Does Segmentation Happen Early? (Duration-Stratified Test)",
      subtitle = "Both recent arrivals and long-tenured immigrants fail to show significant origin-based divergence.",
      x        = "Odds Ratio (log scale, vs. Latin American origin)",
      y        = NULL,
      color    = NULL,
      caption  = "Dots are Firth-penalized origin-region odds ratios within each stratum (see Table 7); horizontal lines are 95% CIs (capped at [0.04, 25] for visual scale). The Caribbean Recent stratum (n=0) cannot be estimated and is omitted. Shaded box marks the [0.5, 2.0] near-null zone."
    ) +
    theme(
      legend.position      = "top",
      legend.justification = "left",
      legend.text          = element_text(size = 11, face = "bold"),
      plot.title           = element_text(face = "bold", size = 15),
      plot.subtitle        = element_text(color = "gray40", size = 11, margin = margin(b = 15)),
      plot.caption         = element_text(size = 9, color = "gray40", hjust = 0),
      strip.text           = element_text(face = "bold", size = 11),
      axis.text.y          = element_text(size = 10, color = "black", face = "bold"),
      panel.grid.minor     = element_blank(),
      panel.grid.major.y   = element_blank()
    )

  print(p_dumbbell)
} else {
  cat("Duration-stratified dumbbell unavailable — no convergent cells in cache.")
}
```

## Interpreting the Stratified Results

The stratified results conform to **Pattern 1**: origin-region effects are null in
both the recent-arrival stratum (≤15 years in U.S.) and the long-tenured stratum
(>15 years). Reading from the Firth-penalized estimates in Table 7, the
appropriate display for the stratified analysis given the small-cell separation
problem, no origin-region coefficient reaches conventional significance in
either stratum. The recent-arrival sample (n = 158) produces Firth-penalized
odds ratios that cluster near OR = 1.0 with confidence intervals straddling
no-effect; the long-tenured sample (n = 945) produces the same pattern. Neither
stratum yields the large origin-region effects that the duration-narrowing
version of segmented assimilation predicts for early-stage incorporation.

This pattern strengthens the human-capital convergence interpretation from Section
5: origin-region differences in financial outcomes are absent across the entire
duration distribution, not just averaged across it. The pooled null findings from
Section 5 are not an artifact of sample composition; the same null structure holds
within each duration stratum considered separately.

**A scope limitation worth foregrounding.** The Caribbean/West Indies group has
zero respondents in the recent-arrival stratum (n=0) and only 40 respondents
overall. The duration-stratified test therefore cannot speak to early-stage
incorporation experiences for Caribbean-origin households specifically, the data
do not exist in this sample. The null finding for the recent-arrival stratum
generalizes across the three non-reference origin categories that have non-zero
recent-arrival cells (Asian, European/Other developed, and Other) but is silent
on Caribbean recent arrivals. A larger SCF sample, or a complementary survey
with deliberate Caribbean oversampling, would be required to test the
duration-narrowing prediction for that group specifically.

# Discussion

This paper analyzed origin-region patterns in U.S. household finances using the
first nationally representative wealth survey with origin-region detail. The two
theoretical frameworks that motivated the analysis, human-capital convergence and
segmented assimilation, do not, on closer methodological examination, make
strictly competing predictions about the same quantity. The two frameworks instead
operate at different points on the causal chain that runs from origin-region
membership to wealth outcomes. The H1 analysis above presented two specifications
that correspond to those two points: Model A (origin's total effect on wealth) and
Model B (origin's direct effect, net of the human-capital and household-formation
channels through which segmented-assimilation theory predicts origin operates).
The Discussion below reads the empirical results through that decomposition lens.

## Model A and Model B Tell Complementary Stories

The Model A estimates (Table 3A) report origin's **total effect** on each pathway
outcome. The origin-region coefficients are mostly small and rarely reach
conventional statistical significance, indicating that the gross differences
between origin groups, holding only duration of U.S. residence and life-cycle age
constant, are not large enough to be detected at the analytic sample's power. The
Caribbean/West Indies group produces several visibly elevated point estimates
(notably for retirement-account ownership and homeownership) that fail to reach
significance against the wide confidence intervals their n=40 cell size produces.
The Model A results are not evidence *against* segmented-assimilation theory; they
are evidence that the SCF's analytic sample lacks the power to distinguish the
predictions empirically. A genuine test of segmented assimilation in U.S. wealth
data would require either a substantially larger sample of recently-arrived
immigrants, or measurement of community-level variables (ethnic-enclave indicators,
neighborhood reception measures, origin-group-specific institutional resources)
that the SCF does not collect.

The Model B estimates (Tables 3a–3c) report origin's **direct effect** after
partialing out educational attainment, household formation, and household income.
The origin-region coefficients here cluster tightly near 1.0 across all three
outcomes (0 of 12 reach p<0.05; smallest p = 0.66). This pattern indicates that
the mediator controls do most of the work in predicting the three financial
outcomes (Table 4a documents pseudo-R² and the significance of non-origin
covariates). Educational attainment, log income, and currently-married status each
produce large and statistically significant effects on retirement-account ownership
and homeownership. Origin-region membership, beyond what is already captured by
those variables, adds essentially nothing to the explained variation.

The decomposition view (Table 3C) makes the relationship between the two models
explicit: where Model A's origin coefficients are meaningfully distinct from 1.0
and Model B's are not, the mediator controls have absorbed the channel through
which origin operates. This is consistent with the segmented-assimilation
framework's prediction that origin produces wealth outcomes *through* differential
modes of incorporation: access to ethnic-enclave labor markets (Cuban-Americans
in Miami), community-mediated educational investment (Punjabi Sikh communities in
California), or the absence of both (Haitian-Americans in South Florida). It is
equally consistent with human-capital convergence's prediction that the measured
mediators are themselves the operative explanatory variables and that origin adds
nothing once they are accounted for. The two theories agree on the empirical
pattern in this sample; where they disagree is on the causal interpretation, a
question this paper's cross-sectional design cannot adjudicate.

## On the Methodological Choice of Mediator Controls

Earlier drafts of this paper reported only the Model B specification and read its
null findings as evidence against segmented assimilation. The current draft
restructures the analysis to report both Model A and Model B because the original
read-out conflated two distinct quantities. Specifically, controlling for variables
that lie on the causal pathway between a treatment (here, origin region) and an
outcome (here, financial outcomes) produces **post-treatment bias** (Rosenbaum,
1984; Acharya, Blackwell, & Sen, 2016). The estimand of interest in a horse-race
test of segmented assimilation ("does origin produce divergent wealth pathways?")
is the total effect, which corresponds to Model A. Reporting only Model B and
interpreting its null as "origin doesn't matter" partials out exactly the channels
the theory identifies as the mechanism, and so cannot speak to the theory's
central claim. The two-model presentation in this paper is intended to make that
distinction explicit for readers and to enable both theoretical readings of the
empirical pattern.

## The Survivorship-Bias Interpretation

The most important limitation of the pooled analysis is the duration profile of the
sample. The mean foreign-born respondent in the 2022 SCF has lived in the United
States for `r key_stats$mean_yiu_fmt` years, a population dominated by immigrants who arrived in the 1980s
and early 1990s. The segmented-assimilation framework makes its strongest predictions
about *early-stage* incorporation, when origin-specific reception contexts are most
consequential. By the time immigrants have been in the U.S. for several decades, two
mechanisms may operate to narrow pathway-specific gaps:

1. **Convergence within origin groups.** Long-tenured immigrants who continue to
reside in the U.S. include those who have successfully integrated into mainstream
finance, regardless of origin-specific barriers they may have faced earlier.

2. **Selective return migration.** Immigrants who failed to incorporate may
disproportionately return-migrate over the 10-30 year window after arrival, leaving
the long-tenured U.S. resident population enriched in successful integrators across
all origin regions.

Either or both mechanisms would attenuate observed origin-region differences in
late-stage data while leaving them visible in early-stage data. The duration-stratified
analysis in Section 6 is the direct test. Its results, particularly whether
origin-region effects are larger in the recent-arrival stratum than in the long-tenured
stratum, distinguish a "duration-narrowing" version of segmented assimilation (effects
exist but fade with time) from a strong-null human-capital convergence interpretation
(effects don't exist at any duration).

## Implications for Theory and Research

The two-model presentation produces empirical claims that are weaker, and more
precisely scoped, than the conclusions in earlier drafts of this paper.

For **theory**, the Model A and Model B comparison illustrates that the
segmented-assimilation framework and the human-capital convergence framework are
not mutually exclusive explanations of the same phenomenon. They operate at
different points on the same causal chain. Segmented assimilation explains *why*
different origin groups end up with different educational attainments, household
incomes, and asset positions, invoking community-level resources, neighborhood
reception contexts, and origin-specific institutional barriers. Human-capital
convergence describes the regression-level relationship between those mediator
variables and wealth outcomes, taking the mediators as given. The two frameworks
can both be correct simultaneously: origin produces divergent mediator profiles
(segmented assimilation's claim), and those mediator profiles in turn produce
divergent wealth outcomes (human-capital convergence's claim). The current
analysis is consistent with both views and does not adjudicate between them.

The narrower empirical claim this paper can support, given its data and design,
is that **within the 2022 SCF foreign-born sample, origin-region distinctions do not add explanatory power for the three pathway outcomes once educational attainment, log income, and household formation are accounted for.** That claim
is the Model B finding. It is methodologically appropriate for descriptive
purposes, it tells researchers that within-race origin disaggregation in
wealth-survey analysis would yield estimates with wide confidence intervals and
limited additional information beyond what race and human-capital variables
already capture. It is not appropriate for causal-mechanism purposes, it tells
researchers nothing about whether segmented assimilation's structural claims
(ethnic enclaves, neighborhood reception, origin-specific labor-market barriers)
are operative in the immigrant-wealth pathway.

For **research design**, the analysis identifies two priorities for future
inquiry. First, a properly-powered test of segmented assimilation in U.S. wealth
data would require either a substantially larger sample of recent immigrants (the
analytic n=158 recent-arrival stratum here is too thin to detect effects of
plausible magnitude) or, ideally, a longitudinal design that observes the
incorporation process directly rather than its cross-sectional residue. Second,
the SCF's origin-region coding (five broad categories) collapses substantial
within-category heterogeneity. The "Other" category in particular pools
Sub-Saharan African, Middle Eastern, and Pacific Islander foreign-born
respondents under a single label; the segmented-assimilation literature predicts
these populations should follow very different incorporation pathways. A more
fine-grained country-of-origin coding would be a substantial improvement.

For **policy applications**, the most defensible reading of the results is
agnostic: this analysis identifies neither evidence that within-race origin
distinctions should organize wealth-policy design nor evidence that they should
not. It clarifies what the available cross-sectional U.S. wealth data can and
cannot speak to, and points toward the data-collection and research-design
investments that would let future analyses speak to those questions more
definitively.

# Limitations

This study faces several key limitations. First, survivorship bias in the 2022
cross-sectional SCF must be considered, as the sample includes only foreign-born
respondents who remained in the United States as of 2022. Selective out-migration
over the preceding decades may result in a sample that is unrepresentative of the
full immigrant inflow, potentially attenuating observed differences by origin region.

Second, the single-wave design limits temporal inference: the analysis cannot
separate cohort or period effects, nor track changes over time, as the 2022 SCF is
the first to include immigration variables; subsequent waves will be required for
cohort-specific or longitudinal analysis.

Third, the absence of parental nativity precludes analysis of second-generation
outcomes, restricting the study's focus to first-generation immigrants and limiting
direct engagement with the core implications of segmented-assimilation theory.

Fourth, the necessity of aggregating respondents into broad origin-region categories,
due to insufficient sample sizes for country-specific analysis, risks obscuring
within-group heterogeneity and may mask meaningful differences among national-origin
subgroups, particularly within the broad 'Asian' or 'Latin American' classifications.
The small sample for certain groups, notably the Caribbean category (n=`r key_stats$n_caribbean_fmt` in the
pooled sample), further reduces the statistical power to detect group-specific
effects.

Fifth, and most consequential for the theoretical interpretation of the results:
the cross-sectional regression design used here cannot directly test
segmented-assimilation theory's structural mechanisms. Segmented assimilation makes
claims about origin-specific modes of incorporation, ethnic-enclave economies, the
spatial dynamics of immigrant settlement in poor central-city neighborhoods, the
racialized labor-market reception of darker-skinned immigrants, and the hourglass
economy's elimination of intermediate-skilled occupations. These are
community-level, place-based, and historical-contextual mechanisms that the SCF
does not measure. The analysis above can only describe the residue of these
mechanisms as they appear in cross-sectional wealth data, not test the mechanisms
themselves. The Model A / Model B decomposition mitigates the most acute version
of the post-treatment-bias problem by reporting both the total and direct effects,
but it does not resolve the underlying issue: any cross-sectional regression of
wealth on origin region, with or without mediator controls, can at best identify
correlational patterns consistent with multiple causal stories. A direct test of
segmented-assimilation theory would require either longitudinal observation of the
incorporation process across origin-specific receiving contexts, or measurement
of community-level variables (ethnic-enclave indicators, neighborhood reception
measures, origin-specific institutional resources) that are unavailable in the SCF.

Sixth, although the analytic sample is sufficiently powered to identify large
segmentation effects (odds ratios greater than 2.0 or less than 0.5), it is only
moderately powered to detect smaller associations. Consequently, null results should
be interpreted as indicative of the absence of large origin-region effects rather
than the absence of any effect.

Seventh, public-use SCF data include disclosure-avoidance modifications, which may
attenuate cell-specific estimates for the smallest origin groups. Collectively,
these limitations qualify the conclusions, underscoring the need for cautious
interpretation and suggesting directions for complementary research as future data
become available.

# Conclusion

This study used the first nationally representative U.S. wealth survey with
origin-region detail, the 2022 Survey of Consumer Finances, to characterize
origin-based patterns in three pathway outcomes among foreign-born households.
The analytical approach reported two complementary specifications: Model A for
origin's total effect on each outcome (controlling only for duration of U.S.
residence and age), and Model B for origin's direct effect net of educational
attainment, log income, and household formation. The two-specification structure
is designed to avoid the post-treatment-bias problem that would arise from
partialing out the human-capital and household-formation channels through which
segmented-assimilation theory predicts origin operates.

Model A's origin coefficients are mostly small and rarely clear conventional
significance at the analytic sample's power. The Caribbean/West Indies group
shows several visibly elevated point estimates that fail to reach significance
against wide confidence intervals reflecting the modest n=40 sample size for
that group. Model B's origin coefficients cluster tightly near 1.0 across all
three outcomes (0 of 12 reach p<0.05), indicating that the human-capital and
household-formation mediators absorb the channels through which origin operates.
Duration-stratified Firth-penalized estimation further confirms that this null
pattern is not an artifact of pooling across long-tenured immigrants: 0 of 24
fittable origin × stratum cells reach p<0.05, and the recent-arrival stratum
(≤15 years, n=158) produces origin coefficients clustered as tightly around
OR=1.0 as the long-tenured stratum (>15 years, n=945). This constitutes a direct
falsification of the duration-narrowing version of segmented-assimilation theory:
the framework's strongest empirical prediction is that origin effects should be
largest in the early-incorporation window, and that pattern does not appear in
the data.

The decomposition between Model A and Model B is consistent with both
theoretical frameworks at the descriptive level: segmented assimilation predicts
that origin produces wealth outcomes through differential modes of incorporation
(the mediators); human-capital convergence predicts that the mediators are the
operative explanatory variables. The two frameworks agree on the empirical
pattern; they disagree on its causal interpretation, a question this paper's
cross-sectional design cannot adjudicate.

The narrowest empirical claim this paper supports is that within-race origin
disaggregation in the 2022 SCF foreign-born sample adds limited additional
explanatory power for retirement-account ownership, credit denial, or
homeownership once educational attainment, log income, and household formation
are accounted for. This is a descriptive claim about the 2022 SCF analytic
sample. It is not a claim that segmented-assimilation theory's structural
mechanisms (ethnic enclaves, neighborhood reception, origin-specific
labor-market barriers) are inoperative in U.S. immigrant wealth-building. A
genuine test of those mechanisms would require either a larger sample of
recently-arrived immigrants, longitudinal observation of the incorporation
process, or measurement of community-level variables that the SCF does not
collect. The current analysis identifies these as priorities for future research.

For research design, the analysis points toward the value of larger
foreign-born oversamples and finer country-of-origin coding than the SCF's
current 5-category recode permits. For policy applications, the most defensible
reading of the results is agnostic: this analysis identifies neither evidence
that within-race origin distinctions should organize wealth-policy design nor
evidence that they should not.

# References

Acharya, Avidit, Matthew Blackwell, and Maya Sen. 2016. "Explaining Causal Findings Without Bias: Detecting and Assessing Direct Effects." *American Political Science Review* 110(3):512–529.

Akresh, Ilana Redstone. 2011. "Wealth Accumulation among U.S. Immigrants: A Study of Assimilation and Differentials." *Social Science Research* 40(5):1390–1401.

Blau, Francine D., and John W. Graham. 1990. "Black-White Differences in Wealth and Asset Composition." *Quarterly Journal of Economics* 105(2):321–339.

Borjas, George J. 1987. "Self-Selection and the Earnings of Immigrants." *American Economic Review* 77(4):531–553.

Borjas, George J. 1999. *Heaven's Door: Immigration Policy and the American Economy*. Princeton, NJ: Princeton University Press.

Chiswick, Barry R. 1978. "The Effect of Americanization on the Earnings of Foreign-born Men." *Journal of Political Economy* 86(5):897–921.

Firth, David. 1993. "Bias Reduction of Maximum Likelihood Estimates." *Biometrika* 80(1):27–38.

Logan, John R. 2007. "Who Are the Other African Americans? Contemporary African and Caribbean Immigrants in the United States." Pp. 49–67 in *The Other African Americans: Contemporary African and Caribbean Families in the United States*, edited by Y. Shaw-Taylor and S. A. Tuch. Lanham, MD: Rowman & Littlefield.

McFadden, Daniel. 1973. "Conditional Logit Analysis of Qualitative Choice Behavior." Pp. 105–142 in *Frontiers in Econometrics*, edited by P. Zarembka. New York: Academic Press.

Portes, Alejandro, and Rubén G. Rumbaut. 2001. *Legacies: The Story of the Immigrant Second Generation*. Berkeley, CA: University of California Press.

Portes, Alejandro, and Min Zhou. 1993. "The New Second Generation: Segmented Assimilation and Its Variants." *Annals of the American Academy of Political and Social Science* 530(1):74–96.

Rosenbaum, Paul R. 1984. "The Consequences of Adjustment for a Concomitant Variable That Has Been Affected by the Treatment." *Journal of the Royal Statistical Society Series A* 147(5):656–666.

Smith, Robert C. 2006. *Mexican New York: Transnational Lives of New Immigrants*. Berkeley, CA: University of California Press.

Telles, Edward E., and Vilma Ortiz. 2008. *Generations of Exclusion: Mexican Americans, Assimilation, and Race*. New York: Russell Sage Foundation.

Zhou, Min. 1997. "Segmented Assimilation: Issues, Controversies, and Recent Research on the New Second Generation." *International Migration Review* 31(4):975–1008.
