---
title: "Segmented Assimilation in the American Financial System: Origin-Based Gradients in Immigrant Household Finances"
subtitle: "Evidence from the 2022 Survey of Consumer Finances"
author: "Delroy Nichols"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 5, dpi = 150
)
rm(list = ls())
gc()
set.seed(42)
```

# Origin-Based Gradients in Immigrant Household Finances

## The Central Question

Segmented assimilation theory predicts that immigrants follow distinct trajectories of
incorporation depending on origin, human capital, and reception context. Three pathways
are commonly identified: *parallel integration* into the white middle class, *downward
assimilation* into the racialized urban underclass, and *selective assimilation* through
ethnic-community resources. While the fertility, educational, and occupational
implications of segmented assimilation have been studied extensively, financial-system
incorporation has received less empirical attention — in part because nationally
representative wealth surveys have lacked detailed national-origin information.

The 2022 Survey of Consumer Finances introduced new questions on respondent and spouse
national origin and years lived in the United States, allowing the first nationally
representative test of segmented-assimilation predictions on financial outcomes for U.S.
household wealth. This paper uses these new variables to test whether financial outcomes
among foreign-born households vary systematically by origin region in ways consistent
with the three-pathway theoretical framework.

**Data Source:** 2022 Survey of Consumer Finances (SCF), Federal Reserve Board. All
estimates are survey-weighted, and all models are pooled across the SCF's five multiply-
imputed datasets using Rubin's Rules.

**Note on Scope:** All findings are correlational and limited to the foreign-born
respondent sample. The SCF's cross-sectional design and absence of parental nativity
preclude causal claims and intergenerational analysis.

## Why This Matters: Beyond Race-Only Frameworks

Wealth and credit-access disparities have generally been studied through the lens of
race rather than origin. Studies of Black-white and Hispanic-white wealth gaps treat the
groups they compare as internally homogeneous — but Black households of Caribbean origin,
Mexican-origin Hispanics, and Cuban-origin Hispanics have very different financial
profiles, shaped by distinct migration histories, reception contexts, and community
resources. Pooling them into single racial categories obscures rather than illuminates
the structural processes that segmented-assimilation theory was developed to explain.

The 2022 SCF, by adding origin-region detail, lets the field begin testing whether the
segmented-assimilation pathways — parallel integration, downward assimilation, selective
assimilation — show up in wealth and credit data. Mexican-origin households may show low
mainstream-finance participation alongside high homeownership (selective assimilation
through co-ethnic mortgage networks). Asian-origin households may show elevated
mainstream participation (parallel integration). These pathway-coherent patterns are
testable with the new data, and they have direct policy implications: interventions
designed for "Hispanic households" or "Asian households" miss the heterogeneity that
actually drives outcomes.

---

```{r load-packages}
required_packages <- c(
  "tidyverse", "haven", "survey", "labelled",
  "ggplot2", "scales", "scf",
  "broom", "knitr", "moments"
)

for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg, repos = "http://cran.us.r-project.org")
    library(pkg, character.only = TRUE)
  }
}
```

```{r download-data}
# Same download pattern as the credit-card-debt analysis (Paper 1).
# scf_download writes .rds files to the current working directory.

dir.create("scf_data", showWarnings = FALSE, recursive = TRUE)
original_wd <- getwd()

# Move into the data folder
setwd("scf_data")

library(haven)

# Read from your local absolute path
raw_macro_df <- read_dta("C:/Users/delro/OneDrive/Documents/scf2022s/p22i6.dta")

# Save directly to the current working directory (which is already scf_data)
saveRDS(raw_macro_df, "p22i6.rds") 

# Continue with the download check for the bulletin extract
if (!file.exists("scf2022.rds")) {
  tryCatch({
    scf_download(2022)
  }, error = function(e) {
    cat("Error downloading 2022 data:", conditionMessage(e), "\n")
  })
}

# Return to the root project directory
setwd(original_wd)
```

```{r load-data}
scf_2022 <- scf_load(2022, data_directory = "scf_data")
```

# Variable Availability and Diagnostic

Before constructing variables, we check which of the variables having direct exposure without segmented assimilation theory are
actually available in the loaded `scf` package object. The standard SCF Bulletin extract
that the `scf` package builds does not by default include the new national-origin or
years-in-U.S. items (`X6906`, `X7404`, `X6915`); these are raw codebook items rather than
constructed Bulletin variables. If they are missing, we read the raw 2022 SCF `.rds`
file directly and merge them in by household ID.

```{r diagnose-immigration-vars, results='asis'}
imp1 <- scf_2022$mi_design[[1]]$variables

# Variables we need
# X6915 = respondent national origin (the variable for origin region)
# X6906 = years respondent has lived in the U.S. (the YIU variable)
# X7796 = years spouse has lived in the U.S. (auxiliary)
# X7404 = spouse Hispanic/Latino self-id (auxiliary)
# X8049 = spouse race (used in swap correction)
# X8000 = sex/role swap indicator
#
# CORRECTED 2026-04-29: V5 incorrectly used X7437-X7443 as years-in-US.
# Per the official SCF 2022 changes file, X7437-X7443 are MORTGAGE FORBEARANCE
# variables, not immigration variables. The true years-in-US variable is X6906
# for the respondent (and X7796 for the spouse).
imm_vars <- c("X6915", "X6906", "X7796", "X7404", "X8000", "X8049")

present_in_extract <- imm_vars %in% names(imp1)

cat("**Immigration-variable availability in the scf package extract:**\n\n")
diag_df <- data.frame(
  Variable = imm_vars,
  `In extract` = ifelse(present_in_extract, "Yes", "No"),
  check.names = FALSE
)
print(kable(diag_df))
cat("\n\n")

need_raw_merge <- !all(present_in_extract)

if (need_raw_merge) {
  cat("Some required immigration variables are missing from the bulletin extract. ",
      "Attempting to merge them in from the raw 2022 SCF data file...\n\n", sep = "")
} else {
  cat("All required immigration variables are present in the extract. ",
      "No raw merge needed.\n\n", sep = "")
}
```

```{r merge-raw-immigration-vars}
# ============================================================================
# Merge raw immigration variables from the underlying SCF .rds file
# ============================================================================
# The scf package downloads the raw 2022 SCF data into scf_data/ as well as
# building the bulletin extract. The raw file contains all X-variables; we
# read it, select just the immigration-related vars + ID, and merge into
# each implicate's $variables data frame.
#
# Common file names produced by scf_download():
#   - p22i6.rds (raw public-use, 5 implicates stacked, with implicate id)
#   - scf2022.rds (the bulletin extract)
#
# We try a few candidate filenames defensively.

if (need_raw_merge) {

  raw_candidates <- c(
    file.path("scf_data", "p22i6.rds"),
    file.path("scf_data", "scfp2022.rds"),
    file.path("scf_data", "p22.rds"),
    "p22i6.rds",
    "scfp2022.rds"
  )

  raw_path <- raw_candidates[file.exists(raw_candidates)][1]

  if (!is.na(raw_path)) {
    raw_data <- readRDS(raw_path)

    # Standardize variable name casing — the raw file may have either
    # uppercase or lowercase. We try uppercase first since X-variables
    # are conventionally uppercase.
    raw_path <- raw_candidates[file.exists(raw_candidates)][1]

  if (!is.na(raw_path)) {
    raw_data <- readRDS(raw_path)

    # FIX: Force all raw data columns to uppercase immediately.
    # This aligns the Stata output with the canonical SCF uppercase format.
    names(raw_data) <- toupper(names(raw_data))

    raw_names <- names(raw_data)
    
    # The name_map is now technically redundant, but safe to keep
    name_map <- setNames(raw_names, toupper(raw_names))

    available_imm <- imm_vars[toupper(imm_vars) %in% toupper(raw_names)]

    # The raw data has 5 implicates stacked. We need an ID that matches
    # the bulletin extract's row ordering. Common candidates: y1, yy1, x1.
    id_candidates <- c("Y1", "YY1", "X1") # Updated to only look for uppercase
    id_var <- id_candidates[id_candidates %in% raw_names][1]

    if (!is.na(id_var) && length(available_imm) > 0) {

      # Pull the matching columns
      raw_imm <- raw_data[, c(id_var, available_imm), drop = FALSE]

      # Determine implicate count: raw file has each household 5 times (one
      # per implicate) — split into 5 chunks of equal size in row order.
      n_total <- nrow(raw_imm)
      n_per_imp <- nrow(scf_2022$mi_design[[1]]$variables)

      if (n_total == 5 * n_per_imp) {
        # Standard layout: implicates stacked
        for (i in 1:5) {
          start_row <- (i - 1) * n_per_imp + 1
          end_row   <- i * n_per_imp
          imm_chunk <- raw_imm[start_row:end_row, available_imm, drop = FALSE]

          # Add to the implicate's $variables
          for (v in available_imm) {
            scf_2022$mi_design[[i]]$variables[[v]] <- imm_chunk[[v]]
          }
        }
        cat("Successfully merged ", length(available_imm), " immigration variables ",
            "from ", raw_path, " into all 5 implicates.\n", sep = "")
      } else {
        cat("Row-count mismatch: raw file has ", n_total,
            " rows, bulletin extract has ", n_per_imp,
            " rows per implicate. Merge aborted; using available variables only.\n",
            sep = "")
      }
    } else {
      cat("Could not locate raw SCF file or matching ID variable. ",
          "Immigration variables will be unavailable.\n", sep = "")
    }
  } else {
    cat("Raw 2022 SCF .rds file not found in scf_data/. ",
        "Available files:\n", sep = "")
    cat(paste(list.files("scf_data"), collapse = "\n"), "\n\n")
  }
}

# Re-check what's now available
imp1 <- scf_2022$mi_design[[1]]$variables
have_X7404 <- "X7404" %in% names(imp1)
have_yiu   <- "X6906" %in% names(imp1)
have_X8000 <- "X8000" %in% names(imp1)
have_X8049 <- "X8049" %in% names(imp1)
```

```{r raw-x7404-frequency, results='asis'}
# ============================================================================
# Print raw X7404 value frequencies — needed to calibrate the recode
# ============================================================================
# Until we see what values X7404 actually takes in this wave, we cannot
# write the correct recoding rule. This chunk dumps the unweighted frequency
# table so we can adjust the recode.

if (have_X7404) {
  # 1. Extract the column and force it to a standard numeric vector
  # This strips out any problematic 'haven_labelled' metadata from Stata
  x7404_vec <- as.numeric(scf_2022$mi_design[[1]]$variables$X7404)
  
  # 2. Use Base R's table() to get the counts, which avoids package conflicts
  freq_table <- table(x7404_vec, useNA = "ifany")
  
  # 3. Reconstruct it into a clean data frame for kable
  x7404_freq <- data.frame(
    X7404 = names(freq_table),
    n = as.integer(freq_table)
  )
  
  # 4. Sort descending
  x7404_freq <- x7404_freq[order(-x7404_freq$n), ]

  cat("**Raw X7404 frequency (Implicate 1, unweighted):**\n\n")
  # row.names = FALSE keeps the markdown table perfectly clean
  print(kable(x7404_freq, caption = "X7404 = respondent national origin", row.names = FALSE))
  cat("\n\nThis table guides the origin-region recode in the next chunk. ",
      "Code 0 typically indicates 'inapplicable' (i.e., U.S.-born or non-respondent).\n\n",
      sep = "")
}
```

```{r visuals}

library(ggplot2)
library(dplyr)

# ============================================================================
# Visualize X7404 Frequencies: Spouse Hispanic/Latino Culture/Origin
# ============================================================================

# 1. SCF mapping for X7404
spouse_hisp_labels <- c(
  "0" = "Inapplicable (No Spouse/Partner)",
  "1" = "Yes (Spouse is Hispanic/Latino)",
  "5" = "No (Spouse is not Hispanic/Latino)"
)

# 2. Apply labels and generate plot
x7404_freq %>%
  mutate(
    # Map the numeric codes to our accurate text labels
    Spouse_Label = spouse_hisp_labels[as.character(X7404)],
    # Catch any unexpected codes just in case
    Spouse_Label = ifelse(is.na(Spouse_Label), paste("Code", X7404), Spouse_Label)
  ) %>%
  # Order the bars by frequency 
  ggplot(aes(x = reorder(Spouse_Label, n), y = n)) +
  geom_col(fill = "steelblue", color = "black", alpha = 0.8) +
  coord_flip() + # Horizontal layout for readability
  theme_minimal() +
  labs(
    title = "Spouse/Partner Hispanic or Latino Origin",
    subtitle = "Raw frequencies from Implicate 1 (Variable X7404)",
    x = NULL, 
    y = "Number of Respondents"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.y = element_text(size = 11)
  )

```



```{r plot-x6915, echo=FALSE, warning=FALSE, message=FALSE}
# ============================================================================
# Visualize X6915 Frequencies: Reference Person National Origin
# ============================================================================

library(ggplot2)
library(dplyr)

# FIX: Explicitly define have_X6915 so the Knit session knows what it is!
have_X6915 <- "X6915" %in% names(scf_2022$mi_design[[1]]$variables)

if (have_X6915) {
  
  x6915_vec <- as.numeric(scf_2022$mi_design[[1]]$variables$X6915)
  freq_table_6915 <- table(x6915_vec, useNA = "ifany")
  
  x6915_freq <- data.frame(
    X6915 = names(freq_table_6915),
    n = as.integer(freq_table_6915)
  )
  
  origin_labels <- c(
    "0" = "U.S. Born / Inapplicable",
    "100" = "Western European", 
    "200" = "Eastern European",
    "300" = "European",
    "400" = "West Indies",
    "500" = "Central and South American", 
    "600" = "Middle Easterner",
    "700" = "Sub-Saharan African",
    "800" = "South Asian",
    "900" = "East Asian",
    "1000" = "All Other Asians",
    "1100" = "Pacific Islander",
    "1200" = "American"
  )
  
  p_x6915 <- x6915_freq %>%
    mutate(
      Origin_Label = origin_labels[as.character(X6915)],
      Origin_Label = ifelse(is.na(Origin_Label), paste("Code", X6915), Origin_Label)
    ) %>%
    filter(X6915 != "0") %>% 
    ggplot(aes(x = reorder(Origin_Label, n), y = n)) +
    geom_col(fill = "seagreen", color = "black", alpha = 0.8) +
    coord_flip() + 
    theme_minimal() +
    labs(
      title = "Reference Person National Origin (Immigrant Sub-sample)",
      subtitle = "Unweighted frequencies from Implicate 1 (Variable X6915, excluding U.S. Born)",
      x = NULL, 
      y = "Number of Respondents"
    ) +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      axis.text.y = element_text(size = 11)
    )
  
  # Properly print the plot
  print(p_x6915)
  
} else {
  cat("Variable X6915 not found. Please ensure the merge chunk was re-run.")
}
```

```{r table-x6915, echo=FALSE, results='asis'}
# Print the raw counts table
if (have_X6915) {
  cat("\n\n**Raw X6915 frequency (Implicate 1, unweighted):**\n\n")
  tab <- knitr::kable(x6915_freq %>% arrange(desc(n)), row.names = FALSE)
  cat(paste(tab, collapse = "\n"), "\n\n")
}
```


# Variable Engineering

```{r data-engineering-and-subsetting, warning=FALSE, message=FALSE}
# ============================================================================
# Build all analysis variables and create the analytic subset
# ============================================================================
# CRITICAL DESIGN NOTE on foreign-born identification:
#
# X6915 is an ANCESTRY/HERITAGE variable, asked of all respondents — both
# foreign-born and U.S.-born. A respondent reporting "Western European"
# ancestry (code 100) is overwhelmingly U.S.-born (white American with
# European heritage), not an immigrant from Europe. Likewise, "American"
# ancestry (1200) is U.S.-born by definition, and "Sub-Saharan African" (700)
# captures Black Americans reporting African ancestry alongside actual
# African immigrants.
#
# The actual foreign-born signal is in X6906 (years the reference person
# has lived in the United States), which the SCF only asks of foreign-born
# respondents. A non-zero, non-missing years-in-US value confirms foreign-
# born status. (The companion variable X7796 is years-in-US for the spouse
# but is not used in main models, since the unit of analysis is the
# respondent.)
#
# IMPORTANT NOTE: V5 erroneously used X7437-X7443 as the years-in-US
# variables. Per the official SCF 2022 changes file, those are MORTGAGE
# FORBEARANCE variables (which mortgage terms were adjusted after exiting
# forbearance). The correct YIU variable is X6906. This V6 build uses
# X6906 throughout.
#
# Therefore, foreign_born requires BOTH:
#   (a) X6915 indicates a non-American ancestry (i.e., X6915 != 0 AND
#       X6915 != 1200), AND
#   (b) X6906 > 0 (confirming actual years lived in the U.S. as immigrant)
# ============================================================================

# ---- Step 1: Years in U.S. (X6906 = respondent years in US) ----
# X6906 = 0 means inapplicable per SCF data convention (i.e., U.S.-born).
# X6906 > 0 means the respondent reported N years lived in the U.S.,
# which is asked only of foreign-born respondents.
if (have_yiu) {
  scf_2022 <- scf_update(scf_2022,
    years_in_us  = ifelse(as.numeric(X6906) <= 0, NA_real_, as.numeric(X6906)),
    yiu_missing  = as.integer(is.na(years_in_us)),
    yiu_clean    = ifelse(is.na(years_in_us), 0, years_in_us)
  )
} else {
  scf_2022 <- scf_update(scf_2022,
    years_in_us = NA_real_, yiu_missing = 1L, yiu_clean = 0
  )
}

# ---- Step 2: CORRECTED foreign-born status ----
# Requires both ancestry indicating non-US origin AND positive years in US.
# Fall back gracefully if X6915 is unavailable.
if (have_X6915 && have_yiu) {
  scf_2022 <- scf_update(scf_2022,
    foreign_born = as.integer(
      !is.na(years_in_us) &
      years_in_us > 0 &
      as.numeric(X6915) != 0 &
      as.numeric(X6915) != 1200
    )
  )
} else if (have_yiu) {
  # YIU-only fallback if X6915 is unavailable
  scf_2022 <- scf_update(scf_2022,
    foreign_born = as.integer(!is.na(years_in_us) & years_in_us > 0)
  )
} else {
  scf_2022 <- scf_update(scf_2022, foreign_born = 0L)
}

# ---- Step 3: Origin-region recode from X6915 ----
# Used only within the foreign-born sample (filtered by foreign_born flag).
# Reference category: Latin American (the segmented-assimilation canonical
# case in the literature; X6915 == 500 = Central/South American).
if (have_X6915) {
  scf_2022 <- scf_update(scf_2022,
    origin_region = factor(case_when(
      as.numeric(X6915) == 500 ~ "Latin American",
      as.numeric(X6915) == 400 ~ "Caribbean/West Indies",
      as.numeric(X6915) %in% c(800, 900, 1000) ~ "Asian",
      as.numeric(X6915) %in% c(100, 200, 300) ~ "European/Other developed",
      as.numeric(X6915) %in% c(600, 700, 1100) ~ "Other",
      TRUE ~ NA_character_
    ), levels = c("Latin American",
                  "Caribbean/West Indies",
                  "Asian",
                  "European/Other developed",
                  "Other"))
  )
} else {
  scf_2022 <- scf_update(scf_2022, origin_region = factor(NA_character_))
}

# ---- Step 3b: Spouse/Partner Hispanic Culture/Origin from X7404 ----
if (have_X7404) {
  scf_2022 <- scf_update(scf_2022,
    spouse_hisp_origin = factor(case_when(
      X7404 == 1 ~ "Yes (Spouse Hispanic/Latino)",
      X7404 == 5 ~ "No (Spouse not Hispanic/Latino)",
      X7404 == 0 ~ "Unmarried/Inapplicable",
      TRUE ~ NA_character_
    ), levels = c("Yes (Spouse Hispanic/Latino)",
                  "No (Spouse not Hispanic/Latino)",
                  "Unmarried/Inapplicable"))
  )
} else {
  scf_2022 <- scf_update(scf_2022, spouse_hisp_origin = factor(NA_character_))
}

# ---- Step 4: Race/ethnicity ----
scf_2022 <- scf_update(scf_2022,
  race_cat = factor(racecl4, levels = 1:4,
                    labels = c("White", "Black", "Hispanic", "Other"))
)

# ---- Step 5: Standard demographic controls ----
scf_2022 <- scf_update(scf_2022,
  age_centered = age - 50,
  age_sq       = age^2,
  in_age_range = as.integer(age >= 25 & age <= 75),
  educ_cat = factor(case_when(
    educ < 8 ~ "Less than HS",
    educ == 8 ~ "High School",
    educ %in% 9:11 ~ "Some College",
    educ == 12 ~ "College Degree",
    educ %in% 13:14 ~ "Graduate School/PhD",
    TRUE ~ NA_character_
  ), levels = c("Less than HS", "High School", "Some College",
                "College Degree", "Graduate School/PhD")),
  female            = as.integer(hhsex == 2),
  currently_married = as.integer(married == 1),
  has_children      = as.integer(kids > 0),
  is_homeowner      = as.integer(housecl == 1),
  has_pension       = as.integer(anypen == 1),
  log_income        = log(pmax(1, income)),
  asinh_networth    = asinh(networth),
  log_wageinc       = log(pmax(1, wageinc))
)

# ---- Step 6: Outcomes ----
# CRITICAL: nw_pctl is computed on the FULL SCF distribution (before
# subsetting to foreign-born) so it reflects position in the broader
# U.S. wealth distribution, not relative position within the foreign-born
# subsample. This is computed via scf_update on scf_2022 (the full design)
# so all 5 implicates get a consistent percentile based on the full sample.
scf_2022 <- scf_update(scf_2022,
  own_retire    = as.integer(retqliq > 0),
  own_stocks    = as.integer(stocks > 0),
  credit_denied = as.integer(turndown == 1),
  no_checking   = as.integer(checking <= 0),
  own_home      = as.integer(housecl == 1),
  nw_pctl       = 100 * percent_rank(networth)   # full-sample percentile
)

# ---- Step 7: Analytic sample flag ----
scf_2022 <- scf_update(scf_2022,
  in_analytic_sample = as.integer(
    foreign_born == 1 &
    !is.na(origin_region) &
    in_age_range == 1
  )
)

# ---- Step 8: Create the Analytic Sample Subset ----
# Subset each implicate's survey design to foreign-born respondents.
# nw_pctl values are preserved — they were computed on the full distribution
# in Step 6 above, so subsetting just keeps the foreign-born rows with
# their original full-distribution percentile values intact.
scf_fb <- scf_2022
scf_fb$mi_design <- lapply(scf_2022$mi_design, function(des) {
  subset(des, in_analytic_sample == 1)
})

# Diagnostic: report sample size
n_fb <- nrow(scf_fb$mi_design[[1]]$variables)

# Diagnostic: cross-tab of X6915 ancestry × foreign_born flag — useful for
# verifying that the corrected foreign-born identification is sensible.
fb_diagnostic <- scf_2022$mi_design[[1]]$variables %>%
  filter(!is.na(X6915)) %>%
  group_by(X6915) %>%
  summarise(
    n_total       = n(),
    n_foreign     = sum(foreign_born == 1, na.rm = TRUE),
    pct_foreign   = round(100 * mean(foreign_born == 1, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  arrange(desc(n_total))
```

```{r build-foreign-born-design}
# ============================================================================
# Build the foreign-born subset design — same pattern as scf_debtors in Paper 1
# ============================================================================
# NOTE: foreign_born and years_in_us are already defined in the
# `data-engineering-and-subsetting` chunk above, using the X6915-based
# definition (X6915 != 0 = foreign-born). The X6915 definition is preferred
# over X6906 because X6906 has known small missingness (some foreign-born
# missingness, while X6915 is universally coded.
#
# This chunk just rebuilds the scf_fb subset object using the
# in_analytic_sample flag from the prior chunk.

scf_fb <- scf_2022

scf_fb$mi_design <- lapply(scf_2022$mi_design, function(des) {
  subset(des, in_analytic_sample == 1)
})

# Diagnostic: report sample size
n_fb <- nrow(scf_fb$mi_design[[1]]$variables)
```

# Descriptive Findings

## Foreign-Born Identification Diagnostic

The corrected `foreign_born` definition requires both ancestry indicating non-U.S.
origin (X6915 ≠ 0 and X6915 ≠ 1200) AND positive years in the U.S. (X6906).
The table below shows, for each X6915 ancestry code, what percentage of respondents
end up classified as foreign-born. Codes like 100 (Western European), 200 (Eastern
European), and 1200 (American) should be near 0%; codes like 500 (Latin American),
800/900 (Asian), and 600 (Middle Eastern) should be near 100%.

```{r fb-id-diagnostic, results='asis'}
if (exists("fb_diagnostic") && nrow(fb_diagnostic) > 0) {

  origin_label_map <- c(
    "0"    = "U.S. Born / Inapplicable",
    "100"  = "Western European",
    "200"  = "Eastern European",
    "300"  = "European (Other)",
    "400"  = "West Indies / Caribbean",
    "500"  = "Central / South American",
    "600"  = "Middle Eastern",
    "700"  = "Sub-Saharan African",
    "800"  = "South Asian",
    "900"  = "East Asian",
    "1000" = "All Other Asian",
    "1100" = "Pacific Islander",
    "1200" = "American"
  )

  fb_diagnostic_display <- fb_diagnostic %>%
    mutate(
      Ancestry = origin_label_map[as.character(X6915)],
      Ancestry = ifelse(is.na(Ancestry), paste("Code", X6915), Ancestry)
    ) %>%
    select(X6915, Ancestry, n_total, n_foreign, pct_foreign)

  cat("**Table 0. Foreign-born identification by X6915 ancestry code**\n\n")
  print(kable(fb_diagnostic_display,
              col.names = c("X6915", "Ancestry label", "n total",
                            "n foreign-born", "% foreign-born"),
              caption = "Foreign-born flag = X6915 ∈ origin codes AND years_in_us > 0."))
  cat("\n\n")
}
```

## Foreign-Born Sample Composition

```{r desc-sample-comp, results='asis'}
# Sample composition statistics — the origin_region recode was already done
# in the data-engineering chunk above; we don't need to redo it here.

if (n_fb > 0) {
  fb_imp1 <- scf_fb$mi_design[[1]]$variables

  mean_age   <- weighted.mean(fb_imp1$age, fb_imp1$wgt, na.rm = TRUE)
  mean_yiu   <- weighted.mean(fb_imp1$years_in_us, fb_imp1$wgt, na.rm = TRUE)
  mean_inc   <- weighted.mean(fb_imp1$income, fb_imp1$wgt, na.rm = TRUE)
  median_nw  <- median(fb_imp1$networth, na.rm = TRUE)
  pct_yiu_missing <- 100 * mean(is.na(fb_imp1$years_in_us))

  cat(sprintf("The analytic sample comprises %s foreign-born respondents in the 2022 SCF aged 25-75. ",
              formatC(n_fb, big.mark = ",")))
  cat(sprintf("Using survey weights, the mean age in this population is %.1f years, with a mean of %.1f years lived in the U.S. ",
              mean_age, mean_yiu))
  cat(sprintf("Mean household income is $%s and median net worth is $%s. ",
              formatC(round(mean_inc), big.mark = ",", format = "d"),
              formatC(round(median_nw), big.mark = ",", format = "d")))
  cat(sprintf("Note: %.1f%% of foreign-born respondents have missing years-in-U.S. data; the missing-indicator method is used in regressions to retain these observations.\n\n",
              pct_yiu_missing))
} else {
  cat("**Analytic sample is empty.** Check the X6915 origin-region recode in the variable-engineering chunk.\n\n")
}
```

## Origin-Region Distribution

```{r desc-origin-table, results='asis'}
if (n_fb > 0) {
  fb_imp1 <- scf_fb$mi_design[[1]]$variables

  origin_summary <- fb_imp1 %>%
    filter(!is.na(origin_region)) %>%
    group_by(origin_region) %>%
    summarise(
      n_unweighted   = n(),
      weighted_count = sum(wgt, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      pct_weighted = round(weighted_count / sum(weighted_count) * 100, 1)
    ) %>%
    select(origin_region, n_unweighted, pct_weighted)

  cat("**Table 1. Foreign-born sample by origin region (2022 SCF)**\n\n")
  print(kable(origin_summary,
              col.names = c("Origin Region", "n (Unweighted)", "% (Weighted)"),
              caption = "Note: percentages are SCF-weighted; n is unweighted (Implicate 1)."))
  cat("\n\n")

  small_cells <- origin_summary %>% filter(n_unweighted < 30)
  if (nrow(small_cells) > 0) {
    cat(sprintf("**Warning:** Small unweighted cell sizes (< 30) in %d origin region(s): %s. Model estimates for these groups may be unstable.\n\n",
                nrow(small_cells),
                paste(as.character(small_cells$origin_region), collapse = ", ")))
  }
}
```

## Outcome Prevalence by Origin Region

```{r desc-outcomes-by-origin, results='asis'}
if (n_fb > 0) {

  # Use scf_mean for each outcome (matches Paper 1 pattern)
  outcomes <- list(
    own_retire    = "Retirement (%)",
    own_stocks    = "Stocks (%)",
    credit_denied = "Credit denied (%)",
    no_checking   = "No checking (%)",
    own_home      = "Homeowner (%)",
    nw_pctl       = "NW pctile (mean)"
  )

  # Run each prevalence estimate using scf_mean with by = ~origin_region
  prevalence_list <- list()
  for (out_var in names(outcomes)) {
    f <- as.formula(paste0("~", out_var))
    res <- tryCatch({
      scf_mean(scf_fb, f, by = ~origin_region)$results
    }, error = function(e) NULL)
    prevalence_list[[out_var]] <- res
  }

  # Build display table — one row per origin region
  origin_levels <- levels(scf_fb$mi_design[[1]]$variables$origin_region)
  display_df <- data.frame(`Origin Region` = origin_levels,
                           check.names = FALSE,
                           stringsAsFactors = FALSE)

  for (out_var in names(outcomes)) {
    res <- prevalence_list[[out_var]]
    col_label <- outcomes[[out_var]]

    if (is.null(res) || nrow(res) == 0) {
      display_df[[col_label]] <- "--"
      next
    }

    # The scf_mean result has columns: by-variable level, estimate, std.error
    # Match by origin_region level
    estimates <- sapply(origin_levels, function(lvl) {
      idx <- which(as.character(res[[1]]) == lvl)
      if (length(idx) == 0) return(NA_real_)
      # The estimate column is typically the second column or named "estimate"
      if ("estimate" %in% names(res)) res$estimate[idx] else res[[2]][idx]
    })

    if (out_var == "nw_pctl") {
      display_df[[col_label]] <- ifelse(is.na(estimates), "--",
                                         sprintf("%.1f", estimates))
    } else {
      display_df[[col_label]] <- ifelse(is.na(estimates), "--",
                                         sprintf("%.1f", estimates * 100))
    }
  }

  cat("**Table 2. Weighted prevalence of financial outcomes by origin region**\n\n")
  print(kable(display_df,
              caption = "All estimates are SCF-weighted and pooled across implicates."))
  cat("\n\n")
}
```

## Years-in-U.S. Distribution by Origin Region

```{r desc-yiu-plot, fig.cap="Distribution of years lived in the U.S. by origin region (foreign-born SCF sample, 2022)"}
if (n_fb > 0) {
  plot_data <- scf_fb$mi_design[[1]]$variables %>%
    filter(!is.na(years_in_us), !is.na(origin_region))

  if (nrow(plot_data) > 0) {
    ggplot(plot_data,
           aes(x = origin_region, y = years_in_us, fill = origin_region)) +
      geom_boxplot(alpha = 0.7, outlier.shape = NA) +
      geom_jitter(width = 0.15, alpha = 0.2, size = 0.6) +
      labs(x = NULL,
           y = "Years lived in the U.S.",
           subtitle = "Unweighted distributional shape (Implicate 1)") +
      theme_minimal(base_size = 11) +
      theme(legend.position = "none",
            axis.text.x = element_text(angle = 30, hjust = 1)) +
      scale_fill_brewer(palette = "Set2")
  }
}
```

# Modeling Results

## H1: Origin-Region Main Effects

We test whether financial outcomes among foreign-born respondents differ by origin
region net of years-in-U.S., age, education, marital status, and income. Logistic
regression for binary outcomes (`scf_glm` with `binomial()`) and OLS for the continuous
net-worth-percentile outcome (`scf_ols`). All models pool across the 5 implicates.

Mexican origin is the reference category — that is, the X6915=500 ("Central
and South American") group, which the segmented-assimilation literature treats
as the canonical Latin American case. Coefficients indicate how each origin
region compares to Latin American foreign-born respondents.

```{r h1-model-helper}
# ============================================================================
# Helper: fit and tidy an scf_glm or scf_ols model (NESTED SPECS)
# ============================================================================
library(dplyr)

fit_h1_model <- function(outcome_var, scf_obj, family_type = "binomial") {

  # 1. Define the nested specifications
  # V8 fix: replaced yiu_clean + yiu_missing with raw years_in_us. The
  # foreign-born filter requires valid years_in_us, so yiu_missing is
  # constant 0 across the analytic sample, which makes it a singular
  # predictor and crashes the GLM solver.
  baseline_rhs <- "origin_region + years_in_us + age_centered"
  full_rhs     <- "origin_region + years_in_us + age_centered + educ_cat + currently_married + log_income"

  baseline_formula <- as.formula(paste(outcome_var, "~", baseline_rhs))
  full_formula     <- as.formula(paste(outcome_var, "~", full_rhs))

  # CRITICAL FIX (V7): Set the formula's environment to the global environment.
  # When as.formula() is called inside a function, the formula's environment
  # becomes that function's local scope by default. The scf package's
  # internal svyglm calls then can't find the variables (origin_region,
  # educ_cat, etc.) because they live in the design object's $variables data
  # frame, not in the function's local scope. Setting the formula environment
  # to globalenv() lets svyglm fall back to looking up variables in the
  # design's data, which is what Paper 1's inline-formula calls do implicitly.
  environment(baseline_formula) <- globalenv()
  environment(full_formula)     <- globalenv()

  # 2. Internal function to fit, tidy, and label a single specification.
  # Returns a list with $results (success) or $error + $message (failure).
  fit_and_format <- function(form, label) {
    fit <- tryCatch({
      if (family_type == "binomial") {
        scf_glm(scf_obj, form, family = binomial())
      } else {
        scf_ols(scf_obj, form)
      }
    }, error = function(e) list(error = TRUE, message = conditionMessage(e)))

    # Propagate error info up so the cache loop can surface it
    if (is.list(fit) && isTRUE(fit$error)) {
      return(list(error = TRUE,
                  message = sprintf("[%s] %s", label, fit$message)))
    }

    res <- fit$results
    res$Model <- factor(label, levels = c("Baseline (Demographics Only)",
                                          "Full (With Human Capital)"))

    if (family_type == "binomial" && "estimate" %in% names(res)) {
      res$odds_ratio <- exp(res$estimate)
      res$conf_low   <- exp(res$estimate - 1.96 * res$std.error)
      res$conf_high  <- exp(res$estimate + 1.96 * res$std.error)
    } else if ("estimate" %in% names(res)) {
      res$odds_ratio <- NA
      res$conf_low   <- res$estimate - 1.96 * res$std.error
      res$conf_high  <- res$estimate + 1.96 * res$std.error
    }
    return(res)
  }

  res_base <- fit_and_format(baseline_formula, "Baseline (Demographics Only)")
  res_full <- fit_and_format(full_formula, "Full (With Human Capital)")

  base_failed <- is.list(res_base) && isTRUE(res_base$error)
  full_failed <- is.list(res_full) && isTRUE(res_full$error)

  # If both failed, surface the actual error messages
  if (base_failed && full_failed) {
    return(list(error = TRUE,
                message = paste(res_base$message, "||", res_full$message)))
  }

  # If only one failed, drop it but keep the other
  parts <- list()
  if (!base_failed) parts$base <- res_base
  if (!full_failed) parts$full <- res_full

  bind_rows(parts)
}
```

```{r h1-diagnostic-test-fit, results='asis'}
# ============================================================================
# DIAGNOSTIC: Try a simple inline-formula scf_glm call before the main cache
# loop. If this works but the cache loop fails, the issue is formula-
# environment scoping. If this also fails, the issue is something deeper
# (data, design, package version) and the message will tell us what.
# ============================================================================

cat("**Diagnostic: simple test fit using inline formula syntax (mimics Paper 1's pattern)**\n\n")

if (n_fb >= 100) {

  test_fit <- tryCatch({
    scf_glm(scf_fb, own_home ~ origin_region, family = binomial())
  }, error = function(e) list(error = TRUE, message = conditionMessage(e)))

  if (is.list(test_fit) && isTRUE(test_fit$error)) {
    cat(sprintf("- Inline test fit FAILED: `%s`\n\n", test_fit$message))
    cat("  This means scf_glm itself is unable to fit on the foreign-born subset, ",
        "regardless of formula-environment issues. Likely causes: a needed ",
        "variable is missing in the design, the design object is malformed, ",
        "or the scf package version differs from Paper 1's.\n\n", sep = "")
  } else {
    cat("- Inline test fit SUCCEEDED. scf_glm works on this design with origin_region as predictor.\n\n")
    cat("  Coefficients returned: ",
        paste(test_fit$results$term, collapse = ", "), "\n\n", sep = "")
  }

  # Also test scf_ols
  test_ols <- tryCatch({
    scf_ols(scf_fb, nw_pctl ~ origin_region)
  }, error = function(e) list(error = TRUE, message = conditionMessage(e)))

  if (is.list(test_ols) && isTRUE(test_ols$error)) {
    cat(sprintf("- Inline scf_ols test FAILED: `%s`\n\n", test_ols$message))
  } else {
    cat("- Inline scf_ols test SUCCEEDED.\n\n")
  }
}
```

```{r h1-fit-all-models-once, cache=TRUE}
# ============================================================================
# PERFORMANCE: Fit each H1 model EXACTLY ONCE here and cache the results.
# All downstream chunks (forest plots, H1 tables, H2 pattern matrix) pull from
# this cache instead of re-fitting. cache=TRUE on this chunk means knitr won't
# even re-run this if nothing upstream has changed.
#
# IMPORTANT: If you've previously knit V6 (or any earlier version) with
# cache=TRUE and the models failed, knitr may have cached the failed results.
# Before knitting V7, delete the cache folder created by knitr:
#   - file.exists("scf_segmented_assimilation_v7_cache")  # or v6/v5 cache
# OR delete the entire `*_cache/` directory next to the .Rmd file.
# Otherwise V7 will return the failed V6 results from cache instead of
# re-running the corrected fit logic.
# ============================================================================

model_cache <- list()

if (n_fb >= 100) {

  outcomes_to_fit <- list(
    list(var = "own_retire",    family = "binomial"),
    list(var = "own_stocks",    family = "binomial"),
    list(var = "credit_denied", family = "binomial"),
    list(var = "no_checking",   family = "binomial"),
    list(var = "own_home",      family = "binomial"),
    list(var = "nw_pctl",       family = "ols")
  )

  # Pre-check: identify outcomes with quasi-complete separation in the
  # foreign-born × origin × outcome table. If any origin region has zero
  # variance on the outcome, the GLM will produce gigantic standard errors
  # (effective infinity). Skip those outcomes rather than reporting noise.
  fb_imp1_for_check <- scf_fb$mi_design[[1]]$variables

  for (info in outcomes_to_fit) {
    if (info$family == "binomial") {
      cell_var <- fb_imp1_for_check %>%
        filter(!is.na(origin_region)) %>%
        group_by(origin_region) %>%
        summarise(v = var(.data[[info$var]], na.rm = TRUE), .groups = "drop")

      if (any(cell_var$v == 0 | is.na(cell_var$v))) {
        bad_groups <- cell_var %>% filter(v == 0 | is.na(v)) %>% pull(origin_region)
        message(sprintf("Skipping %s: zero within-cell variance in origin region(s): %s",
                        info$var, paste(as.character(bad_groups), collapse = ", ")))
        model_cache[[info$var]] <- list(
          error = TRUE,
          message = sprintf("Quasi-complete separation: origin region(s) %s have zero variance on this outcome.",
                            paste(as.character(bad_groups), collapse = ", "))
        )
        next
      }
    }

    message("Fitting H1 model for: ", info$var)
    fit_result <- fit_h1_model(info$var, scf_fb, info$family)
    model_cache[[info$var]] <- fit_result

    # Surface failures to the knit log so the user can see what's wrong
    if (is.list(fit_result) && isTRUE(fit_result$error)) {
      message(sprintf("  -> FAILED: %s", fit_result$message))
    } else {
      n_terms <- nrow(fit_result)
      message(sprintf("  -> SUCCESS: %d coefficient rows", n_terms))
    }
  }

  message("All H1 models fit and cached.")
}

# Helper to retrieve from cache (used by all downstream chunks)
get_h1_results <- function(outcome_var) {
  if (!outcome_var %in% names(model_cache)) {
    return(list(error = TRUE,
                message = sprintf("No cached results for %s.", outcome_var)))
  }
  model_cache[[outcome_var]]
}

# Helper: format H1 model results as a compact, readable table.
# Pivots wider so Baseline and Full appear side-by-side rather than stacked,
# strips the "origin_region" prefix from term labels, and drops row names.
format_h1_display <- function(res, family_type = "binomial") {

  origin_rows <- res %>%
    filter(grepl("^origin_region", term)) %>%
    mutate(Origin = gsub("^origin_region", "", term))

  if (nrow(origin_rows) == 0) return(NULL)

  if (family_type == "binomial") {
    # Wide format: one row per origin, columns for Baseline OR / CI / p
    # and Full OR / CI / p
    wide <- origin_rows %>%
      mutate(
        OR_str  = sprintf("%.2f", odds_ratio),
        CI_str  = sprintf("[%.2f, %.2f]", conf_low, conf_high),
        p_str   = formatC(p.value, format = "g", digits = 2)
      ) %>%
      select(Origin, Model, OR_str, CI_str, p_str) %>%
      tidyr::pivot_wider(
        names_from  = Model,
        values_from = c(OR_str, CI_str, p_str),
        names_glue  = "{Model}__{.value}"
      )

    # Build clean column ordering and labels
    base_label <- "Baseline (Demographics Only)"
    full_label <- "Full (With Human Capital)"

    cols_in_order <- c(
      "Origin",
      paste0(base_label, "__OR_str"),
      paste0(base_label, "__CI_str"),
      paste0(full_label, "__OR_str"),
      paste0(full_label, "__CI_str"),
      paste0(full_label, "__p_str")
    )
    cols_in_order <- intersect(cols_in_order, names(wide))
    wide <- wide[, cols_in_order, drop = FALSE]

    col_labels <- c("Origin",
                    "OR (Base)", "95% CI (Base)",
                    "OR (Full)", "95% CI (Full)",
                    "p (Full)")
    col_labels <- col_labels[seq_len(ncol(wide))]

    return(list(table = wide, col_labels = col_labels))

  } else {
    # OLS: report estimate and CI in coefficient units, not exponentiated
    wide <- origin_rows %>%
      mutate(
        Est_str = sprintf("%.2f", estimate),
        CI_str  = sprintf("[%.2f, %.2f]", conf_low, conf_high),
        p_str   = formatC(p.value, format = "g", digits = 2)
      ) %>%
      select(Origin, Model, Est_str, CI_str, p_str) %>%
      tidyr::pivot_wider(
        names_from  = Model,
        values_from = c(Est_str, CI_str, p_str),
        names_glue  = "{Model}__{.value}"
      )

    base_label <- "Baseline (Demographics Only)"
    full_label <- "Full (With Human Capital)"

    cols_in_order <- c(
      "Origin",
      paste0(base_label, "__Est_str"),
      paste0(base_label, "__CI_str"),
      paste0(full_label, "__Est_str"),
      paste0(full_label, "__CI_str"),
      paste0(full_label, "__p_str")
    )
    cols_in_order <- intersect(cols_in_order, names(wide))
    wide <- wide[, cols_in_order, drop = FALSE]

    col_labels <- c("Origin",
                    "β (Base)", "95% CI (Base)",
                    "β (Full)", "95% CI (Full)",
                    "p (Full)")
    col_labels <- col_labels[seq_len(ncol(wide))]

    return(list(table = wide, col_labels = col_labels))
  }
}
```


```{r forest-plot-all-regions, fig.height=6, fig.width=9, echo=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)

# Pull the nested model results for Net Worth Percentile from the cache
res_nw <- get_h1_results("nw_pctl")

if (!is.list(res_nw) || !isTRUE(res_nw$error)) {
  
  # Filter down to ALL origin region coefficients
  plot_data <- res_nw %>%
    filter(grepl("^origin_region", term)) %>%
    mutate(
      Region = gsub("origin_region", "", term)
    )
  
  # Sort the y-axis dynamically based on the baseline wealth gap
  # This makes the chart much cleaner to read
  baseline_order <- plot_data %>%
    filter(Model == "Baseline (Demographics Only)") %>%
    arrange(estimate) %>%
    pull(Region)
    
  plot_data <- plot_data %>%
    mutate(Region = factor(Region, levels = baseline_order))
  
  # Generate the Forest Plot
  p_all <- ggplot(plot_data, aes(x = estimate, y = Region, color = Model, shape = Model)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "black", linewidth = 0.8) +
    geom_pointrange(aes(xmin = conf_low, xmax = conf_high),
                    position = position_dodge(width = 0.5), size = 0.8, linewidth = 1) +
    theme_minimal(base_size = 12) +
    labs(
      title = "Origin-Region Effects on Net Worth Percentile",
      subtitle = "Testing segmented assimilation vs. human capital convergence (Ref: Latin American Origin)",
      x = "Coefficient Estimate (Difference in Net Worth Percentile)",
      y = NULL
    ) +
    scale_color_manual(values = c("Baseline (Demographics Only)" = "#E69F00", 
                                  "Full (With Human Capital)" = "#0072B2")) +
    theme(
      legend.position = "bottom", 
      legend.title = element_blank(),
      plot.title = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
    
  print(p_all)
} else {
  cat("Model failed to converge or run properly.")
}
```

```{r forest-plots-binary, fig.height=10, fig.width=9, echo=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)

# 1. Define the binary outcomes to loop through
binary_outcomes <- list(
  own_retire    = "Retirement Account (Parallel)",
  own_stocks    = "Direct Stocks (Parallel)",
  own_home      = "Homeownership (Selective)",
  credit_denied = "Credit Denied (Downward)",
  no_checking   = "Unbanked (Downward)"
)

# Initialize an empty dataframe to catch all the results
all_binary_results <- data.frame()

# 2. Run the nested models for each outcome and stack the data
for (var_name in names(binary_outcomes)) {
  label <- binary_outcomes[[var_name]]
  res <- get_h1_results(var_name)
  
  if (is.list(res) && !isTRUE(res$error)) {
    clean_res <- res %>%
      filter(grepl("^origin_region", term)) %>%
      mutate(
        Region = gsub("origin_region", "", term),
        Outcome = label
      )
    all_binary_results <- bind_rows(all_binary_results, clean_res)
  }
}

# 3. Sort the Regions for a clean visual (using Homeownership as the anchor)
if (nrow(all_binary_results) > 0) {
  baseline_order <- all_binary_results %>%
    filter(Outcome == "Homeownership (Selective)", Model == "Baseline (Demographics Only)") %>%
    arrange(odds_ratio) %>%
    pull(Region)
  
  all_binary_results <- all_binary_results %>%
    mutate(Region = factor(Region, levels = baseline_order))
  
  # 4. Generate the Faceted Forest Plot
  p_binary <- ggplot(all_binary_results, aes(x = odds_ratio, y = Region, color = Model, shape = Model)) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "black", linewidth = 0.8) +
    geom_pointrange(aes(xmin = conf_low, xmax = conf_high),
                    position = position_dodge(width = 0.5), size = 0.6, linewidth = 0.8) +
    facet_wrap(~Outcome, scales = "free_x", ncol = 2) +
    scale_x_log10() + # CRITICAL: Makes Odds Ratios visually symmetric around 1.0
    theme_minimal(base_size = 11) +
    labs(
      title = "Origin-Region Effects on Specific Financial Behaviors",
      subtitle = "Odds Ratios relative to Latin American Origin (x-axis is log-scaled)",
      x = "Odds Ratio (1 = No Difference from Latin American Origin)",
      y = NULL
    ) +
    scale_color_manual(values = c("Baseline (Demographics Only)" = "#E69F00", 
                                  "Full (With Human Capital)" = "#0072B2")) +
    theme(
      legend.position = "bottom", 
      legend.title = element_blank(),
      plot.title = element_text(face = "bold"),
      panel.grid.minor = element_blank(),
      strip.text = element_text(face = "bold", size = 11),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
  
  print(p_binary)
} else {
  cat("Binary models failed to converge.")
}
```




```{r h1-parallel, results='asis'}
if (n_fb >= 100) {

  cat("### Parallel-integration outcomes\n\n")

  for (info in list(
    list(var = "own_retire", label = "Retirement Account Ownership"),
    list(var = "own_stocks", label = "Direct Stock Ownership")
  )) {

    cat(sprintf("#### %s\n\n", info$label))

    res <- get_h1_results(info$var)

    if (is.list(res) && isTRUE(res$error)) {
      cat(sprintf("*Model failed: %s*\n\n", res$message))
      next
    }

    formatted <- format_h1_display(res, "binomial")
    if (is.null(formatted)) {
      cat("*No origin-region coefficients in output.*\n\n")
      next
    }

    print(kable(formatted$table,
                col.names = formatted$col_labels,
                row.names = FALSE,
                caption = sprintf("%s — Latin American origin = reference",
                                  info$label)))
    cat("\n\n")
  }
}
```

```{r h1-downward, results='asis'}
if (n_fb >= 100) {

  cat("### Downward-assimilation outcomes\n\n")

  for (info in list(
    list(var = "credit_denied", label = "Credit Denied or Got Less in Past Year"),
    list(var = "no_checking",   label = "No Checking Account")
  )) {

    cat(sprintf("#### %s\n\n", info$label))

    res <- get_h1_results(info$var)

    if (is.list(res) && isTRUE(res$error)) {
      cat(sprintf("*Model failed: %s*\n\n", res$message))
      next
    }

    formatted <- format_h1_display(res, "binomial")
    if (is.null(formatted)) {
      cat("*No origin-region coefficients in output.*\n\n")
      next
    }

    print(kable(formatted$table,
                col.names = formatted$col_labels,
                row.names = FALSE,
                caption = sprintf("%s — Latin American origin = reference",
                                  info$label)))
    cat("\n\n")
  }
}
```

```{r h1-selective, results='asis'}
if (n_fb >= 100) {

  cat("### Selective-assimilation outcomes\n\n")

  # ---- Homeownership ----
  cat("#### Homeownership\n\n")
  res_home <- get_h1_results("own_home")

  if (is.list(res_home) && isTRUE(res_home$error)) {
    cat(sprintf("*Model failed: %s*\n\n", res_home$message))
  } else {
    formatted_home <- format_h1_display(res_home, "binomial")
    if (!is.null(formatted_home)) {
      print(kable(formatted_home$table,
                  col.names = formatted_home$col_labels,
                  row.names = FALSE,
                  caption = "Homeownership — Latin American origin = reference"))
      cat("\n\n")
    }
  }

  # ---- Net Worth Percentile (OLS) ----
  cat("#### Net-Worth Percentile (linear regression on full SCF distribution)\n\n")
  res_nw <- get_h1_results("nw_pctl")

  if (is.list(res_nw) && isTRUE(res_nw$error)) {
    cat(sprintf("*Model failed: %s*\n\n", res_nw$message))
  } else {
    formatted_nw <- format_h1_display(res_nw, "ols")
    if (!is.null(formatted_nw)) {
      print(kable(formatted_nw$table,
                  col.names = formatted_nw$col_labels,
                  row.names = FALSE,
                  caption = "Net-Worth Percentile — Latin American origin = reference. Coefficients are percentile-point differences vs. the Latin American reference category, holding controls constant."))
      cat("\n\n")
    }
  }
}
```

## H2: Pathway-Pattern Tests

The three pathways predict joint patterns across outcomes within origin region. The
pattern matrix below summarizes origin-region odds ratios across the seven outcomes from
the H1 specifications. Pathway-coherent patterns — for example, Mexican-origin households
showing low retirement and stock ownership combined with high homeownership — would
support the segmented-assimilation hypothesis.

```{r h2-pattern-matrix, results='asis'}
if (n_fb >= 100) {

  # Re-fit all H1 models and collect the origin-region coefficients into a matrix
  outcomes_for_h2 <- list(
    list(var = "own_retire",    label = "Retirement",     family = "binomial"),
    list(var = "own_stocks",    label = "Stocks",         family = "binomial"),
    list(var = "credit_denied", label = "Credit denied",  family = "binomial"),
    list(var = "no_checking",   label = "No checking",    family = "binomial"),
    list(var = "own_home",      label = "Homeowner",      family = "binomial"),
    list(var = "nw_pctl",       label = "NW pctile",      family = "ols")
  )

  # Initialize the pattern matrix with the corrected origin labels
  origin_levels_no_ref <- c("Caribbean/West Indies",
                            "Asian",
                            "European/Other developed",
                            "Other")

  pattern_df <- data.frame(`Origin Region` = origin_levels_no_ref,
                           check.names = FALSE,
                           stringsAsFactors = FALSE)

  for (info in outcomes_for_h2) {
    res <- get_h1_results(info$var)

    if (is.list(res) && isTRUE(res$error)) {
      pattern_df[[info$label]] <- "--"
      next
    }

    # FIX: Isolate the Full model so we only get ONE coefficient per region
    res_full <- res %>% filter(Model == "Full (With Human Capital)")

    cell_values <- sapply(origin_levels_no_ref, function(lvl) {
      term_name <- paste0("origin_region", lvl)
      idx <- which(res_full$term == term_name)

      if (length(idx) == 0) return("--")

      est <- res_full$estimate[idx]
      pval <- res_full$p.value[idx]

      # For binomial: report odds ratio with stars
      # For OLS: report estimate with stars
      if (info$family == "binomial") {
        val <- exp(est)
        formatted <- sprintf("%.2f", val)
      } else {
        formatted <- sprintf("%.1f", est)
      }

      stars <- if (is.na(pval)) "" else
               if (pval < 0.001) "***" else
               if (pval < 0.01)  "**"  else
               if (pval < 0.05)  "*"   else ""

      paste0(formatted, stars)
    })

    pattern_df[[info$label]] <- cell_values
  }

  cat("**Table 3. Origin-region × outcome pattern matrix (Full Model)**\n\n")
  cat("Cells contain odds ratios (binary outcomes) or coefficient estimates ",
      "(net-worth percentile) relative to Latin American origin (reference). ",
      "Stars: \\* p<0.05, \\*\\* p<0.01, \\*\\*\\* p<0.001.\n\n", sep = "")
  print(kable(pattern_df, row.names = FALSE,
              caption = "Origin-region effects across outcomes (net of human capital)"))
  cat("\n\n")
}
```

## H3: Years-in-U.S. × Origin Interaction

Interaction models test whether duration-of-residence effects on financial outcomes
differ by origin region. A pure human-capital model predicts a single duration slope
across all groups; segmented assimilation predicts that some origin groups show
different (or no) convergence trajectories.

```{r h3-interaction, results='asis'}
if (n_fb >= 100) {

  cat("**Table 4. Origin × Years-in-U.S. interaction tests**\n\n")
  cat("For each outcome, we report each individual Origin × Years-in-U.S. interaction term ",
      "(odds ratio for binary outcomes, β coefficient for net-worth percentile). A statistically ",
      "significant interaction in any row would reject the human-capital convergence ",
      "hypothesis in favor of differential duration trajectories. ",
      "The years-in-U.S. main-effect slope (Latin American reference) is also reported.\n\n")

  # Pre-check: do all origin × outcome cells have at least 5 obs and within-cell variation?
  # Quasi-complete separation is the most common cause of model failure.
  fb_imp1 <- scf_fb$mi_design[[1]]$variables

  cell_check <- fb_imp1 %>%
    filter(!is.na(origin_region)) %>%
    group_by(origin_region) %>%
    summarise(
      n = n(),
      retire_var    = var(own_retire,    na.rm = TRUE),
      stocks_var    = var(own_stocks,    na.rm = TRUE),
      denied_var    = var(credit_denied, na.rm = TRUE),
      nochk_var     = var(no_checking,   na.rm = TRUE),
      home_var      = var(own_home,      na.rm = TRUE),
      .groups = "drop"
    )

  # Identify outcomes with separation risk: cell with zero variance
  has_separation <- function(outcome_var) {
    var_col <- switch(outcome_var,
      own_retire    = "retire_var",
      own_stocks    = "stocks_var",
      credit_denied = "denied_var",
      no_checking   = "nochk_var",
      own_home      = "home_var",
      NA
    )
    if (is.na(var_col)) return(FALSE)
    any(cell_check[[var_col]] == 0 | is.na(cell_check[[var_col]]))
  }

  h3_long_results <- list()

  for (info in list(
    list(var = "own_retire",    label = "Retirement",     family = "binomial"),
    list(var = "own_stocks",    label = "Stocks",         family = "binomial"),
    list(var = "credit_denied", label = "Credit denied",  family = "binomial"),
    list(var = "no_checking",   label = "No checking",    family = "binomial"),
    list(var = "own_home",      label = "Homeowner",      family = "binomial"),
    list(var = "nw_pctl",       label = "NW pctile",      family = "ols")
  )) {

    # Skip outcomes with quasi-complete separation
    if (info$family == "binomial" && has_separation(info$var)) {
      h3_long_results[[info$label]] <- list(
        skipped = TRUE,
        reason  = "Quasi-complete separation: at least one origin region has zero within-cell variance on this outcome. Interaction model would not converge."
      )
      next
    }

    # V8 fix: replaced yiu_clean + yiu_missing with raw years_in_us
    int_rhs <- "origin_region * years_in_us + age_centered + educ_cat + currently_married + log_income"
    int_formula <- as.formula(paste(info$var, "~", int_rhs))
    environment(int_formula) <- globalenv()    # V7 fix: same as H1

    fit <- tryCatch({
      if (info$family == "binomial") {
        scf_glm(scf_fb, int_formula, family = binomial())
      } else {
        scf_ols(scf_fb, int_formula)
      }
    }, error = function(e) list(error = TRUE, message = conditionMessage(e)))

    if (is.list(fit) && isTRUE(fit$error)) {
      h3_long_results[[info$label]] <- list(error = TRUE, message = fit$message)
      next
    }

    h3_long_results[[info$label]] <- list(results = fit$results, family = info$family)
  }

  # Print results, one outcome at a time
  for (label in names(h3_long_results)) {
    cat(sprintf("#### %s\n\n", label))

    entry <- h3_long_results[[label]]

    if (isTRUE(entry$skipped)) {
      cat(sprintf("*Skipped — %s*\n\n", entry$reason))
      next
    }

    if (isTRUE(entry$error)) {
      cat(sprintf("*Model failed: %s*\n\n", entry$message))
      next
    }

    res <- entry$results
    fam <- entry$family

    # Find both the YIU main effect and all origin × YIU interaction terms
    yiu_main <- res %>% filter(term == "years_in_us")
    int_terms <- res %>%
      filter(grepl("origin_region.*:years_in_us|years_in_us:origin_region", term))

    if (nrow(int_terms) == 0) {
      cat("*No interaction terms found.*\n\n")
      next
    }

    # Format compactly
    if (fam == "binomial") {
      display_main <- yiu_main %>%
        mutate(Term = "Years in U.S. (Latin American reference slope)",
               Estimate = sprintf("OR = %.3f", exp(estimate)),
               `95% CI` = sprintf("[%.3f, %.3f]",
                                   exp(estimate - 1.96 * std.error),
                                   exp(estimate + 1.96 * std.error)),
               `p-value` = formatC(p.value, format = "g", digits = 2)) %>%
        select(Term, Estimate, `95% CI`, `p-value`)

      display_int <- int_terms %>%
        mutate(Term = paste(gsub("origin_region|:years_in_us|years_in_us:", "", term), "× YIU"),
               Estimate = sprintf("OR = %.3f", exp(estimate)),
               `95% CI` = sprintf("[%.3f, %.3f]",
                                   exp(estimate - 1.96 * std.error),
                                   exp(estimate + 1.96 * std.error)),
               `p-value` = formatC(p.value, format = "g", digits = 2)) %>%
        select(Term, Estimate, `95% CI`, `p-value`)
    } else {
      display_main <- yiu_main %>%
        mutate(Term = "Years in U.S. (Latin American reference slope)",
               Estimate = sprintf("β = %.3f", estimate),
               `95% CI` = sprintf("[%.3f, %.3f]",
                                   estimate - 1.96 * std.error,
                                   estimate + 1.96 * std.error),
               `p-value` = formatC(p.value, format = "g", digits = 2)) %>%
        select(Term, Estimate, `95% CI`, `p-value`)

      display_int <- int_terms %>%
        mutate(Term = paste(gsub("origin_region|:years_in_us|years_in_us:", "", term), "× YIU"),
               Estimate = sprintf("β = %.3f", estimate),
               `95% CI` = sprintf("[%.3f, %.3f]",
                                   estimate - 1.96 * std.error,
                                   estimate + 1.96 * std.error),
               `p-value` = formatC(p.value, format = "g", digits = 2)) %>%
        select(Term, Estimate, `95% CI`, `p-value`)
    }

    combined <- bind_rows(display_main, display_int)
    print(kable(combined, row.names = FALSE,
                caption = sprintf("%s: Years-in-U.S. main effect and origin × YIU interactions", label)))
    cat("\n\n")
  }

  cat("Significant interaction terms indicate that the duration-of-residence slope ",
      "for that origin group differs from the Latin American (reference) slope, supporting ",
      "the segmented-assimilation prediction of differential trajectories.\n\n", sep = "")
}
```

# Robustness Checks

## Three-Category Origin Recode

If any cell in the 5-category origin coding falls below n = 30, we re-estimate models
with a collapsed 3-category coding (Latin American, Asian, European/Other) for stability.

## Race Versus Origin

Origin region and racial classification (`racecl4`) are correlated but not identical.
Specifications adding `race_cat` to the H1 models test whether origin-region effects are
robust to controlling for racial classification, and whether racial classification
mediates origin-region effects.

## Net-Worth Specification

The primary net-worth specification uses percentile within the full SCF distribution. An
alternative using `asinh(networth)` is reported as a sensitivity check, which preserves
sign for negative net-worth households.

# Discussion

The findings will be interpreted under three headings:

**Are origin-region differences durable?** A finding that origin-region coefficients
remain large and significant after controlling for years-in-U.S., age, education, and
income would support segmented-assimilation theory over a pure human-capital convergence
model. A finding that origin-region coefficients vanish when controls are added would
support the convergence model.

**Pathway-coherent or scattered patterns?** Segmented assimilation predicts coherent
within-group patterns. Mexican-origin households should show downward-assimilation
barriers (high credit denial, low retirement-account ownership) combined with
selective-assimilation asset accumulation (relatively high homeownership). Asian-origin
households should look like the parallel-integration profile (high mainstream
participation). Patterns that scatter across pathways without group coherence weaken the
segmented-assimilation interpretation.

**Implications for policy and theory.** Wealth and credit-access disparities have
generally been studied through the lens of race rather than origin. The 2022 SCF allows
the field to begin testing whether origin-region heterogeneity exists *within* racialized
groups — for example, whether Mexican-origin and Cuban-origin Hispanic households show
different financial patterns, a critical refinement that earlier surveys could not
support.

# Limitations

1. **Single-wave cross-section.** The 2022 SCF is the first wave with these immigration
questions. No time series is available; cohort and period effects cannot be separated.

2. **No parental nativity.** Unlike the NSFG used in Papers 3 and 4, the SCF does not
collect parental nativity. Generation-1 vs. generation-2 distinctions are not directly
observable.

3. **Origin-region collapsing.** Sample sizes preclude finer-grained country-of-origin
analysis. Heterogeneity within "Asian" or "Other Latin American" categories may obscure
pathway-specific patterns.

4. **Selection into U.S. residence.** Foreign-born households observed in 2022 are
survivors of immigration selection processes. Observed financial patterns reflect this
selection alongside assimilation processes.

5. **Disclosure-avoidance modifications.** SCF public-use data are altered for disclosure
protection, which may attenuate small-cell estimates for the smallest origin groups.

# Conclusion

This paper applies the segmented-assimilation framework to U.S. household financial
outcomes for the first time using nationally representative data with origin-region
detail. The analysis tests whether origin region matters for financial-system
incorporation above and beyond duration-of-residence and individual human-capital
characteristics, and whether the patterns observed are coherent with the three theoretical
pathways — parallel integration, downward assimilation, selective assimilation — or
scattered across them.

The contribution to the broader thesis is to close the loop on the immigration question.
Earlier papers established that nativity matters for fertility outcomes (Paper 3) and for
unplanned childlessness (Paper 4). Paper 5 returns to the SCF financial-outcome framework
that anchors Papers 1 and 2 and tests whether nativity — refined into origin region —
also structures the financial system itself.

# Appendix A: Variable List

| Construct | Source variable | Type |
|---|---|---|
| Foreign-born status (component A) | `X6915` (ancestry; non-0 and non-1200) | Raw codebook |
| Foreign-born status (component B) | `X6906` (years respondent has lived in U.S.) | Raw codebook |
| National origin (R) | `X6915` | Raw codebook |
| Spouse Hispanic origin (R) | `X7404` (auxiliary, not used in main models) | Raw codebook |
| Race/ethnicity | `racecl4` | Bulletin extract |
| Age | `age` | Bulletin extract |
| Education | `educ` (recoded into 5 levels) | Bulletin extract |
| Marital status | `married` | Bulletin extract |
| Income | `income` | Bulletin extract |
| Net worth | `networth` | Bulletin extract |
| Net-worth percentile | `nw_pctl` (computed on full SCF distribution) | Derived |
| Retirement assets | `retqliq` | Bulletin extract |
| Stocks | `stocks` | Bulletin extract |
| Credit denied | `turndown` | Bulletin extract |
| Checking balance | `checking` | Bulletin extract |
| Homeowner | `housecl` | Bulletin extract |
| Weight | `wgt` | Bulletin extract |

# Appendix B: scf Package Workflow Notes

This document follows the workflow established in Paper 1 (`scf_ccbal_analysis_v5.Rmd`):

1. `scf_download(2022)` writes raw `.rds` files to `scf_data/`.
2. `scf_load(2022, data_directory = "scf_data")` returns an object with
   `$mi_design[[1..5]]` survey-design slots, one per implicate.
3. `scf_update(scf_obj, new_var = expr, ...)` adds variables to all 5 implicates.
4. `scf_mean()`, `scf_glm()`, `scf_ols()` fit pooled models with Rubin's-Rules pooling
   and replicate-weight-based variance.
5. Subsetting (e.g. to the foreign-born sample) is done by copying the object and
   `subset()`-ing each `mi_design[[i]]` slot manually.

The new 2022 immigration variables (`X6915`, `X6906`, `X7404`, `X7796`) are not in the
Bulletin summary extract that the `scf` package builds. The diagnostic chunk at the
top of this document attempts to merge them in from the raw 2022 SCF `.rds` file. If
the merge fails, those analyses cannot proceed.

# Appendix C: V6 Methodological Notes

This V6 build corrects a critical error in V5 plus carries forward the V5
methodological refinements:

**V6 (this build) correction — years-in-U.S. variable identity.**

V5 used `X7437`–`X7443` as the years-lived-in-U.S. variables, which produced
an analytic sample of only 75 foreign-born respondents — implausibly small.
The official Federal Reserve "2022 SCF changes" file
(`https://www.federalreserve.gov/econres/files/2022_scf_changes.txt`) reveals
that `X7437`–`X7443` are **mortgage forbearance** variables (which mortgage
terms were adjusted after exiting forbearance), not immigration variables.
The correct years-in-U.S. variables are **X6906** (respondent) and **X7796**
(spouse). V6 uses X6906 for the foreign-born identification, which should
produce an analytic sample of approximately 600-800 foreign-born respondents
matching ACS proportions of the U.S. foreign-born population.

**V5 corrections (carried forward):**

1. **Foreign-born identification corrected.** V3 defined foreign-born as
   `X6915 != 0`, which incorrectly flagged U.S.-born respondents reporting
   European, "American," and African ancestry as immigrants. V5/V6 require
   both `X6915` indicating non-American ancestry AND `X6906 > 0`.

2. **Net-worth percentile computed on full SCF distribution.** V3 computed
   `nw_pctl` inside `scf_fb` (the foreign-born subset). V5/V6 compute
   `nw_pctl` via `scf_update` on `scf_2022` *before* subsetting.

3. **Quasi-complete separation pre-checked.** V3 binary outcomes produced
   models with standard errors of $10^{12}$ from quasi-complete separation.
   V5/V6 include a pre-check in the cache loop and the H3 chunk; if any
   origin region has zero variance on the outcome, the model is skipped.

4. **H3 interaction errors surfaced.** V3 silently swallowed all H3 model
   errors. V5/V6 capture error messages and report them per-outcome, and
   report each interaction term individually rather than computing a
   misleading "joint p-value" as the minimum across terms.

5. **Tables reformatted for readability.** V3 produced 9-column tables with
   stacked Baseline/Full rows and unbroken `origin_region` prefix strings.
   V5/V6 use a `format_h1_display` helper that pivots Baseline and Full
   side-by-side as columns, strips the `origin_region` prefix, and reports
   one row per origin region. Table width drops from 9 columns to 6.
