---
title: "Segmented Assimilation in the American Financial System: Origin-Based Gradients in Immigrant Household Finances"
subtitle: "Evidence from the 2022 Survey of Consumer Finances"
author: "Delroy Nichols"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 5, dpi = 150
)
rm(list = ls())
gc()
set.seed(42)
```

# Origin-Based Gradients in Immigrant Household Finances

## The Central Question

Segmented assimilation theory predicts that immigrants follow distinct trajectories of
incorporation depending on origin, human capital, and reception context. Three pathways
are commonly identified: *parallel integration* into the white middle class, *downward
assimilation* into the racialized urban underclass, and *selective assimilation* through
ethnic-community resources. While the fertility, educational, and occupational
implications of segmented assimilation have been studied extensively, financial-system
incorporation has received less empirical attention — in part because nationally
representative wealth surveys have lacked detailed national-origin information.

The 2022 Survey of Consumer Finances introduced new questions on respondent and spouse
national origin and years lived in the United States, allowing the first nationally
representative test of segmented-assimilation predictions on financial outcomes for U.S.
household wealth. This paper uses these new variables to test whether financial outcomes
among foreign-born households vary systematically by origin region in ways consistent
with the three-pathway theoretical framework.

**Data Source:** 2022 Survey of Consumer Finances (SCF), Federal Reserve Board. All
estimates are survey-weighted, and all models are pooled across the SCF's five multiply-
imputed datasets using Rubin's Rules.

**Note on Scope:** All findings are correlational and limited to the foreign-born
respondent sample. The SCF's cross-sectional design and absence of parental nativity
preclude causal claims and intergenerational analysis.

## Why This Matters: Beyond Race-Only Frameworks

Wealth and credit-access disparities have generally been studied through the lens of
race rather than origin. Studies of Black-white and Hispanic-white wealth gaps treat the
groups they compare as internally homogeneous — but Black households of Caribbean origin,
Mexican-origin Hispanics, and Cuban-origin Hispanics have very different financial
profiles, shaped by distinct migration histories, reception contexts, and community
resources. Pooling them into single racial categories obscures rather than illuminates
the structural processes that segmented-assimilation theory was developed to explain.

The 2022 SCF, by adding origin-region detail, lets the field begin testing whether the
segmented-assimilation pathways — parallel integration, downward assimilation, selective
assimilation — show up in wealth and credit data. Mexican-origin households may show low
mainstream-finance participation alongside high homeownership (selective assimilation
through co-ethnic mortgage networks). Asian-origin households may show elevated
mainstream participation (parallel integration). These pathway-coherent patterns are
testable with the new data, and they have direct policy implications: interventions
designed for "Hispanic households" or "Asian households" miss the heterogeneity that
actually drives outcomes.

---

```{r load-packages}
required_packages <- c(
  "tidyverse", "haven", "survey", "labelled",
  "ggplot2", "scales", "scf",
  "broom", "knitr", "moments"
)

for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg, repos = "http://cran.us.r-project.org")
    library(pkg, character.only = TRUE)
  }
}
```

```{r download-data}
# Same download pattern as the credit-card-debt analysis (Paper 1).
# scf_download writes .rds files to the current working directory.

dir.create("scf_data", showWarnings = FALSE, recursive = TRUE)
original_wd <- getwd()
setwd("scf_data")

if (!file.exists("scf2022.rds")) {
  tryCatch({
    scf_download(2022)
  }, error = function(e) {
    cat("Error downloading 2022 data:", conditionMessage(e), "\n")
  })
}

setwd(original_wd)
```

```{r load-data}
scf_2022 <- scf_load(2022, data_directory = "scf_data")
```

# Variable Availability and Diagnostic

Before constructing variables, we check which of the new 2022 immigration variables are
actually exposed in the loaded `scf` package object. The standard SCF Bulletin extract
that the `scf` package builds does not by default include the new national-origin or
years-in-U.S. items (`X7404`, `X7437`–`X7443`); these are raw codebook items rather than
constructed Bulletin variables. If they are missing, we read the raw 2022 SCF `.rds`
file directly and merge them in by household ID.

```{r diagnose-immigration-vars, results='asis'}
imp1 <- scf_2022$mi_design[[1]]$variables

# Variables we need
imm_vars <- c("X7404", "X7437", "X7438", "X7439", "X7440",
              "X7441", "X7442", "X7443", "X8000", "X8049")

present_in_extract <- imm_vars %in% names(imp1)

cat("**Immigration-variable availability in the scf package extract:**\n\n")
diag_df <- data.frame(
  Variable = imm_vars,
  `In extract` = ifelse(present_in_extract, "Yes", "No"),
  check.names = FALSE
)
print(kable(diag_df))
cat("\n\n")

need_raw_merge <- !all(present_in_extract)

if (need_raw_merge) {
  cat("Some required immigration variables are missing from the bulletin extract. ",
      "Attempting to merge them in from the raw 2022 SCF data file...\n\n", sep = "")
} else {
  cat("All required immigration variables are present in the extract. ",
      "No raw merge needed.\n\n", sep = "")
}
```

```{r merge-raw-immigration-vars}
# ============================================================================
# Merge raw immigration variables from the underlying SCF .rds file
# ============================================================================
# The scf package downloads the raw 2022 SCF data into scf_data/ as well as
# building the bulletin extract. The raw file contains all X-variables; we
# read it, select just the immigration-related vars + ID, and merge into
# each implicate's $variables data frame.
#
# Common file names produced by scf_download():
#   - p22i6.rds (raw public-use, 5 implicates stacked, with implicate id)
#   - scf2022.rds (the bulletin extract)
#
# We try a few candidate filenames defensively.

if (need_raw_merge) {

  raw_candidates <- c(
    file.path("scf_data", "p22i6.rds"),
    file.path("scf_data", "scfp2022.rds"),
    file.path("scf_data", "p22.rds"),
    "p22i6.rds",
    "scfp2022.rds"
  )

  raw_path <- raw_candidates[file.exists(raw_candidates)][1]

  if (!is.na(raw_path)) {
    raw_data <- readRDS(raw_path)

    # Standardize variable name casing — the raw file may have either
    # uppercase or lowercase. We try uppercase first since X-variables
    # are conventionally uppercase.
    raw_names <- names(raw_data)
    name_map <- setNames(raw_names, toupper(raw_names))

    available_imm <- imm_vars[toupper(imm_vars) %in% toupper(raw_names)]

    # The raw data has 5 implicates stacked. We need an ID that matches
    # the bulletin extract's row ordering. Common candidates: y1, yy1, x1.
    id_candidates <- c("y1", "yy1", "Y1", "YY1", "X1")
    id_var <- id_candidates[id_candidates %in% raw_names][1]

    if (!is.na(id_var) && length(available_imm) > 0) {

      # Pull the matching columns
      raw_imm <- raw_data[, c(id_var, available_imm), drop = FALSE]

      # Determine implicate count: raw file has each household 5 times (one
      # per implicate) — split into 5 chunks of equal size in row order.
      n_total <- nrow(raw_imm)
      n_per_imp <- nrow(scf_2022$mi_design[[1]]$variables)

      if (n_total == 5 * n_per_imp) {
        # Standard layout: implicates stacked
        for (i in 1:5) {
          start_row <- (i - 1) * n_per_imp + 1
          end_row   <- i * n_per_imp
          imm_chunk <- raw_imm[start_row:end_row, available_imm, drop = FALSE]

          # Add to the implicate's $variables
          for (v in available_imm) {
            scf_2022$mi_design[[i]]$variables[[v]] <- imm_chunk[[v]]
          }
        }
        cat("Successfully merged", length(available_imm), "immigration variables ",
            "from", raw_path, "into all 5 implicates.\n", sep = "")
      } else {
        cat("Row-count mismatch: raw file has ", n_total,
            " rows, bulletin extract has ", n_per_imp,
            " rows per implicate. Merge aborted; using available variables only.\n",
            sep = "")
      }
    } else {
      cat("Could not locate raw SCF file or matching ID variable. ",
          "Immigration variables will be unavailable.\n", sep = "")
    }
  } else {
    cat("Raw 2022 SCF .rds file not found in scf_data/. ",
        "Available files:\n", sep = "")
    cat(paste(list.files("scf_data"), collapse = "\n"), "\n\n")
  }
}

# Re-check what's now available
imp1 <- scf_2022$mi_design[[1]]$variables
have_X7404 <- "X7404" %in% names(imp1)
have_yiu   <- any(c("X7437", "X7438", "X7439", "X7440",
                    "X7441", "X7442", "X7443") %in% names(imp1))
have_X8000 <- "X8000" %in% names(imp1)
have_X8049 <- "X8049" %in% names(imp1)
```

```{r raw-x7404-frequency, results='asis'}
# ============================================================================
# Print raw X7404 value frequencies — needed to calibrate the recode
# ============================================================================
# Until we see what values X7404 actually takes in this wave, we cannot
# write the correct recoding rule. This chunk dumps the unweighted frequency
# table so we can adjust the recode.

if (have_X7404) {
  x7404_freq <- scf_2022$mi_design[[1]]$variables %>%
    count(X7404, name = "n") %>%
    arrange(desc(n))

  cat("**Raw X7404 frequency (Implicate 1, unweighted):**\n\n")
  print(kable(x7404_freq, caption = "X7404 = respondent national origin"))
  cat("\n\nThis table guides the origin-region recode in the next chunk. ",
      "Code 0 typically indicates 'inapplicable' (i.e., U.S.-born or non-respondent).\n\n",
      sep = "")
}
```

# Variable Engineering

```{r variable-engineering}
# ============================================================================
# Build all analysis variables using scf_update — same pattern as Paper 1
# ============================================================================
# scf_update takes the scf_2022 object and adds variables to all 5 implicates
# at once. Variable names from the bulletin extract are lowercase
# (income, networth, ccbal, kids, edcl, racecl4, hhsex, age, married, etc.)

# ---- Step 1: Foreign-born status from years-in-US variables ----
# X7437–X7443 are reported across multiple roster positions; we take the
# parallel maximum (a non-zero value indicates foreign-born status).

if (have_yiu) {
  # Build the pmax expression dynamically over whichever YIU variables
  # actually loaded.
  yiu_vars_present <- c("X7437", "X7438", "X7439", "X7440",
                        "X7441", "X7442", "X7443")
  yiu_vars_present <- yiu_vars_present[yiu_vars_present %in% names(imp1)]

  scf_2022 <- scf_update(scf_2022,
    years_in_us_raw = suppressWarnings(
      do.call(pmax, c(mget(yiu_vars_present), list(na.rm = TRUE)))
    )
  )

  scf_2022 <- scf_update(scf_2022,
    years_in_us  = ifelse(years_in_us_raw <= 0 | is.infinite(years_in_us_raw),
                          NA_real_, years_in_us_raw),
    foreign_born = as.integer(!is.na(years_in_us) & years_in_us > 0)
  )
} else {
  scf_2022 <- scf_update(scf_2022,
    years_in_us  = NA_real_,
    foreign_born = 0L
  )
}

# ---- Step 2: Origin-region recode from X7404 ----
# IMPORTANT: this is a placeholder mapping. After the first knit, examine
# the X7404 frequency table printed above and adjust the case_when ranges
# to match the actual codebook. The five-category structure is fixed by
# thesis design; only the numeric ranges below need calibration.

if (have_X7404) {
  scf_2022 <- scf_update(scf_2022,
    origin_region = factor(case_when(
      X7404 %in% 100:199 ~ "Mexican",
      X7404 %in% 200:399 ~ "Other Latin American/Caribbean",
      X7404 %in% 400:599 ~ "Asian",
      X7404 %in% 600:799 ~ "European/Other developed",
      X7404 %in% 800:999 ~ "Other",
      TRUE               ~ NA_character_
    ), levels = c("Mexican",
                  "Other Latin American/Caribbean",
                  "Asian",
                  "European/Other developed",
                  "Other"))
  )
} else {
  scf_2022 <- scf_update(scf_2022,
    origin_region = factor(NA_character_,
      levels = c("Mexican",
                 "Other Latin American/Caribbean",
                 "Asian",
                 "European/Other developed",
                 "Other"))
  )
}

# ---- Step 3: Race/ethnicity (existing bulletin variable + swap correction) ----
# The bulletin extract supplies racecl4 directly (1=White, 2=Black, 3=Hispanic,
# 4=Other). The X8000=1 swap correction is already applied inside the bulletin
# macro for the public file, so racecl4 is already corrected.

scf_2022 <- scf_update(scf_2022,
  race_cat = factor(racecl4,
    levels = 1:4,
    labels = c("White", "Black", "Hispanic", "Other")
  )
)

# ---- Step 4: Standard demographic controls ----
scf_2022 <- scf_update(scf_2022,
  age_centered = age - 50,
  age_sq       = age^2,
  in_age_range = as.integer(age >= 25 & age <= 75),

  educ_cat = factor(edcl,
    levels = 1:4,
    labels = c("Less than HS", "High School", "Some College", "College Degree")
  ),

  female            = as.integer(hhsex == 2),
  currently_married = as.integer(married == 1),
  has_children      = as.integer(kids > 0),
  is_homeowner      = as.integer(housecl == 1),
  has_pension       = as.integer(anypen == 1),

  log_income        = log(pmax(1, income)),
  asinh_networth    = asinh(networth),
  log_wageinc       = log(pmax(1, wageinc))
)

# ---- Step 5: Outcomes — three-pathway design ----
# All outcomes are constructed from bulletin-extract variables that exist
# in every wave, so no presence checks needed.

scf_2022 <- scf_update(scf_2022,

  # Parallel integration: mainstream financial-system participation
  own_retire = as.integer(retqliq > 0),       # any retirement-account assets
  own_stocks = as.integer(stocks > 0),         # direct stock ownership

  # Downward assimilation: structural barriers
  credit_denied = as.integer(turndown == 1),   # turned down or got less in past 12 mo
  no_checking   = as.integer(checking <= 0),   # no checking-account balance

  # Selective assimilation: asset accumulation
  own_home  = as.integer(housecl == 1),        # owns primary residence
  nw_pctl   = 100 * percent_rank(networth)     # net-worth percentile, full sample
)

# ---- Step 6: Analytic sample flag ----
scf_2022 <- scf_update(scf_2022,
  in_analytic_sample = as.integer(
    foreign_born == 1 &
    !is.na(origin_region) &
    in_age_range == 1
  )
)
```

```{r build-foreign-born-design}
# ============================================================================
# Build the foreign-born subset design — same pattern as scf_debtors in Paper 1
# ============================================================================
# We copy the full design and subset each implicate manually, since scf
# package doesn't expose a built-in subset method.

scf_fb <- scf_2022
for (i in 1:5) {
  idx <- scf_2022$mi_design[[i]]$variables$in_analytic_sample == 1
  idx[is.na(idx)] <- FALSE
  scf_fb$mi_design[[i]] <- subset(scf_2022$mi_design[[i]], idx)
}

# Diagnostic: report sample size
n_fb <- nrow(scf_fb$mi_design[[1]]$variables)
```

# Descriptive Findings

## Foreign-Born Sample Composition

```{r desc-sample-comp, results='asis'}
if (n_fb > 0) {
  fb_imp1 <- scf_fb$mi_design[[1]]$variables

  mean_age   <- weighted.mean(fb_imp1$age, fb_imp1$wgt, na.rm = TRUE)
  mean_yiu   <- weighted.mean(fb_imp1$years_in_us, fb_imp1$wgt, na.rm = TRUE)
  mean_inc   <- weighted.mean(fb_imp1$income, fb_imp1$wgt, na.rm = TRUE)
  median_nw  <- median(fb_imp1$networth, na.rm = TRUE)

  cat(sprintf("The analytic sample comprises %s foreign-born respondents in the 2022 SCF aged 25-75. ",
              formatC(n_fb, big.mark = ",")))
  cat(sprintf("Using survey weights, the mean age in this population is %.1f years, with a mean of %.1f years lived in the U.S. ",
              mean_age, mean_yiu))
  cat(sprintf("Mean household income is $%s and median net worth is $%s.\n\n",
              formatC(round(mean_inc), big.mark = ","),
              formatC(round(median_nw), big.mark = ",")))
} else {
  cat("**Analytic sample is empty.** The most likely cause is that the X7404 origin-region recode mapping does not match the actual codebook values. Check the X7404 frequency table printed in the diagnostic chunk and update the `case_when` ranges in the variable-engineering chunk.\n\n")
}
```

## Origin-Region Distribution

```{r desc-origin-table, results='asis'}
if (n_fb > 0) {
  fb_imp1 <- scf_fb$mi_design[[1]]$variables

  origin_summary <- fb_imp1 %>%
    filter(!is.na(origin_region)) %>%
    group_by(origin_region) %>%
    summarise(
      n_unweighted   = n(),
      weighted_count = sum(wgt, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      pct_weighted = round(weighted_count / sum(weighted_count) * 100, 1)
    ) %>%
    select(origin_region, n_unweighted, pct_weighted)

  cat("**Table 1. Foreign-born sample by origin region (2022 SCF)**\n\n")
  print(kable(origin_summary,
              col.names = c("Origin Region", "n (Unweighted)", "% (Weighted)"),
              caption = "Note: percentages are SCF-weighted; n is unweighted (Implicate 1)."))
  cat("\n\n")

  small_cells <- origin_summary %>% filter(n_unweighted < 30)
  if (nrow(small_cells) > 0) {
    cat(sprintf("**Warning:** Small unweighted cell sizes (< 30) in %d origin region(s): %s. Model estimates for these groups may be unstable.\n\n",
                nrow(small_cells),
                paste(as.character(small_cells$origin_region), collapse = ", ")))
  }
}
```

## Outcome Prevalence by Origin Region

```{r desc-outcomes-by-origin, results='asis'}
if (n_fb > 0) {

  # Use scf_mean for each outcome (matches Paper 1 pattern)
  outcomes <- list(
    own_retire    = "Retirement (%)",
    own_stocks    = "Stocks (%)",
    credit_denied = "Credit denied (%)",
    no_checking   = "No checking (%)",
    own_home      = "Homeowner (%)",
    nw_pctl       = "NW pctile (mean)"
  )

  # Run each prevalence estimate using scf_mean with by = ~origin_region
  prevalence_list <- list()
  for (out_var in names(outcomes)) {
    f <- as.formula(paste0("~", out_var))
    res <- tryCatch({
      scf_mean(scf_fb, f, by = ~origin_region)$results
    }, error = function(e) NULL)
    prevalence_list[[out_var]] <- res
  }

  # Build display table — one row per origin region
  origin_levels <- levels(scf_fb$mi_design[[1]]$variables$origin_region)
  display_df <- data.frame(`Origin Region` = origin_levels,
                           check.names = FALSE,
                           stringsAsFactors = FALSE)

  for (out_var in names(outcomes)) {
    res <- prevalence_list[[out_var]]
    col_label <- outcomes[[out_var]]

    if (is.null(res) || nrow(res) == 0) {
      display_df[[col_label]] <- "--"
      next
    }

    # The scf_mean result has columns: by-variable level, estimate, std.error
    # Match by origin_region level
    estimates <- sapply(origin_levels, function(lvl) {
      idx <- which(as.character(res[[1]]) == lvl)
      if (length(idx) == 0) return(NA_real_)
      # The estimate column is typically the second column or named "estimate"
      if ("estimate" %in% names(res)) res$estimate[idx] else res[[2]][idx]
    })

    if (out_var == "nw_pctl") {
      display_df[[col_label]] <- ifelse(is.na(estimates), "--",
                                         sprintf("%.1f", estimates))
    } else {
      display_df[[col_label]] <- ifelse(is.na(estimates), "--",
                                         sprintf("%.1f", estimates * 100))
    }
  }

  cat("**Table 2. Weighted prevalence of financial outcomes by origin region**\n\n")
  print(kable(display_df,
              caption = "All estimates are SCF-weighted and pooled across implicates."))
  cat("\n\n")
}
```

## Years-in-U.S. Distribution by Origin Region

```{r desc-yiu-plot, fig.cap="Distribution of years lived in the U.S. by origin region (foreign-born SCF sample, 2022)"}
if (n_fb > 0) {
  plot_data <- scf_fb$mi_design[[1]]$variables %>%
    filter(!is.na(years_in_us), !is.na(origin_region))

  if (nrow(plot_data) > 0) {
    ggplot(plot_data,
           aes(x = origin_region, y = years_in_us, fill = origin_region)) +
      geom_boxplot(alpha = 0.7, outlier.shape = NA) +
      geom_jitter(width = 0.15, alpha = 0.2, size = 0.6) +
      labs(x = NULL,
           y = "Years lived in the U.S.",
           subtitle = "Unweighted distributional shape (Implicate 1)") +
      theme_minimal(base_size = 11) +
      theme(legend.position = "none",
            axis.text.x = element_text(angle = 30, hjust = 1)) +
      scale_fill_brewer(palette = "Set2")
  }
}
```

# Modeling Results

## H1: Origin-Region Main Effects

We test whether financial outcomes among foreign-born respondents differ by origin
region net of years-in-U.S., age, education, marital status, and income. Logistic
regression for binary outcomes (`scf_glm` with `binomial()`) and OLS for the continuous
net-worth-percentile outcome (`scf_ols`). All models pool across the 5 implicates.

Mexican origin is the reference category — coefficients indicate how each origin region
compares to Mexican-origin foreign-born respondents.

```{r h1-model-helper}
# ============================================================================
# Helper: fit and tidy an scf_glm or scf_ols model
# ============================================================================
# Matches the pattern in Paper 1: scf_glm(scf_obj, formula, family) returns
# an object with $results (a data frame of pooled coefficients).

fit_h1_model <- function(outcome_var, scf_obj, family_type = "binomial") {

  base_rhs <- "origin_region + years_in_us + age_centered + educ_cat + currently_married + log_income"
  full_formula <- as.formula(paste(outcome_var, "~", base_rhs))

  fit <- tryCatch({
    if (family_type == "binomial") {
      scf_glm(scf_obj, full_formula, family = binomial())
    } else {
      scf_ols(scf_obj, full_formula)
    }
  }, error = function(e) {
    list(error = TRUE, message = conditionMessage(e))
  })

  if (is.list(fit) && isTRUE(fit$error)) {
    return(fit)
  }

  # Extract the pooled results
  results <- fit$results

  # For binomial models, add odds-ratio columns (matches Paper 1 pattern)
  if (family_type == "binomial" && "estimate" %in% names(results)) {
    results$odds_ratio <- exp(results$estimate)
    results$or_lower   <- exp(results$estimate - 1.96 * results$std.error)
    results$or_upper   <- exp(results$estimate + 1.96 * results$std.error)
  }

  results
}
```

```{r h1-parallel, results='asis'}
if (n_fb >= 100) {

  cat("### Parallel-integration outcomes\n\n")

  for (info in list(
    list(var = "own_retire", label = "Retirement Account Ownership"),
    list(var = "own_stocks", label = "Direct Stock Ownership")
  )) {

    cat(sprintf("#### %s\n\n", info$label))

    res <- fit_h1_model(info$var, scf_fb, "binomial")

    if (is.list(res) && isTRUE(res$error)) {
      cat(sprintf("*Model failed: %s*\n\n", res$message))
      next
    }

    origin_rows <- res %>% filter(grepl("^origin_region", term))

    if (nrow(origin_rows) == 0) {
      cat("*No origin-region coefficients in output.*\n\n")
      next
    }

    display <- origin_rows %>%
      mutate(across(c(estimate, std.error, odds_ratio,
                      or_lower, or_upper), ~round(., 3)),
             p.value = formatC(p.value, format = "e", digits = 2)) %>%
      select(term, estimate, std.error, odds_ratio,
             or_lower, or_upper, p.value)

    print(kable(display,
                col.names = c("Term", "Log-Odds", "SE",
                              "Odds Ratio", "OR Lower", "OR Upper", "p-value"),
                caption = sprintf("%s — Mexican origin = reference", info$label)))
    cat("\n\n")
  }
}
```

```{r h1-downward, results='asis'}
if (n_fb >= 100) {

  cat("### Downward-assimilation outcomes\n\n")

  for (info in list(
    list(var = "credit_denied", label = "Credit Denied or Got Less in Past Year"),
    list(var = "no_checking",   label = "No Checking Account")
  )) {

    cat(sprintf("#### %s\n\n", info$label))

    res <- fit_h1_model(info$var, scf_fb, "binomial")

    if (is.list(res) && isTRUE(res$error)) {
      cat(sprintf("*Model failed: %s*\n\n", res$message))
      next
    }

    origin_rows <- res %>% filter(grepl("^origin_region", term))

    if (nrow(origin_rows) == 0) {
      cat("*No origin-region coefficients in output.*\n\n")
      next
    }

    display <- origin_rows %>%
      mutate(across(c(estimate, std.error, odds_ratio,
                      or_lower, or_upper), ~round(., 3)),
             p.value = formatC(p.value, format = "e", digits = 2)) %>%
      select(term, estimate, std.error, odds_ratio,
             or_lower, or_upper, p.value)

    print(kable(display,
                col.names = c("Term", "Log-Odds", "SE",
                              "Odds Ratio", "OR Lower", "OR Upper", "p-value"),
                caption = sprintf("%s — Mexican origin = reference", info$label)))
    cat("\n\n")
  }
}
```

```{r h1-selective, results='asis'}
if (n_fb >= 100) {

  cat("### Selective-assimilation outcomes\n\n")

  cat("#### Homeownership\n\n")
  res_home <- fit_h1_model("own_home", scf_fb, "binomial")

  if (is.list(res_home) && isTRUE(res_home$error)) {
    cat(sprintf("*Model failed: %s*\n\n", res_home$message))
  } else {
    origin_rows <- res_home %>% filter(grepl("^origin_region", term))
    if (nrow(origin_rows) > 0) {
      display <- origin_rows %>%
        mutate(across(c(estimate, std.error, odds_ratio,
                        or_lower, or_upper), ~round(., 3)),
               p.value = formatC(p.value, format = "e", digits = 2)) %>%
        select(term, estimate, std.error, odds_ratio,
               or_lower, or_upper, p.value)
      print(kable(display,
                  col.names = c("Term", "Log-Odds", "SE",
                                "Odds Ratio", "OR Lower", "OR Upper", "p-value"),
                  caption = "Homeownership — Mexican origin = reference"))
      cat("\n\n")
    }
  }

  cat("#### Net-Worth Percentile (linear regression)\n\n")
  res_nw <- fit_h1_model("nw_pctl", scf_fb, "ols")

  if (is.list(res_nw) && isTRUE(res_nw$error)) {
    cat(sprintf("*Model failed: %s*\n\n", res_nw$message))
  } else {
    origin_rows <- res_nw %>% filter(grepl("^origin_region", term))
    if (nrow(origin_rows) > 0) {
      display <- origin_rows %>%
        mutate(across(c(estimate, std.error), ~round(., 3)),
               p.value = formatC(p.value, format = "e", digits = 2)) %>%
        select(term, estimate, std.error, p.value)
      print(kable(display,
                  col.names = c("Term", "Estimate", "SE", "p-value"),
                  caption = "Net-Worth Percentile — Mexican origin = reference"))
      cat("\n\n")
    }
  }
}
```

## H2: Pathway-Pattern Tests

The three pathways predict joint patterns across outcomes within origin region. The
pattern matrix below summarizes origin-region odds ratios across the seven outcomes from
the H1 specifications. Pathway-coherent patterns — for example, Mexican-origin households
showing low retirement and stock ownership combined with high homeownership — would
support the segmented-assimilation hypothesis.

```{r h2-pattern-matrix, results='asis'}
if (n_fb >= 100) {

  # Re-fit all H1 models and collect the origin-region coefficients into a matrix
  outcomes_for_h2 <- list(
    list(var = "own_retire",    label = "Retirement",     family = "binomial"),
    list(var = "own_stocks",    label = "Stocks",         family = "binomial"),
    list(var = "credit_denied", label = "Credit denied",  family = "binomial"),
    list(var = "no_checking",   label = "No checking",    family = "binomial"),
    list(var = "own_home",      label = "Homeowner",      family = "binomial"),
    list(var = "nw_pctl",       label = "NW pctile",      family = "ols")
  )

  # Initialize the pattern matrix
  origin_levels_no_ref <- c("Other Latin American/Caribbean",
                             "Asian",
                             "European/Other developed",
                             "Other")

  pattern_df <- data.frame(`Origin Region` = origin_levels_no_ref,
                           check.names = FALSE,
                           stringsAsFactors = FALSE)

  for (info in outcomes_for_h2) {
    res <- fit_h1_model(info$var, scf_fb, info$family)

    if (is.list(res) && isTRUE(res$error)) {
      pattern_df[[info$label]] <- "--"
      next
    }

    cell_values <- sapply(origin_levels_no_ref, function(lvl) {
      term_name <- paste0("origin_region", lvl)
      idx <- which(res$term == term_name)

      if (length(idx) == 0) return("--")

      est <- res$estimate[idx]
      pval <- res$p.value[idx]

      # For binomial: report odds ratio with stars
      # For OLS: report estimate with stars
      if (info$family == "binomial") {
        val <- exp(est)
        formatted <- sprintf("%.2f", val)
      } else {
        formatted <- sprintf("%.1f", est)
      }

      stars <- if (is.na(pval)) "" else
               if (pval < 0.001) "***" else
               if (pval < 0.01)  "**"  else
               if (pval < 0.05)  "*"   else ""

      paste0(formatted, stars)
    })

    pattern_df[[info$label]] <- cell_values
  }

  cat("**Table 3. Origin-region × outcome pattern matrix**\n\n")
  cat("Cells contain odds ratios (binary outcomes) or coefficient estimates ",
      "(net-worth percentile) relative to Mexican origin (reference). ",
      "Stars: \\* p<0.05, \\*\\* p<0.01, \\*\\*\\* p<0.001.\n\n", sep = "")
  print(kable(pattern_df, caption = "Origin-region effects across outcomes"))
  cat("\n\n")
}
```

## H3: Years-in-U.S. × Origin Interaction

Interaction models test whether duration-of-residence effects on financial outcomes
differ by origin region. A pure human-capital model predicts a single duration slope
across all groups; segmented assimilation predicts that some origin groups show
different (or no) convergence trajectories.

```{r h3-interaction, results='asis'}
if (n_fb >= 100) {

  # Fit interaction model for each outcome and report the interaction terms
  cat("**Table 4. Origin × Years-in-U.S. interaction tests**\n\n")
  cat("For each outcome, we report the joint significance of the four ",
      "Origin × Years-in-U.S. interaction terms. A significant joint test ",
      "indicates that duration-of-residence effects differ by origin region.\n\n")

  interaction_results <- data.frame(
    Outcome = character(),
    `Joint p-value` = character(),
    `Largest interaction OR/coef` = character(),
    check.names = FALSE,
    stringsAsFactors = FALSE
  )

  for (info in list(
    list(var = "own_retire",    label = "Retirement",     family = "binomial"),
    list(var = "own_stocks",    label = "Stocks",         family = "binomial"),
    list(var = "credit_denied", label = "Credit denied",  family = "binomial"),
    list(var = "no_checking",   label = "No checking",    family = "binomial"),
    list(var = "own_home",      label = "Homeowner",      family = "binomial"),
    list(var = "nw_pctl",       label = "NW pctile",      family = "ols")
  )) {

    int_rhs <- "origin_region * years_in_us + age_centered + educ_cat + currently_married + log_income"
    int_formula <- as.formula(paste(info$var, "~", int_rhs))

    fit <- tryCatch({
      if (info$family == "binomial") {
        scf_glm(scf_fb, int_formula, family = binomial())
      } else {
        scf_ols(scf_fb, int_formula)
      }
    }, error = function(e) NULL)

    if (is.null(fit)) {
      interaction_results <- rbind(interaction_results,
        data.frame(Outcome = info$label,
                   `Joint p-value` = "--",
                   `Largest interaction OR/coef` = "--",
                   check.names = FALSE))
      next
    }

    int_terms <- fit$results %>%
      filter(grepl("origin_region.*:years_in_us|years_in_us:origin_region", term))

    if (nrow(int_terms) == 0) {
      interaction_results <- rbind(interaction_results,
        data.frame(Outcome = info$label,
                   `Joint p-value` = "--",
                   `Largest interaction OR/coef` = "--",
                   check.names = FALSE))
      next
    }

    # Approximate joint p-value: minimum p-value across interaction terms
    # (Bonferroni-conservative; a proper Wald joint test would use vcov)
    min_p <- min(int_terms$p.value, na.rm = TRUE)
    largest_idx <- which.max(abs(int_terms$estimate))
    largest_est <- int_terms$estimate[largest_idx]

    if (info$family == "binomial") {
      largest_str <- sprintf("OR = %.3f", exp(largest_est))
    } else {
      largest_str <- sprintf("β = %.3f", largest_est)
    }

    interaction_results <- rbind(interaction_results,
      data.frame(Outcome = info$label,
                 `Joint p-value` = formatC(min_p, format = "e", digits = 2),
                 `Largest interaction OR/coef` = largest_str,
                 check.names = FALSE))
  }

  print(kable(interaction_results,
              caption = "Note: Joint p-value is the minimum p across the four interaction terms (conservative)."))
  cat("\n\n")
  cat("Significant interactions in any cell indicate divergent duration trajectories ",
      "by origin region — a key segmented-assimilation prediction.\n\n", sep = "")
}
```

# Robustness Checks

## Three-Category Origin Recode

If any cell in the 5-category origin coding falls below n = 30, we re-estimate models
with a collapsed 3-category coding (Latin American, Asian, European/Other) for stability.

## Race Versus Origin

Origin region and racial classification (`racecl4`) are correlated but not identical.
Specifications adding `race_cat` to the H1 models test whether origin-region effects are
robust to controlling for racial classification, and whether racial classification
mediates origin-region effects.

## Net-Worth Specification

The primary net-worth specification uses percentile within the full SCF distribution. An
alternative using `asinh(networth)` is reported as a sensitivity check, which preserves
sign for negative net-worth households.

# Discussion

The findings will be interpreted under three headings:

**Are origin-region differences durable?** A finding that origin-region coefficients
remain large and significant after controlling for years-in-U.S., age, education, and
income would support segmented-assimilation theory over a pure human-capital convergence
model. A finding that origin-region coefficients vanish when controls are added would
support the convergence model.

**Pathway-coherent or scattered patterns?** Segmented assimilation predicts coherent
within-group patterns. Mexican-origin households should show downward-assimilation
barriers (high credit denial, low retirement-account ownership) combined with
selective-assimilation asset accumulation (relatively high homeownership). Asian-origin
households should look like the parallel-integration profile (high mainstream
participation). Patterns that scatter across pathways without group coherence weaken the
segmented-assimilation interpretation.

**Implications for policy and theory.** Wealth and credit-access disparities have
generally been studied through the lens of race rather than origin. The 2022 SCF allows
the field to begin testing whether origin-region heterogeneity exists *within* racialized
groups — for example, whether Mexican-origin and Cuban-origin Hispanic households show
different financial patterns, a critical refinement that earlier surveys could not
support.

# Limitations

1. **Single-wave cross-section.** The 2022 SCF is the first wave with these immigration
questions. No time series is available; cohort and period effects cannot be separated.

2. **No parental nativity.** Unlike the NSFG used in Papers 3 and 4, the SCF does not
collect parental nativity. Generation-1 vs. generation-2 distinctions are not directly
observable.

3. **Origin-region collapsing.** Sample sizes preclude finer-grained country-of-origin
analysis. Heterogeneity within "Asian" or "Other Latin American" categories may obscure
pathway-specific patterns.

4. **Selection into U.S. residence.** Foreign-born households observed in 2022 are
survivors of immigration selection processes. Observed financial patterns reflect this
selection alongside assimilation processes.

5. **Disclosure-avoidance modifications.** SCF public-use data are altered for disclosure
protection, which may attenuate small-cell estimates for the smallest origin groups.

# Conclusion

This paper applies the segmented-assimilation framework to U.S. household financial
outcomes for the first time using nationally representative data with origin-region
detail. The analysis tests whether origin region matters for financial-system
incorporation above and beyond duration-of-residence and individual human-capital
characteristics, and whether the patterns observed are coherent with the three theoretical
pathways — parallel integration, downward assimilation, selective assimilation — or
scattered across them.

The contribution to the broader thesis is to close the loop on the immigration question.
Earlier papers established that nativity matters for fertility outcomes (Paper 3) and for
unplanned childlessness (Paper 4). Paper 5 returns to the SCF financial-outcome framework
that anchors Papers 1 and 2 and tests whether nativity — refined into origin region —
also structures the financial system itself.

# Appendix A: Variable List

| Construct | Source variable | Type |
|---|---|---|
| Foreign-born status | `X7437`–`X7443` (max non-zero) | Raw codebook |
| National origin (R) | `X7404` | Raw codebook |
| Race/ethnicity | `racecl4` | Bulletin extract |
| Age | `age` | Bulletin extract |
| Education | `edcl` | Bulletin extract |
| Marital status | `married` | Bulletin extract |
| Income | `income` | Bulletin extract |
| Net worth | `networth` | Bulletin extract |
| Retirement assets | `retqliq` | Bulletin extract |
| Stocks | `stocks` | Bulletin extract |
| Credit denied | `turndown` | Bulletin extract |
| Checking balance | `checking` | Bulletin extract |
| Homeowner | `housecl` | Bulletin extract |
| Weight | `wgt` | Bulletin extract |

# Appendix B: scf Package Workflow Notes

This document follows the workflow established in Paper 1 (`scf_ccbal_analysis_v5.Rmd`):

1. `scf_download(2022)` writes raw `.rds` files to `scf_data/`.
2. `scf_load(2022, data_directory = "scf_data")` returns an object with
   `$mi_design[[1..5]]` survey-design slots, one per implicate.
3. `scf_update(scf_obj, new_var = expr, ...)` adds variables to all 5 implicates.
4. `scf_mean()`, `scf_glm()`, `scf_ols()` fit pooled models with Rubin's-Rules pooling
   and replicate-weight-based variance.
5. Subsetting (e.g. to the foreign-born sample) is done by copying the object and
   `subset()`-ing each `mi_design[[i]]` slot manually.

The new 2022 immigration variables (`X7404`, `X7437`–`X7443`) are not in the Bulletin
summary extract that the `scf` package builds. The diagnostic chunk at the top of this
document attempts to merge them in from the raw 2022 SCF `.rds` file. If the merge
fails, those analyses cannot proceed.
